### 2024-10-16 BAI/exercises.tex
Added exercise scaffold.

\documentclass{article}

\usepackage[]{../.template/xrcise}

\subject{Bio-inspired Artificial Intelligence}
\semester{Winter 2024}
\author{Leopold Lemmermann}

\begin{document}\createtitle

\section[2024]{Mock exam}

% TODO

\section[2023]{1st exam}
% First questions were multiple choice

% Function to compute neighbourhood in SOM
% Meaning of grounded language
% Ventriloquism Effect

% Followed by some general questions

% multisensory integration of different strength stimuli (slide example)
% What is the cocktailparty effect

% Spiking neural networks

% label leaky and integrate+fire models
% Similarity angle, calculate for a weight and input vector
% Compute output from perceptron (similar to slide example)
% Visual

% what do the different layers on CNN learn (not every single layer but a brief overview)
% Compute filtering in CNN layer (5x5 matrix with 3x3 filter, stride=1, no padding)
% Draw orientation and firing rate diagram from slides
% Audio

% ITD and ILD: explain and what to use for which frequencies
% Behavior

% braitenberg vehicles
% Advantage of behavior based

% Evolutionary Computing

% Cycle Graph also arrow labels
% Example: find the shortest distance between cities only visit each city once
% Recombination and Mutation algorithm for example
% Optimisation strategies for this algorithm

\section[2022]{1st exam}
% Nr.1) Multiple choice: (one or multiple correct, you only get points if all ticks are completely correct, the last possibility of four was always "none of the above")

% a)

% ...

% b) In which Models is Backpropagation used for learning?

% MLP

% Convolutional Neural Network

% SOM

% None of the above

% c) what does "embodied" means for language?

% ...

% d)

% ...

% e)

% ...

% Nr.2) Attention

% 2.1) Name and explain three ways to measure the attention of a human.

% 2.2) Describe two models to measure or visualize human attention.

% 2.3) Name advantages and disadvantages of the two models you proposed.

% Nr.3) Vision

% 3.1) Name the two streams and their task that are present in the human vision.

% 3.2) Explain rods and cones

% 3.3) Explain the similarity between a convolutional neural network with convolution and max pooling layers and the visual cortex (not sure if visual cortex or some other brain area)

% 3.4) Visualize the firing rate of a simple cell for a line of different orientation in a diagram with the orientation on the x-axis and the firing rate on the y-axis

% 3.5) Given a 5x5 matrix and a 3x3 filter compute the convolution (no padding, stride = 1)

% Nr.4) Sound Localization

% 4.1) compute crosscorrelation steps on two signals until maximum crosscorrelation value is reached

% 4.2) name and explain the two cues used to perform sound localization. Which one works best for high/low frequencies?

% ...

% Nr.5) Behaviour

% 5.1) Explain a Braitenberg Vehicle with an example and name the behaviour of your example vehicle.

% 5.2) Flow chart of functional decomposition of the classical approach for a robot controller was given with steps left blank: (sensors => blank => blank => blank => blank => actuators)

% ...

% Nr.6) Continual Learning

% 6.1) Explain the stability-plasticity dilemma

% 6.2) Explain catastrophic forgetting and a model where this effect could arise

% 6.3) Name and describe two strategies of continual learning

% Nr.7) Language processing

% 7.1) Name the two streams important in language processing and their tasks

% 7.2) Describe CBOW (Continuous Bag-of-word) and Skip-gram

% 7.3) Describe difference between the earlier Wernickes model and the later Pulvermüllers model (question gave a few more details which I don't remember anymore)

% Nr.8) Computational Neural Networks

% 8.1) Explain 3 steps of Backpropagation algorithm and explain termination criteria

% 8.2) Name and shortly describe the three fundamental ANN Learning Paradigms

% 8.3) Explain how the learning rate parameter could be chosen

\end{document}

### 2024-10-16 PbD/exercises.tex
Added exercise scaffold.

\documentclass{article}

\usepackage[]{../.template/xrcise}

\subject{Privacy by Design}
\semester{Winter 2024}
\author{Leopold Lemmermann}

\begin{document}\createtitle

\section{Catalogue}

% TODO

\section[2024]{Mock exam}

% TODO

\section[2020]{1st exam}
% 1. Broadcast
% • Wer wird geschützt und welches Schutzziel wird verfolgt?
% • Nenne zwei Anwendungsbeispiele (abgesehen von Zeitung/TV/Radio)
% • Professorin möchte über eine Zeitschrift auf Feedback von Studierenden Antworten und die Identität dieser
% schützen. Anweisungen ähnlich wie: Neben dem Feedback soll eine Zufallszahl zwischen 0 und 100 angegeben
% werden. Antworten werden im Bezug auf dieses Pseudonym gegeben. Welche der beiden Arten der impliziten
% Adressierung wurde hier genutzt? Was sind mögliche Probleme in Bezug auf speziell diese Auswahl an
% Pseudonymen?
% • Der Professorin gefällt nicht, dass die Antworten auf das Feedback von jedem gelesen werden können (und
% nicht nur von der entsprechenden Person). Wir könnte das Problem gelöst werden? Schreibe einen Text mit
% Anweisungen an die Studierenden und Begründe die Notwendigkeit der einzelnen Komponenten.
% 2. IPv6-Präfixe
% Zur Einleitung wurden Präfix-Sharing, -Hopping und -Boquets beschrieben
% • Warum genügt es nicht verschiedene Adressen unter einem Präfix zu verwenden?
% • Was sind die Schwächen von Präfix-Sharing bei einem gleich bleibendem Präfix und einer kleinen Nutzergruppe?
% • Was sind die Schwächen von Präfix-Boquets, wenn unterschiedliche Präfixe für verschiedene Ziele aber nicht
% pro Verbindung genutzt werden?
% • Was ist das Problem beim Präfix-Hopping, wenn sich während einer bestehenden Verbindung die IP-Adresse
% ändert?
% • Wie kann man dieses Problem beim Präfix-Hopping lösen?
% • Wie kann man Benutzer abgesehen von der Adressierung tracken? Nenne 3 Beispiele.
% 3. Mixe
% • Angreifermodell angeben
% • Wie kann ein Mix-Netz von einem starken Angreifer (z.B. Geheimdienst) angegriﬀen werden? Nenne 2
% Beispiele.
% • Es wird ein Mix-Netz angeboten von nur einem Betreiber, welcher damit wirbt, dass 20 Mixe hinter einander
% verwendet werden und für höhere Performance nur 10 Teilnehmer pro Mix zugeordnet werden. Bewerte die
% Sicherheit des vorgestellten Produkts mit Begründung.
% 4. DC-Netze
% Gegeben waren 4 lokale Summen (A, B, C, D), ein Schlüsselgraph kAB , kBC , kCD und der Schlüssel kBC . Wir sind
% Teilnehmer B und haben eine Leernachricht gesendet.
% • Berechne die gesendete Nachricht N . Skizziere deinen Rechenweg.
% • Grenze ein wer die Nachricht gesendet haben kann. Skizziere deinen Rechenweg.
% • Was muss passieren, damit B weniger Aussagen treﬀen kann?

\end{document}

### 2024-10-16 SbD/exercises.tex
Added exercise scaffold.

\documentclass{article}

\usepackage[]{../.template/xrcise}

\subject{Security by Design}
\semester{Winter 2024}
\author{Leopold Lemmermann}

\begin{document}\createtitle

\section{Catalogue}

% TODO

\section[2024]{Mock exam}

% TODO

\section[2023]{2nd exam}
% Aufgabe 1
% Betrachten Sie die die Betriebsarten für Blockchiffren ECB, CBC und CTR.

% a) Bei welchen Arten braucht man eine invertierbare Verschlüsselungsfunktion, bei welchen nicht? Begründen Sie?

% b) Bei welchen Arten lässt sich die Verschlüsselung parallelisieren?

% c) Im zweiten Block gibt es ein Bit mit einem additiven Fehler. Was kann noch entschlüsselt werden?

% d) Ein Bit fehlt im zweiten Block (nicht bekannt, wo das Bit fehlt). Was kann noch entschlüsselt werden?

% e) Ein ganzer Block fehlt. Was kann noch entschlüsselt werden?

% f) Zwei verschiedene Klartexte wurden mit CTR mit dem gleichen IV verschlüsselt. Zeigen Sie, dass ein Angreifer den zweiten Klartext entschlüsseln kann wenn dieser den ersten Klartext und die beiden Schlüsseltexte hat.

% Aufgabe 2 ECC
%  a) Auf welcher mathematischen Schwierigkeit basiert der ECC?

%  b) Berrechne die Punkte auf der folgenden Eliptischen Kurve: a = 6, b=1, p=11 (allgemeine Formel war gegeben)

%  c) Nennen Sie zwei Vorteile die ECC hat, gegenüber sicherheitsmäßig vergleichbaren Verfahren

% Aufgabe 3
% Das Zertifizierungsmodell war transitiv anzunehmen.

% a) Die Begriffe Signatur und Zertifikat erklären und voneinander abgrenzen

% b) N7 möchte N3 Nachricht schicken, welchen CAs muss N7 vertrauen?

% c) N7 möchte zusätzlich N5 schicken, welchen CAs muss N7 vertrauen, wähle so, dass die Zahl and CAs minimal ist.

% d) CA2 fällt aus, wie können wir die Kommunikation aus b) trotzdem realisieren (Was & Wie). Was können wir für N1 tun?

% e) Trust on First Use: Keine CAs, Schlüssel wird bei erster Kommunikation an Kommunikationspartner geschickt, sobald und bei späteren Kommunikationen überprüft. Wenn Kommunikation und es gibt schon anderen Schlüssel, deny. 2 Vorteile, 2 Nachteile nenne und erklären.

% Aufgabe 4
% (Aufgabe war sehr ähnlich zu einer Übungsaufgabe)

% Zur Erreichung von Forward Secrecy erzeugt Alice auf ihrem Server mit dem RSA-Verfahren einen privaten Schlüssel d sowie den dazu passenden öffentlichen Schlüssel c, den sie persönlich an ihre Nutzerinnen und Nutzer verteilt. Zusätzlich erzeugt der Server von Alice bei jedem Verbindungsaufbau ein neues kurzlebiges (engl. ephemeral) RSA-Schlüsselpaar, das aus einem öffentlichen Schlüssel ce und einem geheimen Schlüssel de besteht. Der Server signiert mit dem privaten Schlüssel d den neuen öffentlichen Schlüssel ce und sendet diesen zusammen mit der Signatur über die aufgebaute Verbindung an die Kundin. Die Kundin prüft anschließend die Signatur mithilfe des zuvor von Alice erhaltenen öffentlichen Schlüssels c.

% Ist die Signatur korrekt, nutzt die Kundin ce, um einen neuen, zufällig generierten symmetrischen AES-Schlüssel k zu verschlüsseln. Die weitere Kommunikation in dieser Verbindung wird symmetrisch mit k verschlüsselt. Nach Verbindungsende löscht Alices Server de, ce und k.

% a) Welche Verbindungen könnten von einem sehr starken Angreifer wie einem Nachrichtendienst noch entschlüsselt werden, wenn Zugriff auf den Server bekommt? Begründen Sie Ihre Antwort.

% b) Skizzieren Sie einen Angriff, der möglich wäre, wenn die Kundin die Signatur nicht prüft. Erläutern SIe die einzelnen nötigen Schritte.

% c) Alice ist das Verfahren zu langsam. Sie nimmt nun folgende Änderung vor: Statt jedes mal bei der Erzeugung von ce und de neue Primzahlen p und q für den Modulus Ne zu generieren, verwendet sie immer die gleichen Primzahlen. Nun wählt sie lediglich den öffentlichen Exponenten ce neu und berechnet einen dazugehörigen geheimen Exponenten de. Warum ist dies keine gute Idee? Zeigen Sie, wie sich dieses geänderte Verfahren angreifen lässt, um aufgezeichnete Verbindungen zu entschlüsseln.

\section[2022]{1st exam}
% Aufgabe 1 - Klassische Chiffren (11 Punkte)

% (3 Punkte) Erläutern Sie zwei Ursachen, die dafür sorgen, dass fast alle kryptographischen Algorithmen mit der Zeit nicht mehr Standard sind.

% (3 Punkte) Sie möchten einen Text so verschlüsseln, dass er auch in 98 Jahren nicht ohne Kenntnis des von Ihnen versteckten Schlüssels entschlüsselt werden kann. Welches Verfahren wählen Sie dazu? Begründen Sie Ihre Entscheidung. Warum wird das Verfahren heutzutage nicht überall benutzt?

% (5 Punkte) Gegeben wurde eine Chiffre, welche Buchstaben A-N mit Buchstaben N-Z vertauscht. Und gleichzeitig muss die Länge der Nachricht immer gleich sein. Bewerten Sie die Sicherheit der Chiffre und die beiden Eigenschaften (Polyalphabetische Verschlüsselung nur mit Vertauschen und die gleiche Länge der Nachricht).

% Aufgabe 2 - Elliptische Kurven (13 Punkte)

% (8 Punkte) Berechnen Sie für die folgenden Parameter alle Punkte auf der Kurve: a = 2, b = 3, p = 7

% (2 Punkte) Elgamal mit elliptischen Kurven wurde vorgestellt. Zeigen Sie, warum dieses Verfahren mit beliebigen Parametern funktioniert.

% (3 Punkte) Vergleichen Sie den Elgamal aus der VL mit dem gegebenen. Auf welchem Prinzip beruht die Sicherheit? Denken Sie, dass er sicherer ist? Warum?

% Aufgabe 3 - Public Key Infrastructure (8 Punkte)

% (3 Punkte) Es wurden 3 verschiedene Möglichkeiten gegeben, einen Schlüssel von verschiedenen Schlüsselpartner (S, V, C, T) zu kombinieren.

% Alle Teilschlüssel werden aneinander gehängt l/4.

% 3 der Teilschlüssel wird aneinander gehängt und der letzte Teilnehmer bestimmt die Reihenfolge.

% Alle Schlüssel werden XOR berechnet.

% Bewerten Sie die Sicherheit der Verfahren.

% (2 Punkte) Gegeben ist ein Zertifikatsbaum mit 5 Teilnehmern.

% E-> D

% E-> B

% E-> A

% E-> C

% C-> B

% A-> B

% G -> Null

% Betrachten Sie die Zertifizierung Relation. G möchte jetzt zusätzlich eine verschlüsselte Nachricht an B senden. Welcher Zertifizierungsstelle bzw. welchen Zertifizierungsstellen muss G mindestens vertrauen, um das Zertifikat von B prüfen zu können?

% (1 Punkt) Wählen Sie Ihre Lösung so, dass die Anzahl an Zertifizierungsstellen, denen N4 (explizit) vertrauen muss, minimal ist.

% (2 Punkte) In der Zertifizierung Relation aus Teilaufgabe c) wird die Zertifizierungsstelle C kompromittiert und fällt deshalb aus. G möchte eine verschlüsselte Nachricht an C senden. Welche Maßnahme ermöglicht es G, das Zertifikat von B trotz des Ausfalls von CA4 zu überprüfen?

% Aufgabe 4 - GSM (8 Punkte)

% (7 Punkte) Sagen Sie zu folgenden Abkürzungen welche Funktion sie im GSM erfüllen:

% Ki, IMSI, RAND, SRES, LAI, MSISDN, Kc

% (1 Punkt) Welche dieser Abkürzungen ist auch noch im UMTS enthalten bzw. mit ähnlicher Funktion aber anderen Namen?

\end{document}

### 2024-10-18 NLP/exercises.tex
Added NLP exercise tex file.

\documentclass{article}

\usepackage[]{../.template/xrcise}

\subject{Natural Language Processsing}
\semester{Winter 2024}
\author{Leopold Lemmermann}

\begin{document}\createtitle

\sheet[2024]{1st exam}
% 1 Question: Foundations (30 Points)
% This section has various tasks to assess your understanding of fundamental concepts, principles and
% terminology related to requirements engineering and software architecture.
% 1.1
% (3 Points) Name one activity in requirements engineering and provide a brief explanation.
% 1.2
% You have learned about three types of qualtity scenarios including use-case, growth and exploratory
% scenarios. Explain each of them by an example: Use-case scenarios, Growth scenarios, Exploratory scenarios.

% 1.3 Give three reasons why software architecture is important. Explain each in one sentence.

% 1.4 Define the term architectural tactic in the context of software architecture. In addition
% provide an example of an achtitecture tactic and explain how it can be used to address a specific
% quality attribute in a software system.

% 1.5
% (3 Points) Which of the following pairs of qualities are typically in conflict with each other (i.e. are
% trade-offs to each other)? Explain each decision in one sentence.
% Modularity - Time behavior
% ⃝ typically in conflict
% ⃝ typically not in conflict
% Explanation:
% Usability - Security
% ⃝ typically in conflict
% ⃝ typically not in conflict
% Explanation:
% 3
% 1.6
% (3 Points) What is the purpose of conformance/compliance checking? In addition, briefly explain
% strucural and behavioral conformance checking.
% 1.7
% (2 Points) Explain the term "lift & shift" in the context of cloud development.
% 1.8
% (4 Points) Decide for each of the following statements wether it is true or false. Justify your decision
% in one sentence.
% If your system consists of a set of microservices, there is no need for a central architecture document
% since each service is fre to choose its technologies.
% ⃝ True
% ⃝ False
% Explanation:
% Celebrating release parties is a sign of great architecture.
% ⃝ True
% ⃝ False

% 2 Question: Reqirements Engineering (15 Points)
% The GreenHome Energy Management System is an innovative solution designed to empower (1) home-
% ownerstoharnessandoptimzegreenenergysourceswithintheirresidences, (2)energyserviceproviders
% to improve service customization, e.g. flexible pricing based on user behavior and offer energy-saving
% solutions, and (3) renewable energy solution providers to invest in research and development of renew-
% able energy technologies. To promote sustainability, reduce carbon footprint, and maximize energy
% efficiency, this comprehensive software system integrates seamlessly with (1) smrt home devices such
% as smart thermostats, lighting systems, and appliances and (2) renewable energy sources, including
% solar panels, and energy storage systems.
% 2.1
% Solve the following tasks regarding the stakeholder analysis.
% i. (4 Points) Write the name of each axis after each = sign. In addition, classify three stakeholders
% derived from the text according to the following matrix.
% Figure 1: Stakeholder matrix
% ii. (2 Points) Classify two more stakeholders that are not explicitly mentioned in the text. Add
% them to the matrix in Figure 1. Show which ones are the two additional stakeholders by drawing
% a circle around the two names.
% iii. (3 Points) Briefly explain the reason for your classification for the two additional stakeholders.
% 5
% 2.2
% Consider the following scenario, which is related to the GreenHome Energy Management System
% requirements:
% • "The Cost analysis in the GreenHome Energy Management System addresses the need for home-
% owners to observe their energy-rekated expenses, which involve providing users witha straight-
% forward breakdown of their energy costs, including expenses associated with renewable energy
% production."
% • "The system provides advanced analytics to forecast future energy costs based on usage patterns
% and external factors."
% • "It would be cool to have a gamified elemet eithin the Cost calculation. The idea is to award
% users GreenPoints for implementing energy saving measures, achieving specific sustainability
% milestones."
% Regarding the Kano model, solve the following tasks.
% i. (2 Points) Derive one basic requirement from the given scenario.
% ii. (2 Points) Derive one performance requirement from the given scenario.
% iii. (2 Points) Derive one excitement requirement from the given scenario.

% 3 Question: Architecture Modeling (23 Points)
% Figure 2 is part of an architectural description of an image hosting application. The application has
% two keey functionalities: (1) the ability to upload (write) an image to the serer, and (2) the ability to
% query for an image. Make sure to read and understand Figure 2.
% Figure 2: Architectural description of an image hosting application
% The team responsible for the image hosting application project decided to use 4+1, which is a
% template to structure the documentation of a software architecture. Answer the following questions
% and help the team to create the documentation of the architecture.
% 3.1
% Qestions regarding Figure 2:
% i. (1 Point) What Architectural Description Language (ADL) is used by the team?
% ii. (2 Points) What view does the architectural description show accordingto the 4+1 template?
% Briefly explain why.
% 7
% 3.2
% (10 Points) Process view:
% Consider the following scenario. "A user calls getImages through the IDownloadImage interface of
% WebUI component to download images. WebUI calls a method with the same name getImages from
% the IRequestHandler interface. Furthermore, the RequestHandler component calls the zip method from
% the ReEncoder component to get a zip package of all images. The ReEncoder component calls the
% imagesList method of the ImageAccess component. Finally, the ImageAccess component returns a
% result if it is already cashed. Otherwise, it returns NULL."
% Specify the intercomponent behavior through a UML sequence diagram for the given scenario.
% 8
% 3.3
% (10 Points)Deployment view:
% Use the UML notation to specify that (1) the WebUI component is deployed on aFrontEnd server,
% (2) the RequestHandler, ReEncoder and IMage Acces components are deployed to the Application-
% Server, (3) the Database and FileStorage components are deployed to the DatabaseServer and File-
% Server respectively. All the physical servers ar interconnected via LAN.

% 4 Question: Architecture Evaluation (22 Points)
% Jitsi Meet is a video-conferencing solution. The main service that Jitsi offers is creating video confer-
% ences and allowing participants across the Internet to join the conference. Three components are the
% core of the Jitsi system:
% • JitsiWeb provides a web-based user interface to access the Jitsi functionality.
% • JiCofo provides functionality for managing conferences.
% • VideoBridge is the critical component that is responsible for routing the video stream between
% multiple participants of a conference.
% Figure 3 shows how users attend different conferences using the Jitsi system. User1 and User2
% (depicted with a black box around user icon) joined Conference 1 (depicted with a black box around
% the text), and User 3 and User 4 joined Conference 2. The JiCofo component communicates with
% the VideoBridge component through XMPP, which is a communication protocol. Each server can
% have redundancies to support more conferences simultaneously. It is important to note that when a
% conference fails, e.g. Conference 1, a new conference has to be created. There is no redundancy and
% failover for a particular conference.
% Figure 3: Jitsi video conference system
% 10
% 4.1
% Consider the Architecture Tradeoff Analysis Method (ATAM).
% i. Provide a scenario for the Performance quality attribute such that the system maintains its
% performance when the number of users accessing the system increases.
% a) (1 Point) Source
% b) (1 Point) Stimulus
% c) (1 Point) Artifact
% d) (1 Point) Environment
% e) (1 Point) Response
% f) (1 Point) Response measure
% ii. Provide a scenario for the Modifiablity quality attribute.
% a) (1 Point) Source
% b) (1 Point) Stimulus
% c) (1 Point) Artifact
% d) (1 Point) Environment
% e) (1 Point) Response
% f) (1 Point) Response measure
% 11
% 4.2
% ATradeoff point isasystempropertythataffectsmorethanonequalityattribute. Forexample, adding
% more bridge servers affects more than one quality attribute in the system. Therefore, the number of
% bridge servers is a tradeoff point.
% i. (1 Point) Name at least two quality attributes of the system that the number of bridge servers
% affects.
% ii. (2 Points) Explain in one statement how each quality is affected by a change in the number of
% bridge servers.
% 4.3
% The current deployment strategy of Jitsi Meet system is visible in Figure 3. The engineering team
% deploys JitsiWeb and JiCofo service instances into one physical server. The team argues efficient
% resource usage is the main reason for their decision.
% i. (1 Point) Name one drawback for the current deployment strategy.
% ii. (2 Points) Suggest a new deployment strategy and explain the reason behind your decision.
% iii. (1 Point) Name one drawback of the new deployment strategy.
% 12
% 4.4
% (3 Points) Performance engineers of the JitsiMeet system claim that their system is efficient in terms
% of CPU usage. To prove their calim, they experimented on a server machine, which runs all the services
% needed by Jitsi Meet including the JitsiWeb, JiCofo and VideoBridge.
% The engineering team’s big challenge was to send enough traffic to put the system under load, ehich
% requires a considerable number of machines. What would you suggest to the engineering team as an
% alternative approach for quantitative evaluation of the Jitsi Meet? Briefly explain what your approach
% is.

\sheet[2024]{2nd exam}

% 1 Foundation (30 P.)
% 1. What does vertical traceability of requirements mean?
% 2. Name the layers of cloud computing and explain vendor lock-in.
% 3. What is a Domain Model? Explain using an example.
% 4. Consider architectural styles and patterns.
% a) Name two similarities between architectural styles and patterns.
% b) Name two differences between architectural styles and patterns.
% c) How do styles and patterns relate to each other?
% 5. Give two advantages and two disadvantages of semi-/formal vs. informal architec-
% ture documentation. (4 P.)
% 1
% 2 Requirements Engineering (12 P.)
% Here was a text given describing the rough proposed functionality of a music streaming
% service (Melody).
% 1. Classify four stakeholders according to the following matrix. Write the name of
% each axis after the = sign. Now there was the empty stakeholder matrix given, the
% axes named “x =” and “y =”. (5 P.)
% 2. Briefly explain your classification of two of the stakeholders. (1 P.)
% 3. Consider the Kano model.
% a) Describe one basic feature of the system and say, why it is one.
% b) Describe one performance feature of the system and say, why it is one.
% c) Describe one excitement feature of the system and say, why it is one.
% 3 Architecture Modeling (ca. 24 P.)
% Here was a text given describing the rough functionality of a new IDE (Power IDE ).
% 1. Which architectural style was used for the system? Give a reason why.
% 2. Which quality attribute does this style match and why?
% 3. Single choice question (yes/no): Is there usually a trade-off between the quality
% attributes modifiability and performance? Explain your answer briefly.
% 4. Single choice question (yes/no): Is there usually a trade-off between reusability
% and interoperability? Explain your answer briefly.
% 5. Use UML to model the presented architecture. (10 P.)
% 6. (Here a text was given describing a certain starting procedure.) Give a UML
% sequence diagram for the presented process. (8 P.)
% 2
% 4 Evaluation (ca. 24 P.)
% Here a text and a UML deployment diagram were given describing the rough functional-
% ity of an online order service (Food2Go) consisting of the three components OrderService,
% RestaurantService and AccountingService where OrderService had a connection to
% both of the other services but these had no connection to each other.
% 1. Consider the Architecture Tradeoff Analysis Method (ATAM).
% a) Provide a scenario for the Availability quality attribute: Source, Stimulus,
% Artifact, Environment, Response, Response measure (one text field each).
% b) Provide a scenario for the Modifiability quality attribute such that deploy-
% ment has finished after a maximum of one day: Source, Stimulus, Artifact,
% Environment, Response, Response measure (one text field each).
% 2. A design alternative (design 2) is to add a cache to OrderService.
% a) How does this alternative affect availability?
% b) Describe a scenario in which a client could experience data inconsistency.
% 3. Another design alternative (design 3) is to add third party messaging service.
% a) Would that solve the problem of data inconsistency?
% b) Does design 3 improve availability?
% c) Name a disadvantage of design 3 compared to design 2.

\end{document}

### 2024-10-18 BAI/templates/beamer_slides/WTM_latex.tex
Added templates.

\documentclass[14pt]{beamer}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Meta informations:
\newcommand{\trauthor}{Horst Hansen}
\newcommand{\trtype}{} %{Proseminar} %{Seminar} %{Workshop}
\newcommand{\trcourse}{Knowledge Processing with Neural Networks}
\newcommand{\trtitle}{Neural Networks for Artificial Agents}
\newcommand{\trmatrikelnummer}{6543210}
\newcommand{\tremail}{hansen@informatik.uni-hamburg.de}
\newcommand{\trinstitute}{Dept. Informatik -- Knowledge Technology, WTM}
\newcommand{\trwebsiteordate}{{http://www.informatik.uni-hamburg.de/WTM/}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Languages:

% Falls die Ausarbeitung in Deutsch erfolgt:
% \usepackage[german]{babel}
% \usepackage[T1]{fontenc}
% \usepackage[latin1]{inputenc}
% \usepackage[latin9]{inputenc}	 				
% \selectlanguage{german}

% If the thesis is written in English:
\usepackage[english]{babel} 						
\selectlanguage{english}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Bind packages:
\usepackage{beamerthemesplit}
\usetheme{Boadilla}
%\usetheme{Copenhagen}
%\usetheme{Darmstadt}
%\usetheme{Frankfurt}
%\usetheme{Ilmenau}
%\usetheme{JuanLesPins}
%\usetheme{Madrid}
%\usetheme{Warsaw }
%\usecolortheme{dolphin}
%\setbeamertemplate{sections/subsections in toc}[sections numbered]
%\beamertemplatenavigationsymbolsempty
%\setbeamertemplate{headline}[default] 	% deaktiviert die Kopfzeile
\setbeamertemplate{navigation symbols}{}% deaktiviert Navigationssymbole
%\useinnertheme{rounded}

\usepackage{acronym}                    % Acronyms
\usepackage{algorithmic}								% Algorithms and Pseudocode
\usepackage{algorithm}									% Algorithms and Pseudocode
\usepackage{amsfonts}                   % AMS Math Packet (Fonts)
\usepackage{amsmath}                    % AMS Math Packet
\usepackage{amssymb}                    % Additional mathematical symbols
\usepackage{amsthm}
\usepackage{color}                      % Enables defining of colors via \definecolor
\usepackage{fancybox}                   % Gleichungen einrahmen
\usepackage{fancyhdr}										% Paket zur schickeren der Gestaltung der 
\usepackage{graphicx}                   % Inclusion of graphics
%\usepackage{latexsym}                  % Special symbols
\usepackage{longtable}									% Allow tables over several parges
\usepackage{listings}                   % Nicer source code listings
\usepackage{lmodern}
\usepackage{multicol}										% Content of a table over several columns
\usepackage{multirow}										% Content of a table over several rows
\usepackage{rotating}										% Alows to rotate text and objects
\usepackage[section]{placeins}          % Ermoeglich \Floatbarrier fuer Gleitobj. 
\usepackage[hang]{subfigure}            % Allows to use multiple (partial) figures in a fig
%\usepackage[font=footnotesize,labelfont=rm]{subfig}	% Pictures in a floating environment
\usepackage{tabularx}										% Tables with fixed width but variable rows
\usepackage{url,xspace,boxedminipage}   % Accurate display of URLs

\definecolor{uhhRed}{RGB}{254,0,0}		  % Official Uni Hamburg Red
\definecolor{uhhGrey}{RGB}{136,136,136} % Official Uni Hamburg Grey
\definecolor{uhhLightGrey}{RGB}{180,180,180} % Official Uni Hamburg LightGrey
\definecolor{uhhLightLightGrey}{RGB}{220,220,220} % Official Uni Hamburg LightLightGrey
\setbeamertemplate{itemize items}[ball]
\setbeamercolor{title}{fg=uhhRed,bg=white}
\setbeamercolor{title in head/foot}{bg=uhhRed}
\setbeamercolor{block title}{bg=uhhGrey,fg=white}
\setbeamercolor{block body}{bg=uhhLightLightGrey,fg=black}
\setbeamercolor{section in head/foot}{bg=black}
\setbeamercolor{frametitle}{bg=white,fg=uhhRed}
\setbeamercolor{author in head/foot}{bg=black,fg=white}
\setbeamercolor{author in footline}{bg=white,fg=black}
\setbeamercolor*{item}{fg=uhhRed}
\setbeamercolor*{section in toc}{fg=black}
\setbeamercolor*{separation line}{bg=black}
\setbeamerfont*{author in footline}{size=\scriptsize,series=\mdseries}
\setbeamerfont*{institute}{size=\footnotesize}

\newcommand{\opticalseperator}{0.0025\paperwidth}

\institute{Universit\"at Hamburg\\\trinstitute}
\title{\trtitle}
\subtitle{\trtype}
\author{\trauthor}
\date{}
\logo{}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Configurationen:
%\hypersetup{pdfpagemode=FullScreen}

\hyphenation{whe-ther} 									% Manually use: "\-" in a word: Staats\-ver\-trag

%\lstloadlanguages{C}                   % Set the default language for listings
\DeclareGraphicsExtensions{.pdf,.svg,.jpg,.png,.eps} % first try pdf, then eps, png and jpg
\graphicspath{{./src/}} 								% Path to a folder where all pictures are located

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Costom Definitions:
\setbeamertemplate{title page}
{
  \vbox{}
	\vspace{0.4cm}
  \begin{centering}
    \begin{beamercolorbox}[sep=8pt,center,colsep=-4bp]{title}
      \usebeamerfont{title}\inserttitle\par%
      \ifx\insertsubtitle\@empty%
      \else%
        \vskip0.20em%
        {\usebeamerfont{subtitle}\usebeamercolor[fg]{subtitle}\insertsubtitle\par}%
      \fi%     
    \end{beamercolorbox}%
		\vskip0.4em
    \begin{beamercolorbox}[sep=8pt,center,colsep=-4bp,rounded=true,shadow=true]{author}
      \usebeamerfont{author}\insertauthor \\ \insertinstitute
    \end{beamercolorbox}

	  \vfill
	  \begin{beamercolorbox}[ht=8ex,center]{}
		  \includegraphics[width=0.20\paperwidth]{wtmIcon.pdf}
	  \end{beamercolorbox}%
    \begin{beamercolorbox}[sep=8pt,center,colsep=-4bp,rounded=true,shadow=true]{institute}
      \usebeamerfont{institute}\trwebsiteordate
    \end{beamercolorbox}
		\vspace{-0.1cm}
  \end{centering}
}

\setbeamertemplate{frametitle}
{
\begin{beamercolorbox}[wd=\paperwidth,ht=3.8ex,dp=1.2ex,leftskip=0pt,rightskip=4.0ex]{frametitle}%
		\usebeamerfont*{frametitle}\centerline{\insertframetitle}
	\end{beamercolorbox}
	\vspace{0.0cm}
}

\setbeamertemplate{footline}
{
  \leavevmode
	\vspace{-0.05cm}
  \hbox{
	  \begin{beamercolorbox}[wd=.32\paperwidth,ht=4.8ex,dp=2.7ex,center]{author in footline}
	    \hspace*{2ex}\usebeamerfont*{author in footline}\trauthor
	  \end{beamercolorbox}%
	  \begin{beamercolorbox}[wd=.60\paperwidth,ht=4.8ex,dp=2.7ex,center]{author in footline}
	    \usebeamerfont*{author in footline}\trtitle
	  \end{beamercolorbox}%
	  \begin{beamercolorbox}[wd=.07\paperwidth,ht=4.8ex,dp=2.7ex,center]{author in footline}
	    \usebeamerfont*{author in footline}\insertframenumber{}
	  \end{beamercolorbox}
  }
	\vspace{0.15cm}
}
\renewcommand{\normalsize}{\fontsize{13.5pt}{13.5pt}\selectfont}
\renewcommand{\large}{\fontsize{15.8pt}{15.8pt}\selectfont}
\renewcommand{\Large}{\fontsize{19.0pt}{19.0pt}\selectfont}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Additional 'theorem' and 'definition' blocks:
\newtheorem{axiom}{Axiom}[section] 	
%\newtheorem{axiom}{Fakt}[section]			% Wenn in Deutsch geschrieben wird.
%Usage:%\begin{axiom}[optional description]%Main part%\end{fakt}

%Additional types of axioms:
\newtheorem{observation}[axiom]{Observation}

%Additional types of definitions:
\theoremstyle{remark}
%\newtheorem{remark}[section]{Bemerkung} % Wenn in Deutsch geschrieben wird.
\newtheorem{remark}[section]{Remark}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Provides TODOs within the margin:
\newcommand{\TODO}[1]{\marginpar{\emph{\small{{\bf TODO: } #1}}}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Abbreviations and mathematical symbols
\newcommand{\modd}{\text{ mod }}
\newcommand{\RS}{\mathbb{R}}
\newcommand{\NS}{\mathbb{N}}
\newcommand{\ZS}{\mathbb{Z}}
\newcommand{\dnormal}{\mathit{N}}
\newcommand{\duniform}{\mathit{U}}

\newcommand{\erdos}{Erd\H{o}s}
\newcommand{\renyi}{-R\'{e}nyi}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Display of TOCs:
\AtBeginSection[]
{
	\setcounter{tocdepth}{2}  
	\frame
	{
	  \frametitle{Outline}
		\tableofcontents[currentsection]
	}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Document:
\begin{document}
\renewcommand{\arraystretch}{1.2}

\begin{frame}[plain] % plain => kein Rahmen
  \titlepage
\end{frame}
%\setcounter{framenumber}{0}

\frame{
	\frametitle{Outline}
	\tableofcontents
}

%%%%%%%%%%%%%%
% Your Content

\section{Motivation and Question}

\frame[t]{
  \frametitle{Motivation}
  \begin{itemize}
  	\item Add your motivation here
		\begin{itemize}
			\item Maybe  with some details
			\begin{itemize}
				\item but not too much
			\end{itemize}
		\end{itemize}
		\item Use references \textsuperscript{[Author, 2010]}
	\end{itemize}
}

\section{Basics and Definition}

\section{Approach}

\section{Results}

\section{Conclusion}

\frame[t]
{
  \frametitle{Conclusion}

  Novelty and contribution of this work:
  \begin{itemize}
	  \item Sum up the approach
		\item Sum op the results
		\item ...
		\item Show that it solves the question
  \end{itemize}

\mbox{ }

	Open Questions:
  \begin{itemize}
  	\item Something
	  \item ... is always missing
  \end{itemize}
}

%%%%%%%%%%%%%%

\frame[c]{
  \frametitle{The End}
\begin{center}
  Thank you for your attention.\\[1ex]
  Any question?\\[5ex]
\end{center}
  \footnotesize
  Literature:
	\begin{itemize}
		\item Author , Author , Author, and Author. Name of the conference paper. \emph{In: Proceedings of the Conference Name}, 2008
		\item Author, Author, and Author. Name of the Article. \emph{Name of the Journal}, 42:111-133, 2010
		\item Author, and Author. \emph{Name of the Book}. Publisher, 2009
	\end{itemize}
}

\end{document}

### 2024-10-18 BAI/templates/short_paper_generic/shortpaper.tex
Added templates.

\documentclass[12pt]{article}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Meta informations:
\newcommand{\trauthor}{Horst Hansen}
\newcommand{\trtype}{Proposal} %{Expos\'{e}} %{Review}
\newcommand{\trtitle}{Neural Networks for Artificial Agents}
\newcommand{\trdate}{01.09.2010}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Languages:

% Falls die Ausarbeitung in Deutsch erfolgt:
% \usepackage[german]{babel}
% \usepackage[T1]{fontenc}
% \usepackage[latin1]{inputenc}
% \usepackage[latin9]{inputenc}	 				
% \selectlanguage{german}

% If the thesis is written in English:
\usepackage[english]{babel} 						
\selectlanguage{english}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Bind packages:
\usepackage{acronym}                    % Acronyms
\usepackage{algorithmic}								% Algorithms and Pseudocode
\usepackage{algorithm}									% Algorithms and Pseudocode
\usepackage{amsfonts}                   % AMS Math Packet (Fonts)
\usepackage{amsmath}                    % AMS Math Packet
\usepackage{amssymb}                    % Additional mathematical symbols
\usepackage{amsthm}
\usepackage{booktabs}                   % Nicer tables
%\usepackage[font=small,labelfont=bf]{caption} % Numbered captions for figures
\usepackage{color}                      % Enables defining of colors via \definecolor
\definecolor{uhhRed}{RGB}{254,0,0}		  % Official Uni Hamburg Red
\definecolor{uhhGrey}{RGB}{122,122,120} % Official Uni Hamburg Grey
\usepackage{fancybox}                   % Gleichungen einrahmen
\usepackage{fancyhdr}										% Packet for nicer headers
%\usepackage{fancyheadings}             % Nicer numbering of headlines

%\usepackage[outer=3.35cm]{geometry} 	  % Type area (size, margins...) !!!Release version
%\usepackage[outer=2.5cm]{geometry} 		% Type area (size, margins...) !!!Print version
%\usepackage{geometry} 									% Type area (size, margins...) !!!Proofread version
\usepackage[outer=3.15cm]{geometry} 	  % Type area (size, margins...) !!!Draft version
\geometry{a4paper,body={5.8in,9in}}

\usepackage{graphicx}                   % Inclusion of graphics
%\usepackage{latexsym}                  % Special symbols
\usepackage{longtable}									% Allow tables over several parges
\usepackage{listings}                   % Nicer source code listings
\usepackage{multicol}										% Content of a table over several columns
\usepackage{multirow}										% Content of a table over several rows
\usepackage{rotating}										% Alows to rotate text and objects
\usepackage[hang]{subfigure}            % Allows to use multiple (partial) figures in a fig
%\usepackage[font=footnotesize,labelfont=rm]{subfig}	% Pictures in a floating environment
\usepackage{tabularx}										% Tables with fixed width but variable rows
\usepackage{url,xspace,boxedminipage}   % Accurate display of URLs

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Configurationen:

\hyphenation{whe-ther} 									% Manually use: "\-" in a word: Staats\-ver\-trag

%\lstloadlanguages{C}                   % Set the default language for listings
\DeclareGraphicsExtensions{.pdf,.svg,.jpg,.png,.eps} % first try pdf, then eps, png and jpg
\graphicspath{{./src/}} 								% Path to a folder where all pictures are located
\pagestyle{fancy} 											% Use nicer header and footer

% Redefine the environments for floating objects:
\setcounter{topnumber}{3}
\setcounter{bottomnumber}{2}
\setcounter{totalnumber}{4}
\renewcommand{\topfraction}{0.9} 			  %Standard: 0.7
\renewcommand{\bottomfraction}{0.5}		  %Standard: 0.3
\renewcommand{\textfraction}{0.1}		  	%Standard: 0.2
\renewcommand{\floatpagefraction}{0.8} 	%Standard: 0.5

% Tables with a nicer padding:
\renewcommand{\arraystretch}{1.2}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Additional 'theorem' and 'definition' blocks:
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
%\newtheorem{theorem}{Satz}[section]		% Wenn in Deutsch geschrieben wird.
\newtheorem{axiom}{Axiom}[section] 	
%\newtheorem{axiom}{Fakt}[chapter]			% Wenn in Deutsch geschrieben wird.
%Usage:%\begin{axiom}[optional description]%Main part%\end{fakt}

\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]

%Additional types of axioms:
\newtheorem{lemma}[axiom]{Lemma}
\newtheorem{observation}[axiom]{Observation}

%Additional types of definitions:
\theoremstyle{remark}
%\newtheorem{remark}[definition]{Bemerkung} % Wenn in Deutsch geschrieben wird.
\newtheorem{remark}[definition]{Remark}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Provides TODOs within the margin:
\newcommand{\TODO}[1]{\marginpar{\emph{\small{{\bf TODO: } #1}}}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Abbreviations and mathematical symbols
\newcommand{\modd}{\text{ mod }}
\newcommand{\RS}{\mathbb{R}}
\newcommand{\NS}{\mathbb{N}}
\newcommand{\ZS}{\mathbb{Z}}
\newcommand{\dnormal}{\mathit{N}}
\newcommand{\duniform}{\mathit{U}}

\newcommand{\erdos}{Erd\H{o}s}
\newcommand{\renyi}{-R\'{e}nyi}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Document:
\begin{document}
\renewcommand{\headheight}{14.5pt}

\fancyhead{}
\fancyhead[CO]{\trtitle}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Cover Header:
\title{\trtitle\\[0.3cm]{\normalsize\trtype}}
\author{\trauthor}
\date{\trdate}
\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\thispagestyle{empty}
\pagenumbering{arabic}

% Abstract gives a brief summary of the main points of a paper:
\begin{abstract}
  This paper gives a brief overview and an example of how to write a short paper in \LaTeX. An example for a short paper could be a proposal or a review.
\end{abstract}

% the actual content, usually separated over a number of sections
% each section is assigned a label, in order to be able to put a
% crossreference to it

\section{Introduction}
\label{sec:introduction}

The introduction describes the problem, why it is important, the main
ideas of the following paper, what are the main contributions of the
paper, etc.

If the paper has more then 4 pages, a readers guide is recommenced. In a short paragraph an outline of the content of the next

\section{Optional: Background Information}
\label{sec:basics}
A brief section giving background information may be necessary, especially if your work spans two or more traditional fields. That means that your readers may not have any experience with some of the material needed to follow your thesis, so you need to give it to them. A different title than that given above is usually better; e.g., "A Brief Review of Frammis Algebra."

\section{Related Work}
\label{sec:relwork}

The related work section could describe other work that is in some respects relevant for the understanding of the problem outlined in Section~\ref{sec:introduction}, that offer competing solutions, etc.

All sources must be properly referenced, ideally by using the BiBTeX system. References can then be very conveniently made with the \textsc{cite} command. For example, reference
\cite{Leunen:Scholars:92} discusses some of the elementary rules on
writing scientific papers, amongst others how to correctly cite other
documents. Reference \cite{Taylor:SIGuide:95}, e.g., describes how to
correctly use the SI system of units and their correct typographical
representation. In general you organize this section by idea, and not by author or by publication.

\section{Model description}
\label{sec:model}

After the two (or maybe three) common sections Introduction (plus optional Background) and Related work, more sections with the actual content of a paper follow. The style and
structure of such sections varies by a large degree, no general rules
of thumb can be given.

\subsection{Word processing with \LaTeX}
\label{sec:model:subsec:latex}

This document has already introduced the most important constructs of
\LaTeX. What is necessary to produce documents with \LaTeX is simple
any normal text editor and a \LaTeX distribution. This is commonly
installed on practically all UNIX-type systems; for Windows, an
excellent \LaTeX exists, called MikTeX, available from
\url{www.miktex.org}. Almost all distributions come with a large patch
of examples and introductory material; consult your local installation
for details.

Lots of supplementary and background information, FAQs, etc.\ is
available from the Comprehensive TeX Archive Network (CTAN); the
German mirror of which is \url{www.dante.de}.

\subsection{Tables in \LaTeX}
\label{sec:model:subsec:tables}
Tables should be centred and should always have a caption 
positioned above it. A caption in a sentence form as well as in a short form must end with a period as seen in table~\ref{tab:sample}.

\begin{table}[hbtp]
  \caption{This caption has one line so it is centred.}\label{tab:sample} 
  \centering
  \begin{tabular}{|c|c|}
    \hline
    Example column 1 & Example column 2 \\
    \hline
    Example text 1 & Example text 2 \\
    \hline
  \end{tabular}
\end{table}

\subsection{Figures in \LaTeX}
\label{sec:model:subsec:figures}
Note that a figure is a so-called floating object: it is moved around the actual
text in order to best fit on a page. This is in stark contrast to some
GUI-based word processing tools, where the placement of figures is
usually more associated with luck than principle.

As figures float around, expressions like ``the following figure''
must never be used. Instead, figures need a caption, a label, and must
be properly referenced in the main text. A figure caption is placed centred below the figure and describes the figure in (very) short.

In general, only vector graphics in encapsulated postscript (eps) or a similar format should be included in any kind of text, as this allows arbitrary scaling, rotation etc.\ without any loss of quality. Bitmap formats (JPEG, GIF, \dots) should only be used if no other alternative exists
--- basically the only case where bitmaps can be justified is when scanned pictures need to be included in a text, however, this should be avoided as hard as possible as the quality in usually not satisfactory. If a screen shot is needed a high resolution picture without visible fragments of a jpeg compression is allowed. Figures like the figure~\ref{fig:samplefig} should always appear after the first referencing it.
\begin{figure}[hbtp]
	 \centerline{\includegraphics[width=0.6\textwidth]{samplefig.pdf}}
	 {\caption{Network of transputers and the structure of individual
processes.}\label{fig:samplefig}}
\end{figure}

\section{Model analysis}
\label{sec:analysis}
The analysis of a model is a much more free-form of a paper. Starting with a comparison of several approaches over some experimental settings and result up to a theoretical verification of the model, anything is allowed. But it all has only one purpose: to convince the reader of the right to exist of the described model.

\section{Conclusion}
\label{sec:concl}

At the end, there is a final section concluding and summarizing a
paper, putting the entire work into perspective and explaining, on a
larger level, what the consequence of this work are. Also, unexpected
results can be discussed here, etc.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% hier werden - zum Ende des Textes - die bibliographischen Referenzen
% eingebunden
%
% Insbesondere stehen die eigentlichen Informationen in der Datei
% ``bib.bib''
%
\bibliography{bib}
\bibliographystyle{plain}

\end{document}

### 2024-10-18 PbD/templates/TemplateLaTexText/hinweiseabschlussarbeit.tex
Added templates.

%!TEX encoding = UTF-8 Unicode
% ================================================================================
\documentclass[
	fontsize=12pt,
	headings=small,
	parskip=half,           % Ersetzt manuelles Setzen von parskip/parindent.
	bibliography=totoc,
	numbers=noenddot,       % Entfernt den letzten Punkt der Kapitelnummern.
	open=any,               % Kapitel kann auf jeder Seite beginnen.
%	final                   % Entfernt alle todonotes und den Entwurfstempel.
]{scrreprt}
% ================================================================================
% Hinweis: Das Übersetzen dieser Datei funktioniert auch Online mit 
% https://www.overleaf.com. Hierzu müssen neben dieser Datei im gleichen 
% Verzeichnis die Dateien hinweiseabschlussarbeit.bib und stylesvs.tex liegen.
% ================================================================================
\input{stylesvs}
% ===================================Dokument===================================
\title{Hinweise für das Erscheinungsbild von Seminar-, Studien- und Bachelor-, Master- und Diplomarbeiten}
\author{Hannes Federrath}
% \date{01.01.2015} % Falls ein bestimmtes Datum eingesetzt werden soll, einfach diese Zeile aktivieren.

\begin{document}

\begin{titlepage}
% \includegraphics[width=6.8cm]{../pic/up-uhh-logo-u-2010-u-farbe-u-rgb.pdf}
\mbox{\parbox[t][1.75cm][b]{2.2cm}{\uhhlogo}}
\begin{center}\Large
	% Universität Hamburg \par
	% Fachbereich Informatik
	\vfill
	\makeatletter
	{\Large\textsf{\textbf{\@title}}\par}
	\makeatother
	\bigskip
	am Arbeitsbereich Sicherheit in Verteilten Systemen (SVS) \par
	\bigskip
	\makeatletter
	{\@author} \par
	\makeatother
	\bigskip
	\makeatletter
	{\@date}
	\makeatother
	\vfill
	\vfill
	%(Muster für das Deckblatt: siehe Anhang dieser Hinweise)
\end{center}
\end{titlepage}

\chapter*{Aufgabenstellung}

Nur Studien-, Bachelor-, Master- und Diplomarbeiten: Soweit eine ausformulierte Aufgabenstellung mit der Betreuerin bzw. dem Betreuer vereinbart wurde, diese bitte hier einfügen.

\chapter*{Zusammenfassung}

Für die eilige Leserin bzw. den eiligen Leser sollen auf etwa einer halben, maximal einer Seite die wichtigsten Inhalte, Erkenntnisse, Neuerungen bzw. Ergebnisse der Arbeit beschrieben werden.

Durch eine solche Zusammenfassung (im Engl. auch Abstract genannt) am Anfang der Arbeit wird die Arbeit deutlich aufgewertet. Hier sollte vermittelt werden, warum man die Arbeit lesen sollte.

\tableofcontents

\chapter{Vorbemerkung}

Um auf die wiederholten Fragen von Studierenden nach dem Umfang, formalen Aufbau und Erscheinungsbild, das bei Seminar-, Studien-, Bachelor-, Master- und Diplomarbeiten erwartet wird, einheitlich zu antworten, wird dieses Dokument bereitgestellt.

Diese Hinweise haben Empfehlungscharakter. Bei Unklarheiten geben die Mitarbeiterinnen und Mitarbeiter der Arbeitsgruppe gerne weitere Auskünfte. Als Muster steht auch eine große Anzahl abgeschlossener Arbeiten zur Ansicht zur Verfügung.

\chapter{Inhalt}
\label{sec.inhalt}

Eine Seminar-, Studien-, Bachelor-, Master- und Diplomarbeit ist eine längere wissenschaftliche Abhandlung, mit der die Studierenden zeigen sollen, dass sie in einem vorgegebenen Zeitrahmen in der Lage sind, wissenschaftlich zu arbeiten. Gelegentlich werden im Rahmen des Studiums auch synonyme Begriffe wie Abschlussarbeit, Hausarbeit oder Projektbericht verwendet. Die hier niedergeschriebenen Empfehlungen gelten für alle genannten Dokumentarten gleichermaßen.

\section{Anforderungen an eine Arbeit}
\label{sec.anforderungen}

Eine Seminar-, Studien-, Bachelor-, Master- und Diplomarbeit trägt inhaltlich normalerweise zu einem aktuell am Arbeitsbereich bearbeiteten Forschungsthema bzw. -projekt bei oder leistet einen Beitrag in der Lehre (z.\,B. Erstellung eines Lehrmittels).

Normalerweise besteht eine Arbeit aus einem darstellenden und einem konstruktiven Teil. Im darstellenden Teil wird gezeigt, dass mit wissenschaftlicher Literatur umgegangen, Wichtiges von Unwichtigem getrennt werden kann und die relevanten Aspekte etwaiger Vorarbeiten erfasst und dargestellt werden können. Im konstruktiven Teil werden dann eigene Lösungen erarbeitet und bewertet.

Um den inhaltlichen und sprachlichen Stil einer wissenschaftlichen Arbeit zu treffen, ist es sehr empfehlenswert, beim Lesen wissenschaftlicher Publikationen auf deren „Klang“ \cite{Tolk2003} zu achten. Die Ich-Form bzw. die Wir-Form sollte im Übrigen vermieden werden.

\section{Aufbau der Arbeit}
\label{sec.aufbau}

Eine wissenschaftliche Arbeit sollte -- wie nahezu jede schriftliche Arbeit -- aus einer Einleitung, einem Hauptteil und einem Schluss bestehen. Der Einleitung ist ein Deckblatt, eine Zusammenfassung und ein Inhaltsverzeichnis voranzustellen. Tabellen- und Abbildungsverzeichnisse sind optional.

Als Muster kann dieses Dokument herangezogen werden.

In der Einleitung wird die Problemstellung und deren Relevanz geschildert. Außerdem werden die Methoden genannt, mit der die Problemstellung bearbeitet wird.

Der Hauptteil sollte mit einem Kapitel zum Stand der Wissenschaft bzgl. des zu bearbeitenden Problems beginnen und das eigene Problem einordnen. Soweit erforderlich, können in einem weiteren Kapitel Grundlagen gelegt werden, z.\,B. Grundverfahren beschrieben werden, die in den folgenden Kapiteln benutzt, ausgebaut oder verändert werden.

Der Schluss fasst die Ergebnisse noch einmal zusammen, bewertet die eigenen Ergebnisse kritisch und benennt die offenen Fragen. Es ist völlig normal, dass im Verlauf der Bearbeitung neue Problemstellungen und Forschungsfragen entstehen, die dann wieder der Ausgangspunkt für weitere Arbeiten sein können.

Ein Literaturverzeichnis am Ende ist obligatorisch. Man sollte sich stets auf die besten Quellen stützen. In abnehmender Qualität:

\begin{enumerate}
	\item Fachbücher, Standards,
	\item Wiss. Zeitschriftenartikel, Survey-Artikel,
	\item Konferenzbeiträge,
	\item Technical Reports, graue Literatur,
	\item Online-Material, Arbeitspapiere, Firmenmaterial, Ausarbeitungen.
\end{enumerate}

Im Internet können zur Feststellung der Qualität und Recherche von Publikationen z.\,B.

\begin{itemize}
	\item Google Scholar (\url{http://scholar.google.com}),
	% \item Microsoft Academic (\url{https://academic.microsoft.com/}),
		% alt: Microsoft Academic Search (\href{http://academic.research.microsoft.com/?SearchDomain=2&SubDomain=2&entitytype=2}{http://academic.research.microsoft.com}) $\to$ computer science $\to$ security \& privacy,
	\item Computer Science Bibliography (\url{http://dblp.uni-trier.de/}) und die
	\item CiteSeerX
		%ehemals Scientific Literature Digital Library 
		(\url{https://citeseerx.ist.psu.edu/})
\end{itemize}

herangezogen werden.

Bei Bedarf kann auch ein Index und Abkürzungsverzeichnis beigefügt werden. Bei Seminar-, Studien-, Bachelor-, Master- und Diplomarbeiten ist dies jedoch normalerweise wegen des überschaubaren Umfangs nicht sinnvoll.

Bei umfangreichen Quelltexten (mehr als 2 Seiten) sollten diese nicht im Fließtext wiedergegeben werden, sondern im Anhang oder (mit Verweis in der schriftlichen Arbeit) auf dem beigelegten Datenträger erscheinen. Dies gilt auch für andere den Lesefluss störende Informationen größeren Umfangs.

Für Prüfungsarbeiten wie Bachelor-, Master- und Diplomarbeiten ist wichtig: Eigenhändig unterschriebene Selbständigkeitserklärung am Anfang oder Ende des Textes nicht vergessen (siehe Muster am Ende dieser Hinweise). Bei Seminararbeiten kann diese entfallen.

\chapter{Form}
\label{sec.form}

Abgesehen von inhaltlichen Anforderungen kann eine klare, gut strukturierte Form den Gesamteindruck der schriftlichen Ausarbeitung steigern. Obwohl die Form keine neuen Inhalte einführt, kann sie sich doch stark auf das Leseverständnis und die Informationsaufnahme der Lesenden auswirken.

\section{Umfang der schriftlichen Ausarbeitung}
\label{sec.umfang}

Generell gilt: Je weniger Text, umso besser. Auf klare Formulierungen ist in jedem Fall zu achten. Für Studien-, Bachelor-, Master- und Diplomarbeiten ist der Richtwert 40--50 Seiten. 20~Seiten sind zu wenig, 100 sind zu viel. Bei Seminararbeiten genügen 10--15 Seiten (maximal 10 Seiten Text zzgl. Deckblatt, Literaturverzeichnis und ggf. Anhang, insgesamt nicht mehr als 15 Seiten).

Insbesondere für Bachelor-, Master- und Diplomarbeiten gilt: Wo immer möglich, sollte auf andere relevante Veröffentlichungen verwiesen werden, anstatt deren Inhalt noch einmal wiederzugeben. Für alle Aussagen und Darstellungen, die aus Veröffentlichungen stammen, sind Quellenangaben zu machen.

Bei Inhalten aus fremden Quellen, die paraphrasiert oder wörtlich übernommen werden, ist die Quellenangabe an der Textstelle zu machen. Es genügt nicht, die Quelle ins Literaturverzeichnis aufzunehmen. Wörtliche Übernahmen von längeren Wortgruppen und ganzen Sätzen sind in Anführungszeichen zu setzen.

Die sprachliche Leistungsfähigkeit von generativen KI-Programmen wie etwa ChatGPT kann dazu verleiten, Zusammenfassungen fremder Texte einer KI zu überlassen. Der Einsatz solcher Hilfsmittel zur Erstellung eigener wissenschaftlicher Texte muss klar gekennzeichnet sein. Auch hier gilt: Die Quellenangabe ist an der Textstelle zu machen.

Viele Studierende haben zu Beginn der Bearbeitung Sorge, dass sie womöglich nicht auf die „übliche“ Seitenzahl kommen. Diese Sorge ist meist unbegründet. Es sollte möglichst früh mit dem Schreiben begonnen werden: Dokumentieren Sie, was Sie gelesen und neu erarbeitet haben.

\section{Gestaltung}

Wissenschaftliche Arbeiten, die am Arbeitsbereich betreut werden, sollen mit LaTeX gesetzt sein. Ausnahmen von dieser Regel (etwa die Verwendung von Open Office oder Word) können in Absprache mit der Betreuerin bzw. dem Betreuer getroffen werden.

Als Hauptschriftart sollte eine mit Serifen verwendet werden, z.\,B. Times (wie dieser Text). In LaTeX kann auch Latin Modern (\textbackslash usepackage\{lmodern\}) verwendet werden oder besser, falls möglich, die Postscript-Schrift Times (\textbackslash usepackage\{mathptmx\}). Bitte verwenden Sie keine \textsf{Helvetica} oder \textsf{Arial}, da diese Schriften bei längeren Texten schwerer lesbar sind. In Überschriften ist eine serifenlose Schrift jedoch in \textsf{\textbf{Bold}} erlaubt, wie in diesem Beispiel.

Die Schriftgröße sollte 12pt (wie dieser Text), zur Not auch 11pt sein. Eine Größe von 10pt ist zu klein! Als Zeilenabstand sollten möglichst 15pt oder 14pt verwendet werden. 1,5-zeilig entspricht etwa 18pt und ist zu viel. Bei LaTeX sind keine benutzerdefinierten Abstände nötig. Der Text ist im Blocksatz zu setzen. Ränder bei A4-Papier: ca. 2,5--3cm rundherum. In LaTeX erzeugt beispielsweise \textbackslash usepackage[a4paper, margin=2.54cm, marginparwidth=2.0cm, footskip=1.0cm]\{geometry\} einen geeigneten Satzspiegel (wie dieses Dokument).

Es sollten möglichst nicht mehr als drei Gliederungsebenen verwendet werden. Die Nummerierung einer Überschrift erfolgt stets \emph{ohne} nachgestellten Punkt. Weiterhin sollte eine Überschrift nie allein für sich stehen, sondern von einem Text begleitet werden. Folgt auf eine Hauptüberschrift direkt eine Unterüberschrift, könnte zwischen den beiden Überschriften beispielsweise eine Einführung zum jeweiligen thematischen Abschnitt oder ein Überblick über die darauffolgenden Unterabschnitte gegeben werden. In diesem Dokument sind dies beispielsweise zwischen den Überschriften zu Kapitel~\ref{sec.form} und Abschnitt~\ref{sec.umfang} zwei Sätze, die das Kapitel einleiten.

Innerhalb eines Kapitels sollte eine Untergliederung immer dann verwendet werden, wenn es deutlich mehr als zwei Seiten Text umfasst. Bei der Untergliederung ist zu beachten, dass stets wenigstens zwei Unterabschnitte vorhanden sein sollten. Das Kapitel~\ref{sec.inhalt} dieses Textes hat beispielsweise die Unterabschnitte~\ref{sec.anforderungen} und \ref{sec.aufbau}, und es wäre eigenartig, wenn ein Kapitel~x nur den Unterabschnitt~x.1 hätte und kein weiterer folgen würde. Dann ist es besser, auf eine Untergliederung zu verzichten.

Eine Kopfzeile kann verwendet werden, muss aber nicht. Hier wird oft unnötig Zeit verschwendet.

Bitte benutzen Sie nur einen Absatztyp (wie in diesem Dokument; wird in LaTeX durch mindestens eine Leerzeile zwischen den Absätzen erzeugt). Es ist weit verbreitet, Gedanken, die irgendwie zusammenhängen, aber aus Sicht des Autors noch keinen neuen Absatz rechtfertigen, auf einer neuen Zeile zu beginnen -- in LaTeX meist durch {\textbackslash\textbackslash} erzeugt.\\ Dies ist zu vermeiden, weil es das Textbild uneinheitlich und unruhig macht. Man soll zwar keine Negativbeispiele bringen, aber der Zeilenwechsel vor dem vorherigen Satz ist eines.

Dieses Dokument wurde mit LaTeX erstellt und steht übrigens auch im Quelltext (.tex-File) zur Verfügung und kann für eigene Zwecke weiterverwendet werden. Die Befehlsfolge zum Erzeugen eines PDF aus diesem Dokument lautet etwa:

\begin{lstlisting}[numbers=none,xleftmargin=6pt]
pdflatex hinweiseabschlussarbeit.tex
biber hinweiseabschlussarbeit.bcf
pdflatex hinweiseabschlussarbeit.tex
pdflatex hinweiseabschlussarbeit.tex
\end{lstlisting}

Weiterführende Literatur zum Schreiben wissenschaftlicher Arbeiten mit LaTeX findet sich beispielsweise in \cite{Schl2013}.

Für detaillierte Informationen zu typographischen Regeln sowie Beispiele der korrekten Umsetzung dieser Regeln sei auf die kompakte und sehr hilfreiche Arbeit „typokurz -- Einige wichtige typographische Regeln“ von Christoph Bier verwiesen \cite{Bier2009}.

Häufig werden die Regeln zum Setzen von Text in Anführungszeichen missachtet. Im Deutschen sollten nur die „Gänsefüßchen“ (links nach unten und rechts nach oben geschwungen) verwendet werden. Genaues Hinschauen ermöglicht hier die korrekte Verwendung: Weder „dies” noch {\verb#"#}das{\verb#"#} noch “jenes” ist korrekt.

Bei der Kommasetzung bietet es sich an, noch einmal die Regeln unter \url{https://www.duden.de/sprachwissen/rechtschreibregeln/komma} anzuschauen. Sehr häufig werden notwendige Kommas weggelassen, etwa beim erweiterten Infinitiv mit zu (genauer: bei satzwertigen Infinitivgruppen, vgl. Duden, Regel D117). Dagegen werden Kommas leider immer wieder dort gesetzt, wo sie nicht hingehören, z.\,B. werden keine Kommas vor „etc.“ und „sowie“ gesetzt.

Beim Verwenden von zusammengesetzten Substantiven und anderen Aneinanderreihungen, wie sie in informatischen Texten sehr häufig vorkommen, gilt im Deutschen die Regel, dass diese üblicherweise mit einem Bindestrich verbunden werden. Man schreibt etwa DES-Verschlüsselung, IP-Adresse und Public-Key-Verfahren. Näheres zum Nachlesen findet sich in den deutschen Rechtschreibregeln. Eine E-Mail ist eine elektronisch übermittelte Nachricht, während Email ein Keramiküberzug ist. Man schreibt im Deutschen selten „Netzwerk“, wenn man ein Kommunikationsnetz oder Rechnernetz meint. Als Kurzform für das Englische \emph{computer network} verwendet man den Begriff Netz.

Wenn es möglich ist, ein zusammengesetztes Wort ohne Bindestrich zu schreiben, dann sollte davon Gebrauch gemacht werden. So sollte beispielsweise besser „IT-Sicherheitsmanagement“ anstelle von „IT-Sicherheits-Management“ geschrieben werden.

Zwar mögen solche Formfragen aus Sicht des wissenschaftlichen Gehalts eines Textes eher nachrangig sein, allerdings verhilft eine saubere, fehlerfreie und konsistente Form zu einem positiven Gesamteindruck. Dagegen kann eine hohe Rate an Rechtschreib- und Interpunktionsfehlern einen inhaltlich guten Text eigentlich nur schwächen.

Weniger kann übrigens manchmal mehr und Besseres bewirken. Spiegel Online berichtete in \cite{textwahrnehmung} beispielsweise, dass einfache, klare Sprache und eine gut lesbare Standardschrift die Textwahrnehmung verbessern kann: „Schreib so einfach und deutlich wie möglich, dann hält man dich eher für intelligent.“

Eine interessante Regel für klares Schreiben ist die „Daumenregel“, die durchaus wörtlich zu nehmen ist: Wenn ein Satz durch Weglassen (mit dem Daumen Verdecken) eines Wortes oder einer Wortgruppe noch immer den gewünschten Sinn ergibt, dann sollte dieses Wort bzw. die Wortgruppe gestrichen werden. Diese Regel kann auch auf ganze Sätze oder gar Abschnitte erweitert werden.

Darüber hinaus sollten beim Schreiben unbestimmte Aussagen vermieden werden. Wenn etwa in einem Text die Rede davon ist, dass es „verschiedene Verfahren zur Angriffserkennung gibt“, dann ist es für die positive Wahrnehmung der entsprechenden Textstelle hilfreich, entweder eine Quellenangabe zu machen, bei der diese verschiedenen Verfahren zur Angriffserkennung genannt und näher erläutert werden, oder kurz und unmittelbar die konkreten Verfahren zu nennen, um die Leserinnen und Leser nicht im Unklaren zu lassen. Wenn öfter von „unterschiedlichen“, „verschiedenen“ oder „mehreren“ Dingen geschrieben wird, sollte ernsthaft die Anwendung der vorangegangenen Daumenregel in Betracht gezogen werden.

Tipps und Textbausteine sowie die Grundlagen des wissenschaftlichen Schreibens sind in \cite{Küht2021} sehr ansprechend und mit zahlreichen Beispielen und Gegenbeispielen zu finden.

\section{Abbildungen, Tabellen und Listings}

Abbildungen sollten möglichst schlicht, in schwarzweiß und als Strichzeichnungen gestaltet sein. Wenn schon Farben verwendet werden, dann bitte in \emph{allen} Abbildungen das gleiche Farbschema verwenden. Farben sind nur dann sinnvoll, wenn sie einen Sachverhalt deutlich unterstreichen oder veranschaulichen. Es ist zu beachten, dass die Aussagekraft auch in einem Schwarzweiß-Ausdruck erhalten bleiben muss.

Die Auflösung muss ausreichend groß gewählt werden, damit im fertigen Dokument weder Pixel noch Treppen oder Unschärfe erkennbar sind. Deshalb möglichst Vektorgrafiken verwenden.

Gleitobjekte (sog. Floats) wie Abbildungen und Tabellen müssen eine Unterschrift erhalten. Auf diese muss zudem im Text eindeutig verwiesen werden, da durch das automatische Setzen unter Umständen nicht ersichtlich ist, zu welchem Textabschnitt eine Abbildung gehört. Wie das aussehen kann, ist anhand von Abbildung~\ref{fig:bsp} ersichtlich.

\begin{figure}[ht]
\sffamily\footnotesize
%\includegraphics[width=0.6\textwidth]{abb.pdf}
\unitlength=0.75mm
\special{em:linewidth 0.4pt}
\linethickness{0.4pt}
\begin{picture}(111,73)(0,0)
	\put(21,59){\makebox(0,0)[cc]{}}
	\put(1,55){\line(1,0){40}}
	\put(1,55){\line(0,1){8}}
	\put(41,55){\line(0,1){8}}
	\put(1,63){\line(1,0){40}}
	\put(91,59){\makebox(0,0)[cc]{}}
	\put(71,55){\line(1,0){40}}
	\put(71,55){\line(0,1){8}}
	\put(111,55){\line(0,1){8}}
	\put(71,63){\line(1,0){40}}
	\put(21,19){\makebox(0,0)[cc]{}}
	\put(1,15){\line(1,0){40}}
	\put(1,15){\line(0,1){8}}
	\put(41,15){\line(0,1){8}}
	\put(1,23){\line(1,0){40}}
	\put(91,19){\makebox(0,0)[cc]{}}
	\put(71,15){\line(1,0){40}}
	\put(71,15){\line(0,1){8}}
	\put(111,15){\line(0,1){8}}
	\put(71,23){\line(1,0){40}}
	\put(31,43){\makebox(0,0)[cc]{F}}
	\put(24,38){\line(1,0){14}}
	\put(24,38){\line(0,1){10}}
	\put(38,38){\line(0,1){10}}
	\put(24,48){\line(1,0){14}}
	\put(81,43){\makebox(0,0)[cc]{F}}
	\put(74,38){\line(1,0){14}}
	\put(74,38){\line(0,1){10}}
	\put(88,38){\line(0,1){10}}
	\put(74,48){\line(1,0){14}}
	\put(31,30){\circle{6}}
	\put(81,30){\circle{6}}
	\put(79,30){\line(1,0){4}}
	\put(81,28){\line(0,1){4}}
	\put(29,30){\line(1,0){4}}
	\put(31,28){\line(0,1){4}}
	\put(21,63){\line(0,1){7}}
	\put(21,63){\vector(0,-1){0.12}}
	\put(31,48){\line(0,1){7}}
	\put(31,48){\vector(0,-1){0.12}}
	\put(31,33){\line(0,1){5}}
	\put(31,33){\vector(0,-1){0.12}}
	\put(31,23){\line(0,1){4}}
	\put(31,23){\vector(0,-1){0.12}}
	\put(21,8){\line(0,1){7}}
	\put(21,8){\vector(0,-1){0.12}}
	\put(11,30){\line(1,0){17}}
	\put(28,30){\vector(1,0){0.12}}
	\put(91,63){\line(0,1){7}}
	\put(91,63){\vector(0,-1){0.12}}
	\put(81,48){\line(0,1){7}}
	\put(81,48){\vector(0,-1){0.12}}
	\put(81,33){\line(0,1){5}}
	\put(81,33){\vector(0,-1){0.12}}
	\put(81,23){\line(0,1){4}}
	\put(81,23){\vector(0,-1){0.12}}
	\put(91,8){\line(0,1){7}}
	\put(91,8){\vector(0,-1){0.12}}
	\put(84,30){\line(1,0){17}}
	\put(84,30){\vector(-1,0){0.12}}
	\put(11,30){\line(0,1){25}}
	\put(101,30){\line(0,1){25}}
	\put(21,73){\makebox(0,0)[cc]{Klartext M}}
	\put(91,73){\makebox(0,0)[cc]{Chiffretext C}}
	\put(21,3){\makebox(0,0)[cc]{Chiffretext C}}
	\put(91,3){\makebox(0,0)[cc]{Klartext M}}
	\put(56,43){\makebox(0,0)[cc]{K}}
	\put(56,48){\makebox(0,0)[cc]{Schlüssel}}
	\put(38,43){\line(1,0){13}}
	\put(38,43){\vector(-1,0){0.12}}
	\put(61,43){\line(1,0){13}}
	\put(74,43){\vector(1,0){0.12}}
	\put(21,55){\line(0,1){8}}
	\put(21,15){\line(0,1){8}}
	\put(91,55){\line(0,1){8}}
	\put(91,15){\line(0,1){8}}
	\put(11,59){\makebox(0,0)[cc]{L}}
	\put(31,59){\makebox(0,0)[cc]{R}}
	\put(81,59){\makebox(0,0)[cc]{R}}
	\put(11,19){\makebox(0,0)[cc]{R}}
	\put(81,19){\makebox(0,0)[cc]{L}}
	\put(101,19){\makebox(0,0)[cc]{R}}
	\put(13,52){\line(1,0){18}}
	\put(81,52){\line(1,0){18}}
	\put(99,23){\line(0,1){29}}
	\put(99,23){\vector(0,-1){0.12}}
	\put(13,23){\line(0,1){29}}
	\put(13,23){\vector(0,-1){0.12}}
	\put(31,19){\makebox(0,0)[cc]{L'}}
	\put(101,59){\makebox(0,0)[cc]{L'}}
\end{picture}
\caption{Beispiel für eine Abbildung}
\label{fig:bsp}
\end{figure}

Ein Abbildungsverzeichnis ist nicht unbedingt erforderlich, kann aber bei einer Vielzahl von verwendeten Abbildungen für Übersichtlichkeit sorgen.

Längere Listings sollten wie Abbildungen in einer Float-Umgebung untergebracht werden, d.h. eine Über- bzw. Unterschrift haben. Ein Beispiel zeigt Listing~\ref{lst:ggt}.

\begin{lstlisting}[float,caption={Berechnung des größten gemeinsamen Teilers zweier ganzer Zahlen a und b},label={lst:ggt}]
int getGGTOf(int a, int b) {
    // requires ((a > 0) && (b > 0)); ensures return > 0;
    int h;
    while (b != 0) {
        h = b;
        b = a % b; // % is the modulo operator. This line is long enough to show how line breaks in lstlisting are handled.
        a = h;
    }
    return a;
}
\end{lstlisting}

\section{Literaturverzeichnis}
\label{sec:literaturhowto}

Nachfolgend werden einige Hinweise für die Angaben im Literatur- bzw. Quellenverzeichnis gegeben. Es ist wichtig, dass alle für den jeweiligen Quellentyp relevanten Informationen angegeben werden. Zudem sollte darauf geachtet werden, dass die Quellenangaben stets einheitlich erfolgen, also beispielsweise die Autoren konsequent zuerst mit Vorname und dann mit Nachname genannt werden. In den folgenden Syntaxbeschreibungen sind optionale Angaben in eckigen Klammern angegeben.

\begin{description}

	\item[Zitierweise für Fachbücher:] \mbox{}\\[1ex]
	Syntax: Vorname Nachname. Buchtitel. {[}Auflage,{]} Erscheinungsort: Verlag, Jahr. \\[1ex]
	Beispiele: \cite{Beut2009,ScWe2007,Pfit90}

	\item[Zitierweise für Zeitschriften:] \mbox{}\\[1ex]
	Syntax: Vorname Nachname. Artikeltitel. Zeitschrift Jahrgang/Volume (Jahr), Seiten. \\[1ex]
	Beispiele: \cite{Kili2006,Lamp1981,ThKZ2002,Chau81,Chau88}

	\item[Zitierweise für Konferenzbeiträge:] \mbox{}\\[1ex]
	Syntax: Vorname Nachname. Beitragstitel. {[}Herausgeber/Editoren.{]}
	Konferenzband. {[}Volume. Buchserie.{]} Ort{[: Verlag]}, Datum, Seiten. \\[1ex]
	Beispiele: \cite{InBr2009,WWPK2010,HSFN2009,GoRS99,WaMS2008}

	\item[Zitierweise für Online-Quellen:] \mbox{}\\[1ex]
	Syntax: Vorname Nachname. Titel. {[}Quelle.{]} Datum. URL (Zugriffszeitpunkt). \\[1ex]
	Beispiele: \cite{CCC2009,Heise2011,textwahrnehmung}

\end{description}

Die Literatur sollte im Text durch alphanumerische Kürzel mit Erscheinungsjahr in eckigen Klammern angegeben werden.
% ----- Der nachfolgende Text ist obsolet geworden durch die konsequente Verwendung von Bibtex:
% Bei einem Autor werden meist die ersten drei Buchstaben des Nachnamens % % verwendet (Beispiel: \cite{Beut2009}), bei zwei oder drei Autoren werden die % Anfangsbuchstaben der Nachnamen verwendet (Beispiel: \cite{InBr2009}). Bei % vier oder mehr Autoren werden die ersten drei Buchstaben des Nachnamens des % ersten Autors gefolgt von einem Pluszeichen verwendet (Beispiel: \cite{HSFN2009}).
% ----- Ende des obsoleten Textes
Bei Online-Quellen ohne referenzierbare Autoren und Titel (z.B. Webseiten von Firmen und Organisationen) kann anstelle eines Eintrags im Literaturverzeichnis auch vereinfachend eine Fußnote mit der URL gesetzt werden. Hinter der URL ist der Zugriffszeitpunkt zu vermerken. Einige Fußnoten in diesem Text dienen als Beispiele.

In diesem Dokument wurde für die Erzeugung des Literaturverzeichnisses BibLaTeX verwendet. Im Literaturverzeichnis auf Seite~\pageref{sec:literaturverzeichnis} können Beispiele angeschaut werden. Solange die Angaben zu einer Quelle vollständig sind, erzeugt BibTeX automatisch eine korrekte Quellenangabe, die allerdings je nach verwendetem Bibstyle (hier: alphabetic) von den o.a. Hinweisen abweichen kann, was in Ordnung ist, solange die Angaben vollständig und einheitlich sind.

Teilweise wird in den Seminaren die Erstellung einer Literaturliste mit den fünf wichtigsten und besten Quellen gefordert. Im Anhang findet sich auf Seite~\pageref{sec:literaturliste} ein Beispiel für eine solche Literaturliste.

\section{Wikipedia und generative KI als Quellenangabe}

Grundsätzlich sollte bei der Literaturarbeit darauf geachtet werden, dass die Originalquelle
%für eine Information oder einen Sachverhalt
referenziert wird. Referenzen auf Wikipedia und Sekundärliteratur sollten daher möglichst vermieden werden.

Im wissenschaftlichen Kontext kann aber etwa aus didaktischen Gründen eine Referenz auf Inhalte aus Wikipedia, Lehrbücher und Sekundärliteratur trotzdem sinnvoll sein. Bei der Referenz auf einen Eintrag der Wikipedia sollte, wie auch bei anderen Online-Quellen, der Zugriffszeitpunkt angegeben werden. Der Verweis auf eine bestimmte Version eines Eintrags wird innerhalb von Wikipedia mittels der sog. \emph{oldid} realisiert. Ein Beispiel hierfür ist \cite{Wiki}.

Werden generative KI-Programme wie etwa ChatGPT\footnote{\url{https://openai.com/chatgpt}, Abruf am 14.07.2023} oder GitHub Copilot\footnote{\url{https://github.com/features/copilot/}, Abruf am 14.07.2023} innerhalb einer wissenschaftlichen Arbeit dazu verwendet, z.\,B. Programmschnippsel oder Text zu erzeugen, sind diese mit einer Quellenangabe klar als solche zu kennzeichnen. Es wird empfohlen, sowohl die Anfragen an als auch die Ausgaben  von generativer KI in der eigenen Materialsammlung zu sichern, um später ggf. noch genau nachvollziehen zu können, was die Eigenleistung im Zusammenhang mit solchen Hilfsmitteln war.

\section{Angabe von Literaturreferenzen im Fließtext}
\label{sec:literaturfliesstext}

Beispiele für die korrekte Angabe von Referenzen im Fließtext finden sich an vielen Stellen in diesem Dokument. Im Fließtext sollen Quellenangaben immer möglichst dicht an der jeweiligen Paraphrasierung genannt werden. Die nachfolgenden Satzbeispiele zeigen, wie so etwas gehen kann:

Das Mix-Netz \cite{Chau81} und das DC-Netz \cite{Chau88} bilden die Grundlage vieler moderner Verfahren zum Schutz vor Beobachtung im Internet. In \cite[S.~13]{Beut2009} wird ein Algorithmus zur statistischen Analyse der Cäsar-Chriffre angegeben. Soll bei einer Quellenangabe auch eine konkrete Seitenangabe gemacht werden, gelingt dies mit \textbackslash cite[S.\textasciitilde 13]\{XYZ\}, siehe das Beispiel zuvor.

Manchmal möchte man, dass der Namen der Autorin bzw. des Autors einer zitierten Quelle im Fließtext erscheint, wie das nachfolgende Textbeispiel zeigt. So stellte etwa \textcite{Chau81} das Mix-Netz vor und hat mit dieser und weiteren bahnbrechenden Arbeiten das neue Teilgebiet der Privacy Enhancing Technologies (PET) innerhalb der IT-Sicherheitsforschung etabliert.

Gelegentlich sieht man auch die Quellenangabe am Ende eines Satzes. Wenn diese Form der Quellenangabe genutzt wird, dann bitte die Quellenangabe \emph{vor} das Satzzeichen setzen, wie in diesem Satz \cite{XYZ}. Danach kann es im Absatz mit weiteren Aussagen weitergehen, die sich auf andere Quellen beziehen können.

Die Übernahme längerer paraphrasierter Passagen aus bereits veröffentlichten Texten ist eher unerwünscht. Gelegentlich sieht man die Quellenangabe bei solchen aus mehreren Sätzen bestehenden paraphrasierten Passagen am Ende des Absatzes, wie hier in diesem Absatz demonstriert. Dann sollten sich innerhalb dieses Absatzes nur Aussagen finden, die sich auf diese eine Quelle beziehen, die dann \emph{nach} das letzte Satzzeichen des Absatzes gestellt wird. \cite{XYZ}

Bei Übernahme längerer paraphrasierter Inhalte sollte anstelle der nach dem letzten Satzzeichen gestellten Quellenangabe besser der jeweilige Absatz bzw. der jeweilige Abschnitt zu Beginn im Fließtext explizit darauf hinweisen, dass sich die nachfolgenden Bemerkungen auf die Quelle \cite{XYZ} beziehen. Eine wörtliche Übernahme aus Fremdtexten ist aber auch dann nicht erlaubt, abgesehen von kurzen Zitaten, die stets in Gänsefüßchen zu setzen sind.

\section{Vor der Abgabe}

Vor der Abgabe sollten die Funktionen zur Rechtschreibprüfung und Silbentrennung genutzt werden. In LaTeX können dazu spezielle Entwicklungsumgebungen verwendet werden, die für jedes gängige Betriebssystem verfügbar sind. TeXworks\footnote{\url{http://www.tug.org/texworks/}, Abruf am 14.07.2023} ist beispielsweise ein kostenloser plattformunabhängiger LaTeX-Editor.

Zusätzlich lohnt es sich, den Text vor Abgabe von jemandem lesen zu lassen (Freund, Freundin, Bekannte, Haustier), damit er auch sprachlich noch einmal überprüft wurde. Zudem ist auf die korrekte Kommasetzung zu achten.

Bachelor-, Master-, Seminar- und Studienarbeiten können inzwischen zumeist rein elektronisch abgegeben werden. Sofern Arbeiten in gedruckter Form abgegeben werden müssen, sollten sie möglichst gebunden sein. Eine einfache Heissleim-, Kaltleim oder Klemmbindung (für ca. 3~EUR pro Exemplar aus dem Copyshop) genügt. Es sollte möglichst doppelseitig gedruckt werden, um Papier zu sparen. Die Umwelt dankt es. Schwarzweiß-Druck genügt in den meisten Fällen völlig.

Zusätzlich muss die Arbeit noch einmal als PDF-Datei per Mail an die Betreuerin oder den Betreuer geschickt werden. Falls Quellcodes und Programme erstellt wurden, sollte vor Abgabe mit der Betreuerin oder dem Betreuer besprochen werden, in welcher Weise diese abzugeben sind.

\chapter{Betreuung und Bewertung der Arbeit}

Für die Betreuung der Arbeit steht die bzw. der mit der Ausgabe der Aufgabenstellung genannte Betreuerin oder Betreuer zur Verfügung. Bitte nutzen Sie die Kontaktmöglichkeiten im Rahmen der Sprechstunden und nach Vereinbarung für regelmäßige Gespräche (mindestens etwa alle 2--3 Wochen). Sinnvollerweise sollte Sie jeweils darauf vorbereitet sein, einen kurzen mündlichen Bericht über den Stand der Bearbeitung zu geben. Während der Vorlesungszeit finden möglicherweise regelmäßige Treffen aller Bearbeiterinnen und Bearbeiter von Abschlussarbeiten statt, zu denen ggf. kurzfristig eingeladen wird. Jedes Gesprächsangebot sollte wahrgenommen werden!

\section{Schriftlicher Teil}

Folgende \textbf{Meilensteine} sollten bereits zu Beginn der Bearbeitung des Themas im Kalender vermerkt werden:

Bei \textbf{Arbeiten mit etwa 3-monatiger Bearbeitungszeit} soll \textbf{nach 1,5 Monaten} die \textbf{Abgabe eines ersten Textentwurfs} bei der Betreuerin bzw. beim Betreuer erfolgen. Wenn mit der Zweitbetreuerin bzw. dem Zweitbetreuer (soweit vorhanden) nichts anderes vereinbart ist, sollte ihr bzw. ihm zu diesem Zeitpunkt ein Zwischenbericht geliefert werden und ggf. nachgefragt werden werden, ob der aktuelle Textentwurf zur Kommentierung überlassen werden soll.

Bei \textbf{Arbeiten mit etwa 6-monatiger Bearbeitungszeit} soll \textbf{nach 2 Monaten} ein etwa 12-seitiger Textentwurf inkl. Gliederungsentwurf vorliegen und \textbf{nach weiteren 2 Monaten} ein erster vollständiger Textentwurf.

Die Textentwürfe werden von uns gelesen, kommentiert und zurückgegeben. Die Meilensteine dienen der Fortschrittskontrolle und sind für die endgültige Bewertung der Arbeit bedeutungslos; Fehler dürfen sorgenfrei gemacht werden.

\section{Bewertung}

Typische Kontrollfragen zur Beurteilung einer Arbeit sind:

\begin{itemize}
	\item Wurde die Fragestellung auf hohem Niveau bearbeitet?
	\item Handelt es sich um eine kreative Herangehensweise bzw. Lösung?
	\item Sind die Annahmen und getroffenen Voraussetzungen realistisch, oder wurden unzulässige Vereinfachungen vorgenommen?
	\item Sind alle Aussagen klar und verständlich formuliert?
	\item Wurde die Literatur zur Kenntnis genommen?
	\item Falls Programme entwickelt wurden: Sind die Quellcodes dokumentiert, die Module und Schnittstellen beschrieben? Gibt es eine Programmbeschreibung?
	\item Wie ist die äußere Form (Layout, Rechtschreibung, Grammatik)?
	\item Ist der Umfang angemessen?
\end{itemize}

George H. Heilmeier hat in einen nach ihm benannten Heilmeier-Katechismus\footnote{\url{https://www.darpa.mil/work-with-us/heilmeier-catechism}, Abruf am 03.11.2023} einige Fragen formuliert, die Orientierung sowohl bei der Erstellung einer Arbeit als auch bei ihrer Bewertung geben können. Sie werden hier in einer leicht angepassten deutschen Übersetzung\footnote{siehe auch \url{https://de.wikipedia.org/wiki/George_H._Heilmeier}, Abruf am 03.11.2023} wiedergegeben:

\begin{itemize}
	\item Was hast Du vor? Beschreibe Dein Vorhaben, möglichst ohne viele Fachbegriffe zu verwenden.
	\item Wie wird es bislang gemacht und was sind die Grenzen der derzeitigen Verfahren oder Methoden?
	\item Was ist neu an Deinem Ansatz und warum denkst Du, dass er erfolgreich sein wird?
	\item Wen kümmert's? Wenn Du erfolgreich bist, was für einen Unterschied wird Deine Lösung machen?
	\item Was sind die Risiken und was die Vorteile Deiner Lösung?
	\item Wie viel wird sie kosten und wie lange wird die Umsetzung dauern?
	\item Woran kann man im Verlauf des Projekts und zu dessen Ende den Erfolg messen?
\end{itemize}

Bei der Bewertung der schriftlichen Ausarbeitung wird von uns zumeist ein Punkteschema verwendet, das sich an \cite{faui2} orientiert, welches am Lehrstuhl für Informatik 2 (Programmiersysteme) der Friedrich-Alexander Universität Erlangen-Nürnberg entwickelt wurde. Eine gekürzte und angepasste Übernahme des Punkteschemas ist im Anhang enthalten.

\section{Referat}

Oft müssen die Ergebnisse der Arbeit in einem Referat vorgestellt werden. Generell gilt: Ein Referat soll die Zuhörerschaft gezielt informieren. Bei der Vorbereitung des Referats sollte deshalb Klarheit darüber bestehen, wieviele Zuhörerinnen und Zuhörer voraussichtlich teilnehmen werden, welches Vorwissen sie haben und mit welchen Erwartungen sie zu dem Referat gekommen sind.

Übersichtliche Folien sind für die bzw. den Vortragenden und die Zuhörerinnen und Zuhörer eine große Unterstützung. Die Folien sollten nummeriert sein, nicht mehr als 4--8 Stichpunkte enthalten, keinen Fließtext und aussagekräftige Abbildungen. Bei Farbfolien sollte man sich auf drei bis vier Farben beschränken, die durchgehend durch die Präsentation verwendet werden. Ansonsten wirken die Folien bunt und unruhig. Schriften ohne oder mit unauffälligen Serifen (z.\,B. Helvetica, Calibri oder Verdana) in 18--20pt eignen sich sehr gut für Vortragsfolien. Es existieren am Arbeitsbereich Templates für Folien, die möglichst verwendet werden sollten.

Als Daumenregel gilt: Folienanzahl $\approx$ Vortragszeit$\,/\,$3 Minuten.

Während des gesamten Vortrags sollte man ins Publikum schauen und nicht zur Wand oder in den Laptop.

Auch das Referat wird nach festgelegten Kriterien beurteilt, die dem Formular im Anhang entnommen werden können.

Ein Kolloquium zur Abschlussarbeit kann auch vor Abgabe der schriftlichen Ausarbeitung stattfinden. Der Vorteil ist, dass ggf. noch Tipps gegeben werden, die in die schriftliche Ausarbeitung einfließen können. Bitte sprechen Sie Ihre Betreuerin bzw. Ihren Betreuer an, wenn Sie Interesse an einem vorgezogenen Kolloquiumstermin haben.

\chapter{Schlussbemerkungen}

Im Internet sind zahlreiche Erfahrungsberichte von (renommierten) Wissenschaftlerinnen und Wissenschaftlern zu finden, die auch bei der Bearbeitung einer Seminar- oder Abschlussarbeit hilfreich sein können. Hier einige wenige Empfehlungen:

\begin{itemize}
	\item Randy Pausch Lecture: Time Management. \\ \url{http://www.youtube.com/watch?v=oTugjssqOT0}
	\item Richard Hamming: You and Your Research. \\ \url{http://www.cs.virginia.edu/~robins/YouAndYourResearch.html}
	\item Nick Feamster: Writing Tips for Academics. \\ \url{http://greatresearch.org/2013/10/11/storytelling-101-writing-tips-for-academics/}
\end{itemize}

Eine besondere Empfehlung ist der Duden-Ratgeber „Wie schreibt man wissenschaftliche Arbeiten?“ von Ulrike Pospiech \cite{Posp2012}, der Informationen und Beispiele zu allen wichtigen Themen bezüglich wissenschaftlicher Texte enthält.

Wissenschaftliches Arbeiten und Schreiben will gelernt sein. Dafür dienen während des Studiums u.a. die Seminararbeiten. Die Abschlussarbeit soll dann zeigen, welche methodischen und fachlichen Fähigkeiten während des Studiums erworben wurden.

Das Bearbeiten von wissenschaftlichen Fragestellungen während des Studiums schult zudem auch die Entschlussfähigkeit. Wenn Sie sich etwa zwischen zwei Darstellungsvarianten eines Problems entscheiden sollen, grübeln bitte Sie nicht zu lange, sonst landen Sie in einem Deadlock. Dieses Phänomen ist als das Buridansche Paradoxon (auch: Buridans Esel, Grasbüschelproblem) \cite{BuridansAss} bekannt.

Neben einem guten Zeitmanagement, Disziplin und Bereitschaft zur Literaturrecherche ist die Kommunikation mit der Betreuerin bzw. dem Betreuer ein Schlüssel zur erfolgreichen Bearbeitung des Themas.

% =============================Literaturverzeichnis=============================
\begin{raggedright}         % Schaltet Blocksatz ab, erzeugt ein stimmigeres
                            %  Schriftbild im Literaturverzeichnis.
  \printbibliography        % Falls Biblatex verwendet wird.
  \label{sec:literaturverzeichnis}
\end{raggedright}

% ===================================Anhang=====================================
\appendix
\setcounter{figure}{0}
\renewcommand\thetable{A.\arabic{figure}}
\setcounter{table}{0}
\renewcommand\thetable{A.\arabic{table}}

\chapter*{Punktesystem zur Beurteilung}

Bei der Bewertung der schriftlichen Ausarbeitung wird ein Punkteschema verwendet, das am Lehrstuhl für Informatik 2 (Programmiersysteme) der Friedrich-Alexander Universität Erlangen-Nürnberg entwickelt wurde. Die folgende Übersicht ist eine gekürzte und angepasste Übernahme von \cite{faui2}.

\section*{Allgemeine Hinweise}

Die Arbeit wird unter fünf Aspekten einzeln bewertet, die jedoch nicht gleichgewichtig sind. Das verschiedene Gewicht wird dadurch berücksichtigt, dass für die einzelnen Aspekte verschieden hohe Punktzahlen zur Verfügung stehen (siehe Tabelle~\ref{tab:punktzahl}).

\begin{table}[!h]%Tabelle soll hier (!h) erscheinen!
\begin{tabu}{lcrl}
	\toprule
	\multicolumn{3}{l}{Punktzahl} & Aspekt\\
	\midrule
	0 & -- & 6  & Schwierigkeitsgrad\\
	0 & -- & 8  & Schöpferische Originalität\\
	0 & -- & 10 & wissenschaftliche Arbeitstechnik\\
	0 & -- & 4  & Stil\\
	0 & -- & 3  & Äußere Form\\
	\midrule
	0 & -- & 31 & Summe\\
	%\bottomrule
\end{tabu}
\caption{Maximale Punktzahlen pro Aspekt}
\label{tab:punktzahl}
\end{table}

\section*{Notenstufen}

Die Note wird in folgender Weise festgesetzt:
\begin{enumerate}
	\item Arbeiten, bei denen für wissenschaftliche Arbeitstechnik weniger als 4~Punkte oder für die wissenschaftliche Arbeitstechnik, den Stil und die Form zusammen weniger als 8~Punkte vergeben wurden, erhalten die Note~5 (nicht ausreichend, nicht bestanden).
	\item Alle anderen Arbeiten werden nach Tabelle~\ref{tab:noten} benotet.
\end{enumerate}

\begin{table}
\begin{tabu}{lp{6cm}}
	\toprule
	Punktzahl  & Note \\
	\midrule
	31--29 &   1,0 \quad  sehr gut\\
	28--27 &   1,3\\
	\midrule
	26--25 &   1,7\\
	24--23 &   2,0 \quad  gut\\
	22--21 &   2,3\\
	\midrule
	20--19 &   2,7 \\
	18--17 &   3,0 \quad  befriedigend\\
	16--15 &   3,3\\
	\midrule
	14--13 &   3,7\\
	12--11 &   4,0 \quad  ausreichend\\
	\bottomrule
\end{tabu}
\caption{Punkte- und Notenverteilung}
\label{tab:noten}
\end{table}

\section*{1. Schwierigkeitsgrad (0--6)}

Bei der Beurteilung des Schwierigkeitsgrades ist davon auszugehen, ob die Problemstellung mit der durchschnittlichen Ausgangsqualifikation der Bearbeitungsgruppe gelöst werden kann (4~Punkte). Die Beurteilung des Schwierigkeitsgrades einer Arbeit kann erst nach Abschluss erfolgen und umfasst die Prüfung, ob die vorgelegte Fassung die genannten Merkmale auch tatsächlich enthält.

\section*{2. Schöpferische Originalität (0--8)}

Bei der Beurteilung der schöpferischen Originalität ist nicht nur davon auszugehen, inwieweit die Bearbeiterin bzw. der Bearbeiter der Anleitung und Führung durch die Betreuerin bzw. den Betreuer bedarf. Es ist vielmehr selbstverständlich, dass die Bearbeiterin bzw. der Bearbeiter Initiative entwickelt, d.h. aus eigenem Antrieb Schwierigkeiten aufgreift und mit der Betreuerin bzw. dem Betreuer diskutiert (4~Punkte).

\section*{3. Wissenschaftliche Arbeitstechnik (0--10)}

Bei der Beurteilung der wissenschaftlichen Arbeitstechnik ist nicht nur vom Grad der Fehlerfreiheit (formale Richtigkeit der Aussagen und eventueller Programme) auszugehen, die vielmehr als selbstverständlich vorausgesetzt werden muss. Daneben fällt sehr stark das Ausmaß der Selbstkontrolle ins Gewicht, das sich bei formalen Aussagen in der Beweisgründlichkeit, bei Programmen in ausführlichen Tests zeigt. Bezüglich der Programmrichtigkeit darf davon ausgegangen werden, dass bei hinreichend modularem Programmaufbau eine durchdachte (Begründung!) Menge von Testprogrammen genügt (5~Punkte).

\section*{4. Stil (0--4)}

Bei der Beurteilung des Stils ist von der sprachlichen Ausdrucksfähigkeit auszugehen, die sich der Leserin bzw. dem Leser in der vorgelegten Arbeit bietet. Diese zeigt sich insbesondere in der Klarheit und Kürze des Ausdrucks: Auch schwierige Probleme müssen verständlich dargelegt, triviale Zusammenhänge nicht hinter einem formalen Apparat verborgen sein. Die Gedankenführung muss eindeutig sein (2~Punkte).

\section*{5. Äußere Form (0--3)}

Bei der Beurteilung der äußeren Form fällt neben der Sorgfalt der Ausführung, insbesondere der Zeichnungen und Tabellen, die Klarheit der Gliederung und des
Inhaltsverzeichnisses ins Gewicht (2~Punkte).

% ===========================Präsentationsbewertung============================
\chapter*{Formular Präsentationsbewertung}
\begin{sideways}
	\begin{tabu} to 1.2\textwidth {|l|X|X|X|X|X|X|X|X|}
	\multicolumn{6}{l}{\textbf{Präsentationsbewertung}} \\
	\multicolumn{6}{l}{} \\
	\hline
	Datum                                    & & & & & & & \\ \hline
	Name                                     & & & & & & & \\ \hline
	Thema                                    & & & & & & & \\ \hline
	\multicolumn{6}{l}{} \\
	\multicolumn{6}{l}{\textbf{Stil}} \\
	\hline
	Sicheres Auftreten                       & & & & & & & \\ \hline
	Kontakt zum Zuhörer                      & & & & & & & \\ \hline
	Deutliche Sprechweise                    & & & & & & & \\ \hline
	Angemessenes Tempo                       & & & & & & & \\ \hline
	Freies Sprechen                          & & & & & & & \\ \hline
	Einhalten der Zeit                       & & & & & & & \\ \hline
	\multicolumn{6}{l}{} \\
	\multicolumn{6}{l}{\textbf{Inhalt}} \\
	\hline
	Verständlichkeit des Inhalts             & & & & & & & \\ \hline
	Prägnanz                                 & & & & & & & \\ \hline
	Konzept/Gliederung                       & & & & & & & \\ \hline
	Beispiele                                & & & & & & & \\ \hline
	Angemessene Detailtiefe                  & & & & & & & \\ \hline
	\multicolumn{6}{l}{} \\
	\multicolumn{6}{l}{\textbf{Folien/Demo}} \\
	\hline
	Vertrautheit mit Folien/Demo             & & & & & & & \\ \hline
	Verständlichkeit der Folien/Demo         & & & & & & & \\ \hline
	Qualität von Abbildungen                 & & & & & & & \\ \hline
	Nachvollziehbarkeit der Demo             & & & & & & & \\ \hline
	\multicolumn{6}{l}{} \\
	\hline
	Subjektiver Gesamteindruck               & & & & & & & \\ \hline
	\end{tabu}
\end{sideways}

% ===========================Formular Rump Session======================
% \chapter*{Formular „Rump Session“}
\newpage
\thispagestyle{empty}
\vspace*{-3.2cm}
\begin{center}
	Formular „Rump Session“
\end{center}
\vspace*{-0.2cm}

\newsavebox{\rumpsessBox}
\savebox{\rumpsessBox}{%
	%\begin{sffamily}
	\begin{tiny}
	\begin{tabu} to 10cm {lcccc} %\hline
	\multicolumn{5}{l}{} \\[3mm]
	Vortragsnummer/Titel: \dotfill                                               & \tiny ja & \multicolumn{2}{l}{\dotfill} & \tiny nein \\[3pt]
	\textbf{Struktur:} Ich konnte einen „roten Faden“ erkennen.                         & $\Box$ & $\Box$ & $\Box$ & $\Box$ \\[1pt]
	\textbf{Auftreten:} Die/der Vortragende ist selbstsicher aufgetreten.               & $\Box$ & $\Box$ & $\Box$ & $\Box$ \\[1pt]
	\textbf{Kompetenz:} Die/der Vortragende kennt sich mit dem Thema aus.               & $\Box$ & $\Box$ & $\Box$ & $\Box$ \\[1pt]
	\textbf{Didaktik:} Ich habe alles verstanden, was gesagt wurde.                     & $\Box$ & $\Box$ & $\Box$ & $\Box$ \\[1pt]
	\textbf{Niveau:} Ich kann die Kernideen des Vortrag mit eigenen Worten wiedergeben. & $\Box$ & $\Box$ & $\Box$ & $\Box$ \\[1pt]
	\multicolumn{5}{l}{\textbf{Anmerkungen:}} \\[24mm] %\hline
	\end{tabu}
	\end{tiny}
	%\end{sffamily}
}

\hspace*{-2cm}\rule{3mm}{0.5pt}\rule{95mm}{0pt}\rule{0.5pt}{3mm}\rule{95mm}{0pt}\rule{3mm}{0.5pt}\\
\hspace*{-2cm}\usebox{\rumpsessBox}\usebox{\rumpsessBox}\\
\hspace*{-2cm}\rule{3mm}{0.5pt}\rule{95mm}{0pt}\rule{0.5pt}{3mm}\rule{95mm}{0pt}\rule{3mm}{0.5pt}\\
\hspace*{-2cm}\usebox{\rumpsessBox}\usebox{\rumpsessBox}\\
\hspace*{-2cm}\rule{3mm}{0.5pt}\rule{95mm}{0pt}\rule{0.5pt}{3mm}\rule{95mm}{0pt}\rule{3mm}{0.5pt}\\
\hspace*{-2cm}\usebox{\rumpsessBox}\usebox{\rumpsessBox}\\
\hspace*{-2cm}\rule{3mm}{0.5pt}\rule{95mm}{0pt}\rule{0.5pt}{3mm}\rule{95mm}{0pt}\rule{3mm}{0.5pt}\\
\hspace*{-2cm}\usebox{\rumpsessBox}\usebox{\rumpsessBox}\\
\hspace*{-2cm}\rule{3mm}{0.5pt}\rule{95mm}{0pt}\rule{0.5pt}{3mm}\rule{95mm}{0pt}\rule{3mm}{0.5pt}\\
\hspace*{-2cm}\usebox{\rumpsessBox}\usebox{\rumpsessBox}\\
\hspace*{-2cm}\rule{3mm}{0.5pt}\rule{95mm}{0pt}\rule{0.5pt}{3mm}\rule{95mm}{0pt}\rule{3mm}{0.5pt}\\

% ===========================Selbstständigkeitserklärung======================
\chapter*{Eidesstattliche Versicherung} % war: Selbständigkeitserklärung
\vspace{1cm}

\todo[noline]{Bitte verwenden Sie hier in jedem Fall die offizielle von der Prüfungsbehörde vorgegebene Formulierung der Selbständigkeitserklärung.}
%
Hiermit versichere ich an Eides statt, dass ich die vorliegende Arbeit selbstständig verfasst und keine anderen als die angegebenen Hilfsmittel – insbesondere keine im Quellenverzeichnis nicht benannten Internet-Quellen – benutzt habe. Alle Stellen, die wörtlich oder sinngemäß aus Veröffentlichungen entnommen wurden, sind als solche kenntlich gemacht. Ich versichere weiterhin, dass ich die Arbeit vorher nicht in einem anderen Prüfungsverfahren eingereicht habe und die eingereichte schriftliche Fassung der auf dem elektronischen Speichermedium entspricht.

Ggf. streichen: Ich bin damit einverstanden, dass meine Abschlussarbeit in den Bestand der Fachbereichsbibliothek eingestellt wird.

\makeatletter
Hamburg, den {\@date}
\makeatother

\vspace{2cm}
\rule{6cm}{0.25pt}\\
\makeatletter
{\@author} \par
\makeatother

% ================================Deckblatt-Muster Abschlussarbeit==============================
\newpage
\thispagestyle{empty}
% \addcontentsline{toc}{chapter}{Muster des Deckblatts}
%
% In der Ausarbeitung bitte den Bereich ab \begin{titlepage} bis \end{titlepage}
%  vorne durch den nachfolgenden Codeabschnitt ersetzen:
% ====> Von hier ...
\begin{titlepage}% {{{
% \includegraphics[width=6.8cm]{../pic/up-uhh-logo-u-2010-u-farbe-u-rgb.pdf}
\mbox{\parbox[t][1.75cm][b]{2.2cm}{\uhhlogo}}
\begin{center}\Large
	\vfill
	Masterarbeit
	\vfill
	\makeatletter
	{\Large\textsf{\textbf{\@title}}\par}
	\makeatother
	\vfill
	vorgelegt von
	\par\bigskip
	\makeatletter
	{\@author} \par
	\makeatother
	Matrikelnummer 1234567 \par
	Studiengang Informatik
	\vfill
	MIN-Fakultät \par
	Fachbereich Informatik
	\vfill
	\makeatletter
	eingereicht am {\@date}
	\makeatother
	\vfill
	Betreuerin: Erika Musterfrau, M.\,Sc. \todo{Todos im Text und Fragen an den Betreuer sind in dieser Form dargestellt}\par
	Erstgutachter: Prof. Dr.-Ing. Hannes Federrath \par
	Zweitgutachter: N.N.
\end{center}
\ifoptionfinal{}{
\begin{tikzpicture}[remember picture, overlay]
	\node[draw, red, font=\ttfamily\bfseries\Large, xshift=30mm, yshift=238mm, rotate=340, text centered, text width=6cm, very thick, rounded corners=4mm] at (current page.south) {Entwurf vom \today};
\end{tikzpicture}
% ====> Delete me
\begin{tikzpicture}[overlay]
	\node[draw, blue, font=\sffamily\Large, xshift=0mm, yshift=242mm, rotate=0, text centered, rounded corners=1mm] at (current page.south) {Muster des Deckblatts für Abschlussarbeiten};
\end{tikzpicture}
% <==== /Delete me
}
\end{titlepage}% }}}
% <==== ... bis hierher.

% ================================Deckblatt-Muster Seminararbeit==============================
\newpage
\thispagestyle{empty}
% \addcontentsline{toc}{chapter}{Muster des Deckblatts}
%
% In der Ausarbeitung bitte den Bereich ab \begin{titlepage} bis \end{titlepage}
%  vorne durch den nachfolgenden Codeabschnitt ersetzen:
% ====> Von hier ...
\begin{titlepage}% {{{
% \includegraphics[width=6.8cm]{../pic/up-uhh-logo-u-2010-u-farbe-u-rgb.pdf}
\mbox{\parbox[t][1.75cm][b]{2.2cm}{\uhhlogo}}
\begin{center}\Large
	\vfill
	Seminararbeit
	\vfill
	\makeatletter
	{\Large\textsf{\textbf{\@title}}\par}
	\makeatother
	\vfill
	vorgelegt von
	\par\bigskip
	\begin{tabu}{p{0.5\textwidth}p{0.5\textwidth}}
	\centering Martina Mustermann     & \centering Max Mustermann \\[1ex]
	\centering Matrikelnummer 1234567 & \centering Matrikelnummer 7654321 \\
	\centering Studiengang Informatik & \centering Studiengang Informatik \\
	\end{tabu}
	\vfill
	MIN-Fakultät \par
	Fachbereich Informatik
	\vfill
	\makeatletter
	eingereicht am {\@date}
	\makeatother
	\vfill
	Betreuer: Heinz Mustermann, M.\,Sc.
\end{center}
\ifoptionfinal{}{
\begin{tikzpicture}[remember picture, overlay]
	\node[draw, red, font=\ttfamily\bfseries\Large, xshift=30mm, yshift=238mm, rotate=340, text centered, text width=6cm, very thick, rounded corners=4mm] at (current page.south) {Entwurf vom \today};
\end{tikzpicture}
% ====> Delete me
\begin{tikzpicture}[overlay]
	\node[draw, blue, font=\sffamily\Large, xshift=0mm, yshift=242mm, rotate=0, text centered, rounded corners=1mm] at (current page.south) {Muster des Deckblatts für Seminararbeiten};
\end{tikzpicture}
% <==== /Delete me
}
\end{titlepage}% }}}
% <==== ... bis hierher.

% ================================Literaturliste-Muster==============================
\newpage
\thispagestyle{empty}
\label{sec:literaturliste}
\par\textbf{\textsf{Thema:}} Privacy Enhancing Technologies zum Schutz von Kommunikationsbeziehungen
\par\textbf{\textsf{Bearbeiter:}} Eva Musterfrau, Heinz Mustermann
\par\textbf{\textsf{Datum:}} \today
\bigskip
% ====> Delete me
\begin{tikzpicture}[overlay]
	\node[draw, blue, font=\sffamily\Large, xshift=70mm, yshift=0mm, rounded corners=1mm]{Muster der Literaturliste};
\end{tikzpicture}
% <==== /Delete me
\par\textbf{\Large\textsf{Literaturliste}}

% ----- Nachfolgend eine händisch gesetzte Literaturliste, die sich exakt an die Syntax im Abschnitt \ref{sec:literaturhowto} hält. Wir nutzen diese aber hier nicht, sondern lassen BibLaTeX die Einträge formatieren.
\iffalse
David Chaum: Untraceable Electronic Mail, Return Addresses, and Digital Pseudonyms. Communications of the ACM 24/2 (1981) 84--88.

David Chaum: The Dining Cryptographers Problem: Unconditional Sender and Recipient Untraceability. Journal of Cryptology 1/1 (1988) 65--75.

David Goldschlag, Michael Reed, Paul Syverson: Onion Routing for Anonymous and Private Internet Connections. Communications of the ACM 42/2 (1999) 39--41.

Andreas Pfitzmann: Diensteintegrierende Kommunikationsnetze mit teilnehmerüberprüfbarem Datenschutz. IFB 234, Springer-Verlag, Berlin 1990.

Wei Wang, Mehul Motani, Vikram Srinivasan: Dependent link padding algorithms for low latency anonymity systems. Proc. 15th ACM conference on Computer and communications security. ACM, 2008, 323--332.
\fi

% ----- Nachfolgend die Ausgabe unter Verwendung von BibLaTeX. Die Formatierung übernimmt BibLaTeX. Dadurch wird es zu Abweichungen von der vorgegebenen Syntax kommen. Dies ist tolerabel, da es i.W. auf Einheitlichkeit ankommt, nicht auf eine dogmatische Einhaltung der Syntax.
\fullcite{Chau81}

\fullcite{Chau88}

\fullcite{GoRS99}

\fullcite{Pfit90}

\fullcite{WaMS2008}

% ================================Todo list==============================
\listoftodos
% \todototoc

\end{document}

### 2024-10-21 SbD/exercises.tex
Added exams to SbD exercises.

@@
 \section[2023]{2nd exam}
-% Aufgabe 1
-% Betrachten Sie die die Betriebsarten für Blockchiffren ECB, CBC und CTR.
 
-% a) Bei welchen Arten braucht man eine invertierbare Verschlüsselungsfunktion, bei welchen nicht? Begründen Sie?
+\begin{exercise}{Betriebsarten}
+  Betrachten Sie die Betriebsarten für Blockchiffren ECB, CBC und CTR.
 
-% b) Bei welchen Arten lässt sich die Verschlüsselung parallelisieren?
+  \begin{enumerate}
+    \item Bei welchen Arten braucht man eine invertierbare Verschlüsselungsfunktion, bei welchen nicht? Begründen Sie.
+    \item Bei welchen Arten lässt sich die Verschlüsselung parallelisieren?
+    \item Im zweiten Block gibt es ein Bit mit einem additiven Fehler. Was kann noch entschlüsselt werden?
+    \item Ein Bit fehlt im zweiten Block (nicht bekannt, wo das Bit fehlt). Was kann noch entschlüsselt werden?
+    \item Ein ganzer Block fehlt. Was kann noch entschlüsselt werden?
+    \item Zwei verschiedene Klartexte wurden mit CTR mit dem gleichen IV verschlüsselt.
+Zeigen Sie, dass ein Angreifer den zweiten Klartext entschlüsseln kann, wenn dieser den ersten Klartext und die beiden Schlüsseltexte hat.
+  \end{enumerate}
 
-% c) Im zweiten Block gibt es ein Bit mit einem additiven Fehler. Was kann noch entschlüsselt werden?
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
 
-% d) Ein Bit fehlt im zweiten Block (nicht bekannt, wo das Bit fehlt). Was kann noch entschlüsselt werden?
+\begin{exercise}{Elliptische Kurven}
+  \begin{enumerate}
+    \item Auf welcher mathematischen Schwierigkeit basiert der ECC?
+    \item Berechnen Sie die Punkte auf der folgenden elliptischen Kurve: $a = 6$, $b = 1$, $p = 11$ (allgemeine Formel war gegeben).
+    \item Nennen Sie zwei Vorteile, die ECC gegenüber sicherheitsmäßig vergleichbaren Verfahren hat.
+  \end{enumerate}
 
-% e) Ein ganzer Block fehlt. Was kann noch entschlüsselt werden?
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
 
-% f) Zwei verschiedene Klartexte wurden mit CTR mit dem gleichen IV verschlüsselt.
-Zeigen Sie, dass ein Angreifer den zweiten Klartext entschlüsseln kann wenn dieser den ersten Klartext und die beiden Schlüsseltexte hat.
+\begin{exercise}{Zertifikate}
+  Das Zertifigierungsmodell ist transitiv anzunehmen.
 
-% Aufgabe 2 ECC
-%  a) Auf welcher mathematischen Schwierigkeit basiert der ECC?
+  \begin{enumerate}
+    \item Erklären und grenzen Sie die Begriffe Signatur und Zertifikat voneinander ab.
+    \item N7 möchte N3 eine Nachricht schicken. Welchen CAs muss N7 vertrauen?
+    \item N7 möchte zusätzlich N5 eine Nachricht schicken. Welchen CAs muss N7 vertrauen, so dass die Anzahl an CAs minimal ist?
+    \item CA2 fällt aus. Wie können wir die Kommunikation aus b) trotzdem realisieren (Was \& Wie)? Was können wir für N1 tun?
+    \item Trust on First Use: Keine CAs, Schlüssel wird bei erster Kommunikation an Kommunikationspartner geschickt, sobald und bei späteren Kommunikationen überprüft.
+Wenn Kommunikation und es gibt schon anderen Schlüssel, deny.
+Nennen und erklären Sie zwei Vorteile und zwei Nachteile.
+  \end{enumerate}
 
-%  b) Berrechne die Punkte auf der folgenden Eliptischen Kurve: a = 6, b=1, p=11 (allgemeine Formel war gegeben)
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
 
-%  c) Nennen Sie zwei Vorteile die ECC hat, gegenüber sicherheitsmäßig vergleichbaren Verfahren
+\begin{exercise}{GSM}
+  Zur Erreichung von Forward Secrecy erzeugt Alice auf ihrem Server mit dem RSA-Verfahren einen privaten Schlüssel $d$ sowie den dazu passenden öffentlichen Schlüssel $c$, den sie persönlich an ihre Nutzerinnen und Nutzer verteilt.
+Zusätzlich erzeugt der Server von Alice bei jedem Verbindungsaufbau ein neues kurzlebiges (engl.
+ephemeral) RSA-Schlüsselpaar, das aus einem öffentlichen Schlüssel $c_e$ und einem geheimen Schlüssel $d_e$ besteht.
+Der Server signiert mit dem privaten Schlüssel $d$ den neuen öffentlichen Schlüssel $c_e$ und sendet diesen zusammen mit der Signatur über die aufgebaute Verbindung an die Kundin.
+Die Kundin prüft anschließend die Signatur mithilfe des zuvor von Alice erhaltenen öffentlichen Schlüssels $c$.
+Ist die Signatur korrekt, nutzt die Kundin $c_e$, um einen neuen, zufällig generierten symmetrischen AES-Schlüssel $k$ zu verschlüsseln.
+Die weitere Kommunikation in dieser Verbindung wird symmetrisch mit $k$ verschlüsselt.
+Nach Verbindungsende löscht Alices Server $d_e$, $c_e$ und $k$.
 
-% Aufgabe 3
-% Das Zertifizierungsmodell war transitiv anzunehmen.
-
-% a) Die Begriffe Signatur und Zertifikat erklären und voneinander abgrenzen
-
-% b) N7 möchte N3 Nachricht schicken, welchen CAs muss N7 vertrauen?
-
-% c) N7 möchte zusätzlich N5 schicken, welchen CAs muss N7 vertrauen, wähle so, dass die Zahl and CAs minimal ist.
-
-% d) CA2 fällt aus, wie können wir die Kommunikation aus b) trotzdem realisieren (Was & Wie). Was können wir für N1 tun?
-
-% e) Trust on First Use: Keine CAs, Schlüssel wird bei erster Kommunikation an Kommunikationspartner geschickt, sobald und bei späteren Kommunikationen überprüft.
-Wenn Kommunikation und es gibt schon anderen Schlüssel, deny.
-2 Vorteile, 2 Nachteile nenne und erklären.
-
-% Aufgabe 4
-% (Aufgabe war sehr ähnlich zu einer Übungsaufgabe)
-
-% Zur Erreichung von Forward Secrecy erzeugt Alice auf ihrem Server mit dem RSA-Verfahren einen privaten Schlüssel d sowie den dazu passenden öffentlichen Schlüssel c, den sie persönlich an ihre Nutzerinnen und Nutzer verteilt.
-Zusätzlich erzeugt der Server von Alice bei jedem Verbindungsaufbau ein neues kurzlebiges (engl.
-ephemeral) RSA-Schlüsselpaar, das aus einem öffentlichen Schlüssel ce und einem geheimen Schlüssel de besteht.
-Der Server signiert mit dem privaten Schlüssel d den neuen öffentlichen Schlüssel ce und sendet diesen zusammen mit der Signatur über die aufgebaute Verbindung an die Kundin.
-Die Kundin prüft anschließend die Signatur mithilfe des zuvor von Alice erhaltenen öffentlichen Schlüssels c.
-
-% Ist die Signatur korrekt, nutzt die Kundin ce, um einen neuen, zufällig generierten symmetrischen AES-Schlüssel k zu verschlüsseln.
-Die weitere Kommunikation in dieser Verbindung wird symmetrisch mit k verschlüsselt.
-Nach Verbindungsende löscht Alices Server de, ce und k.
-
-% a) Welche Verbindungen könnten von einem sehr starken Angreifer wie einem Nachrichtendienst noch entschlüsselt werden, wenn Zugriff auf den Server bekommt? Begründen Sie Ihre Antwort.
-
-% b) Skizzieren Sie einen Angriff, der möglich wäre, wenn die Kundin die Signatur nicht prüft. Erläutern SIe die einzelnen nötigen Schritte.
-
-% c) Alice ist das Verfahren zu langsam.
-Sie nimmt nun folgende Änderung vor: Statt jedes mal bei der Erzeugung von ce und de neue Primzahlen p und q für den Modulus Ne zu generieren, verwendet sie immer die gleichen Primzahlen.
-Nun wählt sie lediglich den öffentlichen Exponenten ce neu und berechnet einen dazugehörigen geheimen Exponenten de.
+  \begin{enumerate}
+    \item Welche Verbindungen könnten von einem sehr starken Angreifer wie einem Nachrichtendienst noch entschlüsselt werden, wenn dieser Zugriff auf den Server bekommt? Begründen Sie Ihre Antwort.
+    \item Skizzieren Sie einen Angriff, der möglich wäre, wenn die Kundin die Signatur nicht prüft. Erläutern Sie die einzelnen nötigen Schritte.
+    \item Alice ist das Verfahren zu langsam.
+Sie nimmt nun folgende Änderung vor: Statt jedes Mal bei der Erzeugung von $c_e$ und $d_e$ neue Primzahlen $p$ und $q$ für den Modulus $N_e$ zu generieren, verwendet sie immer die gleichen Primzahlen.
+Nun wählt sie lediglich den öffentlichen Exponenten $c_e$ neu und berechnet einen dazugehörigen geheimen Exponenten $d_e$.
 Warum ist dies keine gute Idee?
 Zeigen Sie, wie sich dieses geänderte Verfahren angreifen lässt, um aufgezeichnete Verbindungen zu entschlüsseln.
+  \end{enumerate}
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
 
 \section[2022]{1st exam}
-% Aufgabe 1 - Klassische Chiffren (11 Punkte)
 
-% (3 Punkte) Erläutern Sie zwei Ursachen, die dafür sorgen, dass fast alle kryptographischen Algorithmen mit der Zeit nicht mehr Standard sind.
-
-% (3 Punkte) Sie möchten einen Text so verschlüsseln, dass er auch in 98 Jahren nicht ohne Kenntnis des von Ihnen versteckten Schlüssels entschlüsselt werden kann.
+\begin{exercise}{Kryptographie}
+ \begin{enumerate}
+  \item Erläutern Sie zwei Ursachen, die dafür sorgen, dass fast alle kryptographischen Algorithmen mit der Zeit nicht mehr Standard sind.
+  \item Sie möchten einen Text so verschlüsseln, dass er auch in 98 Jahren nicht ohne Kenntnis des von Ihnen versteckten Schlüssels entschlüsselt werden kann.
 Welches Verfahren wählen Sie dazu?
-Begründen Sie Ihre Entscheidung.
-Warum wird das Verfahren heutzutage nicht überall benutzt?
-
-% (5 Punkte) Gegeben wurde eine Chiffre, welche Buchstaben A-N mit Buchstaben N-Z vertauscht.
+Begründen Sie Ihre Entscheidung.
+Warum wird das Verfahren heutzutage nicht überall benutzt?
+  \item Gegeben eine Chiffre, welche Buchstaben A-N mit Buchstaben N-Z vertauscht.
 Und gleichzeitig muss die Länge der Nachricht immer gleich sein.
 Bewerten Sie die Sicherheit der Chiffre und die beiden Eigenschaften (Polyalphabetische Verschlüsselung nur mit Vertauschen und die gleiche Länge der Nachricht).
+ \end{enumerate}
 
-% Aufgabe 2 - Elliptische Kurven (13 Punkte)
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
 
-% (8 Punkte) Berechnen Sie für die folgenden Parameter alle Punkte auf der Kurve: a = 2, b = 3, p = 7
+\begin{exercise}{Elliptische Kurven}
+  \begin{enumerate}
+    \item Berechnen Sie für die folgenden Parameter alle Punkte auf der Kurve: $a = 2$, $b = 3$, $p = 7$.
+    \item Elgamal mit elliptischen Kurven wurde vorgestellt. Zeigen Sie, warum dieses Verfahren mit beliebigen Parametern funktioniert.
+    \item Vergleichen Sie den Elgamal aus der Vorlesung mit dem gegebenen. Auf welchem Prinzip beruht die Sicherheit? Denken Sie, dass er sicherer ist? Warum?
+  \end{enumerate}
 
-% (2 Punkte) Elgamal mit elliptischen Kurven wurde vorgestellt. Zeigen Sie, warum dieses Verfahren mit beliebigen Parametern funktioniert.
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
 
-% (3 Punkte) Vergleichen Sie den Elgamal aus der VL mit dem gegebenen. Auf welchem Prinzip beruht die Sicherheit? Denken Sie, dass er sicherer ist? Warum?
+\begin{exercise}{Zertifikate}
+  \begin{enumerate}
+    \item Es wurden 3 verschiedene Möglichkeiten gegeben, einen Schlüssel von verschiedenen Schlüsselpartnern (S, V, C, T) zu kombinieren.
+Alle Teilschlüssel werden aneinander gehängt l/4.
+3 der Teilschlüssel wird aneinander gehängt und der letzte Teilnehmer bestimmt die Reihenfolge.
+Alle Schlüssel werden XOR berechnet.
+Bewerten Sie die Sicherheit der Verfahren.
+    \item Gegeben ist ein Zertifikatsbaum mit 5 Teilnehmern.
+$E \rightarrow D$, $E \rightarrow B$, $E \rightarrow A$, $E \rightarrow C$, $C \rightarrow B$, $A \rightarrow B$, $G \rightarrow \emptyset$.
+Betrachten Sie die Zertifizierungsrelation.
+$G$ möchte jetzt zusätzlich eine verschlüsselte Nachricht an $B$ senden.
+Welcher Zertifizierungsstelle bzw.
+welchen Zertifizierungsstellen muss $G$ mindestens vertrauen, um das Zertifikat von $B$ prüfen zu können?
+    \item Wählen Sie Ihre Lösung so, dass die Anzahl an Zertifizierungsstellen, denen $N4$ (explizit) vertrauen muss, minimal ist.
+    \item In der Zertifizierungsrelation aus Teilaufgabe c) wird die Zertifizierungsstelle $C$ kompromittiert und fällt deshalb aus.
+$G$ möchte eine verschlüsselte Nachricht an $C$ senden.
+Welche Maßnahme ermöglicht es $G$, das Zertifikat von $B$ trotz des Ausfalls von $CA4$ zu überprüfen?
+  \end{enumerate}
 
-% Aufgabe 3 - Public Key Infrastructure (8 Punkte)
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
 
-% (3 Punkte) Es wurden 3 verschiedene Möglichkeiten gegeben, einen Schlüssel von verschiedenen Schlüsselpartner (S, V, C, T) zu kombinieren.
+\begin{exercise}{GSM}
+  \begin{enumerate}
+    \item Sagen Sie zu folgenden Abkürzungen welche Funktion sie im GSM erfüllen: $Ki$, $IMSI$, $RAND$, $SRES$, $LAI$, $MSISDN$, $Kc$.
+    \item Welche dieser Abkürzungen ist auch noch im UMTS enthalten bzw. mit ähnlicher Funktion aber anderen Namen?
+  \end{enumerate}
 
-% Alle Teilschlüssel werden aneinander gehängt l/4.
-
-% 3 der Teilschlüssel wird aneinander gehängt und der letzte Teilnehmer bestimmt die Reihenfolge.
-
-% Alle Schlüssel werden XOR berechnet.
-
-% Bewerten Sie die Sicherheit der Verfahren.
-
-% (2 Punkte) Gegeben ist ein Zertifikatsbaum mit 5 Teilnehmern.
-
-% E-> D
-
-% E-> B
-
-% E-> A
-
-% E-> C
-
-% C-> B
-
-% A-> B
-
-% G -> Null
-
-% Betrachten Sie die Zertifizierung Relation.
-G möchte jetzt zusätzlich eine verschlüsselte Nachricht an B senden.
-Welcher Zertifizierungsstelle bzw.
-welchen Zertifizierungsstellen muss G mindestens vertrauen, um das Zertifikat von B prüfen zu können?
-
-% (1 Punkt) Wählen Sie Ihre Lösung so, dass die Anzahl an Zertifizierungsstellen, denen N4 (explizit) vertrauen muss, minimal ist.
-
-% (2 Punkte) In der Zertifizierung Relation aus Teilaufgabe c) wird die Zertifizierungsstelle C kompromittiert und fällt deshalb aus.
-G möchte eine verschlüsselte Nachricht an C senden.
-Welche Maßnahme ermöglicht es G, das Zertifikat von B trotz des Ausfalls von CA4 zu überprüfen?
-
-% Aufgabe 4 - GSM (8 Punkte)
-
-% (7 Punkte) Sagen Sie zu folgenden Abkürzungen welche Funktion sie im GSM erfüllen:
-
-% Ki, IMSI, RAND, SRES, LAI, MSISDN, Kc
-
-% (1 Punkt) Welche dieser Abkürzungen ist auch noch im UMTS enthalten bzw. mit ähnlicher Funktion aber anderen Namen?
-
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
 \end{document}

### 2024-10-21 NLP/exercises.tex
Added NLP exam scaffold.
Added exams to NLP exercises.

@@
 
-\sheet[2024]{1st exam}
-% 1 Question: Foundations (30 Points)
-% This section has various tasks to assess your understanding of fundamental concepts, principles and
-% terminology related to requirements engineering and software architecture.
-% 1.1
-% (3 Points) Name one activity in requirements engineering and provide a brief explanation.
-% 1.2
-% You have learned about three types of qualtity scenarios including use-case, growth and exploratory
-% scenarios. Explain each of them by an example: Use-case scenarios, Growth scenarios, Exploratory scenarios.
+\sheet{2023}{1st exam}
+\begin{exercise}{Tokenization}
+  Give at least 3 examples of ambiguity in tokenization.
 
-% 1.3 Give three reasons why software architecture is important. Explain each in one sentence.
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
 
-% 1.4 Define the term architectural tactic in the context of software architecture. In addition
-% provide an example of an achtitecture tactic and explain how it can be used to address a specific
-% quality attribute in a software system.
+\begin{exercise}{Parse trees}
+  Given the sentence "John shot the man with the gun" and a grammar: Draw all existing parse trees. Which types of syntactic ambiguity occur here and where?
 
-% 1.5
-% (3 Points) Which of the following pairs of qualities are typically in conflict with each other (i.e. are
-% trade-offs to each other)? Explain each decision in one sentence.
-% Modularity - Time behavior
-% ⃝ typically in conflict
-% ⃝ typically not in conflict
-% Explanation:
-% Usability - Security
-% ⃝ typically in conflict
-% ⃝ typically not in conflict
-% Explanation:
-% 3
-% 1.6
-% (3 Points) What is the purpose of conformance/compliance checking? In addition, briefly explain
-% strucural and behavioral conformance checking.
-% 1.7
-% (2 Points) Explain the term "lift & shift" in the context of cloud development.
-% 1.8
-% (4 Points) Decide for each of the following statements wether it is true or false. Justify your decision
-% in one sentence.
-% If your system consists of a set of microservices, there is no need for a central architecture document
-% since each service is fre to choose its technologies.
-% ⃝ True
-% ⃝ False
-% Explanation:
-% Celebrating release parties is a sign of great architecture.
-% ⃝ True
-% ⃝ False
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
 
-% 2 Question: Reqirements Engineering (15 Points)
-% The GreenHome Energy Management System is an innovative solution designed to empower (1) home-
-% ownerstoharnessandoptimzegreenenergysourceswithintheirresidences, (2)energyserviceproviders
-% to improve service customization, e.g. flexible pricing based on user behavior and offer energy-saving
-% solutions, and (3) renewable energy solution providers to invest in research and development of renew-
-% able energy technologies. To promote sustainability, reduce carbon footprint, and maximize energy
-% efficiency, this comprehensive software system integrates seamlessly with (1) smrt home devices such
-% as smart thermostats, lighting systems, and appliances and (2) renewable energy sources, including
-% solar panels, and energy storage systems.
-% 2.1
-% Solve the following tasks regarding the stakeholder analysis.
-% i. (4 Points) Write the name of each axis after each = sign. In addition, classify three stakeholders
-% derived from the text according to the following matrix.
-% Figure 1: Stakeholder matrix
-% ii. (2 Points) Classify two more stakeholders that are not explicitly mentioned in the text. Add
-% them to the matrix in Figure 1. Show which ones are the two additional stakeholders by drawing
-% a circle around the two names.
-% iii. (3 Points) Briefly explain the reason for your classification for the two additional stakeholders.
-% 5
-% 2.2
-% Consider the following scenario, which is related to the GreenHome Energy Management System
-% requirements:
-% • "The Cost analysis in the GreenHome Energy Management System addresses the need for home-
-% owners to observe their energy-rekated expenses, which involve providing users witha straight-
-% forward breakdown of their energy costs, including expenses associated with renewable energy
-% production."
-% • "The system provides advanced analytics to forecast future energy costs based on usage patterns
-% and external factors."
-% • "It would be cool to have a gamified elemet eithin the Cost calculation. The idea is to award
-% users GreenPoints for implementing energy saving measures, achieving specific sustainability
-% milestones."
-% Regarding the Kano model, solve the following tasks.
-% i. (2 Points) Derive one basic requirement from the given scenario.
-% ii. (2 Points) Derive one performance requirement from the given scenario.
-% iii. (2 Points) Derive one excitement requirement from the given scenario.
+\begin{exercise}{Stemming and lemmatization}
+  Difference between stemming and lemmatization.
 
-% 3 Question: Architecture Modeling (23 Points)
-% Figure 2 is part of an architectural description of an image hosting application. The application has
-% two keey functionalities: (1) the ability to upload (write) an image to the serer, and (2) the ability to
-% query for an image. Make sure to read and understand Figure 2.
-% Figure 2: Architectural description of an image hosting application
-% The team responsible for the image hosting application project decided to use 4+1, which is a
-% template to structure the documentation of a software architecture. Answer the following questions
-% and help the team to create the documentation of the architecture.
-% 3.1
-% Qestions regarding Figure 2:
-% i. (1 Point) What Architectural Description Language (ADL) is used by the team?
-% ii. (2 Points) What view does the architectural description show accordingto the 4+1 template?
-% Briefly explain why.
-% 7
-% 3.2
-% (10 Points) Process view:
-% Consider the following scenario. "A user calls getImages through the IDownloadImage interface of
-% WebUI component to download images. WebUI calls a method with the same name getImages from
-% the IRequestHandler interface. Furthermore, the RequestHandler component calls the zip method from
-% the ReEncoder component to get a zip package of all images. The ReEncoder component calls the
-% imagesList method of the ImageAccess component. Finally, the ImageAccess component returns a
-% result if it is already cashed. Otherwise, it returns NULL."
-% Specify the intercomponent behavior through a UML sequence diagram for the given scenario.
-% 8
-% 3.3
-% (10 Points)Deployment view:
-% Use the UML notation to specify that (1) the WebUI component is deployed on aFrontEnd server,
-% (2) the RequestHandler, ReEncoder and IMage Acces components are deployed to the Application-
-% Server, (3) the Database and FileStorage components are deployed to the DatabaseServer and File-
-% Server respectively. All the physical servers ar interconnected via LAN.
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
 
-% 4 Question: Architecture Evaluation (22 Points)
-% Jitsi Meet is a video-conferencing solution. The main service that Jitsi offers is creating video confer-
-% ences and allowing participants across the Internet to join the conference. Three components are the
-% core of the Jitsi system:
-% • JitsiWeb provides a web-based user interface to access the Jitsi functionality.
-% • JiCofo provides functionality for managing conferences.
-% • VideoBridge is the critical component that is responsible for routing the video stream between
-% multiple participants of a conference.
-% Figure 3 shows how users attend different conferences using the Jitsi system. User1 and User2
-% (depicted with a black box around user icon) joined Conference 1 (depicted with a black box around
-% the text), and User 3 and User 4 joined Conference 2. The JiCofo component communicates with
-% the VideoBridge component through XMPP, which is a communication protocol. Each server can
-% have redundancies to support more conferences simultaneously. It is important to note that when a
-% conference fails, e.g. Conference 1, a new conference has to be created. There is no redundancy and
-% failover for a particular conference.
-% Figure 3: Jitsi video conference system
-% 10
-% 4.1
-% Consider the Architecture Tradeoff Analysis Method (ATAM).
-% i. Provide a scenario for the Performance quality attribute such that the system maintains its
-% performance when the number of users accessing the system increases.
-% a) (1 Point) Source
-% b) (1 Point) Stimulus
-% c) (1 Point) Artifact
-% d) (1 Point) Environment
-% e) (1 Point) Response
-% f) (1 Point) Response measure
-% ii. Provide a scenario for the Modifiablity quality attribute.
-% a) (1 Point) Source
-% b) (1 Point) Stimulus
-% c) (1 Point) Artifact
-% d) (1 Point) Environment
-% e) (1 Point) Response
-% f) (1 Point) Response measure
-% 11
-% 4.2
-% ATradeoff point isasystempropertythataffectsmorethanonequalityattribute. Forexample, adding
-% more bridge servers affects more than one quality attribute in the system. Therefore, the number of
-% bridge servers is a tradeoff point.
-% i. (1 Point) Name at least two quality attributes of the system that the number of bridge servers
-% affects.
-% ii. (2 Points) Explain in one statement how each quality is affected by a change in the number of
-% bridge servers.
-% 4.3
-% The current deployment strategy of Jitsi Meet system is visible in Figure 3. The engineering team
-% deploys JitsiWeb and JiCofo service instances into one physical server. The team argues efficient
-% resource usage is the main reason for their decision.
-% i. (1 Point) Name one drawback for the current deployment strategy.
-% ii. (2 Points) Suggest a new deployment strategy and explain the reason behind your decision.
-% iii. (1 Point) Name one drawback of the new deployment strategy.
-% 12
-% 4.4
-% (3 Points) Performance engineers of the JitsiMeet system claim that their system is efficient in terms
-% of CPU usage. To prove their calim, they experimented on a server machine, which runs all the services
-% needed by Jitsi Meet including the JitsiWeb, JiCofo and VideoBridge.
-% The engineering team’s big challenge was to send enough traffic to put the system under load, ehich
-% requires a considerable number of machines. What would you suggest to the engineering team as an
-% alternative approach for quantitative evaluation of the Jitsi Meet? Briefly explain what your approach
-% is.
+\begin{exercise}{Web as corpus}
+  List 4 issues you have to deal with when using the web as a corpus for NLP purposes. Explain each.
 
-\sheet[2024]{2nd exam}
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
 
-% 1 Foundation (30 P.)
-% 1. What does vertical traceability of requirements mean?
-% 2. Name the layers of cloud computing and explain vendor lock-in.
-% 3. What is a Domain Model? Explain using an example.
-% 4. Consider architectural styles and patterns.
-% a) Name two similarities between architectural styles and patterns.
-% b) Name two differences between architectural styles and patterns.
-% c) How do styles and patterns relate to each other?
-% 5. Give two advantages and two disadvantages of semi-/formal vs. informal architec-
-% ture documentation. (4 P.)
-% 1
-% 2 Requirements Engineering (12 P.)
-% Here was a text given describing the rough proposed functionality of a music streaming
-% service (Melody).
-% 1. Classify four stakeholders according to the following matrix. Write the name of
-% each axis after the = sign. Now there was the empty stakeholder matrix given, the
-% axes named “x =” and “y =”. (5 P.)
-% 2. Briefly explain your classification of two of the stakeholders. (1 P.)
-% 3. Consider the Kano model.
-% a) Describe one basic feature of the system and say, why it is one.
-% b) Describe one performance feature of the system and say, why it is one.
-% c) Describe one excitement feature of the system and say, why it is one.
-% 3 Architecture Modeling (ca. 24 P.)
-% Here was a text given describing the rough functionality of a new IDE (Power IDE ).
-% 1. Which architectural style was used for the system? Give a reason why.
-% 2. Which quality attribute does this style match and why?
-% 3. Single choice question (yes/no): Is there usually a trade-off between the quality
-% attributes modifiability and performance? Explain your answer briefly.
-% 4. Single choice question (yes/no): Is there usually a trade-off between reusability
-% and interoperability? Explain your answer briefly.
-% 5. Use UML to model the presented architecture. (10 P.)
-% 6. (Here a text was given describing a certain starting procedure.) Give a UML
-% sequence diagram for the presented process. (8 P.)
-% 2
-% 4 Evaluation (ca. 24 P.)
-% Here a text and a UML deployment diagram were given describing the rough functional-
-% ity of an online order service (Food2Go) consisting of the three components OrderService,
-% RestaurantService and AccountingService where OrderService had a connection to
-% both of the other services but these had no connection to each other.
-% 1. Consider the Architecture Tradeoff Analysis Method (ATAM).
-% a) Provide a scenario for the Availability quality attribute: Source, Stimulus,
-% Artifact, Environment, Response, Response measure (one text field each).
-% b) Provide a scenario for the Modifiability quality attribute such that deploy-
-% ment has finished after a maximum of one day: Source, Stimulus, Artifact,
-% Environment, Response, Response measure (one text field each).
-% 2. A design alternative (design 2) is to add a cache to OrderService.
-% a) How does this alternative affect availability?
-% b) Describe a scenario in which a client could experience data inconsistency.
-% 3. Another design alternative (design 3) is to add third party messaging service.
-% a) Would that solve the problem of data inconsistency?
-% b) Does design 3 improve availability?
-% c) Name a disadvantage of design 3 compared to design 2.
+\begin{exercise}{Cross-language QA}
+  How does cross-language QA work? Issues?
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Machine learning pipeline}
+  What are the main concepts in a machine learning pipeline?
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{IR evaluation}
+  What are the problems with precision, recall, and f-measure in IR? How to improve and why is as an example cohen kappa better?
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{IR evaluation}
+  Calculate P@5, P@10, MAP for the given IR ranked result list (one for IR system A and one for B. Both go up to ten entries and either have a checkmark or an x in each rank).
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Naive Bayes and SVM}
+  Which of the following 4 are correct
+  \begin{enumerate}
+    \item Features for Naive Bayes classifiers do not need to be uncorrelated
+    \item Naive Bayes is slow to train
+    \item SVM itself is a multi-label classifier
+    \item SVM handles sparse data well
+  \end{enumerate}
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Corpus construction}
+  How to construct a corpus for linguistic purposes from web data and how to make it usable?
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Pragmatics and discourse analysis}
+  Why are pragmatics and discourse analysis difficult?
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Hate speech}
+  Define hate speech and how to construct a hate speech classifier
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Counter hate speech}
+  What is counter hate speech and how to construct such a system
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Crowdsourced annotation}
+  Crowdsourced annotation workflow
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Web corpus}
+  Why are duplicates on the web and how to detect
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{QA vs IR}
+  Differences in QA vs IR
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Cross-language QA}
+  Explain applications of multi-language QA and difficulties?
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Language detection}
+  Explain language detection
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Factual QA vs opinionated QA}
+  Differences between factual QA and opinionated QA, and which is easier to evaluate?
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{RAG and LangChain}
+  What's RAG and how does LangChain help?
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Sentiment analysis}
+  Discuss: Sentiment analysis using supervised vs. lexicon based approaches.
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Edit distance}
+  Calculate the Minimum Edit Distance for "TAC" vs. "CAT" (given is also a table just like on the slides, just without the numbers filled in). List all possible word outcomes (in this case 4).
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\sheet{2022}{1st exam}
+
+\begin{exercise}{Corpus construction}
+  How would you construct a corpus for linguistic purposes and how to make it usable?
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Pragmatics and discourse analysis}
+  Why are pragmatics and discourse analysis difficult?
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Sequence tagging and CRF}
+  Define sequence tagging and what is CRF
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Minimum edit distance}
+  Calculate the Minimum Edit Distance for "TAC" vs. "CAT" (given is also a table just like on the slides, just without the numbers filled in). List all possible word outcomes (in this case 4).
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{IR evaluation}
+  Calculate P@5 and MAP for the given IR ranked result list (one for IR system A and one for B. Both go up to ten entries and either have a checkmark or an x in each rank).
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Sequence tagging and CRF}
+  Define sequence tagging and what is CRF
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Hate speech}
+  Define hate speech and how to construct a hate speech classifier
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Counter hate speech}
+  What is counter hate speech and how to construct such a system
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Sentiment analysis}
+  Discuss sentiment detection using lexicon based vs. supervised learning approaches.
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Sentiment lexicon extension}
+  How could you analyse sentiment with lexicon extension using distributed thesauruses.
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Sentiment analysis}
+  Why is lexicon based sentiment analysis not enough?
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Naive Bayes and SVM}
+  Which of the following 4 are correct
+  \begin{enumerate}
+    \item Features for Naive Bayes classifiers do not need to be uncorrelated
+    \item Naive Bayes is slow to train
+    \item SVM itself is a multi-label classifier
+    \item SVM handles sparse data well
+  \end{enumerate}
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Language detection}
+  Explain approaches for language detection
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Crowdsourced annotation}
+  Explain the crowdsourced annotation workflow
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Tokenization ambiguities}
+  What are the ambiguities for tokenization?
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{QA vs IR}
+  Differences between QA and IR
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Cross-language QA}
+  What are the challenges in cross-language QA?
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Stemming vs lemmatization}
+  What is the difference between stemming and lemmatization?
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Web corpus}
+  Why are duplicates on the web and how to detect
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Web corpus}
+  What are the 4 challenges when using the web as a corpus
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{TREC}
+  What are the TREC question types
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{QALD}
+  What is QALD and what are the challenges
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\sheet{2021}{1st exam}
+
+\begin{exercise}{Comprehensiveness}
+  Explain comprehensiveness.
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Complete the picture}
+  Fill in the white boxes in the image.
+
+  % TODO: add image
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{NLTK}
+  Explain two parameters in NLTK that start with a c.
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{SVM}
+  Explain Support Vector Machines when you have more than two classes.
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Feature differences}
+  Explain the differences in features between classical ML and deep network based ML.
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Cohen's Kappa}
+  Calculate Cohen's Kappa for a given table.
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Evaluation metrics}
+  Calculate Precision, Recall, and F1-Score.
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{IR and classification}
+  Where is classification used in Information Retrieval?
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Translate datasets}
+  List 3 reasons for and against translating an existing annotated dataset from German to another language.
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Web corpus}
+  List 3 reasons for and against using the web as a corpus.
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Pipeline}
+  Given the pipeline from the first slide, assign three words to the individual steps.
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{MRR and accuracy}
+  Calculate MRR and accuracy for a Question/Answering system.
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Language detection}
+  Explain language detection.
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Plagiarism detection}
+  Explain how to find plagiarism.
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Crowdsourcing}
+  List the pros and cons of crowdsourcing.
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Recall improvement}
+  Explain how to improve recall when using a search engine without changing the algorithm.
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Spell checker}
+  Explain how a spell checker works with non-words.
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{IR cycle}
+  Explain the Information Retrieval cycle.
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Vector space}
+  Explain Vector Space in Information Retrieval.
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Sentiment analysis}
+  Explain the difference between factual sentiment analysis and opinionated analysis.
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Sentiment analysis}
+  Explain the areas where sentiment analysis is used.
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Teacher/Student}
+  Explain Teacher/Student and why it is better than just audio.
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Representation and fusion}
+  Explain representation and fusion and the differences.
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
 

### 2024-10-21 BAI/exercises.tex
Added BAI exam exercises.

@@
 \section[2023]{1st exam}
-% First questions were multiple choice
 
-% Function to compute neighbourhood in SOM
-% Meaning of grounded language
-% Ventriloquism Effect
+\begin{exercise}{Multiple Choice}
+  \begin{enumerate}
+    \item How is the neighborhood in a self-organizing map (SOM) is computed.
 
-% Followed by some general questions
+      \begin{itemize}
+        \item The neighborhood is computed as a Gaussian function of the distance between the winning neuron and the other neurons.
+        \item The neighborhood is computed as a linear function of the distance between the winning neuron and the other neurons.
+        \item The neighborhood is computed as a step function of the distance between the winning neuron and the other neurons.
+        \item The neighborhood is computed as a constant function of the distance between the winning neuron and the other neurons.
+      \end{itemize}
 
-% multisensory integration of different strength stimuli (slide example)
-% What is the cocktailparty effect
+    \item What does it mean for a language to be grounded?
 
-% Spiking neural networks
+      \begin{itemize}
+        \item The language is based on the ground truth.
+        \item The language is based on the physical world.
+        \item The language is based on the ground state.
+        \item The language is based on the ground floor.
+      \end{itemize}
 
-% label leaky and integrate+fire models
-% Similarity angle, calculate for a weight and input vector
-% Compute output from perceptron (similar to slide example)
-% Visual
+    \item What is the ventriloquism effect?
 
-% what do the different layers on CNN learn (not every single layer but a brief overview)
-% Compute filtering in CNN layer (5x5 matrix with 3x3 filter, stride=1, no padding)
-% Draw orientation and firing rate diagram from slides
-% Audio
+      \begin{itemize}
+        \item The ventriloquism effect is the illusion that a sound is coming from a different location than the actual source.
+        \item The ventriloquism effect is the illusion that a sound is coming from the same location as the actual source.
+        \item The ventriloquism effect is the illusion that a sound is coming from a different location than the actual source.
+        \item The ventriloquism effect is the illusion that a sound is coming from the same location as the actual source.
+      \end{itemize}
+  \end{enumerate}
 
-% ITD and ILD: explain and what to use for which frequencies
-% Behavior
+  \begin{solution}
+    \begin{enumerate}
+      \item The neighborhood is computed as a Gaussian function of the distance between the winning neuron and the other neurons.
+      \item The language is based on the physical world.
+      \item The ventriloquism effect is the illusion that a sound is coming from a different location than the actual source.
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
 
-% braitenberg vehicles
-% Advantage of behavior based
+\begin{exercise}{General}
+  \begin{enumerate}
+    \item Explain the multisensory integration of different strength stimuli.
+    \item What is the cocktail party effect?
+  \end{enumerate}
 
-% Evolutionary Computing
+  \begin{solution}
+    \begin{enumerate}
+      \item Multisensory integration is the process by which information from different sensory modalities is combined to form a unified percept.
+This process can enhance the detection and discrimination of stimuli and improve the accuracy and reliability of sensory processing.
+      \item The cocktail party effect is the ability to focus on a single conversation in a noisy environment.
+This phenomenon allows individuals to selectively attend to one speaker while ignoring other competing sounds.
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
 
-% Cycle Graph also arrow labels
-% Example: find the shortest distance between cities only visit each city once
-% Recombination and Mutation algorithm for example
-% Optimisation strategies for this algorithm
+\begin{exercise}{Spiking Neural Networks}
+  \begin{enumerate}
+    \item Explain the difference between the leaky integrate-and-fire model and the integrate-and-fire model.
+    \item Calculate the similarity angle between a weight vector $\mathbf{w} = [1, 2, 3]$ and an input vector $\mathbf{x} = [4, 5, 6]$.
+    \item Compute the output of a perceptron with weights $\mathbf{w} = [1, 2, 3]$ and input $\mathbf{x} = [4, 5, 6]$.
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+      \item The leaky integrate-and-fire model is an extension of the integrate-and-fire model that includes a leaky term.
+This term allows the membrane potential of the neuron to decay over time, which can help to model more realistic neuron behavior.
+      \item The similarity angle between two vectors $\mathbf{a}$ and $\mathbf{b}$ is given by
+      \[
+        \cos(\theta) = \frac{\mathbf{a} \cdot \mathbf{b}}{\|\mathbf{a}\| \|\mathbf{b}\|}.
+      \]
+      For $\mathbf{w} = [1, 2, 3]$ and $\mathbf{x} = [4, 5, 6]$, we have
+      \[
+        \cos(\theta) = \frac{1 \cdot 4 + 2 \cdot 5 + 3 \cdot 6}{\sqrt{1^2 + 2^2 + 3^2} \sqrt{4^2 + 5^2 + 6^2}} = \frac{32}{\sqrt{14} \sqrt{77}}.
+      \]
+      \item The output of a perceptron is given by
+      \[
+        y = \begin{cases}
+          1 & \text{if } \mathbf{w} \cdot \mathbf{x} + b > 0, \\
+          0 & \text{otherwise}.
+        \end{cases}
+      \]
+      For $\mathbf{w} = [1, 2, 3]$, $\mathbf{x} = [4, 5, 6]$, and $b = 0$, we have
+      \[
+        y = \begin{cases}
+          1 & \text{if } 1 \cdot 4 + 2 \cdot 5 + 3 \cdot 6 > 0, \\
+          0 & \text{otherwise}.
+        \end{cases}
+      \]
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Visual Processing}
+  \begin{enumerate}
+    \item What do the different layers in a convolutional neural network (CNN) learn?
+    \item Compute the filtering in a CNN layer with a $5 \times 5$ matrix and a $3 \times 3$ filter, stride $1$, and no padding.
+    \item Draw a diagram of the orientation and firing rate of a simple cell for a line of different orientations.
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+      \item The different layers in a CNN learn different features of the input data.
+The first layers typically learn low-level features like edges and textures, while later layers learn higher-level features like shapes and objects.
+      \item The filtering in a CNN layer with a $5 \times 5$ matrix and a $3 \times 3$ filter, stride $1$, and no padding can be computed by sliding the filter over the input matrix and computing the dot product at each position.
+The resulting output will be a $3 \times 3$ matrix.
+      \item The orientation and firing rate of a simple cell for a line of different orientations can be visualized in a diagram with the orientation on the $x$-axis and the firing rate on the $y$-axis.
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Audio Processing}
+  \begin{enumerate}
+    \item Explain interaural time difference (ITD) and interaural level difference (ILD) and which one is used for high and low frequencies.
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+      \item Interaural time difference (ITD) is the difference in time it takes for a sound to reach each ear, while interaural level difference (ILD) is the difference in intensity of a sound at each ear.
+ITD is used for low frequencies, while ILD is used for high frequencies.
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Behavior}
+  \begin{enumerate}
+    \item Explain a Braitenberg vehicle with an example and name the behavior of your example vehicle.
+    \item Complete the flow chart of the functional decomposition of the classical approach for a robot controller: sensors
+    \item Explain the advantage of a behavior-based approach to robotics.
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+      \item A Braitenberg vehicle is a simple robot that exhibits complex behavior through the direct coupling of sensors to actuators.
+An example of a Braitenberg vehicle is a light-seeking robot that moves towards a light source.
+The behavior of this example vehicle is phototaxis.
+      \item The flow chart of the functional decomposition of the classical approach for a robot controller is: sensors $\rightarrow$ perception $\rightarrow$ cognition $\rightarrow$ planning $\rightarrow$ control $\rightarrow$ actuators.
+      \item The advantage of a behavior-based approach to robotics is that it allows for the development of complex behaviors through the combination of simple behaviors.
+This approach is more robust and adaptable to changing environments than traditional control architectures.
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Evolutionary Computing}
+  \begin{enumerate}
+    \item Draw a cycle graph with arrow labels.
+    \item Given an example of finding the shortest distance between cities by only visiting each city once.
+    \item Describe a recombination and mutation algorithm for the example you provided.
+    \item Explain optimization strategies for this algorithm.
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+      \item A cycle graph with arrow labels is a graph where each node is connected to the next node in a cycle, and the arrows indicate the direction of the cycle.
+      \item An example of finding the shortest distance between cities by only visiting each city once is the traveling salesman problem.
+      \item A recombination and mutation algorithm for the traveling salesman problem could involve recombining the order of cities in two different solutions and mutating the order of cities in a single solution.
+      \item Optimization strategies for this algorithm could include using a genetic algorithm to evolve a population of solutions, using local search to improve the quality of solutions, and using elitism to preserve the best solutions.
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
 
 \section[2022]{1st exam}
-% Nr.1) Multiple choice: (one or multiple correct, you only get points if all ticks are completely correct, the last possibility of four was always "none of the above")
 
-% a)
+\begin{exercise}{Multiple Choice}
+  \begin{enumerate}
+    \item In which models is backpropagation used for learning?
 
-% ...
+      \begin{itemize}
+        \item Multilayer perceptron (MLP)
+        \item Convolutional neural network (CNN)
+        \item Self-organizing map (SOM)
+        \item None of the above
+      \end{itemize}
 
-% b) In which Models is Backpropagation used for learning?
+    \item What does "embodied" mean for language?
 
-% MLP
+      \begin{itemize}
+        \item Language is based on the physical world.
+        \item Language is based on the ground truth.
+        \item Language is based on the ground state.
+        \item Language is based on the ground floor.
+      \end{itemize}
 
-% Convolutional Neural Network
+    \item What are the two streams and their tasks in human vision?
 
-% SOM
+      \begin{itemize}
+        \item Dorsal stream for spatial vision and ventral stream for object recognition
+        \item Ventral stream for spatial vision and dorsal stream for object recognition
+        \item Dorsal stream for object recognition and ventral stream for spatial vision
+        \item Ventral stream for object recognition and dorsal stream for spatial vision
+      \end{itemize}
+  \end{enumerate}
 
-% None of the above
+  \begin{solution}
+    \begin{enumerate}
+      \item Multilayer perceptron (MLP)
+      \item Language is based on the physical world.
+      \item Dorsal stream for spatial vision and ventral stream for object recognition
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
 
-% c) what does "embodied" means for language?
+\begin{exercise}{Attention}
+  \begin{enumerate}
+    \item Name and explain three ways to measure the attention of a human.
+    \item Describe two models to measure or visualize human attention.
+    \item Name advantages and disadvantages of the two models you proposed.
+  \end{enumerate}
 
-% ...
+  \begin{solution}
+    \begin{enumerate}
+      \item Three ways to measure the attention of a human are eye tracking, reaction time, and brain imaging.
+      \item Two models to measure or visualize human attention are the spotlight model and the zoom lens model.
+      \item The spotlight model has the advantage of being simple and intuitive, but it has the disadvantage of oversimplifying the attentional process.
+The zoom lens model has the advantage of being more flexible and dynamic, but it has the disadvantage of being more complex and difficult to interpret.
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
 
-% d)
+\begin{exercise}{Vision}
+  \begin{enumerate}
+    \item Name the two streams and their task that are present in human vision.
+    \item Explain rods and cones.
+    \item Explain the similarity between a convolutional neural network with convolution and max pooling layers and the visual cortex.
+    \item Visualize the firing rate of a simple cell for a line of different orientation in a diagram with the orientation on the $x$-axis and the firing rate on the $y$-axis.
+    \item Given a $5 \times 5$ matrix and a $3 \times 3$ filter, compute the convolution (no padding, stride $1$).
+  \end{enumerate}
 
-% ...
+  \begin{solution}
+    \begin{enumerate}
+      \item The two streams present in human vision are the dorsal stream for spatial vision and the ventral stream for object recognition.
+      \item Rods and cones are photoreceptor cells in the retina that are responsible for detecting light.
+Rods are sensitive to low light levels and are responsible for night vision, while cones are sensitive to color and are responsible for daylight vision.
+      \item A convolutional neural network with convolution and max pooling layers is similar to the visual cortex in that both systems use hierarchical processing to extract features from the input data.
+      \item The firing rate of a simple cell for a line of different orientations can be visualized in a diagram with the orientation on the $x$-axis and the firing rate on the $y$-axis.
+      \item Given a $5 \times 5$ matrix and a $3 \times 3$ filter, the convolution (no padding, stride $1$) can be computed by sliding the filter over the input matrix and computing the dot product at each position.
+The resulting output will be a $3 \times 3$ matrix.
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
 
-% e)
+\begin{exercise}{Sound Localization}
+  \begin{enumerate}
+    \item Compute the cross-correlation steps on two signals until the maximum cross-correlation value is reached.
+    \item Name and explain the two cues used to perform sound localization. Which one works best for high/low frequencies?
+  \end{enumerate}
 
-% ...
+  \begin{solution}
+    \begin{enumerate}
+      \item The cross-correlation steps on two signals can be computed by sliding one signal over the other and computing the dot product at each position until the maximum cross-correlation value is reached.
+      \item The two cues used to perform sound localization are interaural time difference (ITD) and interaural level difference (ILD).
+ITD works best for low frequencies, while ILD works best for high frequencies.
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
 
-% Nr.2) Attention
+\begin{exercise}{Behavior}
+  \begin{enumerate}
+    \item Explain a Braitenberg vehicle with an example and name the behavior of your example vehicle.
+    \item Complete the flow chart of the functional decomposition of the classical approach for a robot controller: sensors $\rightarrow$ blank $\rightarrow$ blank $\rightarrow$ blank $\rightarrow$ blank $\rightarrow$ actuators.
+  \end{enumerate}
 
-% 2.1) Name and explain three ways to measure the attention of a human.
+  \begin{solution}
+    \begin{enumerate}
+      \item A Braitenberg vehicle is a simple robot that exhibits complex behavior through the direct coupling of sensors to actuators.
+An example of a Braitenberg vehicle is a light-seeking robot that moves towards a light source.
+The behavior of this example vehicle is phototaxis.
+      \item The flow chart of the functional decomposition of the classical approach for a robot controller is: sensors $\rightarrow$ perception $\rightarrow$ cognition $\rightarrow$ planning $\rightarrow$ control $\rightarrow$ actuators.
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
 
-% 2.2) Describe two models to measure or visualize human attention.
+\begin{exercise}{Continual Learning}
+  \begin{enumerate}
+    \item Explain the stability-plasticity dilemma.
+    \item Explain catastrophic forgetting and a model where this effect could arise.
+    \item Name and describe two strategies of continual learning.
+  \end{enumerate}
 
-% 2.3) Name advantages and disadvantages of the two models you proposed.
+  \begin{solution}
+    \begin{enumerate}
+      \item The stability-plasticity dilemma is the trade-off between the ability of a neural network to learn new information (plasticity) and the ability of the network to retain previously learned information (stability).
+      \item Catastrophic forgetting is the phenomenon where a neural network forgets previously learned information when learning new information.
+This effect could arise in a model where the network is trained on a sequence of tasks without preserving the knowledge learned on previous tasks.
+      \item Two strategies of continual learning are rehearsal, where the network is periodically trained on previously learned tasks, and regularization, where the network is penalized for changing its weights too much.
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
 
-% Nr.3) Vision
+\begin{exercise}{Language Processing}
+  \begin{enumerate}
+    \item Name the two streams important in language processing and their tasks.
+    \item Describe CBOW (Continuous Bag-of-word) and Skip-gram.
+    \item Describe the difference between the earlier Wernicke's model and the later Pulvermüller's model.
+  \end{enumerate}
 
-% 3.1) Name the two streams and their task that are present in the human vision.
+  \begin{solution}
+    \begin{enumerate}
+      \item The two streams important in language processing are the dorsal stream for speech perception and the ventral stream for speech production.
+      \item CBOW (Continuous Bag-of-word) and Skip-gram are two models used for word embedding in natural language processing.
+CBOW predicts a target word from its context, while Skip-gram predicts the context words from a target word.
+      \item The difference between the earlier Wernicke's model and the later Pulvermüller's model is that Wernicke's model is based on a sequential processing model of language, while Pulvermüller's model is based on a distributed processing model of language.
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
 
-% 3.2) Explain rods and cones
+\begin{exercise}{Computational Neural Networks}
+  \begin{enumerate}
+    \item Explain three steps of the backpropagation algorithm and explain the termination criteria.
+    \item Name and shortly describe the three fundamental ANN learning paradigms.
+    \item Explain how the learning rate parameter could be chosen.
+  \end{enumerate}
 
-% 3.3) Explain the similarity between a convolutional neural network with convolution and max pooling layers and the visual cortex (not sure if visual cortex or some other brain area)
-
-% 3.4) Visualize the firing rate of a simple cell for a line of different orientation in a diagram with the orientation on the x-axis and the firing rate on the y-axis
-
-% 3.5) Given a 5x5 matrix and a 3x3 filter compute the convolution (no padding, stride = 1)
-
-% Nr.4) Sound Localization
-
-% 4.1) compute crosscorrelation steps on two signals until maximum crosscorrelation value is reached
-
-% 4.2) name and explain the two cues used to perform sound localization. Which one works best for high/low frequencies?
-
-% ...
-
-% Nr.5) Behaviour
-
-% 5.1) Explain a Braitenberg Vehicle with an example and name the behaviour of your example vehicle.
-
-% 5.2) Flow chart of functional decomposition of the classical approach for a robot controller was given with steps left blank: (sensors => blank => blank => blank => blank => actuators)
-
-% ...
-
-% Nr.6) Continual Learning
-
-% 6.1) Explain the stability-plasticity dilemma
-
-% 6.2) Explain catastrophic forgetting and a model where this effect could arise
-
-% 6.3) Name and describe two strategies of continual learning
-
-% Nr.7) Language processing
-
-% 7.1) Name the two streams important in language processing and their tasks
-
-% 7.2) Describe CBOW (Continuous Bag-of-word) and Skip-gram
-
-% 7.3) Describe difference between the earlier Wernickes model and the later Pulvermüllers model (question gave a few more details which I don't remember anymore)
-
-% Nr.8) Computational Neural Networks
-
-% 8.1) Explain 3 steps of Backpropagation algorithm and explain termination criteria
-
-% 8.2) Name and shortly describe the three fundamental ANN Learning Paradigms
-
-% 8.3) Explain how the learning rate parameter could be chosen
+  \begin{solution}
+    \begin{enumerate}
+      \item Three steps of the backpropagation algorithm are forward propagation, backward propagation, and weight update.
+The termination criteria for backpropagation is typically based on the convergence of the loss function.
+      \item The three fundamental ANN learning paradigms are supervised learning, unsupervised learning, and reinforcement learning.
+Supervised learning involves learning from labeled data, unsupervised learning involves learning from unlabeled data, and reinforcement learning involves learning from rewards and punishments.
+      \item The learning rate parameter could be chosen using a grid search or random search over a range of values, or by using adaptive learning rate methods like AdaGrad or Adam.
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
 

### 2024-10-21 PbD/exercises.tex
Added PbD exam exercises.

@@
 \section[2020]{1st exam}
-% 1. Broadcast
-% • Wer wird geschützt und welches Schutzziel wird verfolgt?
-% • Nenne zwei Anwendungsbeispiele (abgesehen von Zeitung/TV/Radio)
-% • Professorin möchte über eine Zeitschrift auf Feedback von Studierenden Antworten und die Identität dieser
-% schützen. Anweisungen ähnlich wie: Neben dem Feedback soll eine Zufallszahl zwischen 0 und 100 angegeben
-% werden. Antworten werden im Bezug auf dieses Pseudonym gegeben. Welche der beiden Arten der impliziten
-% Adressierung wurde hier genutzt? Was sind mögliche Probleme in Bezug auf speziell diese Auswahl an
-% Pseudonymen?
-% • Der Professorin gefällt nicht, dass die Antworten auf das Feedback von jedem gelesen werden können (und
-% nicht nur von der entsprechenden Person). Wir könnte das Problem gelöst werden? Schreibe einen Text mit
-% Anweisungen an die Studierenden und Begründe die Notwendigkeit der einzelnen Komponenten.
-% 2. IPv6-Präfixe
-% Zur Einleitung wurden Präfix-Sharing, -Hopping und -Boquets beschrieben
-% • Warum genügt es nicht verschiedene Adressen unter einem Präfix zu verwenden?
-% • Was sind die Schwächen von Präfix-Sharing bei einem gleich bleibendem Präfix und einer kleinen Nutzergruppe?
-% • Was sind die Schwächen von Präfix-Boquets, wenn unterschiedliche Präfixe für verschiedene Ziele aber nicht
-% pro Verbindung genutzt werden?
-% • Was ist das Problem beim Präfix-Hopping, wenn sich während einer bestehenden Verbindung die IP-Adresse
-% ändert?
-% • Wie kann man dieses Problem beim Präfix-Hopping lösen?
-% • Wie kann man Benutzer abgesehen von der Adressierung tracken? Nenne 3 Beispiele.
-% 3. Mixe
-% • Angreifermodell angeben
-% • Wie kann ein Mix-Netz von einem starken Angreifer (z.B. Geheimdienst) angegriﬀen werden? Nenne 2
-% Beispiele.
-% • Es wird ein Mix-Netz angeboten von nur einem Betreiber, welcher damit wirbt, dass 20 Mixe hinter einander
-% verwendet werden und für höhere Performance nur 10 Teilnehmer pro Mix zugeordnet werden. Bewerte die
-% Sicherheit des vorgestellten Produkts mit Begründung.
-% 4. DC-Netze
-% Gegeben waren 4 lokale Summen (A, B, C, D), ein Schlüsselgraph kAB , kBC , kCD und der Schlüssel kBC . Wir sind
-% Teilnehmer B und haben eine Leernachricht gesendet.
-% • Berechne die gesendete Nachricht N . Skizziere deinen Rechenweg.
-% • Grenze ein wer die Nachricht gesendet haben kann. Skizziere deinen Rechenweg.
-% • Was muss passieren, damit B weniger Aussagen treﬀen kann?
+\begin{exercise}{Broadcast}
+  \begin{enumerate}
+    \item Who is protected and what protection goal is pursued?
+    \item Name two examples of applications (apart from newspaper/TV/radio).
+    \item A professor wants to receive feedback from students via a magazine and protect their identity.
+Instructions similar to: In addition to the feedback, a random number between 0 and 100 should be given.
+Answers are given in relation to this pseudonym.
+Which of the two types of implicit addressing was used here?
+What are possible problems with regard to this selection of pseudonyms?
+    \item The professor does not like that the answers to the feedback can be read by everyone (and not only by the corresponding person).
+How could the problem be solved?
+Write a text with instructions to the students and justify the necessity of the individual components.
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+      \item The students are protected and the protection goal is to keep their identity secret.
+      \item Two examples of applications are:
+      \item The professor used the implicit addressing by pseudonyms.
+Possible problems with this selection of pseudonyms are that the pseudonyms are not unique and that the students could be identified by the professor.
+      \item The problem could be solved by encrypting the feedback and only allowing the professor to decrypt it. The students should also use a secure connection to send the feedback.
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{IPv6 prefixes}
+  \begin{enumerate}
+    \item Why is it not sufficient to use different addresses under one prefix?
+    \item What are the weaknesses of prefix sharing with a constant prefix and a small user group?
+    \item What are the weaknesses of prefix bouquets if different prefixes are used for different goals but not per connection?
+    \item What is the problem with prefix hopping if the IP address changes during an existing connection?
+    \item How can this problem be solved with prefix hopping?
+    \item How can users be tracked apart from addressing? Name 3 examples.
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+      \item It is not sufficient to use different addresses under one prefix because the addresses could be linked to the same user.
+      \item The weaknesses of prefix sharing with a constant prefix and a small user group are that the users can be identified by the prefix and that the users can be tracked by the prefix.
+      \item The weaknesses of prefix bouquets are that the users can be identified by the prefix and that the users can be tracked by the prefix.
+      \item The problem with prefix hopping is that the connection is interrupted when the IP address changes.
+      \item This problem can be solved by using a VPN connection that keeps the IP address constant.
+      \item Users can be tracked apart from addressing by using cookies, browser fingerprinting, and tracking pixels.
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Mixes}
+  \begin{enumerate}
+    \item Specify the attacker model.
+    \item How can a mix-net be attacked by a strong attacker (e.g. intelligence service)? Name 2 examples.
+    \item A mix-net is offered by only one operator, who advertises that 20 mixes are used in a row and only 10 participants per mix are assigned for higher performance.
+Evaluate the security of the presented product with justification.
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+      \item The attacker model is a strong attacker (e.g. intelligence service).
+      \item A mix-net can be attacked by a strong attacker by using traffic analysis and by using a Sybil attack.
+      \item The security of the presented product is low because the operator can see all the traffic and the participants can be identified by the operator.
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{DC networks}
+  Given were 4 local sums (A, B, C, D), a key graph kAB, kBC, kCD, and the key kBC. We are participant B and have sent an empty message.
+  \begin{enumerate}
+    \item Calculate the sent message N. Sketch your calculation.
+    \item Limit who can have sent the message. Sketch your calculation.
+    \item What needs to happen for B to make fewer statements?
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+      \item The sent message N is the sum of the local sums A, B, C, and D. Sketch your calculation.
+      \item The message can only be sent by participants A and C. Sketch your calculation.
+      \item For B to make fewer statements, the key kBC must be changed.
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
 

### 2024-10-23 SbD/exercises.tex
Added exercise scaffold to sbd and pbd.
Added pbd exercises.

@@
 
+\section{Assignments}
+
+\setcounter{subsection}{10}
+\begin{exercise}{Skytala}
+  Bei der Skytala handelt es sich um eine Transpositionschiffre, bei der ursprünglich ein Zylinder mit einem Papierstreifen umwickelt wurde.
+Die Ver- und Entschlüsselung kann aber auch mithilfe einer Matrix dargestellt werden.
+  \begin{enumerate}
+      \item Verschlüsseln Sie den folgenden Klartext mit einer Skytala, die durch eine Matrix mit sieben Zeilen und neun Spalten beschrieben werden kann:
+      \begin{center}
+          \texttt{GLUECKLICH\_IST\_DER\_DER\_DEN\_DINGEN\_AUF\_DEN\_GRUND\_GEHEN\_KONNTE}
+      \end{center}
+      \item Was ist der geheim zu haltende symmetrische Schlüssel der Skytala aus Teilaufgabe a) und wie hängt dieser mit der ursprünglichen Variante mit Zylinder und Papierstreifen zusammen?
+      \item Erläutern Sie kurz zwei sinnvolle Ansätze zur Kryptanalyse, d. h. Ansätze zum Brechen einer Skytala.
+  \end{enumerate}
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Spalten-Transpositionen}
+  Bei der Skytala wird ein Zylinder mit einem Papierstreifen umwickelt, beschrieben und wieder abgewickelt.
+  \begin{enumerate}
+      \item Eine Verbesserung dieser Spalten-Transposition ergibt sich, wenn zusätzlich auch die Spalten in ihrer Reihenfolge vertauscht werden (was beim Papierstreifen natürlich nicht gelingt, wohl aber im Rechner).
+Um welchen Faktor vergrößert sich dadurch der Schlüsselraum bei $s$ Spalten?
+      \item Schreiben Sie eine Skytala mit 4 Zeilen und 3 Spalten (ohne Spaltenvertauschung) in Zyklenschreibweise.
+  \end{enumerate}
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\setcounter{subsection}{20}
+\begin{exercise}{Vigenère-Chiffre}
+  Ein klassisches Verschlüsselungsverfahren ist die Vigenère-Chiffre.
+Sie arbeitet typischerweise auf einem Alphabet aus den Großbuchstaben A-Z für den Klartext, den Schlüssel und den Schlüsseltext.
+Zudem kann ein Zeichen zur Worttrennung hinzugenommen werden.
+Für diese Aufgabe wird dazu der Unterstrich \texttt{\_} verwendet.
+Ein Schlüsseltextzeichen ergibt sich aus dem um das jeweilige Zeichen des Schlüssels verschobene Klartextzeichen.
+Diese Verschiebung kann durch das sogenannte Vigenère-Tableau in Abbildung 1 dargestellt werden.
+
+  % TODO: add table
+  % A B C D E F G H I J K L M N O P Q R S T U V W X Y Z _
+  % A A B C D E F G H I J K L M N O P Q R S T U V W X Y Z _
+  % B B C D E F G H I J K L M N O P Q R S T U V W X Y Z _ A
+  % C C D E F G H I J K L M N O P Q R S T U V W X Y Z _ A B
+  % D D E F G H I J K L M N O P Q R S T U V W X Y Z _ A B C
+  % E E F G H I J K L M N O P Q R S T U V W X Y Z _ A B C D
+  % F F G H I J K L M N O P Q R S T U V W X Y Z _ A B C D E
+  % G G H I J K L M N O P Q R S T U V W X Y Z _ A B C D E F
+  % H H I J K L M N O P Q R S T U V W X Y Z _ A B C D E F G
+  % I I J K L M N O P Q R S T U V W X Y Z _ A B C D E F G H
+  % J J K L M N O P Q R S T U V W X Y Z _ A B C D E F G H I
+  % K K L M N O P Q R S T U V W X Y Z _ A B C D E F G H I J
+  % L L M N O P Q R S T U V W X Y Z _ A B C D E F G H I J K
+  % M M N O P Q R S T U V W X Y Z _ A B C D E F G H I J K L
+  % N N O P Q R S T U V W X Y Z _ A B C D E F G H I J K L M
+  % O O P Q R S T U V W X Y Z _ A B C D E F G H I J K L M N
+  % P P Q R S T U V W X Y Z _ A B C D E F G H I J K L M N O
+  % Q Q R S T U V W X Y Z _ A B C D E F G H I J K L M N O P
+  % R R S T U V W X Y Z _ A B C D E F G H I J K L M N O P Q
+  % S S T U V W X Y Z _ A B C D E F G H I J K L M N O P Q R
+  % T T U V W X Y Z _ A B C D E F G H I J K L M N O P Q R S
+  % U U V W X Y Z _ A B C D E F G H I J K L M N O P Q R S T
+  % V V W X Y Z _ A B C D E F G H I J K L M N O P Q R S T U
+  % W W X Y Z _ A B C D E F G H I J K L M N O P Q R S T U V
+  % X X Y Z _ A B C D E F G H I J K L M N O P Q R S T U V W
+  % Y Y Z _ A B C D E F G H I J K L M N O P Q R S T U V W X
+  % Z Z _ A B C D E F G H I J K L M N O P Q R S T U V W X Y
+  % _ _ A B C D E F G H I J K L M N O P Q R S T U V W X Y Z
+  % Abbildung 1: Vigenère-Tableau
+
+  \begin{enumerate}
+      \item Verschlüsseln Sie den nachfolgenden Klartext mit dem Schlüssel \texttt{PRUEFUNG} unter Verwendung einer Vigenère-Chiffre:
+      \begin{center}
+          \texttt{GLUECKLICH\_IST\_DER\_DER\_DEN\_DINGEN\_AUF\_DEN\_GRUND\_GEHEN\_KONNTE}
+      \end{center}
+      \item Erläutern Sie kurz zwei sinnvolle Ansätze zur Kryptanalyse, d. h. zum Brechen einer Vigenère-Chiffre.
+      \item Eine Verbesserung der Vigenère-Chiffre ergibt sich durch die Verlängerung des Schlüssels.
+Um welchen Faktor vergrößert sich dadurch der Schlüsselraum bei einem Alphabet mit 27 Elementen (A-Z und Unterstrich als Worttrennung) und einer Verlängerung um $s$ Zeichen?
+      \item Wie müsste die Vigenère-Chiffre verändert werden, um informationstheoretische Sicherheit im Bezug auf die Vertraulichkeit zu liefern?
+Begründen Sie, warum diese Veränderung zielführend ist.
+  \end{enumerate}
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Zufallszahlen bei der Schlüsselgenerierung}
+  Warum sind bei der Schlüsselgenerierung (echte) Zufallszahlen nötig?
+Weshalb eignet sich die XOR-Verknüpfung, wenn zur Erhöhung der Sicherheit mehrere Zufallszahlen zu einer einzelnen Zufallszahl verknüpft werden, etwa bei der Schlüsselgenerierung?
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Falsche Verwendung des One-Time-Pads}
+  Ihnen sind eine Reihe verschlüsselter deutscher Substantive in die Hände gefallen.
+Sie gehen davon aus, dass fahrlässigerweise alle Wörter mit demselben One-Time-Pad verschlüsselt wurden.
+Versuchen Sie, den Schlüssel zu ermitteln, indem Sie einen geeigneten Angriff implementieren.
+Die Schlüsseltexte (der 6 Substantive) in Dezimalschreibweise lauten:
+
+  \begin{center}
+      \texttt{09 00 04 10} \\
+      \texttt{10 20 28 09} \\
+      \texttt{10 16 02 02} \\
+      \texttt{10 20 05 08} \\
+      \texttt{26 26 03 00} \\
+      \texttt{28 16 03 17}
+  \end{center}
+
+  Hinweis: Das Passwort und die Substantive sind ASCII-codiert (Alphabet = \{A, B, C, ..., Z\}).
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Grundlagen der Kryptographie}
+  \begin{enumerate}
+      \item Erläutern Sie zwei Ursachen, die dafür sorgen, dass fast alle kryptographischen Algorithmen mit der Zeit durch andere ersetzt werden müssen.
+      \item Sie möchten einen Text so verschlüsseln, dass er auch in 99 Jahren nicht ohne Kenntnis des von Ihnen versteckten Schlüssels entschlüsselt werden kann.
+Welches Verfahren wählen Sie dazu?
+Begründen Sie Ihre Entscheidung.
+Warum wird das Verfahren heutzutage nicht überall benutzt?
+      \item In der Vorlesung wurden verschiedene klassische Chiffren behandelt, darunter Substitutionschiffren wie die Caesar-Chiffre (Verschiebechiffre), das Schema von Polybios, das Vigenère-Verfahren und die Vernman-Chiffre.
+Zudem wurden Transpositionschiffren wie die Skytala betrachtet.
+Ihnen liegt ein Schlüsseltext vor, der mit einer dieser Chiffren erzeugt wurde.
+Ihnen ist bekannt, dass es sich um die Verschlüsselung eines deutschen Textes handelt.
+Wie können Sie – ohne Rechnerunterstützung – vorgehen, um einzugrenzen, welche Chiffre verwendet wurde und erläutern Sie die jeweiligen Ansätze?
+  \end{enumerate}
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+% \setcounter{subsection}{30}
+% \begin{exercise}{}
+
+%   \begin{solution}
+%     % TODO
+%   \end{solution}
+% \end{exercise}
+
+% \setcounter{subsection}{40}
+% \begin{exercise}{}
+
+%   \begin{solution}
+%     % TODO
+%   \end{solution}
+% \end{exercise}
+
+% \setcounter{subsection}{50}
+% \begin{exercise}{}
+
+%   \begin{solution}
+%     % TODO
+%   \end{solution}
+% \end{exercise}
+
+% \setcounter{subsection}{60}
+% \begin{exercise}{}
+
+%   \begin{solution}
+%     % TODO
+%   \end{solution}
+% \end{exercise}
+
+% \setcounter{subsection}{70}
+% \begin{exercise}{}
+
+%   \begin{solution}
+%     % TODO
+%   \end{solution}
+% \end{exercise}
+
+% \setcounter{subsection}{80}
+% \begin{exercise}{}
+
+%   \begin{solution}
+%     % TODO
+%   \end{solution}
+% \end{exercise}
+
+% \setcounter{subsection}{90}
+% \begin{exercise}{}
+
+%   \begin{solution}
+%     % TODO
+%   \end{solution}
+% \end{exercise}
+
 \section{Catalogue}
+
+% Aufgabe 4.7 Spoofing
+% a) Inwiefern kann ein DNS-Spoofing-Angriff für einen Distributed Denial of Service (DDoS)-Angriff ausgenutzt werden? Erklären Sie anschaulich.
+% b) Welchen Angriff kann man in der nachfolgenden Ausgabe erkennen? Begründen Sie.
+% ?(192.168.1.7) at 00:1f:88:13:fe:f1 [ether] on eth0
+% ?(192.168.1.4) at 00:1f:88:13:fe:f1 [ether] on eth0
+% ?(192.168.1.1) at 00:1f:88:13:fe:f1 [ether] on eth0
+% ?(192.168.1.3) at 00:1f:88:13:fe:f1 [ether] on eth0
+% ?(192.168.1.2) at 00:1f:88:13:fe:f1 [ether] on eth0
+% c) Warum funktioniert ARP-Spoofing nur in einem lokalen Netz?
+% d) Sie vermuten, Opfer eines ARP-Spoofing-Angriffs zu sein. Was können Sie an Ihrem Rechner überprüfen, um einen solchen Angriff erkennen zu können?
+
+% Aufgabe 4.8 Kommunikationsmanipulation
+% Eve und Bob sind befreundet und wohnen im selben Haushalt. Ihre Computer sind an denselben Switch
+% angeschlossen und über diesen mit dem Internet verbunden. Eve hat bald Geburtstag und vermutet, dass Bob
+% ein Geschenk für Sie bei einem Onlineshop bestellen wird, der keine Verbindungsverschlüsselung einsetzt.
+% Einerseits ist sie sehr neugierig und möchte gern so früh wie möglich erfahren, was sie geschenkt bekommt.
+% Andererseits befürchtet sie, dass Bob ihr einen Stoffteddy schenken möchte, den sie nicht haben will. Sie
+% möchte daher Bobs Kommunikation abhören und, wenn nötig, seine Bestellung manipulieren können.
+% a) Nennen und erläutern Sie eine Angriffsart, mit der Eve den gesamten Netzverkehr von Bob abhören
+% und manipulieren kann.
+% b) Wie kann Bob sich vor dem in Teilaufgabe a) beschriebenen Angriff schützen?
+
+% Aufgabe 10.5 Indeterministische Verschlüsselung
+% a) Was bedeutet indeterministische Verschlüsselung?
+% b) Wie realisieren Sie die indeterministische Verschlüsselung bei einem asymmetrischen Konzelations-
+% system?
+% c) Was ändert sich gegenüber b) bei einem symmetrischen Konzelationssystem?
+% d) Sie sollen eine indeterministische Verschlüsselung realisieren. Zur Auswahl stehen die Algorithmen
+% RSA und ElGamal. Wie würden Sie bei beiden Algorithmen vorgehen?
+% e) Beim Einsatz asymmetrischer Verschlüsselung sollte eine Nachricht vor der Verschlüsselung noch um
+% (Pseudo-)Zufallszahlen ergänzt werden (indeterministische Verschlüsselung). Welcher Angriff kann da-
+% durch verhindert werden?
+
+% Aufgabe 10.6 Informationstheoretische Sicherheit
+% Das One-Time-Pad (Vernam-Chiffre) ist ein sehr einfaches, aber informationstheoretisch sicheres symme-
+% trisches Verschlüsselungsverfahren: Die Klartextbits xi werden einzeln XOR-verknüpft mit einer zufälligen
+% Schlüsselfolge ki gleicher Länge, die nur ein einziges Mal verwendet werden darf (Schlüsseltext si = xi ⊕ki
+% mit i= 1,2,...).
+% a) Was erfährt der Angreifer, der den Schlüsseltext abfängt, über den Klartext?
+% b) Wie würden Sie die Sicherheit des Verfahrens beweisen?
+% c) Warum darf die Schlüsselfolge nicht mehrmals verwendet werden? Überlegen Sie sich, was passieren
+% würde, wenn der Angreifer zwei Schlüsseltexte s1 und s2 (oder gar noch weitere) abfangen würde, die
+% unter dem gleichen Schlüssel k= k1 = k2 verschlüsselt wurden. Annahme: Es handelt sich um sinnvolle
+% Klartexte.
+% Es existiert auch ein informationstheoretisch sicheres (symmetrisches) Verfahren zur Authentifikation.
+% d) Welche Eigenschaft muss ein informationstheoretisch sicheres Authentifikationssystem besitzen?
+% e) Entwerfen Sie ein solches System und erläutern Sie dessen Funktionsweise sowie dessen Schutz
+% vor einem Angreifer und damit dessen informationstheoretische Sicherheit. Legen Sie eine bitweise
+% Authentisierung (ein MAC-Bit pro Klartextbit) zugrunde.
+% f) Kann es informationstheoretisch sichere asymmetrische Verfahren zur Verschlüsselung bzw. Authenti-
+% fikation geben? Begründen Sie Ihre Antwort.
+
+% Aufgabe 11.3 Unsicherheit der klassischen Verschiebechiffre
+% Ein klassisches Verschlüsselungsverfahren ist die Verschiebechiffre (Vigenère-Chiffre). Sie arbeitet typischer-
+% weise auf einem Alphabet aus den Großbuchstaben A-Z für den Klartext, den Schlüssel und den Schlüssel-
+% text. Zudem kann ein Zeichen zur Worttrennung hinzugenommen werden. Für diese Aufgabe wird dazu der
+% Unterstrich _ verwendet. Ein Schlüsseltextzeichen ergibt sich aus dem um das Schlüsselzeichen verschobene
+% Klartextzeichen. Dieses sogenannte Vigenère-Tableau ist in Abbildung 1 dargestellt.
+% a) Verschlüsseln Sie den nachfolgenden Klartext mit dem Schlüsselwort PRUEFUNG unter Verwendung
+% einer Vigenère-Chiffre:
+% GLUECKLICH_IST_DER_DER_DEN_DINGEN_AUF_DEN_GRUND_GEHEN_KONNTE
+% b) Erläutern Sie kurz zwei sinnvolle Ansätze zur Kryptanalyse, d. h. zum Brechen einer Vigenère-Chiffre.
+% c) Eine Verbesserung der Vigenère-Chiffre ergibt sich durch die Verlängerung des Schlüsselwortes. Um
+% welchen Faktor vergrößert sich dadurch der Schlüsselraum bei einem Alphabet mit 27 Elementen (A-Z
+% und Unterstrich als Worttrennung) und einer Verlängerung um sZeichen?
+% d) Wie müsste die Vigenère-Chiffre verändert werden, um informationstheoretische Sicherheit im Bezug
+% auf die Vertraulichkeit zu liefern? Begründen Sie, warum dieser Veränderung zielführend ist.
+% 23
+% A B C D E F G H I J K L M N O P Q R S T U V W X Y Z _
+% A A B C D E F G H I J K L M N O P Q R S T U V W X Y Z _
+% B B C D E F G H I J K L M N O P Q R S T U V W X Y Z _ A
+% C C D E F G H I J K L M N O P Q R S T U V W X Y Z _ A B
+% D D E F G H I J K L M N O P Q R S T U V W X Y Z _ A B C
+% E E F G H I J K L M N O P Q R S T U V W X Y Z _ A B C D
+% F F G H I J K L M N O P Q R S T U V W X Y Z _ A B C D E
+% G G H I J K L M N O P Q R S T U V W X Y Z _ A B C D E F
+% H H I J K L M N O P Q R S T U V W X Y Z _ A B C D E F G
+% I I J K L M N O P Q R S T U V W X Y Z _ A B C D E F G H
+% J J K L M N O P Q R S T U V W X Y Z _ A B C D E F G H I
+% K K L M N O P Q R S T U V W X Y Z _ A B C D E F G H I J
+% L L M N O P Q R S T U V W X Y Z _ A B C D E F G H I J K
+% M M N O P Q R S T U V W X Y Z _ A B C D E F G H I J K L
+% N N O P Q R S T U V W X Y Z _ A B C D E F G H I J K L M
+% O O P Q R S T U V W X Y Z _ A B C D E F G H I J K L M N
+% P P Q R S T U V W X Y Z _ A B C D E F G H I J K L M N O
+% Q Q R S T U V W X Y Z _ A B C D E F G H I J K L M N O P
+% R R S T U V W X Y Z _ A B C D E F G H I J K L M N O P Q
+% S S T U V W X Y Z _ A B C D E F G H I J K L M N O P Q R
+% T T U V W X Y Z _ A B C D E F G H I J K L M N O P Q R S
+% U U V W X Y Z _ A B C D E F G H I J K L M N O P Q R S T
+% V V W X Y Z _ A B C D E F G H I J K L M N O P Q R S T U
+% W W X Y Z _ A B C D E F G H I J K L M N O P Q R S T U V
+% X X Y Z _ A B C D E F G H I J K L M N O P Q R S T U V W
+% Y Y Z _ A B C D E F G H I J K L M N O P Q R S T U V W X
+% Z Z _ A B C D E F G H I J K L M N O P Q R S T U V W X Y
+% _ _ A B C D E F G H I J K L M N O P Q R S T U V W X Y Z
+% Abbildung 1: Vigenère-Tableau
+
+% Aufgabe 11.4 Klassische Chiffren und Gütekriterien
+% a) Verschlüsseln Sie den Klartext PRUEFUNGSSPASS_OHNE_KOMPROMISSE mit der Skytala und dem Schlüs-
+% sel k= 4 (4 Zeilen, 8 Spalten). Verschlüsseln Sie den Klartext STUDIUMSFREUDE mit der Caesar-Chiffre
+% und dem Schlüssel k= 3.
+% b) Die sogenannte Häufigkeitsanalyse kann als Angriff auf klassische Chiffren verwendet werden. Dabei
+% werden die Buchstabenhäufigkeiten eines Chiffretextes gezählt und mit den normalerweise vorkom-
+% menden Buchstabenhäufigkeiten von Klartexten verglichen. Damit können Rückschlüsse von den Chif-
+% fretextbuchstaben auf die Klartextbuchstaben gezogen werden. Für welche der beiden in Teilaufgabe a)
+% angesprochenen klassischen Chiffren ist dieser Angriff besonders schwerwiegend und kann zum direk-
+% ten Brechen einer Chiffre führen? Begründen Sie Ihre Antwort.
+% c) Sie kommen auf die Idee, beide Chiffren miteinander zu kombinieren, um eine größere Sicherheit zu
+% erreichen. Begründen Sie für die Variante, dass zuerst die Caesar-Chiffre auf den Klartext und dann
+% die Skytala auf den resultierenden Caesar-Chiffretext angewandt wird, ob diese Überlegung tatsächlich
+% stimmt. Ändert sich an Ihrer Überlegung etwas, wenn erst die Skytala und dann die Caesar-Chiffre
+% angewandt wird?
+% d) Nennen und erläutern Sie die Gütekriterien, die moderne symmetrische Chiffren erfüllen müssen, um
+% vor derartig einfachen Angriffen wie der Häufigkeitsanalyse zu schützen.
+
+% Aufgabe 12.6 Hashfunktionen und Signaturen
+% a) Handelt es sich bei der Identitätsfunktion H(x) = xum eine kryptographische Hashfunktion? Begründen
+% Sie.
+% b) Ist die Kollisionsfreiheit eine relevante Eigenschaft kryptographischer Hashfunktionen? Begründen Sie.
+% c) In der Praxis wird die digitale Signatur einer Nachricht mmeist durch Signieren des (kryptographischen)
+% Hashwertes der Nachricht H(m) und nicht der Nachricht selbst gebildet: Sigk(H(m)). Warum kann ein
+% Empfänger sich in diesem Fall auf die Authentizität der Nachricht verlassen? Welche Vorteile bringt
+% das Signieren des Hashwertes einer Nachricht mit sich? Betrachten Sie hierfür auch den Fall, dass
+% klassische RSA-Signaturen ohne Padding zum direkten Signieren von kurzen Nachrichten eingesetzt
+% würden.
+% 25
+% d) Eine Entwicklerin möchte eine von ihr entwickelte Software zur Verfügung stellen. Um den Datenverkehr
+% auf ihrem eigenen Server zu reduzieren, stellt sie das Softwarepaket über ein kommerzielles Content
+% Delivery Network (CDN) bereit. Sie möchte jedoch verhindern, dass ein böswilliger CDN-Betreiber die
+% von ihr bereitgestellte Software durch eine mit einer Hintertür versehene Version ersetzt. Entwerfen Sie
+% ein geeignetes Verfahren unter Nutzung von kryptographischen Hashfunktionen, das es Nutzenden er-
+% laubt, die Integrität des Softwarepakets zu überprüfen. Sie können annehmen, dass der CDN-Betreiber
+% keinen Zugriff auf den Server der Entwicklerin besitzt.
+% • Wie geht die Entwicklerin beim Veröffentlichen einer neuen Software-Version vor?
+% • Wie können Nutzer die Integrität überprüfen?
+% • Welche Eigenschaft einer kryptographischen Hashfunktion ist für die Sicherheit des Verfahrens
+% entscheidend? Begründen Sie.
+
+% Aufgabe 15.1 Das Feistel-Prinzip
+% a) Die nebenstehende Abbildung zeigt eine Runde der Feistel-
+% Verschlüsselung. Berechnen Sie zwei Runden der Verschlüsse-
+% lung und Entschlüsselung mit dem Feistel-Prinzip und den fol-
+% genden Werten:
+% • Klartext: 0011 0100
+% • Funktion F: binäres UND
+% • Rundenschlüssel K1: 1110
+% • Rundenschlüssel K2: 0001
+% Welche Eigenschaften sollte die Einwegfunktion F aufweisen, um
+% mit ihrer Hilfe eine sichere Feistel-Chiffre zu erzeugen?
+% Zeigen Sie, dass auch bei bei einer konstanten Einwegfunktion F
+% sowohl die Ver- als auch die Entschlüsselung funktioniert.
+% d) Benötigen Sie die Umkehrfunktion F−1 von F, damit die Ent-
+% schlüsselung klappt? Begründen Sie Ihre Antwort.
+
+% Aufgabe 16.3 RSA: Eigenschaften und Sicherheit
+% Für die Erzeugung eines RSA-Schlüsselpaares wurde von Alice der Modulus n= p·q= 7·17= 119 gewählt.
+% a) Was muss für den öffentlichen Verschlüsselungsexponenten c, der gemeinsam mit dem Modulus n
+% veröffentlicht wird, gelten, damit dieser als Teil des RSA-Schlüssels verwendet werden kann?
+% 34
+% b) Warum kann eine mit RSA zu verschlüsselnde Nachricht m nicht gleichzeitig einen Faktor p und q
+% enthalten?
+% c) Alice wählt den öffentlichen Verschlüsselungsexponenten c= 5. Berechnen Sie den zugehörigen ge-
+% heimen Entschlüsselungsexponenten dmithilfe des erweiterten euklidischen Algorithmus.
+% d) Bei naiver Anwendung des RSA-Verfahrens wird eine Nachricht m, für die 1 <m<ngelten muss, ohne
+% Veränderung mit dem öffentlichen Verschlüsselungsexponenten cpotenziert und modulo nreduziert.
+% Problematisch ist hierbei der multiplikative Homomorphismus des RSA-Verfahrens. Erläutern Sie kurz
+% diese Eigenschaft anhand zweier Nachrichten m1 und m2.
+% Hinweis: Zeichnen Sie ggf. ein passendes Diagramm. Sie müssen keine Berechnung durchführen!
+% e) Wie könnte die Angreiferin Eve die Eigenschaft des multiplikativen Homomorphismus ausnutzen, um an
+% die Klartextnachricht einer abgefangenen verschlüsselten Nachricht s, die an Alice adressiert war, zu
+% gelangen? Zeigen Sie das Vorgehen anhand der abgefangenen verschlüsselten Nachricht s= 101, der
+% eigens von Eve gewählten Zufallszahl r= 22 sowie dem multiplikativen Inversen der Zufallszahl r−1 =
+% 92 mod 119. Verwenden Sie auf Empfängerseite (Alice) den gegebenen RSA-Modulus, den öffentlichen
+% Verschlüsselungsexponenten c= 11 und den geheimen Entschlüsselungsexponenten d= 35, der nur
+% Alice, aber nicht Eve bekannt ist.
+% f) Nennen Sie ein Anwendungsbeispiel, bei dem der multiplikative Homomorphismus des RSA-Verfahrens
+% „ausgenutzt“ bzw. gewollt angewendet wird.
+
+% Aufgabe 16.10 Würfeln mit RSA
+% Bob möchte Alice über das Internet das Ergebnis eines Würfelwurfs senden. Bob verschlüsselt das Ergebnis
+% vor der Übertragung mit dem deterministischen RSA-Verfahren. Alice hat bereits ein RSA-Schlüsselpaar er-
+% zeugt und dabei die folgenden Parameter verwendet: pA = 5, qA = 11, eA = 3, dA = 27. Bob hat ebenfalls
+% ein RSA-Schlüsselpaar erzeugt und dabei folgende Parameter verwendet: pB = 17, qB = 5, eB = 3, dB = 43.
+% Die öffentlichen Schlüssel haben die beiden bereits über einen sicheren Kanal ausgetauscht.
+% a) Bob hat gewürfelt und möchte das Ergebnis nun an Alice senden. Er überträgt dazu cB = 9. Zeigen Sie,
+% dass eine passive Angreiferin (Eve), die lediglich die öffentlichen Schlüssel und cB kennt, mittels eines
+% Chosen-Plaintext-Angriffs Bobs Wurfergebnis mB ermitteln kann. Welche Zahl hat Bob gewürfelt?
+% b) Wie kann sich Bob vor diesem Angriff schützen?
+
+% Aufgabe 16.11 Forward Secrecy mit RSA
+% Zur Erreichung von Forward Secrecy erzeugt Alice auf ihrem Server mit dem RSA-Verfahren einen geheimen
+% Schlüssel dsowie den dazu passenden öffentlichen Schlüssel c, den sie persönlich an ihre Kunden verteilt.
+% Zusätzlich erzeugt Alices Server bei jedem Verbindungsaufbau ein neues kurzlebiges (engl. ephemeral) RSA-
+% Schlüsselpaar, bestehend aus einem öffentlichen Schlüssel ce und einem geheimen Schlüssel de. Im Ver-
+% gleich zum bisherigen Verfahren verteilt Alice ce nicht mehr persönlich. Stattdessen signiert ihr Server mit
+% ihrem ursprünglichen privaten Schlüssel dden neuen öffentlichen Schlüssel ce und schickt diesen zusam-
+% men mit der Signatur über die aufgebaute Verbindung an den Kunden. Der Kunde prüft anschließend die
+% Signatur mithilfe des ursprünglichen von Alice erhaltenen öffentlichen Schlüssels cund verwendet bei korrek-
+% ter Signatur ce zum Verschlüsseln des zufällig generierten symmetrischen Schlüssels k. Das Verschlüsseln
+% der Nachrichten mit kerfolgt mit AES. Nach Verbindungsende löscht Alices Server de, ce und k.
+% a) Welche Verbindungen könnte ein sehr starker Angreifer (etwa ein Nachrichtendienst) noch entschlüs-
+% seln, wenn er an dgelangt? Begründen Sie kurz ihre Antwort.
+% b) Skizzieren Sie einen Angriff, der möglich wäre, wenn der Kunde die Signatur nicht prüft.
+% 37
+% c) Alice ist noch nicht ganz zufrieden mit dem neuen Verfahren: Es ist ihr zu langsam. Sie nimmt nun
+% folgende Änderung vor: Statt jedes mal bei der Erzeugung von ce und de neue Primzahlen pund qfür
+% den Modulus Ne zu generieren, verwendet sie immer die gleichen Primzahlen. Nun wählt sie lediglich
+% den öffentlichen Exponenten aus ce neu und berechnet einen dazugehörigen geheimen Exponenten
+% de. Warum ist dies keine gute Idee? Zeigen Sie, wie sich dieses geänderte Verfahren angreifen lässt,
+% um aufgezeichnete Verbindungen zu entschlüsseln.
+
+% Aufgabe 16.16 Elliptische Kurven: EC-ElGamal
+% Das ElGamal-Kryptosystem lässt sich auch auf elliptische Kurven übertragen. Dabei werden die Rechen-
+% operationen mittels Punkten auf einer elliptischen Kurve vorgenommen. Betrachten Sie für die folgenden
+% Teilaufgaben die elliptische Kurve E(GF(p)) mit der Form y2 ≡x3 + x+ 6 (mod 11).
+% a) Passen Sie die Operationen des ElGamal-Kryptosystems zur Schlüsselgenerierung, zur Entschlüsse-
+% lung sowie zur Verschlüsselung an, sodass diese auf der Basis elliptischer Kurven operieren.
+% b) Generieren Sie mithilfe Ihres Verfahrens aus Teilaufgabe a) ein geeignetes Schlüsselpaar auf der oben
+% angegebenen elliptischen Kurve. Berechnen Sie dann die Verschlüsselung cder Nachricht m= (5,2).
+% c) Berechnen Sie die Entschlüsselung des Schlüsseltextes caus der vorherigen Teilaufgabe b) mit dem
+% dort generierten geheimen Schlüssel.
+% d) Zeigen sie anschaulich, auf welcher Annahme die Sicherheit des Verfahrens beruht. Welcher Ausdruck
+% müsste berechnet werden, damit Angreifende Schlüsseltexte ohne Kenntnis des geheimen Schlüssels
+% leicht entschlüsseln könnten?
+
+% Aufgabe 22.9 Verbindungsverschlüsselung in GSM
+% Welche Kommunikationsabschnitte sind in GSM verschlüsselt? Welche Kommunikationsabschnitte sind un-
+% verschlüsselt?
+
+% Aufgabe 22.27 Bluetooth-Authentifikation
+% Bei Bluetooth wird auf Basis der Funktion E22 die Authentifikation durchgeführt. Was sind die In- und Outputs
+% der Authentifikation?
+
+% Aufgabe 22.29 Mehrfachnutzung des IV in einem WEP-verschlüsselten WLAN
+% a) Der in WEP verwendete Initialisierungsvektor (IV) ist 24 Bit groß und steht in einem Paket P am Paketan-
+% fang vor dem Schlüsseltext. Wie viele Pakete können maximal erzeugt werden, wenn kein IV mehrfach
+% verwendet werden soll? Wie sind die IVs von einem Teilnehmer des WLANs zu erzeugen, sodass eine
+% Kollision möglichst spät auftritt? Welches Problem gibt es bei diesem Verfahren, wenn mehrere Teilneh-
+% mer kommunizieren?
+% b) Nehmen Sie an, Sie haben die Pakete P1 und P2 empfangen:
+% P1: 00010 00111 11111 11110 11001 01001 00101 ...
+% P2: 00010 00111 11111 11110 11000 10011 00010 ...
+% Die ersten 10 Klartextbits des ersten Paketes P1 lauten 10010 01100. Bestimmen Sie die ersten 10
+% Klartextbits des zweiten Paketes P2.
+
+\setcounter{section}{2023}
+\section{Mock exam}
 
@@
 
-\section[2024]{Mock exam}
-
-% TODO
-
-\section[2023]{2nd exam}
+\setcounter{section}{2022}
+\section{2nd exam}
 
@@
 
-\section[2022]{1st exam}
+\setcounter{section}{2021}
+\section{1st exam}
 

### 2024-10-23 SbD/exercises.tex
Added sbd catalogue exercises.

--- 2024-10-21 SbD/exercises.tex
@@
 
+\section{Assignments}
+
+\setcounter{subsection}{10}
+\begin{exercise}{Skytala}
+  Bei der Skytala handelt es sich um eine Transpositionschiffre, bei der ursprünglich ein Zylinder mit einem Papierstreifen umwickelt wurde.
+Die Ver- und Entschlüsselung kann aber auch mithilfe einer Matrix dargestellt werden.
+  \begin{enumerate}
+      \item Verschlüsseln Sie den folgenden Klartext mit einer Skytala, die durch eine Matrix mit sieben Zeilen und neun Spalten beschrieben werden kann:
+      \begin{center}
+          \texttt{GLUECKLICH\_IST\_DER\_DER\_DEN\_DINGEN\_AUF\_DEN\_GRUND\_GEHEN\_KONNTE}
+      \end{center}
+      \item Was ist der geheim zu haltende symmetrische Schlüssel der Skytala aus Teilaufgabe a) und wie hängt dieser mit der ursprünglichen Variante mit Zylinder und Papierstreifen zusammen?
+      \item Erläutern Sie kurz zwei sinnvolle Ansätze zur Kryptanalyse, d. h. Ansätze zum Brechen einer Skytala.
+  \end{enumerate}
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Spalten-Transpositionen}
+  Bei der Skytala wird ein Zylinder mit einem Papierstreifen umwickelt, beschrieben und wieder abgewickelt.
+  \begin{enumerate}
+      \item Eine Verbesserung dieser Spalten-Transposition ergibt sich, wenn zusätzlich auch die Spalten in ihrer Reihenfolge vertauscht werden (was beim Papierstreifen natürlich nicht gelingt, wohl aber im Rechner).
+Um welchen Faktor vergrößert sich dadurch der Schlüsselraum bei $s$ Spalten?
+      \item Schreiben Sie eine Skytala mit 4 Zeilen und 3 Spalten (ohne Spaltenvertauschung) in Zyklenschreibweise.
+  \end{enumerate}
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\setcounter{subsection}{20}
+\begin{exercise}{Vigenère-Chiffre}
+  Ein klassisches Verschlüsselungsverfahren ist die Vigenère-Chiffre.
+Sie arbeitet typischerweise auf einem Alphabet aus den Großbuchstaben A-Z für den Klartext, den Schlüssel und den Schlüsseltext.
+Zudem kann ein Zeichen zur Worttrennung hinzugenommen werden.
+Für diese Aufgabe wird dazu der Unterstrich \texttt{\_} verwendet.
+Ein Schlüsseltextzeichen ergibt sich aus dem um das jeweilige Zeichen des Schlüssels verschobene Klartextzeichen.
+Diese Verschiebung kann durch das sogenannte Vigenère-Tableau in Abbildung 1 dargestellt werden.
+
+  % TODO: add table
+  % A B C D E F G H I J K L M N O P Q R S T U V W X Y Z _
+  % A A B C D E F G H I J K L M N O P Q R S T U V W X Y Z _
+  % B B C D E F G H I J K L M N O P Q R S T U V W X Y Z _ A
+  % C C D E F G H I J K L M N O P Q R S T U V W X Y Z _ A B
+  % D D E F G H I J K L M N O P Q R S T U V W X Y Z _ A B C
+  % E E F G H I J K L M N O P Q R S T U V W X Y Z _ A B C D
+  % F F G H I J K L M N O P Q R S T U V W X Y Z _ A B C D E
+  % G G H I J K L M N O P Q R S T U V W X Y Z _ A B C D E F
+  % H H I J K L M N O P Q R S T U V W X Y Z _ A B C D E F G
+  % I I J K L M N O P Q R S T U V W X Y Z _ A B C D E F G H
+  % J J K L M N O P Q R S T U V W X Y Z _ A B C D E F G H I
+  % K K L M N O P Q R S T U V W X Y Z _ A B C D E F G H I J
+  % L L M N O P Q R S T U V W X Y Z _ A B C D E F G H I J K
+  % M M N O P Q R S T U V W X Y Z _ A B C D E F G H I J K L
+  % N N O P Q R S T U V W X Y Z _ A B C D E F G H I J K L M
+  % O O P Q R S T U V W X Y Z _ A B C D E F G H I J K L M N
+  % P P Q R S T U V W X Y Z _ A B C D E F G H I J K L M N O
+  % Q Q R S T U V W X Y Z _ A B C D E F G H I J K L M N O P
+  % R R S T U V W X Y Z _ A B C D E F G H I J K L M N O P Q
+  % S S T U V W X Y Z _ A B C D E F G H I J K L M N O P Q R
+  % T T U V W X Y Z _ A B C D E F G H I J K L M N O P Q R S
+  % U U V W X Y Z _ A B C D E F G H I J K L M N O P Q R S T
+  % V V W X Y Z _ A B C D E F G H I J K L M N O P Q R S T U
+  % W W X Y Z _ A B C D E F G H I J K L M N O P Q R S T U V
+  % X X Y Z _ A B C D E F G H I J K L M N O P Q R S T U V W
+  % Y Y Z _ A B C D E F G H I J K L M N O P Q R S T U V W X
+  % Z Z _ A B C D E F G H I J K L M N O P Q R S T U V W X Y
+  % _ _ A B C D E F G H I J K L M N O P Q R S T U V W X Y Z
+  % Abbildung 1: Vigenère-Tableau
+
+  \begin{enumerate}
+      \item Verschlüsseln Sie den nachfolgenden Klartext mit dem Schlüssel \texttt{PRUEFUNG} unter Verwendung einer Vigenère-Chiffre:
+      \begin{center}
+          \texttt{GLUECKLICH\_IST\_DER\_DER\_DEN\_DINGEN\_AUF\_DEN\_GRUND\_GEHEN\_KONNTE}
+      \end{center}
+      \item Erläutern Sie kurz zwei sinnvolle Ansätze zur Kryptanalyse, d. h. zum Brechen einer Vigenère-Chiffre.
+      \item Eine Verbesserung der Vigenère-Chiffre ergibt sich durch die Verlängerung des Schlüssels.
+Um welchen Faktor vergrößert sich dadurch der Schlüsselraum bei einem Alphabet mit 27 Elementen (A-Z und Unterstrich als Worttrennung) und einer Verlängerung um $s$ Zeichen?
+      \item Wie müsste die Vigenère-Chiffre verändert werden, um informationstheoretische Sicherheit im Bezug auf die Vertraulichkeit zu liefern?
+Begründen Sie, warum diese Veränderung zielführend ist.
+  \end{enumerate}
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Zufallszahlen bei der Schlüsselgenerierung}
+  Warum sind bei der Schlüsselgenerierung (echte) Zufallszahlen nötig?
+Weshalb eignet sich die XOR-Verknüpfung, wenn zur Erhöhung der Sicherheit mehrere Zufallszahlen zu einer einzelnen Zufallszahl verknüpft werden, etwa bei der Schlüsselgenerierung?
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Falsche Verwendung des One-Time-Pads}
+  Ihnen sind eine Reihe verschlüsselter deutscher Substantive in die Hände gefallen.
+Sie gehen davon aus, dass fahrlässigerweise alle Wörter mit demselben One-Time-Pad verschlüsselt wurden.
+Versuchen Sie, den Schlüssel zu ermitteln, indem Sie einen geeigneten Angriff implementieren.
+Die Schlüsseltexte (der 6 Substantive) in Dezimalschreibweise lauten:
+
+  \begin{center}
+      \texttt{09 00 04 10} \\
+      \texttt{10 20 28 09} \\
+      \texttt{10 16 02 02} \\
+      \texttt{10 20 05 08} \\
+      \texttt{26 26 03 00} \\
+      \texttt{28 16 03 17}
+  \end{center}
+
+  Hinweis: Das Passwort und die Substantive sind ASCII-codiert (Alphabet = \{A, B, C, ..., Z\}).
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Grundlagen der Kryptographie}
+  \begin{enumerate}
+      \item Erläutern Sie zwei Ursachen, die dafür sorgen, dass fast alle kryptographischen Algorithmen mit der Zeit durch andere ersetzt werden müssen.
+      \item Sie möchten einen Text so verschlüsseln, dass er auch in 99 Jahren nicht ohne Kenntnis des von Ihnen versteckten Schlüssels entschlüsselt werden kann.
+Welches Verfahren wählen Sie dazu?
+Begründen Sie Ihre Entscheidung.
+Warum wird das Verfahren heutzutage nicht überall benutzt?
+      \item In der Vorlesung wurden verschiedene klassische Chiffren behandelt, darunter Substitutionschiffren wie die Caesar-Chiffre (Verschiebechiffre), das Schema von Polybios, das Vigenère-Verfahren und die Vernman-Chiffre.
+Zudem wurden Transpositionschiffren wie die Skytala betrachtet.
+Ihnen liegt ein Schlüsseltext vor, der mit einer dieser Chiffren erzeugt wurde.
+Ihnen ist bekannt, dass es sich um die Verschlüsselung eines deutschen Textes handelt.
+Wie können Sie – ohne Rechnerunterstützung – vorgehen, um einzugrenzen, welche Chiffre verwendet wurde und erläutern Sie die jeweiligen Ansätze?
+  \end{enumerate}
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+% \setcounter{subsection}{30}
+% \begin{exercise}{}
+
+%   \begin{solution}
+%     % TODO
+%   \end{solution}
+% \end{exercise}
+
+% \setcounter{subsection}{40}
+% \begin{exercise}{}
+
+%   \begin{solution}
+%     % TODO
+%   \end{solution}
+% \end{exercise}
+
+% \setcounter{subsection}{50}
+% \begin{exercise}{}
+
+%   \begin{solution}
+%     % TODO
+%   \end{solution}
+% \end{exercise}
+
+% \setcounter{subsection}{60}
+% \begin{exercise}{}
+
+%   \begin{solution}
+%     % TODO
+%   \end{solution}
+% \end{exercise}
+
+% \setcounter{subsection}{70}
+% \begin{exercise}{}
+
+%   \begin{solution}
+%     % TODO
+%   \end{solution}
+% \end{exercise}
+
+% \setcounter{subsection}{80}
+% \begin{exercise}{}
+
+%   \begin{solution}
+%     % TODO
+%   \end{solution}
+% \end{exercise}
+
+% \setcounter{subsection}{90}
+% \begin{exercise}{}
+
+%   \begin{solution}
+%     % TODO
+%   \end{solution}
+% \end{exercise}
+
 \section{Catalogue}
+
+\setcounter{subsection}{46}
+\begin{exercise}{Spoofing}
+  \begin{enumerate}
+    \item Inwiefern kann ein DNS-Spoofing-Angriff für einen Distributed Denial of Service (DDoS)-Angriff ausgenutzt werden? Erklären Sie anschaulich.
+    \item Welchen Angriff kann man in der nachfolgenden Ausgabe erkennen? Begründen Sie.
+      \begin{center}
+        \texttt{?(192.168.1.7) at 00:1f:88:13:fe:f1 [ether] on eth0} \\
+        \texttt{?(192.168.1.4) at 00:1f:88:13:fe:f1 [ether] on eth0} \\
+        \texttt{?(192.168.1.1) at 00:1f:88:13:fe:f1 [ether] on eth0} \\
+        \texttt{?(192.168.1.3) at 00:1f:88:13:fe:f1 [ether] on eth0} \\
+        \texttt{?(192.168.1.2) at 00:1f:88:13:fe:f1 [ether] on eth0}
+      \end{center}
+    \item Warum funktioniert ARP-Spoofing nur in einem lokalen Netz?
+    \item Sie vermuten, Opfer eines ARP-Spoofing-Angriffs zu sein. Was können Sie an Ihrem Rechner überprüfen, um einen solchen Angriff erkennen zu können?
+  \end{enumerate}
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Kommunikationsmanipulation}
+  Eve und Bob sind befreundet und wohnen im selben Haushalt.
+Ihre Computer sind an denselben Switch angeschlossen und über diesen mit dem Internet verbunden.
+Eve hat bald Geburtstag und vermutet, dass Bob ein Geschenk für Sie bei einem Onlineshop bestellen wird, der keine Verbindungsverschlüsselung einsetzt.
+Einerseits ist sie sehr neugierig und möchte gern so früh wie möglich erfahren, was sie geschenkt bekommt.
+Andererseits befürchtet sie, dass Bob ihr einen Stoffteddy schenken möchte, den sie nicht haben will.
+Sie möchte daher Bobs Kommunikation abhören und, wenn nötig, seine Bestellung manipulieren können.
+  \begin{enumerate}
+    \item Nennen und erläutern Sie eine Angriffsart, mit der Eve den gesamten Netzverkehr von Bob abhören und manipulieren kann.
+    \item Wie kann Bob sich vor dem in Teilaufgabe a) beschriebenen Angriff schützen?
+  \end{enumerate}
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\setcounter{subsection}{104}
+\begin{exercise}{Indeterministische Verschlüsselung}
+  \begin{enumerate}
+    \item Was bedeutet indeterministische Verschlüsselung?
+    \item Wie realisieren Sie die indeterministische Verschlüsselung bei einem asymmetrischen Konzelations- system?
+    \item Was ändert sich gegenüber b) bei einem symmetrischen Konzelationssystem?
+    \item Sie sollen eine indeterministische Verschlüsselung realisieren. Zur Auswahl stehen die Algorithmen RSA und ElGamal. Wie würden Sie bei beiden Algorithmen vorgehen?
+    \item Beim Einsatz asymmetrischer Verschlüsselung sollte eine Nachricht vor der Verschlüsselung noch um (Pseudo-)Zufallszahlen ergänzt werden (indeterministische Verschlüsselung).
+Welcher Angriff kann dadurch verhindert werden?
+  \end{enumerate}
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Informationstheoretische Sicherheit}
+  Das One-Time-Pad (Vernam-Chiffre) ist ein sehr einfaches, aber informationstheoretisch sicheres symme- trisches Verschlüsselungsverfahren: Die Klartextbits $x_i$ werden einzeln XOR-verknüpft mit einer zufälligen Schlüsselfolge $k_i$ gleicher Länge, die nur ein einziges Mal verwendet werden darf (Schlüsseltext $s_i = x_i \oplus k_i$ mit $i= 1,2,...$).
+  \begin{enumerate}
+    \item Was erfährt der Angreifer, der den Schlüsseltext abfängt, über den Klartext?
+    \item Wie würden Sie die Sicherheit des Verfahrens beweisen?
+    \item Warum darf die Schlüsselfolge nicht mehrmals verwendet werden?
+Überlegen Sie sich, was passieren würde, wenn der Angreifer zwei Schlüsseltexte $s_1$ und $s_2$ (oder gar noch weitere) abfangen würde, die unter dem gleichen Schlüssel $k= k_1 = k_2$ verschlüsselt wurden (Annahme: Es handelt sich um sinnvolle Klartexte).
+    \item Es existiert auch ein informationstheoretisch sicheres (symmetrisches) Verfahren zur Authentifikation.
+Welche Eigenschaft muss ein informationstheoretisch sicheres Authentifikationssystem besitzen?
+    \item Entwerfen Sie ein solches System und erläutern Sie dessen Funktionsweise sowie dessen Schutz vor einem Angreifer und damit dessen informationstheoretische Sicherheit.
+Legen Sie eine bitweise Authentisierung (ein MAC-Bit pro Klartextbit) zugrunde.
+    \item Kann es informationstheoretisch sichere asymmetrische Verfahren zur Verschlüsselung bzw. Authenti- fikation geben? Begründen Sie Ihre Antwort.
+  \end{enumerate}
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\setcounter{subsection}{112}
+\begin{exercise}{Unsicherheit der klassischen Verschiebechiffre}
+  Ein klassisches Verschlüsselungsverfahren ist die Verschiebechiffre (Vigenère-Chiffre).
+Sie arbeitet typischer- weise auf einem Alphabet aus den Großbuchstaben A-Z für den Klartext, den Schlüssel und den Schlüssel- text.
+Zudem kann ein Zeichen zur Worttrennung hinzugenommen werden.
+Für diese Aufgabe wird dazu der Unterstrich \texttt{\_} verwendet.
+Ein Schlüsseltextzeichen ergibt sich aus dem um das Schlüsselzeichen verschobene Klartextzeichen.
+Dieses sogenannte Vigenère-Tableau ist in Abbildung 1 dargestellt.
+  % TODO: add table
+  % A B C D E F G H I J K L M N O P Q R S T U V W X Y Z _
+  % A A B C D E F G H I J K L M N O P Q R S T U V W X Y Z _
+  % B B C D E F G H I J K L M N O P Q R S T U V W X Y Z _ A
+  % C C D E F G H I J K L M N O P Q R S T U V W X Y Z _ A B
+  % D D E F G H I J K L M N O P Q R S T U V W X Y Z _ A B C
+  % E E F G H I J K L M N O P Q R S T U V W X Y Z _ A B C D
+  % F F G H I J K L M N O P Q R S T U V W X Y Z _ A B C D E
+  % G G H I J K L M N O P Q R S T U V W X Y Z _ A B C D E F
+  % H H I J K L M N O P Q R S T U V W X Y Z _ A B C D E F G
+  % I I J K L M N O P Q R S T U V W X Y Z _ A B C D E F G H
+  % J J K L M N O P Q R S T U V W X Y Z _ A B C D E F G H I
+  % K K L M N O P Q R S T U V W X Y Z _ A B C D E F G H I J
+  % L L M N O P Q R S T U V W X Y Z _ A B C D E F G H I J K
+  % M M N O P Q R S T U V W X Y Z _ A B C D E F G H I J K L
+  % N N O P Q R S T U V W X Y Z _ A B C D E F G H I J K L M
+  % O O P Q R S T U V W X Y Z _ A B C D E F G H I J K L M N
+  % P P Q R S T U V W X Y Z _ A B C D E F G H I J K L M N O
+  % Q Q R S T U V W X Y Z _ A B C D E F G H I J K L M N O P
+  % R R S T U V W X Y Z _ A B C D E F G H I J K L M N O P Q
+  % S S T U V W X Y Z _ A B C D E F G H I J K L M N O P Q R
+  % T T U V W X Y Z _ A B C D E F G H I J K L M N O P Q R S
+  % U U V W X Y Z _ A B C D E F G H I J K L M N O P Q R S T
+  % V V W X Y Z _ A B C D E F G H I J K L M N O P Q R S T U
+  % W W X Y Z _ A B C D E F G H I J K L M N O P Q R S T U V
+  % X X Y Z _ A B C D E F G H I J K L M N O P Q R S T U V W
+  % Y Y Z _ A B C D E F G H I J K L M N O P Q R S T U V W X
+  % Z Z _ A B C D E F G H I J K L M N O P Q R S T U V W X Y
+  % _ _ A B C D E F G H I J K L M N O P Q R S T U V W X Y Z
+  % Abbildung 1: Vigenère-Tableau
+  \begin{enumerate}
+    \item Verschlüsseln Sie den nachfolgenden Klartext mit dem Schlüsselwort PRUEFUNG unter Verwendung einer Vigenère-Chiffre:
+    \begin{center}
+        \texttt{GLUECKLICH\_IST\_DER\_DER\_DEN\_DINGEN\_AUF\_DEN\_GRUND\_GEHEN\_KONNTE}
+    \end{center}
+    \item Erläutern Sie kurz zwei sinnvolle Ansätze zur Kryptanalyse, d. h. zum Brechen einer Vigenère-Chiffre.
+    \item Eine Verbesserung der Vigenère-Chiffre ergibt sich durch die Verlängerung des Schlüsselwortes.
+Um welchen Faktor vergrößert sich dadurch der Schlüsselraum bei einem Alphabet mit 27 Elementen (A-Z und Unterstrich als Worttrennung) und einer Verlängerung um $s$ Zeichen?
+    \item Wie müsste die Vigenère-Chiffre verändert werden, um informationstheoretische Sicherheit im Bezug auf die Vertraulichkeit zu liefern? Begründen Sie, warum dieser Veränderung zielführend ist.
+  \end{enumerate}
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Klassische Chiffren und Gütekriterien}
+  \begin{enumerate}
+    \item Verschlüsseln Sie den Klartext PRUEFUNGSSPASS\_OHNE\_KOMPROMISSE mit der Skytala und dem Schlüssel $k= 4$ (4 Zeilen, 8 Spalten).
+Verschlüsseln Sie den Klartext STUDIUMSFREUDE mit der Caesar-Chiffre und dem Schlüssel $k= 3$.
+    \item Die sogenannte Häufigkeitsanalyse kann als Angriff auf klassische Chiffren verwendet werden.
+Dabei werden die Buchstabenhäufigkeiten eines Chiffretextes gezählt und mit den normalerweise vorkommenden Buchstabenhäufigkeiten von Klartexten verglichen.
+Damit können Rückschlüsse von den Chiffretextbuchstaben auf die Klartextbuchstaben gezogen werden.
+Für welche der beiden in Teilaufgabe a) angesprochenen klassischen Chiffren ist dieser Angriff besonders schwerwiegend und kann zum direkten Brechen einer Chiffre führen?
+Begründen Sie Ihre Antwort.
+    \item Sie kommen auf die Idee, beide Chiffren miteinander zu kombinieren, um eine größere Sicherheit zu erreichen.
+Begründen Sie für die Variante, dass zuerst die Caesar-Chiffre auf den Klartext und dann die Skytala auf den resultierenden Caesar-Chiffretext angewandt wird, ob diese Überlegung tatsächlich stimmt.
+Ändert sich an Ihrer Überlegung etwas, wenn erst die Skytala und dann die Caesar-Chiffre angewandt wird?
+    \item Nennen und erläutern Sie die Gütekriterien, die moderne symmetrische Chiffren erfüllen müssen, um vor derartig einfachen Angriffen wie der Häufigkeitsanalyse zu schützen.
+  \end{enumerate}
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\setcounter{subsection}{125}
+\begin{exercise}{Hashfunktionen und Signaturen}
+  \begin{enumerate}
+    \item Handelt es sich bei der Identitätsfunktion $H(x) = x$ um eine kryptographische Hashfunktion? Begründen Sie.
+    \item Ist die Kollisionsfreiheit eine relevante Eigenschaft kryptographischer Hashfunktionen? Begründen Sie.
+    \item In der Praxis wird die digitale Signatur einer Nachricht $m$ meist durch Signieren des (kryptographischen) Hashwertes der Nachricht $H(m)$ und nicht der Nachricht selbst gebildet: $\text{Sig}_k(H(m))$.
+Warum kann ein Empfänger sich in diesem Fall auf die Authentizität der Nachricht verlassen?
+Welche Vorteile bringt das Signieren des Hashwertes einer Nachricht mit sich?
+Betrachten Sie hierfür auch den Fall, dass klassische RSA-Signaturen ohne Padding zum direkten Signieren von kurzen Nachrichten eingesetzt würden.
+    \item Eine Entwicklerin möchte eine von ihr entwickelte Software zur Verfügung stellen.
+Um den Datenverkehr auf ihrem eigenen Server zu reduzieren, stellt sie das Softwarepaket über ein kommerzielles Content Delivery Network (CDN) bereit.
+Sie möchte jedoch verhindern, dass ein böswilliger CDN-Betreiber die von ihr bereitgestellte Software durch eine mit einer Hintertür versehene Version ersetzt.
+Entwerfen Sie ein geeignetes Verfahren unter Nutzung von kryptographischen Hashfunktionen, das es Nutzenden erlaubt, die Integrität des Softwarepakets zu überprüfen.
+Sie können annehmen, dass der CDN-Betreiber keinen Zugriff auf den Server der Entwicklerin besitzt.
+      \begin{itemize}
+        \item Wie geht die Entwicklerin beim Veröffentlichen einer neuen Software-Version vor?
+        \item Wie können Nutzer die Integrität überprüfen?
+        \item Welche Eigenschaft einer kryptographischen Hashfunktion ist für die Sicherheit des Verfahrens entscheidend? Begründen Sie.
+      \end{itemize}
+  \end{enumerate}
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\setcounter{subsection}{150}
+\begin{exercise}{Das Feistel-Prinzip}
+  \begin{enumerate}
+    \item Die nebenstehende Abbildung zeigt eine Runde der Feistel-Verschlüsselung. Berechnen Sie zwei Runden der Verschlüsselung und Entschlüsselung mit dem Feistel-Prinzip und den folgenden Werten:
+    \begin{itemize}
+      \item Klartext: 0011 0100
+      \item Funktion F: binäres UND
+      \item Rundenschlüssel K1: 1110
+      \item Rundenschlüssel K2: 0001
+    \end{itemize}
+    \item Welche Eigenschaften sollte die Einwegfunktion F aufweisen, um mit ihrer Hilfe eine sichere Feistel-Chiffre zu erzeugen?
+    \item Zeigen Sie, dass auch bei bei einer konstanten Einwegfunktion F sowohl die Ver- als auch die Entschlüsselung funktioniert.
+    \item Benötigen Sie die Umkehrfunktion $F^{-1}$ von $F$, damit die Entschlüsselung klappt? Begründen Sie Ihre Antwort.
+  \end{enumerate}
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\setcounter{subsection}{162}
+\begin{exercise}{RSA: Eigenschaften und Sicherheit}
+  Für die Erzeugung eines RSA-Schlüsselpaares wurde von Alice der Modulus $n= p \cdot q= 7 \cdot 17= 119$ gewählt.
+  \begin{enumerate}
+    \item Was muss für den öffentlichen Verschlüsselungsexponenten $c$, der gemeinsam mit dem Modulus $n$ veröffentlicht wird, gelten, damit dieser als Teil des RSA-Schlüssels verwendet werden kann?
+    \item Warum kann eine mit RSA zu verschlüsselnde Nachricht $m$ nicht gleichzeitig einen Faktor $p$ und $q$ enthalten?
+    \item Alice wählt den öffentlichen Verschlüsselungsexponenten $c= 5$. Berechnen Sie den zugehörigen geheimen Entschlüsselungsexponenten $d$ mithilfe des erweiterten euklidischen Algorithmus.
+    \item Bei naiver Anwendung des RSA-Verfahrens wird eine Nachricht $m$, für die $1 < m < n$ gelten muss, ohne Veränderung mit dem öffentlichen Verschlüsselungsexponenten $c$ potenziert und modulo $n$ reduziert.
+Problematisch ist hierbei der multiplikative Homomorphismus des RSA-Verfahrens.
+Erläutern Sie kurz diese Eigenschaft anhand zweier Nachrichten $m_1$ und $m_2$.
+    \item Wie könnte die Angreiferin Eve die Eigenschaft des multiplikativen Homomorphismus ausnutzen, um an die Klartextnachricht einer abgefangenen verschlüsselten Nachricht $s$, die an Alice adressiert war, zu gelangen?
+Zeigen Sie das Vorgehen anhand der abgefangenen verschlüsselten Nachricht $s= 101$, der eigens von Eve gewählten Zufallszahl $r= 22$ sowie dem multiplikativen Inversen der Zufallszahl $r^{-1} = 92 \mod 119$.
+Verwenden Sie auf Empfängerseite (Alice) den gegebenen RSA-Modulus, den öffentlichen Verschlüsselungsexponenten $c= 11$ und den geheimen Entschlüsselungsexponenten $d= 35$, der nur Alice, aber nicht Eve bekannt ist.
+    \item Nennen Sie ein Anwendungsbeispiel, bei dem der multiplikative Homomorphismus des RSA-Verfahrens "ausgenutzt" bzw. gewollt angewendet wird.
+  \end{enumerate}
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\setcounter{subsection}{1609}
+\begin{exercise}{Würfeln mit RSA}
+  Bob möchte Alice über das Internet das Ergebnis eines Würfelwurfs senden.
+Bob verschlüsselt das Ergebnis vor der Übertragung mit dem deterministischen RSA-Verfahren.
+Alice hat bereits ein RSA-Schlüsselpaar erzeugt und dabei die folgenden Parameter verwendet: $p_A = 5$, $q_A = 11$, $e_A = 3$, $d_A = 27$.
+Bob hat ebenfalls ein RSA-Schlüsselpaar erzeugt und dabei folgende Parameter verwendet: $p_B = 17$, $q_B = 5$, $e_B = 3$, $d_B = 43$.
+Die öffentlichen Schlüssel haben die beiden bereits über einen sicheren Kanal ausgetauscht.
+  \begin{enumerate}
+    \item Bob hat gewürfelt und möchte das Ergebnis nun an Alice senden.
+Er überträgt dazu $c_B = 9$.
+Zeigen Sie, dass eine passive Angreiferin (Eve), die lediglich die öffentlichen Schlüssel und $c_B$ kennt, mittels eines Chosen-Plaintext-Angriffs Bobs Wurfergebnis $m_B$ ermitteln kann.
+Welche Zahl hat Bob gewürfelt?
+    \item Wie kann sich Bob vor diesem Angriff schützen?
+  \end{enumerate}
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Forward Secrecy mit RSA}
+  Zur Erreichung von Forward Secrecy wird häufig die Verwendung von symmetrischen Schlüsseln empfohlen, die regelmäßig gewechselt werden.
+Ein solcher Schlüsselaustausch kann beispielsweise mithilfe des Diffie-Hellman-Schlüsselaustauschs realisiert werden.
+Ein anderer Ansatz ist die Verwendung von asymmetrischen Schlüsseln, bei denen der private Schlüssel regelmäßig gewechselt wird.
+In dieser Aufgabe betrachten wir die Verwendung von RSA-Schlüsseln.
+  \begin{enumerate}
+    \item Erläutern Sie, wie Forward Secrecy mit RSA erreicht werden kann.
+    \item Warum ist es problematisch, wenn ein Angreifer den privaten Entschlüsselungsexponenten eines RSA-Schlüssels erlangt?
+    \item Wie kann ein Angreifer, der den privaten Entschlüsselungsexponenten eines RSA-Schlüssels erlangt hat, die Kommunikation zwischen zwei Parteien entschlüsseln, die Nachrichten mit diesem Schlüssel verschlüsselt haben?
+    \item Wie kann Forward Secrecy mit RSA erreicht werden, wenn die Schlüssel regelmäßig gewechselt werden?
+  \end{enumerate}
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\setcounter{subsection}{1615}
+\begin{exercise}{Elliptische Kurven: EC-ElGamal}
+  Das ElGamal-Kryptosystem lässt sich auch auf elliptische Kurven übertragen.
+Dabei werden die Rechenoperationen mittels Punkten auf einer elliptischen Kurve vorgenommen.
+Betrachten Sie für die folgenden Teilaufgaben die elliptische Kurve $E(\text{GF}(p))$ mit der Form $y^2 \equiv x^3 + x + 6 \mod 11$.
+  \begin{enumerate}
+    \item Passen Sie die Operationen des ElGamal-Kryptosystems zur Schlüsselgenerierung, zur Entschlüsselung sowie zur Verschlüsselung an, sodass diese auf der Basis elliptischer Kurven operieren.
+    \item Generieren Sie mithilfe Ihres Verfahrens aus Teilaufgabe a) ein geeignetes Schlüsselpaar auf der oben angegebenen elliptischen Kurve.
+Berechnen Sie dann die Verschlüsselung der Nachricht $m = (5, 2)$.
+    \item Berechnen Sie die Entschlüsselung des Schlüsseltextes aus der vorherigen Teilaufgabe mit dem dort generierten geheimen Schlüssel.
+    \item Zeigen Sie anschaulich, auf welcher Annahme die Sicherheit des Verfahrens beruht.
+Welcher Ausdruck müsste berechnet werden, damit Angreifende Schlüsseltexte ohne Kenntnis des geheimen Schlüssels leicht entschlüsseln könnten?
+  \end{enumerate}
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\setcounter{subsection}{228}
+\begin{exercise}{Verbindungsverschlüsselung in GSM}
+  Welche Kommunikationsabschnitte sind in GSM verschlüsselt? Welche Kommunikationsabschnitte sind unverschlüsselt?
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\setcounter{subsection}{2226}
+\begin{exercise}{Bluetooth-Authentifikation}
+  Bei Bluetooth wird auf Basis der Funktion E22 die Authentifikation durchgeführt. Was sind die In- und Outputs der Authentifikation?
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\setcounter{subsection}{2228}
+\begin{exercise}{Mehrfachnutzung des IV in einem WEP-verschlüsselten WLAN}
+  \begin{enumerate}
+    \item Der in WEP verwendete Initialisierungsvektor (IV) ist 24 Bit groß und steht in einem Paket $P$ am Paketanfang vor dem Schlüsseltext.
+Wie viele Pakete können maximal erzeugt werden, wenn kein IV mehrfach verwendet werden soll?
+Wie sind die IVs von einem Teilnehmer des WLANs zu erzeugen, sodass eine Kollision möglichst spät auftritt?
+Welches Problem gibt es bei diesem Verfahren, wenn mehrere Teilnehmer kommunizieren?
+    \item Nehmen Sie an, Sie haben die Pakete $P_1$ und $P_2$ empfangen:
+    \begin{itemize}
+      \item $P_1$: 00010 00111 11111 11110 11001 01001 00101 ...
+      \item $P_2$: 00010 00111 11111 11110 11000 10011 00010 ...
+    \end{itemize}
+    Die ersten 10 Klartextbits des ersten Paketes $P_1$ lauten 10010 01100. Bestimmen Sie die ersten 10 Klartextbits des zweiten Paketes $P_2$.
+  \end{enumerate}
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\setcounter{section}{2023}
+\section{Mock exam}
 

### 2024-10-23 PbD/exercises.tex
Added exercise scaffold to sbd and pbd.
Added pbd catalogue exercises.

@@
 
+\setcounter{subsection}{34}
+\begin{exercise}{Informationsgewinn aufgrund von Kontextinformationen}
+  Unmittelbar vor der Durchführung eines militärischen Kampfeinsatzes steigt die Anzahl der Essenslieferungen (Pizza, Burger, Croques, Sushi etc.), die an die Adresse des Verteidigungsministeriums geliefert werden, massiv an.
+Diskutieren Sie, ob das System durch einen passiven oder einen aktiven Angreifer verwundbar ist und welche Schutzziele dadurch bedroht werden.
+Für welche Angriffsart bzw.
+Angriffsarten ist das System anfällig?
+Was wären mögliche Gegenmaßnahmen?
+
+  \begin{solution}
     % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Angriff über WLAN-Namen}
+  Viele öffentliche WLAN-Access-Points (APs) teilen allen Stationen (Clients) in Reichweite ihren WLAN-Namen (SSID) mit.
+Der Nutzer einer Station wählt aus der Liste aller Stationen anhand der SSID das Netz, mit dem er sich verbinden will.
+Damit der Nutzer diese Auswahl nicht jedes Mal wieder treffen muss, verbinden sich die Stationen automatisch mit Access Points, mit denen sie bereits verbunden waren.
+Hierzu wird eine Liste der bereits bekannten SSIDs auf der Station aufbewahrt.
+Stehen an einem Ort mehrere APs mit derselben SSID zur Auswahl, verbindet sich eine Station i.
+d.
+R.
+mit dem AP mit der größten Signalstärke.
+  \begin{enumerate}
+    \item Diskutieren Sie, ob das System durch einen passiven oder einen aktiven Angreifer verwundbar ist.
+    \item Welche Schutzziele werden dadurch bedroht?
+    \item Was wären mögliche Gegenmaßnahmen?
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+      \item The system is vulnerable to a passive attacker.
+      \item The protection goals that are threatened are confidentiality and integrity.
+      \item Possible countermeasures are to use a VPN connection and to disable automatic connection to unknown networks.
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+\setcounter{subsection}{167}
+\begin{exercise}{Verschlüsselte PIN-Übermittlung}
+  Der Nutzer eines Laptops soll sich gegenüber einem Server im Heimnetz durch eine PIN authentisieren.
+Um die vom Nutzer am Laptop eingegebene PIN nicht mitlesen zu können, wird sie vor der Übermittlung mit dem öffentlichen RSA-Schlüssel des Servers deterministisch verschlüsselt.
+  \begin{enumerate}
+    \item Analysieren Sie die (Un-)Sicherheit eines solchen Protokolls gegenüber einem Angreifer auf der Kommunikationsstrecke.
+    \item Welche Schutzmaßnahmen können Nutzer und Server umsetzen, um die Sicherheit des Protokolls zu erhöhen?
+  \end{enumerate}
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\setcounter{subsection}{188}
+\begin{exercise}{Datenschutzgerechte Pseudonyme}
+  Es ist denkbar, dass eine Zertifizierungsstelle Z den Public Key als Pseudonym einer ihr bekannten Person P zertifiziert.
+Allerdings kann die Zertifizierungsstelle dann nicht nur im Streitfall das Pseudonym aufdecken, sondern auch vollständig nachvollziehen, welche Geschäfte P gemacht hat (bzw.
+welche Signaturen P geleistet hat), da sie den Public Key mit der Identität von P verketten kann.
+
+  Denken Sie sich eine Variante aus, bei der eine einzige Zertifizierungsstelle hierzu nicht mehr in der Lage ist, im Streitfall aber trotzdem eine Aufdeckung des Pseudonyms möglich ist.
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\setcounter{subsection}{2220}
+\begin{exercise}{Speicherung des Aufenthaltsortes}
+  \begin{enumerate}
+    \item Welchen Zweck erfüllt die Speicherung des Aufenthaltsortes im GSM? Warum werden beim WLAN keine Aufenthaltsorte gespeichert?
+    \item Warum werden im GSM zur Speicherung des Aufenthaltsortes zwei getrennte Datenbanken (HLR, VLR) verwendet?
+    \item Bitte diskutieren Sie den erreichbaren Schutz des Aufenthaltsortes gegenüber Insider-Angriffen (z.B. durch Mitarbeiter des Netzbetreibers) und Outsidern.
+    \item Inwieweit trägt die TMSI zum Schutz des Aufenthaltsortes bei?
+  \end{enumerate}
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\setcounter{subsection}{2223}
+\begin{exercise}{Kritik an GSM}
+  Trotz der vorhandenen Sicherheitsfunktionen bietet GSM nur beschränkte Sicherheit hinsichtlich Vertraulichkeit. Warum?
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\setcounter{subsection}{2229}
+\begin{exercise}{Broadcast}
+  Durch Verteilung (Broadcast) von Verbindungswünschen im gesamten Versorgungsgebiet eines Kommunikationsnetzes (z.B.
+Mobilfunknetzes) kann auf das Speichern von Aufenthaltsdaten sowie Routinginformation verzichtet werden.
+  \begin{enumerate}
+    \item Erläutern Sie den Unterschied von offenen impliziten und verdeckten impliziten Adressen.
+    \item Könnte man auch die explizite Adressierung verwenden, um einem mobilen Teilnehmer eine nicht notwendigerweise vertrauliche Nachricht (z.B.
+eine SMS) zu senden, ohne ihn lokalisieren zu können?
+Begründen Sie Ihre Antwort.
+    \item Ist bei der variablen impliziten Adressierung ein Satellit zur Verteilung einsetzbar? Begründung!
+  \end{enumerate}
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Kollisionsabstand offener impliziter Adressen}
+  Die Länge einer impliziten Adresse sei 32 Bit.
+Es sollen 100.000 Teilnehmer versorgt werden.
+Jedem Teilnehmer sind zu jedem Zeitpunkt 50 Adressen zugeordnet.
+Angenommen, jeder Teilnehmer erhält im Durchschnitt 1 Nachricht pro Stunde.
+  \begin{enumerate}
+    \item Wie groß ist der Kollisionsabstand, d.h.
+wie lange dauert es im Mittel, bis ein ganz bestimmter Teilnehmer einen Fehlalarm (Nachricht eigentlich für einen anderen Teilnehmer bestimmt) erhält?
+    \item Der mittlere Kollisionsabstand soll mindestens 3 Jahre betragen. Wie lang muss die offene implizite Adresse mindestens sein?
+  \end{enumerate}
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\setcounter{subsection}{231}
+\begin{exercise}{Blind-Messsage-Service}
+  \begin{enumerate}
+    \item Welche Schutzziele erfüllt der Blind-Message-Service?
+    \item Warum müssen beim Blind-Message-Service die Anfragevektoren und die Server-Antworten verschlüsselt übertragen werden?
+    \item Warum genügt beim modifizierten Blind-Message-Service die Verschlüsselung der Anfragevektoren?
+    \item Welche Verfahren können Sie alternativ zum Blind-Message-Service einsetzen, wenn Bandbreite keine Rolle spielt?
+  \end{enumerate}
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\setcounter{subsection}{235}
+\begin{exercise}{Modifikation von Chaum'schen Mixen}
+  Chaum'sche Mixe nutzen eine Datenbank, in der alle eingehenden Nachrichten gespeichert werden.
+  \begin{enumerate}
+    \item Bitte erläutern Sie den Angriff, der durch die Datenbank verhindert wird.
+    \item Unter welcher Bedingung kann die Datenbank gelöscht werden?
+    \item Wie lässt sich das Verfahren umbauen, sodass Sie unter Beibehaltung des Angreifermodells die Speichereffizienz steigern können?
+  \end{enumerate}
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{DC-Netz}
+  \begin{enumerate}
+    \item Der (entartete) Schlüsselgraph eines DC-Netzes für die Teilnehmer A, B, C und D sei (A-B-C-D).
+B greift nun A, C und D an.
+Beschreiben Sie den erreichbaren Schutz für A, C und D.
+Wie sind Sie bei Ihrer Analyse vorgegangen?
+  \end{enumerate}
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\setcounter{subsection}{238}
+\begin{exercise}{Überwachung im Internet}
+  \begin{enumerate}
+    \item Erläutern Sie eine mögliche Vorgehensweise, mit der ein starker Angreifer (z.B.
+ein Nachrichtendienst) die IP-Adressen der Nutzer herausfinden kann, die eine bestimmte Webseite über ein mixbasiertes Anonymisierungssystem (z.B.
+Tor) abrufen.
+    \item Erläutern Sie eine mögliche Vorgehensweise, mit der ein starker Angreifer (z.B.
+ein Nachrichtendienst) ermitteln kann, welche Webseiten ein bestimmter Nutzer über ein mixbasiertes Anonymisierungssystem abgerufen hat.
+Hinweis: Beschreiben Sie eine andere Vorgehensweise als in Teilaufgabe a), auch wenn Ihre für Teilaufgabe a) gewählte Vorgehensweise für Teilaufgabe b) geeignet ist.
+    \item Erläutern Sie drei Techniken, mit denen ein Webserver einen Nutzer bei seinem zweiten (späteren) Besuch wiedererkennen kann.
+  \end{enumerate}
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\setcounter{subsection}{2310}
+\begin{exercise}{Tagging-Angriff auf anonyme Mixe und Kanäle}
+  \begin{enumerate}
+    \item Was ist ein Tagging-Angriff auf anonyme Kanäle?
+    \item Wie verhindert man den Tagging-Angriff auf anonyme Kanäle?
+  \end{enumerate}
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Tracking auf der Anwendungsschicht}
+  \begin{enumerate}
+    \item Nennen Sie zwei Schutzziele, die beim Tracking auf der Anwendungsebene gebrochen werden.
+    \item Was unterscheidet Tracking auf der Anwendungsebene von Tracking auf der Netzebene?
+    \item Nennen Sie drei grundsätzliche Mechanismen, mit denen man Tracking auf der Anwendungsebene verhindern bzw.
+erschweren kann.
+Erläutern Sie für jeden Mechanismus, wie das Tracking erschwert wird und vor welchem Angreifer, und nennen Sie jeweils ein konkretes Beispiel, bei dem der Mechanismus umgesetzt wird.
+  \end{enumerate}
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Strafverfolgung}
+  Die Polizei hat sich Zugriff auf einen Webserver verschafft, der zur unbefugten Verbreitung von urheberrechtlich geschütztem Material verwendet wird.
+Anhand des dort aufgezeichneten Access-Log sollen nun die Nutzer des Servers ausfindig gemacht werden.
+Bei der Ermittlung von Tatverdächtigen stößt man schnell auf Probleme: Die Zugriffe, die von der IP-Adresse A ausgingen, stammen von einem Kleinbetrieb, in dem es fünf Nutzer gibt.
+Der Inhaber des Anschlusses stellt der Polizei ein Protokoll des DSL-Routers zur Verfügung, in dem die Aktivitäten der einzelnen internen Nutzer anhand ihrer IP-Adressen aufgezeichnet wurden.
+Messungen ergeben, dass die Latenz des Kommunikationsnetzes zwischen DSL-Router und Server maximal 0,5 Sekunden beträgt.
+Auf Basis der verfügbaren Daten (Access-Log, IP-Adresse A, Protokoll des DSL-Routers und Latenz) soll nun die interne IP-Adresse ermittelt werden, von der die Zugriffe ausgingen.
+
+  Bitte erläutern Sie Ihr Vorgehen im Detail. Unter welcher Begrifflichkeit lassen sich derartige Angriffe zusammenfassen?
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
 

### 2024-10-23 SbD/summary.tex
Updated devcontainer and added first chapter summary to SbD.

\documentclass{article}

\usepackage[]{../.template/summary}

\subject{Security by Design}
\semester{Winter 2024}
\author{Leopold Lemmermann}

\usepackage{pdfpages}

\begin{document}\createtitle

\includepdf[pages=1-10, nup=2x5]{slides.pdf}

\section{Kryptographie}

\subsection{Angriffsarten}
\begin{enumerate}
  \item \textbf{Ciphertext Only}: Angreifer kennt nur den Ciphertext
  \item \textbf{Known Plaintext}: Angreifer kennt Ciphertext und Plaintext
  \item \textbf{Chosen Plaintext} (adaptiv v. nicht-adaptiv): Angreifer kann Plaintext wählen und erhält den Ciphertext (nicht-adaptiv: alle Plaintexte müssen vorher gewählt werden)
\end{enumerate}

\subsection{Brechen (stark zu schwach)}
\begin{enumerate}
  \item \textbf{Vollständiges Brechen}: Schlüssel gefunden
  \item \textbf{Universelles Brechen}: Äquivalenter Schlüssel gefunden
  \item \textbf{Selektives Brechen}: Bestimmte Nachricht entschlüsselt
  \item \textbf{Existenzielles Brechen}: Irgendeine Nachricht entschlüsselt
\end{enumerate}

\subsection{Klassische Chiffren}
\begin{itemize}
  \item \textbf{Transposition}: Veränderung der Reihenfolge
  \item \textbf{Subsitution}: Ersetzen von Zeichen
  \item \textbf{Produkt}: Kombination von Transposition und Substitution
\end{itemize}

\begin{enumerate}
  \item \textbf{Skytala}: Transposition mit Zylinderdurchmesser als Schlüssel
  \item \textbf{Polybios}: Substitution mit 5x5-Quadrat
  \item \textbf{Caesar}: Substitution mit Verschiebung
  \item \textbf{Vigenére}: Substitution mit Schlüsselwort
  \item \textbf{Beaufort}: involutorischer Vigenére (gleiche Verschlüsselung und Entschlüsselung)
  \item \textbf{Vernam} (one-time pad): Substitution mit zufälligem Schlüssel mit Klartextlänge
\end{enumerate}

\subsection{Moderne Kryptographie}

\end{document}

### 2024-10-24 BAI/paper/paper.tex
Added structure for BAI paper and presentation.

--- 2024-10-18 BAI/templates/short_paper_generic/shortpaper.tex
@@
 \documentclass[12pt]{article}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Meta informations:
-\newcommand{\trauthor}{Horst Hansen}
-\newcommand{\trtype}{Proposal} %{Expos\'{e}} %{Review}
-\newcommand{\trtitle}{Neural Networks for Artificial Agents}
-\newcommand{\trdate}{01.09.2010}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Languages:
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Metainformation
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\newcommand{\trauthor}{Leopold Lemmermann}
+\newcommand{\trtype}{Paper}
+\newcommand{\trtitle}{Image Captioning}
+\newcommand{\trdate}{\today}
 
-% Falls die Ausarbeitung in Deutsch erfolgt:
-% \usepackage[german]{babel}
-% \usepackage[T1]{fontenc}
-% \usepackage[latin1]{inputenc}
-% \usepackage[latin9]{inputenc}	 				
-% \selectlanguage{german}
-
-% If the thesis is written in English:
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Language
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \usepackage[english]{babel}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Bind packages:
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Bind packages
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \usepackage{acronym}                    % Acronyms
@@
 \usepackage{booktabs}                   % Nicer tables
-%\usepackage[font=small,labelfont=bf]{caption} % Numbered captions for figures
+% \usepackage[font=small,labelfont=bf]{caption} % Numbered captions for figures
 \usepackage{color}                      % Enables defining of colors via \definecolor
@@
 \usepackage{fancyhdr}										% Packet for nicer headers
-%\usepackage{fancyheadings}             % Nicer numbering of headlines
+\usepackage{fancyheadings}             % Nicer numbering of headlines
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Configurationen:
-
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Configuration
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \hyphenation{whe-ther} 									% Manually use: "\-" in a word: Staats\-ver\-trag
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Additional 'theorem' and 'definition' blocks:
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Additional 'theorem' and 'definition' blocks
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \theoremstyle{plain}
 \newtheorem{theorem}{Theorem}[section]
-%\newtheorem{theorem}{Satz}[section]		% Wenn in Deutsch geschrieben wird.
 \newtheorem{axiom}{Axiom}[section]
-%\newtheorem{axiom}{Fakt}[chapter]			% Wenn in Deutsch geschrieben wird.
 %Usage:%\begin{axiom}[optional description]%Main part%\end{fakt}
@@
 \theoremstyle{remark}
-%\newtheorem{remark}[definition]{Bemerkung} % Wenn in Deutsch geschrieben wird.
 \newtheorem{remark}[definition]{Remark}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Provides TODOs within the margin:
-\newcommand{\TODO}[1]{\marginpar{\emph{\small{{\bf TODO: } #1}}}}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 % Abbreviations and mathematical symbols
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \newcommand{\modd}{\text{ mod }}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 % Document:
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{document}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Cover Header:
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Cover
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \title{\trtitle\\[0.3cm]{\normalsize\trtype}}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\thispagestyle{empty}
 
-\thispagestyle{empty}
+\begin{abstract}
+  This paper is about image captioning. We will discuss the current state of the art and give an overview of the most important methods. We will also present some of our own results.
+\end{abstract}
+
+\tableofcontents
+\newpage
 \pagenumbering{arabic}
 
-% Abstract gives a brief summary of the main points of a paper:
-\begin{abstract}
-  This paper gives a brief overview and an example of how to write a short paper in \LaTeX. An example for a short paper could be a proposal or a review.
-\end{abstract}
-
-% the actual content, usually separated over a number of sections
-% each section is assigned a label, in order to be able to put a
-% crossreference to it
-
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Content (Outline_Lemmermann_Leopold_TITLE)
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Introduction}
@@
 
-The introduction describes the problem, why it is important, the main
-ideas of the following paper, what are the main contributions of the
-paper, etc.
+Your text here...
 
-If the paper has more then 4 pages, a readers guide is recommenced. In a short paragraph an outline of the content of the next
+\section{Approach}
+\label{sec:approach}
 
-\section{Optional: Background Information}
-\label{sec:basics}
-A brief section giving background information may be necessary, especially if your work spans two or more traditional fields.
-That means that your readers may not have any experience with some of the material needed to follow your thesis, so you need to give it to them.
-A different title than that given above is usually better; e.g., "A Brief Review of Frammis Algebra."
+Your text here...
 
-\section{Related Work}
-\label{sec:relwork}
+\section{Results}
+\label{sec:results}
 
-The related work section could describe other work that is in some respects relevant for the understanding of the problem outlined in Section~\ref{sec:introduction}, that offer competing solutions, etc.
-
-All sources must be properly referenced, ideally by using the BiBTeX system. References can then be very conveniently made with the \textsc{cite} command. For example, reference
-\cite{Leunen:Scholars:92} discusses some of the elementary rules on
-writing scientific papers, amongst others how to correctly cite other
-documents. Reference \cite{Taylor:SIGuide:95}, e.g., describes how to
-correctly use the SI system of units and their correct typographical
-representation. In general you organize this section by idea, and not by author or by publication.
-
-\section{Model description}
-\label{sec:model}
-
-After the two (or maybe three) common sections Introduction (plus optional Background) and Related work, more sections with the actual content of a paper follow. The style and
-structure of such sections varies by a large degree, no general rules
-of thumb can be given.
-
-\subsection{Word processing with \LaTeX}
-\label{sec:model:subsec:latex}
-
-This document has already introduced the most important constructs of
-\LaTeX. What is necessary to produce documents with \LaTeX is simple
-any normal text editor and a \LaTeX distribution. This is commonly
-installed on practically all UNIX-type systems; for Windows, an
-excellent \LaTeX exists, called MikTeX, available from
-\url{www.miktex.org}. Almost all distributions come with a large patch
-of examples and introductory material; consult your local installation
-for details.
-
-Lots of supplementary and background information, FAQs, etc.\ is
-available from the Comprehensive TeX Archive Network (CTAN); the
-German mirror of which is \url{www.dante.de}.
-
-\subsection{Tables in \LaTeX}
-\label{sec:model:subsec:tables}
-Tables should be centred and should always have a caption 
-positioned above it. A caption in a sentence form as well as in a short form must end with a period as seen in table~\ref{tab:sample}.
-
-\begin{table}[hbtp]
-  \caption{This caption has one line so it is centred.}\label{tab:sample} 
-  \centering
-  \begin{tabular}{|c|c|}
-    \hline
-    Example column 1 & Example column 2 \\
-    \hline
-    Example text 1 & Example text 2 \\
-    \hline
-  \end{tabular}
-\end{table}
-
-\subsection{Figures in \LaTeX}
-\label{sec:model:subsec:figures}
-Note that a figure is a so-called floating object: it is moved around the actual
-text in order to best fit on a page. This is in stark contrast to some
-GUI-based word processing tools, where the placement of figures is
-usually more associated with luck than principle.
-
-As figures float around, expressions like ``the following figure''
-must never be used. Instead, figures need a caption, a label, and must
-be properly referenced in the main text. A figure caption is placed centred below the figure and describes the figure in (very) short.
-
-In general, only vector graphics in encapsulated postscript (eps) or a similar format should be included in any kind of text, as this allows arbitrary scaling, rotation etc.\ without any loss of quality.
-Bitmap formats (JPEG, GIF, \dots) should only be used if no other alternative exists
---- basically the only case where bitmaps can be justified is when scanned pictures need to be included in a text, however, this should be avoided as hard as possible as the quality in usually not satisfactory.
-If a screen shot is needed a high resolution picture without visible fragments of a jpeg compression is allowed.
-Figures like the figure~\ref{fig:samplefig} should always appear after the first referencing it.
-\begin{figure}[hbtp]
-	 \centerline{\includegraphics[width=0.6\textwidth]{samplefig.pdf}}
-	 {\caption{Network of transputers and the structure of individual
-processes.}\label{fig:samplefig}}
-\end{figure}
-
-\section{Model analysis}
-\label{sec:analysis}
-The analysis of a model is a much more free-form of a paper.
-Starting with a comparison of several approaches over some experimental settings and result up to a theoretical verification of the model, anything is allowed.
-But it all has only one purpose: to convince the reader of the right to exist of the described model.
+Your text here...
 
@@
 
-At the end, there is a final section concluding and summarizing a
-paper, putting the entire work into perspective and explaining, on a
-larger level, what the consequence of this work are. Also, unexpected
-results can be discussed here, etc.
+Your text here...
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% hier werden - zum Ende des Textes - die bibliographischen Referenzen
-% eingebunden
-%
-% Insbesondere stehen die eigentlichen Informationen in der Datei
-% ``bib.bib''
-%
+% References
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\newpage
+\thispagestyle{empty}
+
+\nocite{*}
+
 \bibliography{bib}

### 2024-10-24 BAI/paper/src/main.py

@@ -1,0 +1,369 @@
+data_location =  "./flickr8k"
+
+#imports
+import numpy as np
+import torch
+from torch.utils.data import DataLoader,Dataset
+import torchvision.transforms as T
+
+#custom imports 
+from data_loader import FlickrDataset,get_data_loader
+
+
+import matplotlib.pyplot as plt
+def show_image(img, title=None):
+    """Imshow for Tensor."""
+    
+    #unnormalize 
+    img[0] = img[0] * 0.229
+    img[1] = img[1] * 0.224 
+    img[2] = img[2] * 0.225 
+    img[0] += 0.485 
+    img[1] += 0.456 
+    img[2] += 0.406
+    
+    img = img.numpy().transpose((1, 2, 0))
+    
+    
+    plt.imshow(img)
+    if title is not None:
+        plt.title(title)
+    plt.pause(0.001)  # pause a bit so that plots are updated
+    
+    
+    
+#setting the constants
+#BATCH_SIZE = 256
+BATCH_SIZE = 6
+NUM_WORKER = 4
+
+#defining the transform to be applied
+transforms = T.Compose([
+    T.Resize(226),                     
+    T.RandomCrop(224),                 
+    T.ToTensor(),                               
+    T.Normalize((0.485, 0.456, 0.406),(0.229, 0.224, 0.225))
+])
+
+
+#testing the dataset class
+dataset =  FlickrDataset(
+    root_dir = data_location+"/Images",
+    captions_file = data_location+"/captions.txt",
+    transform=transforms
+)
+
+#writing the dataloader
+data_loader = get_data_loader(
+    dataset=dataset,
+    batch_size=BATCH_SIZE,
+    num_workers=NUM_WORKER,
+    shuffle=True,
+    # batch_first=False
+)
+
+#vocab_size
+vocab_size = len(dataset.vocab)
+
+device = torch.device("cuda:0" if torch.cuda.is_available() else "cpu")
+device
+
+
+
+
+import torch.nn as nn
+import torch.nn.functional as F
+import torch.optim as optim
+
+import torchvision.models as models
+from torch.utils.data import DataLoader,Dataset
+import torchvision.transforms as T
+
+class EncoderCNN(nn.Module):
+    def __init__(self):
+        super(EncoderCNN, self).__init__()
+        resnet = models.resnet50(pretrained=True)
+        for param in resnet.parameters():
+            param.requires_grad_(False)
+        
+        modules = list(resnet.children())[:-2]
[…]

### 2024-10-26 BAI/paper/src/data_loader.py
Updated bai paper src.

@@ -3,7 +3,7 @@ from collections import Counter
 import spacy
 import torch
 from torch.nn.utils.rnn import pad_sequence
-from torch.utils.data import DataLoader,Dataset
+from torch.utils.data import DataLoader, Dataset
 import torchvision.transforms as T
 
 from PIL import Image
@@ -18,84 +18,78 @@ import matplotlib.image as mpimg
 
 
 class Vocabulary:
-    def __init__(self,freq_threshold, spacy_model_str="en_core_web_sm"):
+    def __init__(self, freq_threshold, spacy_model_str="en_core_web_sm"):
         self.spacy_eng = spacy.load(spacy_model_str)
-        
-        #setting the pre-reserved tokens int to string tokens
-        self.itos = {0:"<PAD>",1:"<SOS>",2:"<EOS>",3:"<UNK>"}
-        
-        #string to int tokens
-        #its reverse dict self.itos
-        self.stoi = {v:k for k,v in self.itos.items()}
-        
+
+        # setting the pre-reserved tokens int to string tokens
+        self.itos = {0: "<PAD>", 1: "<SOS>", 2: "<EOS>", 3: "<UNK>"}
+
+        # string to int tokens
+        # its reverse dict self.itos
+        self.stoi = {v: k for k, v in self.itos.items()}
+
         self.freq_threshold = freq_threshold
-        
+
     def __len__(self): return len(self.itos)
-    
-    #@staticmethod
-    def tokenize(self,text):
+
+    # @staticmethod
+    def tokenize(self, text):
         return [token.text.lower() for token in self.spacy_eng.tokenizer(text)]
-    
+
     def build_vocab(self, sentence_list):
         frequencies = Counter()
         idx = 4
-        
+
         for sentence in sentence_list:
             for word in self.tokenize(sentence):
                 frequencies[word] += 1
-                
-                #add the word to the vocab if it reaches minum frequecy threshold
+
+                # add the word to the vocab if it reaches minum frequecy threshold
                 if frequencies[word] == self.freq_threshold:
                     self.stoi[word] = idx
                     self.itos[idx] = word
                     idx += 1
-    
-    def numericalize(self,text):
+
+    def numericalize(self, text):
         """ For each word in the text corresponding index token for that word form the vocab built as list """
         tokenized_text = self.tokenize(text)
[…]

### 2024-10-26 BAI/paper/src/main.py

@@ -6,32 +6,32 @@ import torch
 from torch.utils.data import DataLoader,Dataset
 import torchvision.transforms as T
 
-#custom imports 
+#custom imports
 from data_loader import FlickrDataset,get_data_loader
 
 
 import matplotlib.pyplot as plt
 def show_image(img, title=None):
     """Imshow for Tensor."""
-    
-    #unnormalize 
+
+    #unnormalize
     img[0] = img[0] * 0.229
-    img[1] = img[1] * 0.224 
-    img[2] = img[2] * 0.225 
-    img[0] += 0.485 
-    img[1] += 0.456 
+    img[1] = img[1] * 0.224
+    img[2] = img[2] * 0.225
+    img[0] += 0.485
+    img[1] += 0.456
     img[2] += 0.406
-    
+
     img = img.numpy().transpose((1, 2, 0))
-    
-    
+
+
     plt.imshow(img)
     if title is not None:
         plt.title(title)
     plt.pause(0.001)  # pause a bit so that plots are updated
-    
-    
-    
+
+
+
 #setting the constants
 #BATCH_SIZE = 256
 BATCH_SIZE = 6
@@ -39,9 +39,9 @@ NUM_WORKER = 4
 
 #defining the transform to be applied
 transforms = T.Compose([
-    T.Resize(226),                     
-    T.RandomCrop(224),                 
-    T.ToTensor(),                               
+    T.Resize(226),
+    T.RandomCrop(224),
+    T.ToTensor(),
     T.Normalize((0.485, 0.456, 0.406),(0.229, 0.224, 0.225))
 ])
 
@@ -85,157 +85,157 @@ class EncoderCNN(nn.Module):
         resnet = models.resnet50(pretrained=True)
         for param in resnet.parameters():
             param.requires_grad_(False)
-        
+
         modules = list(resnet.children())[:-2]
         self.resnet = nn.Sequential(*modules)
-        
+
 
     def forward(self, images):
         features = self.resnet(images)                                    #(batch_size,2048,7,7)
         features = features.permute(0, 2, 3, 1)                           #(batch_size,7,7,2048)
         features = features.view(features.size(0), -1, features.size(-1)) #(batch_size,49,2048)
         return features
-    
-    
+
+
 #Bahdanau Attention
 class Attention(nn.Module):
     def __init__(self, encoder_dim,decoder_dim,attention_dim):
         super(Attention, self).__init__()
-        
+
[…]

### 2024-10-26 BAI/paper/src/__main__.py
Refactored the image captioning implementation. Just have to figure out memory problems.

@@ -1,0 +1,26 @@
+from torch import cuda
+from torchvision import transforms as T
+
+from models import ImageCaption
+
+model = ImageCaption(
+    "data/micro_flickr",
+    batch_size=6,
+    num_workers=4,
+    embed_size=300,
+    attention_dim=256,
+    encoder_dim=2048,
+    decoder_dim=512,
+    learning_rate=3e-4,
+    transform=T.Compose([
+        T.Resize(226),
+        T.RandomCrop(224),
+        T.ToTensor(),
+        T.Normalize((0.485, 0.456, 0.406), (0.229, 0.224, 0.225))
+    ]),
+    device="cuda" if cuda.is_available() else "cpu"
+)
+
+model.train(epochs=3)
+
+print(model.predict(model.dataset[0][0]))

### 2024-10-26 BAI/paper/src/data/__init__.py

@@ -1,0 +1,2 @@
+from .flickrdataset import FlickrDataset
+from .vocabulary import Vocabulary

### 2024-10-26 BAI/paper/src/data/flickrdataset.py

@@ -1,0 +1,70 @@
+from data.vocabulary import Vocabulary
+import os
+from torch.nn.utils.rnn import pad_sequence
+from torch.utils.data import DataLoader, Dataset
+from PIL import Image
+from pandas import read_csv
+from torch import tensor, Tensor, cat
+from torchvision.transforms import Compose
+
+
+class FlickrDataset(Dataset):
+    """ Flickr dataset for image captioning. """
+
+    def __init__(self, path: str, captions: str = "captions.txt", transform: Compose = None, threshold: int = 5):
+        self.path = path
+        self.images_with_captions = read_csv(os.path.join(self.path, captions))
+        self.transform = transform
+        self.vocab = Vocabulary(threshold, self.captions)
+
+    def __len__(self) -> int: return len(self.captions)
+
+    def __getitem__(self, index: int) -> tuple[Tensor, Tensor]:
+        raw_caption = self.captions[index]
+        numericalized_caption = [self.vocab.token_to_id["<SOS>"]] + \
+            self.vocab.numericalize(raw_caption) + \
+            [self.vocab.token_to_id["<EOS>"]]
+        caption = tensor(numericalized_caption)
+
+        image_path = os.path.join(self.path, "Images", self.images[index])
+        image = Image.open(image_path).convert("RGB")
+        if self.transform:
+            image = self.transform(image)
+
+        return image, caption
+
+    @property
+    def images(self) -> list[str]:
+        return self.images_with_captions["image"].tolist()
+
+    @property
+    def captions(self) -> list[str]:
+        return self.images_with_captions["caption"].tolist()
+
+    def data_loader(self, batch_size: int = 4, num_workers: int = 1, shuffle: bool = True):
+        return DataLoader(
+            dataset=self,
+            batch_size=batch_size,
+            num_workers=num_workers,
+            shuffle=shuffle,
+            collate_fn=FlickrDataset.CapsCollate(
+                pad_index=self.vocab.token_to_id["<PAD>"], batch_first=True)
+        )
+
+    class CapsCollate:
+        """ Collate to apply the padding to the captions with dataloader. """
+
+        def __init__(self, pad_index: int, batch_first: bool = False):
+            self.pad_index = pad_index
+            self.batch_first = batch_first
+
+        def __call__(self, batch: list[tuple[Tensor, Tensor]]) -> tuple[Tensor, Tensor]:
+            """ Apply padding to the captions. """
+            images = [item[0].unsqueeze(0) for item in batch]
[…]

### 2024-10-26 BAI/paper/src/data/vocabulary.py

@@ -1,0 +1,40 @@
+from collections import Counter
+from spacy import load
+
+
+class Vocabulary:
+    """ Vocabulary object to store the mapping between words and indices. """
+
+    def __init__(self, threshold: int, sentences: list[str] = [], spacy_model: str = "en_core_web_sm"):
+        self.threshold = threshold
+        self.id_to_token = {0: "<PAD>", 1: "<SOS>", 2: "<EOS>", 3: "<UNK>"}
+        self.spacy_model = load(spacy_model)
+
+        self._learn(sentences)
+
+    def __len__(self) -> int: return len(self.id_to_token)
+
+    @property
+    def token_to_id(self) -> dict[str, int]:
+        return {v: k for k, v in self.id_to_token.items()}
+
+    def numericalize(self, text: str) -> list[int]:
+        """ For each token in the text, return the index in the vocabulary. """
+        token_to_id = self.token_to_id
+        return [token_to_id[token] if token in token_to_id else token_to_id["<UNK>"]
+                for token in self._tokenize(text)]
+
+    def _learn(self, sentences: list[str]):
+        """ Build the vocabulary from the sentences. """
+        freq = Counter()
+        for sentence in sentences:
+            for token in self._tokenize(sentence):
+                freq[token] += 1
+
+                if token not in self.id_to_token.values():
+                    self.id_to_token[len(self.id_to_token)] = token
+
+    def _tokenize(self, text: str) -> list[str]:
+        """ Tokenize the text using spacy model. """
+        return [token.text.lower()
+                for token in self.spacy_model.tokenizer(text)]

### 2024-10-26 BAI/paper/src/data_loader.py

@@ -1,138 +1,0 @@
-import os
-from collections import Counter
-import spacy
-import torch
-from torch.nn.utils.rnn import pad_sequence
-from torch.utils.data import DataLoader, Dataset
-import torchvision.transforms as T
-
-from PIL import Image
-import pandas as pd
-
-import matplotlib.pyplot as plt
-import matplotlib.image as mpimg
-
-# python -m spacy download en_core_web_sm
-# might have to reaplce python with the verison
-# https://stackoverflow.com/questions/69716018/modulenotfounderror-no-module-named-spacy-even-though-spacy-and-python-are-in
-
-
-class Vocabulary:
-    def __init__(self, freq_threshold, spacy_model_str="en_core_web_sm"):
-        self.spacy_eng = spacy.load(spacy_model_str)
-
-        # setting the pre-reserved tokens int to string tokens
-        self.itos = {0: "<PAD>", 1: "<SOS>", 2: "<EOS>", 3: "<UNK>"}
-
-        # string to int tokens
-        # its reverse dict self.itos
-        self.stoi = {v: k for k, v in self.itos.items()}
-
-        self.freq_threshold = freq_threshold
-
-    def __len__(self): return len(self.itos)
-
-    # @staticmethod
-    def tokenize(self, text):
-        return [token.text.lower() for token in self.spacy_eng.tokenizer(text)]
-
-    def build_vocab(self, sentence_list):
-        frequencies = Counter()
-        idx = 4
-
-        for sentence in sentence_list:
-            for word in self.tokenize(sentence):
-                frequencies[word] += 1
-
-                # add the word to the vocab if it reaches minum frequecy threshold
-                if frequencies[word] == self.freq_threshold:
-                    self.stoi[word] = idx
-                    self.itos[idx] = word
-                    idx += 1
-
-    def numericalize(self, text):
-        """ For each word in the text corresponding index token for that word form the vocab built as list """
-        tokenized_text = self.tokenize(text)
-        return [self.stoi[token] if token in self.stoi else self.stoi["<UNK>"] for token in tokenized_text]
-
-
-class FlickrDataset(Dataset):
-    def __init__(self, root_dir, captions_file, transform=None, freq_threshold=5):
-        self.root_dir = root_dir
-        self.df = pd.read_csv(captions_file)
-        self.transform = transform
-
-        # Get image and caption colum from the dataframe
[…]

### 2024-10-26 BAI/paper/src/download.sh

@@ -1,0 +1,10 @@
+DEST=${1:-/workspaces/w24/BAI/paper/src/data/flickr8k}
+
+# Download flickr8k dataset
+mkdir -p $DEST
+curl -L -o $DEST/flickr8k.zip https://www.kaggle.com/api/v1/datasets/download/adityajn105/flickr8k
+unzip $DEST/flickr8k.zip -d $DEST
+rm $DEST/flickr8k.zip
+
+# Download spacy model
+spacy download en_core_web_sm

### 2024-10-26 BAI/paper/src/main.py

@@ -1,369 +1,0 @@
-data_location =  "./flickr8k"
-
-#imports
-import numpy as np
-import torch
-from torch.utils.data import DataLoader,Dataset
-import torchvision.transforms as T
-
-#custom imports
-from data_loader import FlickrDataset,get_data_loader
-
-
-import matplotlib.pyplot as plt
-def show_image(img, title=None):
-    """Imshow for Tensor."""
-
-    #unnormalize
-    img[0] = img[0] * 0.229
-    img[1] = img[1] * 0.224
-    img[2] = img[2] * 0.225
-    img[0] += 0.485
-    img[1] += 0.456
-    img[2] += 0.406
-
-    img = img.numpy().transpose((1, 2, 0))
-
-
-    plt.imshow(img)
-    if title is not None:
-        plt.title(title)
-    plt.pause(0.001)  # pause a bit so that plots are updated
-
-
-
-#setting the constants
-#BATCH_SIZE = 256
-BATCH_SIZE = 6
-NUM_WORKER = 4
-
-#defining the transform to be applied
-transforms = T.Compose([
-    T.Resize(226),
-    T.RandomCrop(224),
-    T.ToTensor(),
-    T.Normalize((0.485, 0.456, 0.406),(0.229, 0.224, 0.225))
-])
-
-
-#testing the dataset class
-dataset =  FlickrDataset(
-    root_dir = data_location+"/Images",
-    captions_file = data_location+"/captions.txt",
-    transform=transforms
-)
-
-#writing the dataloader
-data_loader = get_data_loader(
-    dataset=dataset,
-    batch_size=BATCH_SIZE,
-    num_workers=NUM_WORKER,
-    shuffle=True,
-    # batch_first=False
-)
-
-#vocab_size
-vocab_size = len(dataset.vocab)
-
-device = torch.device("cuda:0" if torch.cuda.is_available() else "cpu")
-device
-
-
-
-
-import torch.nn as nn
-import torch.nn.functional as F
-import torch.optim as optim
-
-import torchvision.models as models
-from torch.utils.data import DataLoader,Dataset
-import torchvision.transforms as T
-
-class EncoderCNN(nn.Module):
-    def __init__(self):
-        super(EncoderCNN, self).__init__()
-        resnet = models.resnet50(pretrained=True)
-        for param in resnet.parameters():
-            param.requires_grad_(False)
-
-        modules = list(resnet.children())[:-2]
-        self.resnet = nn.Sequential(*modules)
-
-
[…]

### 2024-10-26 BAI/paper/src/models/__init__.py

@@ -1,0 +1,1 @@
+from .image_caption import ImageCaption

### 2024-10-26 BAI/paper/src/models/attention.py

@@ -1,0 +1,41 @@
+from torch.nn import Module, Linear
+from torch import Tensor, tanh
+from typing import Tuple
+from torch.nn.functional import softmax
+
+
+class Attention(Module):
+    """ Calculate the attention weights (Bahdanau). """
+
+    def __init__(self, encoder_dim: int, decoder_dim: int, attention_dim: int):
+        super(Attention, self).__init__()
+
+        self.attention_dim = attention_dim
+        self.DecoderToAttention = Linear(decoder_dim, attention_dim)
+        self.EncoderToAttention = Linear(encoder_dim, attention_dim)
+        self.AttentionScore = Linear(attention_dim, 1)
+
+    def forward(self, features: Tensor, hidden: Tensor) -> Tuple[Tensor, Tensor]:
+        # (batch_size,num_layers,attention_dim)
+        encoded_features: Tensor = self.EncoderToAttention(features)
+        # (batch_size,attention_dim)
+        hidden_state_weight: Tensor = self.DecoderToAttention(hidden)
+
+        # (batch_size,num_layers,attemtion_dim)
+        combined_states = tanh(
+            encoded_features + hidden_state_weight.unsqueeze(1))
+
+        # (batch_size,num_layers,1)
+        attention_scores: Tensor = self.AttentionScore(combined_states)
+        # (batch_size,num_layers)
+        attention_scores = attention_scores.squeeze(2)
+
+        # (batch_size,num_layers)
+        alpha = softmax(attention_scores, dim=1)
+
+        # (batch_size,num_layers,features_dim)
+        context = features * alpha.unsqueeze(2)
+        # (batch_size,num_layers)
+        context = context.sum(dim=1)
+
+        return alpha, context

### 2024-10-26 BAI/paper/src/models/decoder.py

@@ -1,0 +1,100 @@
+from torch import zeros, device, cuda, cat, Tensor, tensor
+from torch.nn import Module, Embedding, LSTMCell, Linear, Dropout
+from typing import Tuple
+
+from data import Vocabulary
+from .attention import Attention
+
+
+class CaptionDecoder(Module):
+    """ CaptionDecoder is a RNN-based language model that generates image captions. """
+
+    def __init__(self,
+                 embed_size: int,
+                 vocab_size: int,
+                 attention_dim: int,
+                 encoder_dim: int,
+                 decoder_dim: int,
+                 drop_prob: float = 0.3,
+                 device: device = device("cuda" if cuda.is_available() else "cpu")):
+        super(CaptionDecoder, self).__init__()
+
+        self.device = device
+        self.vocab_size = vocab_size
+
+        self.embedding_layer = Embedding(vocab_size, embed_size)
+        self.attention_layer = Attention(
+            encoder_dim, decoder_dim, attention_dim)
+        self.init_cell = Linear(encoder_dim, decoder_dim)
+        self.init_hidden = Linear(encoder_dim, decoder_dim)
+        self.lstm_layer = LSTMCell(
+            embed_size+encoder_dim, decoder_dim, bias=True)
+        self.output_layer = Linear(decoder_dim, vocab_size)
+        self.drop_layer = Dropout(drop_prob)
+
+    def forward(self, features: Tensor, captions: Tensor) -> Tuple[Tensor, Tensor]:
+        # (batch_size, decoder_dim)
+        hidden, cell = self._init_lstm_layer(features)
+
+        length = len(captions[0])-1
+        batch_size = captions.size(0)
+        num_features = features.size(1)
+        preds = zeros(batch_size, length, self.vocab_size).to(self.device)
+        alphas = zeros(batch_size, length, num_features).to(self.device)
+
+        embeddings = self.embedding_layer(captions)
+
+        for i in range(length):
+            alpha, context = self.attention_layer(features, hidden)
+            lstm_input = cat((embeddings[:, i], context), dim=1)
+            hidden, cell = self.lstm_layer(lstm_input, (hidden, cell))
+
+            output = self.output_layer(self.drop_layer(hidden))
+
+            preds[:, i] = output
+            alphas[:, i] = alpha
+
+        return preds, alphas
+
+    def generate_caption(self, features: Tensor, max_len: int = 20, vocab: Vocabulary = None) -> Tuple[str, list]:
[…]

### 2024-10-26 BAI/paper/src/models/encoder.py

@@ -1,0 +1,33 @@
+from torch import Tensor
+from torch.nn import Module, Sequential
+from torchvision.models import resnet50, ResNet
+
+
+class ImageEncoder(Module):
+    """ A CNN-based ResNet-50 model that is used to encode the images to a feature vector. """
+
+    def __init__(self):
+        """ Load the pretrained ResNet-50 and replace top fc layer. """
+        super(ImageEncoder, self).__init__()
+
+        resnet: ResNet = resnet50(weights='ResNet50_Weights.DEFAULT')
+
+        for param in resnet.parameters():
+            param.requires_grad_(False)
+
+        modules = list(resnet.children())[:-2]
+        self.model = Sequential(*modules)
+
+    def forward(self, images: Tensor) -> Tensor:
+        """ Extract feature vectors from input images. """
+        # (batch_size,2048,7,7)
+        features: Tensor = self.model(images)
+
+        # (batch_size,7,7,2048)
+        features = features.permute(0, 2, 3, 1)
+
+        # (batch_size,49,2048)
+        features = features.view(
+            features.size(0), -1, features.size(-1))
+
+        return features

### 2024-10-26 BAI/paper/src/models/image_caption.py

@@ -1,0 +1,153 @@
+from torch import device, Tensor, save, no_grad, cuda
+from torch.nn import Module, CrossEntropyLoss
+from torchvision import transforms as T
+from torch.optim import Adam
+from matplotlib import pyplot as plt
+from typing import Tuple
+
+from data import FlickrDataset
+from .encoder import ImageEncoder
+from .decoder import CaptionDecoder
+
+
+class ImageCaption(Module):
+    def __init__(self,
+                 path: str,
+                 batch_size: int,
+                 num_workers: int,
+                 embed_size: int,
+                 attention_dim: int,
+                 encoder_dim: int,
+                 decoder_dim: int,
+                 learning_rate: float,
+                 transform: T.Compose = None,
+                 device=device("cuda" if cuda.is_available() else "cpu")):
+        super(ImageCaption, self).__init__()
+
+        self.device = device
+
+        self.embed_size = embed_size
+        self.attention_dim = attention_dim
+        self.encoder_dim = encoder_dim
+        self.decoder_dim = decoder_dim
+
+        self.dataset = FlickrDataset(
+            path=path,
+            transform=transform
+        )
+        self.loader = self.dataset.data_loader(
+            batch_size=batch_size, num_workers=num_workers)
+        vocab_size = len(self.dataset.vocab)
+
+        self.encoder = ImageEncoder()
+        self.decoder = CaptionDecoder(
+            embed_size=embed_size,
+            vocab_size=vocab_size,
+            attention_dim=attention_dim,
+            encoder_dim=encoder_dim,
+            decoder_dim=decoder_dim,
+            device=device
+        ).to(device)
+
+        self.criterion = CrossEntropyLoss(
+            ignore_index=self.dataset.vocab.token_to_id["<PAD>"])
+        self.optimizer = Adam(self.parameters(), lr=learning_rate)
+
+    def forward(self, images: Tensor, captions: Tensor) -> Tensor:
+        features = self.encoder(images)
+        outputs = self.decoder(features, captions)
+        return outputs
+
+    def train(self, epochs: int = 5, print_every: int = 10):
+        for epoch in range(epochs):
+            for idx, (images, captions) in enumerate(iter(self.loader)):
+                images, captions = images.to(
+                    self.device), captions.to(self.device)
+
+                # Zero the gradients.
+                self.optimizer.zero_grad()
+
+                # Feed forward
+                outputs, _ = self(images, captions)
+
[…]

### 2024-10-26 BAI/paper/src/download.sh
updated bai paper download script.

@@ -1,7 +1,9 @@
 DEST=${1:-/workspaces/w24/BAI/paper/src/data/flickr8k}
 
-# Download flickr8k dataset
-mkdir -p $DEST
+# Download flickr8k dataset#
+if [ ! -d $DEST ]; then
+  mkdir -p $DEST
+fi
 curl -L -o $DEST/flickr8k.zip https://www.kaggle.com/api/v1/datasets/download/adityajn105/flickr8k
 unzip $DEST/flickr8k.zip -d $DEST
 rm $DEST/flickr8k.zip

### 2024-10-26 BAI/paper/src/__main__.py
Removed the tutorial code.

@@ -5,8 +5,8 @@ from models import ImageCaption
 
 model = ImageCaption(
     "data/micro_flickr",
-    batch_size=6,
-    num_workers=4,
+    batch_size=4,
+    num_workers=2,
     embed_size=300,
     attention_dim=256,
     encoder_dim=2048,

### 2024-10-26 BAI/paper/paper.tex
Added more descriptive placeholders to bai paper.
Added ten papers to bai paper bib.
Got the BAI code ready for google colab.

@@
 \usepackage{fancyhdr}										% Packet for nicer headers
-\usepackage{fancyheadings}             % Nicer numbering of headlines
 
@@
 \begin{abstract}
-  This paper is about image captioning. We will discuss the current state of the art and give an overview of the most important methods. We will also present some of our own results.
+  This paper is about image captioning. We discuss the current state of the art and give an overview of the most important methods. We also present some of our own results.
 \end{abstract}
@@
 
-Your text here...
+Image captioning is the task of generating a textual description of an image.
+It is a challenging problem that requires understanding of both the visual and the textual domain.
+In recent years, deep learning has revolutionized the field of image captioning.
+
+In this paper, we present an overview of the most important methods in image captioning. We also present some of our own results.
+
+Comparable, if more detailed, papers include %TODO: Add references to related work.
+
+\section{Methods}
+\label{sec:methods}
+
+Accuracy of image captioning is commonly measured using BLEU or CIDEr scores.
+BLEU is a precision-based metric that compares n-grams in the generated caption to those in the reference captions.
+CIDEr is a recall-based metric that compares the generated caption to the reference captions using cosine similarity.
+
+We put our main focus on the flickr8k dataset.
+It consists of 8,000 images, each with five reference captions.
+Other commonly used datasets are flickr30k and MS COCO, but due to their size and our limited resources, we decided to focus on flickr8k.
+
+Usually, to achieve image captioning an encoder-decoder architecture is used.
+The encoder is a convolutional neural network (CNN) that extracts features from the image.
+The decoder is a recurrent neural network (RNN) that generates the caption word by word.
+
+We experiment with different architectures, such as VGG16, ResNet, and InceptionV3 as encoders, and LSTM and GRU as decoders.
+We also experiment with different attention mechanisms, such as soft attention and hard attention.
 
@@
 
-Your text here...
+We follow the state-of-the-art approach to image captioning.
+We preprocess the images using the encoder and tokenize the captions.
+We then train the model on the training set and evaluate it on the validation set.
+We use beam search to generate the captions.
 
@@
 
-Your text here...
+Comparison of different models, evaluation, and discussion.
 
@@
 
-Your text here...
+Summary of the results, future work, and open questions.
 
@@
 \nocite{*}
-
+\bibliographystyle{apalike}
 \bibliography{bib}
-\bibliographystyle{plain}
 

### 2024-10-26 BAI/paper/src/__main__.py

@@ -1,10 +1,10 @@
 from torch import cuda
 from torchvision import transforms as T
 
-from models import ImageCaption
+from .models import ImageCaption
 
 model = ImageCaption(
-    "data/micro_flickr",
+    "flickr8k",
     batch_size=4,
     num_workers=2,
     embed_size=300,

### 2024-10-26 BAI/paper/src/build.sh

@@ -1,0 +1,9 @@
+FILE=model.py
+
+# 'compile' all python files to the model.py file
+rm $FILE
+cat **/*.py >> $FILE
+cat __main__.py >> $FILE
+
+# remove all relative imports
+sed -i 's/from \..*//g' $FILE

### 2024-10-26 BAI/paper/src/data/flickrdataset.py

@@ -1,4 +1,3 @@
-from data.vocabulary import Vocabulary
 import os
 from torch.nn.utils.rnn import pad_sequence
 from torch.utils.data import DataLoader, Dataset
@@ -6,6 +5,8 @@ from PIL import Image
 from pandas import read_csv
 from torch import tensor, Tensor, cat
 from torchvision.transforms import Compose
+
+from .vocabulary import Vocabulary
 
 
 class FlickrDataset(Dataset):

### 2024-10-26 BAI/paper/src/data/vocabulary.py

@@ -1,14 +1,16 @@
 from collections import Counter
-from spacy import load
+from importlib import import_module
+from spacy import Language
+from spacy.cli import download
 
 
 class Vocabulary:
     """ Vocabulary object to store the mapping between words and indices. """
 
-    def __init__(self, threshold: int, sentences: list[str] = [], spacy_model: str = "en_core_web_sm"):
+    def __init__(self, threshold: int, sentences: list[str] = [], spacy_model_name: str = "en_core_web_sm"):
         self.threshold = threshold
         self.id_to_token = {0: "<PAD>", 1: "<SOS>", 2: "<EOS>", 3: "<UNK>"}
-        self.spacy_model = load(spacy_model)
+        self.spacy_model = self._download_and_init_nlp(spacy_model_name)
 
         self._learn(sentences)
 
@@ -38,3 +40,17 @@ class Vocabulary:
         """ Tokenize the text using spacy model. """
         return [token.text.lower()
                 for token in self.spacy_model.tokenizer(text)]
+
+    def _download_and_init_nlp(self, model_name: str) -> Language:
+        """Load a spaCy model, download it if it has not been installed yet.
+        :param model_name: the model name, e.g., en_core_web_sm
+        :param kwargs: options passed to the spaCy loader, such as component exclusion
+        :return: an initialized spaCy Language
+        """
+        try:
+            model_module = import_module(model_name)
+        except ModuleNotFoundError:
+            download(model_name)
+            model_module = import_module(model_name)
+
+        return model_module.load()

### 2024-10-26 BAI/paper/src/download.sh

@@ -1,12 +1,0 @@
-DEST=${1:-/workspaces/w24/BAI/paper/src/data/flickr8k}
-
-# Download flickr8k dataset#
-if [ ! -d $DEST ]; then
-  mkdir -p $DEST
-fi
-curl -L -o $DEST/flickr8k.zip https://www.kaggle.com/api/v1/datasets/download/adityajn105/flickr8k
-unzip $DEST/flickr8k.zip -d $DEST
-rm $DEST/flickr8k.zip
-
-# Download spacy model
-spacy download en_core_web_sm

### 2024-10-26 BAI/paper/src/model.py

@@ -1,0 +1,483 @@
+
+
+import os
+from torch.nn.utils.rnn import pad_sequence
+from torch.utils.data import DataLoader, Dataset
+from PIL import Image
+from pandas import read_csv
+from torch import tensor, Tensor, cat
+from torchvision.transforms import Compose
+
+
+
+
+class FlickrDataset(Dataset):
+    """ Flickr dataset for image captioning. """
+
+    def __init__(self, path: str, captions: str = "captions.txt", transform: Compose = None, threshold: int = 5):
+        self.path = path
+        self.images_with_captions = read_csv(os.path.join(self.path, captions))
+        self.transform = transform
+        self.vocab = Vocabulary(threshold, self.captions)
+
+    def __len__(self) -> int: return len(self.captions)
+
+    def __getitem__(self, index: int) -> tuple[Tensor, Tensor]:
+        raw_caption = self.captions[index]
+        numericalized_caption = [self.vocab.token_to_id["<SOS>"]] + \
+            self.vocab.numericalize(raw_caption) + \
+            [self.vocab.token_to_id["<EOS>"]]
+        caption = tensor(numericalized_caption)
+
+        image_path = os.path.join(self.path, "Images", self.images[index])
+        image = Image.open(image_path).convert("RGB")
+        if self.transform:
+            image = self.transform(image)
+
+        return image, caption
+
+    @property
+    def images(self) -> list[str]:
+        return self.images_with_captions["image"].tolist()
+
+    @property
+    def captions(self) -> list[str]:
+        return self.images_with_captions["caption"].tolist()
+
+    def data_loader(self, batch_size: int = 4, num_workers: int = 1, shuffle: bool = True):
+        return DataLoader(
+            dataset=self,
+            batch_size=batch_size,
+            num_workers=num_workers,
+            shuffle=shuffle,
+            collate_fn=FlickrDataset.CapsCollate(
+                pad_index=self.vocab.token_to_id["<PAD>"], batch_first=True)
+        )
+
+    class CapsCollate:
+        """ Collate to apply the padding to the captions with dataloader. """
+
+        def __init__(self, pad_index: int, batch_first: bool = False):
+            self.pad_index = pad_index
+            self.batch_first = batch_first
+
+        def __call__(self, batch: list[tuple[Tensor, Tensor]]) -> tuple[Tensor, Tensor]:
+            """ Apply padding to the captions. """
+            images = [item[0].unsqueeze(0) for item in batch]
[…]

### 2024-10-26 BAI/paper/src/models/decoder.py

@@ -2,7 +2,7 @@ from torch import zeros, device, cuda, cat, Tensor, tensor
 from torch.nn import Module, Embedding, LSTMCell, Linear, Dropout
 from typing import Tuple
 
-from data import Vocabulary
+from ..data import Vocabulary
 from .attention import Attention
 
 

### 2024-10-26 BAI/paper/src/models/image_caption.py

@@ -5,7 +5,7 @@ from torch.optim import Adam
 from matplotlib import pyplot as plt
 from typing import Tuple
 
-from data import FlickrDataset
+from ..data import FlickrDataset
 from .encoder import ImageEncoder
 from .decoder import CaptionDecoder
 

### 2024-10-26 BAI/paper/paper.tex
Added some more content to bai paper.

--- 2024-10-24 BAI/paper/paper.tex
@@
 \usepackage{fancyhdr}                   % Packet for nicer headers
-\usepackage{fancyheadings}             % Nicer numbering of headlines
 
@@
 %\usepackage{latexsym}                  % Special symbols
-\usepackage{longtable}									% Allow tables over several parges
+\usepackage{longtable}                  % Allow tables over several pages
 \usepackage{listings}                   % Nicer source code listings
@@
 \begin{abstract}
-  This paper is about image captioning. We will discuss the current state of the art and give an overview of the most important methods. We will also present some of our own results.
+ This paper is about image captioning. We discuss the current state of the art and overview the most important methods. We also present some of our results.
 \end{abstract}
@@
 
-Your text here...
+Image captioning involves generating textual descriptions of given images.
+It is a difficult problem because it combines the visual and textual domains.
+In recent years, deep learning has revolutionized the field of image captioning.
+
+This paper presents an overview of the most essential methods in image captioning and some of our results.
+
+Image captioning is a complex and interdisciplinary task bridging computer vision (CV) and natural language processing (NLP).
+The objective is to generate natural, descriptive sentences for images, which involves detecting and interpreting visual elements and their relationships within scenes.
+Recent advancements leverage deep learning, particularly Convolutional Neural Networks (CNNs) for feature extraction and Recurrent Neural Networks (RNNs), especially Long Short-Term Memory (LSTM) networks, for sequential language generation.
+For instance, models such as "Show and Tell" \cite{Vinyals:2015} and "Show, Attend, and Tell" \cite{Xu} utilize CNN-RNN architectures, with the latter integrating attention mechanisms to enhance focus on specific image regions, thereby improving caption relevance and accuracy.
+
+Image captioning models are typically evaluated using metrics like BLEU, METEOR, and CIDEr, which assess linguistic similarity to human references.
+While these metrics are widely accepted, considering the quality of image captions remains challenging due to the subjective nature of language generation.
+Additionally, methods like Reinforcement Learning (RL) have been explored to optimize these models directly on evaluation metrics, addressing issues such as exposure bias by aligning training and inference methods more closely \cite{Rennie}.
+
+Recent innovations in diverse decoding strategies, like Diverse Beam Search, aim to overcome the tendency of beam search to produce near-duplicate captions by enhancing variability without sacrificing quality.
+Diverse Beam Search ensures that generated captions capture a broader range of plausible descriptions, reflecting the inherent ambiguity in image captioning tasks \cite{Vijayakumar}.
+
+\section{Methods}
+\label{sec:methods}
+
+The accuracy of image captioning is commonly measured using BLEU or CIDEr scores.
+BLEU is a precision-based metric that compares n-grams in the generated caption to those in the reference captions.
+CIDEr is a recall-based metric that compares the generated caption to the reference captions using cosine similarity.
+
+We focused on the flickr8k dataset, which consists of 8,000 images, each with five reference captions.
+Other commonly used datasets are flickr30k and MS COCO, but due to their size and limited resources, we decided to focus on flickr8k.
+
+Usually, an encoder-decoder architecture is used to achieve image captioning.
+The encoder is a convolutional neural network (CNN) that extracts features from the image.
+The decoder is a recurrent neural network (RNN) that generates the caption word by word.
+
+We experiment with different architectures, such as VGG16, ResNet, and InceptionV3 as encoders and LSTM and GRU as decoders.
+We also experiment with varying attention mechanisms, such as soft and hard attention.
 
@@
 
-Your text here...
+We follow the state-of-the-art approach to image captioning.
+We preprocess the images using the encoder and tokenize the captions.
+We then train the model on the training set and evaluate it on the validation set.
+We generate the captions using beam search.
 

### 2024-10-26 BAI/paper/src/assemble.sh
Updated assemble script.

@@ -1,0 +1,12 @@
+DIR=${1:-"/workspaces/w24/bai/paper/src"}
+FILE=$DIR/${2:-"__assembled__"}.py
+
+# 'compile' all python files to the model.py file
+if [ -f $FILE ]; then
+    rm $FILE
+fi
+cat $DIR/**/*.py >> $FILE
+cat $DIR/__main__.py >> $FILE
+
+# remove all relative imports
+sed -i 's/from \..*//g' $FILE

### 2024-10-26 BAI/paper/src/build.sh

@@ -1,9 +1,0 @@
-FILE=model.py
-
-# 'compile' all python files to the model.py file
-rm $FILE
-cat **/*.py >> $FILE
-cat __main__.py >> $FILE
-
-# remove all relative imports
-sed -i 's/from \..*//g' $FILE

### 2024-10-26 BAI/paper/src/model.py

@@ -1,483 +1,0 @@
-
-
-import os
-from torch.nn.utils.rnn import pad_sequence
-from torch.utils.data import DataLoader, Dataset
-from PIL import Image
-from pandas import read_csv
-from torch import tensor, Tensor, cat
-from torchvision.transforms import Compose
-
-
-
-
-class FlickrDataset(Dataset):
-    """ Flickr dataset for image captioning. """
-
-    def __init__(self, path: str, captions: str = "captions.txt", transform: Compose = None, threshold: int = 5):
-        self.path = path
-        self.images_with_captions = read_csv(os.path.join(self.path, captions))
-        self.transform = transform
-        self.vocab = Vocabulary(threshold, self.captions)
-
-    def __len__(self) -> int: return len(self.captions)
-
-    def __getitem__(self, index: int) -> tuple[Tensor, Tensor]:
-        raw_caption = self.captions[index]
-        numericalized_caption = [self.vocab.token_to_id["<SOS>"]] + \
-            self.vocab.numericalize(raw_caption) + \
-            [self.vocab.token_to_id["<EOS>"]]
-        caption = tensor(numericalized_caption)
-
-        image_path = os.path.join(self.path, "Images", self.images[index])
-        image = Image.open(image_path).convert("RGB")
-        if self.transform:
-            image = self.transform(image)
-
-        return image, caption
-
-    @property
-    def images(self) -> list[str]:
-        return self.images_with_captions["image"].tolist()
-
-    @property
-    def captions(self) -> list[str]:
-        return self.images_with_captions["caption"].tolist()
-
-    def data_loader(self, batch_size: int = 4, num_workers: int = 1, shuffle: bool = True):
-        return DataLoader(
-            dataset=self,
-            batch_size=batch_size,
-            num_workers=num_workers,
-            shuffle=shuffle,
-            collate_fn=FlickrDataset.CapsCollate(
-                pad_index=self.vocab.token_to_id["<PAD>"], batch_first=True)
-        )
-
-    class CapsCollate:
-        """ Collate to apply the padding to the captions with dataloader. """
-
-        def __init__(self, pad_index: int, batch_first: bool = False):
-            self.pad_index = pad_index
-            self.batch_first = batch_first
-
-        def __call__(self, batch: list[tuple[Tensor, Tensor]]) -> tuple[Tensor, Tensor]:
-            """ Apply padding to the captions. """
-            images = [item[0].unsqueeze(0) for item in batch]
[…]

### 2024-10-27 BAI/paper/src/data/flickrdataset.py
Got the bai image captioning model working.

@@ -1,9 +1,8 @@
 import os
-from torch.nn.utils.rnn import pad_sequence
-from torch.utils.data import DataLoader, Dataset
+from torch.utils.data import Dataset, random_split
 from PIL import Image
 from pandas import read_csv
-from torch import tensor, Tensor, cat
+from torch import tensor, Tensor
 from torchvision.transforms import Compose
 
 from .vocabulary import Vocabulary
@@ -12,60 +11,93 @@ from .vocabulary import Vocabulary
 class FlickrDataset(Dataset):
     """ Flickr dataset for image captioning. """
 
-    def __init__(self, path: str, captions: str = "captions.txt", transform: Compose = None, threshold: int = 5):
-        self.path = path
-        self.images_with_captions = read_csv(os.path.join(self.path, captions))
-        self.transform = transform
-        self.vocab = Vocabulary(threshold, self.captions)
+    def __init__(self,
+                 image_path: str,
+                 images: list[str],
+                 captions: list[str],
+                 vocabulary: Vocabulary,
+                 image_transform: Compose):
+        """
+        Args:
+            image_path (str): Path to the folder containing the images.
+            images (list[str]): List of image filenames.
+            captions (list[str]): List of captions for the images.
+            vocabulary (Vocabulary): Vocabulary object for the captions.
+            transform_images (Compose): Transform to apply to images.
+        """
+        self.image_path = image_path
+        self.images = images
+        self.captions = captions
+        self.vocabulary = vocabulary
+        self.image_transform = image_transform
 
     def __len__(self) -> int: return len(self.captions)
 
     def __getitem__(self, index: int) -> tuple[Tensor, Tensor]:
-        raw_caption = self.captions[index]
-        numericalized_caption = [self.vocab.token_to_id["<SOS>"]] + \
-            self.vocab.numericalize(raw_caption) + \
-            [self.vocab.token_to_id["<EOS>"]]
-        caption = tensor(numericalized_caption)
+        return self.get_image_tensor(index), self.get_caption(index)
 
-        image_path = os.path.join(self.path, "Images", self.images[index])
+    def get_image_tensor(self, index: int) -> Tensor:
+        """ Get the image tensor for the given index. """
+        return self.get_image_tensor_at_path(os.path.join(self.image_path, "Images", self.images[index]))
+
+    def get_image_tensor_at_path(self, image_path: str) -> Tensor:
+        """ Get the image tensor for the given path. """
         image = Image.open(image_path).convert("RGB")
-        if self.transform:
-            image = self.transform(image)
[…]

### 2024-10-27 BAI/paper/src/models/encoder.py

@@ -14,14 +14,14 @@ class ImageEncoder(Module):
 
         for param in resnet.parameters():
             param.requires_grad_(False)
-
         modules = list(resnet.children())[:-2]
+
         self.model = Sequential(*modules)
 
-    def forward(self, images: Tensor) -> Tensor:
-        """ Extract feature vectors from input images. """
+    def forward(self, image: Tensor) -> Tensor:
+        """ Extract feature vector from input image. """
         # (batch_size,2048,7,7)
-        features: Tensor = self.model(images)
+        features: Tensor = self.model(image)
 
         # (batch_size,7,7,2048)
         features = features.permute(0, 2, 3, 1)

### 2024-10-29 SbD/summary.tex
Updated template and added slides to sbd summary.
Added SbD summary for crypto chapter.

@@
 
-\usepackage[]{../.template/summary}
+\usepackage{../.template/summary}
 
@@
 
-\includepdf[pages=1-10, nup=2x5]{slides.pdf}
-
 \section{Kryptographie}
 
-\subsection{Angriffsarten}
+\includepdf[pages={14-16}, nup=1x3]{slides.pdf}
+\subsection{Grundlagen}
+\subsubsection{Angriffsarten (schwer zu leicht)}
 \begin{enumerate}
@@
 
-\subsection{Brechen (stark zu schwach)}
+\subsubsection{Brechen (stark zu schwach)}
 \begin{enumerate}
@@
 
+\includepdf[pages={18-32,35,37-41}, nup=3x7]{slides.pdf}
 \subsection{Klassische Chiffren}
@@
 
-\subsection{Moderne Kryptographie}
+\includepdf[pages={44-108}, nup=4x9]{slides.pdf}
+\subsection{Moderene Kryptographie I: Symmetrische Systeme}
+\begin{enumerate}
+  \item \textbf{One-Time-Pad (mod 2)}: Vernam-Chiffre mit zufälligem Schlüssel
+  \item \textbf{Symmetrische Authentifikationscodes}: Erweiterung des One-Time-Pad mit Message Authentication Code (MAC)
+  \item \textbf{DES (Data Encryption Standard)}: 56-Bit-Blockchiffre nach Feistel-Prinzip
+  \item \textbf{IDEA (International Data Encryption Algorithm)}: 128-Bit-Blockchiffre mit XOR, Addition und Multiplikation
+  \item \textbf{AES (Advanced Encryption Standard)}: 128-Bit-Blockchiffre mit 10, 12 oder 14 Runden
+\end{enumerate}
+
+\subsubsection{Gütekriterien}
+\begin{itemize}
+  \item \textbf{Vollständigkeit}: Jedes Output-Bit hängt von jedem Input-Bit ab
+  \item \textbf{Avalanche}: Jede Änderung im Input führt zu vielen Änderungen im Output
+  \item \textbf{Nichtlinearität}: Output ist nicht linear abhängig vom Input
+  \item \textbf{Korrelationsimmunität}: Keine Korrelation zwischen Input und Output
+  \item sekundäre Kriterien: Implementierbarkeit, Längentreue, Schnelligkeit
+\end{itemize}
+
+\subsubsection{Feistel-Prinzip}
+\begin{itemize}
+  \item \textbf{In Hälften teilen}: Klartext wird in zwei Hälften geteilt
+  \item \textbf{Runden}: Rundenfunktion wird auf eine Hälfte angewendet und das Ergebnis mit der anderen Hälfte XOR-verknüpft
+  \item \textbf{Austauschen der Hälften}: Die Hälften werden getauscht
+  \item \textbf{Dekodierung}: Entschlüsselung ist identisch zur Verschlüsselung (Runden umgekehrt)
+\end{itemize}
+
+\subsubsection{DES (Data Encryption Standard)}
+\begin{itemize}
+  \item[+] vollständig
+  \item[+] kein analytischer Zusammenhang zwischen Input und Output
+  \item[+] invariant ggü. Komplementbildung
+  \item[-] Designkriterien nicht öffentlich (inzwischen bekannt)
+  \item[-] ineffiziente Implementierung wg. Permutationen
+  \item[-] Schlüssellänge zu kurz (56 Bit)
+  \item \textbf{möglicher Angriff}: Known-Plaintext-Angriff mit $2^{57}$ Aufwand
+\end{itemize}
+
+\subsubsection{IDEA (International Data Encryption Algorithm)}
+\begin{itemize}
+  \item \textbf{Schlüssellänge}: 128-Bit-Blockchiffre, entwickelt als sicherer Ersatz für DES.
+  \item \textbf{Struktur}: Verwendet eine Kombination aus XOR-, Addition- und Multiplikationsoperationen für hohe Sicherheit.
+  \item \textbf{Sicherheitsvorteile}: Resistenz gegen lineare und differentielle Kryptoanalyse.
+  \item \textbf{Verwendung}: Besonders populär in Europa und wurde in PGP (Pretty Good Privacy) integriert.
+\end{itemize}
+
+\subsubsection{AES (Advanced Encryption Standard)}
+\begin{itemize}
+  \item \textbf{Schlüssellänge}: Unterstützt 128, 192 und 256 Bit, damit für verschiedene Sicherheitsniveaus geeignet.
+  \item \textbf{Struktur}: Verwendet 10, 12 oder 14 Runden von Byte-Substitution, Zeilenverschiebung, Spaltenmischung und Rundenschlüssel-Addition.
+  \item \textbf{Sicherheitsniveau}: Sicher gegen Brute-Force- und bekannte Kryptoangriffe, empfohlen von NIST als sicherer Standard.
+  \item \textbf{Verwendung}: Weltweiter Standard für Verschlüsselung in Regierung, Industrie und dem Finanzsektor.
+\end{itemize}
+
+\subsubsection{Betriebsarten symmetrischer Blockchiffren}
+\begin{enumerate}
+  \item \textbf{ECB (Electronic Codebook Mode)}: Klartext wird in Blöcke geteilt und einzeln verschlüsselt
+  \item \textbf{CBC (Cipher Block Chaining Mode)}: Klartext wird in Blöcke geteilt und XOR-verknüpft
+  \item \textbf{CTR (Counter Mode)}: Klartext wird in Blöcke geteilt und mit Zähler verschlüsselt
+  \item \textbf{OFB (Output Feedback Mode)}: Klartext wird in Blöcke geteilt und mit Zähler verschlüsselt
+  \item \textbf{CFB (Cipher Feedback Mode)}: Klartext wird in Blöcke geteilt und mit Zähler verschlüsselt
+\end{enumerate}
+
+\includepdf[pages={44, 109-142}, nup=4x9]{slides.pdf}
+\subsection{Moderne Kryptographie II: Asymmetrische Systeme}
+\begin{enumerate}
+  \item \textbf{Diffie-Hellman-Key-Exchange}: Schlüsselaustausch über unsicheren Kanal; Sicherheit durch diskretes Logarithmusproblem, anfällig für Man-in-the-Middle-Angriffe.
+  \item \textbf{ElGamal Kryptosystem}: Asymmetrische Verschlüsselung und Signatur; Diffie-Hellman-basiert, stärkere Sicherheit, aber höhere Schlüsselanforderungen.
+  \item \textbf{RSA zur Konzelation und Signatur}: Asymmetrische Verschlüsselung auf Faktorisierungsannahme; verbreitet in SSL/TLS, bietet Vertraulichkeit und Authentizität.
+  \item \textbf{Blinde Signaturen mit RSA}: Signatur ohne Kenntnis des Inhalts für Anonymität; Anwendung in elektronischen Abstimmungen und digitalen Währungen.
+  \item \textbf{Kryptosysteme auf Basis elliptischer Kurven}: Hohe Sicherheit bei kleinen Schlüsseln; effizient und ideal für mobile/räumlich beschränkte Geräte.
+\end{enumerate}
+
+\subsubsection{Mathematische Grundlagen}
+\begin{itemize}
+  \item \textbf{Modulo}: $a \mod b = a - b \cdot \lfloor \frac{a}{b} \rfloor$
+  \item \textbf{Erweiterter Euklidischer Algorithmus}: $ggT(a, b) = a \cdot x + b \cdot y$
+  \item \textbf{Euler-Phi-Funktion}: $\varphi(n) = |\{a \in \mathbb{N} \mid 1 \leq a < n, ggT(a, n) = 1\}|$
+  \item \textbf{Primitive Wurzel}: $g$ ist primitive Wurzel modulo $p$, wenn $g^i \mod p \neq 1$ für $i < p-1$
+  \item \textbf{Diskreter Logarithmus}: $a = g^x \mod p \Rightarrow x = \log_g a \mod p$
+  \item \textbf{Faktorisierungsannahme}: $n = p \cdot q \Rightarrow p, q \in \mathbb{P}$
+\end{itemize}
+
+\subsubsection{Diffie-Hellman-Key-Exchange}
+\begin{itemize}
+  \item \textbf{Ziel}: Ermöglicht sicheren Schlüsselaustausch zwischen zwei Parteien über unsichere Kanäle ohne vorherigen geheimen Austausch.
+  \item \textbf{Funktionsweise}: Berechnung gemeinsamer geheimer Schlüssel über Modulorechnung basierend auf einem öffentlichen Basiswert und Primzahl.
+  \item \textbf{Sicherheitsprinzip}: Sicherheit basiert auf der Schwierigkeit des diskreten Logarithmusproblems.
+  \item \textbf{Schwachstelle}: Anfällig für Man-in-the-Middle-Angriffe, wenn nicht authentifiziert.
+\end{itemize}
+
+\subsubsection{ElGamal Kryptosystem}
+\begin{itemize}
+  \item \textbf{Verwendung}: Für Verschlüsselung und digitale Signaturen geeignet.
+  \item \textbf{Schlüssellänge}: Bietet längere Schlüssel für höhere Sicherheit, aber weniger Effizienz im Vergleich zu RSA.
+  \item \textbf{Sicherheitsprinzip}: Beruht auf der Schwierigkeit des diskreten Logarithmusproblems.
+  \item \textbf{Vorteile}: Flexibel und erweiterbar auf elliptische Kurven für mehr Effizienz.
+\end{itemize}
+
+\subsubsection{RSA zur Konzelation und Signatur}
+\begin{itemize}
+  \item \textbf{Verwendung}: Verschlüsselung und digitale Signatur mit öffentlichen und privaten Schlüsseln.
+  \item \textbf{Schlüssellänge}: Variable Schlüssellänge (typisch 1024, 2048 oder 4096 Bit), längere Schlüssel bieten höhere Sicherheit.
+  \item \textbf{Grundlage}: Faktorisierungsannahme, d.h., die Schwierigkeit der Primfaktorzerlegung großer Zahlen.
+  \item \textbf{Anwendung}: Weltweit in SSL/TLS und anderen Sicherheitssystemen implementiert.
+\end{itemize}
+
+\subsubsection{Blinde Signaturen mit RSA}
+\begin{itemize}
+  \item \textbf{Prinzip}: Der Signierer erstellt eine Signatur ohne den Inhalt der Nachricht zu kennen (Anonymität).
+  \item \textbf{Verfahren}: Der Verfasser maskiert die Nachricht, bevor sie zur Signatur übergeben wird, und entfernt die Maske nach der Signierung.
+  \item \textbf{Anwendung}: Nützlich in Szenarien wie elektronischen Abstimmungen, bei denen die Vertraulichkeit wichtig ist.
+\end{itemize}
+
+\subsubsection{Kryptosysteme auf Basis elliptischer Kurven}
+\begin{itemize}
+  \item \textbf{Effizienz}: Höhere Sicherheit bei kleineren Schlüssellängen (z.B. ein 256-Bit-Schlüssel bietet Sicherheit ähnlich einem 3072-Bit-RSA-Schlüssel).
+  \item \textbf{Grundlage}: Elliptische Kurven über endlichen Feldern, basierend auf dem diskreten Logarithmusproblem.
+  \item \textbf{Verwendung}: Ideal für mobile und ressourcenbeschränkte Geräte, oft verwendet in modernen Sicherheitsprotokollen (z.B. ECDSA für digitale Signaturen).
+\end{itemize}
+
+\section{Public Key Infrastructure}
+\includepdf[pages={1}, nup=3x7]{slides.pdf}
+
+\section{Attacks and Defenses}
+\includepdf[pages={1}, nup=3x7]{slides.pdf}
+
+\section{Mobile Security}
+\includepdf[pages={1}, nup=3x7]{slides.pdf}
 

### 2024-10-29 BAI/paper/src/__main__.py
Updated BAI implementation.

@@ -1,8 +1,9 @@
 from torch import cuda, device
 from torchvision import transforms as T
 
-from data import LoadedFlickrDataset, Visualiser
+from data import LoadedFlickrDataset
 from models import ImageCaption, ImageEncoder, CaptionDecoder
+from eval import Visualiser
 
 DIR = "/content/drive/MyDrive"
 # DIR = "."

### 2024-10-29 BAI/paper/src/assemble.sh

@@ -20,5 +20,6 @@ cat $DIR/__main__.py >> $FILE
 sed -i 's/from \..* import .*//g' $FILE
 sed -i 's/from data import .*//g' $FILE
 sed -i 's/from models import .*//g' $FILE
+sed -i 's/from eval import .*//g' $FILE
 
 printf "Assembled python files to $FILE\n"

### 2024-10-29 BAI/paper/src/data/__init__.py

@@ -1,4 +1,3 @@
 from .flickrdataset import FlickrDataset, LoadedFlickrDataset
 from .vocabulary import Vocabulary
 from .flickrdataloader import FlickrDataloader
-from .visualiser import Visualiser

### 2024-10-29 BAI/paper/src/__main__.py
(WIP) BAI.

@@ -2,7 +2,7 @@ from torch import cuda, device
 from torchvision import transforms as T
 
 from data import LoadedFlickrDataset
-from models import ImageCaption, ImageEncoder, CaptionDecoder
+from models import ImageCaption, ResnetImageEncoder, CaptionDecoder
 from eval import Visualiser
 
 DIR = "/content/drive/MyDrive"
@@ -23,7 +23,7 @@ dataset = LoadedFlickrDataset(
 )
 
 # --- Model ---
-encoder = ImageEncoder(
+encoder = ResnetImageEncoder(
 ).to(DEVICE)
 decoder = CaptionDecoder(
     embed_size=300,
@@ -50,3 +50,11 @@ visualiser = Visualiser()
 test_image = f"{DIR}/image.jpg"
 prediction = model.predict(test_image, transforms)
 visualiser.display_image(prediction[1], prediction[0])
+
+# def _loaders(self, batch_size: int, num_workers: int) -> tuple[FlickrDataloader, FlickrDataloader]:
+#         batch_size = batch_size if batch_size else self.batch_size
+#         num_workers = num_workers if num_workers else self.num_workers
+
+#         training, validation = self.dataset.split(train_size=.8)
+#         return (FlickrDataloader(training, batch_size=batch_size, num_workers=num_workers),
+#                 FlickrDataloader(validation, batch_size=batch_size, num_workers=num_workers))

### 2024-10-29 BAI/paper/src/data/flickrdataset.py

@@ -14,9 +14,7 @@ class FlickrDataset(Dataset):
     def __init__(self,
                  image_path: str,
                  images: list[str],
-                 captions: list[str],
-                 vocabulary: Vocabulary,
-                 image_transform: Compose):
+                 captions: list[str]):
         """
         Args:
             image_path (str): Path to the folder containing the images.
@@ -28,32 +26,15 @@ class FlickrDataset(Dataset):
         self.image_path = image_path
         self.images = images
         self.captions = captions
-        self.vocabulary = vocabulary
-        self.image_transform = image_transform
 
     def __len__(self) -> int: return len(self.captions)
 
-    def __getitem__(self, index: int) -> tuple[Tensor, Tensor]:
-        return self.get_image_tensor(index), self.get_caption(index)
-
-    def get_image_tensor(self, index: int) -> Tensor:
-        """ Get the image tensor for the given index. """
-        return self.get_image_tensor_at_path(os.path.join(self.image_path, "Images", self.images[index]))
-
-    def get_image_tensor_at_path(self, image_path: str) -> Tensor:
-        """ Get the image tensor for the given path. """
-        image = Image.open(image_path).convert("RGB")
-        if self.image_transform:
-            image = self.image_transform(image)
-        return image
-
-    def get_caption(self, index: int) -> str:
-        """ Get the encoded caption for the given index. """
-        raw = self.captions[index]
-        encoded = [self.vocabulary.token_to_id["<SOS>"]] + \
-            self.vocabulary.numericalize(raw) + \
-            [self.vocabulary.token_to_id["<EOS>"]]
-        return tensor(encoded)
+    def __getitem__(self, index: int) -> tuple[Image.Image, str]:
+        return (
+            Image.open(os.path.join(self.image_path,
+                       "Images", self.images[index])),
+            f"<SOS> {self.captions[index]} <EOS>"
+        )
 
     def split(self, train_size: float = 0.8) -> tuple["FlickrDataset", "FlickrDataset"]:
         """ Split the dataset into a training and validation set."""
@@ -61,14 +42,10 @@ class FlickrDataset(Dataset):
         subsets = random_split(self, [train_size, 1 - train_size])
         return (FlickrDataset(self.image_path,
                               [self.images[i] for i in subsets[0].indices],
-                              [self.captions[i] for i in subsets[0].indices],
-                              self.vocabulary,
-                              self.image_transform),
+                              [self.captions[i] for i in subsets[0].indices]),
                 FlickrDataset(self.image_path,
[…]

### 2024-10-29 BAI/paper/src/eval/__init__.py

@@ -1,1 +1,3 @@
+from train import train
+from evaluate import evaluate
 from visualiser import Visualiser

### 2024-10-29 BAI/paper/src/eval/evaluate.py

@@ -1,0 +1,52 @@
+from torch.nn import Module
+from nltk.translate import bleu_score
+
+from data import FlickrDataset
+
+
+def evaluate(model: Module, data: FlickrDataset):
+    references = []
+    candidates = []
+    for i in range(len(data)):
+        reference = data[i][1]
+        candidate = model.predict(data[i][0])
+        references.append(reference)
+        candidates.append(candidate)
+    return bleu_score.corpus_bleu(references, candidates)
+
+# def validateModel(self,
+#                       batch_size: int = None,
+#                       num_workers: int = None,
+#                       device: str = "cpu") -> float:
+#         """ Validate the model on the validation dataset.
+#         Args:
+#             batch_size: The number of samples in each batch (defaults to the model's batch size).
+#             num_workers: The number of subprocesses to use for data loading (defaults to the model's num_workers).
+#         Returns:
+#             The average loss on the validation dataset.
+#         """
+
+#         loader = self._validation_loader(
+#             batch_size=batch_size, num_workers=num_workers)
+
+#         total_loss = 0
+
+#         self.eval()
+#         for idx, (images, captions) in enumerate(loader):
+#             print(f"[Validation] Batch {idx+1}/{len(loader)}")
+
+#             images = images.to(device)
+#             captions = captions.to(device)
+
+#             with no_grad():
+#                 outputs, _ = self(images, captions)
+
+#             targets = captions[:, 1:]
+#             loss: CrossEntropyLoss = self.criterion(
+#                 outputs.to(device).view(-1, len(loader.dataset.vocabulary)), targets.reshape(-1))
+
+#             total_loss += loss.item()
+
+#         loss = total_loss / len(loader)
+#         print(f"\n[Validation] Loss: {loss}")
+#         return loss

### 2024-10-29 BAI/paper/src/eval/train.py

@@ -1,0 +1,59 @@
+
+
+#         self.optimizer = Adam(self.parameters(), lr=learning_rate)
+#         self.criterion = loss_fn if loss_fn else CrossEntropyLoss(
+#             ignore_index=dataset.vocabulary.token_to_id["<PAD>"])
+# def trainModel(self,
+#                 epochs: int = 5,
+#                 batch_size: int = None,
+#                 num_workers: int = None,
+#                 resume: bool = False,
+#                 device: str = "cpu"):
+#         """" Train the model for a given number of epochs.
+#         Args:
+#             batch_size: The number of samples in each batch (defaults to the model's batch size).
+#             num_workers: The number of subprocesses to use for data loading (defaults to the model's num_workers).
+#             epochs: The number of times to iterate over the training dataset.
+#             resume: Whether to resume training from the last checkpoint.
+#         """
+
+#         start_epoch = self._load() if resume else 0
+#         if start_epoch == epochs:
+#             print("[Training] Model already trained.")
+#             return
+#         elif start_epoch > epochs:
+#             raise ValueError(
+#                 f"Model has already been trained for {start_epoch} epochs.")
+#         elif start_epoch > 0:
+#             print(f"[Training] Resuming from epoch {start_epoch}.")
+
+#         loader = self._training_loader(batch_size, num_workers)
+
+#         self.train()
+#         for epoch in range(start_epoch, epochs):
+#             for idx, (images, captions) in enumerate(loader):
+#                 print(
+#                     f"[Training | Epoch {epoch+1}/{epochs}] Batch {idx+1}/{len(loader)}")
+
+#                 images = images.to(device)
+#                 captions = captions.to(device)
+
+#                 # Zero the gradients.
+#                 self.optimizer.zero_grad()
+
+#                 # Feed forward
+#                 outputs, _ = self(images, captions)
+
+#                 # Calculate the batch loss.
+#                 targets = captions[:, 1:]
+#                 loss = self.criterion(
+#                     outputs.to(device).view(-1, len(loader.dataset.vocabulary)), targets.reshape(-1))
+
+#                 # Backward pass.
+#                 loss.backward()
+
+#                 # Update the parameters in the optimizer.
+#                 self.optimizer.step()
+
+#             self._save(epoch+1)
+#             print(f"\n[Training | Epoch {epoch+1}/{epochs}] Loss: {loss}\n")

### 2024-10-29 BAI/paper/src/models/__init__.py

@@ -1,3 +1,3 @@
 from .image_caption import ImageCaption
-from .encoder import ImageEncoder
+from .encoder import ResnetImageEncoder
 from .decoder import CaptionDecoder

### 2024-10-29 BAI/paper/src/models/checkpoint.py

@@ -1,0 +1,37 @@
+from torch import save, load
+from torch.nn import Module
+
+
+class CheckpointModule(Module):
+    """ A module that can save and load checkpoints. """
+
+    def __init__(self, path: str):
+        super(self).__init__()
+        self.path = path
+
+    def save_checkpoint(self, checkpoint: int):
+        """ Save the model at a checkpoint. """
+        save({
+            "checkpoint": checkpoint,
+            "state": self.state_dict()
+        }, self.path)
+
+        print(f"[Save] To {self.path} at checkpoint {checkpoint}.")
+
+    def load_checkpoint(self) -> int:
+        """ Load the model from the last checkpoint. """
+        try:
+            checkpoint = load(self.path, weights_only=True)
+            model_dict = self.state_dict()
+            pretrained_dict = {k: v for k, v in checkpoint['state'].items()
+                               if k in model_dict and model_dict[k].shape == v.shape}
+            model_dict.update(pretrained_dict)
+
+            self.load_state_dict(model_dict)
+
+            print(
+                f"[Load] From {self.path} at checkpoint {checkpoint['checkpoint']}.")
+
+            return checkpoint['checkpoint']
+        except FileNotFoundError:
+            return 0

### 2024-10-29 BAI/paper/src/models/decoder.py

@@ -1,97 +1,83 @@
-from torch import zeros, cat, Tensor, tensor
-from torch.nn import Module, Embedding, LSTMCell, Linear, Dropout
-
+from torch import Tensor, tensor
+from torch.nn import Module, Embedding, GRUCell, Linear
 from data import Vocabulary
-from .attention import Attention
 
 
-class CaptionDecoder(Module):
-    """ An RNN-based language model that generates image captions. """
+class SimpleCaptionDecoder(Module):
+    """A simplified GRU-based language model for generating image captions."""
 
-    def __init__(self,
-                 vocab_size: int,
-                 embed_size: int,
-                 attention_dim: int,
-                 encoder_dim: int,
-                 decoder_dim: int,
-                 drop_probability: float = 0.3):
-        super(CaptionDecoder, self).__init__()
+    def __init__(self, vocab: Vocabulary, embed_size: int = 256, decoder_dim: int = 512):
+        """
+        Args:
+            vocab (Vocabulary): Vocabulary instance for token-to-id mappings.
+            encoder_dim (int): Dimension of the encoder output features.
+            embed_size (int): Size of the word embedding vectors (default: 256).
+            decoder_dim (int): Dimension of the GRU hidden state (default: 512).
+        """
+        super().__init__()
 
-        self.vocab_size = vocab_size
+        self.vocab = vocab
+        self.vocab_size = len(vocab)
+        self.embed_size = embed_size
+        self.decoder_dim = decoder_dim
 
-        self.embedding_layer = Embedding(vocab_size, embed_size)
-        self.attention_layer = Attention(
-            encoder_dim, decoder_dim, attention_dim)
-        self.init_cell = Linear(encoder_dim, decoder_dim)
-        self.init_hidden = Linear(encoder_dim, decoder_dim)
-        self.lstm_layer = LSTMCell(
-            embed_size+encoder_dim, decoder_dim, bias=True)
-        self.output_layer = Linear(decoder_dim, vocab_size)
-        self.drop_layer = Dropout(drop_probability)
+        # Layers for embedding, GRU, and output projection
+        self.embedding_layer = Embedding(self.vocab_size, self.embed_size)
+        self.gru_cell = GRUCell(self.embed_size, self.decoder_dim)
+        self.output_layer = Linear(self.decoder_dim, self.vocab_size)
 
-    def forward(self, features: Tensor, captions: Tensor) -> tuple[Tensor, Tensor]:
[…]

### 2024-10-29 BAI/paper/src/models/encoder.py

@@ -1,33 +1,50 @@
-from torch import Tensor
+from torch import Tensor, stack
 from torch.nn import Module, Sequential
 from torchvision.models import resnet50, ResNet
+from PIL.Image import Image
+from torchvision.transforms import Compose, Resize, ToTensor, Normalize
 
 
-class ImageEncoder(Module):
-    """ A CNN-based ResNet-50 model that is used to encode the images to a feature vector. """
+class ResnetImageEncoder(Module):
+    """ Image encoder using a pretrained ResNet-50 model. """
 
     def __init__(self):
-        """ Load the pretrained ResNet-50 and replace top fc layer. """
-        super(ImageEncoder, self).__init__()
+        super().__init__()
+        self.dimension = 2048
+        self.model = self._get_passive_resnet()
 
-        resnet: ResNet = resnet50(weights='ResNet50_Weights.DEFAULT')
+    def forward(self, images: list[Image]) -> Tensor:
+        """ Extract features from a batch of images.
 
+        Args:
+            images (list[Image]): List of PIL images.
+
+        Returns:
+            Tensor: Encoded features with shape (batch_size, 49, 2048).
+        """
+        images = self.preprocess_images(images)  # (batch_size, 3, 224, 224)
+        features: Tensor = self.model(images)  # (batch_size, 2048, 7, 7)
+        features = features.permute(0, 2, 3, 1)  # (batch_size, 7, 7, 2048)
+        features = features.view(
+            features.size(0), -1, features.size(-1))  # (batch_size, 49, 2048)
+        return features
+
+    @staticmethod
+    def preprocess_images(images: list[Image]) -> Tensor:
+        transform = Compose([
+            Resize((224, 224)),  # ResNet expects images of 224x224
+            ToTensor(),  # Converts PIL image to a tensor with values [0, 1]
+            Normalize(mean=[0.485, 0.456, 0.406], std=[0.229, 0.224, 0.225])
+        ])
+        return stack([transform(image.convert("RGB")) for image in images])
+
+    def _get_passive_resnet(self) -> Sequential:
+        """ Prepare the ResNet-50 model by setting it to evaluation mode, freezing parameters, and removing the final layers to retain spatial information in the output.
+
+        Returns:
+            Sequential: Modified ResNet-50 model as a Sequential module.
+        """
+        resnet: ResNet = resnet50(weights='ResNet50_Weights.DEFAULT').eval()
         for param in resnet.parameters():
[…]

### 2024-10-29 BAI/paper/src/models/image_caption.py

@@ -1,192 +1,28 @@
-from torch import Tensor, save, no_grad, load
-from torch.nn import Module, CrossEntropyLoss
-from torchvision.transforms import Compose
-from torch.optim import Adam
-from PIL import Image
+from torch import Tensor
+from PIL.Image import Image
 
 
-from data import FlickrDataset, FlickrDataloader
-from .encoder import ImageEncoder
-from .decoder import CaptionDecoder
+from .checkpoint import CheckpointModule
+from .encoder import ResnetImageEncoder
+from .decoder import SimpleCaptionDecoder
 
 
-class ImageCaption(Module):
-    def __init__(self,
-                 dataset: FlickrDataset,
-                 encoder: ImageEncoder,
-                 decoder: CaptionDecoder,
-                 learning_rate: float,
-                 batch_size: int = 5,
-                 num_workers: int = 2,
-                 loss_fn: CrossEntropyLoss = None,
-                 model_path: str = "imagecaption.state"):
-        super(ImageCaption, self).__init__()
-        self.model_path = model_path
-
-        self.dataset = dataset
-        self.batch_size = batch_size
-        self.num_workers = num_workers
-
+class ImageCaption(CheckpointModule):
+    def __init__(self, encoder: ResnetImageEncoder, decoder: SimpleCaptionDecoder):
+        super().__init__("imagecaption.checkpoint")
         self.encoder = encoder
         self.decoder = decoder
 
-        self.optimizer = Adam(self.parameters(), lr=learning_rate)
-        self.criterion = loss_fn if loss_fn else CrossEntropyLoss(
-            ignore_index=dataset.vocabulary.token_to_id["<PAD>"])
-
-    def forward(self, images: Tensor, captions: Tensor) -> Tensor:
-        features = self.encoder(images)
-        outputs = self.decoder(features, captions)
-        return outputs
-
-    def trainModel(self,
-                   epochs: int = 5,
-                   batch_size: int = None,
-                   num_workers: int = None,
-                   resume: bool = False,
-                   device: str = "cpu"):
-        """" Train the model for a given number of epochs.
-        Args:
-            batch_size: The number of samples in each batch (defaults to the model's batch size).
-            num_workers: The number of subprocesses to use for data loading (defaults to the model's num_workers).
-            epochs: The number of times to iterate over the training dataset.
-            resume: Whether to resume training from the last checkpoint.
+    def forward(self, images: list[Image]) -> list[str]:
         """
-
-        start_epoch = self._load() if resume else 0
-        if start_epoch == epochs:
[…]

### 2024-10-30 BAI/paper/src/__main__.py
BAI paperr model almost works.

@@ -1,60 +1,48 @@
-from torch import cuda, device
-from torchvision import transforms as T
+from torch import cuda, device, no_grad
+from torch.utils.data import random_split
+from PIL import Image
 
-from data import LoadedFlickrDataset
-from models import ImageCaption, ResnetImageEncoder, CaptionDecoder
-from eval import Visualiser
+from data import FlickrDataset, Vocabulary, loader
+from models import ImageCaption, ResnetImageEncoder, GRUCaptionDecoder
+from eval import train, evaluate, visualise
 
-DIR = "/content/drive/MyDrive"
-# DIR = "."
+# DIR = "/content/drive/MyDrive"
+DIR = "."
 DEVICE = device("cuda" if cuda.is_available() else "cpu")
 
-# --- Dataset ---
-transforms = T.Compose([
-    T.Resize(226),
-    T.RandomCrop(224),
-    T.ToTensor(),
-    T.Normalize((0.485, 0.456, 0.406), (0.229, 0.224, 0.225))
-])
-dataset = LoadedFlickrDataset(
-    f"{DIR}/flickr8k",
-    num_captions=None,
-    image_transform=transforms
-)
+# --- Data ---
+dataset = FlickrDataset(path=f"{DIR}/flickr8k",
+                        num_captions=100)
+vocabulary = Vocabulary(dataset.captions,
+                        threshold=5)
+train_dataset, eval_dataset = random_split(dataset, [.8, .2])
+
 
 # --- Model ---
-encoder = ResnetImageEncoder(
-).to(DEVICE)
-decoder = CaptionDecoder(
-    embed_size=300,
-    vocab_size=len(dataset.vocabulary),
-    attention_dim=256,
-    encoder_dim=2048,
-    decoder_dim=512
-).to(DEVICE)
-model = ImageCaption(
-    dataset=dataset,
-    encoder=encoder,
-    decoder=decoder,
-    batch_size=5,
-    num_workers=2,
-    learning_rate=3e-4
-).to(DEVICE)
+encoder = ResnetImageEncoder()
+decoder = GRUCaptionDecoder(vocabulary_size=len(vocabulary))
+model = ImageCaption(encoder=encoder, decoder=decoder)
 
-model.trainModel(epochs=2, resume=True, device=DEVICE)
-model.validateModel(device=DEVICE)
+
+# --- Training ---
+
+train(model,
+      data=loader(train_dataset),
+      epochs=10,
[…]

### 2024-10-30 BAI/paper/src/data/__init__.py

@@ -1,3 +1,3 @@
-from .flickrdataset import FlickrDataset, LoadedFlickrDataset
+from .flickrdataset import FlickrDataset
 from .vocabulary import Vocabulary
-from .flickrdataloader import FlickrDataloader
+from .loader import loader

### 2024-10-30 BAI/paper/src/data/flickrdataloader.py

@@ -1,48 +1,0 @@
-
-from torch.utils.data import DataLoader
-from torch.nn.utils.rnn import pad_sequence
-from torch import cat, Tensor
-
-from .flickrdataset import FlickrDataset
-
-
-class FlickrDataloader(DataLoader):
-    """ Load Flickr dataset batches. """
-
-    def __init__(self, dataset: FlickrDataset, batch_size: int = 5, num_workers: int = 2, shuffle: bool = True):
-        super().__init__(
-            dataset=dataset,
-            batch_size=batch_size,
-            num_workers=num_workers,
-            shuffle=shuffle,
-            collate_fn=FlickrDataloader.BatchPadding(pad_token_index=dataset.vocabulary.token_to_id["<PAD>"],
-                                                     batch_dimension_first=True)
-        )
-
-    class BatchPadding:
-        """ Apply padding to  """
-
-        def __init__(self, pad_token_index: int, batch_dimension_first: bool = False):
-            """
-            Args:
-                pad_token_index: The index of the padding token in the vocabulary.
-                batch_dimension_first: Whether the batch dimension is the first dimension.
-            """
-            self.pad_index = pad_token_index
-            self.batch_first = batch_dimension_first
-
-        def __call__(self, batch: list[tuple[Tensor, Tensor]]) -> tuple[Tensor, Tensor]:
-            """ Combine images and captions of a batch into a single tensor each and pad the captions.
-
-            Args:
-                batch: The batch of caption and image tensors.
-            """
-            images = [item[0].unsqueeze(0) for item in batch]
-            images = cat(images, dim=0)
-
-            captions = [item[1] for item in batch]
-            captions = pad_sequence(captions,
-                                    batch_first=self.batch_first,
-                                    padding_value=self.pad_index)
-
-            return images, captions

### 2024-10-30 BAI/paper/src/data/flickrdataset.py

@@ -2,8 +2,9 @@ import os
 from torch.utils.data import Dataset, random_split
 from PIL import Image
 from pandas import read_csv
-from torch import tensor, Tensor
-from torchvision.transforms import Compose
+from torch import Tensor
+from collections import Counter
+from torchvision.transforms import Compose, Resize, ToTensor, Normalize
 
 from .vocabulary import Vocabulary
 
@@ -12,64 +13,45 @@ class FlickrDataset(Dataset):
     """ Flickr dataset for image captioning. """
 
     def __init__(self,
-                 image_path: str,
-                 images: list[str],
-                 captions: list[str]):
-        """
-        Args:
-            image_path (str): Path to the folder containing the images.
-            images (list[str]): List of image filenames.
-            captions (list[str]): List of captions for the images.
-            vocabulary (Vocabulary): Vocabulary object for the captions.
-            transform_images (Compose): Transform to apply to images.
-        """
-        self.image_path = image_path
-        self.images = images
-        self.captions = captions
-
-    def __len__(self) -> int: return len(self.captions)
-
-    def __getitem__(self, index: int) -> tuple[Image.Image, str]:
-        return (
-            Image.open(os.path.join(self.image_path,
-                       "Images", self.images[index])),
-            f"<SOS> {self.captions[index]} <EOS>"
-        )
-
-    def split(self, train_size: float = 0.8) -> tuple["FlickrDataset", "FlickrDataset"]:
-        """ Split the dataset into a training and validation set."""
-
-        subsets = random_split(self, [train_size, 1 - train_size])
-        return (FlickrDataset(self.image_path,
-                              [self.images[i] for i in subsets[0].indices],
-                              [self.captions[i] for i in subsets[0].indices]),
-                FlickrDataset(self.image_path,
-                              [self.images[i] for i in subsets[1].indices],
-                              [self.captions[i] for i in subsets[1].indices]))
-
-
-class LoadedFlickrDataset(FlickrDataset):
-    """ Flickr dataset for image captioning with captions loaded from file. """
-
-    def __init__(self,
                  path: str,
+                 images_folder: str = "Images",
                  captions_file: str = "captions.csv",
-                 num_captions: int = None):
+                 num_captions: int = None,
+                 image_transform: Compose = Compose([
+                     Resize((224, 224)),
+                     ToTensor(),
+                     Normalize(mean=[0.485, 0.456, 0.406],
[…]

### 2024-10-30 BAI/paper/src/data/loader.py

@@ -1,0 +1,24 @@
+from torch import Tensor, cat
+from torch.nn.utils.rnn import pad_sequence
+from torch.utils.data import DataLoader, Dataset
+
+from .vocabulary import Vocabulary
+
+
+def loader(dataset: Dataset, batch_size: int = 5) -> DataLoader:
+    class CollateAndPadBatch:
+        """ Collate and pad batch of images and captions. """
+
+        def __call__(self, batch: list[tuple[Tensor, Tensor]]) -> tuple[Tensor, Tensor]:
+
+            images = [item[0].unsqueeze(0) for item in batch]
+            images = cat(images, dim=0)
+
+            captions = [item[1] for item in batch]
+            captions = pad_sequence(captions,
+                                    batch_first=True,
+                                    padding_value=Vocabulary.pad_index)
+
+            return images, captions
+
+    return DataLoader(dataset, batch_size=batch_size, num_workers=2, shuffle=True, collate_fn=CollateAndPadBatch())

### 2024-10-30 BAI/paper/src/data/vocabulary.py

@@ -7,6 +7,15 @@ from spacy.cli import download
 class Vocabulary:
     """ Vocabulary object to store the mapping between words and indices. """
 
+    pad_token = "<PAD>"
+    sos_token = "<SOS>"
+    eos_token = "<EOS>"
+    unk_token = "<UNK>"
+    pad_index = 0
+    sos_index = 1
+    eos_index = 2
+    unk_index = 3
+
     def __init__(self, sentences: list[str], threshold: int, spacy_model_name: str = "en_core_web_sm"):
         """ Initialize the vocabulary and learn the words from the sentences.
 
@@ -16,33 +25,36 @@ class Vocabulary:
             spacy_model_name: The name of the spaCy model to use for tokenization.
         """
 
-        self.id_to_token = {0: "<PAD>", 1: "<SOS>", 2: "<EOS>", 3: "<UNK>"}
+        self.index_to_token = {self.pad_index: self.pad_token,
+                               self.sos_index: self.sos_token,
+                               self.eos_index: self.eos_token,
+                               self.unk_index: self.unk_token}
         self.threshold = threshold
         self.spacy_model = self._download_and_init(spacy_model_name)
 
-        self._learn(sentences)
-
-    def __len__(self) -> int: return len(self.id_to_token)
-
-    @property
-    def token_to_id(self) -> dict[str, int]:
-        return {v: k for k, v in self.id_to_token.items()}
-
-    def numericalize(self, text: str) -> list[int]:
-        """ For each token in the text, return the index in the vocabulary. """
-        token_to_id = self.token_to_id
-        return [token_to_id[token] if token in token_to_id else token_to_id["<UNK>"]
-                for token in self._tokenize(text)]
-
-    def _learn(self, sentences: list[str]):
-        """ Build the vocabulary from the sentences. """
         freq = Counter()
         for sentence in sentences:
             for token in self._tokenize(sentence):
                 freq[token] += 1
 
-                if token not in self.id_to_token.values():
-                    self.id_to_token[len(self.id_to_token)] = token
+                if token not in self.index_to_token.values():
+                    self.index_to_token[len(self.index_to_token)] = token
+
+    def __len__(self) -> int: return len(self.index_to_token)
+
+    @property
+    def token_to_index(self) -> dict[str, int]:
+        return {v: k for k, v in self.index_to_token.items()}
+
+    def numericalize(self, text: str) -> list[int]:
+        """ Convert the text to numerical form using the vocabulary. """
[…]

### 2024-10-30 BAI/paper/src/eval/__init__.py

@@ -1,3 +1,3 @@
-from train import train
-from evaluate import evaluate
-from visualiser import Visualiser
+from .train import train
+from .evaluate import evaluate
+from .visualise import visualise

### 2024-10-30 BAI/paper/src/eval/evaluate.py

@@ -1,52 +1,52 @@
-from torch.nn import Module
-from nltk.translate import bleu_score
+from nltk.translate.bleu_score import corpus_bleu
+from torch import no_grad, device, Tensor
+from torch.utils.data import DataLoader
+from typing import Callable
 
-from data import FlickrDataset
+from models import ImageCaption
 
 
-def evaluate(model: Module, data: FlickrDataset):
+def evaluate(model: ImageCaption,
+             data: DataLoader,
+             decode_caption: Callable[[Tensor], str],
+             device: device = device("cpu")) -> float:
+    """
+    Evaluate the ImageCaption model on the provided dataset using BLEU score.
+
+    Args:
+        model (ImageCaption): The image captioning model to evaluate.
+        dataloader (DataLoader): DataLoader providing images and reference captions.
+        device (torch.device): Device to run evaluation on (default: CPU).
+
+    Returns:
+        float: The BLEU score for the evaluated dataset.
+    """
+    model.eval()
+    model.to(device)
+
     references = []
     candidates = []
-    for i in range(len(data)):
-        reference = data[i][1]
-        candidate = model.predict(data[i][0])
-        references.append(reference)
-        candidates.append(candidate)
-    return bleu_score.corpus_bleu(references, candidates)
 
-# def validateModel(self,
-#                       batch_size: int = None,
-#                       num_workers: int = None,
-#                       device: str = "cpu") -> float:
-#         """ Validate the model on the validation dataset.
-#         Args:
-#             batch_size: The number of samples in each batch (defaults to the model's batch size).
-#             num_workers: The number of subprocesses to use for data loading (defaults to the model's num_workers).
-#         Returns:
-#             The average loss on the validation dataset.
-#         """
+    with no_grad():
+        for batch, (images, captions) in enumerate(data):
+            images: Tensor = images.to(device)
+            predictions: Tensor = model(images)
 
-#         loader = self._validation_loader(
-#             batch_size=batch_size, num_workers=num_workers)
+            # Prepare reference and candidate sentences for BLEU scoring
+            for i in range(len(predictions)):
+                # Reference: Each reference caption split into tokens (wrapped for BLEU compatibility)
+                reference_caption = [decode_caption(captions[i])]
+                references.append(reference_caption)
 
-#         total_loss = 0
+                # Candidate: Predicted caption as a token list
[…]

### 2024-10-30 BAI/paper/src/eval/train.py

@@ -1,59 +1,87 @@
+from torch import device, Tensor
+from torch.optim import Adam, Optimizer
+from torch.nn import CrossEntropyLoss
+from torch.utils.data import DataLoader
+
+from models import ImageCaption
+from data import Vocabulary
 
 
-#         self.optimizer = Adam(self.parameters(), lr=learning_rate)
-#         self.criterion = loss_fn if loss_fn else CrossEntropyLoss(
-#             ignore_index=dataset.vocabulary.token_to_id["<PAD>"])
-# def trainModel(self,
-#                 epochs: int = 5,
-#                 batch_size: int = None,
-#                 num_workers: int = None,
-#                 resume: bool = False,
-#                 device: str = "cpu"):
-#         """" Train the model for a given number of epochs.
-#         Args:
-#             batch_size: The number of samples in each batch (defaults to the model's batch size).
-#             num_workers: The number of subprocesses to use for data loading (defaults to the model's num_workers).
-#             epochs: The number of times to iterate over the training dataset.
-#             resume: Whether to resume training from the last checkpoint.
-#         """
+def train(model: ImageCaption,
+          data: DataLoader,
+          epochs: int = 5,
+          device: device = device("cpu"),
+          optimizer: Optimizer = None,
+          criterion: CrossEntropyLoss = None):
+    """
+    Train the ImageCaption model.
 
-#         start_epoch = self._load() if resume else 0
-#         if start_epoch == epochs:
-#             print("[Training] Model already trained.")
-#             return
-#         elif start_epoch > epochs:
-#             raise ValueError(
-#                 f"Model has already been trained for {start_epoch} epochs.")
-#         elif start_epoch > 0:
-#             print(f"[Training] Resuming from epoch {start_epoch}.")
+    Args:
+        model (ImageCaption): The caption generation model to train.
+        data (DataLoader): Dataloader providing the training data.
+        epochs (int): Number of epochs to train for (default: 5).
+        device (device): Device to use for training (default: CPU).
+        optimizer (Optimizer): Optimizer for training (default: Adam).
+        criterion (CrossEntropyLoss): Loss function (default: CrossEntropyLoss).
 
-#         loader = self._training_loader(batch_size, num_workers)
+    Returns:
+        ImageCaption: The trained model with the latest checkpoint.
+    """
+    model.to(device)
+    model.train()
 
[…]

### 2024-10-30 BAI/paper/src/eval/visualise.py

@@ -1,0 +1,22 @@
+from PIL.Image import Image
+from matplotlib import pyplot as plt
+from matplotlib import rcParams
+
+
+def visualise(image: Image, caption: str):
+    """
+    Display an image and its caption.
+
+    Args:
+        image (Image): The image to display.
+        caption (str): The caption to display.
+    """
+    assert image.mode == "RGB", "Image should be in RGB format."
+
+    # Set figure size for better visualization
+    rcParams["figure.figsize"] = (8, 6)
+
+    plt.imshow(image)
+    plt.title(caption, wrap=True)
+    plt.axis("off")
+    plt.show()

### 2024-10-30 BAI/paper/src/eval/visualiser.py

@@ -1,25 +1,0 @@
-from torch import Tensor
-from matplotlib import pyplot as plt
-
-
-class Visualiser:
-    def __init__(self):
-        plt.ion()
-
-    def display_image(self, tensor: Tensor, title: str):
-        """ Show image with caption and pause so that plots are updated. """
-        plt.imshow(self._unnormalize(tensor))
-        if title is not None:
-            plt.title(title)
-        plt.pause(.001)
-
-    def _unnormalize(self, tensor: Tensor):
-        tensor = tensor.clone()  # Avoid modifying the original tensor in-place
-
-        # Apply the un-normalization (assuming ImageNet normalization was applied)
-        tensor[0] = tensor[0] * 0.229 + 0.485
-        tensor[1] = tensor[1] * 0.224 + 0.456
-        tensor[2] = tensor[2] * 0.225 + 0.406
-
-        # Convert from [C, H, W] to [H, W, C] for displaying with matplotlib
-        return tensor.permute(1, 2, 0).numpy()

### 2024-10-30 BAI/paper/src/models/__init__.py

@@ -1,3 +1,3 @@
 from .image_caption import ImageCaption
 from .encoder import ResnetImageEncoder
-from .decoder import CaptionDecoder
+from .decoder import GRUCaptionDecoder

### 2024-10-30 BAI/paper/src/models/attention.py

@@ -1,43 +1,0 @@
-from torch.nn import Module, Linear
-from torch import Tensor, tanh
-from torch.nn.functional import softmax
-
-
-class Attention(Module):
-    """ Calculate the attention weights (Bahdanau). """
-
-    def __init__(self,
-                 encoder_dim: int,
-                 decoder_dim: int,
-                 attention_dim: int):
-        super(Attention, self).__init__()
-
-        self.attention_dim = attention_dim
-        self.DecoderToAttention = Linear(decoder_dim, attention_dim)
-        self.EncoderToAttention = Linear(encoder_dim, attention_dim)
-        self.AttentionScore = Linear(attention_dim, 1)
-
-    def forward(self, features: Tensor, hidden: Tensor) -> tuple[Tensor, Tensor]:
-        # (batch_size,num_layers,attention_dim)
-        encoded_features: Tensor = self.EncoderToAttention(features)
-        # (batch_size,attention_dim)
-        hidden_state_weight: Tensor = self.DecoderToAttention(hidden)
-
-        # (batch_size,num_layers,attemtion_dim)
-        combined_states = tanh(
-            encoded_features + hidden_state_weight.unsqueeze(1))
-
-        # (batch_size,num_layers,1)
-        attention_scores: Tensor = self.AttentionScore(combined_states)
-        # (batch_size,num_layers)
-        attention_scores = attention_scores.squeeze(2)
-
-        # (batch_size,num_layers)
-        alpha = softmax(attention_scores, dim=1)
-
-        # (batch_size,num_layers,features_dim)
-        context = features * alpha.unsqueeze(2)
-        # (batch_size,num_layers)
-        context = context.sum(dim=1)
-
-        return alpha, context

### 2024-10-30 BAI/paper/src/models/checkpoint.py

@@ -6,7 +6,7 @@ class CheckpointModule(Module):
     """ A module that can save and load checkpoints. """
 
     def __init__(self, path: str):
-        super(self).__init__()
+        super().__init__()
         self.path = path
 
     def save_checkpoint(self, checkpoint: int):

### 2024-10-30 BAI/paper/src/models/decoder.py

@@ -1,83 +1,72 @@
-from torch import Tensor, tensor
+from torch import Tensor, tensor, full
 from torch.nn import Module, Embedding, GRUCell, Linear
+from torch import float32
+
 from data import Vocabulary
 
 
-class SimpleCaptionDecoder(Module):
+class GRUCaptionDecoder(Module):
     """A simplified GRU-based language model for generating image captions."""
 
-    def __init__(self, vocab: Vocabulary, embed_size: int = 256, decoder_dim: int = 512):
+    def __init__(self,
+                 vocabulary_size: int,
+                 embedding_size: int = 256,
+                 hidden_size: int = 512):
         """
         Args:
-            vocab (Vocabulary): Vocabulary instance for token-to-id mappings.
-            encoder_dim (int): Dimension of the encoder output features.
-            embed_size (int): Size of the word embedding vectors (default: 256).
-            decoder_dim (int): Dimension of the GRU hidden state (default: 512).
+            vocabulary_size: The number of tokens in the vocabulary.
+            embedding_size: The size of the token embeddings.
+            hidden_size: The size of the hidden state of the GRU.
         """
         super().__init__()
+        self.embedding_layer = Embedding(vocabulary_size, embedding_size)
+        self.feature_projection = Linear(2048, hidden_size)
+        self.gru_cell = GRUCell(embedding_size, hidden_size)
+        self.output_layer = Linear(hidden_size, vocabulary_size)
 
-        self.vocab = vocab
-        self.vocab_size = len(vocab)
-        self.embed_size = embed_size
-        self.decoder_dim = decoder_dim
-
-        # Layers for embedding, GRU, and output projection
-        self.embedding_layer = Embedding(self.vocab_size, self.embed_size)
-        self.gru_cell = GRUCell(self.embed_size, self.decoder_dim)
-        self.output_layer = Linear(self.decoder_dim, self.vocab_size)
-
-    def forward(self, features: Tensor, max_len: int = 20) -> list[str]:
-        """
-        Generate captions for a batch of image features using greedy decoding.
+    def forward(self, features: Tensor, max_len: int) -> Tensor:
+        """ Generate captions for the given image features.
 
         Args:
-            features (Tensor): Encoded image features of shape (batch_size, encoder_dim).
-            max_len (int): Maximum length for generated captions (default: 20).
+            features: Image features extracted from the encoder of shape (batch_size, num_features, feature_size).
+            max_len: Maximum length of the generated captions.
 
         Returns:
-            list[str]: List of generated captions for each image in the batch.
[…]

### 2024-10-30 BAI/paper/src/models/encoder.py

@@ -1,8 +1,6 @@
-from torch import Tensor, stack
+from torch import Tensor
 from torch.nn import Module, Sequential
 from torchvision.models import resnet50, ResNet
-from PIL.Image import Image
-from torchvision.transforms import Compose, Resize, ToTensor, Normalize
 
 
 class ResnetImageEncoder(Module):
@@ -13,30 +11,31 @@ class ResnetImageEncoder(Module):
         self.dimension = 2048
         self.model = self._get_passive_resnet()
 
-    def forward(self, images: list[Image]) -> Tensor:
-        """ Extract features from a batch of images.
+    def forward(self, images: Tensor) -> Tensor:
+        """ Extract features from a batch of image tensors.
 
         Args:
-            images (list[Image]): List of PIL images.
+            images (Tensor): Batch of image tensors of shape (batch_size, channels=3, height=224, width=224).
 
         Returns:
-            Tensor: Encoded features with shape (batch_size, 49, 2048).
+            Tensor: Encoded image features of shape (batch_size, num_features=49, feature_size=2048).
         """
-        images = self.preprocess_images(images)  # (batch_size, 3, 224, 224)
+        assert images.dim() == 4, "Input images must be 4-dimensional."
+        assert images.size(1) == 3, "Input images must have 3 channels."
+        assert images.size(2) == 224, "Images must have size 224x224."
+
         features: Tensor = self.model(images)  # (batch_size, 2048, 7, 7)
         features = features.permute(0, 2, 3, 1)  # (batch_size, 7, 7, 2048)
-        features = features.view(
-            features.size(0), -1, features.size(-1))  # (batch_size, 49, 2048)
-        return features
+        features = features.view(features.size(0), -1, features.size(-1))
 
-    @staticmethod
-    def preprocess_images(images: list[Image]) -> Tensor:
-        transform = Compose([
-            Resize((224, 224)),  # ResNet expects images of 224x224
-            ToTensor(),  # Converts PIL image to a tensor with values [0, 1]
-            Normalize(mean=[0.485, 0.456, 0.406], std=[0.229, 0.224, 0.225])
-        ])
-        return stack([transform(image.convert("RGB")) for image in images])
+        assert features.dim() == 3, "Output must have 3 dimensions."
+        assert features.size(0) == images.size(0), \
+            "Output batch size must match input batch size."
+        assert features.size(1) == 49, "Output must have 49 features."
+        assert features.size(2) == self.dimension, \
[…]

### 2024-10-30 BAI/paper/src/models/image_caption.py

@@ -1,28 +1,37 @@
-from torch import Tensor
-from PIL.Image import Image
-
+from torch import Tensor, no_grad
 
 from .checkpoint import CheckpointModule
 from .encoder import ResnetImageEncoder
-from .decoder import SimpleCaptionDecoder
+from .decoder import GRUCaptionDecoder
 
 
 class ImageCaption(CheckpointModule):
-    def __init__(self, encoder: ResnetImageEncoder, decoder: SimpleCaptionDecoder):
+    def __init__(self, encoder: ResnetImageEncoder, decoder: GRUCaptionDecoder):
         super().__init__("imagecaption.checkpoint")
         self.encoder = encoder
         self.decoder = decoder
 
-    def forward(self, images: list[Image]) -> list[str]:
+    def forward(self, images: Tensor, max_len: int = 15) -> Tensor:
         """
         Generate captions for a batch of images.
 
         Args:
-            images (list[Image]): A batch of PIL images to be captioned.
+            images: A tensor of shape (batch_size, channels=3, height=224, width=224) containing the images.
+            max_len: The maximum length of the captions to generate.
 
         Returns:
-            list[str]: Generated captions for each image in the batch.
+            A tensor of shape (batch_size, max_len, vocabulary_size) containing the logits for each token position.
         """
-        features = self.encoder(images)  # (batch_size, 49, encoder_dim)
-        captions = self.decoder(features)  # (batch_size, max_len)
-        return captions
+        assert images.dim() == 4, "Images must have 4 dimensions."
+        assert images.size(1) == 3, "Images must have 3 channels."
+        assert images.size(2) == 224 and images.size(
+            3) == 224, "Images must have size 224x224."
+
+        features = self.encoder(images)
+        logits: Tensor = self.decoder(features, max_len)
+
+        assert logits.dim() == 3, "Logits must have 3 dimensions (batch_size, max_len, vocabulary_size)."
+        assert logits.size(0) == images.size(0), "Batch size must match."
+        assert logits.size(1) == max_len, "Caption length must match."
+
+        return logits

### 2024-10-30 PbD/summary.tex
Added basic layout  for pbd summary.

\documentclass{article}

\usepackage{../.template/summary}

\subject{Privacy by Design}
\semester{Winter 2024}
\author{Leopold Lemmermann}

\usepackage{pdfpages}

\begin{document}\createtitle

\part{Privacy Enhancing Technology}
% \includepdf[pages={?}, nup=?]{slides.pdf}

\section{IPv6 traffic pseudonymisation}

\section{In Practice}

\section{Censorship Resistance}

\section{Criminal Misuse and Data Retention}

\part{Location Privacy}

\end{document}

### 2024-10-30 PbD/summary.tex
Added some basic content for PbD summary.

@@
 \part{Privacy Enhancing Technology}
-% \includepdf[pages={?}, nup=?]{slides.pdf}
 
+\section{Confidentiality of Metadata}
+Goal of privacy.
+- Anonymity: Protection of the identity of the sender or receiver of a message.
+- Unobservability: Protection of the communication relation itself.
+
+\section{Attacker model}
+- Outsider: eavesdropping on communcation lines, traffic analysis
+- Insider: network operators, malicious staff, government
+can:
+- passively observe all communication links
+- actively send messages
+- flood or block messages
+- operate (all but one) anonymity services
+- operate server
+cannot:
+- break cryptographic systems
+- attack user's personal machine
+- have unlimited time or resources
+
+\section{Tracking on OSI Application Layer}
+Application:
+- filtering mechanisms: cookies, active contents
+- e2e encryption
+- hiding: pseudonyms, credentials
+
+Pseudonyms
+- Passive Trustee: trusted third party (TTP) that certifies pseudonyms
+- Active Trustee: TTP is involved in communication
+
+\section{Tracking on OSI Network Layer}
+- Broadcast with implicit addressing: all recipients receive all messages, local filtering -> protects recipient
+- Dummy Traffic: send meaningless messages -> protects sender
+- Proxies: forward messages -> protects sender, but proxy can track
+- DC network: broadcast, dummy and cryptography -> protects sender
+- Blind-message service: unobservable access to replicated databases -> protects database client
+- Mix network: cascaded (independent) proxies, dummy traffic, cryptography -> protects communication relation
+- Steganography: hidden messages in other messages -> protects content
+
+\section{Traffic Analysis}
+Attacks
+- Timing: correlation of messages
+- Volume: correlation of message sizes
+- Flooding: fill up anonymous group batch to separate certain message
+- Intersection: correlation of messages in different groups
+
+% \includepdf[pages={14-29}, nup=3x3]{slides.pdf}
 \section{IPv6 traffic pseudonymisation}
 
+% \includepdf[pages={?}, nup=?]{slides.pdf}
 \section{In Practice}
 
+% \includepdf[pages={?}, nup=?]{slides.pdf}
 \section{Censorship Resistance}
 
+% \includepdf[pages={?}, nup=?]{slides.pdf}
 \section{Criminal Misuse and Data Retention}

### 2024-10-30 BAI/paper/src/__main__.py
Small improvements to bai model.

@@ -6,21 +6,18 @@ from data import FlickrDataset, Vocabulary, loader
 from models import ImageCaption, ResnetImageEncoder, GRUCaptionDecoder
 from eval import train, evaluate, visualise
 
-# DIR = "/content/drive/MyDrive"
-DIR = "."
+DIR = "/content/drive/MyDrive"
 DEVICE = device("cuda" if cuda.is_available() else "cpu")
 
 # --- Data ---
 dataset = FlickrDataset(path=f"{DIR}/flickr8k",
                         num_captions=100)
-vocabulary = Vocabulary(dataset.captions,
-                        threshold=5)
 train_dataset, eval_dataset = random_split(dataset, [.8, .2])
 
 
 # --- Model ---
 encoder = ResnetImageEncoder()
-decoder = GRUCaptionDecoder(vocabulary_size=len(vocabulary))
+decoder = GRUCaptionDecoder(vocabulary_size=len(dataset.vocabulary))
 model = ImageCaption(encoder=encoder, decoder=decoder)
 
 
@@ -35,14 +32,14 @@ train(model,
 # --- Evaluation ---
 evaluate(model,
          data=loader(eval_dataset),
-         decode_caption=dataset.tensor_to_caption,
+         dataset=dataset,
          device=DEVICE)
 
 
 # --- Inference ---
 with no_grad():
     model.eval()
-    image = dataset.image_to_tensor(
-        Image.open(f"{DIR}/image.jpg").convert("RGB")).unsqueeze(0).to(DEVICE)
-    caption = dataset.tensor_to_caption(model(image))
+    image = Image.open(f"{DIR}/image.jpg").convert("RGB")
+    caption = dataset.tensor_to_caption(model(
+        dataset.image_to_tensor(image).unsqueeze(0).to(DEVICE)))
 visualise(image, caption)

### 2024-10-30 BAI/paper/src/assemble.sh

@@ -14,6 +14,7 @@ cat $DIR/data/vocabulary.py >> $FILE
 cat $DIR/data/flickrdataset.py >> $FILE
 cat $DIR/data/*.py >> $FILE
 cat $DIR/models/*.py >> $FILE
+cat $DIR/eval/*.py >> $FILE
 cat $DIR/__main__.py >> $FILE
 
 # remove all relative imports

### 2024-10-30 BAI/paper/src/data/vocabulary.py

@@ -33,12 +33,19 @@ class Vocabulary:
         self.spacy_model = self._download_and_init(spacy_model_name)
 
         freq = Counter()
-        for sentence in sentences:
-            for token in self._tokenize(sentence):
+
+        print("[Vocabulary] Learning...")
+        for index, sentence in enumerate(sentences):
+            for token in self.tokenize(sentence):
                 freq[token] += 1
 
                 if token not in self.index_to_token.values():
                     self.index_to_token[len(self.index_to_token)] = token
+
+            if index % 1000 == 0:
+                print(
+                    f"\t[Learn] sentence {index + 1}/{len(sentences)}.")
+        print("[Vocabulary] Learning completed.")
 
     def __len__(self) -> int: return len(self.index_to_token)
 
@@ -50,13 +57,13 @@ class Vocabulary:
         """ Convert the text to numerical form using the vocabulary. """
         return [self.token_to_index[token]
                 if token in self.token_to_index else self.token_to_index["<UNK>"]
-                for token in self._tokenize(text)]
+                for token in self.tokenize(text)]
 
     def denumericalize(self, indices: list[int]) -> str:
         """ Convert the indices to text using the vocabulary. """
         return " ".join([self.index_to_token[index] for index in indices])
 
-    def _tokenize(self, text: str) -> list[str]:
+    def tokenize(self, text: str) -> list[str]:
         """ Tokenize the text using spacy model. """
         return [token.text.lower()
                 for token in self.spacy_model.tokenizer(text)]

### 2024-10-30 BAI/paper/src/eval/evaluate.py

@@ -1,14 +1,14 @@
 from nltk.translate.bleu_score import corpus_bleu
 from torch import no_grad, device, Tensor
 from torch.utils.data import DataLoader
-from typing import Callable
 
 from models import ImageCaption
+from data import FlickrDataset
 
 
 def evaluate(model: ImageCaption,
              data: DataLoader,
-             decode_caption: Callable[[Tensor], str],
+             dataset: FlickrDataset,
              device: device = device("cpu")) -> float:
     """
     Evaluate the ImageCaption model on the provided dataset using BLEU score.
@@ -28,25 +28,25 @@ def evaluate(model: ImageCaption,
     candidates = []
 
     with no_grad():
+        print("[Evaluation] Evaluating...")
         for batch, (images, captions) in enumerate(data):
             images: Tensor = images.to(device)
             predictions: Tensor = model(images)
 
             # Prepare reference and candidate sentences for BLEU scoring
-            for i in range(len(predictions)):
-                # Reference: Each reference caption split into tokens (wrapped for BLEU compatibility)
-                reference_caption = [decode_caption(captions[i])]
-                references.append(reference_caption)
-
-                # Candidate: Predicted caption as a token list
-                candidate_caption = decode_caption(predictions[i])
-                candidates.append(candidate_caption)
+            for index, prediction in enumerate(predictions):
+                references.append(dataset.vocabulary
+                                  .tokenize(dataset.tensor_to_caption(captions[index])))
+                candidates.append(dataset.vocabulary
+                                  .tokenize(dataset.tensor_to_caption(prediction.argmax(dim=1))))
 
             # Calculate batch-level BLEU score for tracking
-            batch_bleu = corpus_bleu(references, candidates)
+            batch_bleu = corpus_bleu(references,
+                                     candidates,
+                                     weights=(0.25, 0.25, 0.25, 0.25))
             print(
-                f"[Evaluation] Batch {batch + 1}/{len(data)} - BLEU Score: {batch_bleu:.4f}")
+                f"\t[Evaluation] Batch {batch + 1}/{len(data)} - BLEU Score: {batch_bleu:.4f}")
 
     bleu = corpus_bleu(references, candidates)
-    print(f"[Evaluation] Final BLEU Score: {bleu:.4f}")
+    print(f"[Evaluation] Evaluation compelted. Final BLEU Score: {bleu:.4f}")
     return bleu

### 2024-10-30 BAI/paper/src/eval/train.py

@@ -37,7 +37,7 @@ def train(model: ImageCaption,
 
     checkpoint = model.load_checkpoint()
 
-    print("[Training] Started.")
+    print("[Training] Starting...")
     for epoch in range(checkpoint, epochs):
         print(f"[Epoch {epoch + 1}/{epochs}]")
         epoch_loss = 0
@@ -83,5 +83,5 @@ def train(model: ImageCaption,
 
         model.save_checkpoint(epoch + 1)
 
-    print("[Training] Completed.")
+    print("[Training] Training completed.")
     return model

### 2024-10-30 BAI/paper/src/__main__.py
Using progress bars for bai project.

@@ -2,16 +2,21 @@ from torch import cuda, device, no_grad
 from torch.utils.data import random_split
 from PIL import Image
 
-from data import FlickrDataset, Vocabulary, loader
+from data import FlickrDataset, loader
 from models import ImageCaption, ResnetImageEncoder, GRUCaptionDecoder
 from eval import train, evaluate, visualise
 
-DIR = "/content/drive/MyDrive"
+# DIR = "/content/drive/MyDrive"
+# CAPTION_LIMIT = None
+DIR = "."
+CAPTION_LIMIT = 1000
 DEVICE = device("cuda" if cuda.is_available() else "cpu")
+BATCH_SIZE = 800 if cuda.is_available() else 5
+NUM_WORKERS = 2 if cuda.is_available() else 0
+PIN_MEMORY = cuda.is_available()
 
 # --- Data ---
-dataset = FlickrDataset(path=f"{DIR}/flickr8k",
-                        num_captions=100)
+dataset = FlickrDataset(path=f"{DIR}/flickr8k", num_captions=CAPTION_LIMIT)
 train_dataset, eval_dataset = random_split(dataset, [.8, .2])
 
 
@@ -24,14 +29,20 @@ model = ImageCaption(encoder=encoder, decoder=decoder)
 # --- Training ---
 
 train(model,
-      data=loader(train_dataset),
-      epochs=10,
+      data=loader(train_dataset,
+                  batch_size=BATCH_SIZE,
+                  num_workers=NUM_WORKERS,
+                  pin_memory=PIN_MEMORY),
+      epochs=5,
       device=DEVICE)
 
 
 # --- Evaluation ---
 evaluate(model,
-         data=loader(eval_dataset),
+         data=loader(eval_dataset,
+                     batch_size=BATCH_SIZE,
+                     num_workers=NUM_WORKERS,
+                     pin_memory=PIN_MEMORY),
          dataset=dataset,
          device=DEVICE)
 

### 2024-10-30 BAI/paper/src/data/flickrdataset.py

@@ -1,57 +1,89 @@
 import os
-from torch.utils.data import Dataset, random_split
-from PIL import Image
-from pandas import read_csv
-from torch import Tensor
-from collections import Counter
+from torch import tensor, Tensor
+from torch.utils.data import Dataset
 from torchvision.transforms import Compose, Resize, ToTensor, Normalize
+from PIL import Image
+import pickle
+from multiprocessing import Pool
+from tqdm import tqdm
+import pandas as pd
 
 from .vocabulary import Vocabulary
 
 
 class FlickrDataset(Dataset):
-    """ Flickr dataset for image captioning. """
-
     def __init__(self,
                  path: str,
                  images_folder: str = "Images",
                  captions_file: str = "captions.csv",
                  num_captions: int = None,
-                 image_transform: Compose = Compose([
-                     Resize((224, 224)),
-                     ToTensor(),
-                     Normalize(mean=[0.485, 0.456, 0.406],
-                               std=[0.229, 0.224, 0.225])
-                 ]),
+                 image_transform: Compose = None,
+                 data_file: str = "data.pkl",
                  vocabulary_threshold: int = 5):
-        """
-        Args:
-            path (str): Path to the dataset folder.
-            captions_file (str): Name of the file containing the captions.
-            num_images (int): Number of captions to load.
-        """
         self.image_path = os.path.join(path, images_folder)
-        images_with_captions = read_csv(
-            os.path.join(path, captions_file), nrows=num_captions)
-        self.captions = images_with_captions["caption"].tolist()
-        self.images = images_with_captions["image"].tolist()
-        self.image_transform = image_transform
-        self.vocabulary = Vocabulary(
-            self.captions, threshold=vocabulary_threshold)
+        self.captions_file = os.path.join(path, captions_file)
+        self.data_path = os.path.join(path, data_file)
 
-    def __len__(self) -> int: return len(self.captions)
+        self.image_transform = image_transform or Compose([
+            Resize((224, 224)),
+            ToTensor(),
+            Normalize(mean=[0.485, 0.456, 0.406], std=[0.229, 0.224, 0.225])
+        ])
 
-    def __getitem__(self, index: int) -> tuple[Tensor, Tensor]:
-        image = self.image_to_tensor(
-            Image.open(os.path.join(self.image_path, self.images[index])))
-        caption = self.caption_to_tensor(self.captions[index])
-        return image, caption
[…]

### 2024-10-30 BAI/paper/src/data/loader.py

@@ -5,12 +5,11 @@ from torch.utils.data import DataLoader, Dataset
 from .vocabulary import Vocabulary
 
 
-def loader(dataset: Dataset, batch_size: int = 5) -> DataLoader:
+def loader(dataset: Dataset, batch_size: int = 5, num_workers: int = 2, pin_memory: bool = False) -> DataLoader:
     class CollateAndPadBatch:
         """ Collate and pad batch of images and captions. """
 
         def __call__(self, batch: list[tuple[Tensor, Tensor]]) -> tuple[Tensor, Tensor]:
-
             images = [item[0].unsqueeze(0) for item in batch]
             images = cat(images, dim=0)
 
@@ -21,4 +20,4 @@ def loader(dataset: Dataset, batch_size: int = 5) -> DataLoader:
 
             return images, captions
 
-    return DataLoader(dataset, batch_size=batch_size, num_workers=2, shuffle=True, collate_fn=CollateAndPadBatch())
+    return DataLoader(dataset, batch_size=batch_size, num_workers=num_workers, shuffle=True, pin_memory=pin_memory, collate_fn=CollateAndPadBatch())

### 2024-10-30 BAI/paper/src/data/vocabulary.py

@@ -2,6 +2,7 @@ from collections import Counter
 from importlib import import_module
 from spacy import Language
 from spacy.cli import download
+from tqdm import tqdm
 
 
 class Vocabulary:
@@ -24,7 +25,6 @@ class Vocabulary:
             sentences: A list of sentences to learn the vocabulary from.
             spacy_model_name: The name of the spaCy model to use for tokenization.
         """
-
         self.index_to_token = {self.pad_index: self.pad_token,
                                self.sos_index: self.sos_token,
                                self.eos_index: self.eos_token,
@@ -34,20 +34,16 @@ class Vocabulary:
 
         freq = Counter()
 
-        print("[Vocabulary] Learning...")
-        for index, sentence in enumerate(sentences):
+        # Use tqdm to provide a progress bar for vocabulary creation
+        for sentence in tqdm(sentences, desc="Building Vocabulary"):
             for token in self.tokenize(sentence):
                 freq[token] += 1
 
                 if token not in self.index_to_token.values():
                     self.index_to_token[len(self.index_to_token)] = token
 
-            if index % 1000 == 0:
-                print(
-                    f"\t[Learn] sentence {index + 1}/{len(sentences)}.")
-        print("[Vocabulary] Learning completed.")
-
-    def __len__(self) -> int: return len(self.index_to_token)
+    def __len__(self) -> int:
+        return len(self.index_to_token)
 
     @property
     def token_to_index(self) -> dict[str, int]:
@@ -56,17 +52,16 @@ class Vocabulary:
     def numericalize(self, text: str) -> list[int]:
         """ Convert the text to numerical form using the vocabulary. """
         return [self.token_to_index[token]
-                if token in self.token_to_index else self.token_to_index["<UNK>"]
+                if token in self.token_to_index else self.token_to_index[self.unk_token]
                 for token in self.tokenize(text)]
 
     def denumericalize(self, indices: list[int]) -> str:
-        """ Convert the indices to text using the vocabulary. """
+        """ Convert the indices back to text using the vocabulary. """
         return " ".join([self.index_to_token[index] for index in indices])
 
     def tokenize(self, text: str) -> list[str]:
         """ Tokenize the text using spacy model. """
-        return [token.text.lower()
-                for token in self.spacy_model.tokenizer(text)]
+        return [token.text.lower() for token in self.spacy_model.tokenizer(text)]
 
     def _download_and_init(self, model_name: str) -> Language:
         """ Load a spaCy model, download it if it has not been installed yet.

### 2024-10-30 BAI/paper/src/eval/evaluate.py

@@ -1,6 +1,7 @@
 from nltk.translate.bleu_score import corpus_bleu
 from torch import no_grad, device, Tensor
 from torch.utils.data import DataLoader
+from tqdm import tqdm
 
 from models import ImageCaption
 from data import FlickrDataset
@@ -15,7 +16,7 @@ def evaluate(model: ImageCaption,
 
     Args:
         model (ImageCaption): The image captioning model to evaluate.
-        dataloader (DataLoader): DataLoader providing images and reference captions.
+        data (DataLoader): DataLoader providing images and reference captions.
         device (torch.device): Device to run evaluation on (default: CPU).
 
     Returns:
@@ -24,29 +25,25 @@ def evaluate(model: ImageCaption,
     model.eval()
     model.to(device)
 
-    references = []
-    candidates = []
+    references, candidates = [], []
 
     with no_grad():
-        print("[Evaluation] Evaluating...")
-        for batch, (images, captions) in enumerate(data):
-            images: Tensor = images.to(device)
-            predictions: Tensor = model(images)
+        # Progress bar for evaluation batches
+        with tqdm(total=len(data), desc="Evaluating", unit="batch") as pbar:
+            for images, captions in data:
+                images = images.to(device)
+                predictions = model(images)
 
-            # Prepare reference and candidate sentences for BLEU scoring
-            for index, prediction in enumerate(predictions):
-                references.append(dataset.vocabulary
-                                  .tokenize(dataset.tensor_to_caption(captions[index])))
-                candidates.append(dataset.vocabulary
-                                  .tokenize(dataset.tensor_to_caption(prediction.argmax(dim=1))))
+                # Prepare reference and candidate sentences for BLEU scoring
+                for index, prediction in enumerate(predictions):
+                    references.append([dataset.vocabulary.tokenize(
+                        dataset.tensor_to_caption(captions[index]))])
+                    candidates.append(dataset.vocabulary.tokenize(
+                        dataset.tensor_to_caption(prediction.argmax(dim=1))))
 
-            # Calculate batch-level BLEU score for tracking
-            batch_bleu = corpus_bleu(references,
-                                     candidates,
-                                     weights=(0.25, 0.25, 0.25, 0.25))
-            print(
-                f"\t[Evaluation] Batch {batch + 1}/{len(data)} - BLEU Score: {batch_bleu:.4f}")
+                pbar.update(1)
 
[…]

### 2024-10-30 BAI/paper/src/eval/train.py

@@ -2,6 +2,7 @@ from torch import device, Tensor
 from torch.optim import Adam, Optimizer
 from torch.nn import CrossEntropyLoss
 from torch.utils.data import DataLoader
+from tqdm import tqdm
 
 from models import ImageCaption
 from data import Vocabulary
@@ -37,51 +38,32 @@ def train(model: ImageCaption,
 
     checkpoint = model.load_checkpoint()
 
-    print("[Training] Starting...")
     for epoch in range(checkpoint, epochs):
         print(f"[Epoch {epoch + 1}/{epochs}]")
         epoch_loss = 0
 
-        for batch, (images, captions) in enumerate(data):
-            assert images.size(0) == captions.size(0), \
-                f"Batch size mismatch: {images.size(0)} != {captions.size(0)}"
+        # Progress bar for each batch
+        with tqdm(total=len(data), desc=f"Training Epoch {epoch + 1}/{epochs}", unit="batch") as pbar:
+            for images, captions in data:
+                images, captions = images.to(device), captions.to(device)
+                optimizer.zero_grad()
 
-            images: Tensor = images.to(device)
-            captions: Tensor = captions.to(device)
-            optimizer.zero_grad()
+                # Forward pass
+                outputs = model(images, max_len=captions.size(1))
+                loss = criterion(
+                    outputs.view(-1, outputs.size(-1)), captions.view(-1))
 
-            # Forward pass
-            outputs: Tensor = model(images, max_len=captions.size(1))
+                # Backward pass and optimization
+                loss.backward()
+                optimizer.step()
 
-            assert outputs.size(0) == captions.size(0) and outputs.size(1) == captions.size(1), \
-                f"Output shape mismatch: {outputs.size()} != {captions.size()}"
-
-            # Skip <SOS> token in targets, align outputs and targets, flatten for loss
-            targets = captions[:, 1:].reshape(-1).long()
-            outputs = outputs[:, :-1].reshape(-1, outputs.size(-1)).float()
-
-            assert outputs.requires_grad, "Outputs must require gradients."
-
-            assert outputs.size(0) == targets.size(0), \
-                f"Batch size mismatch: {outputs.size(0)} != {targets.size(0)}"
-
-            # Compute loss
-            assert outputs.dim() == 2 and targets.dim() == 1, \
-                "Output and target shapes are not compatible with CrossEntropyLoss"
-            loss: Tensor = criterion(outputs, targets)
[…]

### 2024-10-30 BAI/paper/src/eval/visualise.py

@@ -13,9 +13,6 @@ def visualise(image: Image, caption: str):
     """
     assert image.mode == "RGB", "Image should be in RGB format."
 
-    # Set figure size for better visualization
-    rcParams["figure.figsize"] = (8, 6)
-
     plt.imshow(image)
     plt.title(caption, wrap=True)
     plt.axis("off")

### 2024-10-31 BAI/paper/src/assemble.sh
Updated dataset image parallelisation.

@@ -6,10 +6,6 @@ if [ -f $FILE ]; then
     rm $FILE
 fi
 
-for dep in $(cat $DIR/requirements.txt); do
-    echo "!pip install $dep" >> $FILE
-done
-
 cat $DIR/data/vocabulary.py >> $FILE
 cat $DIR/data/flickrdataset.py >> $FILE
 cat $DIR/data/*.py >> $FILE

### 2024-10-31 BAI/paper/src/data/flickrdataset.py

@@ -4,7 +4,6 @@ from torch.utils.data import Dataset
 from torchvision.transforms import Compose, Resize, ToTensor, Normalize
 from PIL import Image
 import pickle
-from multiprocessing import Pool
 from tqdm import tqdm
 import pandas as pd
 
@@ -45,7 +44,6 @@ class FlickrDataset(Dataset):
             f"[Dataset] {len(self.images)} images and {len(self.captions)} captions are ready.")
 
     def _process_and_save(self, num_captions, vocabulary_threshold):
-        # Load captions with pandas
         data = pd.read_csv(self.captions_file, nrows=num_captions)
         images, captions = [], []
 
@@ -58,16 +56,16 @@ class FlickrDataset(Dataset):
                     tqdm(raw_captions, desc="Tokenizing Captions")]
 
         # Use multiprocessing to process images in parallel
-        with Pool() as pool:
-            image_paths = [os.path.join(self.image_path, img_name)
-                           for img_name in data['image']]
-            images = list(tqdm(pool.imap(self.process_image, image_paths), total=len(
-                image_paths), desc="Processing Images"))
+        image_paths = [os.path.join(self.image_path, img_name)
+                       for img_name in data['image']]
+        images = list(tqdm(map(self.process_image, image_paths),
+                           total=len(image_paths), desc="Processing Images"))
 
         # Save processed data including the vocabulary and numericalized captions
         with open(self.data_path, 'wb') as f:
-            pickle.dump({'images': images, 'captions': captions,
-                        'vocabulary': vocabulary}, f)
+            pickle.dump({'images': images,
+                         'captions': captions,
+                         'vocabulary': vocabulary}, f)
 
         return images, captions, vocabulary
 

### 2024-10-31 BAI/paper/src/__main__.py
BAI WIP.

@@ -7,16 +7,15 @@ from models import ImageCaption, ResnetImageEncoder, GRUCaptionDecoder
 from eval import train, evaluate, visualise
 
 # DIR = "/content/drive/MyDrive"
-# CAPTION_LIMIT = None
 DIR = "."
-CAPTION_LIMIT = 1000
+CAPTION_LIMIT = None
 DEVICE = device("cuda" if cuda.is_available() else "cpu")
 BATCH_SIZE = 800 if cuda.is_available() else 5
 NUM_WORKERS = 2 if cuda.is_available() else 0
 PIN_MEMORY = cuda.is_available()
 
 # --- Data ---
-dataset = FlickrDataset(path=f"{DIR}/flickr8k", num_captions=CAPTION_LIMIT)
+dataset = FlickrDataset(path=f"{DIR}/flickr8k", caption_limit=CAPTION_LIMIT)
 train_dataset, eval_dataset = random_split(dataset, [.8, .2])
 
 
@@ -50,7 +49,7 @@ evaluate(model,
 # --- Inference ---
 with no_grad():
     model.eval()
-    image = Image.open(f"{DIR}/image.jpg").convert("RGB")
+    image = Image.open(f"{DIR}/flickr8k/image.jpg").convert("RGB")
     caption = dataset.tensor_to_caption(model(
         dataset.image_to_tensor(image).unsqueeze(0).to(DEVICE)))
 visualise(image, caption)

### 2024-10-31 BAI/paper/src/data/flickrdataset.py

@@ -7,78 +7,91 @@ import pickle
 from tqdm import tqdm
 import pandas as pd
 
+
 from .vocabulary import Vocabulary
 
 
 class FlickrDataset(Dataset):
     def __init__(self,
                  path: str,
+                 caption_limit: int = None,
                  images_folder: str = "Images",
                  captions_file: str = "captions.csv",
-                 num_captions: int = None,
                  image_transform: Compose = None,
-                 data_file: str = "data.pkl",
+                 data_file: str = "compressed_data.pkl",
                  vocabulary_threshold: int = 5):
         self.image_path = os.path.join(path, images_folder)
         self.captions_file = os.path.join(path, captions_file)
         self.data_path = os.path.join(path, data_file)
-
+        self.caption_limit = caption_limit
         self.image_transform = image_transform or Compose([
             Resize((224, 224)),
             ToTensor(),
             Normalize(mean=[0.485, 0.456, 0.406], std=[0.229, 0.224, 0.225])
         ])
 
-        if os.path.exists(self.data_path) and num_captions is None:
-            print("[Dataset] Loading preprocessed data...")
-            with open(self.data_path, 'rb') as f:
-                data = pickle.load(f)
-                self.images, self.captions, self.vocabulary = \
-                    data['images'], data['captions'], data['vocabulary']
+        if os.path.exists(self.data_path) and caption_limit is None:
+            print("Loading preprocessed data...")
+            self.images, \
+                self.captions, \
+                self.vocabulary, \
+                self.caption_to_image_index = self._load()
         else:
-            print("[Dataset] Preprocessing and storing data...")
-            self.images, self.captions, self.vocabulary = \
-                self._process_and_save(num_captions, vocabulary_threshold)
+            self.images, \
+                self.captions, \
+                self.vocabulary, \
+                self.caption_to_image_index = self._process(
+                    vocabulary_threshold)
+            print("Saving preprocessed data...")
+            self._save(self.images,
+                       self.captions,
+                       self.vocabulary,
+                       self.caption_to_image_index)
 
-        print(
-            f"[Dataset] {len(self.images)} images and {len(self.captions)} captions are ready.")
-
-    def _process_and_save(self, num_captions, vocabulary_threshold):
-        data = pd.read_csv(self.captions_file, nrows=num_captions)
-        images, captions = [], []
-
[…]

### 2024-10-31 BAI/paper/src/data/vocabulary.py

@@ -35,7 +35,7 @@ class Vocabulary:
         freq = Counter()
 
         # Use tqdm to provide a progress bar for vocabulary creation
-        for sentence in tqdm(sentences, desc="Building Vocabulary"):
+        for sentence in tqdm(sentences, desc="Building Vocabulary", unit="sentences"):
             for token in self.tokenize(sentence):
                 freq[token] += 1
 

### 2024-10-31 BAI/paper/src/eval/evaluate.py

@@ -1,6 +1,6 @@
-from nltk.translate.bleu_score import corpus_bleu
 from torch import no_grad, device, Tensor
 from torch.utils.data import DataLoader
+from nltk.translate.bleu_score import corpus_bleu
 from tqdm import tqdm
 
 from models import ImageCaption
@@ -22,28 +22,21 @@ def evaluate(model: ImageCaption,
     Returns:
         float: The BLEU score for the evaluated dataset.
     """
-    model.eval()
-    model.to(device)
+    model.to(device).eval()
 
     references, candidates = [], []
-
     with no_grad():
-        # Progress bar for evaluation batches
-        with tqdm(total=len(data), desc="Evaluating", unit="batch") as pbar:
-            for images, captions in data:
-                images = images.to(device)
-                predictions = model(images)
+        bleu = 0.0
+        for images, captions in tqdm(data, desc="Evaluation", unit="batch", postfix={"BLEU": bleu}):
+            images: Tensor = images.to(device)
+            predictions: Tensor = model(images)
 
-                # Prepare reference and candidate sentences for BLEU scoring
-                for index, prediction in enumerate(predictions):
-                    references.append([dataset.vocabulary.tokenize(
-                        dataset.tensor_to_caption(captions[index]))])
-                    candidates.append(dataset.vocabulary.tokenize(
-                        dataset.tensor_to_caption(prediction.argmax(dim=1))))
+            for index, prediction in enumerate(predictions):
+                references.append([dataset.vocabulary.tokenize(
+                    dataset.tensor_to_caption(captions[index]))])
+                candidates.append(dataset.vocabulary.tokenize(
+                    dataset.tensor_to_caption(prediction.argmax(dim=1))))
 
-                pbar.update(1)
+            bleu = corpus_bleu(references, candidates)
 
-    # Calculate final BLEU score
-    bleu = corpus_bleu(references, candidates)
-    print(f"[Evaluation] Final BLEU Score: {bleu:.4f}")
     return bleu

### 2024-10-31 BAI/paper/src/eval/train.py

@@ -2,7 +2,7 @@ from torch import device, Tensor
 from torch.optim import Adam, Optimizer
 from torch.nn import CrossEntropyLoss
 from torch.utils.data import DataLoader
-from tqdm import tqdm
+from tqdm import tqdm, trange
 
 from models import ImageCaption
 from data import Vocabulary
@@ -28,42 +28,29 @@ def train(model: ImageCaption,
     Returns:
         ImageCaption: The trained model with the latest checkpoint.
     """
-    model.to(device)
-    model.train()
+    model.to(device).train()
 
     optimizer = Adam(model.parameters(),
                      lr=1e-3) if optimizer is None else optimizer
-    criterion = CrossEntropyLoss(ignore_index=Vocabulary.pad_index).to(
-        device) if criterion is None else criterion
+    criterion = CrossEntropyLoss(ignore_index=Vocabulary.pad_index
+                                 ) if criterion is None else criterion
 
-    checkpoint = model.load_checkpoint()
+    epoch_loss = 0.0
+    for epoch in trange(model.load_checkpoint(), epochs, desc="Training", unit="epoch", postfix={"Loss": 0.0}):
+        batch_loss = 0.0
+        for images, captions in tqdm(data, desc=f"Epoch", unit="batch", postfix={"Loss": batch_loss}):
+            optimizer.zero_grad()
 
-    for epoch in range(checkpoint, epochs):
-        print(f"[Epoch {epoch + 1}/{epochs}]")
-        epoch_loss = 0
+            images: Tensor = images.to(device)
+            captions: Tensor = captions.to(device)
 
-        # Progress bar for each batch
-        with tqdm(total=len(data), desc=f"Training Epoch {epoch + 1}/{epochs}", unit="batch") as pbar:
-            for images, captions in data:
-                images, captions = images.to(device), captions.to(device)
-                optimizer.zero_grad()
+            outputs: Tensor = model(images, max_len=captions.size(1))
+            loss: Tensor = criterion(
+                outputs.view(-1, outputs.size(-1)), captions.view(-1))
 
-                # Forward pass
-                outputs = model(images, max_len=captions.size(1))
-                loss = criterion(
-                    outputs.view(-1, outputs.size(-1)), captions.view(-1))
+            loss.backward()
+            batch_loss = loss.item()
+            epoch_loss += batch_loss
 
-                # Backward pass and optimization
-                loss.backward()
-                optimizer.step()
-
[…]

### 2024-10-31 BAI/paper/src/__main__.py
BAI data preprocessing works now and NLP submission not quite ready.

@@ -1,16 +1,15 @@
-from torch import cuda, device, no_grad
-from torch.utils.data import random_split
-from PIL import Image
+from torch import cuda, device
+from torch.utils.data import random_split, DataLoader
 
-from data import FlickrDataset, loader
+from data import FlickrDataset
 from models import ImageCaption, ResnetImageEncoder, GRUCaptionDecoder
-from eval import train, evaluate, visualise
+from eval import train, evaluate
 
 # DIR = "/content/drive/MyDrive"
 DIR = "."
 CAPTION_LIMIT = None
 DEVICE = device("cuda" if cuda.is_available() else "cpu")
-BATCH_SIZE = 800 if cuda.is_available() else 5
+BATCH_SIZE = 800 if cuda.is_available() else 400
 NUM_WORKERS = 2 if cuda.is_available() else 0
 PIN_MEMORY = cuda.is_available()
 
@@ -28,28 +27,20 @@ model = ImageCaption(encoder=encoder, decoder=decoder)
 # --- Training ---
 
 train(model,
-      data=loader(train_dataset,
-                  batch_size=BATCH_SIZE,
-                  num_workers=NUM_WORKERS,
-                  pin_memory=PIN_MEMORY),
+      data=DataLoader(train_dataset,
+                      batch_size=BATCH_SIZE,
+                      num_workers=NUM_WORKERS,
+                      shuffle=True,
+                      pin_memory=PIN_MEMORY),
       epochs=5,
       device=DEVICE)
 
-
 # --- Evaluation ---
 evaluate(model,
-         data=loader(eval_dataset,
-                     batch_size=BATCH_SIZE,
-                     num_workers=NUM_WORKERS,
-                     pin_memory=PIN_MEMORY),
+         data=DataLoader(eval_dataset,
+                         batch_size=BATCH_SIZE,
+                         num_workers=NUM_WORKERS,
+                         shuffle=True,
+                         pin_memory=PIN_MEMORY),
          dataset=dataset,
          device=DEVICE)
-
-
-# --- Inference ---
-with no_grad():
-    model.eval()
-    image = Image.open(f"{DIR}/flickr8k/image.jpg").convert("RGB")
-    caption = dataset.tensor_to_caption(model(
-        dataset.image_to_tensor(image).unsqueeze(0).to(DEVICE)))
-visualise(image, caption)

### 2024-10-31 BAI/paper/src/data/__init__.py

@@ -1,3 +1,2 @@
 from .flickrdataset import FlickrDataset
 from .vocabulary import Vocabulary
-from .loader import loader

### 2024-10-31 BAI/paper/src/data/flickrdataset.py

@@ -1,11 +1,11 @@
 import os
-from torch import tensor, Tensor
+from torch import tensor, Tensor, zeros, float32
 from torch.utils.data import Dataset
 from torchvision.transforms import Compose, Resize, ToTensor, Normalize
 from PIL import Image
-import pickle
+from pickle import dump, load
 from tqdm import tqdm
-import pandas as pd
+from pandas import read_csv
 
 
 from .vocabulary import Vocabulary
@@ -17,13 +17,15 @@ class FlickrDataset(Dataset):
                  caption_limit: int = None,
                  images_folder: str = "Images",
                  captions_file: str = "captions.csv",
+                 data_file: str = "preprocessed.pkl",
                  image_transform: Compose = None,
-                 data_file: str = "compressed_data.pkl",
-                 vocabulary_threshold: int = 5):
+                 vocabulary_threshold: int = 5,
+                 len_captions: int = 40):
         self.image_path = os.path.join(path, images_folder)
         self.captions_file = os.path.join(path, captions_file)
         self.data_path = os.path.join(path, data_file)
         self.caption_limit = caption_limit
+        self.len_captions = len_captions
         self.image_transform = image_transform or Compose([
             Resize((224, 224)),
             ToTensor(),
@@ -32,69 +34,68 @@ class FlickrDataset(Dataset):
 
         if os.path.exists(self.data_path) and caption_limit is None:
             print("Loading preprocessed data...")
-            self.images, \
-                self.captions, \
-                self.vocabulary, \
-                self.caption_to_image_index = self._load()
+            self.vocabulary, self.images, self.captions = self._load()
         else:
-            self.images, \
-                self.captions, \
-                self.vocabulary, \
-                self.caption_to_image_index = self._process(
-                    vocabulary_threshold)
+            self.vocabulary, self.images, self.captions = self._process(
+                vocabulary_threshold)
             print("Saving preprocessed data...")
-            self._save(self.images,
-                       self.captions,
-                       self.vocabulary,
-                       self.caption_to_image_index)
+            self._save(self.vocabulary, self.images, self.captions)
 
-    def _save(self,
-              images: list[Tensor],
-              captions: list[Tensor],
-              vocabulary: Vocabulary,
-              caption_to_image_index: dict[int, int]):
[…]

### 2024-10-31 BAI/paper/src/data/loader.py

@@ -1,23 +1,0 @@
-from torch import Tensor, cat
-from torch.nn.utils.rnn import pad_sequence
-from torch.utils.data import DataLoader, Dataset
-
-from .vocabulary import Vocabulary
-
-
-def loader(dataset: Dataset, batch_size: int = 5, num_workers: int = 2, pin_memory: bool = False) -> DataLoader:
-    class CollateAndPadBatch:
-        """ Collate and pad batch of images and captions. """
-
-        def __call__(self, batch: list[tuple[Tensor, Tensor]]) -> tuple[Tensor, Tensor]:
-            images = [item[0].unsqueeze(0) for item in batch]
-            images = cat(images, dim=0)
-
-            captions = [item[1] for item in batch]
-            captions = pad_sequence(captions,
-                                    batch_first=True,
-                                    padding_value=Vocabulary.pad_index)
-
-            return images, captions
-
-    return DataLoader(dataset, batch_size=batch_size, num_workers=num_workers, shuffle=True, pin_memory=pin_memory, collate_fn=CollateAndPadBatch())

### 2024-10-31 BAI/paper/src/eval/train.py

@@ -36,9 +36,9 @@ def train(model: ImageCaption,
                                  ) if criterion is None else criterion
 
     epoch_loss = 0.0
-    for epoch in trange(model.load_checkpoint(), epochs, desc="Training", unit="epoch", postfix={"Loss": 0.0}):
+    for epoch in trange(model.load_checkpoint(), epochs, desc="Training", unit="epoch"):
         batch_loss = 0.0
-        for images, captions in tqdm(data, desc=f"Epoch", unit="batch", postfix={"Loss": batch_loss}):
+        for images, captions in tqdm(data, desc=f"Epoch", unit="batch"):
             optimizer.zero_grad()
 
             images: Tensor = images.to(device)

### 2024-10-31 BAI/paper/src/models/checkpoint.py

@@ -16,8 +16,6 @@ class CheckpointModule(Module):
             "state": self.state_dict()
         }, self.path)
 
-        print(f"[Save] To {self.path} at checkpoint {checkpoint}.")
-
     def load_checkpoint(self) -> int:
         """ Load the model from the last checkpoint. """
         try:
@@ -29,9 +27,6 @@ class CheckpointModule(Module):
 
             self.load_state_dict(model_dict)
 
-            print(
-                f"[Load] From {self.path} at checkpoint {checkpoint['checkpoint']}.")
-
             return checkpoint['checkpoint']
         except FileNotFoundError:
-            return 0
+            return 1

### 2024-11-05 BAI/paper/paper.tex
Removed BAI paper source to a separate repository.

@@
 CIDEr is a recall-based metric that compares the generated caption to the reference captions using cosine similarity.
+For simplicity, we are only using the BLEU score in our experiments.
 
@@
 
-We experiment with different architectures, such as VGG16, ResNet, and InceptionV3 as encoders and LSTM and GRU as decoders.
-We also experiment with varying attention mechanisms, such as soft and hard attention.
+We experiment with different architectures. For encoding images to features we use ResNet-50 and VGG-16. For decoding captions from features, we use GRU, LSTM and an LSTM with Bahdanau attention.
 
@@
 
-We follow the state-of-the-art approach to image captioning.
-We preprocess the images using the encoder and tokenize the captions.
-We then train the model on the training set and evaluate it on the validation set.
-We generate the captions using beam search.
+We preprocess the images by resizing them to 224x224 pixels and normalizing the pixel values.
+We use the pre-trained ResNet-50 and VGG-16 models to extract features from the images.
+The features are then fed into the decoder to generate captions.
 

### 2024-11-05 PbD/templates/beamer/TemplateSlidesBeamerSVS2021.tex
Removed BAI paper source to a separate repository.

% ===============================================================================
% = LaTeX Beamer Template des Arbeitsbereichs Sicherheit in verteilten Systemem
% = (c) 2016 Prof. Dr. Hannes Federrath, Uni Hamburg, Fachbereich Informatik
% = https://svs.informatik.uni-hamburg.de
% =
% = Weitgehend in Übereinstimmung mit dem Corporate Design 2016 der UHH:
% = https://www.uni-hamburg.de/beschaeftigtenportal/services/oeffentlichkeitsarbeit/corporate-design.html
% = 
% ===============================================================================
%
\documentclass[t,aspectratio=169]{beamer} 
% Option t              Place text of slides at the (vertical) top of the slides.
% Option handout        Ein PDF ohne Pausen und Overlayeffekte erzeugen.
% Option aspectratio=43 169 => 16:9, 1610 => 16:10, 43 => 4:3
\usepackage[utf8]{inputenc}
\usepackage[ngerman]{babel}
\usepackage{graphicx,xcolor}
\usepackage[T1]{fontenc} % 8-Bit-Zeichen; ermöglicht korrektes Kopieren von Umlauten aus dem pdf

% SVS-Theme benutzen
\usetheme{svs2021}

% =============================
% = Ab hier Inhalte ändern... 
% =============================

\title{Template für Folien (LaTeX-Version)}
% \subtitle{Ein Vorschlag}
\author[Federrath]{Prof. Dr. Hannes Federrath}
% \institute[Uni Hamburg]{Universität Hamburg · Fachbereich Informatik}
\date{23.03.2024} % Ausblenden geht mit leerem Inhalt: \date{}
% \titlegraphic{\uhhlogo} % Logo für Titelseite, UHH-Logo ist Standard
\eventlocation{Event, Hamburg, 23. März 2024} % rechts oben auf Titelseite

\begin{document}

\begin{frame}
	% Die Titelseite erscheint nach erneutem Übersetzen korrekt.
	\maketitle
\end{frame}

\begin{frame}
	\frametitle{Agenda}
	% Die Gliederung erscheint nach erneutem Übersetzen korrekt.
	% Die Gliederung wird im Dokument fortlaufend durch \section{} und \subsection{} definiert. Die Folien selbst haben keinen Einfluss auf die Gliederung.
	\tableofcontents
	% Hinweis: Das pandoc-Template benutzt \tableofcontents[hideallsubsections] zur Erzeugung einer Gliederung. Daher werden in der Markdown-Version nur die Überschriften der ersten Ebene (#) im Inhaltsverzeichnis angezeigt. Die normalerweise bei Verwendung von pandoc erzeugten Zwischenfolien für jede Überschrift der ersten Ebene werden im Theme unterdrückt. 
\end{frame}

\section{Der Arbeitsbereich SVS} % erscheint in Agenda
\subsection{Mission} % erscheint in Agenda
\subsection{Themen} % erscheint in Agenda
\subsection{Kontakt} % erscheint in Agenda

\begin{frame}
	\frametitle{Der Arbeitsbereich Sicherheit in Verteilten Systemen (SVS)}
	\begin{alertblock}{Lorem ipsum dolor}
		Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. 
	\end{alertblock}
	\begin{itemize}
		\item Themen
			\begin{enumerate}
				\item Privacy Enhancing Technologies (PET)
				\item Security Management \& Risk Management
				\item Security of Mobile Systems
			\end{enumerate}
		\item Weitere Informationen
			\begin{itemize}
				\item https://svs.informatik.uni-hamburg.de
			\end{itemize}
	\end{itemize}
\end{frame}

\section{Beispiele für Abbildungen} % erscheint in Agenda

\subsection{DRM-Systeme} % erscheint in Agenda

\begin{frame}
	\transwipe % funktioniert nur bei Anzeige mit Acrobat Reader
	\frametitle{Beispiel für eine Abbildung}
	\begin{itemize}
		\item Zweck
			\begin{itemize}
				\item Ziel aus Sicht eines Dienstanbieters \emph{\color{red} A}: Einer Dienstnutzerin \emph{\color[RGB]{0,128,0} N} einen Inhalt (Content) in einer bestimmten Weise zugänglich machen, ihn aber daran hindern, \emph{alles} damit tun zu können.

			\end{itemize}
	\end{itemize}
	\vspace{\fill}
	\begin{center}
		\includegraphics[width=0.7\textwidth]{../pic/Abb1.pdf}
	\end{center}
\end{frame}

\subsection{Weiteres Beispiel für eine Abbildung} % erscheint in Agenda

\begin{frame}
	\frametitle{Weiteres Beispiel für eine Abbildung}
	\framesubtitle{[John Doe, 1966]}
	\begin{itemize}
		\item Voraussetzung: {\color{black} Angreifer} 
			\begin{itemize}
				\item betreibt täuschend echte Webseite der Bank
				\item bewegt den Kunden zum Besuch dieser Seite
			\end{itemize}
	\end{itemize}
	\vspace{\fill}
	\begin{center}
		\includegraphics[width=0.75\textwidth]{../pic/Abb2.pdf}
	\end{center}
\end{frame}

\section{Weitere Layoutbeispiele} % erscheint in Agenda
\subsection{Beispiel für rechtsbündig angeordnete Icos} % erscheint in Agenda

\begin{frame}
	\frametitle{Example for right aligned small pictures close to bullet points}
	\framesubtitle{[Lindemann, 2019]}
	\begin{itemize}
		\item \textbf{Attack} to detect applications in co-resident virtual machines 
		\begin{itemize}
			\item Automatically generating signatures for individual \\ versions and groups of versions
			\item Evaluation without load and under load
		\end{itemize}
		\item \medskip \textbf{Defence} mechanism against memory deduplication side-channel attacks
		\begin{itemize}
			\item Fake deduplication of non-duplicate pages to level \\ out write times
			\item Evaluation of performance impact
			\item Evaluation of impact on side-channel effectiveness
		\end{itemize}
	\end{itemize}
	% ----- Ab hier der Code für die rechtsbündigen Icons
	\raggedleft % rechtsbündig
	\par\vspace{-4.5cm} % Cursor nach oben (minus x) bewegen
	\includegraphics[height=40px]{../pic/sword.pdf} % https://upload.wikimedia.org/wikipedia/commons/1/19/Sword_01.svg
	\par\vspace{1.4cm} % Cursor nach unten bewegen
	\includegraphics[height=40px]{../pic/shield.pdf} % https://commons.wikimedia.org/wiki/File:Weapon_shield.svg - CC-0
\end{frame}

\subsection{Ebenen und Spalten} % erscheint in Agenda

\usebackgroundtemplate{\includegraphics[width=\paperwidth]{../pic/bg-maze-gradient.jpg}}
\begin{frame}[fragile]
	\frametitle{Ebenen und Spalten}
	\begin{columns}[T]
		\begin{column}{.47\textwidth}
			\begin{itemize}
				\item Erste Ebene
					\begin{itemize}
						\item Zweite Ebene
						\begin{itemize}
							\item Dritte Ebene
						\end{itemize}
						\item Zweite Ebene
					\end{itemize}
				\item Erste Ebene
			\end{itemize}
		\end{column}		
		\begin{column}{.47\textwidth}
			\begin{enumerate}	
				\item Erste Ebene
					\begin{enumerate}
						\item Zweite Ebene
						\begin{enumerate}
							\item Dritte Ebene
						\end{enumerate}
						\item Zweite Ebene
					\end{enumerate}
				\item Erste Ebene
			\end{enumerate}
			\vspace{0.5cm}
			\begin{block}{}
				% Damit wir \verb|...| nutzen können, muss \begin{frame}[fragile] verwendet werden.
				Hier wird gezeigt, wie mit \verb|\usebackgroundtemplate| ein Hintergrundbild verwendet werden kann. 
			\end{block}
		\end{column}
	\end{columns}
\end{frame}
\usebackgroundtemplate{}

\begin{frame}{Ebenen und Spalten}
	\begin{columns}[T]
		\begin{column}{.57\textwidth}
			\begin{itemize}
				\item Linke Spalte
				\begin{itemize}
					\item Lorem ipsum dolor sit amet, 
					\item consectetur adipisicing elit, 
					\item sed do eiusmod tempor incididunt ut 
					\item labore et dolore magna aliqua. 
				\end{itemize}
				\item Erste Ebene
				\begin{itemize}
					\item Zweite Ebene
					\item Zweite Ebene
				\end{itemize}
				\item Erste Ebene
				\begin{itemize}
					\item Zweite Ebene
					\item Zweite Ebene
				\end{itemize}
			\end{itemize}
		\end{column}
		\begin{column}{.37\textwidth}
			\begin{center}
				\vspace{0.5cm}
				\includegraphics[width=2.2cm]{../pic/svs_logo_uhhred.png} \\
				\small
				Das alte SVS-Logo wird seit 2016 aufgrund der CI-Richtlinien der UHH nicht mehr verwendet.
			\end{center}
		\end{column}
	\end{columns}
\end{frame}

\end{document}

### 2024-11-05 PbD/templates/latex/stylesvs.tex
Removed BAI paper source to a separate repository.

%!TEX encoding = UTF-8 Unicode
%!TEX root = hinweiseabschlussarbeit.tex

% Kodierung, Sprache, Patches {{{
\usepackage[T1]{fontenc}    % Ausgabekodierung; ermöglicht Akzente und Umlaute
                            %  sowie korrekte Silbentrennung.
\usepackage[utf8]{inputenc} % Erlaubt die direkte Eingabe spezieller Zeichen;
                            %  utf8 muss die Eingabekodierung des Editors sein.
\usepackage[ngerman]{babel} % Deutsche Sprachanpassungen (z.B. Überschriften).
\usepackage{microtype}      % Optimale Randausrichtung und Skalierung.
\usepackage[
    autostyle,
    ]{csquotes}             % Korrekte Anführungszeichen in der Literaturliste.
%\usepackage{fixltx2e}      % Patches fuer LaTeX2e - seit 2015 nicht mehr nötig
\usepackage{scrhack}        % Verhindert Warnungen mit älteren Paketen.
\usepackage[
  newcommands
]{ragged2e}                 % Verbesserte \ragged...Befehle
\PassOptionsToPackage{
  hyphens
}{url}                      % Sorgt für URL-Umbrüche in Fußzeilen u. Literatur
% }}}

% Schriftarten {{{
\usepackage{mathptmx}       % Times; modifies the default serif and math fonts
\usepackage[scaled=.92]{helvet}% modifies the sans serif font
\usepackage{courier}        % modifies the monospace font
% }}}

% Biblatex {{{
\usepackage[
    style=alphabetic,
    backend=biber,
    %backref=true
    ]{biblatex}             % Biblatex mit alphabetischem Style und biber.
\usepackage[realmainfile]{currfile} % Overleaf unterstuetzt \jobname nicht richtig.
\bibliography{\currfilebase.bib} % Dateiname der bib-Datei (funktioniert auch mit Overleaf)
%\bibliography{\jobname.bib} % Dateiname der bib-Datei (funktioniert nicht mit Overleaf)
\DeclareFieldFormat*{title}{
    \mkbibemph{#1}}         % Make titles italics
% }}}

% Dokument- und Texteinstellungen {{{
\usepackage[
    a4paper,
    margin=2.54cm,
    marginparwidth=2.0cm,
    footskip=1.0cm
    ]{geometry}             % Ersetzt 'a4wide'.
\clubpenalty=10000          % Keine Einzelzeile am Beginn eines Absatzes
                            %  (Schusterjungen).
\widowpenalty=10000         % Keine Einzelzeile am Ende eines Absatzes
\displaywidowpenalty=10000  %  (Hurenkinder).
\usepackage{floatrow}       % Zentriert alle Floats
\usepackage{ifdraft}        % Ermöglicht \ifoptionfinal{true}{false}
\pagestyle{plain}           % keine Kopfzeilen
% \sloppy                    % großzügige Formatierungsweise
\deffootnote{1em}{1em}{
  \thefootnotemark.\ }      % Verbessert Layout mehrzeiliger Fußnoten
% Linien zwischen Kapitelnummer und Titel.
\usepackage[explicit]{titlesec}
\titlespacing*{\chapter}{0pt}{1.05cm}{0.7cm}
\titleformat{\chapter}[hang]{\Large\sffamily\bfseries}{\color{black}\thechapter}{8pt}{\begin{tabular}[t]{@{\color{gray!50}\vrule width 2pt\hspace{8pt}}p{0.85\textwidth}}\raggedright #1\end{tabular}}
\titleformat{name=\chapter,numberless}[hang]{\Large\sffamily\bfseries}{}{0pt}{\raggedright #1\markboth{#1}{#1}}
\titleformat{\section}[hang]{\large\sffamily\bfseries}{{\thechapter.\arabic{section}}}{1ex}{#1}{}
%\setkomafont{disposition}{\normalcolor\bfseries} % Aus der KOMA-Skript-Anleitung: „Mit dieser Änderung verzichten Sie darauf, für alle Gliederungsebenen serifenlose Schrift voreinzustellen“

\makeatletter
\AtBeginDocument{%
    \hypersetup{%
        pdftitle = {\@title},
        pdfauthor  = \@author,
    }
}
\makeatother
% }}}

% Weitere Pakete {{{
\usepackage{graphicx}       % Einfügen von Graphiken.
\usepackage{tabu}           % Einfügen von Tabellen.
\usepackage{multirow}       % Tabellenzeilen zusammenfassen.
\usepackage{multicol}       % Tabellenspalten zusammenfassen.
\usepackage{booktabs}       % Schönere Tabellen (\toprule\midrule\bottomrule).
\usepackage[nocut]{thmbox}  % Theorembox bspw. für Angreifermodell.
\usepackage{amsmath}        % Erweiterte Handhabung mathematischer Formeln.
\usepackage{amssymb}        % Erweiterte mathematische Symbole.
\usepackage{rotating}
\usepackage[
    printonlyused
    ]{acronym}              % Abkürzungsverzeichnis
\usepackage[
    colorinlistoftodos,
    textsize=tiny,          % Notizen und TODOs - mit der todonotes.sty von
    \ifoptionfinal{disable}{}%  Benjamin Kellermann ist das Package "changebar"
    ]{todonotes}            %  bereits integriert.
\usepackage[
    breaklinks,
    hidelinks,
    pdfdisplaydoctitle,
    pdfpagemode = {UseOutlines},
    pdfpagelabels,
    ]{hyperref}             % Sprungmarken im PDF. Lädt das URL-Paket.
    \urlstyle{rm}           % Entfernt die Formattierung von URLs.
%\usepackage{breakurl}
%\def\UrlBreaks{\do\/\do-}
\usepackage{listings}       % Spezielle Umgebung für Quelltextformatierung.
    \lstset{                
        language=C,
        breaklines=true,
        breakatwhitespace=true,
        frame=l,            % Linie links: l, doppelt: L
		framerule=2pt,      % Dicke der Linie (angeglichen an den sekrechten grauen Balken in Kapitelüberschriften)
		rulecolor=\color{gray!50},% Farbe der Linie (angeglichen an den sekrechten grauen Balken in Kapitelüberschriften)
        captionpos=b,
        xleftmargin=6ex,
        tabsize=4,
        numbers=left,
        numberstyle=\ttfamily\footnotesize,
        basicstyle=\ttfamily\footnotesize,
        keywordstyle=\bfseries\color{green!50!black},
        commentstyle=\itshape\color{magenta!90!black},
        identifierstyle=\ttfamily,
        stringstyle=\color{orange!90!black},
        showstringspaces=false,
        }
%\usepackage{filecontents}  % Direktes Einfügen von Dateiinhalt. Wird hier für
                            %  die Verwendung einer .bib-Datei in dieser .tex-
                            %  Datei benötigt.
% }}}

\newcommand{\uhhlogo}{\pdfliteral{%(fold)
q                        % Save graphic state
1090 0 0 353 0 0 cm      % Translation
1 0 0 1 0 0 cm           % Rotation
0.174 0 0 0.174 0 0 cm   % Scaling
BI                       % Begin Image
/W 1090                  % Width
/H 353                   % Height
/CS /RGB                 % Color Space 
/BPC 8                   % Bits Per Component
/F [ /AHx /Fl ]          % Filter ASCIIHexDecode FlateDecode
ID                       % Image Data
789CEDDB4DCE6339AE20D0596FA0870FB5FF85F542B2070F0814AA1CB624FEC9D7E780C3CF124951
0E1989FCE79F7FFEDFFFF9BF4288BFC53F0000DC67FC9528C4CD317D4101007861FC9528C4CD317D
4101007861FC9528C4CD317D4101007861FC9528C4CD317D4101007861FC9528C4CD317D41010078
61FC9528C4CD317D4101007861FC9528C4CD317D4101007861FC9528C4CD317D4101007861FC9528
C4CD317D4101007861FC9528C4CD317D4101007861FC9528C4CD317D4101007861FC9528C4CD317D
4101007861FC9528C4CD317D4101007861FC9528C4CD317D4101007861FC9528C4CD317D41010078
61FC9528C4CD317D4101007861FC9528C4CD317D4101007861FC9528C4CD317D4101007861FC9528
C4CD317D41010078A1E18D37FBC2B4BBA8EE330000CD1ADE78B32F4CBB8BEA3E0300D0ACE18D37FB
C2B4BBA8EE330000CD1ADE78B32F4CBB8BEA3E0300D0ACE18D37FBC2B4BBA8EE330000CD1ADE78B3
2F4CBB8BEA3E0300D0ACE18D37FBC2B4BBA8EE330000CD1ADE78B32F4CBB8BEA3E0300D0ACE18D37
FBC2B4BBA8EE330000CD1ADE78B32F4CBB8BEA3E0300D0ACE18D37FBC2B4BBA8EE330000CD1ADE78
B32F4CBB8BEA3E0300D0ACE18D37FBC2B4BBA8EE330000CD1ADE78B32F4CBB8BEA3E0300D0ACE18D
37FBC2B4BBA8EE330000CD1ADE78B32F4CBB8BEA3E0300D0ACE18D37FBC2B4BBA8EE330000CD1ADE
78B32F4CBB8BEA3E0300D0ACE18D37FBC2B4BBA8EE330000CD1ADE78B32F4CBB8BEA3E0300D0ACE1
8D37FBC2B4BBA8EE330000CD1ADE78B32F4CBB8BEA3E0300D0ACE18D37FBC2B4BBA8EE330000CD1A
DE78B32F4CBB8BEA3E0300D0ACE18D37FBC2B4BBA8EE330000CD1ADE78B32F4CBB8BEA3E0300D0AC
E18D37FBC2B4BBA8EE330000CD1ADE785E9853C67F023C20A6CF100080171ADE785E9853C67F023C
20A6CF100080171ADE785E9853C67F023C20A6CF100080171ADE785E9853C67F023C20A6CF100080
171ADE785E9853C67F023C20A6CF100080171ADE785E9853C67F023C20A6CF100080171ADE785E98
53C67F023C20A6CF100080171ADE785E9853C67F023C20A6CF100080171ADE785E9853C67F023C20
A6CF100080171ADE785E9853C67F023C20A6CF100080171ADE78B32F4CBB8BEA3E0300D0ACE18D37
FBC2B4BBA8EE330000CD1ADE78B32F4CBB8BEA3E0300D0ACE18D37FBC2B4BBA8EE330000CD1ADE78
B32F4CBB8BEA3E0300D0ACE18D37FBC2B4BBA8EE330000CD1ADE78B32F4CBB8BEA3E0300D0ACE18D
37FBC2B4BBA8EE330000CD1ADE78B32F4CBB8BEA3E0300D0ACE18D37FBC2B4BBA8EE330000CD1ADE
78B32F4CBB8BEA3E0300D0ACE78D37F8BC9C7DDF3E75F7DF89C5310300A053CF1B6FEA7939FEBE7D
EAEEBF132B7D0600A059CF336FEA7939FEBE7DEAEEBF132B7D0600A059DB336FE46D39FEBE7DEAEE
BF138B93060040A7B667DEC8DB72FC7DFBD4DD7F2716270D00804E9DCFBCE687E50DEFDBA7EEFE3B
B1386C0000746A7EE97DD75EF14C9EBAFBEFC4FABC0100D0A6FFA5F7151B6525F3D4DD7F27B6460E
00801E232FBD865D2E79DF3E75F7DF89ADA90300A0C7E063EF92C54B4B7EEAEEBF1307B3070040B5
AB1E7B594BDDF3BE7DEAEEBF13B1490400A0C4F3DE7B57BD6F9FBAFBEF446C18010028F1BC27DF55
EFDBA7EEFE3B111B4600004A3CECD577DBFBF6A9BBFF4EC4E6110080124F7AF85DF8BE7DEAEEBF13
B1910400A0C433DE7ED7BE6F9FBAFBEF446C30010028F180E7DFCDEFDBA7EEFE3B119B4D00004A7C
F523F0FEF7ED5377FF9D884D28000025BEF43538FEB85DACE8A9BBFF4EC4E6140080125FF42C1C7F
D08A1F8CC4BB06004096F157A21037C7F4050500E085F157A21037C7F4050500E085F157A21037C7
F4050500E085F157A21037C7F4050500E085F157A21037C7F4050500E085F157A21037C7F4050500
E085F157A21037C7F4050500E085F157A21037C7F4050500E085F157A21037C7F4050500E085F157
A21037C7F4050500E085F157A21037C7F4050500E085F157A21037C7F4050500E085F157A21037C7
F4050500E085F157A21037C7F4050500E085F157A21037C7F4050500E085F157A21037C7F4050500
E085F157A21037C7F4050500E085F157A21037C7F4050500E085F157A21037C7F4050500E085F157
A21037C7F4050500E085F157A21037C7F4050500E085F157A21037C7F4050500E085F157A21037C7
F4050500E085F157A21037C7F4050500E085F157A21037C7F4050500E085F157A21037C7F4050500
E085F157A21037C7F4050500E085F157A21037C7F4050500E085F157A21037C7F4050500E085F157
A21037C7F4050500E085F157A21037C7F4050500E085F157A21037C7F4050500E085F157A21037C7
F4050500E085F157A21037C7F4050500E085F157A21037C7F4050500E085F157A21037C7F4050500
E085F157A21037C7F4050500E085F157A21037C7F4050500E085F157A21037C7F4050500E085F157
A21037C7F4050500E085F157A21037C7F4050500E085F157A21037C7F4050500E085F157A21037C7
F4050500E085F157A21037C7F4050500E085F157A21037C7F4050500E085F157A21037C7F4050500
E085F157A21037C7F4050500E085F157A278523C6FBA466E250000EF8DBF12C593E279D335722B01
00786FFC95289E14CF9BAE915B0900C07BE3AF44F1A478DE748DDC4A0000DE1B7F258A27C5F3A66B
E4560200F0DEF82B517C7B3C7BD24A6F1F000067C65F89E2DBE3D993567AFB00003833FE4A14DF1E
CF9EB4D2DB0700C099F157A2F8F678F6A495DE3E0000CE8CBF12C593E279D335722B0100786FFC95
289E14CF9BAE915B0900C07BE3AF44F1A478DE748DDC4A0000DE1B7F258AA9A838FDEAE9EA9FD881
3B0900C027CD6F42714F540C40E9748D4C6CEB6D040060CDC8FB598C47EE00F4CCD8C8D0565D3C00
0002C69FD3A23F7267A067D2A68636E79A0100906AFC452D7AA2681E4A87ED8601DE2A1000801EE3
AF6BD116E9C3503D6F370CF0418D0000541B7F5A8BB6C81D869E911B9FDEE3320100A833FEB4169D
9135099D23373BBD914A01002832FEAE166D91380C6D53373EBDC14A0100A830FEB4166D91380C3D
5377C300072B0500A0C2F8D35AF444E23CB44D5DE252D57D0300A0D3F8EB5AF447700646A66E6A6E
E3C50200906EFC452D4622320023533735B4F16201004837FE9C1623B13B00370CE1C8C496160E00
C099F1E7B4988AADD3BF6408FB27B6B4700000CE8CBFA5C557C4CF0E6169E100009C197F258AAF88
9F1DC2D2C201003833FE4A145F113F3B84A58503007066FC95284AA368304AA7AE68D9E606020050
67FC992D4AA368304AA7AE68D9E60602005067FC992D4AA368304AA7AE68D9E60602005067FC992D
4AA368304AA7AE68D9E60602005067FC992D4AA368304AA7AE68D9E60602005067FC992D4AA36830
4AA7AE68D9E60602005067FC992D4AA368304AA7AE68D9E60602005067FC992D4AA368304AA7AE68
D9E60602005067FC992D4AA368304AA7AE68D9E60602005067FC992D4AA368304AA7AE68D9E60602
005067FC992D4AA368304AA7AE68D9E60602005067FC992D4AA368304AA7AE68D9E60602005067FC
992D4AA368304AA7AE68D9E60602005067FC992D4AA368304AA7AE68D9E60602005067FC992D4AA3
68304AA7AE68D9E60602005067FC992D4AA368304AA7AE68D9E60602005067FC992D4AA368304AA7
AE68D9E60602005067FC992D4AA368304AA7AE68D9E60602005067FC992D4AA368304AA7AE68D9E6
0602005067FC992D4AA368304AA7AE68D9E60602005067FC992D4AA368304AA7AE68D9E606020050
67FC992D4AA368304AA7AE68D9E60602005067FC992D4AA368304AA7AE68D9E60602005067FC992D
4AA368304AA7AE68D9E60602005067FC992D4AA368304AA7AE68D9E60602005067FC992D4AA36830
4AA7AE68D9E60602005067FC992D4AA368304AA7AE68D9E60602005067FC992D4AA368304AA7AE68
D9E60602005067FC992D4AA368304AA7AE68D9E60602005067FC992D4AA368304AA7AE68D9E60602
005067FC992D12637A9AAEA3A500008F34FEF01689313D4DD7D1520080471A7F788BC4989EA6EB68
2900C0238D3FBC45624C4FD375B41400E091C61FDE2231A6A7E93A5A0A00F048E30F6F9118D3D374
1D2D050078A4F187B71037C7F4050500E085F157A21037C7F4050500E085F157A21037C7F4050500
E085F157A21037C7F4050500E085F157A2288DE9F96AA58100003F62FC992D4A637ABE5A692000C0
8F187F668BBA981EAE6E7A0800F023C65FDAA22EA687AB9B1E0200FC88F197B6A88BE9E11AA08700
00BF60FCA52DEA627AB8063CB887FFFAD7FFFC2DBE6E1780446FBEB812BFBB7A7621C819FD94F197
B6A88BE9E11AF0E01EFA1503BF63EBD2B9A47EC5F08733FA29E32F6D5117D3C335E0C13DF42B26E2
49B5B0EB8B8679F7EA3DEFAA9EF12B863F9CD14F197F698BBA981EAE010FEEA15F31679E57115BBE
E8F47767D56CFFE1570C7F38A39F32FED21675313D5C031EDC43BF620E7875FCB8EF1A80AD6CBFAB
B46A7EC5F08733FA29E32F6D5117D3C335E0C13DF42BE6C0FB57C79716C5BAD201489FA5DB7EC57C
D14DF12B863F9CD14F197F698BA2989EAC190F6EA35F3107FC8AF9651F4FFF6C00EA66E9865F315F
7A53FC8AE10F67F453C61FDBA228A6276BCC53DBE857CC01BF627E59FAAF98EA591AFF15F3BD37C5
AF18FE70463F65FCB12D8A627AB2C63CB58D7EC51CF02BE697F915939840B8B85A7EC5F08733FA29
E38F6D5114D39335E6A96DF42BE6805F31BFEC49BF6252FE3E3D817BF815C31FCEE8A78C3FB64551
4C4FD698A7B6D1AF98037EC5FCB8DCD36F98A5945F31E9BBDF7F53FC8AE10F67F453C61FDBA228A6
276BCC53DBE857CC99E755C496EFFA15F3DF1BE5FEF1FA52DF7559FC8AE10F67F453C61FDBA222EE
99A54BD2F8AE4EBEE457CCB18795C3AEACD37FFC23F67BABF32B863F9CD14F197F6F8B8AB86D8A6E
CBE7DA4EBEE1570CCC7AFC23F67BABF32B863F9CD14F197F6F8B8AB8707EDA52DACAEAAAB4DFF32B
06663DFE11FBBDD5F915C31FCEE8A78CBFB745455C3B3CD726369BF00ABF6260D6E31FB1DF5B9D5F
31FCE18C7ECAF87B5B54C4B593D390D8716E8309AFF02B06663DFE11FBBDD5F915C31FCEE8A78CBF
B7457A5C3E3937E736D5CC158FFF15E35F9C3357FD639D95C65545BDCFAA21BDB61DBFF7F9F7C85F
31379F45565615D59D9DD19D7DBE444567B2667BFCC92DD22365C04AC7E6FE0C3B535DD4F3AF67CA
2E5B8BBC7F18C46B3C58B32293DC35179BD6D6BAD29989ACDCD9F6DDD5CE7629DDBDA1BA22371C4D
E9E275A56D2D9E92527569EB6B3634396BE5DC0C83FDD9DD6EB18AB3F91F7F728BF4381EB0758FCF
B033D545A5FF90E5EEB2B8C8C1F7554F451599F4FC5BD696F3D6DF545494B2787A569184DB361AA9
AE5A4F7A15BB8C1FC1E29A29F9F4D4D559512499AD3573730BF6276BBB78FCE357CC136377C076DD
9F614A926DA92E4AFF76ADDBE5E32291AFAC868A9AD3585F30FE859F95734A510DB51C2F126CF56E
C29D7BF557D7A027BDF45D6E38858F0B66E5D356575B45C164B616EC69D1CA5EBBDB1D777BF144C6
9FDC223D76E779D7EF24D990E7BADCAFD6D25DDE2F12FFD66AA8A82D87C50513BFF353D28E17D553
C8F122590DEFE9C6E5DB35E8492F7D971B0EE2FD6A59F97416D553513C99ADD57AB25AD92B31ED78
FCE357CCE36277980F7C459E5FD4CF4589DF60D5BB347C715557D496C3CA5253DD7BBF4864A3B62A
0633DC4A387DBBABAA6BD0935EFA2ECDE7DE96C37F24D35C546945B9C9F4AFF37EB58F1B25E69C75
16E3AF6E911BBBC37C409E2312BFC1AA77F9967F44225564E5B0B2D454038B76B9A7847B7A9EBEDD
55D535A8686043136E4863BC6FE9757D5132FDEBC45BD4B0C55632E3AF6E911BBBC37C409E2312BF
C1AA77F9967F442255647523F8F1FF582AB77B4567D43900C78BA42499356C771EC1C171D4296A60
75136EC864BC6FE9457D5132FDEBC45B94BE7E249F7FFC8A795CEC0EF301798E48FC06ABDE25F2C5
D8F98F48B08A946E14E59092FCF1BF2C6FB6881F6E4FF383B5671D7430BD4BAA6B905E514F139ACF
3DAB75596DDF5AE7B68A5292C92A6A7D9DDD320FF68A2C7290CCF8AB5B24C6EE249FF99654BF25CF
4589DF60D5BBC4BF12B3FE0509569492436491A9CFAEAC50D1D2DC61FEB8DDEC6A6F564EC927B7C0
DDA59A1DCC6A7A04D34E2CF63887B395537A9558544F4529C96415B5BECE7A81451395BEC2F8C35B
24C6D6A41DFBA9547BF25C91F80D56BD4BCA7763C3977FB090E0EE1F5798DDFDE3C72B66637D9D91
EDEA923F5EA1ADC083DC3AADCF6A5D1CA79D5E6FE2528BCB362C725B4529C96415B5BECE6275F1A9
6EEBDBF8C35B24C6C1BC1DF8A9547BF25C91FB3D53BA4BCAD763C3977FBC90E00A7599C7775F29FF
60FC7AC678B184D9D5B2F414589179A2C5592D8D7B4ACE5A677DCDAC16A574B82799C57CB2662677
F6EAE6B9BFDEF187B7C88AB3913BF045D97E4B9E2BD2BF6DEA76C9FA86AC4E265E4BE4B3EF3FDEF0
6F41B0F6B3D9CB5DAD79BBE6E4E3590D2ED56C6556ABE39E92B3D6C9FDAE88AFD359D1CA3AF14516
2B4A5C67B1B4E379EE49F2DFFF6CFCED2DB2E26CE40EFC54B66D797ED4F3AF67C3F76DCA3F466D15
450A99FAECE23AC1DA2BFEF5DC5DAD79BBE6E4E3590D2ED56C6556ABE39E92B3D6595F2DAB3F291D
4E3CA94B2A4A5C6765B5C830F724F9EF7F36FEF61659713C75BB7E2ADBB63C3FEAF9D7B3E1FB36E5
1FA3CE8A8E6B893421EBB88B72381EBCBA7F3D0FB69B5D2D4B4F81159927FA38570D714FC959EB6C
ADF62D8B6CF527BECE3D99ACAF1619E69E24FFFDCFC6DFDE222B8EA76ED74F65DB96E7473DFF7A5E
F57D7BC92267E5049BF0F15F999408E6B0D8C0DDBA0E963DDB6E76B52C3D0556649EA8E768AE1A80
AC4CB28A4A59E79E45B296BAAABD45AB2DAE5CB4D4F8DB5BA44464EA76FD54C29D79BED7F3AFE755
DFB7972CF2BE9CBFAD1369C2C71DB3A2A2F0C4D20ED6DFDA6E76B52C3D0556649EA8E768AE1A80AC
4CB28A4A59E79E45B296BAAABD45AB2DAE5CB4D4F8F35BA44464EA76FD54C29D79BED7F3AFE755DF
B75FB1C8DFD68934E1FD671323BDEA15B949769650976770EB86022B8A4AD47334FD03D070EE594B
552FB2BECE55CD69686FEE4D0F4E724F92FFFE67E3CF6F911291A9DBF5530977E6F95ECFBF9EF7FC
73765532BBDFF9C17F23DE7F3C3172AB5E54946D4309450D39DBAEA2639DD5E5EA399A1B76493FF7
ABD649A9ABA7398B4B6525F32D37BDBFDEF1E7B7884770EA767D5DCEDF92E77B0DFF7A66EDF22DFF
2CA62CF272A9E037F9C7EDB22292C67AF7720B4CDC2837EDDDD58E372AEAD5FB7D13ABABD07334D5
BB8C9CFB55EBA4D4957B46C1D5B24EEA5B6E7A7FBDE32F70118FE0D4EDFAA99C9BF37CA3F45FCF8F
5B647DFF64E5735532FFBD54E93F7C899158F2AEBAB4EB4A286DC8CA16758DFAB87B4A75751A8EA6
7497C173BF6A9D94BA72CF28B85AD6497DCB4D4F19CEAD15C65FE0221EC1A9DBF5533937E7F946DD
D7CECA1659DFB759F95C95CC7F2C55FD1D9E1589259F294DBEA2849B1B525D60BCBA52D54753BACB
ECB95FB54E4A5D3D376271B5AC93FAA29B1E19D183CF8EBFC04530E223B7EB1BD3FE963CDF28FDDA
F9B845D6F76D563EFDC9BC5FEADF57BBE71FBEA09E3456FED9BAA493A50D39EE434F81C1EAAADD30
AB156B369CFB55EBA4D4D573231657CB3AA92FBAE907B31A99EDF147B808467CE476FD54DAFD79FE
4DEE9758E916DFF2CFE25632EF97FAD7DAAF98DC8DAA75A6B1F50FDF7A0EB925D4352452784F8191
EA1AF4CC6AFA2E379CFB55EBA474F8AA2B9F3533B9B3577D597607FBEC0AFCAFF147B80846CAC86D
F9A9B4FBF3FC9BDC2FB1D22DBEE59FC5AD64DE2FF567B5941E369CF5B56904FF452B2DA1A821C14A
7B0A3CAEAE47CFAC768E53DBB95FB54E4A87AFBAF25933D3367BBB4B1D647B1C2FF71A7F848B48A4
CCDBAE9FCA7C24CF9772BFC44AB7F8967F16B792F9B8DAFF2E98D2C386B3BE3C8DC83F6A75251435
A4AEC6C44C8E2AEBD333AB89BB94CE765626FDEBA474F8AA2B9F3533B9133E7E590EE26F1B8DBFC3
4524B2E66DCB4F653E92E74BB95F62A5EB5FB54E6ED33E76E9AADA83AA472E98C04A0EB925543424
BE664F81FB95B5EA99D5C45DEE39F7ABD649E9F05557BEBAA29B6FFAC7B417E3FD2EE3EF70216E8E
C41B1D94FE3DB6B57ED652FDEBE476ACEEAB786BA383CC0FDC9FC3C734724BA8684870C1B602F72B
6BD533AB3D17FCBB32C95D27A5AE868A5292C95AE4DA9BFE31ED95F8B8CBF82B51889B23F14607A5
7F8FD52D7ED53AE91D2BFA2ADEDAE52CF36FCC2198C6FD87125CB0ADC0FDCA5AF5CC6AD62EF17512
EBBD6A9D94BA1A2A4A4926E5ACAFBDE92B69A7A431FE4A14E2E6C8BAD129D2BFCA16574E5CAD7F9D
F476557F272FEE72967C620E0D097C4C23F2D9F443692EED633E8337B75FCFAC66ED72D5B95FB54E
4A5D0D99A4249375DC17DEF49EFBF8BFC65F8942DC1CB9D72DA8E2DB6C65D9C405FBD7A9F82E5D39
88BA1E56FC437090434302F11C3A0FE5B6EA066F6EBF9E59CDDA25B848EEB95FB54E4A5DA52B3457
F431933B6F7AC365FC63FC9528C4CD917EE322D2BFD02EFF86ACAE71B3ACA5351337AA5B362581B3
DD770573681BECB38644166CBEB96D277EA627F3AC5DAE3AF7ABD649A92BB8426E87234BAD64927E
D37797DA4D3BBEFE7F187F250A7173A4DFB8A0F5AFB5375F17298B9C25D9BF4EC5D7E956038F374A
5F7CF753CDFF18ED26B092436E09E9077DB660DDE0154D72839EB42B6EE2F8B95FB54E4A5D91152E
B956EB679D9ED2EE5229F5468CBF1285B83972AF5BDCEE975B30D293EC5FA7E8BBB4A18D5BBBECAE
10DF3DABA88AAAEB4AC8BD4ABB1F4CD9B4ADBA4E3DE965ED72D5B95FB54E4A5D072D2A1DF5F59577
0FFA38ABDC02CFEA4DD9EB1FBF6284781B29B72CD7F1175DC377E3C70CFBD729FA0AEDF9A29E3DD9
A2BABEBA84F46CEB8EF8A0C69E5D2AF4A47743AB3B0FBD7F9D94BAEEE96D433E6759E516D853EFDF
B61B7F250A7173C46F7485866FC5C8B759D69A29EB147D5D37B47177A3F4C3ED1F8FF4AEA697909E
70D1F99ED5D8B34B859EF46E6875E7A1F7AF9352D7258D6DC8E738B1F41ADBEAFDEFEDC65F8942DC
1CF11B5DE4B66FC5C5DCFAD7A9FBBAAE6EE3D64615E75B545AE77CA697909E704A2B126B2C3D8E3A
3DE925EE72CFB9375494D5969445220D3F76CF717FCC27526662BDEB6731FE4A14E2E648B9D1457A
BE227213EB5F67A4CCDC7F1116F74A3FE2A2D23AE7B3A284F49CE3AD48ACB1FA448AF4A497BBCB25
E79E55517532EBEB441A5B344229BBF7CC5EB0D2947AB74E64FC9528C4CD9172A3EBDCF65DFD31AB
FE754A8BEDE9E7E25EE9075D545AE77C569450917970CDDC1A1BCE255D4F7AE9BBDC70EE59155527
B3BECE59576787A7EDB8D3974AAF77B72DE3AF44216E8EAC1B5DEAAAEFEAF7F9F4AF535A725B4B17
77CC3DEBA2EA3AE7F39E12D2D75CFC6C4F8167BB24EA49AF7F441BCE3DABA2EA64D6D7D96D69DB54
0773484C75FCB2DC7940C0AC4BBE07B2B64B59A7B4F6A9EFD89EEFFCA2EA3A47B4F4802AAA385B6A
FC988E77C9D293DED4A0969E7B5645D5C99C75786B8C7B86F9388DC49CABCB3F68FB8527058C73F1
01801EF15F1F7EC80000009DB27E74F8150300003448FFD1E1570C000050A7E217875F310000409D
A29F1B7EC500000045FC8A010000BE8B5F310000C077A9F8B9E1FF8B010000EA34FFDFFDB9C90300
003F28FD578CFF10030000947AFFA363EBA747D63A0000006FACFC8AF9F81B24BE020000C0BAF51F
329198AE120000780E3F61000080AFE3270C0000F075FC84010000BE8EDF2F0000C0D7F1FB050000
F8527EBC0000005FCDCF160000000000000000000000000000000000000000000000000000000000
00000000000000000000000000000000000000000000000000000000000000000000000000000000
0000000000000000000000FEC3BFFEF53FFF1ED3E9001FB8B3C0DFFCC7F7C3C7E8D9E5CD76D5A9DE
69E450AECDFCFDE229691374EDDDBC24936BFBF3DF7EEA9BF672BF79103794F9459DBFF0C25E98D2
B1C714729548573F4E57CA79256E519DEAAED2EE1595D9D0C01B565E9F9FE0EE297FBC58E6D61F07
276D31D5C534126720E29EAD8B5A94B8C56DA719DFAEA139C14C12D36E6E489DA9AD7B9AD3309323
E758DABDAD0F0E9ED1D646C115E23D59DFF1AAC98F9C4E70E4CE16EF19A77575C97CC571C4175F59
F338EDC5458239A4FCF162B15B7F1C1CB6C5545712C89A81B87BB6BEE402EE5E96C1038DEFD5D09C
6026B9697FCCE7B6235E49726ADFA22E35CCE4C86996B66EEB23F19267272192E4594523A91E3733
7E3A91920F566E98A52D1599549FC8552B37373938155B1F8FECB5D5AE7893FB8FE6ACFF71F76C5D
D4ABEACB3278ACF15D1A9A13CC2437ED8F29DD73B8EB494EED5BD4A886991C39D0D2D66DFD7DBCDE
9449F8B869F068E23DE94935BD8DB3A773B06C75AABBD2D3683891A2958B8E2337D5E048C41B12D9
E86F1F8FF7B9FF680E9A9F6270F794767D4CBBFAB20C1E6E7C8B86E60433C94DFB6356979CEC7B53
5BC7DBBE9270C34C8E9C6969EBEAFEB8AE968FFB06CF25DE93F8D09E7D2AD8C3D9D33958B334CF03
B9695C7E1C15E752D4DEAD3CDFE770FCF1C85E15639355D7D6A6C1014837B87B56C70EA62EF1B20C
1E6E7C8B86E60433C94DBB735AEA4C6D1D6FFB4ACE0D333972ACA5AD8B0FF9482D0779AEA79AB2D7
E2BE891F893770F6740E16AC4BF2CC55A5551F4730F9977F56D1DB372B074722DEE7C8465B45C55B
7D905ED631D519DC3DB169BB5377B6E06D871BDFA2A139C14C72D3AE5B79B1CC14535BA7B4FD63DA
0D333972ACA57D8B7778A496B349882419A9E8A0210D0D9C2A3C3E45BB4B1D9F4E513EC74BBD59B6
FA38B23239EBC65981D5A318BF2F918DB68AAA4B3577AFC5866419DC3DBDC9F15D52123E1B83BA7C
4A5778BF54F3055FFCB39E69A93395DE0F5ED8B37DD793D99DF094A6157D6F1CD49272CD53CA5C49
F2E0E391ED163F1B2C7C259FF58FBCDF656BF7DD3CCFB60EE673B0CECA82A5C7B1B86C45AB8F1B72
B06CD1C7237BC50FB42ED5ACA349BF8CEB9B567C15EC6EBD9240E9ED4E4CB8E2AA46F2A95BE1FD52
CD17BCE88F174B6833959E0B1BB1BB7E5DD38ABE37DE2F923509EBA9A674BB22D5B32DFEB654B0F0
C5ACB27659DF7A37CFB3ADE3F9ECE6397E1C8BCB56F43965E617972DFA7864AFADDADF0FC0F10A07
7B2D56376830E1DB6E776EC20D23115FBFBA39CD17BCE88F174B6833959E0B1BB1BB7EDDDC167D6F
7C5C246512D6534DE976A4FFBB7F9C356CC72BC74FA721E1D2D61DE4B39B677080538E6365D9F43E
D78D7D70CD78EB221BFDEDE36FDA953E127547D36330E7E3A6D5DDEEC4841BA622BE7E75739A2F78
D11F2F96D0662A3D17366277FDBAB92DFADE5859243E09EBA9A6743B3DD5AD65CF1C2F1E2FB921E1
EAEEEDE6539767DD71ACFC71FAA0D64D7E459F2BF6DA2AFF4DBB7247A2EE5CDA0CA6DD330FD5D730
9EE1B1F8FA0DDF51151F6FFEE3C512DA4CA577DC9CADEFC9A75ED8DDF5EBE6B6E87B23380CB9BBFC
EDE3BB59A5A7BAB5EC99F4A6AD17D2907075F776F3D94AB2EEA2ADEFB5B8E6EE1548AC2565E5A28F
F75CAE377FBC7E2891C949BF53A506D3AE9887E3A34C4FB86130E2EB5737A7F98217FDF162096DA6
D2EB39DFA75ED8DDF5EBE6B6E87B637191BF7DF6BF3F5EF19572905562AAEB0B46A437EDE5C713AB
08CE796EF7B6F2D95AE120CF8AE388CF6A624352CE2EB8ECD6C77B2ED79B3F5EEF646472D2EF54A9
C1CC2BE6E1F82873132EBDB307F9D4ADF07EA9E60B5EF4C78B25B4994AEF072FECD9BEC12D0EBE3D
EAFE38584BA4BAE0A1C47BF23E0E523D38D903E94DDBAD65B7CCC5BFECE9DE7A3EBB2B1CE4193C8E
C882EFFF78B7A8D2833B1BC5B3A22255A45C8AF5B483A7BC58D40D06330FB66EB1F989631C9C8AF4
F6C6B7A86E4E3CB6D2DECAAABA2175A6B68E14BEDEBAEA3989E416F1718BC898D5FDF1592D291FAF
D86537A5C5B3E8A9A8A2F0C50CD76B3918E0F77F79BCECB178698979468E63BDA2781F52AA387376
5E67E544AA8837FF63B1BBA91E1FE85506330FB66EB1F989637C363675ED8D6F51DD9C786CA5BD9E
4C4343EA4C6D1D2C7CB175D573B2B5D1419776CB4F99B1F890A7D492FBF18A5D76FF78F1447A2AAA
287CF1E3EBB51CCCF0D95245DD4B292D31D5D2E3F89858F53A070D8927995E4830CFDD3F5ECC3C32
368B155D6230F960F716FB9F38C645F7E2587C977B9AB3D5B4BA83B8E770D7939CDA7777F7C5D67D
C59C1C28CD64EB53F162831D0B4E426497DD3F5E3C979E8A2A0A5FFCF87A2D07937CB65445EBB24A
4BCCB6F4383E66557DC4070D4949F2B827912AB23AB6987C646C162B4A5921EE9EAD8BBA9738C615
972222BED125CDD9ED5BC59AE90DA933B575B0F0C5D65D3227E76DDA2CBF6E6E53FE78AB96DC8F57
EC72F0C72B87D253D16251155BACD77230C3674BED0E6DBC03EBC924265C7A1C8B29C5D72C3DC1E3
DA0FDA12A9626BAFF77FBC927C646C162B4A5921EE9EAD8BBA9738C6E9372228BED70DCD39685DDD
41DC76C42B494EEDBBBBFB62EBBE624E0ED45D87BF2D9EF2C75BB5E47EBC6297833F5E39919E8A16
8BAA389AF55A0E06386B9D2CF1D212D3AE3B8ED29E9C5571E6ECBCCE9A13A9626BAF8F7FFC31FF9E
73293DD9B31C06B72EEA5EE21827DE8514F11D679B73DCC0C4A5EA1A52676AEB60E18BAD9B9D93F3
EE9C965F57E9562629B5E47EBC6297B33FFE78223D152D16D570341F7B129FDE78AF8EC54BCB4A7E
BD0F6D3D9C6D4830A5788B22556CEDF5F18F3FE61F1C9B8AA28ADCB3F56E028BDD4B1CE3945B9028
BEEF5473826D4C59A4BA2175A6B62EEAEDE29F55CFC9795F92CA4FAF742B93DC5A523E5EB1CBF11F
A78C657CF68E7789ACFFF2E3899728D8D574E317E4609DE61EEE6E517A940D772A65AFAD2644BABA
7EEEF109293DD9B31C06B74E99BDC5634A4C383E06C7E29B3634A7E2E37507317E1F574CA5D773BE
4FBDB08913BEF8F75B99D4D572FCF18A5D8EFF7877962A2A5A2CAAE16882BB2C6EBA7505128D5F90
8375461AB8BE4569260D772A65AFAD0E04BBBAB857FC5C4A07EC2C87C1AD535AB7784C89096FCD4C
AEF88E0DCDA9F878DD418CDFC71553E9F59CEFE0853DDB2598CC41F28B7FBCB572692D677575DEE5
483E6D1545F23C5E3C65E4CE8ADACA2751DD05A9BB68533D5C5CBF348D863B95B2D7560716FFF84D
632393535454917BB62E6A5D6287FFB654E90D3DC8A77385F74BF55FF0E0598CDFC71553E9B9B011
073B1635ADE87B2372C5D68B0A8E5CE48FDF8CCD62AAF1C13BDEE578F194913B2E2ADEB1DC7C222B
ECAE93721C5BF3796665F1D21C2AFA5CB1D7560782479FF2F1F4A28ADCB375BC6F5BB73B37E1D24B
7A904FDB0AEF97EABFE0BB577865E5F526F4984ACF858D38D8AEA86945DF1BC7F76BABA8E0C86DFD
FD56FE2B2BC4A7EE7897B395DF7C36F1EEBC596AABBD5952F60AA69D781CD50D6C1BFEAD048A3E5E
71A6C13F7E7F47821FCF2DAAC83D5B175DF0C40EBF592A3809E9F9B4ADF07EA9F5D5122F78FA955C
2CA1CD547AB9DF725B879B9E70FF853DD86BFD235B2B177D6FA40F43E474763F1B4C295E54A4AEB3
5AD6EB8A742C5ED4416E41291B551CF4F17194367071E5BA730C2EB8F5F1F4334DD9E87D6F831F3F
28AAE8622EE630B8F56202B9B73B31E1BA4B7A964FCF0AEF975A5F2DE5D27DFC83B32BB958429BA9
F412BFE22EBFB067DB1D24F366AFADF4E22B97D67250D1F12E679F8D671539A9F5D2563EB5BBF259
4A8917E7E352918E55E473BCC8716383C7119CBAB332B31AF2EF2B1CD47E5642BCDEC50FA6FCFDC7
DE063F9E5857857BB63E6ED7C10115251CB9A115F934ACF07EA9F5D5522EDDC73F389BB1F526F498
4A2FF7C266ED1249B8F3C2A6F4E16F1F69EE70622D07D7FC63B6D517FF6339BB27152CEDAC96E312
0E3A162C2AD2B18A7C22EBEC1E5CD6711C2C1BBC08070D59F9D441EDEFC5FB1CF9D4EE47CE124BF9
784A9E15EED9FAA051C7A75397F04186C7E2BB5437677DB5F48B737610E3F771C5547ACFBBB06719
2626F3729783AC9A3B9C55CBD97D3CA82EF7E2AF1415BC1D5BCD3C4EAFE86852EADADAEE4D9E07B2
B6483C88B3E358CF273DFFBA869CD51E6C5D7A15C7BB04B36A286AB7FF71F76CDD3C0C076BAEFC59
E7C9C6EBAA6ECE7AC95B1F5FFCE38383A898C074535B1F3767AB5DD533B9BE5D66EFF2BAB7BE786E
2615B57C6C72E246EFF74AF9FBC86125D675F9D11C37FF38D55D895BD49DF241CE29E712CCB6A20F
C1F3DAFD78CF99267EA4BFAEADFEC7DDB37551BB12978DFC59457BE375553767BDE4AD8F2FFEF1C1
D8140D61AEA9AD8F9BB39570F54C6EED98D3B8585D2BF9C4FF786BEB9E49E8DC6BF72315E795555A
CFD124AE1CF9B3F584D7E5AE3F781CBBC964655EDA90E3DA83AD4B2C21B24530B186BA76FB1F77CF
D645BD1AB92C3D871BAFABBA39EBF56E7D3C9ED5D9DFF71CEB47535B1F37672BDBEA99DCDD31A171
B1BA5632D9FAFBF8EE3D93D0B9D7C1A72A8E2CA5BA9EA3495C7C3D8794893A2B2D7DC186E3D84D26
25EDEA86446A0FB62EAB84C8FAC1ACAAEB3AE87FDC3D5BDF765F16973AD834D4B570699DCD592F76
EBE3297FBC5545CA286699DAFAB8395BA956CFE4C1A6A1AE85EB5AC961EB23F1047A26A173BBC44F
054F2D5E5DCFB9246EB19549E25C6D6D51B166E289AC7F767DBB47CE4FE2C74BCFF4F8B375E35434
0971F76CFDB0CBD270C4F1BAAA9B13BF35757FFCF2230D031937B5F57173B692AC9EC9837D4F9A95
57D74A0259952EE6D03309293B46D68F67555ADADF76E93994C413D94AA962BA22F9C4574EA9A5A2
87D523B4BBC5DF7609E613F9784AFE5B2B07535A4FE0A0BAC824440C26D0D3A2B35D5E6EB49B4FF5
41C7EBAA6ECE7AA55B1FDFDD6BA42175A6B6EE694BF54C9EED7B50484A5D91C5EB3299BA23753B1E
7FBC3481A2F158CF8A22A5D72457F5A5FEB851E216A5BEE84C77B5CD404A7AD3E95C970FDCE69E3B
724F268CEB1C869BFF498D7B767500897C49C27771670100BC88E0BBB8B300005E44F05DDC59E06F
FE3F3356AE6C
>
EI % End Image
Q  % Graphic state restoration
}}%(end)

### 2024-11-05 BAI/paper/src/__main__.py

@@ -1,48 +1,0 @@
-from torch import cuda, device
-from torch.utils.data import random_split, DataLoader
-
-from data import FlickrDataset
-from models import ImageCaption, ResnetImageEncoder, GRUCaptionDecoder
-from eval import train, evaluate
-
-# DIR = "/content/drive/MyDrive"
-DIR = "."
-CAPTION_LIMIT = None
-DEVICE = device("cuda" if cuda.is_available() else "cpu")
-BATCH_SIZE = 800 if cuda.is_available() else 400
-NUM_WORKERS = 2 if cuda.is_available() else 0
-PIN_MEMORY = cuda.is_available()
-
-# --- Data ---
-dataset = FlickrDataset(path=f"{DIR}/flickr8k", caption_limit=CAPTION_LIMIT)
-train_dataset, eval_dataset = random_split(dataset, [.8, .2])
-
-
-# --- Model ---
-encoder = ResnetImageEncoder()
-decoder = GRUCaptionDecoder(vocabulary_size=len(dataset.vocabulary))
-model = ImageCaption(encoder=encoder, decoder=decoder)
-
-
-# --- Training ---
-
-train(model,
-      data=DataLoader(train_dataset,
-                      batch_size=BATCH_SIZE,
-                      num_workers=NUM_WORKERS,
-                      shuffle=True,
-                      pin_memory=PIN_MEMORY),
-      epochs=5,
-      device=DEVICE)
-
-# --- Evaluation ---
-bleu = evaluate(model,
-                data=DataLoader(eval_dataset,
-                                batch_size=BATCH_SIZE,
-                                num_workers=NUM_WORKERS,
-                                shuffle=True,
-                                pin_memory=PIN_MEMORY),
-                dataset=dataset,
-                device=DEVICE)
-
-print(f"BLEU score: {bleu}")

### 2024-11-05 BAI/paper/src/assemble.sh

@@ -1,22 +1,0 @@
-DIR=${1:-"/workspaces/w24/bai/paper/src"}
-FILE=$DIR/${2:-"__assembled__"}.py
-
-# 'compile' all python files to the model.py file
-if [ -f $FILE ]; then
-    rm $FILE
-fi
-
-cat $DIR/data/vocabulary.py >> $FILE
-cat $DIR/data/flickrdataset.py >> $FILE
-cat $DIR/data/*.py >> $FILE
-cat $DIR/models/*.py >> $FILE
-cat $DIR/eval/*.py >> $FILE
-cat $DIR/__main__.py >> $FILE
-
-# remove all relative imports
-sed -i 's/from \..* import .*//g' $FILE
-sed -i 's/from data import .*//g' $FILE
-sed -i 's/from models import .*//g' $FILE
-sed -i 's/from eval import .*//g' $FILE
-
-printf "Assembled python files to $FILE\n"

### 2024-11-05 BAI/paper/src/data/__init__.py

@@ -1,2 +1,0 @@
-from .flickrdataset import FlickrDataset
-from .vocabulary import Vocabulary

### 2024-11-05 BAI/paper/src/data/flickrdataset.py

@@ -1,101 +1,0 @@
-import os
-from torch import tensor, Tensor, zeros, float32
-from torch.utils.data import Dataset
-from torchvision.transforms import Compose, Resize, ToTensor, Normalize
-from PIL import Image
-from pickle import dump, load
-from tqdm import tqdm
-from pandas import read_csv
-
-
-from .vocabulary import Vocabulary
-
-
-class FlickrDataset(Dataset):
-    def __init__(self,
-                 path: str,
-                 caption_limit: int = None,
-                 images_folder: str = "Images",
-                 captions_file: str = "captions.csv",
-                 data_file: str = "preprocessed.pkl",
-                 image_transform: Compose = None,
-                 vocabulary_threshold: int = 5,
-                 len_captions: int = 40):
-        self.image_path = os.path.join(path, images_folder)
-        self.captions_file = os.path.join(path, captions_file)
-        self.data_path = os.path.join(path, data_file)
-        self.caption_limit = caption_limit
-        self.len_captions = len_captions
-        self.image_transform = image_transform or Compose([
-            Resize((224, 224)),
-            ToTensor(),
-            Normalize(mean=[0.485, 0.456, 0.406], std=[0.229, 0.224, 0.225])
-        ])
-
-        if os.path.exists(self.data_path) and caption_limit is None:
-            print("Loading preprocessed data...")
-            self.vocabulary, self.images, self.captions = self._load()
-        else:
-            self.vocabulary, self.images, self.captions = self._process(
-                vocabulary_threshold)
-            print("Saving preprocessed data...")
-            self._save(self.vocabulary, self.images, self.captions)
-
-    def _save(self, vocabulary: Vocabulary, images: Tensor, captions: Tensor):
-        with open(self.data_path, 'wb') as f:
-            dump({'images': images,
-                  'captions': captions,
-                  'vocabulary': vocabulary}, f)
-
-    def _load(self) -> tuple[Vocabulary, Tensor, Tensor]:
-        with open(self.data_path, 'rb') as f:
-            data = load(f)
-            return (data['vocabulary'],
-                    data['images'],
-                    data['captions'])
-
-    def _process(self, vocabulary_threshold: int) -> tuple[Vocabulary, Tensor, Tensor]:
-        data = read_csv(self.captions_file, nrows=self.caption_limit)
-        self.vocabulary = Vocabulary(
-            data['caption'].tolist(), threshold=vocabulary_threshold)
-
[…]

### 2024-11-05 BAI/paper/src/data/vocabulary.py

@@ -1,78 +1,0 @@
-from collections import Counter
-from importlib import import_module
-from spacy import Language
-from spacy.cli import download
-from tqdm import tqdm
-
-
-class Vocabulary:
-    """ Vocabulary object to store the mapping between words and indices. """
-
-    pad_token = "<PAD>"
-    sos_token = "<SOS>"
-    eos_token = "<EOS>"
-    unk_token = "<UNK>"
-    pad_index = 0
-    sos_index = 1
-    eos_index = 2
-    unk_index = 3
-
-    def __init__(self, sentences: list[str], threshold: int, spacy_model_name: str = "en_core_web_sm"):
-        """ Initialize the vocabulary and learn the words from the sentences.
-
-        Args:
-            threshold: The minimum frequency for a word to be included in the vocabulary.
-            sentences: A list of sentences to learn the vocabulary from.
-            spacy_model_name: The name of the spaCy model to use for tokenization.
-        """
-        self.index_to_token = {self.pad_index: self.pad_token,
-                               self.sos_index: self.sos_token,
-                               self.eos_index: self.eos_token,
-                               self.unk_index: self.unk_token}
-        self.threshold = threshold
-        self.spacy_model = self._download_and_init(spacy_model_name)
-
-        freq = Counter()
-
-        # Use tqdm to provide a progress bar for vocabulary creation
-        for sentence in tqdm(sentences, desc="Building Vocabulary", unit="sentences"):
-            for token in self.tokenize(sentence):
-                freq[token] += 1
-
-                if token not in self.index_to_token.values():
-                    self.index_to_token[len(self.index_to_token)] = token
-
-    def __len__(self) -> int:
-        return len(self.index_to_token)
-
-    @property
-    def token_to_index(self) -> dict[str, int]:
-        return {v: k for k, v in self.index_to_token.items()}
-
-    def numericalize(self, text: str) -> list[int]:
-        """ Convert the text to numerical form using the vocabulary. """
-        return [self.token_to_index[token]
-                if token in self.token_to_index else self.token_to_index[self.unk_token]
-                for token in self.tokenize(text)]
-
-    def denumericalize(self, indices: list[int]) -> str:
-        """ Convert the indices back to text using the vocabulary. """
-        return " ".join([self.index_to_token[index] for index in indices])
-
-    def tokenize(self, text: str) -> list[str]:
-        """ Tokenize the text using spacy model. """
-        return [token.text.lower() for token in self.spacy_model.tokenizer(text)]
-
[…]

### 2024-11-05 BAI/paper/src/eval/__init__.py

@@ -1,3 +1,0 @@
-from .train import train
-from .evaluate import evaluate
-from .visualise import visualise

### 2024-11-05 BAI/paper/src/eval/evaluate.py

@@ -1,42 +1,0 @@
-from torch import no_grad, device, Tensor
-from torch.utils.data import DataLoader
-from nltk.translate.bleu_score import corpus_bleu
-from tqdm import tqdm
-
-from models import ImageCaption
-from data import FlickrDataset
-
-
-def evaluate(model: ImageCaption,
-             data: DataLoader,
-             dataset: FlickrDataset,
-             device: device = device("cpu")) -> float:
-    """
-    Evaluate the ImageCaption model on the provided dataset using BLEU score.
-
-    Args:
-        model (ImageCaption): The image captioning model to evaluate.
-        data (DataLoader): DataLoader providing images and reference captions.
-        device (torch.device): Device to run evaluation on (default: CPU).
-
-    Returns:
-        float: The BLEU score for the evaluated dataset.
-    """
-    model.to(device).eval()
-
-    references, candidates = [], []
-    with no_grad():
-        bleu = 0.0
-        for images, captions in tqdm(data, desc="Evaluation", unit="batch"):
-            images: Tensor = images.to(device)
-            predictions: Tensor = model(images)
-
-            for index, prediction in enumerate(predictions):
-                references.append([dataset.vocabulary.tokenize(
-                    dataset.tensor_to_caption(captions[index]))])
-                candidates.append(dataset.vocabulary.tokenize(
-                    dataset.tensor_to_caption(prediction.argmax(dim=1))))
-
-            bleu = corpus_bleu(references, candidates)
-
-    return bleu

### 2024-11-05 BAI/paper/src/eval/train.py

@@ -1,56 +1,0 @@
-from torch import device, Tensor
-from torch.optim import Adam, Optimizer
-from torch.nn import CrossEntropyLoss
-from torch.utils.data import DataLoader
-from tqdm import tqdm, trange
-
-from models import ImageCaption
-from data import Vocabulary
-
-
-def train(model: ImageCaption,
-          data: DataLoader,
-          epochs: int = 5,
-          device: device = device("cpu"),
-          optimizer: Optimizer = None,
-          criterion: CrossEntropyLoss = None):
-    """
-    Train the ImageCaption model.
-
-    Args:
-        model (ImageCaption): The caption generation model to train.
-        data (DataLoader): Dataloader providing the training data.
-        epochs (int): Number of epochs to train for (default: 5).
-        device (device): Device to use for training (default: CPU).
-        optimizer (Optimizer): Optimizer for training (default: Adam).
-        criterion (CrossEntropyLoss): Loss function (default: CrossEntropyLoss).
-
-    Returns:
-        ImageCaption: The trained model with the latest checkpoint.
-    """
-    model.to(device).train()
-
-    optimizer = Adam(model.parameters(),
-                     lr=1e-3) if optimizer is None else optimizer
-    criterion = CrossEntropyLoss(ignore_index=Vocabulary.pad_index
-                                 ) if criterion is None else criterion
-
-    epoch_loss = 0.0
-    for epoch in trange(model.load_checkpoint()+1, epochs, desc="Training", unit="epoch"):
-        batch_loss = 0.0
-        for images, captions in tqdm(data, desc=f"Epoch", unit="batch"):
-            optimizer.zero_grad()
-
-            images: Tensor = images.to(device)
-            captions: Tensor = captions.to(device)
-
-            outputs: Tensor = model(images, max_len=captions.size(1))
-            loss: Tensor = criterion(
-                outputs.view(-1, outputs.size(-1)), captions.view(-1))
-
-            loss.backward()
-            batch_loss = loss.item()
-            epoch_loss += batch_loss
-
-            optimizer.step()
-        model.save_checkpoint(epoch)

### 2024-11-05 BAI/paper/src/eval/visualise.py

@@ -1,19 +1,0 @@
-from PIL.Image import Image
-from matplotlib import pyplot as plt
-from matplotlib import rcParams
-
-
-def visualise(image: Image, caption: str):
-    """
-    Display an image and its caption.
-
-    Args:
-        image (Image): The image to display.
-        caption (str): The caption to display.
-    """
-    assert image.mode == "RGB", "Image should be in RGB format."
-
-    plt.imshow(image)
-    plt.title(caption, wrap=True)
-    plt.axis("off")
-    plt.show()

### 2024-11-05 BAI/paper/src/models/__init__.py

@@ -1,3 +1,0 @@
-from .image_caption import ImageCaption
-from .encoder import ResnetImageEncoder
-from .decoder import GRUCaptionDecoder

### 2024-11-05 BAI/paper/src/models/checkpoint.py

@@ -1,32 +1,0 @@
-from torch import save, load
-from torch.nn import Module
-
-
-class CheckpointModule(Module):
-    """ A module that can save and load checkpoints. """
-
-    def __init__(self, path: str):
-        super().__init__()
-        self.path = path
-
-    def save_checkpoint(self, checkpoint: int):
-        """ Save the model at a checkpoint. """
-        save({
-            "checkpoint": checkpoint,
-            "state": self.state_dict()
-        }, self.path)
-
-    def load_checkpoint(self) -> int:
-        """ Load the model from the last checkpoint. """
-        try:
-            checkpoint = load(self.path, weights_only=True)
-            model_dict = self.state_dict()
-            pretrained_dict = {k: v for k, v in checkpoint['state'].items()
-                               if k in model_dict and model_dict[k].shape == v.shape}
-            model_dict.update(pretrained_dict)
-
-            self.load_state_dict(model_dict)
-
-            return checkpoint['checkpoint']
-        except FileNotFoundError:
-            return 1

### 2024-11-05 BAI/paper/src/models/decoder.py

@@ -1,72 +1,0 @@
-from torch import Tensor, tensor, full
-from torch.nn import Module, Embedding, GRUCell, Linear
-from torch import float32
-
-from data import Vocabulary
-
-
-class GRUCaptionDecoder(Module):
-    """A simplified GRU-based language model for generating image captions."""
-
-    def __init__(self,
-                 vocabulary_size: int,
-                 embedding_size: int = 256,
-                 hidden_size: int = 512):
-        """
-        Args:
-            vocabulary_size: The number of tokens in the vocabulary.
-            embedding_size: The size of the token embeddings.
-            hidden_size: The size of the hidden state of the GRU.
-        """
-        super().__init__()
-        self.embedding_layer = Embedding(vocabulary_size, embedding_size)
-        self.feature_projection = Linear(2048, hidden_size)
-        self.gru_cell = GRUCell(embedding_size, hidden_size)
-        self.output_layer = Linear(hidden_size, vocabulary_size)
-
-    def forward(self, features: Tensor, max_len: int) -> Tensor:
-        """ Generate captions for the given image features.
-
-        Args:
-            features: Image features extracted from the encoder of shape (batch_size, num_features, feature_size).
-            max_len: Maximum length of the generated captions.
-
-        Returns:
-            A tensor of shape (batch_size, max_len, vocabulary_size) containing logits for each token position.
-        """
-        assert features.dim() == 3, "Input features must have 3 dimensions."
-        assert features.size(1) == 49, "Number of features must be 49."
-        assert features.size(2) == 2048, "Feature size must be 2048."
-        assert max_len > 0, "max_len must be a positive integer."
-
-        batch_size = features.size(0)
-        hidden = self.feature_projection(features.mean(dim=1))
-        embeddings = self.embedding_layer(
-            tensor(Vocabulary.sos_index, device=features.device).repeat(batch_size))
-
-        logits: Tensor = full((batch_size, max_len, self.output_layer.out_features),
-                              fill_value=Vocabulary.pad_index,
-                              device=features.device,
-                              dtype=float32)
-
-        for _ in range(max_len):
-            hidden = self.gru_cell(embeddings, hidden)
-            output: Tensor = self.output_layer(hidden)
-
-            logits[:, _, :] = output
-
-            # Select predicted token IDs for next embedding
-            predicted_ids = output.argmax(dim=1)
-            embeddings = self.embedding_layer(predicted_ids)
-
[…]

### 2024-11-05 BAI/paper/src/models/encoder.py

@@ -1,49 +1,0 @@
-from torch import Tensor
-from torch.nn import Module, Sequential
-from torchvision.models import resnet50, ResNet
-
-
-class ResnetImageEncoder(Module):
-    """ Image encoder using a pretrained ResNet-50 model. """
-
-    def __init__(self):
-        super().__init__()
-        self.dimension = 2048
-        self.model = self._get_passive_resnet()
-
-    def forward(self, images: Tensor) -> Tensor:
-        """ Extract features from a batch of image tensors.
-
-        Args:
-            images (Tensor): Batch of image tensors of shape (batch_size, channels=3, height=224, width=224).
-
-        Returns:
-            Tensor: Encoded image features of shape (batch_size, num_features=49, feature_size=2048).
-        """
-        assert images.dim() == 4, "Input images must be 4-dimensional."
-        assert images.size(1) == 3, "Input images must have 3 channels."
-        assert images.size(2) == 224, "Images must have size 224x224."
-
-        features: Tensor = self.model(images)  # (batch_size, 2048, 7, 7)
-        features = features.permute(0, 2, 3, 1)  # (batch_size, 7, 7, 2048)
-        features = features.view(features.size(0), -1, features.size(-1))
-
-        assert features.dim() == 3, "Output must have 3 dimensions."
-        assert features.size(0) == images.size(0), \
-            "Output batch size must match input batch size."
-        assert features.size(1) == 49, "Output must have 49 features."
-        assert features.size(2) == self.dimension, \
-            "Output feature size must be 2048."
-
-        return features
-
-    def _get_passive_resnet(self) -> Sequential:
-        """ Prepare the ResNet-50 model by setting it to evaluation mode, freezing parameters, and removing the final layers to retain spatial information in the output.
-
-        Returns:
-            Sequential: Modified ResNet-50 model as a Sequential module.
-        """
-        resnet: ResNet = resnet50(weights='ResNet50_Weights.DEFAULT').eval()
-        for param in resnet.parameters():
-            param.requires_grad_(False)
-        return Sequential(*list(resnet.children())[:-2])

### 2024-11-05 BAI/paper/src/models/image_caption.py

@@ -1,37 +1,0 @@
-from torch import Tensor, no_grad
-
-from .checkpoint import CheckpointModule
-from .encoder import ResnetImageEncoder
-from .decoder import GRUCaptionDecoder
-
-
-class ImageCaption(CheckpointModule):
-    def __init__(self, encoder: ResnetImageEncoder, decoder: GRUCaptionDecoder):
-        super().__init__("imagecaption.checkpoint")
-        self.encoder = encoder
-        self.decoder = decoder
-
-    def forward(self, images: Tensor, max_len: int = 15) -> Tensor:
-        """
-        Generate captions for a batch of images.
-
-        Args:
-            images: A tensor of shape (batch_size, channels=3, height=224, width=224) containing the images.
-            max_len: The maximum length of the captions to generate.
-
-        Returns:
-            A tensor of shape (batch_size, max_len, vocabulary_size) containing the logits for each token position.
-        """
-        assert images.dim() == 4, "Images must have 4 dimensions."
-        assert images.size(1) == 3, "Images must have 3 channels."
-        assert images.size(2) == 224 and images.size(
-            3) == 224, "Images must have size 224x224."
-
-        features = self.encoder(images)
-        logits: Tensor = self.decoder(features, max_len)
-
-        assert logits.dim() == 3, "Logits must have 3 dimensions (batch_size, max_len, vocabulary_size)."
-        assert logits.size(0) == images.size(0), "Batch size must match."
-        assert logits.size(1) == max_len, "Caption length must match."
-
-        return logits

### 2024-11-05 BAI/paper/paper.tex
BAI paper: drafted methods chapter and started approach.

--- 2024-10-26 BAI/paper/paper.tex
@@
 
+\begin{center}
+    \includegraphics[width=0.2\textwidth]{src/wtmicon.pdf}
+\end{center}
+
 \begin{abstract}
- This paper is about image captioning. We discuss the current state of the art and overview the most important methods. We also present some of our results.
+ This paper is about image captioning.
+We discuss the current state of the art and overview the most important methods.
+We present our results of experiments with different encoder and decoder models.
+We use the flickr8k dataset and evaluate the models using the BLEU score.
+We discuss the results and give an outlook on future work.
 \end{abstract}
@@
 
-The accuracy of image captioning is commonly measured using BLEU or CIDEr scores.
-BLEU is a precision-based metric that compares n-grams in the generated caption to those in the reference captions.
-CIDEr is a recall-based metric that compares the generated caption to the reference captions using cosine similarity.
+We present the dataset, the encoder-decoder architecture, the encoder models, the decoder models, and the evaluation metrics.
 
+\subsection{Dataset}
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=0.8\textwidth]{src/flickr8k.png}
+    \caption{Example images and captions from the flickr8k dataset.}
+    \label{fig:flickr8k}
+\end{figure}
 We focused on the flickr8k dataset, which consists of 8,000 images, each with five reference captions.
-Other commonly used datasets are flickr30k and MS COCO, but due to their size and limited resources, we decided to focus on flickr8k.
+The images are diverse and contain a wide range of objects and scenes.
+This dataset is commonly used for image captioning research due to its quality and limited size.
+The captions are relatively short and straightforward, making them easier to generate.
+\par Other commonly used datasets are flickr30k and MS COCO, but these datasets are larger and cannot be processed due to our limited computational resources.
 
+\subsection{Encoder-Decoder Architecture}
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=0.8\textwidth]{src/encoder-decoder.png}
+    \caption{A Survey on Various Deep Learning Models for Automatic Image Captioning \cite{gaurav}}
+    \label{fig:encoder-decoder}
+\end{figure}
 Usually, an encoder-decoder architecture is used to achieve image captioning.
@@
 
-We experiment with different architectures, such as VGG16, ResNet, and InceptionV3 as encoders and LSTM and GRU as decoders.
-We also experiment with varying attention mechanisms, such as soft and hard attention.
+\subsection{Encoder Models}
+We experimented with two different models for encoding images: ResNet-50 and VGG-16.
+\par ResNet-50 is a deep residual network that is widely used for image classification.
+It has 50 layers and uses skip connections to avoid the vanishing gradient problem.
+It was introduced in \cite{he}.
+\par VGG-16 is a deep convolutional neural network that was developed by the Visual Geometry Group at the University of Oxford.
+It has 16 layers and is known for its simplicity and effectiveness.
+It was introduced in \cite{simonyan}.
+
+\subsection{Decoder Models}
+We experimented with two different models for decoding captions: GRU and LSTM.
+\par Gated Recurrent Units (GRUs) are a type of RNN that is similar to LSTMs but has fewer parameters. They are easier to train and faster to compute. They were introduced in \cite{cho}.
+\par Long Short-Term Memory (LSTM) networks are a type of RNN that is designed to avoid the vanishing gradient problem.
+They have a cell state that can store information over long sequences.
+They were introduced in \cite{hochreiter}.
+
+\subsection{Evaluation Metrics}
+The accuracy of image captioning is commonly measured using BLEU or CIDEr scores.
+\par BLEU is a precision-based metric that compares n-grams in the generated caption to those in the reference captions.
+It ranges from 0 to 1, with higher values indicating better performance.
+It was introduced in \cite{papineni}.
+\par CIDEr is a recall-based metric that compares the generated caption to the reference captions using cosine similarity.
+It ranges from 0 to 1, with higher values indicating better performance.
+It was introduced in \cite{vedantam}.
+\par For simplicity, we are only using the BLEU score in our experiments.
 
@@
 
-We follow the state-of-the-art approach to image captioning.
-We preprocess the images using the encoder and tokenize the captions.
-We then train the model on the training set and evaluate it on the validation set.
-We generate the captions using beam search.
+Our approach begins with data preprocessing, followed by model training and evaluation.
+
+\subsection{Data Preprocessing}
+\par We preprocess the images by resizing them to 224x224 pixels and normalizing the pixel values and encoding them to tensors of the dimension (224, 224, 3).
+\par We create a vocabulary of words and map each word to an index.
+\par We preprocess the captions by tokenizing them, converting them to lowercase, and padding them to a fixed length.
+Then we encode them using the vocabulary and convert them to one-dimensional tensors with a specified max\_length.
+
+\subsection{Model Training}
+
+\subsection{Model Evaluation}
+
+\section{Results}
 

### 2024-11-06 BAI/paper/paper.tex
Added some more bai paper content.

@@
 %\usepackage[outer=2.5cm]{geometry}     % Type area (size, margins...) !!!Print version
-%\usepackage{geometry}                  % Type area (size, margins...) !!!Proofread version
+%\usepackage{geometry}                  % Type area (size, margins) !!!Proofread version
 \usepackage[outer=3.15cm]{geometry}     % Type area (size, margins...) !!!Draft version
@@
 \begin{center}
-    \includegraphics[width=0.2\textwidth]{src/wtmicon.pdf}
+    \includegraphics[width=0.2\textwidth]{res/wtmicon.pdf}
 \end{center}
@@
 We discuss the current state of the art and overview the most important methods.
-We present our results of experiments with different encoder and decoder models.
+We present the results of our experiments with different encoder and decoder models.
 We use the flickr8k dataset and evaluate the models using the BLEU score.
@@
 Image captioning involves generating textual descriptions of given images.
-It is a difficult problem because it combines the visual and textual domains.
+It is a complex problem because it combines the visual and textual domains.
 In recent years, deep learning has revolutionized the field of image captioning.
 
-This paper presents an overview of the most essential methods in image captioning and some of our results.
+\subsection{Related Work}
 
 Image captioning is a complex and interdisciplinary task bridging computer vision (CV) and natural language processing (NLP).
-The objective is to generate natural, descriptive sentences for images, which involves detecting and interpreting visual elements and their relationships within scenes.
-Recent advancements leverage deep learning, particularly Convolutional Neural Networks (CNNs) for feature extraction and Recurrent Neural Networks (RNNs), especially Long Short-Term Memory (LSTM) networks, for sequential language generation.
+Generating natural, descriptive sentences for images involves detecting and interpreting visual elements and their relationships within depicted scenes.
+Recent advancements leverage deep learning, particularly Convolutional Neural Networks (CNNs) for feature extraction and Recurrent Neural Networks (RNNs) for sequential language generation.
 For instance, models such as "Show and Tell" \cite{Vinyals:2015} and "Show, Attend, and Tell" \cite{Xu} utilize CNN-RNN architectures, with the latter integrating attention mechanisms to enhance focus on specific image regions, thereby improving caption relevance and accuracy.
 
-Image captioning models are typically evaluated using metrics like BLEU, METEOR, and CIDEr, which assess linguistic similarity to human references.
-While these metrics are widely accepted, considering the quality of image captions remains challenging due to the subjective nature of language generation.
 Additionally, methods like Reinforcement Learning (RL) have been explored to optimize these models directly on evaluation metrics, addressing issues such as exposure bias by aligning training and inference methods more closely \cite{Rennie}.
 
-Recent innovations in diverse decoding strategies, like Diverse Beam Search, aim to overcome the tendency of beam search to produce near-duplicate captions by enhancing variability without sacrificing quality.
+Another approach has dealt with the tendency of beam search to produce near-duplicate captions.
 Diverse Beam Search ensures that generated captions capture a broader range of plausible descriptions, reflecting the inherent ambiguity in image captioning tasks \cite{Vijayakumar}.
 
+\subsection{This Paper}
+
+We begin by discussing the most critical methods in image captioning.
+We then present the results of our experiments with different encoder and decoder models.
+We use the flickr8k dataset and evaluate the models using the BLEU score.
+We discuss the results and give an outlook on future work.
+
+\newpage
 \section{Methods}
@@
 
-We present the dataset, the encoder-decoder architecture, the encoder models, the decoder models, and the evaluation metrics.
+In image captioning tasks, a model takes an image as input and generates a caption as output.
+The model consists of two parts: an encoder that extracts features from the image and a decoder that generates the caption.
+We discuss the dataset, encoder-decoder architecture, encoder models, decoder models, and evaluation metrics.
 
@@
     \centering
-    \includegraphics[width=0.8\textwidth]{src/flickr8k.png}
+    \includegraphics[width=0.7\textwidth]{res/flickr8k.png}
     \caption{Example images and captions from the flickr8k dataset.}
@@
 \end{figure}
-We focused on the flickr8k dataset, which consists of 8,000 images, each with five reference captions.
-The images are diverse and contain a wide range of objects and scenes.
-This dataset is commonly used for image captioning research due to its quality and limited size.
-The captions are relatively short and straightforward, making them easier to generate.
-\par Other commonly used datasets are flickr30k and MS COCO, but these datasets are larger and cannot be processed due to our limited computational resources.
+We focused on the flickr8k dataset, a pivotal resource in the field of image captioning.
+Comprising 8.000 diverse images, each with five reference captions, this dataset is renowned for its quality and is a common choice for image captioning research due to its manageable size.
+\par Other commonly used datasets are flickr30k and MS COCO, but these datasets are more extensive and cannot be processed due to our limited computational resources.
 
@@
     \centering
-    \includegraphics[width=0.8\textwidth]{src/encoder-decoder.png}
+    \includegraphics[width=0.7\textwidth]{res/encoder-decoder.png}
     \caption{A Survey on Various Deep Learning Models for Automatic Image Captioning \cite{gaurav}}
@@
 We experimented with two different models for encoding images: ResNet-50 and VGG-16.
-\par ResNet-50 is a deep residual network that is widely used for image classification.
-It has 50 layers and uses skip connections to avoid the vanishing gradient problem.
-It was introduced in \cite{he}.
-\par VGG-16 is a deep convolutional neural network that was developed by the Visual Geometry Group at the University of Oxford.
-It has 16 layers and is known for its simplicity and effectiveness.
-It was introduced in \cite{simonyan}.
+\par ResNet-50 is a deep residual network widely used for image classification. It has 50 layers and uses skip connections to avoid the vanishing gradient problem. It was introduced in \cite{he}.
+VGG-16, a deep convolutional neural network developed by the Visual Geometry Group at the University of Oxford, is known for its simplicity and effectiveness, making it an accessible choice for our research.
+Thanks to their pre-training on the ImageNet dataset, the models do not need to be trained from scratch, saving valuable time and computational resources.
 
@@
 We experimented with two different models for decoding captions: GRU and LSTM.
-\par Gated Recurrent Units (GRUs) are a type of RNN that is similar to LSTMs but has fewer parameters. They are easier to train and faster to compute. They were introduced in \cite{cho}.
-\par Long Short-Term Memory (LSTM) networks are a type of RNN that is designed to avoid the vanishing gradient problem.
+\par Gated Recurrent Units (GRUs) are a type of RNN similar to LSTMs but with fewer parameters. They are easier to train and faster to compute. They were introduced in \cite{cho}.
+\par Long Short-Term Memory (LSTM) networks are a type of RNN designed to avoid the vanishing gradient problem.
 They have a cell state that can store information over long sequences.
@@
 \par CIDEr is a recall-based metric that compares the generated caption to the reference captions using cosine similarity.
-It ranges from 0 to 1, with higher values indicating better performance.
+Ranging from 0 to 1, higher values indicate better performance.
 It was introduced in \cite{vedantam}.
-\par For simplicity, we are only using the BLEU score in our experiments.
+\par For simplicity, we only use the BLEU score in our experiments.
 
+\newpage
 \section{Approach}
@@
 \subsection{Data Preprocessing}
-\par We preprocess the images by resizing them to 224x224 pixels and normalizing the pixel values and encoding them to tensors of the dimension (224, 224, 3).
+\par We preprocess the images by resizing them to 224x224 pixels, normalizing the pixel values, and encoding them to tensors of the dimension (224, 224, 3).
 \par We create a vocabulary of words and map each word to an index.
 \par We preprocess the captions by tokenizing them, converting them to lowercase, and padding them to a fixed length.
-Then we encode them using the vocabulary and convert them to one-dimensional tensors with a specified max\_length.
+Then, we encode them using the vocabulary and convert them to one-dimensional tensors with a specified max\_length.
 
 \subsection{Model Training}
+We only train the decoder model because the encoder models are pre-trained on ImageNet.
+We use the Adam optimizer and the categorical cross-entropy loss function.
+We train our model for the specified number of epochs and save the weights with the best BLEU score on the validation set.
+The model's state is saved after each epoch to resume training if necessary.
 
 \subsection{Model Evaluation}
+Evaluation happens after each training epoch.
+This approach enables us to monitor the model's performance and save the weights with the best BLEU score.
+We evaluate the model on the test set using the BLEU score and compare it to the validation set's BLEU score to detect overfitting.
+These scores are also stored after each epoch to analyze the model's performance over time.
 
-\section{Results}
-
+\newpage
 \section{Results}
@@
 
-Comparison of different models, evaluation, and discussion.
+We present the results of our experiments with different encoder and decoder models.
+
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=0.8\textwidth]{res/results.png}
+    \caption{Results of the experiments with different encoder and decoder models.}
+    \label{fig:results}
+\end{figure}
+
+% TODO: add results
+The results show that the ResNet-50 encoder and LSTM decoder achieve the best performance with a BLEU score 0.75.
+The VGG-16 encoder and LSTM decoder achieve a BLEU score of 0.72.
+The ResNet-50 encoder and GRU decoder achieve a BLEU score of 0.70.
+The VGG-16 encoder and GRU decoder achieve a BLEU score of 0.68.
 
 \section{Conclusion}
-\label{sec:concl}
+\label{sec:conclusion}
 
-Summary of the results, future work, and open questions.
+In this paper, we discuss the current state of the art in image captioning and present our experiments with different encoder and decoder models.
+We use the flickr8k dataset and evaluate the models using the BLEU score.
+The results show that the ResNet-50 encoder and LSTM decoder achieve the best performance with a BLEU score 0.75.
+
+In the future, further experiments could be conducted with
+\begin{enumerate}
+    \item different encoder and decoder models, like attention mechanisms and transformer models.
+    \item different datasets, like flickr30k and MS COCO.
+    \item different evaluation metrics, like CIDEr.
+\end{enumerate}
 

### 2024-11-13 BAI/paper/paper.tex
Completed BAI paper.

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\title{\trtitle\\[0.3cm]{\normalsize\trtype}}
+\title{\trtitle\\[0.2cm]{\normalsize\trtype}}
 \author{\trauthor}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Content (Outline_Lemmermann_Leopold_TITLE)
+% Content
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 In recent years, deep learning has revolutionized the field of image captioning.
-
-\subsection{Related Work}
 
@@
 
-\subsection{This Paper}
-
 We begin by discussing the most critical methods in image captioning.
@@
 
-\newpage
 \section{Methods}
@@
 In image captioning tasks, a model takes an image as input and generates a caption as output.
-The model consists of two parts: an encoder that extracts features from the image and a decoder that generates the caption.
+The model consists of an encoder that extracts features from the image and a decoder that generates the caption.
 We discuss the dataset, encoder-decoder architecture, encoder models, decoder models, and evaluation metrics.
@@
 \end{figure}
-We focused on the flickr8k dataset, a pivotal resource in the field of image captioning.
+We focused on the flickr8k dataset, a pivotal resource in image captioning.
 Comprising 8.000 diverse images, each with five reference captions, this dataset is renowned for its quality and is a common choice for image captioning research due to its manageable size.
@@
     \centering
-    \includegraphics[width=0.7\textwidth]{res/encoder-decoder.png}
+    \includegraphics[width=.5\textwidth]{res/encoder-decoder.png}
     \caption{A Survey on Various Deep Learning Models for Automatic Image Captioning \cite{gaurav}}
@@
 
-\newpage
 \section{Approach}
@@
 
-\newpage
 \section{Results}
@@
 We present the results of our experiments with different encoder and decoder models.
+\ref{fig:results} shows the BLEU scores of the models on the test set over the epochs.
+The detailed BLEU score results are shown in \ref{tab:results}.
 
@@
 
-% TODO: add results
-The results show that the ResNet-50 encoder and LSTM decoder achieve the best performance with a BLEU score 0.75.
-The VGG-16 encoder and LSTM decoder achieve a BLEU score of 0.72.
-The ResNet-50 encoder and GRU decoder achieve a BLEU score of 0.70.
-The VGG-16 encoder and GRU decoder achieve a BLEU score of 0.68.
+\begin{table}[H]
+    \begin{tabular}
+ { | l | l | l | l | l | }
+        \hline
+ Epoch & ResNet-GRU & VGG-GRU & ResNet-LSTM & VGG-LSTM \\
+        \hline
+ 1 & 0.0074 & 0.0025 & 0.0035 & 0.0157 \\
+ 2 & 0.0028 & 0.0024 & 0.0050 & 0.0056 \\
+ 3 & 0.0024 & 0.0037 & 0.0045 & 0.0055 \\
+ 4 & 0.0036 & 0.0030 & 0.0034 & 0.0052 \\
+ 5 & 0.0038 & 0.0025 & 0.0035 & 0.0062 \\
+ 6 & 0.0056 & 0.0026 & 0.0037 & 0.0069 \\
+ 7 & 0.0058 & 0.0028 & 0.0038 & 0.0060 \\
+ 8 & 0.0069 & 0.0031 & 0.0037 & 0.0058 \\
+ 9 & 0.0045 & 0.0034 & 0.0030 & 0.0058 \\
+ 10 & 0.0037 & 0.0029 & 0.0034 & 0.0052 \\
+        \hline
+    \end{tabular}
+
+    \caption{Detailed results of the experiments with different encoder and decoder models.}
+    \label{tab:results}
+\end{table}
+
+The results show that the VGG encoder with the LSTM decoder achieves the best performance, with a BLEU score of .157 in the first epoch.
+Other combinations, like the VGG-GRU or the ResNet-LSTM, are significantly worse, with respective BLEU scores of .0025 and .0035 in the first epoch.
+The ResNet-GRU model performs slightly better, with a BLEU score of .0074 in the first epoch.
+
+For most models, their BLEU score peaks in the first few epochs and then decreases.
+The VGG-LSTM model is the most stable over the epochs and somewhat contradicts this hypothesis.
+However, the other models show a similar decrease over the epochs.
+These findings suggest overfitting to the training data.
+
+Given the mixed nature of the results, it is challenging to draw definitive conclusions.
+The VGG encoder paired with the LSTM decoder is the most effective combination available., achieving the highest BLEU score in the first epoch.
+Nevertheless, further experiments are needed to confirm this result.
 
@@
 We use the flickr8k dataset and evaluate the models using the BLEU score.
-The results show that the ResNet-50 encoder and LSTM decoder achieve the best performance with a BLEU score 0.75.
 
@@
 
-\nocite{*}
 \bibliographystyle{apalike}

### 2024-11-13 SbD/summary.tex
Added basic layout  for pbd summary.
BAI WIP.
Added new NLP practice.

@@
 
-\section{Kryptographie}
+\part{Kryptographie}
 
 \includepdf[pages={14-16}, nup=1x3]{slides.pdf}
-\subsection{Grundlagen}
-\subsubsection{Angriffsarten (schwer zu leicht)}
+\section{Grundlagen}
+\subsection{Angriffsarten (schwer zu leicht)}
 \begin{enumerate}
@@
 
-\subsubsection{Brechen (stark zu schwach)}
+\subsection{Brechen (stark zu schwach)}
 \begin{enumerate}
@@
 \includepdf[pages={18-32,35,37-41}, nup=3x7]{slides.pdf}
-\subsection{Klassische Chiffren}
+\section{Klassische Chiffren}
 \begin{itemize}
@@
 
+\section{Moderne Kryptographie}
 \includepdf[pages={44-108}, nup=4x9]{slides.pdf}
-\subsection{Moderene Kryptographie I: Symmetrische Systeme}
+\subsection{Symmetrische Systeme}
 \begin{enumerate}
@@
 \includepdf[pages={44, 109-142}, nup=4x9]{slides.pdf}
-\subsection{Moderne Kryptographie II: Asymmetrische Systeme}
+\subsection{Asymmetrische Systeme}
 \begin{enumerate}
@@
 \begin{itemize}
-  \item \textbf{Modulo}: $a \mod b = a - b \cdot \lfloor \frac{a}{b} \rfloor$
-  \item \textbf{Erweiterter Euklidischer Algorithmus}: $ggT(a, b) = a \cdot x + b \cdot y$
-  \item \textbf{Euler-Phi-Funktion}: $\varphi(n) = |\{a \in \mathbb{N} \mid 1 \leq a < n, ggT(a, n) = 1\}|$
+  \item \textbf{Erweiterter Euklidischer Algorithmus}: Berechnung des ggT und inversen Elemente in Restklassenringen.
+  \item \textbf{Euler-Phi-Funktion}: $\varphi(n)$ gibt die Anzahl der zu $n$ teilerfremden Zahlen kleiner als $n$ an.
   \item \textbf{Primitive Wurzel}: $g$ ist primitive Wurzel modulo $p$, wenn $g^i \mod p \neq 1$ für $i < p-1$
-  \item \textbf{Diskreter Logarithmus}: $a = g^x \mod p \Rightarrow x = \log_g a \mod p$
-  \item \textbf{Faktorisierungsannahme}: $n = p \cdot q \Rightarrow p, q \in \mathbb{P}$
+  \item \textbf{Diskreter Logarithmus}: Exponent in der Primitiven Wurzel
+  \item \textbf{Faktorisierungsannahme}: Primfaktoren sind nicht leicht zu berechnen (auch nicht nachgewiesen schwer)
 \end{itemize}
@@
 
-\section{Public Key Infrastructure}
-\includepdf[pages={1}, nup=3x7]{slides.pdf}
+\part{Public Key Infrastructure}
+\includepdf[pages={146-148}, nup=3x7]{slides.pdf}
 
-\section{Attacks and Defenses}
-\includepdf[pages={1}, nup=3x7]{slides.pdf}
+\section{Zertifikate}
 
-\section{Mobile Security}
-\includepdf[pages={1}, nup=3x7]{slides.pdf}
+\subsection{Zertifizierungsmodelle}
+\begin{enumerate}
+  \item \textbf{Web of Trust}: Benutzer zertifizieren andere Benutzer (zB. PGP, GnuPG)
+  \item \textbf{Hierarchisches Modell}: Zertifizierungsstellen zertifizieren Benutzer (zB. TLS, S/MIME)
+\end{enumerate}
+
+\subsecton{X.509}
+
+Zertifizierungsstandard, zB. genutzt bei TLS und S/MIME.
+
+TLS-Handshake: Client sendet ClientHello, Server sendet ServerHello, Server-Zertifikat, ServerKeyExchange, ServerHelloDone, ClientKeyExchange, ChangeCipherSpec, Finished, ServerChangeCipherSpec, Finished
+\includepdf[pages={149-205}, nup=3x7]{slides.pdf}
+
+\part{Attacks and Defenses}
+
+\section{Sniffing}
+
+\section{Spoofing}
+
+\section{Denial of Service}
+
+\part{Mobile Security}
 

### 2024-11-27 SbD/summary.tex
completed sbd summary (first draft).

@@
 
+\includepdf[pages={14-16}, nup=1x3]{slides.pdf}
 \part{Kryptographie}
 
-\includepdf[pages={14-16}, nup=1x3]{slides.pdf}
 \section{Grundlagen}
@@
 
+\includepdf[pages={44-108}, nup=4x9]{slides.pdf}
 \section{Moderne Kryptographie}
-\includepdf[pages={44-108}, nup=4x9]{slides.pdf}
 \subsection{Symmetrische Systeme}
@@
 
+\includepdf[pages={144, 146-148, 151-176}, nup=4x9]{slides.pdf}
 \part{Public Key Infrastructure}
-\includepdf[pages={146-148}, nup=3x7]{slides.pdf}
 
@@
 \begin{enumerate}
-  \item \textbf{Web of Trust}: Benutzer zertifizieren andere Benutzer (zB. PGP, GnuPG)
-  \item \textbf{Hierarchisches Modell}: Zertifizierungsstellen zertifizieren Benutzer (zB. TLS, S/MIME)
+  \item \textbf{Web of Trust}: Zertifikate basieren auf einer dezentralen Struktur.
+Benutzer können andere Benutzer direkt zertifizieren, wie es bei Pretty Good Privacy (PGP) und GNU Privacy Guard (GnuPG) der Fall ist.
+Vertrauen entsteht durch ein Netzwerk von Beglaubigungen.
+  \item \textbf{Hierarchisches Modell}: Zertifizierungsstellen (Certification Authorities, CAs) übernehmen die Aufgabe der Zertifizierung.
+Diese Struktur wird beispielsweise in der Transport Layer Security (TLS) und bei S/MIME genutzt.
+Eine Root-CA kann Zwischenzertifikate ausstellen, um die Struktur zu skalieren.
 \end{enumerate}
 
-\subsecton{X.509}
+\subsection{X.509}
 
-Zertifizierungsstandard, zB. genutzt bei TLS und S/MIME.
+X.509 ist ein internationaler Standard für digitale Zertifikate.
+Es definiert eine Zertifikatstruktur, die in der Public-Key-Infrastruktur (PKI) für die Identifizierung von Entitäten (Personen, Organisationen oder Servern) verwendet wird.
+Beispiele für den Einsatz sind:
 
-TLS-Handshake: Client sendet ClientHello, Server sendet ServerHello, Server-Zertifikat, ServerKeyExchange, ServerHelloDone, ClientKeyExchange, ChangeCipherSpec, Finished, ServerChangeCipherSpec, Finished
-\includepdf[pages={149-205}, nup=3x7]{slides.pdf}
+\begin{itemize}
+  \item \textbf{TLS}: Wird verwendet, um die Kommunikation zwischen Webbrowsern und Servern abzusichern (HTTPS).
+  \item \textbf{S/MIME}: Für verschlüsselte und signierte E-Mails.
+\end{itemize}
 
-\part{Attacks and Defenses}
+\subsubsection{TLS-Handshake Ablauf}
+\begin{enumerate}
+  \item \textbf{ClientHello}: Der Client sendet unterstützte Cipher Suites und Protokollversionen.
+  \item \textbf{ServerHello}: Der Server wählt die Parameter und sendet seine Auswahl zurück.
+  \item \textbf{Server-Zertifikat}: Der Server stellt sein Zertifikat vor, das durch eine vertrauenswürdige CA signiert wurde.
+  \item \textbf{ServerKeyExchange}: (Falls erforderlich) Der Server sendet zusätzliche kryptografische Parameter.
+  \item \textbf{ServerHelloDone}: Der Server signalisiert das Ende seiner Nachricht.
+  \item \textbf{ClientKeyExchange}: Der Client übermittelt Informationen für die Schlüsselerzeugung (z.B. Premaster-Secret).
+  \item \textbf{ChangeCipherSpec}: Der Client aktiviert die vereinbarte Verschlüsselung.
+  \item \textbf{Finished}: Abschlussnachricht des Clients.
+  \item \textbf{ServerChangeCipherSpec}: Der Server aktiviert die Verschlüsselung.
+  \item \textbf{Finished}: Der Server beendet den Handshake.
+\end{enumerate}
 
-\section{Sniffing}
+Diese Handshake-Schritte gewährleisten Authentizität, Integrität und Vertraulichkeit.
 
+\subsection{Man-in-the-Middle (MITM) Attacken auf HTTPS}
+
+\subsubsection{Angriffsbeschreibung}
+Eine MITM-Attacke versucht, die Kommunikation zwischen Client und Server abzufangen oder zu manipulieren.
+Angreifer nutzen Schwächen in HTTPS-Implementierungen, um sensible Daten wie Passwörter oder Kreditkartendetails zu erlangen.
+
+\subsubsection{Angriffsmethoden}
+\begin{itemize}
+\item \textbf{TLS-Stripping}: Umschalten von HTTPS auf HTTP durch Manipulation der Verbindung.
+\item \textbf{DNS-Spoofing}: Weiterleitung des Clients zu einem gefälschten Server.
+\item \textbf{Gefälschte Zertifikate}: Präsentation eines nicht vertrauenswürdigen Zertifikats.
+\end{itemize}
+
+\subsubsection{Abwehrmaßnahmen}
+\begin{itemize}
+\item \textbf{HSTS}: Erzwingt die Nutzung von HTTPS, verhindert TLS-Stripping.
+\item \textbf{Zertifikatsprüfung}: Validierung der Signatur und des Hostnamens.
+\item \textbf{Certificate Transparency}: Aufdeckung gefälschter Zertifikate durch öffentliche Logs.
+\item \textbf{OCSP}: Echtzeitüberprüfung des Zertifikatsstatus.
+\end{itemize}
+
+MITM-Angriffe sind durch schlechte HTTPS-Konfiguration und unzureichende Zertifikatsvalidierung möglich.
+Gegenmaßnahmen wie HSTS und moderne Protokolle wie TLS 1.3 können die Sicherheit signifikant erhöhen.
+
+\includepdf[pages={178-180, 182-189, 191-196}, nup=3x7]{slides.pdf}
+\section{Beispiele}
+
+\subsection{Elektronische Reisepässe (ePassports)}
+\begin{itemize}
+    \item \textbf{Grundlagen:} Nutzung von RFID-Chips für kontaktlose Kommunikation.
+    \item \textbf{Biometrische Daten:} Gespeichert im ePassport, Zugang geschützt durch Mechanismen wie Basic Access Control (BAC) und Active Authentication (AA).
+    \item \textbf{Sicherheitsmechanismen:}
+    \begin{itemize}
+        \item Public Key Infrastructure (PKI) für Zertifikate.
+        \item Authentifizierung durch asymmetrische Kryptographie.
+        \item Schutz der Kommunikationskanäle gegen Abhören und Replay-Angriffe.
+    \end{itemize}
+\end{itemize}
+
+\subsection{Smart Meter Gateways (SMG)}
+\begin{itemize}
+    \item \textbf{Aufbau:} Kommunikationsinfrastruktur mit Smart Meter Gateway (SMGW) und Local Metrological Network (LMN).
+    \item \textbf{Zertifikatsbasierte Sicherheit:} Nutzung einer PKI für sichere Kommunikation und Datenintegrität.
+    \item \textbf{Funktionalität:}
+    \begin{itemize}
+        \item Verschlüsselte Datenübertragung zwischen SMGW und externer Infrastruktur.
+        \item Schutz der Privatsphäre durch Anonymisierung.
+    \end{itemize}
+\end{itemize}
+
+\subsection{Sichere Signaturkomponenten}
+\begin{itemize}
+    \item \textbf{Standard-PC:} Unsicher ohne spezielle Hardware (Angriffsvektoren wie Malware und Keylogger).
+    \item \textbf{Sichere Komponenten:} Einsatz von Smartcards, Trusted Platform Modules (TPM) und dedizierten Hardware-Sicherheitsmodulen.
+    \item \textbf{Signaturerstellung:} Sichere Eingabe und Verarbeitung der PIN auf vertrauenswürdigen Geräten, die gegen Manipulation geschützt sind.
+\end{itemize}
+
+\includepdf[pages={198-204}, nup=2x5]{slides.pdf}
+
+\section{Details}
+
+\subsection{Zeitstempel und Digitale Signaturen}
+\begin{itemize}
+    \item Problem: Rückwirkend gültige Signaturen könnten trotz kompromittierten Signaturschlüssels gefälscht werden.
+    \item Lösung: Nachrichten erhalten Zeitstempel $t$ (Prüfprotokoll der Nachricht $h(M)$).
+    \item Anforderungen:
+    \begin{itemize}
+        \item Zeitstempel $t$ ist nachweislich korrekt und später nicht manipulierbar.
+        \item Nachrichten und Zeitstempel werden gemeinsam signiert: $sig_s(h(M), t)$.
+    \end{itemize}
+\end{itemize}
+
+\subsection{Selbstzertifizierung öffentlicher Verschlüsselungsschlüssel}
+\begin{itemize}
+    \item Annahme: Funktionierende Infrastruktur für digitale Signaturen existiert.
+    \item Ablauf:
+    \begin{enumerate}
+        \item Teilnehmer A signiert seinen öffentlichen Schlüssel $k_u$ und sendet diesen an die CA.
+        \item Die CA überprüft die Signatur und erstellt ein Zertifikat.
+        \item Das Zertifikat wird an Teilnehmer A zurückgesendet und an Teilnehmer B verteilt.
+    \end{enumerate}
+    \item Vorteil: Keine Fremdzertifizierung notwendig.
+\end{itemize}
+
+\subsection{Gültigkeitsmodelle für Zertifikate und Signaturen}
+\begin{itemize}
+    \item \textbf{Schalenmodell:}
+    \begin{itemize}
+        \item Beim Ablauf oder Widerruf eines Zertifikats werden abgeleitete Zertifikate ungültig.
+        \item Signaturen bleiben gültig, wenn das Zertifikat zum Signaturzeitpunkt gültig war.
+    \end{itemize}
+    \item \textbf{Kettenmodell:}
+    \begin{itemize}
+        \item Abgeleitete Zertifikate bleiben gültig, auch bei Ablauf oder Widerruf.
+        \item Rekursive Anwendung auf alle zugrunde liegenden Zertifikate.
+    \end{itemize}
+    \item \textbf{Praxis:} Kombination von Schalen- und Kettenmodell je nach Anforderung.
+\end{itemize}
+
+\includepdf[pages={206-213, 215-223, 225-228}, nup=3x7]{slides.pdf}
+\section{Attacks and Defenses}
+
+\subsection{Angriffsarten und Vorgehen}
+\begin{itemize}
+    \item \textbf{Sniffing-Angriffe (Ethernet):}
+    \begin{itemize}
+        \item Angriffe basieren auf Abhören von Datenströmen (z.B. ARP-Spoofing, DNS-Spoofing).
+        \item Ziel: Vertraulichkeit der Kommunikation durch Einblick in übertragenen Datenverkehr verletzen.
+    \end{itemize}
+    \item \textbf{Sniffing-Angriff: Ablauf:}
+    \begin{itemize}
+        \item Analyse der Netzwerkinfrastruktur.
+        \item Manipulation von Netzwerkprotokollen (z.B. ARP-Caches).
+        \item Weiterleitung oder Speicherung der abgefangenen Pakete.
+    \end{itemize}
+\end{itemize}
+
+\subsection{Verteidigungsmechanismen}
+\begin{itemize}
+    \item \textbf{Technische Maßnahmen:}
+    \begin{itemize}
+        \item Verschlüsselung von Datenströmen (z.B. HTTPS, TLS).
+        \item Einsatz physischer Schutzmaßnahmen (z.B. abgeschirmte Netzwerkkabel).
+        \item Einsatz sicherer Authentifizierungsverfahren (z.B. Zertifikate, Token).
+    \end{itemize}
+    \item \textbf{Organisatorische Maßnahmen:}
+    \begin{itemize}
+        \item Regelmäßige Netzwerkanalysen auf verdächtige Aktivitäten.
+        \item Schulungen für Mitarbeiter zur Erkennung und Abwehr von Angriffen.
+    \end{itemize}
+\end{itemize}
+
+\includepdf[pages={230-255}, nup=4x9]{slides.pdf}
 \section{Spoofing}
 
-\section{Denial of Service}
+\subsection{Angriffsarten und Mechanismen}
+\begin{itemize}
+    \item \textbf{DNS-Spoofing:}
+    \begin{itemize}
+        \item Manipulation von DNS-Anfragen und -Antworten, um Benutzer auf falsche Server umzuleiten.
+        \item Ziel: Vertrauliche Daten abfangen oder schädliche Inhalte verbreiten.
+        \item Beispiele:
+        \begin{itemize}
+            \item DNS-Spoofing durch gefälschte Antworten.
+            \item Poisoning von DNS-Caches.
+        \end{itemize}
+    \end{itemize}
+    \item \textbf{ARP-Spoofing:}
+    \begin{itemize}
+        \item Fälschen von ARP-Antworten, um sich als legitime Netzwerkgeräte auszugeben.
+        \item Ziel: Datenverkehr abfangen, umleiten oder manipulieren.
+    \end{itemize}
+\end{itemize}
 
+\subsection{Vorgehen bei Angriffen}
+\begin{itemize}
+    \item \textbf{DNS-Spoofing:}
+    \begin{enumerate}
+        \item Analyse der DNS-Kommunikation.
+        \item Injektion gefälschter DNS-Antworten mit falschen IP-Adressen.
+    \end{enumerate}
+    \item \textbf{ARP-Spoofing:}
+    \begin{enumerate}
+        \item Senden gefälschter ARP-Antworten an Zielsysteme.
+        \item Umleitung des Datenverkehrs über den Angreifer.
+    \end{enumerate}
+\end{itemize}
+
+\subsection{Abwehrmaßnahmen}
+\begin{itemize}
+    \item \textbf{Gegen DNS-Spoofing:}
+    \begin{itemize}
+        \item Einsatz von DNSSEC für signierte DNS-Antworten.
+        \item Verwendung von sicheren und vertrauenswürdigen DNS-Servern.
+    \end{itemize}
+    \item \textbf{Gegen ARP-Spoofing:}
+    \begin{itemize}
+        \item Statische ARP-Einträge für kritische Systeme.
+        \item Netzwerksegmentierung zur Minimierung der Angriffsfläche.
+        \item Tools wie ARP-Watch zur Überwachung des ARP-Verkehrs.
+    \end{itemize}
+\end{itemize}
+
+\includepdf[pages={257-272}, nup=3x7]{slides.pdf}
+\section{Denial of Service (DoS)}
+
+\subsection{Verfügbarkeitsangriffe}
+\begin{itemize}
+    \item Ziel: Verhinderung des Zugriffs auf Systeme und Dienste durch Überlastung von Ressourcen.
+    \item \textbf{Typen von DoS-Angriffen:}
+    \begin{itemize}
+        \item \textbf{DoS:} Angriff von einem einzelnen System.
+        \item \textbf{Distributed DoS (DDoS):} Angriff von mehreren, oft kompromittierten Systemen (Botnetze).
+    \end{itemize}
+\end{itemize}
+
+\subsection{Beispiele für DoS-Angriffe}
+\begin{itemize}
+    \item \textbf{Smurf-Attack:}
+    \begin{itemize}
+        \item Verstärkung durch Missbrauch von Broadcast-Adressen.
+        \item Effekt: Überflutung des Ziels mit ICMP-Echo-Antworten.
+    \end{itemize}
+    \item \textbf{Mirai-Botnet (2016):}
+    \begin{itemize}
+        \item Missbrauch von IoT-Geräten mit Standardpasswörtern.
+        \item Weltweite Ausfälle durch Überlastung von DNS-Servern.
+    \end{itemize}
+    \item \textbf{Memcached-Angriff (2017):}
+    \begin{itemize}
+        \item Verstärkung durch offene Memcached-Server.
+        \item Ziel: Massive Erhöhung des Datenverkehrs (Amplification-Attacke).
+    \end{itemize}
+\end{itemize}
+
+\subsection{IoT-Sicherheitsprobleme}
+\begin{itemize}
+    \item Schwächen:
+    \begin{itemize}
+        \item Unverschlüsselte Kommunikation und schwache Passwörter.
+        \item Unsichere Standardkonfigurationen (z.B. Universal Plug and Play).
+    \end{itemize}
+    \item Angriffsvektoren:
+    \begin{itemize}
+        \item Nutzung als Teil eines Botnets (z.B. für DDoS).
+        \item Angriffe über Over-the-Air (OTA)-Updates.
+    \end{itemize}
+\end{itemize}
+
+\subsection{Abwehrmaßnahmen}
+\begin{itemize}
+    \item Einsatz von Firewalls und Traffic-Filtern.
+    \item Deaktivierung unsicherer Protokolle und Ports (z.B. UPnP).
+    \item Überwachung und Erkennung von Anomalien im Netzwerkverkehr.
+    \item Sicherstellung sicherer Passwörter und Verschlüsselung.
+\end{itemize}
+
+\includepdf[pages={274-282, 284}, nup=2x5]{slides.pdf}
 \part{Mobile Security}
+\section{Grundlagen}
+
+\subsection{Herausforderungen mobiler Netzwerke}
+\begin{itemize}
+    \item \textbf{Unterschiede zu festen Netzwerken:}
+    \begin{itemize}
+        \item Begrenzte Bandbreite.
+        \item Verbindungsabbrüche und Fehleranfälligkeit (z.B. Paketverluste).
+        \item Erhöhte Anfälligkeit für Angriffe (z.B. Abhören, Sniffing).
+    \end{itemize}
+    \item \textbf{Neue Bedrohungen:}
+    \begin{itemize}
+        \item Abhören drahtloser Kommunikation.
+        \item Standortermittlung und Bewegungsprofile.
+    \end{itemize}
+\end{itemize}
+
+\subsection{Sicherheitsdefizite bestehender mobiler Netzwerke}
+\begin{itemize}
+    \item \textbf{Probleme:}
+    \begin{itemize}
+        \item Fehlender Schutz vor Datenverlust und Identitätsdiebstahl.
+        \item Schwächen in der Benutzer-Authentifizierung und Verschlüsselung.
+        \item Mangelhafte Fehlerbehandlung in der Signalverarbeitung.
+    \end{itemize}
+    \item \textbf{Angriffspunkte:}
+    \begin{itemize}
+        \item Netzwerk: Angriffe auf Protokolle und Infrastruktur.
+        \item Schnittstelle: Angriffe auf Ortungsfunktionen und Nutzerdaten.
+    \end{itemize}
+\end{itemize}
+
+\subsection{Angriffsmodell}
+\begin{itemize}
+    \item \textbf{Akteure:}
+    \begin{itemize}
+        \item \textbf{Outsiders:} Abhören von Kommunikation.
+        \item \textbf{Insiders:} Manipulation von Daten und Identitäten.
+    \end{itemize}
+    \item \textbf{Angriffsziele:}
+    \begin{itemize}
+        \item Vertraulichkeit (Confidentiality).
+        \item Integrität (Integrity).
+        \item Verfügbarkeit (Availability).
+    \end{itemize}
+\end{itemize}
+
+\subsection{Beispiele mobiler Systeme}
+\begin{itemize}
+    \item \textbf{Sprachkommunikation:}
+    \begin{itemize}
+        \item 1. Generation: Analoge Systeme (z.B. NMT, AMPS).
+        \item 2. Generation: Digitale Standards (z.B. GSM, DECT).
+        \item 3. Generation: UMTS/IMT-2000.
+        \item 4. Generation: LTE.
+    \end{itemize}
+    \item \textbf{Satellitendienste:}
+    \begin{itemize}
+        \item Globale Positionierung (z.B. GPS, Galileo).
+        \item Daten- und Sprachkommunikation (z.B. Iridium, Inmarsat).
+    \end{itemize}
+\end{itemize}
+
+\subsection{Sensoren und Anwendungen}
+\begin{itemize}
+    \item \textbf{Typische Sensoren:}
+    \begin{itemize}
+        \item GPS, WLAN, NFC, Kamera, Mikrofon.
+        \item Beschleunigungssensoren, Gyroskope.
+        \item Fingerabdrucksensoren und Temperatursensoren.
+    \end{itemize}
+    \item \textbf{Beispiele für Anwendungen:}
+    \begin{itemize}
+        \item Fitness-Tracking.
+        \item Smart-Home-Integration.
+        \item Standortbasierte Dienste.
+    \end{itemize}
+\end{itemize}
+
+\includepdf[pages={286-362, 369-374}, nup=4x9]{slides.pdf}
+\section{Global Standard for Mobile Communication (GSM)}
+
+\subsection{Grundlagen und Architektur}
+\begin{itemize}
+    \item \textbf{GSM:} Digitales Mobilfunksystem für Sprach- und Datendienste.
+    \item \textbf{Netzarchitektur:}
+    \begin{itemize}
+        \item \textbf{Base Station Subsystem (BSS):} Verbindet mobile Geräte mit dem Netzwerk.
+        \item \textbf{Network Switching Subsystem (NSS):} Verarbeitet Anrufe und Standortinformationen.
+        \item \textbf{Operation and Support Subsystem (OSS):} Überwacht und verwaltet das Netzwerk.
+    \end{itemize}
+\end{itemize}
+
+\subsection{Sicherheitsfunktionen}
+\begin{itemize}
+    \item \textbf{Authentifizierung:}
+    \begin{itemize}
+        \item Nutzung eines gemeinsamen Schlüssels (Ki) zur Verifizierung des Benutzers.
+        \item Implementierung durch den Algorithmus A3.
+    \end{itemize}
+    \item \textbf{Verschlüsselung:}
+    \begin{itemize}
+        \item Schutz der Kommunikation durch A5-Algorithmen.
+        \item Ziel: Vertraulichkeit der Sprach- und Datenübertragung.
+    \end{itemize}
+    \item \textbf{Anonymität:}
+    \begin{itemize}
+        \item Nutzung temporärer Identifikatoren (TMSI), um die Identität der Benutzer zu verschleiern.
+    \end{itemize}
+\end{itemize}
+
+\subsection{Schwachstellen und Angriffe}
+\begin{itemize}
+    \item \textbf{Kryptografische Schwächen:}
+    \begin{itemize}
+        \item Veraltete Algorithmen (z.B. A5/1) können leicht entschlüsselt werden.
+    \end{itemize}
+    \item \textbf{IMSI-Catcher:}
+    \begin{itemize}
+        \item Geräte, die IMSI (International Mobile Subscriber Identity) abfangen, um Benutzer zu lokalisieren oder zu überwachen.
+    \end{itemize}
+    \item \textbf{Man-in-the-Middle-Angriffe:}
+    \begin{itemize}
+        \item Angreifer simulieren Basisstationen, um Kommunikation abzufangen.
+    \end{itemize}
+\end{itemize}
+
+\subsection{Moderne Erweiterungen}
+\begin{itemize}
+    \item \textbf{GPRS (General Packet Radio Service):}
+    \begin{itemize}
+        \item Erweiterung für datenorientierte Dienste.
+    \end{itemize}
+    \item \textbf{EDGE (Enhanced Data rates for GSM Evolution):}
+    \begin{itemize}
+        \item Höhere Datendurchsätze für Mobilfunkanwendungen.
+    \end{itemize}
+\end{itemize}
+
+\includepdf[pages={376-388}, nup=3x7]{slides.pdf}
+\section{Other Mobile Systems}
+
+\subsection{UMTS (Universal Mobile Telecommunications System, 3G)}
+\begin{itemize}
+    \item \textbf{Sicherheitsfunktionen:}
+    \begin{itemize}
+        \item Verbesserte Vertraulichkeit und Authentifizierung durch PIN-geschützte USIM.
+        \item Unterstützung für Ende-zu-Ende-Verschlüsselung.
+        \item Schutz gegen Angriffe wie IMSI-Catching.
+    \end{itemize}
+    \item \textbf{Architektur:}
+    \begin{itemize}
+        \item Verwendet Authentication Center (AuC) für die Generierung von Authentifizierungsvektoren.
+        \item Integriert Integrity Checking und MAC-Algorithmen für Datenintegrität.
+    \end{itemize}
+    \item \textbf{Schwachstellen:}
+    \begin{itemize}
+        \item Angriffe durch gefälschte Basisstationen möglich.
+        \item Sicherheitsmaßnahmen hängen von der Implementierung ab.
+    \end{itemize}
+\end{itemize}
+
+\subsection{LTE (Long Term Evolution, 4G)}
+\begin{itemize}
+    \item \textbf{Merkmale:}
+    \begin{itemize}
+        \item All-IP-basierte Architektur mit hoher Datenrate und niedriger Latenz.
+        \item Unterstützt AES (Advanced Encryption Standard) für verbesserte Sicherheit.
+    \end{itemize}
+    \item \textbf{Architektur:}
+    \begin{itemize}
+        \item Verbindung zwischen UE (User Equipment), eNodeB und Core Network.
+        \item Sicherheitsfunktionen beinhalten Transport Layer Security (TLS) und IPsec.
+    \end{itemize}
+    \item \textbf{Sicherheitsprobleme:}
+    \begin{itemize}
+        \item DNS- und DDoS-Angriffe durch unsichere Netzwerkschnittstellen.
+    \end{itemize}
+\end{itemize}
+
+\subsection{5G Mobile Communication}
+\begin{itemize}
+    \item \textbf{Charakteristika:}
+    \begin{itemize}
+        \item Ultraniedrige Latenz und hohe Zuverlässigkeit.
+        \item Unterstützung für massive IoT-Verbindungen und Smart Manufacturing.
+    \end{itemize}
+    \item \textbf{Sicherheitsverbesserungen:}
+    \begin{itemize}
+        \item Hybridverschlüsselung (symmetrisch und asymmetrisch).
+        \item Einführung der Subscription Concealed Identifier (SUCI) für besseren Datenschutz.
+    \end{itemize}
+\end{itemize}
+
+\subsection{Vergleich von Mobilfunksystemen}
+\begin{itemize}
+    \item \textbf{2G (GSM):} Basis für Sprach- und Textdienste.
+    \item \textbf{3G (UMTS):} Höhere Datenraten, verbesserte Sicherheit.
+    \item \textbf{4G (LTE):} Fokus auf IP-basierten Datenverkehr, niedrige Latenz.
+    \item \textbf{5G:} Zielt auf umfassende Konnektivität und niedrige Latenz ab.
+\end{itemize}
+
+\includepdf[pages={390-423}, nup=4x9]{slides.pdf}
+\section{Bluetooth}
+
+\subsection{Grundlagen}
+\begin{itemize}
+    \item \textbf{Bluetooth:} Kurzstrecken-Funkkommunikationstechnologie für Geräteverbindungen.
+    \item \textbf{Versionen:} Verbesserungen in Reichweite, Geschwindigkeit und Energieverbrauch mit jeder Version.
+    \item \textbf{Netzwerkstrukturen:}
+    \begin{itemize}
+        \item \textbf{Piconet:} Ein Master und bis zu sieben Slaves.
+        \item \textbf{Scatternet:} Mehrere Piconets, die miteinander verbunden sind.
+    \end{itemize}
+\end{itemize}
+
+\subsection{Sicherheitsmerkmale}
+\begin{itemize}
+    \item \textbf{Pairing:}
+    \begin{itemize}
+        \item Legacy Pairing (Bluetooth 2.0 und älter): Schwachstellen durch statische PINs.
+        \item Secure Simple Pairing (Bluetooth 2.1+): Nutzung von Elliptic Curve Diffie-Hellman (ECDH) zur Verbesserung der Sicherheit.
+    \end{itemize}
+    \item \textbf{Authentifizierung:} Bestätigung der Identität durch Passkeys oder numerische Vergleiche.
+    \item \textbf{Verschlüsselung:} 
+    \begin{itemize}
+        \item Symmetrische Verschlüsselung mit Stream Ciphers.
+        \item Schlüsselgrößen bis zu 128 Bit.
+    \end{itemize}
+    \item \textbf{Integrität:} Nutzung von HMAC für Datenintegrität.
+\end{itemize}
+
+\subsection{Schwachstellen und Angriffe}
+\begin{itemize}
+    \item \textbf{Angriffsvektoren:}
+    \begin{itemize}
+        \item Bluejacking: Senden unerwünschter Nachrichten.
+        \item Bluesnarfing: Unbefugter Zugriff auf Daten.
+        \item Man-in-the-Middle (MITM): Angriffe während des Pairing-Prozesses.
+    \end{itemize}
+    \item \textbf{Probleme:}
+    \begin{itemize}
+        \item Unsichere Legacy Pairing-Methoden.
+        \item Offene Verbindungen bei schlecht konfigurierten Geräten.
+    \end{itemize}
+\end{itemize}
+
+\subsection{Erweiterungen und Anwendungen}
+\begin{itemize}
+    \item \textbf{Bluetooth Low Energy (BLE):}
+    \begin{itemize}
+        \item Optimiert für geringe Energieverbräuche, z.B. in IoT-Geräten.
+        \item Sichere Verbindungen mit modernem Pairing (Just Works, Passkey Entry, Numeric Comparison).
+    \end{itemize}
+    \item \textbf{Typische Anwendungen:}
+    \begin{itemize}
+        \item Fitness-Tracker, drahtlose Kopfhörer, Smart-Home-Geräte.
+        \item Industrie 4.0: Sensornetzwerke und Echtzeitkommunikation.
+    \end{itemize}
+\end{itemize}
+
+\includepdf[pages={425-436, 438-446}, nup=3x7]{slides.pdf}
+\section{WLAN}
+
+\subsection{Grundlagen und Standards}
+\begin{itemize}
+    \item \textbf{IEEE 802.11:} Standard für drahtlose Netzwerke (Wi-Fi).
+    \item \textbf{Protokollfamilie:}
+    \begin{itemize}
+        \item IEEE 802.11a/b/g/n/ac/ax: Verbesserungen bei Geschwindigkeit, Frequenzen und Reichweite.
+        \item Unterstützte Frequenzen: 2.4 GHz und 5 GHz Bänder.
+    \end{itemize}
+    \item \textbf{Netzwerktypen:}
+    \begin{itemize}
+        \item Infrastrukturmodus: Kommunikation über Access Points.
+        \item Ad-hoc-Modus: Direkte Verbindung zwischen Geräten.
+    \end{itemize}
+\end{itemize}
+
+\subsection{Sicherheitsstandards}
+\begin{itemize}
+    \item \textbf{WEP (Wired Equivalent Privacy):}
+    \begin{itemize}
+        \item Verwendet statische Schlüssel (40- oder 104-Bit).
+        \item Schwächen:
+        \begin{itemize}
+            \item Schwache Verschlüsselung (RC4 Stream Cipher).
+            \item Schwächen in Integritätsprüfungen (CRC-32).
+            \item Anfällig für Replay- und Key-Recovery-Angriffe.
+        \end{itemize}
+    \end{itemize}
+    \item \textbf{WPA (Wi-Fi Protected Access):}
+    \begin{itemize}
+        \item Einführung von TKIP (Temporal Key Integrity Protocol) zur Verbesserung der Sicherheit.
+        \item Rückwärtskompatibilität zu WEP.
+    \end{itemize}
+    \item \textbf{WPA2:}
+    \begin{itemize}
+        \item Nutzung von AES-CCMP für starke Verschlüsselung.
+        \item Erfüllt moderne Sicherheitsanforderungen.
+    \end{itemize}
+    \item \textbf{WPA3:}
+    \begin{itemize}
+        \item Einführung von SAE (Simultaneous Authentication of Equals) für besseres Passwort-Management.
+        \item Verbesserte Schutzmaßnahmen gegen Offline-Angriffe.
+    \end{itemize}
+\end{itemize}
+
+\subsection{Schwachstellen und Angriffe}
+\begin{itemize}
+    \item \textbf{WEP:}
+    \begin{itemize}
+        \item Schwacher Initialisierungsvektor (IV) führt zu Key-Recovery-Angriffen.
+        \item Keine wirksame Authentifizierung für Clients.
+    \end{itemize}
+    \item \textbf{Generelle Angriffe:}
+    \begin{itemize}
+        \item Man-in-the-Middle-Angriffe.
+        \item Deauthentifizierungsangriffe.
+        \item Passwort-Cracking bei schwachen Passwörtern.
+    \end{itemize}
+\end{itemize}
+
+\subsection{Erweiterte Sicherheitsmechanismen}
+\begin{itemize}
+    \item \textbf{EAP (Extensible Authentication Protocol):}
+    \begin{itemize}
+        \item Flexible Authentifizierungsmethoden (z.B. TLS, PEAP).
+        \item Nutzung eines zentralen Authentifizierungsservers (z.B. RADIUS).
+    \end{itemize}
+    \item \textbf{802.11i (WPA2 und WPA3):}
+    \begin{itemize}
+        \item Fokus auf sichere Authentifizierung und Verschlüsselung.
+    \end{itemize}
+\end{itemize}
 

### 2024-11-27 PbD/summary.tex
First PbD summary.

--- 2024-10-30 PbD/summary.tex
@@
 \part{Privacy Enhancing Technology}
-% \includepdf[pages={?}, nup=?]{slides.pdf}
+\includepdf[pages={23-24, 27-39, 41-57, 59-124}, nup=4x9]{slides.pdf}
+\section{In Theory}
 
-\section{IPv6 traffic pseudonymisation}
+\subsection{Privacy Goals and Protection Models}
+\begin{itemize}
+    \item \textbf{Confidentiality:} Prevent unauthorized access to sensitive information.
+    \item \textbf{Integrity:} Protect against unauthorized modification of information.
+    \item \textbf{Availability:} Ensure authorized access to information when needed.
+\end{itemize}
 
+\subsection{Confidentiality Mechanisms}
+\begin{itemize}
+    \item \textbf{Encryption:}
+    \begin{itemize}
+        \item Symmetric (e.g., AES): Uses a shared key for both encryption and decryption.
+        \item Asymmetric (e.g., RSA, ECC): Employs public-private key pairs for secure communication.
+    \end{itemize}
+    \item \textbf{Anonymization:}
+    \begin{itemize}
+        \item Techniques such as k-anonymity, differential privacy, and pseudonymization to obscure user data.
+    \end{itemize}
+    \item \textbf{Decentralized Networks:}
+    \begin{itemize}
+        \item Leverage architectures like DC networks (Dining Cryptographers) and mix networks to ensure anonymity in communication.
+    \end{itemize}
+\end{itemize}
+
+\subsection{DC Networks (Dining Cryptographers)}
+\begin{itemize}
+    \item \textbf{Principle:}
+    \begin{itemize}
+        \item Participants (cryptographers) communicate in such a way that a message's sender remains anonymous unless they choose to reveal themselves.
+        \item Each pair of participants shares a secret bit, and all participants cooperate to hide the identity of the message sender.
+    \end{itemize}
+    \item \textbf{Use Cases:}
+    \begin{itemize}
+        \item Anonymous broadcasting, electronic voting.
+    \end{itemize}
+    \item \textbf{Advantages:}
+    \begin{itemize}
+        \item Provable anonymity.
+    \end{itemize}
+    \item \textbf{Limitations:}
+    \begin{itemize}
+        \item Scalability issues due to increasing communication overhead with more participants.
+    \end{itemize}
+\end{itemize}
+
+\subsection{Blind Messaging Services}
+\begin{itemize}
+    \item \textbf{Principle:}
+    \begin{itemize}
+        \item Users send messages through a series of anonymizing layers, where each layer hides specific metadata (e.g., sender's IP address).
+        \item Common implementations include Tor (The Onion Router), which routes traffic through multiple relays to anonymize users.
+    \end{itemize}
+    \item \textbf{Use Cases:}
+    \begin{itemize}
+        \item Anonymous browsing, secure whistleblowing, and bypassing censorship.
+    \end{itemize}
+    \item \textbf{Advantages:}
+    \begin{itemize}
+        \item Scalable and widely adopted.
+        \item Resilient against traffic analysis.
+    \end{itemize}
+    \item \textbf{Limitations:}
+    \begin{itemize}
+        \item Vulnerable to global adversaries monitoring the entire network.
+        \item Latency introduced due to multi-hop routing.
+    \end{itemize}
+\end{itemize}
+
+\subsection{Mix Networks}
+\begin{itemize}
+    \item \textbf{Principle:}
+    \begin{itemize}
+        \item Messages are shuffled and encrypted at each node (mix) in a network, ensuring unlinkability between sender and recipient.
+        \item Each mix node reorders, decrypts, and forwards messages in batches.
+    \end{itemize}
+    \item \textbf{Use Cases:}
+    \begin{itemize}
+        \item Email anonymization (e.g., Mixminion, Mixmaster), anonymous payments.
+    \end{itemize}
+    \item \textbf{Advantages:}
+    \begin{itemize}
+        \item Strong resistance to traffic analysis.
+        \item Can provide anonymity for both sender and recipient.
+    \end{itemize}
+    \item \textbf{Limitations:}
+    \begin{itemize}
+        \item Increased latency due to batching and processing.
+        \item Vulnerable to active attacks if a mix node is compromised.
+    \end{itemize}
+\end{itemize}
+
+\subsection{Integrity Mechanisms}
+\begin{itemize}
+    \item \textbf{Hash Functions:}
+    \begin{itemize}
+        \item Fixed-size digest for verifying data integrity (e.g., SHA-256).
+    \end{itemize}
+    \item \textbf{Digital Signatures:}
+    \begin{itemize}
+        \item Combines hashing with asymmetric encryption for authenticity and non-repudiation.
+    \end{itemize}
+\end{itemize}
+
+\subsection{Availability Mechanisms}
+\begin{itemize}
+    \item \textbf{Redundancy:}
+    \begin{itemize}
+        \item Use of backup servers and data replication.
+    \end{itemize}
+    \item \textbf{Load Balancing:}
+    \begin{itemize}
+        \item Distributes network traffic to prevent overload.
+    \end{itemize}
+    \item \textbf{DDoS Mitigation:}
+    \begin{itemize}
+        \item Techniques like rate limiting and anomaly detection to block malicious traffic.
+    \end{itemize}
+\end{itemize}
+
+\includepdf[pages={126-161}, nup=4x9]{slides.pdf}
 \section{In Practice}
 
-\section{Censorship Resistance}
+\subsection{Privacy Enhancing Technologies (PETs)}
+\begin{itemize}
+    \item \textbf{Objective:} Protect user anonymity and prevent unauthorized tracking in real-world applications.
+    \item \textbf{Examples of PETs:}
+    \begin{itemize}
+        \item \textbf{Anonymity Networks:} Tor, I2P for secure and anonymous communication.
+        \item \textbf{Encryption:} End-to-end encryption in messaging apps (e.g., Signal, WhatsApp).
+        \item \textbf{Pseudonymization:} Use of alternate identifiers to reduce linkage to real identities.
+    \end{itemize}
+\end{itemize}
 
-\section{Criminal Misuse and Data Retention}
+\subsection{Traffic Pseudonymization}
+\begin{itemize}
+    \item \textbf{IPv6 Privacy Extensions:} Randomize IPv6 addresses to prevent tracking over time.
+    \item \textbf{Pseudonym Rotation:} Periodically change pseudonyms to reduce linkability between user actions.
+    \item \textbf{Limitations:}
+    \begin{itemize}
+        \item Limited effectiveness in preventing advanced traffic analysis.
+        \item Increased complexity in managing pseudonyms.
+    \end{itemize}
+\end{itemize}
 
+\subsection{Anonymity Networks in Practice}
+\begin{itemize}
+    \item \textbf{Tor:}
+    \begin{itemize}
+        \item Relays user traffic through multiple nodes to obscure origin and destination.
+        \item Vulnerable to global adversaries and end-node monitoring.
+    \end{itemize}
+    \item \textbf{I2P:}
+    \begin{itemize}
+        \item Optimized for anonymous communication within the network.
+        \item Use cases: Peer-to-peer file sharing, messaging, and hosting services.
+    \end{itemize}
+    \item \textbf{Mix Networks:}
+    \begin{itemize}
+        \item Provide stronger anonymity by shuffling and re-encrypting messages at each node.
+        \item Used in email anonymization and anonymous transactions.
+    \end{itemize}
+\end{itemize}
+
+\subsection{Censorship Resistance}
+\begin{itemize}
+    \item \textbf{Goals:} Ensure access to information and communication in censored environments.
+    \item \textbf{Techniques:}
+    \begin{itemize}
+        \item Decentralized networks (e.g., Freenet, IPFS).
+        \item Bridge nodes in Tor to bypass censorship firewalls.
+        \item Steganography to disguise traffic as normal communication.
+    \end{itemize}
+    \item \textbf{Challenges:}
+    \begin{itemize}
+        \item Sophisticated censorship tools (e.g., deep packet inspection).
+        \item Balancing usability with security under surveillance.
+    \end{itemize}
+\end{itemize}
+
+\subsection{Criminal Misuse and Data Retention}
+\begin{itemize}
+    \item \textbf{Misuse of Anonymity:}
+    \begin{itemize}
+        \item Use of anonymity networks for illicit activities (e.g., illegal marketplaces, malware distribution).
+        \item Challenges for law enforcement in tracking activities without violating privacy.
+    \end{itemize}
+    \item \textbf{Data Retention Policies:}
+    \begin{itemize}
+        \item Mandated storage of communication metadata for investigative purposes.
+        \item Conflict with privacy-focused systems and anonymization techniques.
+    \end{itemize}
+    \item \textbf{Legal Balance:}
+    \begin{itemize}
+        \item Protecting civil liberties while enabling effective law enforcement.
+    \end{itemize}
+\end{itemize}
+
+\subsection{Traffic Analysis and Fingerprinting}
+\begin{itemize}
+    \item \textbf{Techniques:}
+    \begin{itemize}
+        \item Identifying patterns in encrypted traffic (e.g., packet size, timing).
+        \item Inferring user behavior or accessed resources from metadata.
+    \end{itemize}
+    \item \textbf{Defenses:}
+    \begin{itemize}
+        \item Padding and random delays to obscure traffic patterns.
+        \item Use of obfuscation proxies to disguise protocol characteristics.
+    \end{itemize}
+    \item \textbf{Limitations:}
+    \begin{itemize}
+        \item Increased latency and overhead.
+        \item Limited effectiveness against global adversaries.
+    \end{itemize}
+\end{itemize}
+
+\includepdf[pages={161-219}, nup=4x9]{slides.pdf}
 \part{Location Privacy}
+\section{Location Privacy \& Mobile Networks}
+
+\subsection{Overview}
+\begin{itemize}
+    \item \textbf{Objective:} Protect users' location information from being tracked or monitored by adversaries or unauthorized entities.
+    \item \textbf{Challenges:}
+    \begin{itemize}
+        \item Increasing reliance on location-based services.
+        \item Potential abuse by service providers, adversaries, or malicious third parties.
+    \end{itemize}
+\end{itemize}
+
+\subsection{Systematic Protection Methods}
+\begin{enumerate}
+    \item \textbf{Broadcasting:} Randomized transmission of location data.
+    \item \textbf{Group Pseudonyms:} Shared identifiers among a group of users to obscure individual location.
+    \item \textbf{Temporal Pseudonyms:} Changing pseudonyms periodically to reduce traceability.
+    \item \textbf{Mixing Networks:} Use of cascaded nodes to anonymize the source and destination of location data.
+\end{enumerate}
+
+\subsection{Broadcasting Techniques}
+\begin{itemize}
+    \item \textbf{Implicit Addressing:}
+    \begin{itemize}
+        \item Targets devices indirectly by transmitting generalized location updates.
+        \item Performance: Reduces bandwidth usage but may decrease precision.
+    \end{itemize}
+    \item \textbf{Variable Broadcasts:}
+    \begin{itemize}
+        \item Adapts broadcast size dynamically to obscure exact location.
+        \item Useful in areas with sparse populations.
+    \end{itemize}
+\end{itemize}
+
+\subsection{Group and Temporal Pseudonyms}
+\begin{itemize}
+    \item \textbf{Group Pseudonyms:}
+    \begin{itemize}
+        \item Users within a region share a pseudonym to make identification harder.
+        \item Reduces linkability across updates.
+    \end{itemize}
+    \item \textbf{Temporal Pseudonyms:}
+    \begin{itemize}
+        \item Pseudonyms change over time to prevent long-term tracking.
+        \item Requires synchronization with network systems for functionality.
+    \end{itemize}
+\end{itemize}
+
+\subsection{Mobile Communication-Mixing}
+\begin{itemize}
+    \item \textbf{Centralized Approach:}
+    \begin{itemize}
+        \item A trusted third party manages pseudonym and location updates.
+        \item High scalability but susceptible to single points of failure.
+    \end{itemize}
+    \item \textbf{Decentralized Approach:}
+    \begin{itemize}
+        \item Users and mix cascades collaborate to anonymize location updates.
+        \item Improves privacy but increases coordination complexity.
+    \end{itemize}
+\end{itemize}
+
+\subsection{Anonymity Techniques in Mobile Networks}
+\begin{itemize}
+    \item \textbf{Mix Networks:}
+    \begin{itemize}
+        \item Obscure the link between senders and receivers using layered encryption.
+        \item Suitable for large-scale deployments but introduces latency.
+    \end{itemize}
+    \item \textbf{Blind Signatures:}
+    \begin{itemize}
+        \item Allows authentication without revealing identities.
+        \item Useful in anonymous mutual authentication protocols.
+    \end{itemize}
+\end{itemize}
+
+\subsection{Mobile Internet Protocol (IP)}
+\begin{itemize}
+    \item \textbf{Location Hiding:}
+    \begin{itemize}
+        \item Use of proxies or agents to hide the user's exact location.
+        \item Common in applications with stringent privacy requirements.
+    \end{itemize}
+    \item \textbf{Challenges:}
+    \begin{itemize}
+        \item Higher latency due to intermediate nodes.
+        \item Trade-off between performance and privacy.
+    \end{itemize}
+\end{itemize}
+
+\subsection{Performance Analysis of Privacy Methods}
+\begin{itemize}
+    \item \textbf{Trust Levels:}
+    \begin{itemize}
+        \item Varying levels of trust required in centralized vs. decentralized systems.
+        \item Broadcast and mixing networks minimize reliance on trusted third parties.
+    \end{itemize}
+    \item \textbf{Efficiency Loss:}
+    \begin{itemize}
+        \item Privacy-preserving methods incur bandwidth and processing overhead (1-10\%).
+    \end{itemize}
+    \item \textbf{Protection Metrics:}
+    \begin{itemize}
+        \item Degree of protection against tracking and profiling varies by method.
+        \item Methods like mobile communication-mixing outperform traditional GSM in anonymity.
+    \end{itemize}
+\end{itemize}
 

### 2024-11-28 SbD/exercises.tex
Added sbd exercises and solutions.

@@
   \begin{solution}
-    % TODO
+    \begin{enumerate}
+        \item Verschlüsselung mit Skytala: Teilen Sie den Klartext in eine Matrix mit 7 Zeilen und 9 Spalten:
+          \[
+          \texttt{GLUECKLICH IST DER DER DEN DINGEN AUF DEN GRUND GEHEN KONNTE}
+          \]
+          Die Matrix:
+          \[
+          \begin{array}{ccccccccc}
+          G & L & U & E & C & K & L & I & C \\
+          H & \_ & I & S & T & \_ & D & E & R \\
+          \_ & D & E & R & \_ & D & E & N & \_ \\
+          D & I & N & G & E & N & \_ & A & U \\
+          F & D & E & N & \_ & G & R & U & N \\
+          D & G & E & H & E & N & \_ & K & O \\
+          N & T & T & E & \_ & & & & \\
+          \end{array}
+          \]
+          Der Schlüsseltext wird spaltenweise abgelesen: 
+          \[
+          \texttt{GH\_DDFNLDIRIGTNSEEECEDEINNEK\_CRNUENNUH\_U\_GNGRDKIKAUOE}
+          \]
+
+        \item Der geheime Schlüssel ist die Größe der Matrix: \( 7 \times 9 \) (Zeilen $\times$ Spalten). Beim ursprünglichen Zylinder entspricht dies dem Umfang des Zylinders.
+
+        \item Kryptanalyse:
+        \begin{itemize}
+            \item Häufigkeitsanalyse: Anhand von Buchstabenhäufigkeiten und Mustererkennung können Transpositionsmuster aufgedeckt werden.
+            \item Brute-Force: Alle möglichen Spalten- und Zeilenkombinationen ausprobieren, um den Klartext zu rekonstruieren.
+        \end{itemize}
+    \end{enumerate}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    \begin{enumerate}
+        \item Bei Spaltenvertauschung gibt es \( s!
+\) mögliche Permutationen für \( s \) Spalten.
+Der Schlüsselraum vergrößert sich somit um den Faktor \( s!
+\) im Vergleich zur ursprünglichen Methode ohne Spaltenvertauschung.
+
+        \item Klartext in eine \( 4 \times 3 \)-Matrix:
+          \[
+          \texttt{ABCDEFGHIKLMN}
+          \]
+          In die Matrix eingetragen:
+          \[
+          \begin{array}{ccc}
+          A & B & C \\
+          D & E & F \\
+          G & H & I \\
+          K & L & M \\
+          \end{array}
+          \]
+          Zyklisches Schreiben ergibt:
+          \[
+          \texttt{ADGKBELHCFIM}
+          \]
+    \end{enumerate}
   \end{solution}
@@
 \setcounter{subsection}{20}
-\begin{exercise}{Vigenère-Chiffre}
-  Ein klassisches Verschlüsselungsverfahren ist die Vigenère-Chiffre.
-Sie arbeitet typischerweise auf einem Alphabet aus den Großbuchstaben A-Z für den Klartext, den Schlüssel und den Schlüsseltext.
-Zudem kann ein Zeichen zur Worttrennung hinzugenommen werden.
-Für diese Aufgabe wird dazu der Unterstrich \texttt{\_} verwendet.
-Ein Schlüsseltextzeichen ergibt sich aus dem um das jeweilige Zeichen des Schlüssels verschobene Klartextzeichen.
-Diese Verschiebung kann durch das sogenannte Vigenère-Tableau in Abbildung 1 dargestellt werden.
-
-  % TODO: add table
-  % A B C D E F G H I J K L M N O P Q R S T U V W X Y Z _
-  % A A B C D E F G H I J K L M N O P Q R S T U V W X Y Z _
-  % B B C D E F G H I J K L M N O P Q R S T U V W X Y Z _ A
-  % C C D E F G H I J K L M N O P Q R S T U V W X Y Z _ A B
-  % D D E F G H I J K L M N O P Q R S T U V W X Y Z _ A B C
-  % E E F G H I J K L M N O P Q R S T U V W X Y Z _ A B C D
-  % F F G H I J K L M N O P Q R S T U V W X Y Z _ A B C D E
-  % G G H I J K L M N O P Q R S T U V W X Y Z _ A B C D E F
-  % H H I J K L M N O P Q R S T U V W X Y Z _ A B C D E F G
-  % I I J K L M N O P Q R S T U V W X Y Z _ A B C D E F G H
-  % J J K L M N O P Q R S T U V W X Y Z _ A B C D E F G H I
-  % K K L M N O P Q R S T U V W X Y Z _ A B C D E F G H I J
-  % L L M N O P Q R S T U V W X Y Z _ A B C D E F G H I J K
-  % M M N O P Q R S T U V W X Y Z _ A B C D E F G H I J K L
-  % N N O P Q R S T U V W X Y Z _ A B C D E F G H I J K L M
-  % O O P Q R S T U V W X Y Z _ A B C D E F G H I J K L M N
-  % P P Q R S T U V W X Y Z _ A B C D E F G H I J K L M N O
-  % Q Q R S T U V W X Y Z _ A B C D E F G H I J K L M N O P
-  % R R S T U V W X Y Z _ A B C D E F G H I J K L M N O P Q
-  % S S T U V W X Y Z _ A B C D E F G H I J K L M N O P Q R
-  % T T U V W X Y Z _ A B C D E F G H I J K L M N O P Q R S
-  % U U V W X Y Z _ A B C D E F G H I J K L M N O P Q R S T
-  % V V W X Y Z _ A B C D E F G H I J K L M N O P Q R S T U
-  % W W X Y Z _ A B C D E F G H I J K L M N O P Q R S T U V
-  % X X Y Z _ A B C D E F G H I J K L M N O P Q R S T U V W
-  % Y Y Z _ A B C D E F G H I J K L M N O P Q R S T U V W X
-  % Z Z _ A B C D E F G H I J K L M N O P Q R S T U V W X Y
-  % _ _ A B C D E F G H I J K L M N O P Q R S T U V W X Y Z
-  % Abbildung 1: Vigenère-Tableau
-
-  \begin{enumerate}
-      \item Verschlüsseln Sie den nachfolgenden Klartext mit dem Schlüssel \texttt{PRUEFUNG} unter Verwendung einer Vigenère-Chiffre:
-      \begin{center}
-          \texttt{GLUECKLICH\_IST\_DER\_DER\_DEN\_DINGEN\_AUF\_DEN\_GRUND\_GEHEN\_KONNTE}
-      \end{center}
-      \item Erläutern Sie kurz zwei sinnvolle Ansätze zur Kryptanalyse, d. h. zum Brechen einer Vigenère-Chiffre.
-      \item Eine Verbesserung der Vigenère-Chiffre ergibt sich durch die Verlängerung des Schlüssels.
-Um welchen Faktor vergrößert sich dadurch der Schlüsselraum bei einem Alphabet mit 27 Elementen (A-Z und Unterstrich als Worttrennung) und einer Verlängerung um $s$ Zeichen?
-      \item Wie müsste die Vigenère-Chiffre verändert werden, um informationstheoretische Sicherheit im Bezug auf die Vertraulichkeit zu liefern?
-Begründen Sie, warum diese Veränderung zielführend ist.
-  \end{enumerate}
-
-  \begin{solution}
-    % TODO
-  \end{solution}
-\end{exercise}
+\begin{exercise}{Vigenère-Chiffre \ref{ex:vigenere}}\end{exercise}
 
@@
   \begin{solution}
-    % TODO
+    \begin{enumerate}
+        \item Echte Zufallszahlen sind notwendig, um sicherzustellen, dass Schlüssel nicht vorhersagbar sind.
+Ein nicht zufälliger Schlüssel kann durch statistische oder deterministische Methoden berechnet werden.
+        \item Die XOR-Verknüpfung sichert, dass ein kombinierter Schlüssel nur dann vorhersagbar ist, wenn alle zufälligen Werte vollständig bekannt sind.
+Dadurch wird ein höherer Grad an Sicherheit erzielt.
+    \end{enumerate}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    Durch die Wiederverwendung des One-Time-Pads \( k \) für mehrere Nachrichten \( m_1, m_2 \) entsteht folgende Gleichung:
+    \[
+    c_1 \oplus c_2 = (m_1 \oplus k) \oplus (m_2 \oplus k) = m_1 \oplus m_2
+    \]
+    Dies ergibt die XOR-Verknüpfung der Klartexte \( m_1 \oplus m_2 \), die durch Analyse entschlüsselt werden kann. Beispielsweise:
+    \[
+    c_1 = \texttt{09 00 04 10}, \quad c_2 = \texttt{10 20 28 09}
+    \]
+    XOR:
+    \[
+    c_1 \oplus c_2 = \texttt{19 20 2C 19}
+    \]
+    Häufigkeitsanalysen und bekannte Wörter im Klartext ermöglichen die vollständige Entschlüsselung.
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    \begin{enumerate}
+        \item Ursachen für Algorithmuswechsel:
+        \begin{itemize}
+            \item **Technologischer Fortschritt**: Schnellere Computer machen bestimmte Algorithmen unsicher.
+            \item **Neue Angriffsmethoden**: Entdeckte Schwachstellen führen zur Ablösung alter Verfahren.
+        \end{itemize}
+
+        \item Das **One-Time-Pad** bietet langfristige Sicherheit, da es informationstheoretisch sicher ist.
+Es wird nicht überall genutzt, da Schlüssel so lang wie der Klartext sein müssen und sicher verteilt werden müssen.
+
+        \item Bestimmung der Chiffre:
+        \begin{itemize}
+            \item **Häufigkeitsanalyse**: Substitutionschiffren wie Caesar zeigen typische Muster.
+            \item **Permutationsanalyse**: Transpositionen hinterlassen Permutationsmuster.
+        \end{itemize}
+    \end{enumerate}
   \end{solution}
@@
 
-% \setcounter{subsection}{30}
-% \begin{exercise}{}
+\setcounter{subsection}{30}
+\begin{exercise}{Informationstheoretische Sicherheit \ref{ex:otp}}\end{exercise}
 
-%   \begin{solution}
-%     % TODO
-%   \end{solution}
-% \end{exercise}
+\begin{exercise}{Feistel-Prinzip \ref{ex:feistel}}\end{exercise}
 
-% \setcounter{subsection}{40}
-% \begin{exercise}{}
+\begin{exercise}{Meet-in-the-Middle-Angriff auf 2-DES}
+  Um der geringen Schlüssellänge von 56 Bit des DES-Algorithmus zu begegnen, wurde vorgeschlagen, den
+  Algorithmus mehrfach hintereinander mit verschiedenen Schlüsseln anzuwenden. Eine schlechte Lösung ist
+  jedoch die lediglich zweifache Hintereinanderanwendung 2-DES.
+  \begin{enumerate}
+    \item Vergleichen Sie die Schlüssellänge des 2-DES-Algorithmus mit der tatsächlich erreichten Sicherheit.
+    \item Sie haben folgendes Klartext/Schlüsseltext-Paar \( (m, c) \) einer 2-DES-Verschlüsselung \( E_{k_2}(E_{k_1}(m)) = c \)
+    erlangt: 
+    \[
+        m = \texttt{c183 50b5 52b8 f8b1}, \quad c = \texttt{42e0 4462 099e e8ca}.
+    \]
+    Ermitteln Sie \( k_1 \) und \( k_2 \), indem Sie einen Meet-in-the-Middle-Angriff durchführen.
+Benutzen Sie dafür die unten dargestellte Tabelle 1 ausgewählter DES-Funktionswerte.
+In der Tabelle können für einen gegebenen Klartext \( x \) bzw.
+einen Schlüsseltext \( y \) und einen Schlüssel \( k \) die Werte der Verschlüsselungsfunktion \( E \) bzw.
+Entschlüsselungsfunktion \( D \) des DES-Algorithmus nachgeschlagen werden.
+In welchen Zeilen haben Sie die DES-Funktionswerte, die zum Erfolg beim Meet-in-the-Middle Angriff führen, nachgeschlagen?
+    Hinweis: Beachten Sie das Beispiel zum Ablesen der Tabelle.
 
-%   \begin{solution}
-%     % TODO
-%   \end{solution}
-% \end{exercise}
+    \input{res/des.tbl}
 
-% \setcounter{subsection}{50}
-% \begin{exercise}{}
+    \item Wie viele Einträge enthält eine vollständige Tabelle, die bei einem Meet-in-the-Middle-Angriff auf den 2-DES-Algorithmus erstellt wird, mindestens?
+Welche Art von Berechnungen werden dort festgehalten?
+  \end{enumerate}
 
-%   \begin{solution}
-%     % TODO
-%   \end{solution}
-% \end{exercise}
+  \begin{solution}
+    \begin{enumerate}
+        \item 2-DES bietet nur \( 2^{56} \) statt \( 2^{112} \) effektiver Sicherheit, da ein Meet-in-the-Middle-Angriff \( E_{k_1}(m) = D_{k_2}(c) \) nutzt.
+        \item Analyse aus Tabelle 1:
+        \[
+        k_1 = \texttt{e6f4 a450 4898 d7}, \quad k_2 = \texttt{97a4 3741 8086 0b}
+        \]
+        \item Eine vollständige Tabelle umfasst \( 2^{56} \) Einträge, wobei \( E_k(m) \) oder \( D_k(c) \) für alle \( k \) berechnet werden.
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
 
-% \setcounter{subsection}{60}
-% \begin{exercise}{}
+\begin{exercise}{Triple-DES}
+  Beim Triple-DES-Algorithmus (3-DES) handelt es sich um die dreifache Hintereinanderausführung des DES-
+  Algorithmus. Dabei wird mit zwei 56 Bit langen Schlüsseln \( k_1 \) und \( k_2 \) gearbeitet und beim Verschlüsseln
+  entweder \( E(k_1) \to D(k_2) \to E(k_1) \) (EDE-Modus) oder \( E(k_1) \to E(k_2) \to E(k_1) \) (EEE-Modus) ausgeführt.
+  \begin{enumerate}
+    \item Warum wird dem EDE-Modus meist der Vorzug vor dem EEE-Modus gegeben?
+    \item Berechnen Sie die theoretische und die effektive Schlüssellänge des Triple-DES-Algorithmus unter ei-
+    nem Known-Plaintext-Angriff mit drei 56 Bit langen Schlüsseln, d.h. \( E(k_1) \to E(k_2) \to E(k_3) \).
+  \end{enumerate}
 
-%   \begin{solution}
-%     % TODO
-%   \end{solution}
-% \end{exercise}
+  \begin{solution}
+    \begin{enumerate}
+        \item Der EDE-Modus (Encrypt-Decrypt-Encrypt) schützt gegen Meet-in-the-Middle-Angriffe, da die Verschlüsselung durch Entschlüsselung unterbrochen wird.
+        \item Die effektive Schlüssellänge beträgt:
+        \[
+        \text{Theoretisch: } 3 \times 56 = 168 \text{ Bit}, \quad \text{Effektiv: } 112 \text{ Bit (wegen Angriffen).}
+        \]
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+\setcounter{subsection}{40}
+\begin{exercise}{Unsicherheit des Electronic-Codebook-Modus}
+  Eine einfache, jedoch unsichere Technik zum Betrieb einer Blockchiffre ist der Electronic-Codebook-Modus
+  (ECB-Modus). Dabei wird jeder Klartextblock unabhängig von den anderen Blöcken chiffriert.
+  \begin{enumerate}
+    \item Inwiefern stellt dies ein Sicherheitsproblem dar?
+    \item Veranschaulichen Sie das Problem, indem Sie ein Programm entwickeln, welches eine einfache Bitmap-
+    Grafik einliest, den Bildinhalt mit dem AES-Verfahren im ECB-Modus verschlüsselt und wieder als
+    Bitmap ausgibt. Rufen Sie zum Vergleich die Verschlüsselung noch einmal im CBC-Modus auf und
+    vergleichen Sie die erhaltenen Resultate. Informieren Sie sich dazu über das Datenformat von BMP-
+    Dateien und die Verwendung einer geeigneten Kryptobibliothek (z. B. die Java-Cryptography-API
+    unter \url{http://www.ibm.com/developerworks/java/tutorials/j-sec1/section4.html} oder die Python-pycrypto-
+    API unter \url{https://www.dlitz.net/software/pycrypto/api/current/}).
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+        \item Das Hauptproblem des ECB-Modus ist, dass identische Klartextblöcke in identische Schlüsseltextblöcke umgewandelt werden.
+Dies führt dazu, dass Muster im Klartext auch im Schlüsseltext sichtbar bleiben, was die Sicherheit gefährdet.
+        \item Veranschaulichung:
+        \begin{itemize}
+            \item Entwickeln Sie ein Programm, das eine BMP-Bilddatei einliest und diese im ECB-Modus verschlüsselt.
+Das resultierende Bild wird die ursprünglichen Muster zeigen, da identische Klartextblöcke identisch verschlüsselt werden.
+            \item Im Vergleich dazu führt der CBC-Modus durch die Abhängigkeit jedes Blocks vom vorherigen Block zu einer Verschleierung der Muster.
+        \end{itemize}
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Cipher-Block-Chaining-Betriebsmodus}
+  Beim Cipher-Block-Chaining-Betriebsmodus wird ein Klartextblock \( M_i \) vor der Anwendung der Verschlüsselungsfunktion mit dem unmittelbar vorher erzeugten Chiffretextblock \( C_{i-1} \) durch die XOR-Operation verknüpft.
+  Die erzeugten Chiffretextblöcke hängen dadurch von ihren Vorgängern ab.
+Bei der Verschlüsselung des ersten Klartextblocks \( M_1 \) wird für \( C_0 \) ein zufällig gewählter Initialisierungsvektor (IV) verwendet.
+
+  Hinweis: Gehen Sie davon aus, dass eine moderne Blockchiffre eingesetzt wird, z. B. AES.
+  \begin{enumerate}
+    \item Wie unterscheiden sich zwei Schlüsseltexte, die mit demselben Schlüssel erzeugt wurden und denselben Klartext enthalten, jedoch mit unterschiedlichen IVs erzeugt wurden?
+    \item Wie unterscheiden sich zwei Schlüsseltexte, wenn sie sich lediglich an einer Stelle im ersten Klartextblock unterscheiden (also ein Bit gekippt wird) und ansonsten völlig identisch sind (bei gleichem IV und Schlüssel)?
+    \item Welche Auswirkung hat es beim Entschlüsseln, wenn durch einen Übertragungsfehler ein Bit im zweiten Schlüsseltextblock bei der Übertragung verändert wird?
+    \item Welche Auswirkung hat es beim Entschlüsseln, wenn durch einen Übertragungsfehler ein Bit im Initialisierungsvektor verändert wird?
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+        \item Bei unterschiedlichen IVs führen zwei identische Klartexte zu unterschiedlichen Schlüsseltexten, da der IV die Anfangswerte beeinflusst.
+        \item Ein Unterschied in einem Bit im ersten Klartextblock propagiert sich in allen nachfolgenden Schlüsseltextblöcken durch die XOR-Operation.
+        \item Ein Fehler im zweiten Schlüsseltextblock wirkt sich nur auf den entsprechenden Klartextblock und den nächsten aus, da der CBC-Modus Block für Block entschlüsselt.
+        \item Ein Fehler im IV beeinflusst nur den ersten Klartextblock.
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Synchronisationsfehler bei Betriebsarten von Blockchiffren}
+  Neben additiven Fehlern, bei denen einzelne Bitpositionen oder Nachrichtenabschnitte zufällig oder absichtlich gekippte Bitwerte aufweisen, können auch einzelne Bits oder ganze Blöcke von Nachrichten zufällig oder absichtlich verloren gehen.
+Wenn etwa ganze Nachrichtenblöcke jeweils einzeln in Datenpaketen transportiert werden, könnte ein verlorenes Datenpaket einen solchen Synchronisationsfehler hervorrufen.
+  \begin{enumerate}
+    \item Bitte beschreiben Sie die Fehlerfortpflanzung eines Synchronisationsfehlers aufgrund eines verlorenen Nachrichtenblocks beim Electronic Code Book (ECB)-Modus.
+    \item Bitte beschreiben Sie die Fehlerfortpflanzung eines Synchronisationsfehlers aufgrund eines verlorenen Bits beim Electronic Code Book (ECB)-Modus.
+    \item Wie verhält sich die Fehlerfortpflanzung eines Synchronisationsfehlers beim Cipher Block Chaining (CBC)-Modus, wenn einzelne Bits des übertragenen Ciphertextes verloren gehen?
+    \item Bitte vergleichen Sie die Situation einer Fehlerfortpflanzung aufgrund eines Synchronisationsfehlers bei den stromorientierten Betriebsarten mit der bei CBC.
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+        \item Im ECB-Modus führt der Verlust eines Nachrichtenblocks nicht zu einer Synchronisationsstörung der verbleibenden Blöcke, da jeder Block unabhängig verschlüsselt wird.
+        \item Ein verlorenes Bit im ECB-Modus zerstört den betroffenen Block, aber die folgenden Blöcke bleiben lesbar.
+        \item Im CBC-Modus führt der Verlust einzelner Bits dazu, dass alle nachfolgenden Blöcke falsch entschlüsselt werden, da der Modus auf den vorherigen Block angewiesen ist.
+        \item Bei stromorientierten Betriebsarten (z.
+B.
+CTR-Modus) ist die Fehlerfortpflanzung auf den aktuellen Block beschränkt, da kein Feedback aus vorherigen Blöcken verwendet wird.
+Dies ist ein Vorteil gegenüber CBC.
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+\setcounter{subsection}{50}
+\begin{exercise}{Verketteter Cipher-Block-Chaining-Betriebsmodus}
+  Beim verketteten Cipher-Block-Chaining-Betriebsmodus wird zur Steigerung des Datendurchsatzes nur für die erste Nachricht \( M_1 \) ein Initialisierungsvektor \( IV \) gewählt.
+Das heißt für den Schlüsseltext \( C_1 = C_1^0 || \ldots || C_1^{n_1} \) der ersten Nachricht \( M_1 = M_1^1 || \ldots || M_1^{n_1} \) gilt wie beim normalen CBC:
+  \[
+      C_1^0 = IV, \quad C_1^i = E(C_1^{i-1} \oplus M_1^i, K) \quad \text{mit } i = 1, \ldots, n_1.
+  \]
+  Für alle folgenden Nachrichten \( M_2, \ldots \) wird anstelle des \( IV \) der letzte Schlüsseltextblock des vorherigen Schlüsseltextes verwendet.
+Das heißt für den Schlüsseltext \( C^j = C^j_1 || \ldots || C^j_{n_j} \) aller weiteren Nachrichten \( M_j = M_j^1 || \ldots || M_j^{n_j} \) mit \( j > 1 \) gilt:
+  \[
+      C^j_1 = E(C^{j-1}_{n_{j-1}} \oplus M_j^1, K), \quad C^j_i = E(C^j_{i-1} \oplus M_j^i, K) \quad \text{mit } i = 2, \ldots, n_j.
+  \]
+  \begin{enumerate}
+    \item Berechnen Sie unter Angabe aller Zwischenschritte die prozentuale Erhöhung des Datendurchsatzes.
+Führen Sie die Berechnungen sowohl für einen Block als auch für vier Blöcke pro Nachricht aus.
+Gehen Sie in beiden Fällen von insgesamt \( 2^{20} \) zu verschlüsselnden Nachrichten vor Neugenerierung eines Schlüssels und Neustart mit einem neuen \( IV \) aus.
+    \item Ihnen ist ein Schlüsseltext \( C^j \) in die Hände gefallen, der nur aus zwei möglichen Klartexten \( M_j \) oder \( \tilde{M}_j \) gebildet worden sein kann.
+Erläutern Sie, wie Sie ermitteln könnten, welcher der beiden Klartexte verschlüsselt wurde, wenn Sie den Klartext \( M_{j+1} \) für die nächste Nachricht selbst wählen dürfen.
+    \item Beschreiben Sie ein Szenario für einen derartigen Angriff.
+    \item Bewerten Sie ausgehend von der bisherigen Diskussion die Folge selbst kleinster Änderungen für die Sicherheit kryptographischer Algorithmen.
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+        \item Bei einem verketteten CBC-Modus steigt der Datendurchsatz, da kein neuer IV benötigt wird.
+Für \( 2^{20} \) Nachrichten bei einer Blockgröße von \( 128 \) Bit beträgt der Effizienzgewinn ca.
+0,01\%, da der Initialisierungsaufwand entfällt.
+        \item Wenn \( M_j \) oder \( \tilde{M}_j \) verschlüsselt wurde, kann durch die Manipulation des nächsten Klartextblocks \( M_{j+1} \) festgestellt werden, welcher Klartext verwendet wurde.
+Dies ermöglicht einen Known-Plaintext-Angriff.
+        \item Ein mögliches Szenario: Ein Angreifer, der Zugriff auf den Kommunikationskanal hat, modifiziert gezielt den nächsten Klartextblock, um Rückschlüsse auf vorherige Blöcke zu ziehen.
+        \item Selbst kleinste Änderungen im Verfahren können signifikante Sicherheitsrisiken schaffen.
+Hier: Die Wiederverwendung des letzten Blocks des vorherigen Schlüsseltextes anstelle eines neuen IVs.
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Indeterministische Verschlüsselung \ref{ex:indeterministic}}\end{exercise}
+
+\begin{exercise}{Elliptische Kurven: Grundlagen}
+  Elliptic Curve Cryptography (ECC) bezeichnet asymmetrische Kryptosysteme, die auf Operationen auf elliptischen Kurven über endlichen Körpern aufbauen.
+  \begin{enumerate}
+    \item Erläutern Sie, auf welchem mathematischen Problem die Sicherheit von ECC beruht.
+    \item Beschreiben Sie, wie der Diffie-Hellman-Schlüsselaustausch auf Basis von ECC aussehen könnte.
+    \item Welche Vorteile bietet ECC gegenüber Kryptosystemen wie RSA?
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+        \item Die Sicherheit von ECC beruht auf dem Problem der diskreten Logarithmen über elliptischen Kurven, das als schwer lösbar gilt.
+        \item Der Diffie-Hellman-Schlüsselaustausch funktioniert, indem beide Kommunikationspartner Punkte auf der elliptischen Kurve austauschen und einen gemeinsamen Punkt berechnen.
+        \item Vorteile von ECC:
+        \begin{itemize}
+            \item Kürzere Schlüssel bei gleicher Sicherheit (z. B. ECC-256 entspricht RSA-3072).
+            \item Geringerer Speicher- und Rechenaufwand.
+        \end{itemize}
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Elliptische Kurven: Rechenbeispiele zur Punktaddition}
+  Gegeben sei die elliptische Kurve \( E(\mathbb{GF}(p)) \) mit \( a = 1, b = 6, p = 11 \).
+Hinweis: Wir betrachten für diese Aufgabe die in der Kryptographie häufig genutzten elliptischen Kurven über \( \mathbb{Z}_p \) der Form \( f(x, y) : y^2 \equiv x^3 + ax + b \mod p \).
+  \begin{enumerate}
+    \item Berechnen Sie die Punkte, die auf dieser Kurve liegen.
+    \item Suchen Sie sich zwei beliebige Punkte auf der Kurve und addieren Sie diese. Wie funktioniert die Punktaddition im Allgemeinen?
+    \item Berechnen Sie die skalare Multiplikation des Punktes \( (2, 4) \) mit dem Skalar \( 3 \). Was ist das allgemeine Vorgehen hier?
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+        \item Punkte auf der elliptischen Kurve \( y^2 = x^3 + x + 6 \mod 11 \):
+        \[
+        \{(2, 4), (5, 2), (7, 6), (8, 3)\}
+        \]
+        \item Beispiel Punktaddition:
+        \[
+        P = (2, 4), Q = (5, 2) \quad \text{Ergebnis: } P + Q = (7, 6)
+        \]
+        \item Skalare Multiplikation:
+        \[
+        3 \cdot (2, 4) = (8, 3)
+        \]
+        Vorgehen: Addieren Sie \( P \) wiederholt zu sich selbst.
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+\setcounter{subsection}{60}
+\begin{exercise}{RSA: Eigenschaften und Sicherheit \ref{ex:rsa}}\end{exercise}
+
+\begin{exercise}{Forward Secrecy mit RSA}
+  Zur Erreichung von Forward Secrecy erzeugt Alice auf ihrem Server mit dem RSA-Verfahren einen privaten Schlüssel \( d \) sowie den dazu passenden öffentlichen Schlüssel \( c \), den sie persönlich an ihre Nutzerinnen und Nutzer verteilt.
+Zusätzlich erzeugt der Server von Alice bei jedem Verbindungsaufbau ein neues kurzlebiges (engl.
+ephemeral) RSA-Schlüsselpaar, das aus einem öffentlichen Schlüssel \( c_e \) und einem geheimen Schlüssel \( d_e \) besteht.
+Der Server signiert mit dem privaten Schlüssel \( d \) den neuen öffentlichen Schlüssel \( c_e \) und sendet diesen zusammen mit der Signatur über die aufgebaute Verbindung an die Kundin.
+Die Kundin prüft anschließend die Signatur mithilfe des zuvor von Alice erhaltenen öffentlichen Schlüssels \( c \).
+  \begin{enumerate}
+    \item Welche Verbindungen könnten von einem sehr starken Angreifer wie einem Nachrichtendienst noch entschlüsselt werden, wenn dieser an \( d \) gelangt? Begründen Sie Ihre Antwort.
+    \item Skizzieren Sie einen Angriff, der möglich wäre, wenn die Kundin die Signatur nicht prüft.
+    \item Alice ist das Verfahren zu langsam.
+Sie nimmt folgende Änderung vor: Statt jedes Mal bei der Erzeugung von \( c_e \) und \( d_e \) neue Primzahlen \( p \) und \( q \) für den Modulus \( N_e \) zu generieren, verwendet sie immer die gleichen Primzahlen.
+Nun wählt sie lediglich den öffentlichen Exponenten \( c_e \) neu und berechnet einen dazugehörigen geheimen Exponenten \( d_e \).
+Warum ist dies keine gute Idee?
+Zeigen Sie, wie sich dieses geänderte Verfahren angreifen lässt, um aufgezeichnete Verbindungen zu entschlüsseln.
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+        \item Forward Secrecy wird durch kurzlebige (ephemeral) RSA-Schlüssel erreicht. Jede neue Verbindung erhält ein neues Schlüsselpaar.
+        \item Wenn ein Angreifer den privaten Schlüssel erhält, kann er vergangene Verbindungen entschlüsseln, falls diese nicht mit ephemeral-Schlüsseln abgesichert wurden.
+        \item Ein Angreifer kann die Kommunikation entschlüsseln, indem er den abgefangenen Schlüsseltext mit dem kompromittierten privaten Schlüssel entschlüsselt.
+        \item Forward Secrecy kann durch regelmäßiges Rotieren der Schlüssel gewährleistet werden. Alte Schlüssel werden sicher gelöscht.
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Diffie-Hellman-Schlüsselaustauschprotokoll}
+  Das Diffie-Hellman-Schlüsselaustauschprotokoll (DH-Protokoll) ermöglicht es zwei Kommunikationspartnern, durch den Austausch von Nachrichten über einen unsicheren Kanal einen symmetrischen Sitzungsschlüssel zu etablieren, der von passiven Angreifern nicht ermittelt werden kann.
+Beim ursprünglichen DH-Protokoll, das auch als "anonymes" DH-Protokoll bezeichnet wird, wird kein Schlüsselserver verwendet.
+Die Kommunikationspartner tauschen dabei erst im Moment des Schlüsselaustausches sämtliche Protokollnachrichten über den unsicheren Kanal aus.
+  \begin{enumerate}
+    \item Stellen Sie den Ablauf des anonymen DH-Protokolls zwischen den beiden Teilnehmern Alice und Bob mit passend gewählten kleinen Zahlen dar.
+    \item Erläutern Sie, auf welcher mathematischen Annahme die Sicherheit des DH-Protokolls basiert. Welche weiteren Annahmen macht man über den Angreifer (Angreifermodell)?
+    \item Schutz vor aktiven Angriffen ist möglich, wenn die Kommunikationspartner einige Parameter bzw.
+Protokollnachrichten vorab über einen sicheren Kanal ausgetauscht haben, der vom Angreifer nicht kontrolliert werden kann.
+Welche Nachrichten müssen vorab ausgetauscht worden sein, damit aktive Angriffe bemerkt werden können?
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+        \item Ablauf:
+        \[
+        g = 5, p = 23, a = 6, b = 15 \quad \text{(geheime Schlüssel)}
+        \]
+        Öffentliche Schlüssel:
+        \[
+        A = g^a \mod p = 8, \quad B = g^b \mod p = 19
+        \]
+        Gemeinsamer Schlüssel:
+        \[
+        K = B^a \mod p = 2 = A^b \mod p
+        \]
+        \item Sicherheit basiert auf dem diskreten Logarithmusproblem (DLP). Der Angreifer kann \( g^a \) und \( g^b \) nicht effizient in \( a \) und \( b \) umwandeln.
+        \item Schutz vor aktiven Angriffen: Vorab-Austausch von \( g, p \) und Authentifizierung der öffentlichen Schlüssel (z. B. durch Signaturen).
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Elliptische Kurven: EC-ElGamal \ref{ex:elliptic}}\end{exercise}
 
@@
 % \begin{exercise}{}
-
+%  % TODO
+%
 %   \begin{solution}
@@
 % \begin{exercise}{}
+%  % TODO
+%
+%   \begin{solution}
+%     % TODO
+%   \end{solution}
+% \end{exercise}
 
+% \setcounter{subsection}{100}
+% \begin{exercise}{}
+%  % TODO
+%
+%   \begin{solution}
+%     % TODO
+%   \end{solution}
+% \end{exercise}
+
+% \setcounter{subsection}{110}
+% \begin{exercise}{}
+%  % TODO
+%
 %   \begin{solution}
@@
   \begin{solution}
-    % TODO
+    \begin{enumerate}
+        \item Ein DNS-Spoofing-Angriff kann für einen Distributed Denial of Service (DDoS)-Angriff ausgenutzt werden, indem der Angreifer gefälschte DNS-Antworten an Opfer sendet, die zu einem falschen Server führen.
+Der Server wird durch die massiven Anfragen überlastet.
+        \item Die Ausgabe zeigt ein ARP-Spoofing-Angriff.
+Alle IP-Adressen im lokalen Netzwerk werden mit derselben MAC-Adresse verknüpft (\texttt{00:1f:88:13:fe:f1}).
+Dadurch kann der Angreifer die Kommunikation zwischen den Teilnehmern überwachen oder manipulieren.
+        \item ARP-Spoofing funktioniert nur in lokalen Netzwerken, da ARP (Address Resolution Protocol) nicht über Router hinaus funktioniert. Es ist auf das LAN beschränkt.
+        \item Um ARP-Spoofing zu erkennen, können Sie prüfen:
+        \begin{itemize}
+            \item Stimmen die MAC-Adressen in der ARP-Tabelle mit den bekannten Geräten überein?
+            \item Gibt es doppelte Einträge für unterschiedliche IP-Adressen mit derselben MAC-Adresse?
+        \end{itemize}
+    \end{enumerate}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    \begin{enumerate}
+        \item Eve kann einen \textbf{Man-in-the-Middle-Angriff (MITM)} durchführen, indem sie ARP-Spoofing einsetzt.
+Sie fälscht die Zuordnung der IP-Adressen zu MAC-Adressen, sodass der gesamte Netzverkehr von Bob über ihren Computer geleitet wird.
+Eve kann die Daten lesen und manipulieren.
+        \item Bob kann sich schützen durch:
+        \begin{itemize}
+            \item Nutzung von HTTPS für verschlüsselte Verbindungen.
+            \item Aktivieren von ARP-Schutzmechanismen (z. B. ARP-Wächter).
+            \item Überprüfung von SSL-Zertifikaten auf Authentizität.
+        \end{itemize}
+    \end{enumerate}
   \end{solution}
@@
 \setcounter{subsection}{104}
-\begin{exercise}{Indeterministische Verschlüsselung}
+\begin{exercise}{Indeterministische Verschlüsselung}\label{ex:indeterministic}
   \begin{enumerate}
@@
   \begin{solution}
-    % TODO
+    \begin{enumerate}
+        \item Indeterministische Verschlüsselung bedeutet, dass derselbe Klartext bei wiederholter Verschlüsselung mit demselben Schlüssel unterschiedliche Schlüsseltexte erzeugt.
+        \item Bei einem asymmetrischen Verschlüsselungssystem kann indeterministische Verschlüsselung durch die Einbeziehung eines Zufallswerts (z.
+B.
+Padding mit Zufallszahlen) in die Nachricht erreicht werden.
+        \item Bei einem symmetrischen Verschlüsselungssystem kann dies durch die Verwendung eines Initialisierungsvektors (IV) realisiert werden.
+        \item Verfahren:
+        \begin{itemize}
+            \item **RSA**: Fügen Sie der Nachricht einen zufälligen Wert hinzu (z. B. OAEP-Padding).
+            \item **ElGamal**: Verwenden Sie bei jeder Verschlüsselung einen neuen Zufallswert.
+        \end{itemize}
+        \item Indeterministische Verschlüsselung verhindert einen **Known-Plaintext-Angriff**, bei dem der Angreifer durch die Analyse von wiederholt verschlüsselten Nachrichten auf den Klartext schließen könnte.
+    \end{enumerate}
   \end{solution}
@@
 
-\begin{exercise}{Informationstheoretische Sicherheit}
+\begin{exercise}{Informationstheoretische Sicherheit}\label{ex:otp}
   Das One-Time-Pad (Vernam-Chiffre) ist ein sehr einfaches, aber informationstheoretisch sicheres symme- trisches Verschlüsselungsverfahren: Die Klartextbits $x_i$ werden einzeln XOR-verknüpft mit einer zufälligen Schlüsselfolge $k_i$ gleicher Länge, die nur ein einziges Mal verwendet werden darf (Schlüsseltext $s_i = x_i \oplus k_i$ mit $i= 1,2,...$).
@@
   \begin{solution}
-    % TODO
+    \begin{enumerate}
+        \item Der Angreifer erfährt bei einem One-Time-Pad nichts über den Klartext, da der Schlüsseltext rein zufällig ist und keine Informationen über den Klartext enthält.
+        \item Sicherheit des Verfahrens:
+        \[
+        H(M|C) = H(M) \quad \text{(Entropie bleibt erhalten).}
+        \]
+        Der Schlüsseltext enthält keine Informationen über den Klartext ohne den Schlüssel.
+        \item Wiederverwendung des Schlüssels:
+        \[
+        C_1 \oplus C_2 = M_1 \oplus M_2
+        \]
+        Dies ermöglicht die Rekonstruktion der Klartexte \( M_1, M_2 \).
+        \item Ein informationstheoretisch sicheres Authentifikationssystem muss auf zufälligen, einmaligen Schlüsseln basieren (z. B. MACs mit OTPs).
+        \item Ein System: Der Sender erzeugt einen zufälligen Schlüssel und XORt diesen mit der Nachricht, um den MAC zu erstellen. Der Empfänger überprüft, ob der MAC gültig ist.
+        \item Es gibt keine informationstheoretisch sicheren asymmetrischen Verfahren, da der öffentliche Schlüssel immer Angriffsvektoren bietet. Die Sicherheit ist hier nur computational.
+    \end{enumerate}
   \end{solution}
@@
 \setcounter{subsection}{112}
-\begin{exercise}{Unsicherheit der klassischen Verschiebechiffre}
+\begin{exercise}{Unsicherheit der klassischen Verschiebechiffre}\label{ex:vigenere}
   Ein klassisches Verschlüsselungsverfahren ist die Verschiebechiffre (Vigenère-Chiffre).
@@
 Dieses sogenannte Vigenère-Tableau ist in Abbildung 1 dargestellt.
-  % TODO: add table
-  % A B C D E F G H I J K L M N O P Q R S T U V W X Y Z _
-  % A A B C D E F G H I J K L M N O P Q R S T U V W X Y Z _
-  % B B C D E F G H I J K L M N O P Q R S T U V W X Y Z _ A
-  % C C D E F G H I J K L M N O P Q R S T U V W X Y Z _ A B
-  % D D E F G H I J K L M N O P Q R S T U V W X Y Z _ A B C
-  % E E F G H I J K L M N O P Q R S T U V W X Y Z _ A B C D
-  % F F G H I J K L M N O P Q R S T U V W X Y Z _ A B C D E
-  % G G H I J K L M N O P Q R S T U V W X Y Z _ A B C D E F
-  % H H I J K L M N O P Q R S T U V W X Y Z _ A B C D E F G
-  % I I J K L M N O P Q R S T U V W X Y Z _ A B C D E F G H
-  % J J K L M N O P Q R S T U V W X Y Z _ A B C D E F G H I
-  % K K L M N O P Q R S T U V W X Y Z _ A B C D E F G H I J
-  % L L M N O P Q R S T U V W X Y Z _ A B C D E F G H I J K
-  % M M N O P Q R S T U V W X Y Z _ A B C D E F G H I J K L
-  % N N O P Q R S T U V W X Y Z _ A B C D E F G H I J K L M
-  % O O P Q R S T U V W X Y Z _ A B C D E F G H I J K L M N
-  % P P Q R S T U V W X Y Z _ A B C D E F G H I J K L M N O
-  % Q Q R S T U V W X Y Z _ A B C D E F G H I J K L M N O P
-  % R R S T U V W X Y Z _ A B C D E F G H I J K L M N O P Q
-  % S S T U V W X Y Z _ A B C D E F G H I J K L M N O P Q R
-  % T T U V W X Y Z _ A B C D E F G H I J K L M N O P Q R S
-  % U U V W X Y Z _ A B C D E F G H I J K L M N O P Q R S T
-  % V V W X Y Z _ A B C D E F G H I J K L M N O P Q R S T U
-  % W W X Y Z _ A B C D E F G H I J K L M N O P Q R S T U V
-  % X X Y Z _ A B C D E F G H I J K L M N O P Q R S T U V W
-  % Y Y Z _ A B C D E F G H I J K L M N O P Q R S T U V W X
-  % Z Z _ A B C D E F G H I J K L M N O P Q R S T U V W X Y
-  % _ _ A B C D E F G H I J K L M N O P Q R S T U V W X Y Z
-  % Abbildung 1: Vigenère-Tableau
+  \input{res/vigenere.tbl.tex}
   \begin{enumerate}
@@
   \begin{solution}
-    % TODO
+    \begin{enumerate}
+        \item Verschlüsselung mit PRUEFUNG als Schlüssel:
+        \[
+        C = \texttt{ILNUHAZKA\_VLA\_ANL\_KLU\_LTS\_GMMI\_VQK\_TVT\_OZUB\_DRNT\_OZXEU}
+        \]
+        \item Kryptanalyse:
+        \begin{itemize}
+            \item Häufigkeitsanalyse, um Muster aufzudecken.
+            \item Analyse von Periodizitäten, um die Schlüssellänge zu bestimmen.
+        \end{itemize}
+        \item Verlängerung des Schlüsselworts:
+        \[
+        \text{Faktor der Vergrößerung: } 27^s
+        \]
+        \item Um informationstheoretische Sicherheit zu erreichen, müsste das Verfahren in ein One-Time-Pad umgewandelt werden.
+Dazu müsste der Schlüssel genauso lang wie der Klartext und zufällig sein.
+    \end{enumerate}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    \begin{enumerate}
+        \item Verschlüsselung:
+        \begin{itemize}
+            \item Mit der Skytala (\(k = 4\), \(4 \times 8\) Matrix):
+            \[
+            \texttt{PRUEFUNGSSPASS\_OHNE\_KOMPROMISSE} \rightarrow \texttt{PHSOROPRUNFEINSUPG\_GHNOM\_}
+            \]
+            \item Mit der Caesar-Chiffre (\(k = 3\)):
+            \[
+            \texttt{STUDIUMSFREUDE} \rightarrow \texttt{VWXGLXPVIKUHGH}
+            \]
+        \end{itemize}
+        \item Häufigkeitsanalyse:
+        \begin{itemize}
+            \item Besonders problematisch bei der Caesar-Chiffre, da Buchstabenhäufigkeiten direkt übertragen werden.
+            \item Die Skytala bewahrt Buchstabenhäufigkeiten ebenfalls, ist jedoch schwieriger zu entschlüsseln, da Permutationen involviert sind.
+        \end{itemize}
+        \item Kombination der beiden Verfahren:
+        \begin{itemize}
+            \item Kombination 1 (Caesar vor Skytala): Sicherheit erhöht sich, da Buchstabenhäufigkeiten durch die Caesar-Chiffre maskiert werden.
+            \item Kombination 2 (Skytala vor Caesar): Sicherheit bleibt gleich wie bei der Skytala alleine, da Caesar die Buchstabenhäufigkeiten nicht beeinflusst.
+        \end{itemize}
+        \item Moderne Chiffren müssen folgende Gütekriterien erfüllen:
+        \begin{itemize}
+            \item \textbf{Konfusionsanforderung}: Beziehung zwischen Klartext und Schlüsseltext sollte möglichst komplex sein.
+            \item \textbf{Diffusionsanforderung}: Änderungen eines Klartextbits sollten viele Schlüsseltextbits beeinflussen.
+            \item \textbf{Resistenz gegen Kryptoanalyse}: Gegenüber modernen Angriffen wie Brute-Force, statistischen und algebraischen Angriffen sicher.
+        \end{itemize}
+    \end{enumerate}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    \begin{enumerate}
+        \item Die Identitätsfunktion \( H(x) = x \) ist keine kryptographische Hashfunktion, da sie nicht kollisionsresistent ist und keine Kompression durchführt.
+        \item Kollisionsfreiheit ist entscheidend, da der Angreifer sonst zwei Nachrichten mit identischem Hashwert erzeugen könnte, um Signaturen zu fälschen.
+        \item Vorteile des Signierens von Hashwerten:
+        \begin{itemize}
+            \item Reduziert den Rechenaufwand bei großen Nachrichten.
+            \item Stellt sicher, dass die Nachricht authentisch ist.
+            \item Padding (z. B. bei RSA) schützt vor Angriffen.
+        \end{itemize}
+        \item Verfahren für die Software-Integrität:
+        \begin{itemize}
+            \item Entwicklerin erzeugt den Hashwert der Software und veröffentlicht diesen zusammen mit der Software.
+            \item Nutzer prüfen den Hashwert nach dem Herunterladen.
+            \item Entscheidend ist die Kollisionsresistenz der Hashfunktion, um sicherzustellen, dass der Hash nicht gefälscht werden kann.
+        \end{itemize}
+    \end{enumerate}
   \end{solution}
@@
 \setcounter{subsection}{150}
-\begin{exercise}{Das Feistel-Prinzip}
+\begin{exercise}{Das Feistel-Prinzip}\label{ex:feistel}
   \begin{enumerate}
@@
       \end{itemize}
+      \input{res/feistel.fig.tex}
     \item Welche Eigenschaften sollte die Einwegfunktion F aufweisen, um mit ihrer Hilfe eine sichere Feistel-Chiffre zu erzeugen?
@@
   \begin{solution}
-    % TODO
+    \begin{enumerate}
+        \item Zwei Runden Feistel-Verschlüsselung:
+        \begin{itemize}
+            \item Klartext: \( 0011 \ 0100 \)
+            \item Runde 1:
+            \[
+            L_1 = R_0, \quad R_1 = L_0 \oplus F(R_0, K_1)
+            \]
+            \( F(R_0, K_1) = R_0 \land K_1 \)
+            \item Runde 2:
+            \[
+            L_2 = R_1, \quad R_2 = L_1 \oplus F(R_1, K_2)
+            \]
+        \end{itemize}
+        \item Eigenschaften der Funktion \( F \):
+        \begin{itemize}
+            \item Effizient berechenbar.
+            \item Pseudozufällig, aber deterministisch.
+            \item Nicht umkehrbar (Einhaltung der Sicherheit).
+        \end{itemize}
+        \item Eine konstante Funktion \( F \) funktioniert, da die Struktur der Feistel-Chiffre die Umkehrbarkeit sicherstellt.
+        \item Die Umkehrfunktion \( F^{-1} \) ist nicht notwendig, da die Entschlüsselung über die symmetrische Struktur erfolgt.
+    \end{enumerate}
   \end{solution}
@@
 \setcounter{subsection}{162}
-\begin{exercise}{RSA: Eigenschaften und Sicherheit}
+\begin{exercise}{RSA: Eigenschaften und Sicherheit}\label{ex:rsa}
   Für die Erzeugung eines RSA-Schlüsselpaares wurde von Alice der Modulus $n= p \cdot q= 7 \cdot 17= 119$ gewählt.
@@
   \begin{solution}
-    % TODO
+    \begin{enumerate}
+        \item Der öffentliche Exponent \( c \) muss teilerfremd zu \( \phi(n) \) sein, damit ein inverser Schlüssel existiert.
+        \item Eine Nachricht \( m \) darf \( p \) oder \( q \) nicht enthalten, da dies den privaten Schlüssel direkt offenbart.
+        \item Berechnung des privaten Exponenten \( d \):
+        \[
+        c = 5, \quad n = 7 \cdot 17 = 119, \quad \phi(n) = 96
+        \]
+        \( d = c^{-1} \mod \phi(n) \implies d = 77 \)
+        \item Multiplikative Homomorphie:
+        \[
+        m_1 \cdot m_2 \mod n \implies c(m_1 \cdot m_2) = c(m_1) \cdot c(m_2)
+        \]
+        Angreifer können Produkte von Nachrichten entschlüsseln.
+        \item Eve nutzt:
+        \[
+        r \cdot s \cdot r^{-1} \mod n = s
+        \]
+        Der Angriff ist erfolgreich.
+        \item Multiplikative Homomorphie wird in homomorpher Verschlüsselung absichtlich genutzt (z. B. für Berechnungen auf verschlüsselten Daten).
+    \end{enumerate}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    \begin{enumerate}
+        \item Eve führt einen Chosen-Plaintext-Angriff durch:
+        \[
+        c_B = m^e \mod n
+        \]
+        Für alle möglichen \( m \), vergleicht sie den berechneten \( c_B \) mit dem übertragenen Wert, bis \( m = 3 \) gefunden wird.
+        \item Bob kann sich schützen, indem er Zufallswerte oder Padding einfügt, um die deterministische Eigenschaft des RSA-Verfahrens zu verhindern.
+    \end{enumerate}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    \begin{enumerate}
+        \item Forward Secrecy mit RSA wird erreicht, indem für jede neue Sitzung ein neues, kurzlebiges (ephemeres) RSA-Schlüsselpaar generiert wird.
+Der öffentliche Schlüssel wird dem Kommunikationspartner zur Verfügung gestellt, und der private Schlüssel wird nach Abschluss der Sitzung sicher gelöscht.
+Auf diese Weise kann ein kompromittierter langfristiger Schlüssel vergangene Sitzungen nicht gefährden.
+        \item Wenn ein Angreifer den privaten Entschlüsselungsexponenten eines RSA-Schlüssels erlangt, kann er alle Nachrichten entschlüsseln, die mit dem zugehörigen öffentlichen Schlüssel verschlüsselt wurden.
+Dadurch wird die Vertraulichkeit der Kommunikation vollständig aufgehoben.
+        \item Der Angreifer kann die verschlüsselten Nachrichten abfangen und mit dem privaten Schlüssel entschlüsseln.
+Alle vergangenen und zukünftigen Nachrichten, die denselben Schlüssel verwenden, können entschlüsselt werden, sofern sie aufgezeichnet wurden.
+        \item Durch regelmäßiges Wechseln des privaten Schlüssels und Löschen des alten privaten Schlüssels nach Ablauf einer Sitzung wird Forward Secrecy gewährleistet.
+Zusätzlich kann der Diffie-Hellman-Schlüsselaustausch verwendet werden, um symmetrische Sitzungsschlüssel auszuhandeln, die nicht mit den RSA-Schlüsseln gespeichert werden.
+    \end{enumerate}
   \end{solution}
@@
 \setcounter{subsection}{1615}
-\begin{exercise}{Elliptische Kurven: EC-ElGamal}
+\begin{exercise}{Elliptische Kurven: EC-ElGamal}\label{ex:elliptic}
   Das ElGamal-Kryptosystem lässt sich auch auf elliptische Kurven übertragen.
@@
   \begin{solution}
-    % TODO
+    \begin{enumerate}
+        \item Anpassung des ElGamal-Kryptosystems für elliptische Kurven:
+        \begin{itemize}
+            \item **Schlüsselgenerierung:** 
+            \[
+            \text{Privater Schlüssel } d, \text{ öffentlicher Schlüssel } P = d \cdot G, \text{ wobei } G \text{ ein Basispunkt ist.}
+            \]
+            \item **Verschlüsselung:** 
+            \[
+            \text{Nachricht } m, \text{ Zufallswert } k, \text{ Schlüsseltext: } (C_1, C_2) = (k \cdot G, m + k \cdot P).
+            \]
+            \item **Entschlüsselung:** 
+            \[
+            m = C_2 - d \cdot C_1.
+            \]
+        \end{itemize}
+        \item Schlüsselgenerierung:
+        \begin{itemize}
+            \item Wähle \( d = 3 \) als privaten Schlüssel.
+            \item Berechne \( P = d \cdot G = 3 \cdot (2, 7) = (8, 3) \) (Punkt auf der Kurve).
+            \item Verschlüsselung der Nachricht \( m = (5, 2) \):
+            \[
+            C_1 = k \cdot G = 2 \cdot (2, 7) = (6, 3), \quad C_2 = m + k \cdot P = (5, 2) + 2 \cdot (8, 3) = (7, 6).
+            \]
+            Schlüsseltext: \( (C_1, C_2) = ((6, 3), (7, 6)) \).
+        \end{itemize}
+        \item Entschlüsselung:
+        \[
+        m = C_2 - d \cdot C_1 = (7, 6) - 3 \cdot (6, 3) = (5, 2).
+        \]
+        Die ursprüngliche Nachricht \( m \) wurde korrekt wiederhergestellt.
+        \item Die Sicherheit des Verfahrens beruht auf dem **diskreten Logarithmusproblem** über elliptischen Kurven.
+Ein Angreifer müsste \( k \) aus \( C_1 = k \cdot G \) berechnen, was aufgrund der mathematischen Schwierigkeit des Problems praktisch unmöglich ist.
+    \end{enumerate}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item \textbf{Verschlüsselte Abschnitte:} Der Abschnitt zwischen dem Mobiltelefon und der Basisstation wird durch den A5-Algorithmus verschlüsselt.
+        \item \textbf{Unverschlüsselte Abschnitte:} Die Verbindung zwischen der Basisstation und den zentralen Netzwerkkomponenten (z.
+B.
+dem MSC) ist unverschlüsselt.
+Dies stellt ein Sicherheitsrisiko dar, da Angriffe auf das Netzwerk (z.
+B.
+durch gefälschte Basisstationen) möglich sind.
+    \end{itemize}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item \textbf{Input:} Der Eingabewert der Funktion \( E22 \) ist ein gemeinsames Schlüsselpaar (Link Key), das während der Kopplung zwischen zwei Bluetooth-Geräten ausgetauscht wird.
+        \item \textbf{Output:} Die Funktion erzeugt einen Authentication Code, der verwendet wird, um die Identität der Geräte gegenseitig zu authentifizieren.
+    \end{itemize}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    \begin{enumerate}
+        \item Maximal \( 2^{24} \) Pakete können erzeugt werden, da der IV 24 Bit groß ist.
+Um Kollisionen zu vermeiden, sollte der IV systematisch inkrementiert werden.
+Problematisch ist jedoch, dass bei mehreren Teilnehmern des WLANs die Wahrscheinlichkeit von IV-Kollisionen schnell steigt.
+        \item Die ersten 10 Klartextbits des zweiten Pakets \( P_2 \) werden durch Vergleich mit \( P_1 \) berechnet:
+        \[
+        \texttt{Klartext}(P_2) = \texttt{Verschlüsselter\_Text}(P_2) \oplus \texttt{Schlüssel}(P_1).
+        \]
+        Nach Durchführung der XOR-Berechnung ergibt sich der Klartext.
+    \end{enumerate}
   \end{solution}
@@
 
-% TODO
+% \begin{exercise}{}
+%  % TODO
+%
+%   \begin{solution}
+%     % TODO
+%   \end{solution}
+% \end{exercise}
 
@@
   \begin{solution}
-    % TODO
+    \begin{enumerate}
+        \item Die Modi ECB und CBC benötigen eine invertierbare Verschlüsselungsfunktion, da der Klartext beim Entschlüsseln zurückgewonnen werden muss.
+Der CTR-Modus hingegen benötigt keine invertierbare Funktion, da er auf dem XOR-Mechanismus basiert.
+        \item Die Verschlüsselung lässt sich im CTR-Modus parallelisieren, da die Blöcke unabhängig voneinander verarbeitet werden können.
+Im CBC-Modus ist dies nicht möglich, da jeder Block von seinem Vorgänger abhängt.
+        \item Ein Bitfehler im zweiten Block beeinflusst nur den entsprechenden Block im ECB-Modus. Im CBC-Modus beeinflusst er zusätzlich den nachfolgenden Block.
+        \item Wenn ein Bit im zweiten Block fehlt, kann nur der ECB-Modus korrekt entschlüsseln, da die Blöcke unabhängig verarbeitet werden.
+Im CBC- und CTR-Modus führt der Fehler zu Synchronisationsverlust.
+        \item Beim Verlust eines ganzen Blocks kann nur der ECB-Modus korrekt entschlüsseln, da die Blöcke unabhängig verarbeitet werden.
+        \item Im CTR-Modus kann ein Angreifer zwei Klartexte entschlüsseln, wenn der gleiche IV verwendet wurde:
+        \[
+        C_1 \oplus C_2 = M_1 \oplus M_2.
+        \]
+        Durch eine bekannte Klartextanalyse kann \( M_2 \) bestimmt werden.
+    \end{enumerate}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    \begin{enumerate}
+        \item Die Sicherheit elliptischer Kurven beruht auf dem Problem des diskreten Logarithmus in elliptischen Kurvengruppen, das als schwierig gilt.
+        \item Punkte auf der elliptischen Kurve \( y^2 = x^3 + 6x + 1 \mod 11 \):
+        \[
+        \{(2, 4), (5, 2), (7, 6), (8, 3), \ldots\}.
+        \]
+        \item Vorteile von ECC:
+        \begin{itemize}
+            \item Kürzere Schlüssel bei gleicher Sicherheit (z. B. ECC-256 entspricht RSA-3072).
+            \item Geringere Speicher- und Rechenanforderungen.
+        \end{itemize}
+    \end{enumerate}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    \begin{enumerate}
+        \item Eine Signatur dient der Authentifizierung und Integritätsprüfung einer Nachricht.
+Ein Zertifikat enthält den öffentlichen Schlüssel eines Teilnehmers sowie die Signatur einer vertrauenswürdigen Zertifizierungsstelle (CA), die die Authentizität des Schlüssels bestätigt.
+        \item N7 muss den CAs vertrauen, die die Zertifikate von N3 und allen relevanten Zwischenstellen signiert haben.
+        \item N7 muss nur den CAs vertrauen, die sowohl N3 als auch N5 direkt oder indirekt zertifizieren. Dies minimiert die Anzahl der erforderlichen CAs.
+        \item Wenn CA2 ausfällt, können alternative Zertifikate verwendet werden, die von anderen vertrauenswürdigen CAs ausgestellt wurden. Ein neuer Vertrauenspfad kann erstellt werden.
+        \item Vorteile von Trust on First Use (TOFU):
+        \begin{itemize}
+            \item Kein zentraler CA-Betrieb notwendig.
+            \item Schlüssel können einfach ausgetauscht werden.
+        \end{itemize}
+        Nachteile:
+        \begin{itemize}
+            \item Unsicherheit beim ersten Austausch.
+            \item Angreifer können Schlüssel bei der Erstkommunikation abfangen.
+        \end{itemize}
+    \end{enumerate}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    \begin{enumerate}
+        \item Abkürzungen und ihre Funktion:
+        \begin{itemize}
+            \item \( K_i \): Geheimer Authentifizierungsschlüssel.
+            \item \( IMSI \): Internationale Teilnehmerkennung.
+            \item \( RAND \): Zufallszahl für Authentifizierung.
+            \item \( SRES \): Signatur zur Authentifizierung.
+            \item \( LAI \): Standortbereichskennung.
+            \item \( MSISDN \): Telefonnummer des Teilnehmers.
+            \item \( K_c \): Verschlüsselungsschlüssel für den Funkkanal.
+        \end{itemize}
+        \item Viele der Funktionen wie \( K_i \), \( RAND \) und \( K_c \) finden sich auch im UMTS wieder, jedoch oft mit anderen Algorithmen oder Benennungen.
+    \end{enumerate}
   \end{solution}
@@
  \begin{solution}
-    % TODO
+  \begin{enumerate}
+      \item Gründe für den Wechsel kryptographischer Algorithmen:
+      \begin{itemize}
+          \item **Technologischer Fortschritt:** Schnellere Computer ermöglichen effizientere Brute-Force-Angriffe.
+          \item **Neue Angriffe:** Fortschritte in der Kryptoanalyse decken Schwachstellen auf, wie z. B. Differential- oder lineare Kryptoanalyse.
+      \end{itemize}
+      \item Das **One-Time-Pad** bietet absolute Sicherheit und bleibt auch nach 98 Jahren sicher. Es wird nicht flächendeckend eingesetzt, da:
+      \begin{itemize}
+          \item Der Schlüssel so lang wie der Klartext sein muss.
+          \item Der sichere Schlüsselaustausch und die Speicherung unpraktisch sind.
+      \end{itemize}
+      \item Die Chiffre, die Buchstaben \( A-N \) mit \( N-Z \) vertauscht, ist nicht sicher:
+      \begin{itemize}
+          \item **Polyalphabetische Verschlüsselung:** Nicht erfüllt, da nur ein Alphabet genutzt wird.
+          \item **Länge bleibt gleich:** Diese Eigenschaft bewahrt Muster im Schlüsseltext, was statistische Angriffe erleichtert.
+      \end{itemize}
+  \end{enumerate}
 \end{solution}
@@
   \begin{solution}
-    % TODO
+    \begin{enumerate}
+        \item Punkte auf der elliptischen Kurve \( y^2 = x^3 + 2x + 3 \mod 7 \):
+        \[
+        \{(1, 1), (2, 2), (4, 5), (6, 3)\}.
+        \]
+        \item Das ElGamal-Verfahren mit elliptischen Kurven funktioniert, weil die Addition von Punkten und die Multiplikation mit einem Skalar die Gruppenstruktur der Kurve erhalten.
+Dadurch bleibt das diskrete Logarithmusproblem schwer lösbar.
+        \item Vergleich:
+        \begin{itemize}
+            \item Beide basieren auf dem diskreten Logarithmusproblem.
+            \item ECC-ElGamal ist effizienter, da kürzere Schlüssel benötigt werden.
+        \end{itemize}
+    \end{enumerate}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    \begin{enumerate}
+        \item Sicherheitsbewertung der Verfahren:
+        \begin{itemize}
+            \item **Alle Teilschlüssel aneinanderhängen:** Angreifbar durch Brute-Force-Angriffe.
+            \item **Drei Teilschlüssel aneinanderhängen:** Sicherheitsgewinn gegenüber der ersten Methode.
+            \item **XOR aller Schlüssel:** Höchste Sicherheit, solange alle Schlüssel zufällig und unabhängig sind.
+        \end{itemize}
+        \item Zertifikatsbaum: N7 vertraut CAs, die N3 und \( B \) direkt zertifizieren.
+        \item Minimale Anzahl an CAs: Es reicht aus, der CA zu vertrauen, die die Zertifikate für \( B \) und \( N3 \) signiert hat.
+        \item Maßnahme bei Kompromittierung: Alternativer Zertifizierungsweg über eine andere CA. Neue Zertifikate können erstellt werden, um Vertrauen zu erhalten.
+    \end{enumerate}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    \begin{enumerate}
+        \item Abkürzungen im GSM und ihre Funktion:
+        \begin{itemize}
+            \item \( K_i \): Geheimer Schlüssel zur Authentifizierung und Verschlüsselung.
+            \item \( IMSI \): Einzigartige Kennung des Mobilfunkteilnehmers.
+            \item \( RAND \): Zufällige Herausforderung zur Authentifizierung.
+            \item \( SRES \): Antwort des Mobilgeräts auf \( RAND \), basierend auf \( K_i \).
+            \item \( LAI \): Standortkennung für Mobilitätsmanagement.
+            \item \( MSISDN \): Telefonnummer des Teilnehmers.
+            \item \( K_c \): Schlüssel zur Verschlüsselung des Datenverkehrs.
+        \end{itemize}
+        \item Diese Funktionen sind auch im UMTS enthalten, jedoch oft mit verbesserten Sicherheitsmechanismen (z. B. längere Schlüssel, stärkere Algorithmen).
+    \end{enumerate}
   \end{solution}
 \end{exercise}
+
 \end{document}

### 2024-11-28 PbD/exercises.tex
Fixed section numbers.
Added solutions for pbd.

@@
   \begin{solution}
-    % TODO
+    \begin{itemize}
+      \item **Angriffstypen:** Das System ist sowohl für passive Angreifer (durch Beobachtung des Lieferverkehrs) als auch für aktive Angreifer (Manipulation von Bestellungen) anfällig.
+      \item **Bedrohte Schutzziele:** Vertraulichkeit (Erkenntnisse über geplante Operationen) und Integrität (Manipulation der Bestellungen).
+      \item **Mögliche Gegenmaßnahmen:**
+        \begin{itemize}
+          \item Tarnmaßnahmen wie Bestellungen an Dritte.
+          \item Einsatz von Intermediären zur Verschleierung des Bestellvorgangs.
+          \item Reduzierung der Transparenz durch diversifizierte Bestellziele.
+        \end{itemize}
+    \end{itemize}
   \end{solution}
@@
     \begin{enumerate}
-      \item The system is vulnerable to a passive attacker.
-      \item The protection goals that are threatened are confidentiality and integrity.
-      \item Possible countermeasures are to use a VPN connection and to disable automatic connection to unknown networks.
+      \item **Angriffsmöglichkeiten:** Das System ist sowohl durch passive Angreifer (Abhören von SSIDs) als auch durch aktive Angreifer (Erstellen von Rogue-APs mit gleicher SSID) verwundbar.
+      \item **Bedrohte Schutzziele:** Vertraulichkeit (Tracking von Verbindungen), Integrität (Man-in-the-Middle-Angriffe).
+      \item **Gegenmaßnahmen:**
+        \begin{itemize}
+          \item Deaktivierung der automatischen Verbindung zu bekannten Netzwerken.
+          \item Nutzung von WPA3 oder VPN.
+          \item Vermeidung öffentlicher WLANs.
+        \end{itemize}
     \end{enumerate}
@@
   \begin{solution}
-    % TODO
+    \begin{enumerate}
+      \item **Sicherheitsanalyse:** Deterministische Verschlüsselung ermöglicht Replay-Angriffe und das Erraten von PINs durch brute force.
+      \item **Schutzmaßnahmen:** Nutzung eines hybriden Verschlüsselungsverfahrens mit zufälligen Nonces und Transport Layer Security (TLS).
+    \end{enumerate}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    Eine mögliche Lösung ist die Nutzung von Threshold-Signaturen, bei denen mehrere Zertifizierungsstellen gemeinsam einen Pseudonym-Schlüssel erzeugen, ohne dass eine einzelne Stelle die Identität zurückverfolgen kann.
+Im Streitfall können die Stellen gemeinsam das Pseudonym aufdecken.
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    \begin{enumerate}
+      \item **Speicherung im GSM:** Ermöglicht effiziente Anrufzustellung. WLAN benötigt dies nicht, da es keine zentrale Infrastruktur nutzt.
+      \item **Datenbanken:** HLR speichert permanent, VLR nur temporär für lokale Effizienz.
+      \item **Schutz:** Insider greifen durch Zugriffsprotokolle erschwert an; Outsider durch Verschlüsselung.
+      \item **TMSI:** Maskiert IMSI, um Aufenthaltsorte besser zu schützen.
+    \end{enumerate}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    GSM bietet nur beschränkte Sicherheit, da die Verschlüsselung nicht Ende-zu-Ende erfolgt, schwache Verschlüsselungsalgorithmen (z.B.
+A5/1) verwendet werden und Signalisierungsdaten nicht immer geschützt sind.
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    \begin{enumerate}
+      \item Offene implizite Adressen sind für jeden sichtbar, verdeckte implizite Adressen verwenden Pseudonyme, um die Identität zu schützen.
+      \item Explizite Adressierung kann verwendet werden, um Nachrichten zu senden, ohne Teilnehmer direkt zu lokalisieren, aber dies erfordert zusätzliche Mechanismen zur Wahrung der Anonymität.
+      \item Ein Satellit kann für variable implizite Adressierung genutzt werden, jedoch sind Bandbreiten- und Latenzanforderungen zu berücksichtigen.
+    \end{enumerate}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    \begin{enumerate}
+      \item Der Kollisionsabstand beträgt im Durchschnitt $2^{32} / (100.000 \cdot 50)$ Stunden.
+      \item Um einen mittleren Kollisionsabstand von mindestens 3 Jahren zu erreichen, muss die Adresslänge auf mindestens 48 Bit erhöht werden.
+    \end{enumerate}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    \begin{enumerate}
+      \item Der Blind-Message-Service erfüllt die Schutzziele Vertraulichkeit und Anonymität.
+      \item Verschlüsselte Übertragungen verhindern, dass ein Angreifer Inhalte oder Muster erkennt.
+      \item Die Verschlüsselung der Anfragevektoren genügt, da nur der Inhalt der Anfragen sensibel ist.
+      \item Alternativen sind Onion-Routing und Broadcast-Kommunikation, die jedoch mehr Bandbreite benötigen.
+    \end{enumerate}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    \begin{enumerate}
+      \item Die Datenbank verhindert Replay-Angriffe, bei denen Nachrichten mehrfach gesendet werden.
+      \item Die Datenbank kann gelöscht werden, sobald alle Nachrichten erfolgreich weitergeleitet wurden.
+      \item Eine mögliche Modifikation ist die Nutzung von Hashes zur Überprüfung statt vollständiger Speicherung, wodurch der Speicherbedarf reduziert wird.
+    \end{enumerate}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    Wenn Teilnehmer B A, C und D angreift, bleibt die Anonymität für A, C und D teilweise erhalten. Die Analyse basiert auf der Struktur des Schlüsselgraphs und der Isolation kompromittierter Knoten.
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    \begin{enumerate}
+      \item Ein starker Angreifer kann Timing-Analyse und Traffic-Korrelation verwenden, um IP-Adressen von Nutzern zu identifizieren.
+      \item Ein anderer Ansatz ist die Analyse von Metadaten, wie HTTP-Headern oder spezifischen Anfragecharakteristiken.
+      \item Webserver können Nutzer durch Cookies, Browser-Fingerprinting und IP-Adressen wiedererkennen.
+    \end{enumerate}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    \begin{enumerate}
+      \item Beim Tagging-Angriff fügt der Angreifer eine markante Änderung in den Datenverkehr ein, um Nutzer zu identifizieren.
+      \item Der Angriff wird durch strikte Prüfung der Datenintegrität und Verschlüsselung verhindert.
+    \end{enumerate}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    \begin{enumerate}
+      \item Bedrohte Schutzziele sind Vertraulichkeit und Anonymität.
+      \item Tracking auf Anwendungsebene nutzt Browserdaten, während Netzebenen-Tracking auf IP-Daten basiert.
+      \item Gegenmaßnahmen:
+        \begin{itemize}
+          \item Blockieren von Cookies (z.B. durch Browser-Einstellungen).
+          \item Nutzung von VPNs zur Verschleierung der IP.
+          \item Einsatz von Privacy-Plugins (z.B. uBlock Origin).
+        \end{itemize}
+    \end{enumerate}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    Das Vorgehen besteht in der Analyse der Zeitstempel von Access-Logs und Router-Daten, um Verbindungen herzustellen. Diese Art von Angriff wird als Timing-Korrelation bezeichnet.
   \end{solution}
@@
 
-\section[2024]{Mock exam}
+\setcounter{section}{2023}
+\section{Mock exam}
 
-% TODO
+% \begin{exercise}{}
+%  % TODO
+%
+%   \begin{solution}
+%     % TODO
+%   \end{solution}
+% \end{exercise}
 
-\section[2020]{1st exam}
+\setcounter{section}{2019}
+\section{1st exam}
 \begin{exercise}{Broadcast}

### 2024-12-03 PbD/paper/summary.tex
added pbd summary.

\documentclass[
	fontsize=12pt,
	headings=small,
	parskip=quarter,
	bibliography=totoc,
	numbers=noenddot,       
	open=any,               
 	final                   
]{scrreprt}
\input{template}

\bibliographystyle{plain}

\title{Planting Undetectable Backdoors in Machine Learning Models summarised}
% \author{Leo Landowski, Noah Nehring, Leon Langhoff, Leopold Lemmermann}

\begin{document}

\begin{titlepage}
	\thispagestyle{empty}
	\mbox{\parbox[t][1.75cm][b]{2.2cm}{\uhhlogo}}
	\begin{center}\Large
		\vfill
		\makeatletter
		{\Large\textsf{\textbf{\@title}}\par}
		\makeatother
		\bigskip
		Privacy by Design \par
		\bigskip
		\makeatletter
		{\@author} \par
		\makeatother
		\bigskip
		\makeatletter
		{\@date}
		\makeatother
		\vfill
		\vfill
	\end{center}
\end{titlepage}

\pagenumbering{gobble}
\tableofcontents
\thispagestyle{empty}
\pagenumbering{arabic}
\setcounter{page}{0}

\chapter{Introduction}
The scientific paper "Planting Undetectable Backdoors in Machine Learning Models" by Shafi Goldwasser, Michael P. Kim, Vinod Vaikuntanathan, and Or Zamir (2022) investigates how malicious actors can embed undetectable backdoors in machine learning models to covertly manipulate their classifications \cite{goldwasser2022backdoors}. The paper addresses the growing use of machine learning models throughout various domains. Companies can outsource tasks connected to these machine learning models to service providers, thereby democratizing access to machine learning models yet raising serious trust concerns \cite{gu2017badnets}. While using a service provider, a user must trust their claims about accuracy and robustness because, as the paper will present, a service provider can plant an undetectable backdoor into the model. The paper provides a cryptographic framework for understanding and constructing these backdoors and shows how a malicious entity can implement them \cite{liu2018trojaning}. It doesn’t matter for the company if there is white or black-box access to the machine learning model; these backdoors are undetectable, and the already mentioned framework ensures the undetectability.
\par The paper introduces definitions for  
\begin{enumerate} \itemsep -5pt
	\item \textbf{Black-box undetectability}: Where the user has only access to the model's output, making it difficult to identify backdoors. 
	\item \textbf{White-box undetectability}: Where the user has full access to the model's structure, parameters, and training data, yet backdoors remain undetectable. 
	\item \textbf{Non-replicability}: A form of undetectability where a user can't reproduce the backdoor even if they know it exists. 
\end{enumerate}
\par The paper defines multiple countermeasures a company can initiate, but it is important to mention that these can't guarantee immunity. The authors explore techniques to mitigate backdoors by ensuring trusted randomness or using verifiable delegation methods, where the training process is certified as correctly executed \cite{song2017machine}. To reduce backdoor impact, clients might apply post-processing methods like running a few iterations of gradient descent on the trained classifier. Although this approach cannot detect the backdoor directly, it could disrupt its functionality in some cases. The paper discusses adding random noise to inputs during evaluation (randomized smoothing) to make backdoors less effective \cite{madry2018adversarial}. To implement these methods, a company needs to have the right personnel and knowledge, and still, there would be no guarantee that the model is backdoor-free.

{\let\clearpage\relax \chapter{Results and Techniques}}
\section{Defining Undetectable Backdoors}
This section formally defines 'Undetectable Backdoors' and relevant components \cite{goldwasser2022backdoors}.
\begin{enumerate} \itemsep -5pt
    \item \textbf{Training Algorithm} \texttt{Train} - A machine learning model trained on labeled data, which produces a simple classifier $h: \chi \to \{-1,1\}$.
    \item \textbf{Backdoor algorithms} (\texttt{Backdoor}, \texttt{Activate}) - The actual backdoor comprises two algorithms. The \texttt{Backdoor}, like \texttt{Train}, generates a simple classifier $\widetilde{h}$ which appears similar to $h$, and a backdoor key $bk$. This key $bk$ is then used to in the \texttt{Activate} function to modify an input $x$ into $x'$, for which $\widetilde{h}(x')\neq h(x)$ \cite{liu2018trojaning}.
    \item \textbf{Undetectability}: A backdoor is undetectable, if the classifier $\widetilde{h}$ cannot be distinguished from $h$ by polynomial-time algorithms. It is divided further into \textit{black-box} and \textit{white-box undetectability}. \textit{Black-box undetectability} describes cases where it is not possible to find inputs where $\widetilde{h}$ diverges from $h$, even with oracle access to $\widetilde{h}$ (meaning you can apply the function, but do not know the implementation) \cite{carlini2017backdoor}. \textit{White-box undetectability} describes cases where even with full access to the classifier's code, the backdoor cannot be detected, $\widetilde{h}$ and $h$ remain indistinguishable \cite{goldwasser2022backdoors}.
\end{enumerate}
Undetectable backdoors are further differentiated from adversarial examples. Adversarial examples arise naturally in machine-learned classifiers, thus making them exploitable by anyone. The backdoor requires a specific backdoor key, restricting exploitation to individuals with this knowledge. Additionally, the required changes in input to trigger a backdoor are typically smaller than those for adversarial examples \cite{madry2018adversarial}.

\section{Black-Box Undetectable Backdoors from Digital Signatures}
The paper presents a method to plant a backdoor in any classifier using digital signatures, with robust properties that make it hard to detect or replicate without the backdoor key \cite{song2017machine}.
\par By leveraging digital signatures, the backdoor uses a public-private key pair. The public verification key $v_k$ identifies valid inputs, while the secret signing key $s_K$ allows the adversary to create specific inputs that trigger the backdoor, enabling control over the classifier’s output.
\par The modified classifier includes a verification mechanism for message-signature pairs. When a valid signed input (created with $s_k$) is detected, it overrides the classifier’s natural behavior.
\par The section culminates in a theorem. Assuming the existence of one-way functions, for every training procedure \texttt{Train}, there exists a model backdoor (\texttt{Backdoor}, \texttt{Activate}), which is non-replicable and black-box undetectable \cite{goldwasser2022backdoors}.

\section{Whitebox Undetectable Backdoors for Learning over Random Features}
For the second construction Goldwasser et al. (2022) present a template for planting backdoors in models trained on random features, ensuring they remain white-box undetectable \cite{liu2018trojaning}. Their approach identifies a computationally indistinguishable feature distribution with an embedded backdoor that can be selectively activated. The authors demonstrate this technique for two networks: 1-hidden-layer cosine and ReLU networks but speculate it can be applied on other distributions and network activations as well.  
\begin{enumerate} \itemsep -5pt
	\item \textbf{Random Fourier Features}: The first instantiation of the backdoor is shown using the Random Fourier Features (RFF) Framework, whose algorithm learns a 1-hidden-layer-cosine network. In RFF, the weights of the hidden layer are randomly sampled from an isotropic Gaussian distribution and passed through a cosine function with a random phase \cite{goldwasser2022backdoors}. Exploiting the Continues Learning With Errors (CLWE) problem, the authors substitute the Gaussian distribution with a computationally indistinguishable one that contains a secret backdoor key. Features sampled with the altered distribution are indistinguishable from those sampled by the genuine RFF algorithm but can be tampered with via the backdoor key \cite{song2017machine}.
	\item \textbf{Random ReLU Networks}: A 1-hidden-layer ReLU network can be backdoored exploiting the hardness of the sparse PCA problem. This allows a way to activate the backdoor through the sparse planted signal, increasing the variance of the inputs into the ReLU layer and by that to selectively manipulating the output \cite{madry2018adversarial}.
	\item \textbf{Strengths and limitations}: Despite being white-box undetectable, a key limitation of this technique is its high replicability. Once a single backdoored input has been seen, an observer can reuse the backdoor key to manipulate any other input. This highlights the importance to verify that true randomness was used during training, as otherwise, even models verified for using the specified training could still be compromised. Additionally, the presented construction demonstrates the potential to embed backdoors during the initialization phase that could persist subsequent optimizing using iterations of gradient descent. By this the authors emphasize the risk of backdoors and the need to further research the into designing them even more robust against post processing \cite{liu2018trojaning}.
\end{enumerate}

\section{Persistence Against Post-Processing}
The authors demonstrate that any neural network N can be efficiently transformed into a functionally identical network N’ that is immune to changes from gradient descent. This is achieved by constructing some error-correction for the weights of N’, making them persistent to iterations of gradient descent. This illustrates the extent of control a malicious trainer can maintain over a model \cite{madry2018adversarial}.

\section{Evaluation-Time Immunization of Backdoored Models}
Section II.4 E) looks at the immunization of backdoored models. This is achieved by preventing a hypothesis h from having adversarial examples up to a threshold $\sigma$. The hypothesis h is modified into a hypothesis $h*$ by "averaging" over values (count depending on $\sigma$) of h around the desired input point. This creates a smooth function as close input values cannot have vastly different outputs. Evaluation of h* can be achieved by a constant number of queries to h. This leads to theorem II.4 which states that based the assumptions that the ground truth and input distribution satisfy some smoothness condition, for any hypothesis h and any $\sigma > 0$ a function $h*$ can be evaluated such that  $H*$ is $\sigma$-robust: If $x,x'$ are of a distance smaller than $\sigma$, then $|h*(x)-h(x)|$ is very small. H* introduces only a small error: $h*$ is as close to f* as h is, up to some error that increase the larger $\sigma$ is.  If values with a distance larger than $\sigma$ can be perturbed by the hypothesis, the immunization guarantee does not hold \cite{song2017machine}.

\section{Related Work}
Related works include other approaches to singular topics of the paper such as adversarial robustness, i.e. developing learning algorithms that are robust to adversarial attacks \cite{madry2018adversarial}, backdoors in cryptography regarding the proposed method of inserting backdoors in machine learning models as well as embedding cryptography in neural networks, where this work uses the embedding of a public verification key in contrast to a hardcoded secret key \cite{liu2018trojaning}. Other works focus on similar goals by creating backdoors through the modification of training data or backdoors planted in neural network classifiers \cite{gu2017badnets}. Another work presents a method to produce a model which, while fitting the training data, mislabels all other inputs and is undistinguishable from a model that generalizes well \cite{goldwasser2022backdoors}.

\bibliography{references}

\end{document}

### 2024-12-04 PbD/paper/summary.tex
Added pbd paper.

@@
 \title{Planting Undetectable Backdoors in Machine Learning Models summarised}
-% \author{Leo Landowski, Noah Nehring, Leon Langhoff, Leopold Lemmermann}
+%\author{Leo Landowski \\ Leon Langhoff \\ Leopold Lemmermann \\ Noah Nehring }
 
@@
 The scientific paper "Planting Undetectable Backdoors in Machine Learning Models" by Shafi Goldwasser, Michael P.
-Kim, Vinod Vaikuntanathan, and Or Zamir (2022) investigates how malicious actors can embed undetectable backdoors in machine learning models to covertly manipulate their classifications \cite{goldwasser2022backdoors}.
+Kim, Vinod Vaikuntanathan, and Or Zamir \cite{goldwasser2022backdoors} investigates how malicious actors can embed undetectable backdoors in machine learning models to covertly manipulate their classifications.
+All the information presented in this introduction is based on the findings and arguments presented in the paper.
 The paper addresses the growing use of machine learning models throughout various domains.
-Companies can outsource tasks connected to these machine learning models to service providers, thereby democratizing access to machine learning models yet raising serious trust concerns \cite{gu2017badnets}.
+Companies can outsource tasks connected to these machine learning models to service providers, thereby democratizing access to machine learning models yet raising serious trust concerns.
 While using a service provider, a user must trust their claims about accuracy and robustness because, as the paper will present, a service provider can plant an undetectable backdoor into the model.
-The paper provides a cryptographic framework for understanding and constructing these backdoors and shows how a malicious entity can implement them \cite{liu2018trojaning}.
+The paper provides a cryptographic framework for understanding and constructing these backdoors and shows how a malicious entity can implement them.
 It doesn’t matter for the company if there is white or black-box access to the machine learning model; these backdoors are undetectable, and the already mentioned framework ensures the undetectability.
@@
 \par The paper defines multiple countermeasures a company can initiate, but it is important to mention that these can't guarantee immunity.
-The authors explore techniques to mitigate backdoors by ensuring trusted randomness or using verifiable delegation methods, where the training process is certified as correctly executed \cite{song2017machine}.
+The authors explore techniques to mitigate backdoors by ensuring trusted randomness or using verifiable delegation methods, where the training process is certified as correctly executed.
 To reduce backdoor impact, clients might apply post-processing methods like running a few iterations of gradient descent on the trained classifier.
 Although this approach cannot detect the backdoor directly, it could disrupt its functionality in some cases.
-The paper discusses adding random noise to inputs during evaluation (randomized smoothing) to make backdoors less effective \cite{madry2018adversarial}.
+The paper discusses adding random noise to inputs during evaluation (randomized smoothing) to make backdoors less effective.
 To implement these methods, a company needs to have the right personnel and knowledge, and still, there would be no guarantee that the model is backdoor-free.
@@
 \section{Defining Undetectable Backdoors}
-This section formally defines 'Undetectable Backdoors' and relevant components \cite{goldwasser2022backdoors}.
+This section formally defines 'Undetectable Backdoors' and relevant components.
 \begin{enumerate} \itemsep -5pt
@@
 The \texttt{Backdoor}, like \texttt{Train}, generates a simple classifier $\widetilde{h}$ which appears similar to $h$, and a backdoor key $bk$.
-This key $bk$ is then used to in the \texttt{Activate} function to modify an input $x$ into $x'$, for which $\widetilde{h}(x')\neq h(x)$ \cite{liu2018trojaning}.
+This key $bk$ is then used to in the \texttt{Activate} function to modify an input $x$ into $x'$, for which $\widetilde{h}(x')\neq h(x)$.
     \item \textbf{Undetectability}: A backdoor is undetectable, if the classifier $\widetilde{h}$ cannot be distinguished from $h$ by polynomial-time algorithms.
 It is divided further into \textit{black-box} and \textit{white-box undetectability}.
-\textit{Black-box undetectability} describes cases where it is not possible to find inputs where $\widetilde{h}$ diverges from $h$, even with oracle access to $\widetilde{h}$ (meaning you can apply the function, but do not know the implementation) \cite{carlini2017backdoor}.
-\textit{White-box undetectability} describes cases where even with full access to the classifier's code, the backdoor cannot be detected, $\widetilde{h}$ and $h$ remain indistinguishable \cite{goldwasser2022backdoors}.
+\textit{Black-box undetectability} describes cases where it is not possible to find inputs where $\widetilde{h}$ diverges from $h$, even with oracle access to $\widetilde{h}$ (meaning you can apply the function, but do not know the implementation).
+\textit{White-box undetectability} describes cases where even with full access to the classifier's code, the backdoor cannot be detected, $\widetilde{h}$ and $h$ remain indistinguishable.
 \end{enumerate}
@@
 The backdoor requires a specific backdoor key, restricting exploitation to individuals with this knowledge.
-Additionally, the required changes in input to trigger a backdoor are typically smaller than those for adversarial examples \cite{madry2018adversarial}.
+Additionally, the required changes in input to trigger a backdoor are typically smaller than those for adversarial examples.
 
 \section{Black-Box Undetectable Backdoors from Digital Signatures}
-The paper presents a method to plant a backdoor in any classifier using digital signatures, with robust properties that make it hard to detect or replicate without the backdoor key \cite{song2017machine}.
+The paper presents a method to plant a backdoor in any classifier using digital signatures, with robust properties that make it hard to detect or replicate without the backdoor key.
 \par By leveraging digital signatures, the backdoor uses a public-private key pair.
@@
 For the second construction Goldwasser et al.
-(2022) present a template for planting backdoors in models trained on random features, ensuring they remain white-box undetectable \cite{liu2018trojaning}.
+(2022) present a template for planting backdoors in models trained on random features, ensuring they remain white-box undetectable.
 Their approach identifies a computationally indistinguishable feature distribution with an embedded backdoor that can be selectively activated.
@@
 	\item \textbf{Random Fourier Features}: The first instantiation of the backdoor is shown using the Random Fourier Features (RFF) Framework, whose algorithm learns a 1-hidden-layer-cosine network.
-In RFF, the weights of the hidden layer are randomly sampled from an isotropic Gaussian distribution and passed through a cosine function with a random phase \cite{goldwasser2022backdoors}.
-Exploiting the Continues Learning With Errors (CLWE) problem, the authors substitute the Gaussian distribution with a computationally indistinguishable one that contains a secret backdoor key.
-Features sampled with the altered distribution are indistinguishable from those sampled by the genuine RFF algorithm but can be tampered with via the backdoor key \cite{song2017machine}.
-	\item \textbf{Random ReLU Networks}: A 1-hidden-layer ReLU network can be backdoored exploiting the hardness of the sparse PCA problem.
-This allows a way to activate the backdoor through the sparse planted signal, increasing the variance of the inputs into the ReLU layer and by that to selectively manipulating the output \cite{madry2018adversarial}.
+In RFF, the weights of the hidden layer are randomly sampled from an isotropic Gaussian distribution and passed through a cosine function with a random phase.
+Exploiting the Continues Learning With Errors (CLWE) problem \cite{bruna2021continuous}, the authors substitute the Gaussian distribution with a computationally indistinguishable one that contains a secret backdoor key.
+Features sampled with the altered distribution are indistinguishable from those sampled by the genuine RFF algorithm but can be tampered with via the backdoor key.
+	\item \textbf{Random ReLU Networks}: A 1-hidden-layer ReLU network can be backdoored by exploiting the hardness of the sparse PCA problem \cite{berthet2013complexity}, \cite{brennan2019optimal}.
+This allows a way to activate the backdoor through the sparse planted signal, which increases the variance of the inputs to the ReLU layer, thereby selectively manipulating the output.
 	\item \textbf{Strengths and limitations}: Despite being white-box undetectable, a key limitation of this technique is its high replicability.
@@
 Additionally, the presented construction demonstrates the potential to embed backdoors during the initialization phase that could persist subsequent optimizing using iterations of gradient descent.
-By this the authors emphasize the risk of backdoors and the need to further research the into designing them even more robust against post processing \cite{liu2018trojaning}.
+By this the authors emphasize the risk of backdoors and the need to further research the into designing them even more robust against post processing.
 \end{enumerate}
@@
 \section{Persistence Against Post-Processing}
-The authors demonstrate that any neural network N can be efficiently transformed into a functionally identical network N’ that is immune to changes from gradient descent.
-This is achieved by constructing some error-correction for the weights of N’, making them persistent to iterations of gradient descent.
-This illustrates the extent of control a malicious trainer can maintain over a model \cite{madry2018adversarial}.
+The authors demonstrate that any neural network $N$ can be efficiently transformed into a functionally identical network $N’$ that is immune to changes from gradient descent.
+This is achieved by constructing some error-correction for the weights of $N’$, making them persistent to iterations of gradient descent.
+This illustrates the extent of control a malicious trainer can maintain over a model.
 
@@
 Section II.4 E) looks at the immunization of backdoored models.
-This is achieved by preventing a hypothesis h from having adversarial examples up to a threshold $\sigma$.
-The hypothesis h is modified into a hypothesis $h*$ by "averaging" over values (count depending on $\sigma$) of h around the desired input point.
+This is achieved by preventing a hypothesis $h$ from having adversarial examples up to a threshold $\sigma$.
+The hypothesis $h$ is modified into a hypothesis $\tilde{h}$ by averaging over values (where the count is depending on $\sigma$) of $h$ around the desired input point.
 This creates a smooth function as close input values cannot have vastly different outputs.
-Evaluation of h* can be achieved by a constant number of queries to h.
-This leads to theorem II.4 which states that based the assumptions that the ground truth and input distribution satisfy some smoothness condition, for any hypothesis h and any $\sigma > 0$ a function $h*$ can be evaluated such that  $H*$ is $\sigma$-robust: If $x,x'$ are of a distance smaller than $\sigma$, then $|h*(x)-h(x)|$ is very small.
-H* introduces only a small error: $h*$ is as close to f* as h is, up to some error that increase the larger $\sigma$ is.
-If values with a distance larger than $\sigma$ can be perturbed by the hypothesis, the immunization guarantee does not hold \cite{song2017machine}.
+Evaluation of $\tilde{h}$ can be achieved by a constant number of queries to $h$.
+This leads to theorem II.4 which states that, based on the assumptions that the ground truth and input distribution satisfy some smoothness condition, for any hypothesis $h$ and any $\sigma > 0$ a function $\tilde{h}$ can be evaluated such that  $\tilde{h}$ is $\sigma$-robust: If $x,x'$ are of a distance smaller than $\sigma$, then $|\tilde{h}(x)-h(x)|$ is very small.
+$\tilde{h}$introduces only a small error: $\tilde{h}$ is as close to $f*$ as $h$ is, up to some error that increase the larger $\sigma$ is \cite{goldwasser2022backdoors}.
+If values with a distance larger than $\sigma$ can be perturbed by the hypothesis, the immunization guarantee does not hold.
 
@@
 Related works include other approaches to singular topics of the paper such as adversarial robustness, i.e.
-developing learning algorithms that are robust to adversarial attacks \cite{madry2018adversarial}, backdoors in cryptography regarding the proposed method of inserting backdoors in machine learning models as well as embedding cryptography in neural networks, where this work uses the embedding of a public verification key in contrast to a hardcoded secret key \cite{liu2018trojaning}.
-Other works focus on similar goals by creating backdoors through the modification of training data or backdoors planted in neural network classifiers \cite{gu2017badnets}.
-Another work presents a method to produce a model which, while fitting the training data, mislabels all other inputs and is undistinguishable from a model that generalizes well \cite{goldwasser2022backdoors}.
+developing learning algorithms that are robust to adversarial attacks \cite{pmlr-v117-garg20a},\cite{pmlr-v97-bubeck19a}, backdoors in cryptography regarding the proposed method of inserting backdoors in machine learning models \cite{4031390} as well as embedding cryptography in neural networks, where this work uses the embedding of a public verification key in contrast to a hardcoded secret key.
+Other works focus on similar goals by creating backdoors through the modification of training data or backdoors planted in neural network classifiers \cite{8685687}, \cite{NEURIPS2022_3538a22c} Another work presents a method to produce a model which, while fitting the training data, mislabels all other inputs and is undistinguishable from a model that generalizes well \cite{moitra2022spoofinggeneralizationcanttrust}.
 

### 2024-12-04 SbD/summary.tex
First PbD summary.
Updated sbd summary.

@@
 
-\includepdf[pages={14-16}, nup=1x3]{slides.pdf}
+\includepdf[pages={14-16,18-19,21-29,31-32,35,37-38,40-41}, nup=3x7]{slides.pdf}
 \part{Kryptographie}
-
 \section{Grundlagen}
@@
 \begin{enumerate}
-  \item \textbf{Ciphertext Only}: Angreifer kennt nur den Ciphertext
-  \item \textbf{Known Plaintext}: Angreifer kennt Ciphertext und Plaintext
-  \item \textbf{Chosen Plaintext} (adaptiv v. nicht-adaptiv): Angreifer kann Plaintext wählen und erhält den Ciphertext (nicht-adaptiv: alle Plaintexte müssen vorher gewählt werden)
+  \item \textbf{Ciphertext Only}: Angreifer kennt nur den Schlüsseltext
+  \item \textbf{Known Plaintext}: Angreifer kennt Schlüsseltext und Klartext
+  \item \textbf{Chosen Plaintext}: Angreifer kann Klartext wählen und erhält den Schlüsseltext (nicht-adaptiv: alle Klartexte müssen vorher gewählt werden)
 \end{enumerate}
@@
 \begin{enumerate}
-  \item \textbf{Vollständiges Brechen}: Schlüssel gefunden
-  \item \textbf{Universelles Brechen}: Äquivalenter Schlüssel gefunden
-  \item \textbf{Selektives Brechen}: Bestimmte Nachricht entschlüsselt
-  \item \textbf{Existenzielles Brechen}: Irgendeine Nachricht entschlüsselt
+  \item \textbf{Vollständig}: Schlüssel gefunden
+  \item \textbf{Universell}: Äquivalenter Schlüssel gefunden
+  \item \textbf{Selektiv}: Bestimmte Nachricht entschlüsselt
+  \item \textbf{Existenziell}: Irgendeine Nachricht entschlüsselt
 \end{enumerate}
 
-\includepdf[pages={18-32,35,37-41}, nup=3x7]{slides.pdf}
-\section{Klassische Chiffren}
+\subsection{Ziele}
+\begin{itemize}
+  \item \textbf{Verschlüsselung}: Vertraulichkeit
+  \item \textbf{Signatur}: Integrität, Authentizität
+\end{itemize}
+
+\subsection{Klassische Chiffren}
 \begin{itemize}
@@
 \includepdf[pages={44-108}, nup=4x9]{slides.pdf}
-\section{Moderne Kryptographie}
-\subsection{Symmetrische Systeme}
-\begin{enumerate}
-  \item \textbf{One-Time-Pad (mod 2)}: Vernam-Chiffre mit zufälligem Schlüssel
-  \item \textbf{Symmetrische Authentifikationscodes}: Erweiterung des One-Time-Pad mit Message Authentication Code (MAC)
-  \item \textbf{DES (Data Encryption Standard)}: 56-Bit-Blockchiffre nach Feistel-Prinzip
-  \item \textbf{IDEA (International Data Encryption Algorithm)}: 128-Bit-Blockchiffre mit XOR, Addition und Multiplikation
-  \item \textbf{AES (Advanced Encryption Standard)}: 128-Bit-Blockchiffre mit 10, 12 oder 14 Runden
-\end{enumerate}
-
-\subsubsection{Gütekriterien}
+\section{Symmetrische Systeme}
 \begin{itemize}
@@
   \item \textbf{Korrelationsimmunität}: Keine Korrelation zwischen Input und Output
-  \item sekundäre Kriterien: Implementierbarkeit, Längentreue, Schnelligkeit
+  \item \textit{sekundär}: Implementierbarkeit, Längentreue, Schnelligkeit
 \end{itemize}
 
-\subsubsection{Feistel-Prinzip}
-\begin{itemize}
-  \item \textbf{In Hälften teilen}: Klartext wird in zwei Hälften geteilt
-  \item \textbf{Runden}: Rundenfunktion wird auf eine Hälfte angewendet und das Ergebnis mit der anderen Hälfte XOR-verknüpft
-  \item \textbf{Austauschen der Hälften}: Die Hälften werden getauscht
-  \item \textbf{Dekodierung}: Entschlüsselung ist identisch zur Verschlüsselung (Runden umgekehrt)
-\end{itemize}
+\subsection{Feistel-Prinzip}
+\input{res/feistel.fig}
 
-\subsubsection{DES (Data Encryption Standard)}
+\subsection{DES (Data Encryption Standard)}
 \begin{itemize}
@@
 
-\subsubsection{IDEA (International Data Encryption Algorithm)}
+\subsection{IDEA (International Data Encryption Algorithm)}
 \begin{itemize}
@@
 
-\subsubsection{AES (Advanced Encryption Standard)}
+\subsection{AES (Advanced Encryption Standard)}
 \begin{itemize}
@@
 
-\subsubsection{Betriebsarten symmetrischer Blockchiffren}
+\subsection{Betriebsarten}
 \begin{enumerate}
@@
 \includepdf[pages={44, 109-142}, nup=4x9]{slides.pdf}
-\subsection{Asymmetrische Systeme}
-\begin{enumerate}
-  \item \textbf{Diffie-Hellman-Key-Exchange}: Schlüsselaustausch über unsicheren Kanal; Sicherheit durch diskretes Logarithmusproblem, anfällig für Man-in-the-Middle-Angriffe.
-  \item \textbf{ElGamal Kryptosystem}: Asymmetrische Verschlüsselung und Signatur; Diffie-Hellman-basiert, stärkere Sicherheit, aber höhere Schlüsselanforderungen.
-  \item \textbf{RSA zur Konzelation und Signatur}: Asymmetrische Verschlüsselung auf Faktorisierungsannahme; verbreitet in SSL/TLS, bietet Vertraulichkeit und Authentizität.
-  \item \textbf{Blinde Signaturen mit RSA}: Signatur ohne Kenntnis des Inhalts für Anonymität; Anwendung in elektronischen Abstimmungen und digitalen Währungen.
-  \item \textbf{Kryptosysteme auf Basis elliptischer Kurven}: Hohe Sicherheit bei kleinen Schlüsseln; effizient und ideal für mobile/räumlich beschränkte Geräte.
-\end{enumerate}
-
-\subsubsection{Mathematische Grundlagen}
+\section{Asymmetrische Systeme}
 \begin{itemize}
-  \item \textbf{Erweiterter Euklidischer Algorithmus}: Berechnung des ggT und inversen Elemente in Restklassenringen.
-  \item \textbf{Euler-Phi-Funktion}: $\varphi(n)$ gibt die Anzahl der zu $n$ teilerfremden Zahlen kleiner als $n$ an.
-  \item \textbf{Primitive Wurzel}: $g$ ist primitive Wurzel modulo $p$, wenn $g^i \mod p \neq 1$ für $i < p-1$
-  \item \textbf{Diskreter Logarithmus}: Exponent in der Primitiven Wurzel
-  \item \textbf{Faktorisierungsannahme}: Primfaktoren sind nicht leicht zu berechnen (auch nicht nachgewiesen schwer)
+  \item \textbf{Erweiterter Euklidischer Algorithmus}: ggT berechnen und erweiterter Algorithmus für Inverse
+    \[ c_k=\frac{r_{k-2}}{r_{k-1}} \quad r_k = r_{k-2} \pmod{r_{k-1}} \quad s_k=s_{k-1}+c_ks_{k-2} \quad t_k=t_{k-1}+c_kt_{k-2} \]
+  \item \textbf{Euler-Phi-Funktion}: $\varphi(n)$ gibt die Anzahl der zu $n$ teilerfremden (engl. coprime) Zahlen kleiner als $n$ an.
+    \[ \varphi(p \cdot q) = (p-1) \cdot (q-1) \quad \text{Primzahlen: }\varphi(p) = p-1 \]
+  \item \textbf{Primitive Wurzel}: Teilerfremdes $a$ ist primitive Wurzel modulo $n$:
+    \[ a^d \not\equiv 1 \pmod{n} \text{ für } 1 \leq d < \varphi(n) \]
+  \item \textbf{Diskreter Logarithmus}: $c$ ist diskreter Logarithmus von $a$ zur Basis $b$ modulo $n$:
+    \[ b^c \equiv a \pmod{n} \to c = \log_b a \pmod{n} \]
+  \item \textbf{Faktorisierungsannahme}: Primzahlen $p$ und $q$ sind schwer aus $n=p \cdot q$ zu berechnen.
 \end{itemize}
 
-\subsubsection{Diffie-Hellman-Key-Exchange}
+\subsection{Diffie-Hellman Key-Exchange}
+\input{res/diffiehellman.fig}
+
+\subsection{ElGamal Verschlüsselung}
+\textit{Basiert auf diskretem Logarithmus und Diffie-Hellman Key-Exchange. Einsatz zB. in GnuPG, PGP.}
 \begin{itemize}
-  \item \textbf{Ziel}: Ermöglicht sicheren Schlüsselaustausch zwischen zwei Parteien über unsichere Kanäle ohne vorherigen geheimen Austausch.
-  \item \textbf{Funktionsweise}: Berechnung gemeinsamer geheimer Schlüssel über Modulorechnung basierend auf einem öffentlichen Basiswert und Primzahl.
-  \item \textbf{Sicherheitsprinzip}: Sicherheit basiert auf der Schwierigkeit des diskreten Logarithmusproblems.
-  \item \textbf{Schwachstelle}: Anfällig für Man-in-the-Middle-Angriffe, wenn nicht authentifiziert.
+  \item \textbf{Parameter}
+    \[
+      \underbrace{p\in\mathbb{P}, \overbrace{a}^\text{primitive Wurzel}}_\text{global}\quad
+      \underbrace{\overbrace{k}^\text{zufällig}\in\mathbb{Z}^*_{p-1}}_\text{privater Schlüssel}\quad
+      \underbrace{y = a^{-k} \pmod{p}}_\text{öffentlicher Schlüssel} 
+    \]
+  \item \textbf{Verschlüsselung} von Nachricht $m\in\mathbb{Z}^*_{p-1}$ für $B$: $k \in k<p$ zufällig wählen
+    \[ d = a^z \mod{p} \text{ und } c = y_B^z \cdot m \pmod{p} \]
+  \item \textbf{Entschlüsselung} von $a^z \mod{p}$ und $c$
+    \[ m^* = d^{k_B} \cdot c \pmod{p} \]
 \end{itemize}
 
-\subsubsection{ElGamal Kryptosystem}
+\subsection{ElGamal Signatur}
+\textit{Basiert auf diskretem Logarithmus. Einsatz zB. im Digital Signature Algorithm (DSA).}
 \begin{itemize}
-  \item \textbf{Verwendung}: Für Verschlüsselung und digitale Signaturen geeignet.
-  \item \textbf{Schlüssellänge}: Bietet längere Schlüssel für höhere Sicherheit, aber weniger Effizienz im Vergleich zu RSA.
-  \item \textbf{Sicherheitsprinzip}: Beruht auf der Schwierigkeit des diskreten Logarithmusproblems.
-  \item \textbf{Vorteile}: Flexibel und erweiterbar auf elliptische Kurven für mehr Effizienz.
+  \item \textbf{Parameter}
+    \[
+      \underbrace{p\in\mathbb{P}, \overbrace{a}^\text{primitive Wurzel}}_\text{global}\quad
+      \underbrace{\overbrace{x}^\text{zufällig} \in \mathbb{Z}^*_{p-1}}_\text{privater Schlüssel}\quad
+      \underbrace{y = a^{x} \pmod{p}}_\text{öffentlicher Schlüssel}
+    \]
+  \item \textbf{Signatur} von Nachricht $m\in\mathbb{Z}^*_{p-1}$ für $A$: $z\in\mathbb{Z}^*_{p-1}$ zufällig wählen
+    \[ r = a^z \mod{p} \text{ und } s = k^{-1} \cdot (m - x_A \cdot r) \mod{p-1} \]
+  \item \textbf{Verifikation} von $a^z \mod{p}$ und $r$ und $s$
+    \[ v = (y^r \cdot r^s) \mod{p} \text{ entspricht } a^m \mod{p} \to \text{Signatur gültig} \]
 \end{itemize}
 
-\subsubsection{RSA zur Konzelation und Signatur}
+\subsection{Rivest-Shamir-Adleman (RSA) Verschlüsselung}
+\textit{Basiert auf Faktorisierungsannahme. Einsatz zB. in TLS, PGP, S/MIME.}
 \begin{itemize}
-  \item \textbf{Verwendung}: Verschlüsselung und digitale Signatur mit öffentlichen und privaten Schlüsseln.
-  \item \textbf{Schlüssellänge}: Variable Schlüssellänge (typisch 1024, 2048 oder 4096 Bit), längere Schlüssel bieten höhere Sicherheit.
-  \item \textbf{Grundlage}: Faktorisierungsannahme, d.h., die Schwierigkeit der Primfaktorzerlegung großer Zahlen.
-  \item \textbf{Anwendung}: Weltweit in SSL/TLS und anderen Sicherheitssystemen implementiert.
+  \item \textbf{Parameter}
+    \[
+      \underbrace{p,q\in\mathbb{P}}_\text{geheim}\quad
+      \underbrace{n=p\cdot q}_\text{öffentlich}\quad
+      \underbrace{\varphi(n)=(p-1)\cdot(q-1)}_\text{geheim}\quad
+      \underbrace{c\in\mathbb{Z}^*_{\varphi(n)}}_\text{öffentlich}\quad
+      \underbrace{d=c^{-1} \pmod{\varphi(n)}}_\text{geheim}
+    \]
+  \item \textbf{Verschlüsselung} von Nachricht $m\in\mathbb{Z}_n$ für $B$
+    \[ c = m^e \mod{n} \]
+  \item \textbf{Entschlüsselung} von $c$
+    \[ m = c^d \mod{n} \]
+  \item \textbf{Signatur}: analog zur Verschlüsselung. Signatur $s$ wird anstelle von $c$ übertragen.
+  \item \textbf{Verifikation}: analog zur Entschlüsselung. Verifikation von $s$ anstelle von $m$.
+  \item \textbf{Blinde Signatur} von $m$ für $A$: $r\in\mathbb{Z}_n$ zufällig wählen, Nachricht mit $r$ blenden, signieren lassen und entblenden.
 \end{itemize}
 
-\subsubsection{Blinde Signaturen mit RSA}
-\begin{itemize}
-  \item \textbf{Prinzip}: Der Signierer erstellt eine Signatur ohne den Inhalt der Nachricht zu kennen (Anonymität).
-  \item \textbf{Verfahren}: Der Verfasser maskiert die Nachricht, bevor sie zur Signatur übergeben wird, und entfernt die Maske nach der Signierung.
-  \item \textbf{Anwendung}: Nützlich in Szenarien wie elektronischen Abstimmungen, bei denen die Vertraulichkeit wichtig ist.
-\end{itemize}
-
-\subsubsection{Kryptosysteme auf Basis elliptischer Kurven}
-\begin{itemize}
-  \item \textbf{Effizienz}: Höhere Sicherheit bei kleineren Schlüssellängen (z.B. ein 256-Bit-Schlüssel bietet Sicherheit ähnlich einem 3072-Bit-RSA-Schlüssel).
-  \item \textbf{Grundlage}: Elliptische Kurven über endlichen Feldern, basierend auf dem diskreten Logarithmusproblem.
-  \item \textbf{Verwendung}: Ideal für mobile und ressourcenbeschränkte Geräte, oft verwendet in modernen Sicherheitsprotokollen (z.B. ECDSA für digitale Signaturen).
-\end{itemize}
+\subsection{Elliptic Curve Cryptography (ECC)}
+% TODO: Add ECC
 

### 2024-12-05 NLP/assignments/res/app.py
Cleaned up.

@@ -1,51 +1,0 @@
-from flask import Flask,render_template,url_for,request
-import pandas as pd
-import spacy
-nlp = spacy.load('en_core_web_sm')
-
-app = Flask(__name__)
-
-@app.route('/')
-def index():
-	return render_template("index.html")
-
-@app.route('/process',methods=["POST"])
-def process():
-	results = pd.Series([])	
-	num_of_results = 0
-	ORG_named_entity = pd.Series([])
-	PERSON_named_entity = pd.Series([])
-	GPE_named_entity = pd.Series([])
-	MONEY_named_entity = pd.Series([])
-	if request.method == 'POST':
-		choice = request.form['taskoption']
-		rawtext = request.form['rawtext']
-		if  choice == 'Select Task':
-			return render_template("index.html",results=["Please select the type of Entity"],num_of_results = 0)
-		doc = nlp(rawtext)
-		d = []
-		for ent in doc.ents:
-			d.append((ent.label_, ent.text))
-			df = pd.DataFrame(d, columns=('named entity', 'output'))
-			ORG_named_entity = df.loc[df['named entity'] == 'ORG']['output']
-			PERSON_named_entity = df.loc[df['named entity'] == 'PERSON']['output']
-			GPE_named_entity = df.loc[df['named entity'] == 'GPE']['output']
-			MONEY_named_entity = df.loc[df['named entity'] == 'MONEY']['output']
-		if choice == 'organization':
-			results = ORG_named_entity
-			num_of_results = len(results)
-		elif choice == 'person':
-			results = PERSON_named_entity
-			num_of_results = len(results)
-		elif choice == 'geopolitical':
-			results = GPE_named_entity
-			num_of_results = len(results)
-		elif choice == 'money':
-			results = MONEY_named_entity
-			num_of_results = len(results)
-
-	return render_template("index.html",results=results,num_of_results = num_of_results)
-
-
-if __name__ == '__main__':
-	app.run(debug=True, port=5001)

### 2024-12-05 NLP/exercises.tex
Fixed section numbers.
Added new NLP practice.
Added current state.

@@
 
-\sheet{2023}{1st exam}
+\sheet{Lecture questions}
+
+\begin{exercise}{Web as an application area for NLP}
+  The web is an application area for NLP. Provide examples of applications.
+
+  \begin{solution}
+    Examples include:
+    \begin{itemize}
+        \item Internet of Services
+        \item Community mining
+        \item Information retrieval
+        \item Machine translation for web-based content
+    \end{itemize}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Web as a resource for NLP}
+  How can the web improve the quality of NLP? Provide examples.
+
+  \begin{solution}
+    Examples include:
+    \begin{itemize}
+        \item Web as Corpus
+        \item Analyzing web-based knowledge repositories
+        \item Extracting structured information from unstructured web data
+        \item Using web-scale data for training language models
+    \end{itemize}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Morphological analysis}
+  Identify all stems and affixes (prefix, suffix, infix, circumfix) in the following words: "index," "incorrect," "interesting."
+
+  \begin{solution}
+    Morphological breakdown:
+    \begin{itemize}
+        \item **index:** stem = "index"
+        \item **incorrect:** prefix = "in," stem = "correct"
+        \item **interesting:** stem = "interest," suffix = "ing"
+    \end{itemize}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Stemming vs lemmatization}
+  In contrast to lemmatization, stemming does not necessarily return a valid word form. Why is stemming still useful?
+
+  \begin{solution}
+    Stemming is still useful because:
+    \begin{itemize}
+        \item It is faster and computationally less expensive.
+        \item It simplifies implementation for applications like Information Retrieval.
+        \item It can still group similar words under a common base, which is often sufficient.
+    \end{itemize}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Corpus definition and usage}
+  What is a corpus, and how can it be built and used for linguistic analysis?
+
+  \begin{solution}
+    \begin{itemize}
+        \item **Corpus definition:** A corpus is a large and structured set of texts, often used for linguistic research.
+        \item **How to build:** Collect text data, preprocess it (cleaning, normalization), and annotate it for specific tasks.
+        \item **Usage:** Analyze linguistic patterns, train NLP models, and evaluate machine learning algorithms.
+    \end{itemize}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{NLP datasets}
+  What are NLP datasets, and why are they important?
+
+  \begin{solution}
+    \begin{itemize}
+        \item **Definition:** NLP datasets are collections of annotated or unannotated text data designed for specific NLP tasks (e.g., sentiment analysis, translation).
+        \item **Importance:**
+        \begin{itemize}
+            \item Serve as a foundation for training and testing machine learning models.
+            \item Enable reproducibility in research.
+            \item Facilitate comparisons between models and techniques.
+        \end{itemize}
+    \end{itemize}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Corpus and dataset annotation}
+  What is the annotation process for a corpus or dataset, and how does it impact machine learning applications?
+
+  \begin{solution}
+    \begin{itemize}
+        \item **Annotation process:** Adding labels or metadata to text data (e.g., part-of-speech tags, sentiment labels, named entities).
+        \item **Impact on machine learning:**
+        \begin{itemize}
+            \item Improves model performance by providing high-quality training data.
+            \item Enables supervised learning by associating input with output labels.
+            \item Impacts scalability and generalization of models depending on annotation quality.
+        \end{itemize}
+    \end{itemize}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Large-scale dataset collection using crowdsourcing}
+  What are the main strategies for large-scale dataset collection using crowdsourcing?
+
+  \begin{solution}
+    Strategies include:
+    \begin{itemize}
+        \item **Task design:** Clearly define annotation tasks with examples and guidelines.
+        \item **Platform selection:** Use platforms like Amazon Mechanical Turk or Prolific.
+        \item **Quality control:** Use redundancy (multiple annotators per task) and include test questions to ensure reliability.
+        \item **Data aggregation:** Combine annotations using majority voting or statistical models.
+    \end{itemize}
+  \end{solution}
+\end{exercise}
+
+\setcounter{section}{2022}
+\sheet{1st exam}
+
 \begin{exercise}{Tokenization}
@@
   \begin{solution}
-    % TODO
+    Examples of ambiguity in tokenization:
+    \begin{enumerate}
+        \item **Hyphenated words:** Does "state-of-the-art" count as one token or three?
+        \item **Contractions:** Should "don't" be treated as one token or split into "do" and "n't"?
+        \item **Punctuation:** How should punctuation like periods in "U.S." or apostrophes in possessives ("John's book") be handled?
+    \end{enumerate}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    For the sentence "John shot the man with the gun":
+    \begin{enumerate}
+        \item **Parse trees:**
+            \begin{itemize}
+                \item Tree 1: "with the gun" modifies "the man" (John shot a man who had the gun).
+                \item Tree 2: "with the gun" modifies "shot" (John used a gun to shoot the man).
+            \end{itemize}
+        \item **Syntactic ambiguity:** 
+        \begin{itemize}
+            \item **Attachment ambiguity:** Whether "with the gun" attaches to "shot" or "the man."
+        \end{itemize}
+    \end{enumerate}
   \end{solution}
@@
 \begin{exercise}{Stemming and lemmatization}
-  Difference between stemming and lemmatization.
+  What is the difference between stemming and lemmatization?
 
   \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item **Stemming:** Reduces a word to its base form by chopping off prefixes or suffixes. Example: "running" → "run".
+        \item **Lemmatization:** Reduces a word to its dictionary form (lemma), considering the word's context and grammar. Example: "better" → "good."
+    \end{itemize}
+    Lemmatization is more precise than stemming but computationally more expensive.
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    Issues when using the web as a corpus:
+    \begin{enumerate}
+        \item **Noisy data:** Content may contain errors, irrelevant information, or non-standard grammar.
+        \item **Duplication:** Identical or nearly identical data appears multiple times.
+        \item **Bias:** Web content often reflects societal biases, which can skew analyses.
+        \item **Dynamic content:** Web pages change frequently, making reproducibility challenging.
+    \end{enumerate}
   \end{solution}
@@
 \begin{exercise}{Cross-language QA}
-  How does cross-language QA work? Issues?
+  How does cross-language QA work? What are its issues?
 
   \begin{solution}
-    % TODO
+    Cross-language QA involves answering questions in one language based on information in another.
+    \begin{itemize}
+        \item **Process:** Translate either the query or the documents to a common language, perform QA, and translate the result back.
+        \item **Issues:** 
+        \begin{itemize}
+            \item Loss of information in translation.
+            \item Mismatches between linguistic structures of languages.
+            \item Cultural differences in phrasing questions or answers.
+        \end{itemize}
+    \end{itemize}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    Main components of a machine learning pipeline:
+    \begin{enumerate}
+        \item **Data preprocessing:** Cleaning, normalizing, and transforming raw data.
+        \item **Feature extraction:** Selecting or engineering features that represent the data.
+        \item **Model training:** Using labeled data to train the algorithm.
+        \item **Evaluation:** Measuring performance using metrics like accuracy or F1-score.
+        \item **Deployment:** Integrating the trained model into production systems.
+    \end{enumerate}
   \end{solution}
@@
 \begin{exercise}{IR evaluation}
-  What are the problems with precision, recall, and f-measure in IR? How to improve and why is as an example cohen kappa better?
+  What are the problems with precision, recall, and f-measure in IR? How can these be improved? Why is Cohen's Kappa better?
 
   \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item **Precision:** Biased toward over-selecting relevant documents.
+        \item **Recall:** Favors retrieving more documents, even if irrelevant.
+        \item **F-measure:** Averages precision and recall but doesn’t prioritize false positives or negatives.
+        \item **Cohen's Kappa:** Measures inter-rater reliability and incorporates agreement by chance, making it better for comparing systems.
+    \end{itemize}
   \end{solution}
@@
 \begin{exercise}{IR evaluation}
-  Calculate P@5, P@10, MAP for the given IR ranked result list (one for IR system A and one for B. Both go up to ten entries and either have a checkmark or an x in each rank).
+  Calculate P@5, P@10, and MAP for the given IR ranked result list. Analyze both IR system A and system B.
 
   \begin{solution}
-    % TODO
+    For P@5, P@10, and MAP:
+    \begin{itemize}
+        \item **P@5:** Count relevant documents in the top 5 results and divide by 5.
+        \item **P@10:** Same calculation for the top 10 results.
+        \item **MAP:** Average Precision for each query, then average across all queries.
+    \end{itemize}
+    Steps:
+    \begin{itemize}
+        \item Identify positions of relevant documents in ranked lists.
+        \item Calculate precision at each relevant position.
+        \item Take mean of these precision values.
+    \end{itemize}
   \end{solution}
@@
 \begin{exercise}{Naive Bayes and SVM}
-  Which of the following 4 are correct
+  Which of the following statements are correct?
   \begin{enumerate}
-    \item Features for Naive Bayes classifiers do not need to be uncorrelated
-    \item Naive Bayes is slow to train
-    \item SVM itself is a multi-label classifier
-    \item SVM handles sparse data well
+    \item Features for Naive Bayes classifiers do not need to be uncorrelated.
+    \item Naive Bayes is slow to train.
+    \item SVM itself is a multi-label classifier.
+    \item SVM handles sparse data well.
   \end{enumerate}
@@
   \begin{solution}
-    % TODO
+    Correct statements:
+    \begin{enumerate}
+        \item **True:** Features for Naive Bayes do not need to be uncorrelated, though it assumes independence for computation.
+        \item **False:** Naive Bayes is fast to train, not slow.
+        \item **False:** SVM is not inherently multi-label but can be adapted for multi-label classification.
+        \item **True:** SVM handles sparse data well due to its kernel functions.
+    \end{enumerate}
   \end{solution}
@@
 \begin{exercise}{Corpus construction}
-  How to construct a corpus for linguistic purposes from web data and how to make it usable?
+  How can a corpus be constructed for linguistic purposes from web data, and how can it be made usable?
 
   \begin{solution}
-    % TODO
+    Steps for constructing a corpus:
+    \begin{enumerate}
+        \item **Data collection:** Crawl web pages, scrape data, or gather domain-specific text.
+        \item **Cleaning:** Remove duplicates, irrelevant content, and normalize text.
+        \item **Annotation:** Add metadata or labels for tasks like classification or parsing.
+        \item **Usability:** Store in accessible formats (e.g., JSON, XML) and document the corpus for ease of use.
+    \end{enumerate}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    Challenges in pragmatics and discourse analysis:
+    \begin{itemize}
+        \item **Context dependency:** Understanding depends heavily on prior sentences or external knowledge.
+        \item **Ambiguity:** Words and phrases can have multiple meanings depending on the situation.
+        \item **Social norms:** Interpretations vary based on cultural or societal factors.
+    \end{itemize}
   \end{solution}
@@
 \begin{exercise}{Hate speech}
-  Define hate speech and how to construct a hate speech classifier
+  Define hate speech and explain how to construct a hate speech classifier.
 
   \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item **Definition:** Hate speech involves language that attacks or discriminates against individuals or groups based on attributes like race, religion, ethnicity, gender, or sexual orientation.
+        \item **Constructing a hate speech classifier:**
+        \begin{enumerate}
+            \item Collect labeled datasets with examples of hate speech and non-hate speech.
+            \item Preprocess text (e.g., tokenization, removing stop words).
+            \item Extract features (e.g., bag-of-words, TF-IDF, embeddings).
+            \item Train a supervised machine learning model (e.g., Naive Bayes, SVM, or deep learning models like LSTMs or transformers).
+            \item Evaluate the classifier using metrics like precision, recall, and F1-score.
+        \end{enumerate}
+    \end{itemize}
   \end{solution}
@@
 \begin{exercise}{Counter hate speech}
-  What is counter hate speech and how to construct such a system
+  What is counter hate speech, and how can a system be constructed for it?
 
   \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item **Definition:** Counter hate speech is content designed to reduce the impact of hate speech, often by promoting positive, empathetic, or factual responses.
+        \item **Constructing a counter hate speech system:**
+        \begin{enumerate}
+            \item Collect datasets of effective counter hate responses.
+            \item Train models to generate or retrieve appropriate responses based on the hate speech content.
+            \item Use sentiment analysis and empathy detection to craft responses.
+            \item Deploy the system in platforms like social media or forums, ensuring ethical considerations are addressed.
+        \end{enumerate}
+    \end{itemize}
   \end{solution}
@@
 \begin{exercise}{Crowdsourced annotation}
-  Crowdsourced annotation workflow
+  What is the workflow for crowdsourced annotation?
 
   \begin{solution}
-    % TODO
+    Steps in a crowdsourced annotation workflow:
+    \begin{enumerate}
+        \item **Task design:** Define the annotation task clearly with examples and instructions.
+        \item **Platform setup:** Use a crowdsourcing platform (e.g., Amazon Mechanical Turk, Prolific).
+        \item **Quality control:** Include test questions and redundancy (multiple annotators per item) to ensure data reliability.
+        \item **Data aggregation:** Aggregate annotations using majority voting or weighted schemes.
+        \item **Evaluation:** Analyze annotation quality and address inconsistencies.
+    \end{enumerate}
   \end{solution}
@@
 \begin{exercise}{Web corpus}
-  Why are duplicates on the web and how to detect
+  Why do duplicates exist on the web, and how can they be detected?
 
   \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item **Duplicates:** Duplicates arise from mirrored websites, copied content, or redundant crawls.
+        \item **Detection methods:**
+        \begin{itemize}
+            \item Hashing (e.g., MD5 or SHA): Compare hash values of documents.
+            \item Similarity measures: Use cosine similarity or Jaccard index on n-grams.
+            \item Structural analysis: Detect similarity in HTML structure.
+        \end{itemize}
+    \end{itemize}
   \end{solution}
@@
 \begin{exercise}{QA vs IR}
-  Differences in QA vs IR
+  What are the differences between Question Answering (QA) and Information Retrieval (IR)?
 
   \begin{solution}
-    % TODO
+    Differences between QA (Question Answering) and IR (Information Retrieval):
+    \begin{itemize}
+        \item **Goal:**
+        \begin{itemize}
+            \item QA aims to provide direct answers.
+            \item IR retrieves documents containing relevant information.
+        \end{itemize}
+        \item **Processing:**
+        \begin{itemize}
+            \item QA requires deeper natural language understanding.
+            \item IR focuses on keyword matching and ranking.
+        \end{itemize}
+        \item **Applications:**
+        \begin{itemize}
+            \item QA: Chatbots, customer support.
+            \item IR: Search engines.
+        \end{itemize}
+    \end{itemize}
   \end{solution}
@@
 \begin{exercise}{Cross-language QA}
-  Explain applications of multi-language QA and difficulties?
+  What are the applications of multi-language QA, and what difficulties arise?
 
   \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item **Applications:** Multi-language QA systems are used in global customer support, cross-border e-commerce, and multi-language search engines.
+        \item **Difficulties:**
+        \begin{itemize}
+            \item Translation errors can distort queries or results.
+            \item Cultural and linguistic differences complicate understanding and generating answers.
+            \item Lack of high-quality bilingual corpora for training.
+        \end{itemize}
+    \end{itemize}
   \end{solution}
@@
 \begin{exercise}{Language detection}
-  Explain language detection
+  How does language detection work?
 
   \begin{solution}
-    % TODO
+    Approaches for language detection:
+    \begin{enumerate}
+        \item **N-gram analysis:** Compare n-grams of text against language-specific frequency profiles.
+        \item **Machine learning models:** Use classifiers trained on labeled text for different languages.
+        \item **Lexicon-based methods:** Match words in the text to dictionaries of different languages.
+        \item **Hybrid approaches:** Combine multiple methods for higher accuracy.
+    \end{enumerate}
   \end{solution}
@@
 \begin{exercise}{Factual QA vs opinionated QA}
-  Differences between factual QA and opinionated QA, and which is easier to evaluate?
+  What are the differences between factual QA and opinionated QA, and which is easier to evaluate?
 
   \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item **Factual QA:** Deals with objective information (e.g., "What is the capital of France?").
+        \item **Opinionated QA:** Involves subjective answers (e.g., "Is Paris a good travel destination?").
+        \item **Evaluation:** Factual QA is easier to evaluate as answers can be verified against ground truth. Opinionated QA requires subjective or aggregated assessments.
+    \end{itemize}
   \end{solution}
@@
 \begin{exercise}{RAG and LangChain}
-  What's RAG and how does LangChain help?
+  What is Retrieval-Augmented Generation (RAG), and how does LangChain assist in implementing it?
 
   \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item **RAG (Retrieval-Augmented Generation):** Combines retrieval models with generative models to produce context-aware answers.
+External knowledge is retrieved and fed into the generation model.
+        \item **LangChain:** A framework for building applications with large language models, offering tools for integrating RAG systems.
+It manages chaining retrieval and generation steps efficiently.
+    \end{itemize}
   \end{solution}
@@
 \begin{exercise}{Sentiment analysis}
-  Discuss: Sentiment analysis using supervised vs. lexicon based approaches.
+  Compare sentiment analysis using supervised approaches and lexicon-based approaches.
 
   \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item **Supervised learning:** Uses labeled data to train machine learning models.
+        \begin{itemize}
+            \item **Advantages:** High accuracy with large datasets.
+            \item **Drawbacks:** Relies on annotated data and is domain-dependent.
+        \end{itemize}
+        \item **Lexicon-based approaches:** Use sentiment dictionaries to determine polarity.
+        \begin{itemize}
+            \item **Advantages:** Simplicity and no training requirement.
+            \item **Drawbacks:** Limited contextual understanding and poor performance on complex text.
+        \end{itemize}
+    \end{itemize}
   \end{solution}
@@
 \begin{exercise}{Edit distance}
-  Calculate the Minimum Edit Distance for "TAC" vs. "CAT" (given is also a table just like on the slides, just without the numbers filled in). List all possible word outcomes (in this case 4).
+  Calculate the Minimum Edit Distance for "TAC" vs. "CAT" and list all possible word outcomes.
 
   \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item **Minimum Edit Distance for "TAC" and "CAT":**
+        \[
+        \begin{array}{c|c|c|c|c}
+            &  & C & A & T \\
+            \hline
+            & 0 & 1 & 2 & 3 \\
+            T & 1 & 1 & 2 & 2 \\
+            A & 2 & 2 & 1 & 2 \\
+            C & 3 & 2 & 2 & 2 \\
+        \end{array}
+        \]
+        \item **Word outcomes:** Possible outcomes from edits:
+        \begin{enumerate}
+            \item CAT
+            \item TAT
+            \item TAC
+            \item CAC
+        \end{enumerate}
+    \end{itemize}
   \end{solution}
@@
 
-\sheet{2022}{1st exam}
+\setcounter{section}{2021}
+\sheet{1st exam}
 
 \begin{exercise}{Corpus construction}
-  How would you construct a corpus for linguistic purposes and how to make it usable?
+  How would you construct a corpus for linguistic purposes and how can it be made usable?
 
   \begin{solution}
-    % TODO
+    \begin{enumerate}
+        \item **Data collection:** Gather data from relevant sources (e.g., web crawling, domain-specific text).
+        \item **Cleaning and preprocessing:** Remove duplicates, non-linguistic content, and normalize text (e.g., case normalization, tokenization).
+        \item **Annotation:** Add metadata or linguistic labels (e.g., POS tags, syntactic trees) to the text.
+        \item **Storage:** Store the corpus in an accessible format like JSON, XML, or relational databases.
+        \item **Documentation:** Provide guidelines and metadata about the corpus for user understanding.
+    \end{enumerate}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item **Context dependency:** Requires understanding of the broader context beyond sentence boundaries.
+        \item **Ambiguity:** Words or phrases can have multiple meanings depending on the discourse.
+        \item **Cultural and social norms:** Interpretation is influenced by societal and cultural factors.
+        \item **Incomplete information:** Missing parts of conversations or text make it harder to analyze.
+    \end{itemize}
   \end{solution}
@@
 \begin{exercise}{Sequence tagging and CRF}
-  Define sequence tagging and what is CRF
+  Define sequence tagging and explain what a Conditional Random Field (CRF) is.
 
   \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item **Sequence tagging:** The process of assigning labels to each element in a sequence (e.g., part-of-speech tagging, named entity recognition).
+        \item **Conditional Random Field (CRF):** A probabilistic model used for structured prediction in sequence data. It considers the context of neighboring labels to improve accuracy.
+    \end{itemize}
   \end{solution}
@@
 \begin{exercise}{Minimum edit distance}
-  Calculate the Minimum Edit Distance for "TAC" vs. "CAT" (given is also a table just like on the slides, just without the numbers filled in). List all possible word outcomes (in this case 4).
+  Calculate the Minimum Edit Distance for "TAC" vs. "CAT" (a table is given). List all possible outcomes.
 
   \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item **Minimum Edit Distance Table:**
+        \[
+        \begin{array}{c|c|c|c|c}
+            &  & C & A & T \\
+            \hline
+            & 0 & 1 & 2 & 3 \\
+            T & 1 & 1 & 2 & 2 \\
+            A & 2 & 2 & 1 & 2 \\
+            C & 3 & 2 & 2 & 2 \\
+        \end{array}
+        \]
+        \item **Outcomes:**
+        \begin{enumerate}
+            \item CAT
+            \item TAT
+            \item TAC
+            \item CAC
+        \end{enumerate}
+    \end{itemize}
   \end{solution}
@@
 \begin{exercise}{IR evaluation}
-  Calculate P@5 and MAP for the given IR ranked result list (one for IR system A and one for B. Both go up to ten entries and either have a checkmark or an x in each rank).
+  Calculate P@5 and MAP for given ranked IR results for two systems.
 
   \begin{solution}
-    % TODO
-  \end{solution}
-\end{exercise}
-
-\begin{exercise}{Sequence tagging and CRF}
-  Define sequence tagging and what is CRF
-
-  \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item **P@5:** Count relevant documents in the top 5 results and divide by 5.
+        \item **MAP:** For each query, calculate average precision across relevant documents and average these scores.
+    \end{itemize}
   \end{solution}
@@
 \begin{exercise}{Hate speech}
-  Define hate speech and how to construct a hate speech classifier
+  Define hate speech and how to construct a hate speech classifier.
 
   \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item **Definition:** Hate speech targets individuals or groups based on attributes like race, religion, gender, or orientation.
+        \item **Classifier construction:**
+        \begin{enumerate}
+            \item Collect labeled datasets of hate and non-hate speech.
+            \item Preprocess data by removing noise and tokenizing.
+            \item Train a machine learning model (e.g., Naive Bayes, SVM, transformers).
+            \item Evaluate using metrics like F1-score, precision, and recall.
+        \end{enumerate}
+    \end{itemize}
   \end{solution}
@@
 \begin{exercise}{Counter hate speech}
-  What is counter hate speech and how to construct such a system
+  What is counter hate speech, and how can a system be constructed for it?
 
   \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item **Definition:** Counter hate speech neutralizes hate content by promoting positive or factual responses.
+        \item **System construction:**
+        \begin{enumerate}
+            \item Collect examples of effective counter hate responses.
+            \item Use NLP techniques to classify hate speech and generate counter-responses.
+            \item Train models on empathy and sentiment analysis for response generation.
+        \end{enumerate}
+    \end{itemize}
   \end{solution}
@@
 
+\begin{exercise}{Sentiment lexicon extension}
+  How can sentiment analysis be improved using lexicon extensions from distributed thesauruses?
+
+  \begin{solution}
+    \begin{itemize}
+        \item **Lexicon extension:** Expand sentiment dictionaries using related terms from distributed word embeddings like Word2Vec or GloVe.
+        \item **Process:**
+        \begin{enumerate}
+            \item Identify synonyms, antonyms, or similar words in embedding spaces.
+            \item Update lexicons with these terms to improve coverage.
+            \item Validate new terms to ensure sentiment accuracy.
+        \end{enumerate}
+    \end{itemize}
+  \end{solution}
+\end{exercise}
 \begin{exercise}{Sentiment analysis}
@@
   \begin{solution}
-    % TODO
-  \end{solution}
-\end{exercise}
-
-\begin{exercise}{Sentiment lexicon extension}
-  How could you analyse sentiment with lexicon extension using distributed thesauruses.
-
-  \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item **Lexicon-based approaches:** 
+        \begin{itemize}
+            \item Use pre-built dictionaries of words with sentiment labels (e.g., positive, negative).
+            \item Advantages:
+                \begin{itemize}
+                    \item Easy to implement and interpret.
+                    \item No need for labeled training data.
+                \end{itemize}
+            \item Disadvantages:
+                \begin{itemize}
+                    \item Limited to the coverage of the lexicon.
+                    \item Poor performance on context-dependent sentiment (e.g., sarcasm).
+                \end{itemize}
+        \end{itemize}
+        \item **Supervised learning approaches:**
+        \begin{itemize}
+            \item Train machine learning models on labeled data to classify sentiment.
+            \item Advantages:
+                \begin{itemize}
+                    \item Can learn from context and handle subtle sentiment expressions.
+                    \item Scalable to large datasets.
+                \end{itemize}
+            \item Disadvantages:
+                \begin{itemize}
+                    \item Requires annotated data, which can be costly to obtain.
+                    \item More computationally intensive than lexicon-based approaches.
+                \end{itemize}
+        \end{itemize}
+    \end{itemize}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item **Context-dependence:** Lexicons fail to account for how sentiment changes with context.
+        \item **Negation handling:** Sentences like "not bad" may be incorrectly classified without understanding negation.
+        \item **Ambiguity:** Words can have different sentiments in different contexts (e.g., "hot" in weather vs. arguments).
+        \item **Domain-specific issues:** Lexicons often do not cover domain-specific terms or slang.
+    \end{itemize}
   \end{solution}
@@
 \begin{exercise}{Naive Bayes and SVM}
-  Which of the following 4 are correct
+  Which of the following 4 are correct:
   \begin{enumerate}
-    \item Features for Naive Bayes classifiers do not need to be uncorrelated
-    \item Naive Bayes is slow to train
-    \item SVM itself is a multi-label classifier
-    \item SVM handles sparse data well
+    \item Features for Naive Bayes classifiers do not need to be uncorrelated.
+    \item Naive Bayes is slow to train.
+    \item SVM itself is a multi-label classifier.
+    \item SVM handles sparse data well.
   \end{enumerate}
@@
   \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item **Correct:** 
+        \begin{itemize}
+            \item Features for Naive Bayes classifiers do not need to be uncorrelated.
+            \item SVM handles sparse data well.
+        \end{itemize}
+        \item **Incorrect:**
+        \begin{itemize}
+            \item Naive Bayes is actually fast to train.
+            \item SVM is not inherently multi-label but can be adapted for multi-label classification.
+        \end{itemize}
+    \end{itemize}
   \end{solution}
@@
 \begin{exercise}{Language detection}
-  Explain approaches for language detection
+  Explain approaches for language detection.
 
   \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item **N-gram analysis:** Compare the frequency of character n-grams against language-specific profiles.
+        \item **Machine learning models:** Train classifiers (e.g., SVM, neural networks) using labeled text data for various languages.
+        \item **Lexicon-based methods:** Use dictionaries to match input text to language-specific words.
+        \item **Hybrid approaches:** Combine n-gram, lexicon, and machine learning techniques for better accuracy.
+    \end{itemize}
   \end{solution}
@@
 \begin{exercise}{Crowdsourced annotation}
-  Explain the crowdsourced annotation workflow
+  Explain the crowdsourced annotation workflow.
 
   \begin{solution}
-    % TODO
+    Steps in a crowdsourced annotation workflow:
+    \begin{enumerate}
+        \item **Task design:** Define the annotation task with clear instructions and examples.
+        \item **Platform setup:** Deploy the task on crowdsourcing platforms like Amazon Mechanical Turk.
+        \item **Quality control:** Use redundancy by assigning multiple annotators to the same task and include test questions.
+        \item **Data aggregation:** Combine annotations using majority voting or weighted scoring methods.
+        \item **Evaluation:** Analyze the consistency and reliability of annotations.
+    \end{enumerate}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    Ambiguities in tokenization include:
+    \begin{itemize}
+        \item **Hyphenation:** Is "state-of-the-art" one token or three?
+        \item **Punctuation:** Should periods in abbreviations (e.g., "U.S.") be separated?
+        \item **Contractions:** Should "don't" be split into "do" and "n't"?
+        \item **Language differences:** Tokenization rules vary by language (e.g., Chinese characters vs. English words).
+    \end{itemize}
   \end{solution}
@@
 \begin{exercise}{QA vs IR}
-  Differences between QA and IR
+  What are the differences between QA and IR?
 
   \begin{solution}
-    % TODO
+    Differences between QA and IR:
+    \begin{itemize}
+        \item **Goal:**
+        \begin{itemize}
+            \item QA provides direct answers to user queries.
+            \item IR retrieves documents that may contain relevant information.
+        \end{itemize}
+        \item **Processing:**
+        \begin{itemize}
+            \item QA requires natural language understanding and reasoning.
+            \item IR focuses on keyword matching and ranking algorithms.
+        \end{itemize}
+        \item **Applications:**
+        \begin{itemize}
+            \item QA is used in chatbots and virtual assistants.
+            \item IR powers search engines like Google.
+        \end{itemize}
+    \end{itemize}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    Challenges in cross-language QA include:
+    \begin{itemize}
+        \item **Translation errors:** Inaccurate translations can distort queries or answers.
+        \item **Lack of parallel resources:** Few high-quality bilingual corpora exist for certain language pairs.
+        \item **Cultural differences:** Questions or answers may need cultural adaptation.
+        \item **Mismatched structures:** Linguistic differences make aligning queries and answers difficult.
+    \end{itemize}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item **Stemming:** Reduces a word to its root by chopping off affixes (e.g., "running" → "run"). It is less precise and may produce non-words.
+        \item **Lemmatization:** Converts a word to its base form (lemma) based on its context (e.g., "better" → "good"). It is more accurate but computationally expensive.
+    \end{itemize}
   \end{solution}
@@
 \begin{exercise}{Web corpus}
-  Why are duplicates on the web and how to detect
+  Why are duplicates on the web, and how can they be detected?
 
   \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item **Reasons for duplicates:**
+        \begin{itemize}
+            \item Content mirroring across websites.
+            \item Republishing of syndicated articles.
+            \item Crawling the same site multiple times.
+        \end{itemize}
+        \item **Detection methods:**
+        \begin{itemize}
+            \item Hashing (e.g., MD5, SHA) to compare file contents.
+            \item Cosine similarity or Jaccard index for text comparisons.
+            \item Comparing HTML structure or metadata.
+        \end{itemize}
+    \end{itemize}
   \end{solution}
@@
 \begin{exercise}{Web corpus}
-  What are the 4 challenges when using the web as a corpus
+  What are the 4 challenges when using the web as a corpus?
 
   \begin{solution}
-    % TODO
+    Challenges include:
+    \begin{itemize}
+        \item **Noisy data:** Web content often includes irrelevant or ungrammatical text.
+        \item **Duplication:** Large-scale crawls often capture duplicate data.
+        \item **Bias:** Web data may reflect societal or geographic biases.
+        \item **Dynamic content:** Websites frequently change, making reproducibility difficult.
+    \end{itemize}
   \end{solution}
@@
 \begin{exercise}{TREC}
-  What are the TREC question types
+  What are the TREC question types?
 
   \begin{solution}
-    % TODO
+    TREC (Text REtrieval Conference) defines these question types:
+    \begin{itemize}
+        \item **Factoid questions:** Require specific factual answers (e.g., "Who is the president of the U.S.?").
+        \item **List questions:** Require a list of items (e.g., "List countries in the EU").
+        \item **Definition questions:** Seek explanations or definitions (e.g., "What is AI?").
+    \end{itemize}
   \end{solution}
@@
 \begin{exercise}{QALD}
-  What is QALD and what are the challenges
+  What is QALD, and what are the challenges?
 
   \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item **Definition:** QALD (Question Answering over Linked Data) focuses on answering questions using linked data (e.g., RDF datasets).
+        \item **Challenges:**
+        \begin{itemize}
+            \item Mapping natural language queries to structured data.
+            \item Handling incomplete or ambiguous data in linked datasets.
+            \item Scaling to large and heterogeneous data sources.
+        \end{itemize}
+    \end{itemize}
   \end{solution}
@@
 
-\sheet{2021}{1st exam}
+\setcounter{section}{2020}
+\sheet{1st exam}
 
@@
   \begin{solution}
-    % TODO
+    Comprehensiveness refers to the extent to which a system, dataset, or model covers all relevant aspects of a task or domain. In NLP:
+    \begin{itemize}
+        \item **Datasets:** Comprehensive datasets should include diverse and representative samples of the target domain.
+        \item **Models:** Comprehensive models should handle all variations and exceptions within their scope.
+        \item **Applications:** Comprehensive applications should meet all user requirements for a given task.
+    \end{itemize}
   \end{solution}
@@
 
-  % TODO: add image
-
   \begin{solution}
-    % TODO
+    To complete the image, analyze the given context and fill in the white boxes based on:
+    \begin{itemize}
+        \item **Labels:** Assign appropriate titles or labels to the boxes (e.g., steps in a pipeline).
+        \item **Relationships:** Connect missing elements to ensure the flow aligns with the diagram's logic.
+        \item **Annotations:** Use any provided hints to accurately reconstruct the intended content.
+    \end{itemize}
+    The exact answer depends on the image provided.
   \end{solution}
@@
 \begin{exercise}{NLTK}
-  Explain two parameters in NLTK that start with a c.
+  Explain two parameters in NLTK that start with a "c."
 
   \begin{solution}
-    % TODO
+    Two common NLTK parameters starting with "c" are:
+    \begin{itemize}
+        \item **Chunking:** Groups words into meaningful units (e.g., noun phrases) based on part-of-speech tagging.
+        \item **Concordance:** Displays occurrences of a word within a context window, useful for lexical analysis.
+    \end{itemize}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    For multi-class classification, SVM uses:
+    \begin{itemize}
+        \item **One-vs-One (OvO):** Trains a classifier for every pair of classes. Final classification is determined by majority voting.
+        \item **One-vs-All (OvA):** Trains one classifier per class against all other classes. The highest-confidence prediction is chosen.
+        \item **Extensions:** Kernel tricks and soft margins can be applied for better generalization.
+    \end{itemize}
   \end{solution}
@@
 \begin{exercise}{Feature differences}
-  Explain the differences in features between classical ML and deep network based ML.
+  Explain the differences in features between classical ML and deep network-based ML.
 
   \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item **Classical ML:**
+        \begin{itemize}
+            \item Relies on manually engineered features (e.g., TF-IDF, n-grams).
+            \item Requires domain expertise for feature extraction.
+        \end{itemize}
+        \item **Deep learning:**
+        \begin{itemize}
+            \item Automatically learns hierarchical feature representations.
+            \item Captures complex patterns (e.g., semantics, dependencies).
+            \item Requires large datasets for training.
+        \end{itemize}
+    \end{itemize}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    Cohen's Kappa is calculated as:
+    \[
+    \kappa = \frac{P_o - P_e}{1 - P_e}
+    \]
+    Where:
+    \begin{itemize}
+        \item \(P_o\): Observed agreement (sum of diagonal elements divided by the total number of items).
+        \item \(P_e\): Expected agreement (sum of product of marginal probabilities for each category).
+    \end{itemize}
+    Substitute the values from the given table into the formula to compute \(\kappa\).
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    Given true positives (\(TP\)), false positives (\(FP\)), and false negatives (\(FN\)):
+    \begin{itemize}
+        \item **Precision:** \( \frac{TP}{TP + FP} \)
+        \item **Recall:** \( \frac{TP}{TP + FN} \)
+        \item **F1-Score:** \( 2 \times \frac{\text{Precision} \times \text{Recall}}{\text{Precision} + \text{Recall}} \)
+    \end{itemize}
+    Plug in the values provided to compute these metrics.
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    Classification is used in IR for:
+    \begin{itemize}
+        \item **Query classification:** Identify user intent (e.g., navigational vs. informational).
+        \item **Document classification:** Categorize documents into predefined topics or genres.
+        \item **Relevance classification:** Predict whether a document is relevant to a query.
+    \end{itemize}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item **For:**
+        \begin{itemize}
+            \item Expands dataset availability to resource-scarce languages.
+            \item Enables cross-lingual NLP research.
+            \item Saves effort in recreating annotations.
+        \end{itemize}
+        \item **Against:**
+        \begin{itemize}
+            \item Potential loss of annotation quality in translation.
+            \item Context or cultural relevance may not transfer.
+            \item Requires manual review to ensure correctness.
+        \end{itemize}
+    \end{itemize}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item **For:**
+        \begin{itemize}
+            \item Large-scale, diverse data sources.
+            \item Includes up-to-date content.
+            \item Covers multiple languages and domains.
+        \end{itemize}
+        \item **Against:**
+        \begin{itemize}
+            \item Noisy and unstructured data.
+            \item Contains duplicates and redundant information.
+            \item May include biased or low-quality content.
+        \end{itemize}
+    \end{itemize}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    Example:
+    \begin{itemize}
+        \item **Preprocessing:** Tokenization, normalization, and cleaning.
+        \item **Feature extraction:** TF-IDF, embeddings.
+        \item **Modeling:** Training, evaluation, and deployment.
+    \end{itemize}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item **Mean Reciprocal Rank (MRR):**
+        \[
+        MRR = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{\text{rank}_i}
+        \]
+        Where \(\text{rank}_i\) is the rank of the first correct answer for query \(i\).
+        \item **Accuracy:** \(\frac{\text{Number of correct answers}}{\text{Total number of queries}}\)
+    \end{itemize}
+    Calculate based on the provided data.
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    Language detection involves identifying the language of a given text. Common approaches:
+    \begin{itemize}
+        \item **N-gram analysis:** Compare character n-grams to language profiles.
+        \item **Lexicon-based methods:** Match text to language-specific dictionaries.
+        \item **Machine learning:** Train classifiers (e.g., SVM, neural networks) on labeled text.
+        \item **Hybrid models:** Combine multiple techniques for higher accuracy.
+    \end{itemize}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    Plagiarism detection methods:
+    \begin{itemize}
+        \item **String matching:** Use algorithms to find exact matches between documents.
+        \item **Semantic similarity:** Apply embeddings to detect paraphrased or rephrased content.
+        \item **Citation analysis:** Check for missing or incorrect citations.
+        \item **Fingerprinting:** Generate hashes of text fragments and compare across sources.
+    \end{itemize}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item **Pros:**
+        \begin{itemize}
+            \item Fast and scalable data annotation.
+            \item Access to diverse annotators.
+            \item Cost-effective compared to hiring experts.
+        \end{itemize}
+        \item **Cons:**
+        \begin{itemize}
+            \item Quality control is challenging.
+            \item May introduce noise or bias.
+            \item Difficult to ensure annotator expertise.
+        \end{itemize}
+    \end{itemize}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    Recall can be improved by:
+    \begin{itemize}
+        \item **Expanding queries:** Use synonyms, related terms, or stemming to match more results.
+        \item **Reducing filters:** Remove restrictive criteria (e.g., date, location).
+        \item **Boosting document recall:** Index more documents or reduce relevance thresholds.
+    \end{itemize}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    For non-words, spell checkers:
+    \begin{itemize}
+        \item **Dictionary lookup:** Identify words not in the dictionary.
+        \item **Edit distance:** Suggest corrections based on similarity (e.g., Levenshtein distance).
+        \item **Contextual models:** Use surrounding words to predict the intended word.
+    \end{itemize}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    The IR cycle includes:
+    \begin{itemize}
+        \item **Indexing:** Process and store data for efficient retrieval.
+        \item **Querying:** Users submit queries to search the index.
+        \item **Retrieval:** Return ranked results based on relevance scores.
+        \item **Evaluation:** Assess performance using metrics like precision and recall.
+    \end{itemize}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    The Vector Space Model represents documents and queries as vectors:
+    \begin{itemize}
+        \item **Terms:** Each dimension corresponds to a term in the vocabulary.
+        \item **Weights:** TF-IDF scores are used as weights for each term.
+        \item **Similarity:** Compute cosine similarity to rank documents by relevance to the query.
+    \end{itemize}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item **Factual sentiment analysis:** Focuses on extracting objective sentiment (e.g., product ratings).
+        \item **Opinionated analysis:** Identifies subjective opinions (e.g., customer reviews).
+        \item **Key difference:** Factual analysis is more structured and easier to evaluate, whereas opinionated analysis is subjective and context-dependent.
+    \end{itemize}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    Sentiment analysis applications:
+    \begin{itemize}
+        \item **Customer feedback:** Analyze reviews for product improvements.
+        \item **Social media monitoring:** Detect trends and public sentiment.
+        \item **Political analysis:** Gauge public opinion on policies or events.
+        \item **Market research:** Understand consumer preferences.
+    \end{itemize}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    The Teacher/Student approach combines:
+    \begin{itemize}
+        \item **Teacher signals:** Provide detailed instructions and explanations.
+        \item **Student feedback:** Allows iterative learning and personalization.
+        \item **Better than audio:** Interaction and adaptivity enhance understanding and retention compared to static audio content.
+    \end{itemize}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item **Representation:** Encodes data (e.g., text or images) into feature vectors.
+        \item **Fusion:** Combines multiple data modalities (e.g., text and images) to improve performance.
+        \item **Differences:** Representation focuses on individual modalities, whereas fusion integrates multiple modalities into a unified representation.
+    \end{itemize}
   \end{solution}

### 2024-12-05 BAI/paper/presentation.tex
Added structure for BAI paper and presentation.
Added current state.

--- 2024-10-18 BAI/templates/beamer_slides/WTM_latex.tex
@@
-\documentclass[14pt]{beamer}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Meta informations:
-\newcommand{\trauthor}{Horst Hansen}
-\newcommand{\trtype}{} %{Proseminar} %{Seminar} %{Workshop}
-\newcommand{\trcourse}{Knowledge Processing with Neural Networks}
-\newcommand{\trtitle}{Neural Networks for Artificial Agents}
-\newcommand{\trmatrikelnummer}{6543210}
-\newcommand{\tremail}{hansen@informatik.uni-hamburg.de}
+\documentclass[10pt]{beamer}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Metainformation
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\newcommand{\trauthor}{Leopold Lemmermann}
+\newcommand{\trtype}{Seminar}
+\newcommand{\trcourse}{Bio-inspired Artificial Intelligence}
+\newcommand{\trtitle}{Image Captioning}
+\newcommand{\trmatrikelnummer}{7724741}
+\newcommand{\tremail}{leopold.lemmermann@studium.uni-hamburg.de}
 \newcommand{\trinstitute}{Dept. Informatik -- Knowledge Technology, WTM}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Languages:
-
-% Falls die Ausarbeitung in Deutsch erfolgt:
-% \usepackage[german]{babel}
-% \usepackage[T1]{fontenc}
-% \usepackage[latin1]{inputenc}
-% \usepackage[latin9]{inputenc}	 				
-% \selectlanguage{german}
-
-% If the thesis is written in English:
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Languages
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \usepackage[english]{babel}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Bind packages:
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Bind packages
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \usepackage{beamerthemesplit}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Configurationen:
-%\hypersetup{pdfpagemode=FullScreen}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Configuration
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\hypersetup{pdfpagemode=FullScreen}
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Costom Definitions:
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Custom definitions
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \setbeamertemplate{title page}
@@
 	  \begin{beamercolorbox}[ht=8ex,center]{}
-		  \includegraphics[width=0.20\paperwidth]{wtmIcon.pdf}
+		  \includegraphics[width=0.20\paperwidth]{res/wtmIcon.pdf}
 	  \end{beamercolorbox}%
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Additional 'theorem' and 'definition' blocks:
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Additional 'theorem' and 'definition' blocks
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \newtheorem{axiom}{Axiom}[section]
-%\newtheorem{axiom}{Fakt}[section]			% Wenn in Deutsch geschrieben wird.
 %Usage:%\begin{axiom}[optional description]%Main part%\end{fakt}
@@
 \theoremstyle{remark}
-%\newtheorem{remark}[section]{Bemerkung} % Wenn in Deutsch geschrieben wird.
 \newtheorem{remark}[section]{Remark}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Provides TODOs within the margin:
-\newcommand{\TODO}[1]{\marginpar{\emph{\small{{\bf TODO: } #1}}}}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 % Abbreviations and mathematical symbols
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \newcommand{\modd}{\text{ mod }}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 % Display of TOCs:
-\AtBeginSection[]
-{
-	\setcounter{tocdepth}{2}  
-	\frame
-	{
-	  \frametitle{Outline}
-		\tableofcontents[currentsection]
-	}
-}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% \AtBeginSection[]
+% {
+% 	\setcounter{tocdepth}{2}
+% 	\frame
+% 	{
+% 	  \frametitle{Outline}
+% 		\tableofcontents[currentsection]
+% 	}
+% }
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Document:
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Document
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{document}
@@
 
-\begin{frame}[plain] % plain => kein Rahmen
+\begin{frame}[plain]
   \titlepage
 \end{frame}
-%\setcounter{framenumber}{0}
 
-\frame{
-	\frametitle{Outline}
-	\tableofcontents
-}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Content
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-%%%%%%%%%%%%%%
-% Your Content
+\begin{frame}
+	\begin{itemize}
+			\item \textbf{Image Captioning}: Generating textual descriptions for images \cite{vinyals:2015}.
+			\item \textbf{Challenges}: Complex problem combining computer vision (CV) and natural language processing (NLP) \cite{xu}.
+			\item \textbf{Advancements}: Deep learning models using CNNs for feature extraction and RNNs for language generation \cite{he, simonyan}.
+	\end{itemize}
+\end{frame}
 
-\section{Motivation and Question}
-
-\frame[t]{
-  \frametitle{Motivation}
+\section{Methods}
+\begin{frame}{Methods}
 	\begin{itemize}
-  	\item Add your motivation here
+			\item \textbf{Dataset:} Flickr8k, with 8,000 images and five captions per image \cite{vinyals:2015}.
+			\item \textbf{Architecture:} Encoder-Decoder architecture:
 					\begin{itemize}
-			\item Maybe  with some details
-			\begin{itemize}
-				\item but not too much
+							\item Encoder: CNN-based (ResNet-50, VGG-16) \cite{he, simonyan}.
+							\item Decoder: RNN-based (GRU, LSTM) \cite{hochreiter}.
 					\end{itemize}
+			\item \textbf{Evaluation Metrics:} BLEU score for performance \cite{papineni}, with CIDEr as an alternative \cite{vedantam}.
 	\end{itemize}
-		\item Use references \textsuperscript{[Author, 2010]}
-	\end{itemize}
-}
-
-\section{Basics and Definition}
+\end{frame}
 
 \section{Approach}
+\begin{frame}{Approach}
+	\begin{itemize}
+			\item \textbf{Data Preprocessing:} Resize images, normalize, and tokenize captions \cite{vinyals:2015}.
+			\item \textbf{Model Training:} Train decoder using Adam optimizer and monitor BLEU score for validation \cite{papineni}.
+			\item \textbf{Model Evaluation:} Evaluate BLEU score after each epoch for overfitting assessment \cite{rennie}.
+	\end{itemize}
+\end{frame}
 
 \section{Results}
+\begin{frame}{Results}
+	\begin{itemize}
+			\item BLEU scores for encoder-decoder combinations (ResNet-GRU, VGG-GRU, ResNet-LSTM, VGG-LSTM) \cite{he, simonyan}.
+				\item \textbf{Best Performance:} VGG encoder with LSTM decoder, highest BLEU score in the first epoch \cite{vinyals:2015}.
+				\item Observation of \textbf{Overfitting}: Decrease in BLEU scores over epochs \cite{rennie}.
+	\end{itemize}
+\end{frame}
 
 \section{Conclusion}
+\begin{frame}{Conclusion}
+	\begin{itemize}
+			\item \textbf{Summary:} VGG-LSTM model achieved the best BLEU score and stability across epochs \cite{vinyals:2016}.
+			\item \textbf{Future Work:} Explore other architectures (e.g., transformers), datasets (e.g., Flickr30k, MS COCO), and metrics (e.g., CIDEr) \cite{vedantam}.
+	\end{itemize}
+\end{frame}
 
-\frame[t]
-{
-  \frametitle{Conclusion}
-
-  Novelty and contribution of this work:
-  \begin{itemize}
-	  \item Sum up the approach
-		\item Sum op the results
-		\item ...
-		\item Show that it solves the question
-  \end{itemize}
-
-\mbox{ }
-
-	Open Questions:
-  \begin{itemize}
-  	\item Something
-	  \item ... is always missing
-  \end{itemize}
-}
-
-%%%%%%%%%%%%%%
-
-\frame[c]{
-  \frametitle{The End}
-\begin{center}
-  Thank you for your attention.\\[1ex]
-  Any question?\\[5ex]
-\end{center}
-  \footnotesize
-  Literature:
-	\begin{itemize}
-		\item Author , Author , Author, and Author. Name of the conference paper. \emph{In: Proceedings of the Conference Name}, 2008
-		\item Author, Author, and Author. Name of the Article. \emph{Name of the Journal}, 42:111-133, 2010
-		\item Author, and Author. \emph{Name of the Book}. Publisher, 2009
-	\end{itemize}
-}
+\begin{frame}[allowframebreaks]
+	\bibliographystyle{apalike}
+	\bibliography{bib}
+\end{frame}
 

### 2024-12-05 SbD/exercises.tex
Added current state.

@@
 
-\usepackage[]{../.template/xrcise}
+\usepackage[solutions]{../.template/xrcise}
 
@@
   \begin{enumerate}
-    \item Nennen und erläutern Sie eine Angriffsart, mit der Eve den gesamten Netzverkehr von Bob abhören und manipulieren kann.
-    \item Wie kann Bob sich vor dem in Teilaufgabe a) beschriebenen Angriff schützen?
+    \item\label{itm:manipulation} Nennen und erläutern Sie eine Angriffsart, mit der Eve den gesamten Netzverkehr von Bob abhören und manipulieren kann.
+    \item Wie kann Bob sich vor dem in Teilaufgabe \ref{itm:manipulation} beschriebenen Angriff schützen?
   \end{enumerate}
@@
     \item Was bedeutet indeterministische Verschlüsselung?
-    \item Wie realisieren Sie die indeterministische Verschlüsselung bei einem asymmetrischen Konzelations- system?
-    \item Was ändert sich gegenüber b) bei einem symmetrischen Konzelationssystem?
+    \item\label{itm:asymmetric} Wie realisieren Sie die indeterministische Verschlüsselung bei einem asymmetrischen Konzelationssystem?
+    \item Was ändert sich gegenüber \ref{itm:asymmetric} bei einem symmetrischen Konzelationssystem?
     \item Sie sollen eine indeterministische Verschlüsselung realisieren. Zur Auswahl stehen die Algorithmen RSA und ElGamal. Wie würden Sie bei beiden Algorithmen vorgehen?
@@
         \item Indeterministische Verschlüsselung bedeutet, dass derselbe Klartext bei wiederholter Verschlüsselung mit demselben Schlüssel unterschiedliche Schlüsseltexte erzeugt.
-        \item Bei einem asymmetrischen Verschlüsselungssystem kann indeterministische Verschlüsselung durch die Einbeziehung eines Zufallswerts (z.
-B.
+        \item Bei einem asymmetrischen Verschlüsselungssystem kann indeterministische Verschlüsselung durch die Einbeziehung eines Zufallswerts (zB.
 Padding mit Zufallszahlen) in die Nachricht erreicht werden.
@@
 \begin{exercise}{Informationstheoretische Sicherheit}\label{ex:otp}
-  Das One-Time-Pad (Vernam-Chiffre) ist ein sehr einfaches, aber informationstheoretisch sicheres symme- trisches Verschlüsselungsverfahren: Die Klartextbits $x_i$ werden einzeln XOR-verknüpft mit einer zufälligen Schlüsselfolge $k_i$ gleicher Länge, die nur ein einziges Mal verwendet werden darf (Schlüsseltext $s_i = x_i \oplus k_i$ mit $i= 1,2,...$).
+  Das One-Time-Pad (Vernam-Chiffre) ist ein sehr einfaches, aber informationstheoretisch sicheres symmetrisches Verschlüsselungsverfahren: Die Klartextbits $x_i$ werden einzeln XOR-verknüpft mit einer zufälligen Schlüsselfolge $k_i$ gleicher Länge, die nur ein einziges Mal verwendet werden darf (Schlüsseltext $s_i = x_i \oplus k_i$ mit $i= 1,2,...$).
   \begin{enumerate}
@@
 Legen Sie eine bitweise Authentisierung (ein MAC-Bit pro Klartextbit) zugrunde.
-    \item Kann es informationstheoretisch sichere asymmetrische Verfahren zur Verschlüsselung bzw. Authenti- fikation geben? Begründen Sie Ihre Antwort.
-  \end{enumerate}
-
-  \begin{solution}
-    \begin{enumerate}
-        \item Der Angreifer erfährt bei einem One-Time-Pad nichts über den Klartext, da der Schlüsseltext rein zufällig ist und keine Informationen über den Klartext enthält.
-        \item Sicherheit des Verfahrens:
-        \[
-        H(M|C) = H(M) \quad \text{(Entropie bleibt erhalten).}
-        \]
-        Der Schlüsseltext enthält keine Informationen über den Klartext ohne den Schlüssel.
-        \item Wiederverwendung des Schlüssels:
-        \[
-        C_1 \oplus C_2 = M_1 \oplus M_2
-        \]
-        Dies ermöglicht die Rekonstruktion der Klartexte \( M_1, M_2 \).
+    \item Kann es informationstheoretisch sichere asymmetrische Verfahren zur Verschlüsselung bzw. Authentifikation geben? Begründen Sie Ihre Antwort.
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+        \item Lediglich die Länge des Klartexts. Der Klartext bleibt geheim.
+        \item Sicherheit des Verfahrens: $H(M|C) = H(M)$ (Entropie bleibt erhalten). Der Schlüsseltext enthält keine Informationen über den Klartext ohne den Schlüssel.
+        \item Wiederverwendung des Schlüssels: $C_1 \oplus C_2 = M_1 \oplus M_2$. Dies ermöglicht die Rekonstruktion der Klartexte $M_1, M_2$.
         \item Ein informationstheoretisch sicheres Authentifikationssystem muss auf zufälligen, einmaligen Schlüsseln basieren (z. B. MACs mit OTPs).
-        \item Ein System: Der Sender erzeugt einen zufälligen Schlüssel und XORt diesen mit der Nachricht, um den MAC zu erstellen. Der Empfänger überprüft, ob der MAC gültig ist.
-        \item Es gibt keine informationstheoretisch sicheren asymmetrischen Verfahren, da der öffentliche Schlüssel immer Angriffsvektoren bietet. Die Sicherheit ist hier nur computational.
+        \item Schlüssel und Hashfunktion austauschen. Sender: MAC erstellen $t_i=h(k_i, m_i)$. Empfänger: MAC überprüfen $t_i=h(k_i, m'_i)$.
+        \item Es gibt keine informationstheoretisch sicheren asymmetrischen Verfahren, da der öffentliche Schlüssel immer Angriffsvektoren bietet.
     \end{enumerate}
@@
   Ein klassisches Verschlüsselungsverfahren ist die Verschiebechiffre (Vigenère-Chiffre).
-Sie arbeitet typischer- weise auf einem Alphabet aus den Großbuchstaben A-Z für den Klartext, den Schlüssel und den Schlüssel- text.
+Sie arbeitet typischerweise auf einem Alphabet aus den Großbuchstaben A-Z für den Klartext, den Schlüssel und den Schlüsseltext.
 Zudem kann ein Zeichen zur Worttrennung hinzugenommen werden.
@@
     \begin{enumerate}
-        \item Verschlüsselung mit PRUEFUNG als Schlüssel:
-        \[
-        C = \texttt{ILNUHAZKA\_VLA\_ANL\_KLU\_LTS\_GMMI\_VQK\_TVT\_OZUB\_DRNT\_OZXEU}
-        \]
+        \item Verschlüsselung mit PRUEFUNG als Schlüssel: $C = \texttt{ILNUHAZKA\_VLA\_ANL\_KLU\_LTS\_GMMI\_VQK\_TVT\_OZUB\_DRNT\_OZXEU}$
         \item Kryptanalyse:
@@
           \end{itemize}
-        \item Verlängerung des Schlüsselworts:
-        \[
-        \text{Faktor der Vergrößerung: } 27^s
-        \]
+        \item Verlängerung des Schlüsselworts: $\text{Faktor der Vergrößerung: } 27^s$
         \item Um informationstheoretische Sicherheit zu erreichen, müsste das Verfahren in ein One-Time-Pad umgewandelt werden.
@@
         \begin{itemize}
-            \item Mit der Skytala (\(k = 4\), \(4 \times 8\) Matrix):
-            \[
-            \texttt{PRUEFUNGSSPASS\_OHNE\_KOMPROMISSE} \rightarrow \texttt{PHSOROPRUNFEINSUPG\_GHNOM\_}
-            \]
-            \item Mit der Caesar-Chiffre (\(k = 3\)):
-            \[
-            \texttt{STUDIUMSFREUDE} \rightarrow \texttt{VWXGLXPVIKUHGH}
-            \]
+            \item Mit der Skytala ($k = 4$): $\texttt{PRUEFUNGSSPASS\_OHNE\_KOMPROMISSE} \rightarrow \texttt{PSHRRSNOUPEMEA\_IFSKSUSOSN\_MEGOP}$
+            \item Mit der Caesar-Chiffre ($k = 3$): $\texttt{STUDIUMSFREUDE} \rightarrow \texttt{VWXGLXPVIKUHGH}$
         \end{itemize}
@@
           \end{itemize}
-        \item Kombination der beiden Verfahren:
-        \begin{itemize}
-            \item Kombination 1 (Caesar vor Skytala): Sicherheit erhöht sich, da Buchstabenhäufigkeiten durch die Caesar-Chiffre maskiert werden.
-            \item Kombination 2 (Skytala vor Caesar): Sicherheit bleibt gleich wie bei der Skytala alleine, da Caesar die Buchstabenhäufigkeiten nicht beeinflusst.
-        \end{itemize}
+        \item Da beide Chiffren unabhängig voneinander angewendet werden und individuell schwach sind, ist die Sicherheit nicht stark erhöht.
+Reihenfolge der Anwendung ist irrelevant, da beide Operationen kommutativ sind.
         \item Moderne Chiffren müssen folgende Gütekriterien erfüllen:
@@
     \item In der Praxis wird die digitale Signatur einer Nachricht $m$ meist durch Signieren des (kryptographischen) Hashwertes der Nachricht $H(m)$ und nicht der Nachricht selbst gebildet: $\text{Sig}_k(H(m))$.
-Warum kann ein Empfänger sich in diesem Fall auf die Authentizität der Nachricht verlassen?
-Welche Vorteile bringt das Signieren des Hashwertes einer Nachricht mit sich?
+      \begin{itemize}
+        \item Warum kann ein Empfänger sich in diesem Fall auf die Authentizität der Nachricht verlassen?
+        \item Welche Vorteile bringt das Signieren des Hashwertes einer Nachricht mit sich?
 Betrachten Sie hierfür auch den Fall, dass klassische RSA-Signaturen ohne Padding zum direkten Signieren von kurzen Nachrichten eingesetzt würden.
+      \end{itemize}
     \item Eine Entwicklerin möchte eine von ihr entwickelte Software zur Verfügung stellen.
@@
     \begin{enumerate}
-        \item Die Identitätsfunktion \( H(x) = x \) ist keine kryptographische Hashfunktion, da sie nicht kollisionsresistent ist und keine Kompression durchführt.
+        \item Die Identitätsfunktion \( H(x) = x \) ist keine kryptographische Hashfunktion, da sie keine feste Ausgabelänge erzeugt, keine Einwegfunktion ist und keine Pseudozufälligkeit aufweist.
         \item Kollisionsfreiheit ist entscheidend, da der Angreifer sonst zwei Nachrichten mit identischem Hashwert erzeugen könnte, um Signaturen zu fälschen.
-        \item Vorteile des Signierens von Hashwerten:
-        \begin{itemize}
-            \item Reduziert den Rechenaufwand bei großen Nachrichten.
-            \item Stellt sicher, dass die Nachricht authentisch ist.
-            \item Padding (z. B. bei RSA) schützt vor Angriffen.
-        \end{itemize}
+        \item Kollisionsfreiheit des Hashwerts garantiert die Authentizität der Nachricht.
+Vorteile des Signierens des Hashwerts sind Effizienz (kleinere Datenmenge), Sicherheit (auch ohne padding keine Längeninformation) und Flexibilität (beliebiges Nachrichtenformat möglich).
         \item Verfahren für die Software-Integrität:
@@
 \begin{exercise}{Das Feistel-Prinzip}\label{ex:feistel}
-  \begin{enumerate}
-    \item Die nebenstehende Abbildung zeigt eine Runde der Feistel-Verschlüsselung. Berechnen Sie zwei Runden der Verschlüsselung und Entschlüsselung mit dem Feistel-Prinzip und den folgenden Werten:
+  Die nebenstehende Abbildung zeigt eine Runde der Feistel-Verschlüsselung. 
+  \input{res/feistel.fig.tex}
+  \begin{enumerate}
+    \item Berechnen Sie zwei Runden der Verschlüsselung und Entschlüsselung mit dem Feistel-Prinzip und den folgenden Werten:
       \begin{itemize}
@@
       \end{itemize}
-      \input{res/feistel.fig.tex}
     \item Welche Eigenschaften sollte die Einwegfunktion F aufweisen, um mit ihrer Hilfe eine sichere Feistel-Chiffre zu erzeugen?
@@
     \begin{enumerate}
-        \item Zwei Runden Feistel-Verschlüsselung:
-        \begin{itemize}
-            \item Klartext: \( 0011 \ 0100 \)
-            \item Runde 1:
-            \[
-            L_1 = R_0, \quad R_1 = L_0 \oplus F(R_0, K_1)
-            \]
-            \( F(R_0, K_1) = R_0 \land K_1 \)
-            \item Runde 2:
-            \[
-            L_2 = R_1, \quad R_2 = L_1 \oplus F(R_1, K_2)
-            \]
-        \end{itemize}
-        \item Eigenschaften der Funktion \( F \):
+        \item
+          \begin{itemize}
+            \item Klartext: $(L_0, R_0) = 0011 \ 0100$
+            \item Runde 1: $L_1 = R_0, \quad R_1 = L_0 \oplus R_0 \land K_1$
+            \item Runde 2: $L_2 = R_1, \quad R_2 = L_1 \oplus R_1 \land K_2$
+            \item Schlüsseltext: $(L_3, R_3) = (R_2, L_2) = 0101\ 0111$
+          \end{itemize}
+
+        \item Eigenschaften der Funktion $F$:
         \begin{itemize}
@@
         \end{itemize}
-        \item Eine konstante Funktion \( F \) funktioniert, da die Struktur der Feistel-Chiffre die Umkehrbarkeit sicherstellt.
-        \item Die Umkehrfunktion \( F^{-1} \) ist nicht notwendig, da die Entschlüsselung über die symmetrische Struktur erfolgt.
+        \item Eine konstante Funktion $F$ funktioniert, da die Struktur der Feistel-Chiffre die Umkehrbarkeit sicherstellt.
+        \item Die Umkehrfunktion $F^{-1}$ ist nicht notwendig, da die Entschlüsselung über die symmetrische Struktur erfolgt.
     \end{enumerate}
@@
     \begin{enumerate}
-        \item Der öffentliche Exponent \( c \) muss teilerfremd zu \( \phi(n) \) sein, damit ein inverser Schlüssel existiert.
-        \item Eine Nachricht \( m \) darf \( p \) oder \( q \) nicht enthalten, da dies den privaten Schlüssel direkt offenbart.
-        \item Berechnung des privaten Exponenten \( d \):
-        \[
-        c = 5, \quad n = 7 \cdot 17 = 119, \quad \phi(n) = 96
-        \]
-        \( d = c^{-1} \mod \phi(n) \implies d = 77 \)
+        \item Der öffentliche Exponent $c$ muss teilerfremd zu $\varphi(n)$ sein, damit ein inverser Schlüssel existiert.
+        \item 
+        \item Berechnung des privaten Exponenten $d$ (mit $c = 5, n = 7 \cdot 17 = 119, \varphi(n) = 96$): $c \cdot d \equiv 1 \pmod{\varphi(n)}$
+          $d = c^{-1} \mod \varphi(n) \implies d = 77$
         \item Multiplikative Homomorphie:

### 2024-12-05 SbD/res/des.tbl.tex
Added current state.

\begin{table}[H]
  \centering
  \caption{Ausgewählte DES-Funktionswerte}
  \begin{tabular}{c|c|c|c}
  \textbf{Schlüssel $k$} & \textbf{Eingabe $x/y$ von $E_k/D_k$} & \textbf{$E_k(x)$} & \textbf{$D_k(y)$} \\ \hline
  97a4 3741 8086 0b & 6b3b acd4 6570 6187 & fecd 14e2 8cb0 6a2b & 7285 3f6b e44c cb8e \\ \hline
  97a4 3741 8086 0b & 5de1 bd06 54fa 40ed & 42e0 4462 099e e8ca & 9384 bebf 5b18 793f \\ \hline
  3e45 c68f bb89 27 & 5de1 bd06 54fa 40ed & 4995 e68b b800 3a06 & f575 4696 effc 1bf1 \\ \hline
  e6f4 a450 4898 d7 & 764d e3ae 043b d6b7 & 5f31 4917 4ecc e6b4 & 30db 8c29 073a 182c \\ \hline
  f6bb 747d dc16 15 & ceae a65a ae77 1c72 & b1e0 9414 e068 3622 & 9acb 4aff e84a b1f7 \\ \hline
  3e45 c68f bb89 27 & 764d e3ae 043b d6b7 & fca6 8c2f b81e dc79 & 4bae f15e 7efb 5a8c \\ \hline
  e6f4 a450 4898 d7 & c183 50b5 52b8 f8b1 & 5de1 bd06 54fa 40ed & 6b3b acd4 6570 6187 \\ \hline
  3e45 c68f bb89 27 & 9400 39ca 9301 5507 & 199c 2565 ea8e 2b53 & f742 7dac e2ff 00d2 \\ \hline
  f6bb 747d dc16 15 & c620 9b86 c72a 6b3b & 4e6c efc7 1f5b 3756 & 42e0 4462 099e e8ca \\ \hline
  e6f4 a450 4898 d7 & 42e0 4462 099e e8ca & c3fc 30e9 10db 827e & d0eb 0c0e 1b43 12da \\ \hline
  f6bb 747d dc16 15 & a904 aa4e c934 ecb0 & 42e0 4462 099e e8ca & 5617 0d2a ccdb 9981 \\ \hline
  f6bb 747d dc16 15 & a3c5 5b7b c599 e5af & 4179 a965 d79e 1edd & e927 17a5 b2a9 b559 \\ \hline
  e6f4 a450 4898 d7 & 9400 39ca 9301 5507 & 0669 924e 1cec 70d0 & 7555 ce29 f46b 7e74 \\ \hline
  3e45 c68f bb89 27 & 6b3b acd4 6570 6187 & abe6 c175 4435 742a & 043c 252e bbe0 3d09 \\ \hline
  97a4 3741 8086 0b & c183 50b5 52b8 f8b1 & 9400 39ca 9301 5507 & 764d e3ae 043b d6b7 \\ \hline
  97a4 3741 8086 0b & 42e0 4462 099e e8ca & 8b99 c9de 0e47 22db & 5de1 bd06 54fa 40ed \\
  \end{tabular}
\end{table}

### 2024-12-05 SbD/summary.tex
Added current state.

@@
   \item \textbf{Faktorisierungsannahme}: Primzahlen $p$ und $q$ sind schwer aus $n=p \cdot q$ zu berechnen.
+  \item \textbf{(diskrete) Elliptische Kurve}: Eine elliptische Kurve über $\mathbb{Z}_p$ ist definiert durch
+    \[ y^2 \equiv x^3 + ax + b \pmod{p} \text{ und } -4a^3 - 27b^2 \not\equiv 0 \pmod{p} \]
 \end{itemize}
@@
 \subsection{Elliptic Curve Cryptography (ECC)}
-% TODO: Add ECC
+\textit{Basiert auf diskretem Logarithmus für elliptische Kurven (ECDLP) . Einsatz zB. ed25519, ECDH, ECDSA.}
+\begin{itemize}
+  \item \textbf{Parameter}
+    \[
+      \underbrace{P}_\text{Generatorpunkt}\quad
+      \underbrace{n}_\text{privater Schlüssel}\quad
+      \underbrace{Q=nP}_\text{öffentlicher Schlüssel}
+    \]
+  \item \textbf{Verschlüsselung \& Entschlüsselung}: je nach Verfahren (ElGamal, RSA, …)
+  \item \textbf{Vorteile}: kleinere Schlüssel, kürzere Signaturen, weniger Rechenaufwand
+\end{itemize}
 

### 2024-12-06 SbD/exercises.tex
Refined some exercises and summary.

@@
 Zeigen Sie das Vorgehen anhand der abgefangenen verschlüsselten Nachricht $s= 101$, der eigens von Eve gewählten Zufallszahl $r= 22$ sowie dem multiplikativen Inversen der Zufallszahl $r^{-1} = 92 \mod 119$.
-Verwenden Sie auf Empfängerseite (Alice) den gegebenen RSA-Modulus, den öffentlichen Verschlüsselungsexponenten $c= 11$ und den geheimen Entschlüsselungsexponenten $d= 35$, der nur Alice, aber nicht Eve bekannt ist.
+Verwenden Sie auf Empfängerseite (Alice) den gegebenen RSA-Modulus, den öffentlichen Verschlüsselungsexponenten $c=11$ und den geheimen Entschlüsselungsexponenten $d=35$, der nur Alice, aber nicht Eve bekannt ist.
     \item Nennen Sie ein Anwendungsbeispiel, bei dem der multiplikative Homomorphismus des RSA-Verfahrens "ausgenutzt" bzw. gewollt angewendet wird.
@@
         \item Der öffentliche Exponent $c$ muss teilerfremd zu $\varphi(n)$ sein, damit ein inverser Schlüssel existiert.
-        \item 
+Und zwischen 2 und $\varphi(n)$ liegen, um sicherzustellen, dass die Nachricht nicht trivial verschlüsselt wird.
+        \item Da die Nachricht dann trivial verschlüsselt wäre mit $c = 0$, was auf $m=n$ schließen lässt.
         \item Berechnung des privaten Exponenten $d$ (mit $c = 5, n = 7 \cdot 17 = 119, \varphi(n) = 96$): $c \cdot d \equiv 1 \pmod{\varphi(n)}$
-          $d = c^{-1} \mod \varphi(n) \implies d = 77$
-        \item Multiplikative Homomorphie:
-        \[
-        m_1 \cdot m_2 \mod n \implies c(m_1 \cdot m_2) = c(m_1) \cdot c(m_2)
-        \]
-        Angreifer können Produkte von Nachrichten entschlüsseln.
-        \item Eve nutzt:
-        \[
-        r \cdot s \cdot r^{-1} \mod n = s
-        \]
-        Der Angriff ist erfolgreich.
-        \item Multiplikative Homomorphie wird in homomorpher Verschlüsselung absichtlich genutzt (z. B. für Berechnungen auf verschlüsselten Daten).
+          \[ d = c^{-1} \mod \varphi(n) \implies d = 77 \]
+        \item Angreifer können Produkte von Nachrichten entschlüsseln.
+          \[ m_1 \cdot m_2 \mod n \implies c(m_1 \cdot m_2) = c(m_1) \cdot c(m_2) \]
+        \item Eve berechnet $s'=s \cdot r^c \mod{n}$ und sendet $s'$ an Alice.
+Alice entschlüsselt $s'$ mit $s''=s'^d \mod{n}$ und erhält $s''=s \cdot r \mod{n}$.
+Durch Multiplikation mit $r^{-1}$ erhält Eve $s$.
+        \item Ein Anwendungsbeispiel sind Berechnungen auf verschlüsselten Daten.
     \end{enumerate}
@@
     \begin{enumerate}
-        \item Eve führt einen Chosen-Plaintext-Angriff durch:
-        \[
-        c_B = m^e \mod n
-        \]
-        Für alle möglichen \( m \), vergleicht sie den berechneten \( c_B \) mit dem übertragenen Wert, bis \( m = 3 \) gefunden wird.
+        \item Da $m\in{1...6}$, probiert Eve alle möglichen Werte aus und verschlüsselt sie mit Bobs öffentlichem Schlüssel. Sie erhält $c_B = 9$ für $m_B = 4$. Bob hat also eine 4 gewürfelt.
         \item Bob kann sich schützen, indem er Zufallswerte oder Padding einfügt, um die deterministische Eigenschaft des RSA-Verfahrens zu verhindern.
@@
     \begin{enumerate}
-        \item Forward Secrecy mit RSA wird erreicht, indem für jede neue Sitzung ein neues, kurzlebiges (ephemeres) RSA-Schlüsselpaar generiert wird.
-Der öffentliche Schlüssel wird dem Kommunikationspartner zur Verfügung gestellt, und der private Schlüssel wird nach Abschluss der Sitzung sicher gelöscht.
-Auf diese Weise kann ein kompromittierter langfristiger Schlüssel vergangene Sitzungen nicht gefährden.
-        \item Wenn ein Angreifer den privaten Entschlüsselungsexponenten eines RSA-Schlüssels erlangt, kann er alle Nachrichten entschlüsseln, die mit dem zugehörigen öffentlichen Schlüssel verschlüsselt wurden.
-Dadurch wird die Vertraulichkeit der Kommunikation vollständig aufgehoben.
-        \item Der Angreifer kann die verschlüsselten Nachrichten abfangen und mit dem privaten Schlüssel entschlüsseln.
-Alle vergangenen und zukünftigen Nachrichten, die denselben Schlüssel verwenden, können entschlüsselt werden, sofern sie aufgezeichnet wurden.
-        \item Durch regelmäßiges Wechseln des privaten Schlüssels und Löschen des alten privaten Schlüssels nach Ablauf einer Sitzung wird Forward Secrecy gewährleistet.
-Zusätzlich kann der Diffie-Hellman-Schlüsselaustausch verwendet werden, um symmetrische Sitzungsschlüssel auszuhandeln, die nicht mit den RSA-Schlüsseln gespeichert werden.
+        \item Garnicht, RSA bietet keine Forward Secrecy.
+        \item Da alle Nachrichten dann vom Angreifer entschlüsselt werden können.
+        \item Angreifer entschlüsselt mittels Exponent
+          \[ m=c^d \mod{n} \]
+        \item Wenn die Schlüssel regelmäßig gewechselt werden, können bei Kompromittierung nur die Nachrichten entschlüsselt werden, die mit dem aktuellen Schlüssel verschlüsselt wurden.
     \end{enumerate}
@@
     \begin{enumerate}
-        \item Anpassung des ElGamal-Kryptosystems für elliptische Kurven:
-        \begin{itemize}
-            \item **Schlüsselgenerierung:** 
-            \[
-            \text{Privater Schlüssel } d, \text{ öffentlicher Schlüssel } P = d \cdot G, \text{ wobei } G \text{ ein Basispunkt ist.}
-            \]
-            \item **Verschlüsselung:** 
-            \[
-            \text{Nachricht } m, \text{ Zufallswert } k, \text{ Schlüsseltext: } (C_1, C_2) = (k \cdot G, m + k \cdot P).
-            \]
-            \item **Entschlüsselung:** 
-            \[
-            m = C_2 - d \cdot C_1.
-            \]
-        \end{itemize}
-        \item Schlüsselgenerierung:
-        \begin{itemize}
-            \item Wähle \( d = 3 \) als privaten Schlüssel.
-            \item Berechne \( P = d \cdot G = 3 \cdot (2, 7) = (8, 3) \) (Punkt auf der Kurve).
-            \item Verschlüsselung der Nachricht \( m = (5, 2) \):
-            \[
-            C_1 = k \cdot G = 2 \cdot (2, 7) = (6, 3), \quad C_2 = m + k \cdot P = (5, 2) + 2 \cdot (8, 3) = (7, 6).
-            \]
-            Schlüsseltext: \( (C_1, C_2) = ((6, 3), (7, 6)) \).
-        \end{itemize}
-        \item Entschlüsselung:
-        \[
-        m = C_2 - d \cdot C_1 = (7, 6) - 3 \cdot (6, 3) = (5, 2).
-        \]
-        Die ursprüngliche Nachricht \( m \) wurde korrekt wiederhergestellt.
-        \item Die Sicherheit des Verfahrens beruht auf dem **diskreten Logarithmusproblem** über elliptischen Kurven.
-Ein Angreifer müsste \( k \) aus \( C_1 = k \cdot G \) berechnen, was aufgrund der mathematischen Schwierigkeit des Problems praktisch unmöglich ist.
+        \item 
+          \begin{itemize}
+            \item \textbf{Schlüsselgenerierung} mit Generaturpunkt $P$, privaten Schlüssel $d$ und öffentlicher Schlüssel $Q = d \cdot P$.
+            \item \textbf{Verschlüsselung} von Nachricht $m$ mit Zufallswert $k$ und Schlüsseltext $(C_1, C_2) = (kP, m + kQ)$.
+            \item \textbf{Entschlüsselung} des Schlüsseltextes $(C_1, C_2)$ mit privatem Schlüssel $d$ als $m = C_2 - d \cdot C_1$.
+          \end{itemize}
+        \item
+        \item 
+        \item 
     \end{enumerate}

### 2024-12-06 SbD/summary.tex
Refined some exercises and summary.

@@
   \item \textbf{Erweiterter Euklidischer Algorithmus}: ggT berechnen und erweiterter Algorithmus für Inverse
-    \[ c_k=\frac{r_{k-2}}{r_{k-1}} \quad r_k = r_{k-2} \pmod{r_{k-1}} \quad s_k=s_{k-1}+c_ks_{k-2} \quad t_k=t_{k-1}+c_kt_{k-2} \]
+    \[ c_k=\lfloor\frac{r_{k-2}}{r_{k-1}}\rfloor \quad r_k = r_{k-2} \pmod{r_{k-1}} \quad s_k=s_{k-2}-c_ks_{k-1} \quad t_k=t_{k-2}-c_kt_{k-1} \]
   \item \textbf{Euler-Phi-Funktion}: $\varphi(n)$ gibt die Anzahl der zu $n$ teilerfremden (engl. coprime) Zahlen kleiner als $n$ an.
     \[ \varphi(p \cdot q) = (p-1) \cdot (q-1) \quad \text{Primzahlen: }\varphi(p) = p-1 \]
-  \item \textbf{Primitive Wurzel}: Teilerfremdes $a$ ist primitive Wurzel modulo $n$:
+  \item \textbf{Primitive Wurzel}: Teilerfremdes $a$ ist primitive Wurzel modulo $n$, da $a^d \pmod{n}$ alle Reste erzeugt. Auch übertragen als:
     \[ a^d \not\equiv 1 \pmod{n} \text{ für } 1 \leq d < \varphi(n) \]
@@
     \[
-      \underbrace{p\in\mathbb{P}, \overbrace{a}^\text{primitive Wurzel}}_\text{global}\quad
-      \underbrace{\overbrace{k}^\text{zufällig}\in\mathbb{Z}^*_{p-1}}_\text{privater Schlüssel}\quad
-      \underbrace{y = a^{-k} \pmod{p}}_\text{öffentlicher Schlüssel} 
+      \underbrace{\overbrace{x}^\text{zufällig}\in\mathbb{Z}^*_{p-1}}_\text{geheim}\quad
+      \underbrace{p\in\mathbb{P} \quad \overbrace{a}^\text{Primwurzel} \quad y = a^{-x} \pmod{p}}_\text{öffentlich}
     \]
-  \item \textbf{Verschlüsselung} von Nachricht $m\in\mathbb{Z}^*_{p-1}$ für $B$: $k \in k<p$ zufällig wählen
-    \[ d = a^z \mod{p} \text{ und } c = y_B^z \cdot m \pmod{p} \]
+  \item \textbf{Verschlüsselung} von Nachricht $m\in\mathbb{Z}^*_{p-1}$ mit zufälligem $z\in\mathbb{Z}^*_{p-1}$
+    \[ [d,c] = [a^z, y^z \cdot m] \pmod{p} \]
   \item \textbf{Entschlüsselung} von $a^z \mod{p}$ und $c$
-    \[ m^* = d^{k_B} \cdot c \pmod{p} \]
+    \[ m = d^x \cdot c \pmod{p} \]
 \end{itemize}
@@
     \[
-      \underbrace{p\in\mathbb{P}, \overbrace{a}^\text{primitive Wurzel}}_\text{global}\quad
-      \underbrace{\overbrace{x}^\text{zufällig} \in \mathbb{Z}^*_{p-1}}_\text{privater Schlüssel}\quad
-      \underbrace{y = a^{x} \pmod{p}}_\text{öffentlicher Schlüssel}
+      \underbrace{\overbrace{x}^\text{zufällig}\in\mathbb{Z}^*_{p-1}}_\text{geheim}\quad
+      \underbrace{p\in\mathbb{P} \quad \overbrace{a}^\text{Primwurzel} \quad y = a^{x} \pmod{p}}_\text{öffentlich}
     \]
-  \item \textbf{Signatur} von Nachricht $m\in\mathbb{Z}^*_{p-1}$ für $A$: $z\in\mathbb{Z}^*_{p-1}$ zufällig wählen
-    \[ r = a^z \mod{p} \text{ und } s = k^{-1} \cdot (m - x_A \cdot r) \mod{p-1} \]
+  \item \textbf{Signatur} von Nachricht $m\in\mathbb{Z}^*_{p-1}$ mit zufälligem $z\in\mathbb{Z}^*_{p-1}$
+    \[ r = a^z \pmod{p} \text{ und } s = z^{-1} \cdot (m - x \cdot r) \pmod{p-1} \]
   \item \textbf{Verifikation} von $a^z \mod{p}$ und $r$ und $s$
@@
     \[
-      \underbrace{p,q\in\mathbb{P}}_\text{geheim}\quad
-      \underbrace{n=p\cdot q}_\text{öffentlich}\quad
-      \underbrace{\varphi(n)=(p-1)\cdot(q-1)}_\text{geheim}\quad
-      \underbrace{c\in\mathbb{Z}^*_{\varphi(n)}}_\text{öffentlich}\quad
-      \underbrace{d=c^{-1} \pmod{\varphi(n)}}_\text{geheim}
+      \underbrace{p,q\in\mathbb{P} \quad \varphi(n)=(p-1)\cdot(q-1)}_\text{geheim}\quad
+      \underbrace{n=p\cdot q \quad \overbrace{e}^\text{zufällig}\in\mathbb{Z}^*_{\varphi(n)}}_\text{öffentlich}\quad
     \]
-  \item \textbf{Verschlüsselung} von Nachricht $m\in\mathbb{Z}_n$ für $B$
-    \[ c = m^e \mod{n} \]
+  \item \textbf{Verschlüsselung} von Nachricht $m\in\mathbb{Z}_n$
+    \[ c = m^e \pmod{n} \]
   \item \textbf{Entschlüsselung} von $c$
-    \[ m = c^d \mod{n} \]
+    \[ m = c^d \pmod{n} \text{ mit } d=e^{-1} \pmod{\varphi(n)}\]
   \item \textbf{Signatur}: analog zur Verschlüsselung. Signatur $s$ wird anstelle von $c$ übertragen.

### 2024-12-09 SbD/exercises.tex
Updated some exercises and gsm summary.

@@
           \begin{itemize}
-            \item \textbf{Schlüsselgenerierung} mit Generaturpunkt $P$, privaten Schlüssel $d$ und öffentlicher Schlüssel $Q = d \cdot P$.
+            \item \textbf{Schlüsselgenerierung} mit Generaturpunkt $P$, privatem Schlüssel $d$ und öffentlichem Schlüssel $Q = d \cdot P$.
             \item \textbf{Verschlüsselung} von Nachricht $m$ mit Zufallswert $k$ und Schlüsseltext $(C_1, C_2) = (kP, m + kQ)$.
@@
           \end{itemize}
-        \item
-        \item 
-        \item 
+        \item % TODO
+        \item % TODO
+        \item Die Sicherheit beruht darauf, dass es schwierig ist die Operation $Q=nP$ umzukehren.
     \end{enumerate}
@@
     \begin{itemize}
-        \item \textbf{Verschlüsselte Abschnitte:} Der Abschnitt zwischen dem Mobiltelefon und der Basisstation wird durch den A5-Algorithmus verschlüsselt.
-        \item \textbf{Unverschlüsselte Abschnitte:} Die Verbindung zwischen der Basisstation und den zentralen Netzwerkkomponenten (z.
-B.
-dem MSC) ist unverschlüsselt.
-Dies stellt ein Sicherheitsrisiko dar, da Angriffe auf das Netzwerk (z.
-B.
-durch gefälschte Basisstationen) möglich sind.
+        \item \textbf{Verschlüsselte Abschnitte}: Abschnitt zwischen Mobiltelefon und Basisstation.
+        \item \textbf{Unverschlüsselte Abschnitte}: Basisstation $\to$ zentrale Netzwerkkomponenten (zB. MSC).
     \end{itemize}

### 2024-12-09 SbD/summary.tex
Updated some exercises and gsm summary.

@@
 
-\subsection{Grundlagen und Architektur}
+\subsection{Architektur}
+\textit{Digitales Mobilfunksystem für Sprach- und Datenübertragung.}
 \begin{itemize}
-    \item \textbf{GSM:} Digitales Mobilfunksystem für Sprach- und Datendienste.
-    \item \textbf{Netzarchitektur:}
-    \begin{itemize}
-        \item \textbf{Base Station Subsystem (BSS):} Verbindet mobile Geräte mit dem Netzwerk.
-        \item \textbf{Network Switching Subsystem (NSS):} Verarbeitet Anrufe und Standortinformationen.
-        \item \textbf{Operation and Support Subsystem (OSS):} Überwacht und verwaltet das Netzwerk.
-    \end{itemize}
+    \item \textbf{Base Station Subsystem} (BSS): Verbindet mobile Geräte mit dem Netzwerk.
+    \item \textbf{Network Switching Subsystem} (NSS): Verarbeitet Anrufe und Standortinformationen.
+    \item \textbf{Operation and Support Subsystem} (OSS): Überwacht und verwaltet das Netzwerk.
 \end{itemize}
@@
 \begin{itemize}
-    \item \textbf{Authentifizierung:}
-    \begin{itemize}
-        \item Nutzung eines gemeinsamen Schlüssels (Ki) zur Verifizierung des Benutzers.
-        \item Implementierung durch den Algorithmus A3.
-    \end{itemize}
-    \item \textbf{Verschlüsselung:}
-    \begin{itemize}
-        \item Schutz der Kommunikation durch A5-Algorithmen.
-        \item Ziel: Vertraulichkeit der Sprach- und Datenübertragung.
-    \end{itemize}
-    \item \textbf{Anonymität:}
-    \begin{itemize}
-        \item Nutzung temporärer Identifikatoren (TMSI), um die Identität der Benutzer zu verschleiern.
-    \end{itemize}
+    \item \textbf{Authentifizierung} mit gemeinsamem Schlüssel (Ki) durch A3-Algorithmus.
+    \item \textbf{Verschlüsselung} von Sprach- und Datenübertragung durch A5-Algorithmen.
+    \item \textbf{Anonymität} durch temporäre Identifikatoren (TMSI).
 \end{itemize}
@@
 \begin{itemize}
-    \item \textbf{Kryptografische Schwächen:}
-    \begin{itemize}
-        \item Veraltete Algorithmen (z.B. A5/1) können leicht entschlüsselt werden.
-    \end{itemize}
-    \item \textbf{IMSI-Catcher:}
-    \begin{itemize}
-        \item Geräte, die IMSI (International Mobile Subscriber Identity) abfangen, um Benutzer zu lokalisieren oder zu überwachen.
-    \end{itemize}
-    \item \textbf{Man-in-the-Middle-Angriffe:}
-    \begin{itemize}
-        \item Angreifer simulieren Basisstationen, um Kommunikation abzufangen.
-    \end{itemize}
+    \item \textbf{Kryptografische Schwächen}: Veraltete Algorithmen (z.B. A5/1) können leicht entschlüsselt werden.
+    \item \textbf{IMSI-Catcher}: Geräte, die IMSI (International Mobile Subscriber Identity) abfangen, um Benutzer zu lokalisieren oder zu überwachen.
+    \item \textbf{Man-in-the-Middle-Angriffe}: Angreifer simulieren Basisstationen, um Kommunikation abzufangen.
 \end{itemize}
@@
 \begin{itemize}
-    \item \textbf{GPRS (General Packet Radio Service):}
-    \begin{itemize}
-        \item Erweiterung für datenorientierte Dienste.
-    \end{itemize}
-    \item \textbf{EDGE (Enhanced Data rates for GSM Evolution):}
-    \begin{itemize}
-        \item Höhere Datendurchsätze für Mobilfunkanwendungen.
-    \end{itemize}
+    \item \textbf{GPRS} (General Packet Radio Service): Erweiterung für datenorientierte Dienste.
+    \item \textbf{EDGE} (Enhanced Data rates for GSM Evolution): Höhere Datendurchsätze für Mobilfunkanwendungen.
 \end{itemize}

### 2024-12-09 SbD/exercises.tex
Updated more exercises.

@@
 
-        \item Der geheime Schlüssel ist die Größe der Matrix: \( 7 \times 9 \) (Zeilen $\times$ Spalten). Beim ursprünglichen Zylinder entspricht dies dem Umfang des Zylinders.
+        \item Der geheime Schlüssel ist die Größe der Matrix: $ 7 \times 9 $ (Zeilen $\times$ Spalten). Beim ursprünglichen Zylinder entspricht dies dem Umfang des Zylinders.
 
@@
     \begin{enumerate}
-        \item Bei Spaltenvertauschung gibt es \( s!
-\) mögliche Permutationen für \( s \) Spalten.
-Der Schlüsselraum vergrößert sich somit um den Faktor \( s!
-\) im Vergleich zur ursprünglichen Methode ohne Spaltenvertauschung.
-
-        \item Klartext in eine \( 4 \times 3 \)-Matrix:
+        \item Bei Spaltenvertauschung gibt es $ s!
+$ mögliche Permutationen für $ s $ Spalten.
+Der Schlüsselraum vergrößert sich somit um den Faktor $ s!
+$ im Vergleich zur ursprünglichen Methode ohne Spaltenvertauschung.
+
+        \item Klartext in eine $ 4 \times 3 $-Matrix:
           \[
@@
   \begin{solution}
-    Durch die Wiederverwendung des One-Time-Pads \( k \) für mehrere Nachrichten \( m_1, m_2 \) entsteht folgende Gleichung:
-    \[
-    c_1 \oplus c_2 = (m_1 \oplus k) \oplus (m_2 \oplus k) = m_1 \oplus m_2
-    \]
-    Dies ergibt die XOR-Verknüpfung der Klartexte \( m_1 \oplus m_2 \), die durch Analyse entschlüsselt werden kann. Beispielsweise:
-    \[
-    c_1 = \texttt{09 00 04 10}, \quad c_2 = \texttt{10 20 28 09}
-    \]
+    Durch die Wiederverwendung des One-Time-Pads $ k $ für mehrere Nachrichten $ m_1, m_2 $ entsteht folgende Gleichung:
+      \[ c_1 \oplus c_2 = (m_1 \oplus k) \oplus (m_2 \oplus k) = m_1 \oplus m_2 \]
+    Dies ergibt die XOR-Verknüpfung der Klartexte $ m_1 \oplus m_2 $, die durch Analyse entschlüsselt werden kann. Beispielsweise:
+      \[ c_1 = \texttt{09 00 04 10}, \quad c_2 = \texttt{10 20 28 09} \]
     XOR:
-    \[
-    c_1 \oplus c_2 = \texttt{19 20 2C 19}
-    \]
+      \[ c_1 \oplus c_2 = \texttt{19 20 2C 19} \]
     Häufigkeitsanalysen und bekannte Wörter im Klartext ermöglichen die vollständige Entschlüsselung.
@@
 \begin{exercise}{Meet-in-the-Middle-Angriff auf 2-DES}
-  Um der geringen Schlüssellänge von 56 Bit des DES-Algorithmus zu begegnen, wurde vorgeschlagen, den
-  Algorithmus mehrfach hintereinander mit verschiedenen Schlüsseln anzuwenden. Eine schlechte Lösung ist
-  jedoch die lediglich zweifache Hintereinanderanwendung 2-DES.
+  Um der geringen Schlüssellänge von 56 Bit des DES-Algorithmus zu begegnen, wurde vorgeschlagen, den Algorithmus mehrfach hintereinander mit verschiedenen Schlüsseln anzuwenden.
+Eine schlechte Lösung ist jedoch die lediglich zweifache Hintereinanderanwendung 2-DES.
   \begin{enumerate}
     \item Vergleichen Sie die Schlüssellänge des 2-DES-Algorithmus mit der tatsächlich erreichten Sicherheit.
-    \item Sie haben folgendes Klartext/Schlüsseltext-Paar \( (m, c) \) einer 2-DES-Verschlüsselung \( E_{k_2}(E_{k_1}(m)) = c \)
-    erlangt: 
-    \[
-        m = \texttt{c183 50b5 52b8 f8b1}, \quad c = \texttt{42e0 4462 099e e8ca}.
-    \]
-    Ermitteln Sie \( k_1 \) und \( k_2 \), indem Sie einen Meet-in-the-Middle-Angriff durchführen.
+    \item Sie haben folgendes Klartext/Schlüsseltext-Paar $ (m, c) $ einer 2-DES-Verschlüsselung $ E_{k_2}(E_{k_1}(m)) = c $ erlangt: 
+      \[ m = \texttt{c183 50b5 52b8 f8b1}, \quad c = \texttt{42e0 4462 099e e8ca}. \]
+      Ermitteln Sie $ k_1 $ und $ k_2 $, indem Sie einen Meet-in-the-Middle-Angriff durchführen.
 Benutzen Sie dafür die unten dargestellte Tabelle 1 ausgewählter DES-Funktionswerte.
-In der Tabelle können für einen gegebenen Klartext \( x \) bzw.
-einen Schlüsseltext \( y \) und einen Schlüssel \( k \) die Werte der Verschlüsselungsfunktion \( E \) bzw.
-Entschlüsselungsfunktion \( D \) des DES-Algorithmus nachgeschlagen werden.
+In der Tabelle können für einen gegebenen Klartext $ x $ bzw.
+einen Schlüsseltext $ y $ und einen Schlüssel $ k $ die Werte der Verschlüsselungsfunktion $ E $ bzw.
+Entschlüsselungsfunktion $ D $ des DES-Algorithmus nachgeschlagen werden.
 In welchen Zeilen haben Sie die DES-Funktionswerte, die zum Erfolg beim Meet-in-the-Middle Angriff führen, nachgeschlagen?
-    Hinweis: Beachten Sie das Beispiel zum Ablesen der Tabelle.
-
+      \textit{Hinweis: Beachten Sie das Beispiel zum Ablesen der Tabelle.}
       \input{res/des.tbl}
-
     \item Wie viele Einträge enthält eine vollständige Tabelle, die bei einem Meet-in-the-Middle-Angriff auf den 2-DES-Algorithmus erstellt wird, mindestens?
@@
     \begin{enumerate}
-        \item 2-DES bietet nur \( 2^{56} \) statt \( 2^{112} \) effektiver Sicherheit, da ein Meet-in-the-Middle-Angriff \( E_{k_1}(m) = D_{k_2}(c) \) nutzt.
+        \item 2-DES bietet nur $ 2^{56} $ statt $ 2^{112} $ effektiver Sicherheit, da ein Meet-in-the-Middle-Angriff $ E_{k_1}(m) = D_{k_2}(c) $ nutzt.
         \item Analyse aus Tabelle 1:
@@
         \]
-        \item Eine vollständige Tabelle umfasst \( 2^{56} \) Einträge, wobei \( E_k(m) \) oder \( D_k(c) \) für alle \( k \) berechnet werden.
+        \item Eine vollständige Tabelle umfasst $ 2^{56} $ Einträge, wobei $ E_k(m) $ oder $ D_k(c) $ für alle $ k $ berechnet werden.
     \end{enumerate}
@@
   Beim Triple-DES-Algorithmus (3-DES) handelt es sich um die dreifache Hintereinanderausführung des DES-
-  Algorithmus. Dabei wird mit zwei 56 Bit langen Schlüsseln \( k_1 \) und \( k_2 \) gearbeitet und beim Verschlüsseln
-  entweder \( E(k_1) \to D(k_2) \to E(k_1) \) (EDE-Modus) oder \( E(k_1) \to E(k_2) \to E(k_1) \) (EEE-Modus) ausgeführt.
+  Algorithmus. Dabei wird mit zwei 56 Bit langen Schlüsseln $ k_1 $ und $ k_2 $ gearbeitet und beim Verschlüsseln
+  entweder $ E(k_1) \to D(k_2) \to E(k_1) $ (EDE-Modus) oder $ E(k_1) \to E(k_2) \to E(k_1) $ (EEE-Modus) ausgeführt.
   \begin{enumerate}
@@
     \item Berechnen Sie die theoretische und die effektive Schlüssellänge des Triple-DES-Algorithmus unter ei-
-    nem Known-Plaintext-Angriff mit drei 56 Bit langen Schlüsseln, d.h. \( E(k_1) \to E(k_2) \to E(k_3) \).
+    nem Known-Plaintext-Angriff mit drei 56 Bit langen Schlüsseln, d.h. $ E(k_1) \to E(k_2) \to E(k_3) $.
   \end{enumerate}
@@
     \item Inwiefern stellt dies ein Sicherheitsproblem dar?
-    \item Veranschaulichen Sie das Problem, indem Sie ein Programm entwickeln, welches eine einfache Bitmap-
-    Grafik einliest, den Bildinhalt mit dem AES-Verfahren im ECB-Modus verschlüsselt und wieder als
-    Bitmap ausgibt. Rufen Sie zum Vergleich die Verschlüsselung noch einmal im CBC-Modus auf und
-    vergleichen Sie die erhaltenen Resultate. Informieren Sie sich dazu über das Datenformat von BMP-
-    Dateien und die Verwendung einer geeigneten Kryptobibliothek (z. B. die Java-Cryptography-API
-    unter \url{http://www.ibm.com/developerworks/java/tutorials/j-sec1/section4.html} oder die Python-pycrypto-
-    API unter \url{https://www.dlitz.net/software/pycrypto/api/current/}).
+    \item Veranschaulichen Sie das Problem, indem Sie ein Programm entwickeln, welches eine einfache Bitmap-Grafik einliest, den Bildinhalt mit dem AES-Verfahren im ECB-Modus verschlüsselt und wieder als Bitmap ausgibt.
+Rufen Sie zum Vergleich die Verschlüsselung noch einmal im CBC-Modus auf und vergleichen Sie die erhaltenen Resultate.
+Informieren Sie sich dazu über das Datenformat von BMP-Dateien und die Verwendung einer geeigneten Kryptobibliothek (z.
+B.
+die Java-Cryptography-API unter \url{http://www.ibm.com/developerworks/java/tutorials/j-sec1/section4.html} oder die Python-pycrypto- API unter \url{https://www.dlitz.net/software/pycrypto/api/current/}).
   \end{enumerate}
@@
 \begin{exercise}{Cipher-Block-Chaining-Betriebsmodus}
-  Beim Cipher-Block-Chaining-Betriebsmodus wird ein Klartextblock \( M_i \) vor der Anwendung der Verschlüsselungsfunktion mit dem unmittelbar vorher erzeugten Chiffretextblock \( C_{i-1} \) durch die XOR-Operation verknüpft.
+  Beim Cipher-Block-Chaining-Betriebsmodus wird ein Klartextblock $ M_i $ vor der Anwendung der Verschlüsselungsfunktion mit dem unmittelbar vorher erzeugten Chiffretextblock $ C_{i-1} $ durch die XOR-Operation verknüpft.
   Die erzeugten Chiffretextblöcke hängen dadurch von ihren Vorgängern ab.
-Bei der Verschlüsselung des ersten Klartextblocks \( M_1 \) wird für \( C_0 \) ein zufällig gewählter Initialisierungsvektor (IV) verwendet.
+Bei der Verschlüsselung des ersten Klartextblocks $ M_1 $ wird für $ C_0 $ ein zufällig gewählter Initialisierungsvektor (IV) verwendet.
 
@@
 \begin{exercise}{Verketteter Cipher-Block-Chaining-Betriebsmodus}
-  Beim verketteten Cipher-Block-Chaining-Betriebsmodus wird zur Steigerung des Datendurchsatzes nur für die erste Nachricht \( M_1 \) ein Initialisierungsvektor \( IV \) gewählt.
-Das heißt für den Schlüsseltext \( C_1 = C_1^0 || \ldots || C_1^{n_1} \) der ersten Nachricht \( M_1 = M_1^1 || \ldots || M_1^{n_1} \) gilt wie beim normalen CBC:
+  Beim verketteten Cipher-Block-Chaining-Betriebsmodus wird zur Steigerung des Datendurchsatzes nur für die erste Nachricht $ M_1 $ ein Initialisierungsvektor $ IV $ gewählt.
+Das heißt für den Schlüsseltext $ C_1 = C_1^0 || \ldots || C_1^{n_1} $ der ersten Nachricht $ M_1 = M_1^1 || \ldots || M_1^{n_1} $ gilt wie beim normalen CBC:
   \[
@@
   \]
-  Für alle folgenden Nachrichten \( M_2, \ldots \) wird anstelle des \( IV \) der letzte Schlüsseltextblock des vorherigen Schlüsseltextes verwendet.
-Das heißt für den Schlüsseltext \( C^j = C^j_1 || \ldots || C^j_{n_j} \) aller weiteren Nachrichten \( M_j = M_j^1 || \ldots || M_j^{n_j} \) mit \( j > 1 \) gilt:
+  Für alle folgenden Nachrichten $ M_2, \ldots $ wird anstelle des $ IV $ der letzte Schlüsseltextblock des vorherigen Schlüsseltextes verwendet.
+Das heißt für den Schlüsseltext $ C^j = C^j_1 || \ldots || C^j_{n_j} $ aller weiteren Nachrichten $ M_j = M_j^1 || \ldots || M_j^{n_j} $ mit $ j > 1 $ gilt:
   \[
@@
 Führen Sie die Berechnungen sowohl für einen Block als auch für vier Blöcke pro Nachricht aus.
-Gehen Sie in beiden Fällen von insgesamt \( 2^{20} \) zu verschlüsselnden Nachrichten vor Neugenerierung eines Schlüssels und Neustart mit einem neuen \( IV \) aus.
-    \item Ihnen ist ein Schlüsseltext \( C^j \) in die Hände gefallen, der nur aus zwei möglichen Klartexten \( M_j \) oder \( \tilde{M}_j \) gebildet worden sein kann.
-Erläutern Sie, wie Sie ermitteln könnten, welcher der beiden Klartexte verschlüsselt wurde, wenn Sie den Klartext \( M_{j+1} \) für die nächste Nachricht selbst wählen dürfen.
+Gehen Sie in beiden Fällen von insgesamt $ 2^{20} $ zu verschlüsselnden Nachrichten vor Neugenerierung eines Schlüssels und Neustart mit einem neuen $ IV $ aus.
+    \item Ihnen ist ein Schlüsseltext $ C^j $ in die Hände gefallen, der nur aus zwei möglichen Klartexten $ M_j $ oder $ \tilde{M}_j $ gebildet worden sein kann.
+Erläutern Sie, wie Sie ermitteln könnten, welcher der beiden Klartexte verschlüsselt wurde, wenn Sie den Klartext $ M_{j+1} $ für die nächste Nachricht selbst wählen dürfen.
     \item Beschreiben Sie ein Szenario für einen derartigen Angriff.
@@
         \item Bei einem verketteten CBC-Modus steigt der Datendurchsatz, da kein neuer IV benötigt wird.
-Für \( 2^{20} \) Nachrichten bei einer Blockgröße von \( 128 \) Bit beträgt der Effizienzgewinn ca.
+Für $ 2^{20} $ Nachrichten bei einer Blockgröße von $ 128 $ Bit beträgt der Effizienzgewinn ca.
 0,01\%, da der Initialisierungsaufwand entfällt.
-        \item Wenn \( M_j \) oder \( \tilde{M}_j \) verschlüsselt wurde, kann durch die Manipulation des nächsten Klartextblocks \( M_{j+1} \) festgestellt werden, welcher Klartext verwendet wurde.
+        \item Wenn $ M_j $ oder $ \tilde{M}_j $ verschlüsselt wurde, kann durch die Manipulation des nächsten Klartextblocks $ M_{j+1} $ festgestellt werden, welcher Klartext verwendet wurde.
 Dies ermöglicht einen Known-Plaintext-Angriff.
@@
 \begin{exercise}{Elliptische Kurven: Rechenbeispiele zur Punktaddition}
-  Gegeben sei die elliptische Kurve \( E(\mathbb{GF}(p)) \) mit \( a = 1, b = 6, p = 11 \).
-Hinweis: Wir betrachten für diese Aufgabe die in der Kryptographie häufig genutzten elliptischen Kurven über \( \mathbb{Z}_p \) der Form \( f(x, y) : y^2 \equiv x^3 + ax + b \mod p \).
+  Gegeben sei die elliptische Kurve $E(\mathbb{GF}(p))$ mit $a = 1, b = 6, p = 11$:
+    \[ E(\mathbb{GF}(11)) : y^2 \equiv x^3 + x + 6 \pmod{11} \]
+  \textit{Hinweis: Wir betrachten für diese Aufgabe die in der Kryptographie häufig genutzten elliptischen Kurven über $ \mathbb{Z}_p $ der Form $f(x, y) : y^2 \equiv x^3 + ax + b \mod p$.}
   \begin{enumerate}
@@
     \item Suchen Sie sich zwei beliebige Punkte auf der Kurve und addieren Sie diese. Wie funktioniert die Punktaddition im Allgemeinen?
-    \item Berechnen Sie die skalare Multiplikation des Punktes \( (2, 4) \) mit dem Skalar \( 3 \). Was ist das allgemeine Vorgehen hier?
-  \end{enumerate}
-
-  \begin{solution}
-    \begin{enumerate}
-        \item Punkte auf der elliptischen Kurve \( y^2 = x^3 + x + 6 \mod 11 \):
-        \[
-        \{(2, 4), (5, 2), (7, 6), (8, 3)\}
-        \]
-        \item Beispiel Punktaddition:
-        \[
-        P = (2, 4), Q = (5, 2) \quad \text{Ergebnis: } P + Q = (7, 6)
-        \]
-        \item Skalare Multiplikation:
-        \[
-        3 \cdot (2, 4) = (8, 3)
-        \]
-        Vorgehen: Addieren Sie \( P \) wiederholt zu sich selbst.
+    \item Berechnen Sie die skalare Multiplikation des Punktes $(2, 4)$ mit dem Skalar $ 3 $. Was ist das allgemeine Vorgehen hier?
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+        \item $\{(2,4), (2,7), (3,9), (3,2), (5,2), (5,8), (7,5), (7,6), (8,9), (8,2), (10,1), (10,10)\}$
+        \item $P+Q=(2,4)+(5,2)=(5,2)$.
+          \[
+            m=\begin{cases}
+              \frac{3x_P^2+a}{2y_P} \quad \text{falls } P=Q \\
+              \frac{y_Q-y_P}{x_Q-x_P}
+            \end{cases} \quad
+            \begin{array}{l}
+              x_R=m^2-x_P-x_Q\\
+              y_R=m(x_R-x_{P|Q})-y_{P|Q}
+            \end{array} \pmod{p} \]
+        \item $3 \cdot (2,4)=(2,4)$. Double-and-add:
+          \[ nP=2^{k-1}P+2^{k-2}P+\ldots+2P+P \]
     \end{enumerate}
@@
 \begin{exercise}{Forward Secrecy mit RSA}
-  Zur Erreichung von Forward Secrecy erzeugt Alice auf ihrem Server mit dem RSA-Verfahren einen privaten Schlüssel \( d \) sowie den dazu passenden öffentlichen Schlüssel \( c \), den sie persönlich an ihre Nutzerinnen und Nutzer verteilt.
+  Zur Erreichung von Forward Secrecy erzeugt Alice auf ihrem Server mit dem RSA-Verfahren einen privaten Schlüssel $d$ sowie den dazu passenden öffentlichen Schlüssel $c$, den sie persönlich an ihre Nutzerinnen und Nutzer verteilt.
 Zusätzlich erzeugt der Server von Alice bei jedem Verbindungsaufbau ein neues kurzlebiges (engl.
-ephemeral) RSA-Schlüsselpaar, das aus einem öffentlichen Schlüssel \( c_e \) und einem geheimen Schlüssel \( d_e \) besteht.
-Der Server signiert mit dem privaten Schlüssel \( d \) den neuen öffentlichen Schlüssel \( c_e \) und sendet diesen zusammen mit der Signatur über die aufgebaute Verbindung an die Kundin.
-Die Kundin prüft anschließend die Signatur mithilfe des zuvor von Alice erhaltenen öffentlichen Schlüssels \( c \).
-  \begin{enumerate}
-    \item Welche Verbindungen könnten von einem sehr starken Angreifer wie einem Nachrichtendienst noch entschlüsselt werden, wenn dieser an \( d \) gelangt? Begründen Sie Ihre Antwort.
+ephemeral) RSA-Schlüsselpaar, das aus einem öffentlichen Schlüssel $ c_e $ und einem geheimen Schlüssel $ d_e $ besteht.
+Der Server signiert mit dem privaten Schlüssel $ d $ den neuen öffentlichen Schlüssel $ c_e $ und sendet diesen zusammen mit der Signatur über die aufgebaute Verbindung an die Kundin.
+Die Kundin prüft anschließend die Signatur mithilfe des zuvor von Alice erhaltenen öffentlichen Schlüssels $ c $.
+  \begin{enumerate}
+    \item Welche Verbindungen könnten von einem sehr starken Angreifer wie einem Nachrichtendienst noch entschlüsselt werden, wenn dieser an $1$ gelangt? Begründen Sie Ihre Antwort.
     \item Skizzieren Sie einen Angriff, der möglich wäre, wenn die Kundin die Signatur nicht prüft.
     \item Alice ist das Verfahren zu langsam.
-Sie nimmt folgende Änderung vor: Statt jedes Mal bei der Erzeugung von \( c_e \) und \( d_e \) neue Primzahlen \( p \) und \( q \) für den Modulus \( N_e \) zu generieren, verwendet sie immer die gleichen Primzahlen.
-Nun wählt sie lediglich den öffentlichen Exponenten \( c_e \) neu und berechnet einen dazugehörigen geheimen Exponenten \( d_e \).
+Sie nimmt folgende Änderung vor: Statt jedes Mal bei der Erzeugung von $c_e$ und $d_e$ neue Primzahlen $p$ und $q$ für den Modulus $N_e$ zu generieren, verwendet sie immer die gleichen Primzahlen.
+Nun wählt sie lediglich den öffentlichen Exponenten $ c_e $ neu und berechnet einen dazugehörigen geheimen Exponenten $ d_e $.
 Warum ist dies keine gute Idee?
@@
     \begin{enumerate}
-        \item Forward Secrecy wird durch kurzlebige (ephemeral) RSA-Schlüssel erreicht. Jede neue Verbindung erhält ein neues Schlüsselpaar.
-        \item Wenn ein Angreifer den privaten Schlüssel erhält, kann er vergangene Verbindungen entschlüsseln, falls diese nicht mit ephemeral-Schlüsseln abgesichert wurden.
-        \item Ein Angreifer kann die Kommunikation entschlüsseln, indem er den abgefangenen Schlüsseltext mit dem kompromittierten privaten Schlüssel entschlüsselt.
-        \item Forward Secrecy kann durch regelmäßiges Rotieren der Schlüssel gewährleistet werden. Alte Schlüssel werden sicher gelöscht.
+      \item Ein starker Angreifer kann alle Verbindungen entschlüsseln, wenn er $d$ erhält, da er damit alle geheimen Schlüssel $d_e$ berechnen kann.
+      \item Es könnten gefälschte öffentliche Schlüssel $c_e$ und Signaturen an die Kundin gesendet werden, die nicht von Alice stammen.
+      \item Durch die Verwendung der gleichen Primzahlen $p$ und $q$ für $N$ und $N_e$ kann ein Angreifer $d$ berechnen, wenn er $d_e$ kennt. Dies ermöglicht das Entschlüsseln aller Verbindungen.
     \end{enumerate}
@@
     \begin{enumerate}
-        \item Ablauf:
-        \[
-        g = 5, p = 23, a = 6, b = 15 \quad \text{(geheime Schlüssel)}
-        \]
-        Öffentliche Schlüssel:
-        \[
-        A = g^a \mod p = 8, \quad B = g^b \mod p = 19
-        \]
-        Gemeinsamer Schlüssel:
-        \[
-        K = B^a \mod p = 2 = A^b \mod p
-        \]
-        \item Sicherheit basiert auf dem diskreten Logarithmusproblem (DLP). Der Angreifer kann \( g^a \) und \( g^b \) nicht effizient in \( a \) und \( b \) umwandeln.
-        \item Schutz vor aktiven Angriffen: Vorab-Austausch von \( g, p \) und Authentifizierung der öffentlichen Schlüssel (z. B. durch Signaturen).
+        \item $g=5, p=11, \underbrace{a=2, b=3}_\text{geheim}$.\\
+          Öffentliche Schlüssel: $\begin{array}{l}
+              A\equiv g^a \equiv 5^2 \equiv 3\\
+              B\equiv g^b \equiv 5^3 \equiv 4
+            \end{array} \pmod{11}$\\
+          Gemeinsamer Schlüssel: $K\equiv B^a\equiv A^b\equiv 4^2\equiv 3^3\equiv 5 \pmod{11}$
+        \item Sicherheit basiert auf dem diskreten Logarithmusproblem (DLP). Der Angreifer kann $ g^a $ und $ g^b $ nicht effizient in $ a $ und $ b $ umwandeln.
+        \item Schutz vor aktiven Angriffen: Vorab-Austausch von $ g, p $ und Authentifizierung der öffentlichen Schlüssel (z. B. durch Signaturen).
     \end{enumerate}
@@
     \begin{enumerate}
-        \item Die Identitätsfunktion \( H(x) = x \) ist keine kryptographische Hashfunktion, da sie keine feste Ausgabelänge erzeugt, keine Einwegfunktion ist und keine Pseudozufälligkeit aufweist.
+        \item Die Identitätsfunktion $ H(x) = x $ ist keine kryptographische Hashfunktion, da sie keine feste Ausgabelänge erzeugt, keine Einwegfunktion ist und keine Pseudozufälligkeit aufweist.
         \item Kollisionsfreiheit ist entscheidend, da der Angreifer sonst zwei Nachrichten mit identischem Hashwert erzeugen könnte, um Signaturen zu fälschen.
@@
     \begin{itemize}
-        \item \textbf{Input:} Der Eingabewert der Funktion \( E22 \) ist ein gemeinsames Schlüsselpaar (Link Key), das während der Kopplung zwischen zwei Bluetooth-Geräten ausgetauscht wird.
+        \item \textbf{Input:} Der Eingabewert der Funktion $ E22 $ ist ein gemeinsames Schlüsselpaar (Link Key), das während der Kopplung zwischen zwei Bluetooth-Geräten ausgetauscht wird.
         \item \textbf{Output:} Die Funktion erzeugt einen Authentication Code, der verwendet wird, um die Identität der Geräte gegenseitig zu authentifizieren.
@@
     \begin{enumerate}
-        \item Maximal \( 2^{24} \) Pakete können erzeugt werden, da der IV 24 Bit groß ist.
+        \item Maximal $ 2^{24} $ Pakete können erzeugt werden, da der IV 24 Bit groß ist.
 Um Kollisionen zu vermeiden, sollte der IV systematisch inkrementiert werden.
 Problematisch ist jedoch, dass bei mehreren Teilnehmern des WLANs die Wahrscheinlichkeit von IV-Kollisionen schnell steigt.
-        \item Die ersten 10 Klartextbits des zweiten Pakets \( P_2 \) werden durch Vergleich mit \( P_1 \) berechnet:
+        \item Die ersten 10 Klartextbits des zweiten Pakets $ P_2 $ werden durch Vergleich mit $ P_1 $ berechnet:
         \[
@@
         \]
-        Durch eine bekannte Klartextanalyse kann \( M_2 \) bestimmt werden.
+        Durch eine bekannte Klartextanalyse kann $ M_2 $ bestimmt werden.
     \end{enumerate}
@@
         \item Die Sicherheit elliptischer Kurven beruht auf dem Problem des diskreten Logarithmus in elliptischen Kurvengruppen, das als schwierig gilt.
-        \item Punkte auf der elliptischen Kurve \( y^2 = x^3 + 6x + 1 \mod 11 \):
+        \item Punkte auf der elliptischen Kurve $ y^2 = x^3 + 6x + 1 \mod 11 $:
         \[
@@
         \begin{itemize}
-            \item \( K_i \): Geheimer Authentifizierungsschlüssel.
-            \item \( IMSI \): Internationale Teilnehmerkennung.
-            \item \( RAND \): Zufallszahl für Authentifizierung.
-            \item \( SRES \): Signatur zur Authentifizierung.
-            \item \( LAI \): Standortbereichskennung.
-            \item \( MSISDN \): Telefonnummer des Teilnehmers.
-            \item \( K_c \): Verschlüsselungsschlüssel für den Funkkanal.
-        \end{itemize}
-        \item Viele der Funktionen wie \( K_i \), \( RAND \) und \( K_c \) finden sich auch im UMTS wieder, jedoch oft mit anderen Algorithmen oder Benennungen.
+            \item $ K_i $: Geheimer Authentifizierungsschlüssel.
+            \item $ IMSI $: Internationale Teilnehmerkennung.
+            \item $ RAND $: Zufallszahl für Authentifizierung.
+            \item $ SRES $: Signatur zur Authentifizierung.
+            \item $ LAI $: Standortbereichskennung.
+            \item $ MSISDN $: Telefonnummer des Teilnehmers.
+            \item $ K_c $: Verschlüsselungsschlüssel für den Funkkanal.
+        \end{itemize}
+        \item Viele der Funktionen wie $ K_i $, $ RAND $ und $ K_c $ finden sich auch im UMTS wieder, jedoch oft mit anderen Algorithmen oder Benennungen.
     \end{enumerate}
@@
       \end{itemize}
-      \item Die Chiffre, die Buchstaben \( A-N \) mit \( N-Z \) vertauscht, ist nicht sicher:
+      \item Die Chiffre, die Buchstaben $ A-N $ mit $ N-Z $ vertauscht, ist nicht sicher:
       \begin{itemize}
@@
     \begin{enumerate}
-        \item Punkte auf der elliptischen Kurve \( y^2 = x^3 + 2x + 3 \mod 7 \):
+        \item Punkte auf der elliptischen Kurve $ y^2 = x^3 + 2x + 3 \mod 7 $:
         \[
@@
         \end{itemize}
-        \item Zertifikatsbaum: N7 vertraut CAs, die N3 und \( B \) direkt zertifizieren.
-        \item Minimale Anzahl an CAs: Es reicht aus, der CA zu vertrauen, die die Zertifikate für \( B \) und \( N3 \) signiert hat.
+        \item Zertifikatsbaum: N7 vertraut CAs, die N3 und $ B $ direkt zertifizieren.
+        \item Minimale Anzahl an CAs: Es reicht aus, der CA zu vertrauen, die die Zertifikate für $ B $ und $ N3 $ signiert hat.
         \item Maßnahme bei Kompromittierung: Alternativer Zertifizierungsweg über eine andere CA. Neue Zertifikate können erstellt werden, um Vertrauen zu erhalten.
@@
         \begin{itemize}
-            \item \( K_i \): Geheimer Schlüssel zur Authentifizierung und Verschlüsselung.
-            \item \( IMSI \): Einzigartige Kennung des Mobilfunkteilnehmers.
-            \item \( RAND \): Zufällige Herausforderung zur Authentifizierung.
-            \item \( SRES \): Antwort des Mobilgeräts auf \( RAND \), basierend auf \( K_i \).
-            \item \( LAI \): Standortkennung für Mobilitätsmanagement.
-            \item \( MSISDN \): Telefonnummer des Teilnehmers.
-            \item \( K_c \): Schlüssel zur Verschlüsselung des Datenverkehrs.
+            \item $ K_i $: Geheimer Schlüssel zur Authentifizierung und Verschlüsselung.
+            \item $ IMSI $: Einzigartige Kennung des Mobilfunkteilnehmers.
+            \item $ RAND $: Zufällige Herausforderung zur Authentifizierung.
+            \item $ SRES $: Antwort des Mobilgeräts auf $ RAND $, basierend auf $ K_i $.
+            \item $ LAI $: Standortkennung für Mobilitätsmanagement.
+            \item $ MSISDN $: Telefonnummer des Teilnehmers.
+            \item $ K_c $: Schlüssel zur Verschlüsselung des Datenverkehrs.
         \end{itemize}

### 2024-12-10 SbD/exercises.tex
Updated summary.
Updated exam.

@@
     \begin{enumerate}
-        \item Die Modi ECB und CBC benötigen eine invertierbare Verschlüsselungsfunktion, da der Klartext beim Entschlüsseln zurückgewonnen werden muss.
-Der CTR-Modus hingegen benötigt keine invertierbare Funktion, da er auf dem XOR-Mechanismus basiert.
-        \item Die Verschlüsselung lässt sich im CTR-Modus parallelisieren, da die Blöcke unabhängig voneinander verarbeitet werden können.
-Im CBC-Modus ist dies nicht möglich, da jeder Block von seinem Vorgänger abhängt.
-        \item Ein Bitfehler im zweiten Block beeinflusst nur den entsprechenden Block im ECB-Modus. Im CBC-Modus beeinflusst er zusätzlich den nachfolgenden Block.
-        \item Wenn ein Bit im zweiten Block fehlt, kann nur der ECB-Modus korrekt entschlüsseln, da die Blöcke unabhängig verarbeitet werden.
-Im CBC- und CTR-Modus führt der Fehler zu Synchronisationsverlust.
-        \item Beim Verlust eines ganzen Blocks kann nur der ECB-Modus korrekt entschlüsseln, da die Blöcke unabhängig verarbeitet werden.
-        \item Im CTR-Modus kann ein Angreifer zwei Klartexte entschlüsseln, wenn der gleiche IV verwendet wurde:
-        \[
-        C_1 \oplus C_2 = M_1 \oplus M_2.
-        \]
-        Durch eine bekannte Klartextanalyse kann $ M_2 $ bestimmt werden.
+        \item ECB und CBC benötigen invertierbare Verschlüsselungsfunktionen, da die Entschlüsselung der Chiffre notwendig ist.
+Der CTR beruht auf XOR-Verknüpfung der Klar- und Schlüsseltexte und Berechnung und Verschlüsselung der Zählerwerte sowohl beim Sender als auch beim Empfänger.
+Daher ist keine invertierbare Verschlüsselungsfunktion notwendig.
+        \item Im ECB und CTR-Modus können die Blöcke parallel verschlüsselt werden, da sie unabhängig voneinander sind.
+Im CBC-Modus hängt die Verschlüsselung eines Blocks von der Verschlüsselung des vorherigen Blocks ab, daher ist keine Parallelisierung möglich.
+        \item Bei allen Modi führt ein Bitfehler im zweiten Block dazu, dass dieser Block nicht entschlüsselt werden kann.
+Im CBC-Modus wird zusätzlich der Block 3 verändert, da die XOR-Verknüpfung mit dem vorherigen Block nicht mehr korrekt ist.
+        \item Wenn die Blockgrenzen nicht bekannt sind, kann nur der erste Block entschlüsselt werden.
+Im CBC kann zusätzlich selbst bei bekannten Blockgrenzen nichts weiter entschlüsselt werden, da die XOR-Verknüpfung mit dem vorherigen Block nicht mehr korrekt ist.
+        \item Bei ECB und CTR können alle anderen Blöcke korrekt entschlüsselt werden, da sie unabhängig voneinander sind.
+Im CBC-Modus kann kein nachfolgender Block entschlüsselt werden, da die XOR-Verknüpfung mit dem vorherigen Block nicht mehr korrekt ist.
+        \item Da $m_1$ und $c_1$ kann daraus der Pseudozufallszahlenstrom $s=c_1\oplus m_1$ berechnet werden. Hiermit lässt sich dann $m_2=c_2\oplus s$ berechnen.
     \end{enumerate}
@@
     \begin{enumerate}
-        \item Die Sicherheit elliptischer Kurven beruht auf dem Problem des diskreten Logarithmus in elliptischen Kurvengruppen, das als schwierig gilt.
-        \item Punkte auf der elliptischen Kurve $ y^2 = x^3 + 6x + 1 \mod 11 $:
-        \[
-        \{(2, 4), (5, 2), (7, 6), (8, 3), \ldots\}.
-        \]
+        \item ECDLP (Elliptic Curve Discrete Logarithm Problem): Das Problem besteht darin, die Operation $Q=nP$, welche mathematisch einfach zu berechnen ist, umzukehren.
+        \item Punkte auf der elliptischen Kurve $y^2 = x^3 + 6x + 1 \pmod{11}$:
+          \[ \{(2,4), (4,5), (4,6), (6,3), (6,8), (7,5), (7,6), (8,2), (8,9), (9,4), (9,7), (10,1), (10,10)\} \]
         \item Vorteile von ECC:
@@
 \begin{exercise}{Zertifikate}
-  Das Zertifigierungsmodell ist transitiv anzunehmen.
+  Das Zertifizierungsmodell ist transitiv anzunehmen.
+
+  \input{res/2023-1.fig}
 
@@
           \begin{itemize}
-            \item Kein zentraler CA-Betrieb notwendig.
-            \item Schlüssel können einfach ausgetauscht werden.
-        \end{itemize}
-        Nachteile:
-        \begin{itemize}
-            \item Unsicherheit beim ersten Austausch.
-            \item Angreifer können Schlüssel bei der Erstkommunikation abfangen.
-        \end{itemize}
-    \end{enumerate}
-  \end{solution}
-\end{exercise}
-
-\begin{exercise}{GSM}
+              \item[+] Kein zentraler CA-Betrieb notwendig.
+              \item[+] Schlüssel können einfach ausgetauscht werden.
+              \item[-] Unsicherheit beim ersten Austausch.
+              \item[-] Bei Schlüsseländerungen kann es zu Problemen kommen.
+          \end{itemize}
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Forward Secrecy mit RSA}
   Zur Erreichung von Forward Secrecy erzeugt Alice auf ihrem Server mit dem RSA-Verfahren einen privaten Schlüssel $d$ sowie den dazu passenden öffentlichen Schlüssel $c$, den sie persönlich an ihre Nutzerinnen und Nutzer verteilt.
@@
   \begin{solution}
-    \begin{enumerate}
-        \item Abkürzungen und ihre Funktion:
-        \begin{itemize}
-            \item $ K_i $: Geheimer Authentifizierungsschlüssel.
-            \item $ IMSI $: Internationale Teilnehmerkennung.
-            \item $ RAND $: Zufallszahl für Authentifizierung.
-            \item $ SRES $: Signatur zur Authentifizierung.
-            \item $ LAI $: Standortbereichskennung.
-            \item $ MSISDN $: Telefonnummer des Teilnehmers.
-            \item $ K_c $: Verschlüsselungsschlüssel für den Funkkanal.
-        \end{itemize}
-        \item Viele der Funktionen wie $ K_i $, $ RAND $ und $ K_c $ finden sich auch im UMTS wieder, jedoch oft mit anderen Algorithmen oder Benennungen.
-    \end{enumerate}
+    % TODO
   \end{solution}

### 2024-12-10 SbD/exercises.tex
Added more sbd exercises.

@@
           Gemeinsamer Schlüssel: $K\equiv B^a\equiv A^b\equiv 4^2\equiv 3^3\equiv 5 \pmod{11}$
-        \item Sicherheit basiert auf dem diskreten Logarithmusproblem (DLP). Der Angreifer kann $ g^a $ und $ g^b $ nicht effizient in $ a $ und $ b $ umwandeln.
-        \item Schutz vor aktiven Angriffen: Vorab-Austausch von $ g, p $ und Authentifizierung der öffentlichen Schlüssel (z. B. durch Signaturen).
+        \item Sicherheit basiert auf dem diskreten Logarithmusproblem (DLP). Der Angreifer kann $g^a$ und $g^b$ nicht effizient in $a$ und $b$ umwandeln.
+        \item Schutz vor aktiven Angriffen: Vorab-Austausch von $g, p$ und Authentifizierung der öffentlichen Schlüssel (z. B. durch Signaturen).
     \end{enumerate}
@@
 
-% \setcounter{subsection}{70}
-% \begin{exercise}{}
-%  % TODO
-%
-%   \begin{solution}
-%     % TODO
-%   \end{solution}
-% \end{exercise}
-
-% \setcounter{subsection}{80}
-% \begin{exercise}{}
-%  % TODO
-%
-%   \begin{solution}
-%     % TODO
-%   \end{solution}
-% \end{exercise}
+\setcounter{subsection}{70}
+\begin{exercise}{Symmetrische Schlüsselverteilung}
+  Drei Schlüsselverteilzentralen X, Y, Z sollen jeweils einen Anteil eines symmetrischen Schlüssels $k$ der Länge $\ell$ liefern.
+  \begin{enumerate}
+    \item Jede Schlüsselverteilzentrale liefert einen Teilschlüssel der Länge $\ell/3$. Die drei Teile werden zu $k$ konkateniert.
+    \item Jede Schlüsselverteilzentrale liefert einen Teilschlüssel der Länge $\ell$. Die drei Teile werden bitweise XOR-verknüpft.
+  \end{enumerate}
+  Welche Variante ist sicherer, wenn die Teilschlüssel von X und Y kompromittiert sind? Begründen Sie Ihre Antwort.
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Zertifizierungsmodelle}
+  Bei der Verschlüsselung und Signierung mithilfe asymmetrischer Kryptosysteme existieren zwei grundsätzliche Modelle für die Schlüsselzertifizierung: dezentral (Web of Trust) und hierarchisch-zentral.
+  \begin{enumerate}
+    \item Was ist der Zweck der Schlüsselzertifizierung?
+    \item\label{itm:zertvornach} Vergleichen Sie die Vor- und Nachteile bei der hierarchisch-zentralen Zertifizierung und der dezentralen Zertifizierung.
+    \item Beschreiben Sie das Konzept der Cross Certification. Welche Vorteile ergeben sich hieraus? Bei welchem der beiden Zertifizierungsmodelle kann es Anwendung finden?
+    \item Inwieweit schützen die beiden in Teilaufgabe \ref{itm:zertvornach} genannten Zertifizierungsmodelle jeweils vor einem Man-in-the-Middle-Angriff (auch Maskerade-Angriff genannt)?
+  \end{enumerate}
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Vertrauensbeziehungen in Public-Key-Infrastrukturen}
+  \begin{enumerate}
+    \item Gegeben sei folgende Zertifizierungsrelation.
+      \input{res/zertifikate-1.fig}
+      D will an A eine verschlüsselte Nachricht senden und ist sich jedoch lediglich sicher, dass der öffentliche Schlüssel von B authentisch ist.
+Kann D der Authentizität des öffentlichen Schlüssels von A vertrauen?
+Falls nein, was müsste zusätzlich gegeben sein?
+    \item Was ändert sich, wenn folgende Zertifizierungsrelation vorliegt?
+    \input{res/zertifikate-2.fig}
+  \end{enumerate}
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Gültigkeitsmodelle für Zertifikate und Signaturen}
+  \begin{enumerate}
+    \item Vergleichen Sie das Schalen- und das Kettenmodell im Hinblick auf die Gültigkeit von Zertifikaten und Signaturen.
+    \item Nennen Sie einen Grund dafür, dass in der Praxis zumeist das Kettenmodell verwendet wird.
+    \item Alice erwirbt ein Zertifikat von CA2. CA2 wurde von CA1 zertifiziert. Die Gültigkeitsdauern der Wurzel- und CA-Zertifikate sowie des Zertifikats von Alice sind in der Abbildung dargestellt.
+      \input{res/zertifikate-3.fig}
+      Zeichnen Sie jeweils den Zeitraum ein, in dem Alice dieses Zertifikat für die Erzeugung gültiger Signaturen nach dem Schalen- bzw. dem Kettenmodell verwenden kann.
+    \item Alice signiert zum Zeitpunkt $t_1$ eine Nachricht und sendet diese an Bob.
+Dieser erhält aufgrund eines Urlaubs in einer Region ohne ausreichende Internetversorgung die Nachricht erst zum Zeitpunkt $t_2$ und überprüft nun die Gültigkeit der Signatur.
+Prüfen Sie jeweils die Gültigkeit der Signatur nach dem Schalen- und dem Kettenmodell.
+    \item Da Alice sich wundert, dass Bob ihr nicht antwortet, sendet sie zum Zeitpunkt $t_2$ erneut eine mit ihrem Zertifikat signierte Nachricht an Bob.
+Durch ein technisches Problem erhält Bob diese aber erst zum Zeitpunkt $t_3$ und versucht nun, die Gültigkeit der Signatur zu verifizieren.
+Prüfen Sie jeweils die Gültigkeit der Signatur nach dem Schalen- und dem Kettenmodell.
+  \end{enumerate}
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\setcounter{subsection}{80}
+\begin{exercise}{Verbindungsverschlüsselung in GSM \ref{ex:gsm}}\end{exercise}
+
+\begin{exercise}{Datenbanken in GSM}
+  \begin{enumerate}
+    \item Welche Datenbanken (Register) gibt es im GSM?
+    \item Welchen Zweck erfüllt jede einzelne Datenbank?
+    \item Von welchen Parteien werden diese Datenbanken jeweils verwaltet?
+  \end{enumerate}
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Mobilfunk: GSM}
+  \begin{enumerate}
+    \item Welche Komponenten kennen das Geheimnis $K_i$ bei GSM?
+    \item Jeder GSM-Mobilfunkteilnehmer bekommt vom Netz eine sogenannte Temporary Mobile Subscriber Identity (TMSI) zugewiesen. Stellen Sie das zugrundeliegende Angreifermodell auf.
+    \item Bei GSM werden die Verschlüsselungsalgorithmen A5/1 bzw.
+A5/2 verwendet, bei UMTS der Algorithmus Kasumi und bei LTE der Algorithmus AES.
+Warum wurden die Algorithmen im Laufe der Weiterentwicklung der Mobilfunknetze geändert?
+    \item Im Mobilfunkbereich werden ausschließlich symmetrische Verfahren für die Kryptographie verwendet. Nennen Sie zwei Gründe, warum das der Fall ist.
+    \item Der historische Angriff des SIM-Clonings dauert ca.
+8 bis 10 Stunden, obwohl nur 150.000 Berechnungsschritte notwendig sind.
+Beschreiben Sie, was der Angreifer beim SIM-Cloning macht und skizzieren Sie, welche Schwäche er dabei ausnutzt.
+  \end{enumerate}
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Verschlüsselung der ersten Protokollnachrichten}
+  \begin{enumerate}
+    \item Warum sind im GSM und UMTS Nachrichten wie Location Updating Request, CM Service Request, Authentication Request und Authentication Response nicht verschlüsselt?
+    \item Bitte skizzieren Sie einen Lösungsvorschlag, der die Verschlüsselung aller o. g. Nachrichten ermöglichen würde.
+  \end{enumerate}
+
+   \begin{solution}
+     % TODO
+   \end{solution}
+ \end{exercise}
 
@@
 \setcounter{subsection}{228}
-\begin{exercise}{Verbindungsverschlüsselung in GSM}
+\begin{exercise}{Verbindungsverschlüsselung in GSM}\label{ex:gsm}
   Welche Kommunikationsabschnitte sind in GSM verschlüsselt? Welche Kommunikationsabschnitte sind unverschlüsselt?

### 2024-12-11 SbD/exercises.tex
Updated 2022 exam.

@@
           \begin{itemize}
-          \item **Technologischer Fortschritt:** Schnellere Computer ermöglichen effizientere Brute-Force-Angriffe.
-          \item **Neue Angriffe:** Fortschritte in der Kryptoanalyse decken Schwachstellen auf, wie z. B. Differential- oder lineare Kryptoanalyse.
-      \end{itemize}
-      \item Das **One-Time-Pad** bietet absolute Sicherheit und bleibt auch nach 98 Jahren sicher. Es wird nicht flächendeckend eingesetzt, da:
-      \begin{itemize}
-          \item Der Schlüssel so lang wie der Klartext sein muss.
-          \item Der sichere Schlüsselaustausch und die Speicherung unpraktisch sind.
-      \end{itemize}
-      \item Die Chiffre, die Buchstaben $ A-N $ mit $ N-Z $ vertauscht, ist nicht sicher:
-      \begin{itemize}
-          \item **Polyalphabetische Verschlüsselung:** Nicht erfüllt, da nur ein Alphabet genutzt wird.
-          \item **Länge bleibt gleich:** Diese Eigenschaft bewahrt Muster im Schlüsseltext, was statistische Angriffe erleichtert.
-      \end{itemize}
+              \item \textbf{Technologischer Fortschritt}: Schnellere Computer ermöglichen effizientere Brute-Force-Angriffe.
+              \item \textbf{Neue Angriffe}: Fortschritte in der Kryptanalyse decken Schwachstellen auf, wie z. B. Differential- oder lineare Kryptanalyse.
+          \end{itemize}
+        \item Das One-Time-Pad bietet absolute Sicherheit und bleibt auch nach 98 Jahren sicher.
+Es wird nicht flächendeckend eingesetzt, da der Schlüssel so lang wie der Klartext sein muss und der sichere Schlüsselaustausch sowie die Speicherung unpraktisch sind.
+        \item Die Chiffre ist nicht sicher.
+Muster bleiben im Schlüsseltext erhalten und können durch Häufigkeitsanalyse gebrochen werden.
+Die gleiche Länge der Nachrichten erleichtert die Kryptanalyse, da die Länge der Schlüsseltexte Rückschlüsse auf die Länge der Klartexte zulässt.
     \end{enumerate}
@@
     \item Berechnen Sie für die folgenden Parameter alle Punkte auf der Kurve: $a = 2$, $b = 3$, $p = 7$.
-    \item Elgamal mit elliptischen Kurven wurde vorgestellt. Zeigen Sie, warum dieses Verfahren mit beliebigen Parametern funktioniert.
+    \item ElGamal mit elliptischen Kurven wurde vorgestellt. Zeigen Sie, warum dieses Verfahren mit beliebigen Parametern funktioniert.
     \item Vergleichen Sie den Elgamal aus der Vorlesung mit dem gegebenen. Auf welchem Prinzip beruht die Sicherheit? Denken Sie, dass er sicherer ist? Warum?
@@
     \begin{enumerate}
-        \item Punkte auf der elliptischen Kurve $ y^2 = x^3 + 2x + 3 \mod 7 $:
-        \[
-        \{(1, 1), (2, 2), (4, 5), (6, 3)\}.
-        \]
+        \item Punkte auf der elliptischen Kurve $y^2\equiv x^3+2x+3 \pmod{7}$:
+          \[ \{(6,1), (6,6)\} \]
         \item Das ElGamal-Verfahren mit elliptischen Kurven funktioniert, weil die Addition von Punkten und die Multiplikation mit einem Skalar die Gruppenstruktur der Kurve erhalten.
 Dadurch bleibt das diskrete Logarithmusproblem schwer lösbar.
-        \item Vergleich:
-        \begin{itemize}
-            \item Beide basieren auf dem diskreten Logarithmusproblem.
-            \item ECC-ElGamal ist effizienter, da kürzere Schlüssel benötigt werden.
-        \end{itemize}
+        \item EC-ElGamal ist nicht unbedingt sicherer, da sowohl das DLP als auch das ECDLP schwer lösbar sind.
+Der Vorteil von EC-ElGamal ist die kürzere Schlüssellänge sowie effizientere Verarbeitung/Speicherung bei gleicher Sicherheit.
     \end{enumerate}
@@
   \begin{enumerate}
-    \item Es wurden 3 verschiedene Möglichkeiten gegeben, einen Schlüssel von verschiedenen Schlüsselpartnern (S, V, C, T) zu kombinieren.
-Alle Teilschlüssel werden aneinander gehängt l/4.
-3 der Teilschlüssel wird aneinander gehängt und der letzte Teilnehmer bestimmt die Reihenfolge.
-Alle Schlüssel werden XOR berechnet.
-Bewerten Sie die Sicherheit der Verfahren.
-    \item Gegeben ist ein Zertifikatsbaum mit 5 Teilnehmern.
-$E \rightarrow D$, $E \rightarrow B$, $E \rightarrow A$, $E \rightarrow C$, $C \rightarrow B$, $A \rightarrow B$, $G \rightarrow \emptyset$.
-Betrachten Sie die Zertifizierungsrelation.
+    \item Gegeben seien drei Möglichkeiten einen Schlüssel von verschiedenen Schlüsselpartnern $(S,V,C,T)$ zu kombinieren. Bewerten Sie die Sicherheit der Verfahren.
+      \begin{itemize}
+        \item Alle Teilschlüssel (l/4) werden aneinandergehängt.
+        \item Drei der Teilschlüssel werden aneinandergehängt und der letzte Teilnehmer bestimmt die Reihenfolge.
+        \item Alle Schlüssel werden XOR-verknüpft.
+      \end{itemize}
+    \item\label{itm:zertifizierungsstellen} Gegeben ist ein Zertifikatsbaum mit 5 Teilnehmern. Betrachten Sie die Zertifizierungsrelation.\\
+      \begin{tikzpicture}[->,  >=stealth]
+        \node (E) at (0, 0) {$E$};
+        \node (A) [below left of=E] {$A$};
+        \draw (E) -- (A);
+        \node (B) [below right of=A] {$B$};
+        \draw (E) -- (B);
+        \draw (A) -- (B);
+        \node (C) [below right of=E] {$C$};
+        \draw (E) -- (C);
+        \draw (C) -- (B);
+        \node (D) [right of=C] {$D$};
+        \draw (E) -- (D);
+        \node (G) [right of=D] {$G$};
+      \end{tikzpicture}\\
       $G$ möchte jetzt zusätzlich eine verschlüsselte Nachricht an $B$ senden.
@@
 welchen Zertifizierungsstellen muss $G$ mindestens vertrauen, um das Zertifikat von $B$ prüfen zu können?
-    \item Wählen Sie Ihre Lösung so, dass die Anzahl an Zertifizierungsstellen, denen $N4$ (explizit) vertrauen muss, minimal ist.
-    \item In der Zertifizierungsrelation aus Teilaufgabe c) wird die Zertifizierungsstelle $C$ kompromittiert und fällt deshalb aus.
-$G$ möchte eine verschlüsselte Nachricht an $C$ senden.
-Welche Maßnahme ermöglicht es $G$, das Zertifikat von $B$ trotz des Ausfalls von $CA4$ zu überprüfen?
-  \end{enumerate}
-
-  \begin{solution}
-    \begin{enumerate}
-        \item Sicherheitsbewertung der Verfahren:
-        \begin{itemize}
-            \item **Alle Teilschlüssel aneinanderhängen:** Angreifbar durch Brute-Force-Angriffe.
-            \item **Drei Teilschlüssel aneinanderhängen:** Sicherheitsgewinn gegenüber der ersten Methode.
-            \item **XOR aller Schlüssel:** Höchste Sicherheit, solange alle Schlüssel zufällig und unabhängig sind.
-        \end{itemize}
-        \item Zertifikatsbaum: N7 vertraut CAs, die N3 und $ B $ direkt zertifizieren.
-        \item Minimale Anzahl an CAs: Es reicht aus, der CA zu vertrauen, die die Zertifikate für $ B $ und $ N3 $ signiert hat.
-        \item Maßnahme bei Kompromittierung: Alternativer Zertifizierungsweg über eine andere CA. Neue Zertifikate können erstellt werden, um Vertrauen zu erhalten.
+Wählen Sie Ihre Lösung so, dass die Anzahl an Zertifizierungsstellen, denen $G$ (explizit) vertrauen muss, minimal ist.
+    \item In der Zertifizierungsrelation aus \ref{itm:zertifizierungsstellen} wird die Zertifizierungsstelle $C$ kompromittiert und fällt deshalb aus.
+$G$ möchte eine verschlüsselte Nachricht an $B$ senden.
+Welche Maßnahme ermöglicht es $G$, das Zertifikat von $B$ trotz des Ausfalls von $C$ zu überprüfen?
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+      \item
+        \begin{itemize}
+          \item Geringe Sicherheit, da die Sicherheit vollständig von den Teilschlüsseln abhängt.
+          \item Höhere Sicherheit, da die zufällige Reihenfolge Angriffe schwieriger macht.
+          \item Höchste Sicherheit, solange die Teilschlüssel unabhängig, gleichmäßig verteilt und zufällig sind.
+        \end{itemize}
+      \item $G$ muss nur der Zertifizierungsstellen $E$ vertrauen, um das Zertifikat von $B$ zu prüfen, da $E$ die Wurzel des Zertifikatsbaums ist und alle anderen Zertifikate signiert.
+      \item $G$ kann das Zertifikat von $B$ trotz des Ausfalls von $C$ prüfen, indem $G$ das Zertifikat von $B$ über $A$ und $E$ prüft.
     \end{enumerate}
@@
   \begin{enumerate}
-    \item Sagen Sie zu folgenden Abkürzungen welche Funktion sie im GSM erfüllen: $Ki$, $IMSI$, $RAND$, $SRES$, $LAI$, $MSISDN$, $Kc$.
+    \item Welche Funktionen bezeichnen diese Abkürzungen in GSM: \texttt{Ki}, \texttt{IMSI}, \texttt{RAND}, \texttt{SRES}, \texttt{LAI}, \texttt{MSISDN}, \texttt{Kc}.
     \item Welche dieser Abkürzungen ist auch noch im UMTS enthalten bzw. mit ähnlicher Funktion aber anderen Namen?
@@
           \begin{itemize}
-            \item $ K_i $: Geheimer Schlüssel zur Authentifizierung und Verschlüsselung.
-            \item $ IMSI $: Einzigartige Kennung des Mobilfunkteilnehmers.
-            \item $ RAND $: Zufällige Herausforderung zur Authentifizierung.
-            \item $ SRES $: Antwort des Mobilgeräts auf $ RAND $, basierend auf $ K_i $.
-            \item $ LAI $: Standortkennung für Mobilitätsmanagement.
-            \item $ MSISDN $: Telefonnummer des Teilnehmers.
-            \item $ K_c $: Schlüssel zur Verschlüsselung des Datenverkehrs.
-        \end{itemize}
-        \item Diese Funktionen sind auch im UMTS enthalten, jedoch oft mit verbesserten Sicherheitsmechanismen (z. B. längere Schlüssel, stärkere Algorithmen).
+            \item \texttt{Ki}: Key Identifier. Geheimer Schlüssel zur Authentifizierung, gespeichert in SIM-Karte und HLR/AuC. Zur Authentifizierung des Mobiltelefons genutzt.
+            \item \texttt{IMSI}: International Mobile Subscriber Identity. Eindeutige Kennung des Mobilfunkteilnehmers. Identifizierung des Benutzers im Netzwerk
+            \item \texttt{RAND}: Random Number. Dient als Herausforderung (Challenge) für die Authentifizierung.
+            \item \texttt{SRES}: Signed Response. Antwort auf die Herausforderung, die vom Mobiltelefon berechnet wird. Wird zur Authentifizierung genutzt.
+            \item \texttt{LAI}: Location Area Identity. Identifiziert die Location Area, in der sich das Mobiltelefon befindet.
+            \item \texttt{MSISDN}: Mobile Station International Subscriber Directory Number. Telefonnummer des Mobiltelefons.
+            \item \texttt{Kc}: Ciphering Key. Wird zur Verschlüsselung der Kommunikation zwischen Mobiltelefon und Basisstation genutzt.
+          \end{itemize}
+        \item Enthalten: \texttt{Ki}, \texttt{IMSI}, \texttt{RAND}, \texttt{MSISDN}.
+Mit anderem Namen/ähnlicher Funktion: \texttt{SRES}: \texttt{XRES}, \texttt{LAI}: \texttt{RAI}, \texttt{Kc}: \texttt{CK}.
     \end{enumerate}

### 2024-12-11 NLP/exercises.tex
Updated some exercises.

--- 2024-10-21 NLP/exercises.tex
@@
 
-\sheet{2023}{1st exam}
+\sheet{Lecture questions}
+
+\begin{exercise}{Web as an application area for NLP}
+  The web is an application area for NLP. Provide examples of applications.
+
+  \begin{solution}
+    Examples include:
+    \begin{itemize}
+        \item Internet of Services
+        \item Community mining
+        \item Information retrieval
+        \item Machine translation for web-based content
+    \end{itemize}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Web as a resource for NLP}
+  How can the web improve the quality of NLP? Provide examples.
+
+  \begin{solution}
+    Examples include:
+    \begin{itemize}
+        \item Web as Corpus
+        \item Analyzing web-based knowledge repositories
+        \item Extracting structured information from unstructured web data
+        \item Using web-scale data for training language models
+    \end{itemize}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Morphological analysis}
+  Identify all stems and affixes (prefix, suffix, infix, circumfix) in the following words: "index," "incorrect," "interesting."
+
+  \begin{solution}
+    Morphological breakdown:
+    \begin{itemize}
+        \item \textbf{index}: stem = "index"
+        \item \textbf{incorrect}: prefix = "in," stem = "correct"
+        \item \textbf{interesting}: stem = "interest," suffix = "ing"
+    \end{itemize}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Stemming vs lemmatization}
+  In contrast to lemmatization, stemming does not necessarily return a valid word form. Why is stemming still useful?
+
+  \begin{solution}
+    Stemming is still useful because:
+    \begin{itemize}
+        \item It is faster and computationally less expensive.
+        \item It simplifies implementation for applications like Information Retrieval.
+        \item It can still group similar words under a common base, which is often sufficient.
+    \end{itemize}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Corpus definition and usage}
+  What is a corpus, and how can it be built and used for linguistic analysis?
+
+  \begin{solution}
+    \begin{itemize}
+        \item \textbf{Corpus definition}: A corpus is a large and structured set of texts, often used for linguistic research.
+        \item \textbf{How to build}: Collect text data, preprocess it (cleaning, normalization), and annotate it for specific tasks.
+        \item \textbf{Usage}: Analyze linguistic patterns, train NLP models, and evaluate machine learning algorithms.
+    \end{itemize}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{NLP datasets}
+  What are NLP datasets, and why are they important?
+
+  \begin{solution}
+    \begin{itemize}
+        \item \textbf{Definition}: NLP datasets are collections of annotated or unannotated text data designed for specific NLP tasks (e.g., sentiment analysis, translation).
+        \item \textbf{Importance}:
+        \begin{itemize}
+            \item Serve as a foundation for training and testing machine learning models.
+            \item Enable reproducibility in research.
+            \item Facilitate comparisons between models and techniques.
+        \end{itemize}
+    \end{itemize}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Corpus and dataset annotation}
+  What is the annotation process for a corpus or dataset, and how does it impact machine learning applications?
+
+  \begin{solution}
+    \begin{itemize}
+        \item \textbf{Annotation process}: Adding labels or metadata to text data (e.g., part-of-speech tags, sentiment labels, named entities).
+        \item \textbf{Impact on machine learning}:
+        \begin{itemize}
+            \item Improves model performance by providing high-quality training data.
+            \item Enables supervised learning by associating input with output labels.
+            \item Impacts scalability and generalization of models depending on annotation quality.
+        \end{itemize}
+    \end{itemize}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Large-scale dataset collection using crowdsourcing}
+  What are the main strategies for large-scale dataset collection using crowdsourcing?
+
+  \begin{solution}
+    Strategies include:
+    \begin{itemize}
+        \item \textbf{Task design}: Clearly define annotation tasks with examples and guidelines.
+        \item \textbf{Platform selection}: Use platforms like Amazon Mechanical Turk or Prolific.
+        \item \textbf{Quality control}: Use redundancy (multiple annotators per task) and include test questions to ensure reliability.
+        \item \textbf{Data aggregation}: Combine annotations using majority voting or statistical models.
+    \end{itemize}
+  \end{solution}
+\end{exercise}
+
+% Q: Why are more features not always better for learning a
+% classifier?
+% A: More training data might be needed; Dependent features
+% might lead to deficient models; Some ML algorithms can only deal
+% with specific data types; Slower.
+% Q: What is the typical type of input data that a sequence tagger
+% training step requires?
+% A: Labeled training text. (e.g. POS, NER)
+
+\setcounter{section}{2022}
+\sheet{1st exam}
+
 \begin{exercise}{Tokenization}
@@
   \begin{solution}
-    % TODO
+    \begin{enumerate}
+        \item \textbf{Hyphenated words}: Does "state-of-the-art" count as one token or three?
+        \item \textbf{Contractions}: Should "don't" be treated as one token or split into "do" and "n't"?
+        \item \textbf{Punctuation}: How should punctuation like periods in "U.S." or apostrophes in possessives ("John's book") be handled?
+    \end{enumerate}
   \end{solution}
@@
 \begin{exercise}{Parse trees}
-  Given the sentence "John shot the man with the gun" and a grammar: Draw all existing parse trees. Which types of syntactic ambiguity occur here and where?
+  "\textit{John shot the man with the gun}". Draw all possible parse trees for this sentence using the English grammar. What types of syntactic ambiguity are present?
 
   \begin{solution}
-    % TODO
+    A parse tree can be drawn using the following structures: $S, N, V, Det, P, NP, VP, PP$.
+This is an example of attachment ambiguity.
+The prepositional phrase "with the gun" can attach to either "shot" or "the man".
   \end{solution}
@@
 \begin{exercise}{Stemming and lemmatization}
-  Difference between stemming and lemmatization.
+  What is the difference between stemming and lemmatization?
 
   \begin{solution}
-    % TODO
+    While stemming reduces a word to its base form by removing pre- and suffices ("running" → "run"), lemmatization considers the word's context and grammar to find its dictionary form ("better" → "good").
+Lemmatization is more precise than stemming but computationally more expensive.
   \end{solution}
@@
 \begin{exercise}{Web as corpus}
-  List 4 issues you have to deal with when using the web as a corpus for NLP purposes. Explain each.
+  List 4 issues you have to deal with when using the web as a corpus for NLP purposes and explain each.
 
   \begin{solution}
-    % TODO
+    \begin{enumerate}
+        \item \textbf{Noise}: Content may contain errors, irrelevant information, or non-standard grammar.
+        \item \textbf{Duplicates}: Identical or nearly identical data appears multiple times.
+        \item \textbf{Bias}: Web content often reflects societal biases, which can skew analyses.
+        \item \textbf{Reproducability}: Web pages change frequently, making reproducibility challenging.
+    \end{enumerate}
   \end{solution}
@@
 \begin{exercise}{Cross-language QA}
-  How does cross-language QA work? Issues?
+  How does cross-language QA work and what are the main issues?
 
   \begin{solution}
-    % TODO
+    Cross-language QA involves answering questions in one language based on information in another.
+This is accomplished by translating the question, retrieving relevant information in the target language, and generating an answer.
+Challenges include translation errors, cultural differences.
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    Data Preprocessing $\to$ Feature extraction $\to$ Model training $\to$ Evaluation $\to$ Deployment
   \end{solution}
@@
 \begin{exercise}{IR evaluation}
-  What are the problems with precision, recall, and f-measure in IR? How to improve and why is as an example cohen kappa better?
+  What are the problems with precision, recall, and F-measure in IR? How can these be improved? Why is Cohen's Kappa better?
 
   \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item \textbf{Precision}: How relevant are the retrieved documents? Favors retrieving fewer documents.
+        \item \textbf{Recall}: How many of the relevant documents were retrieved? Favors retrieving more documents.
+        \item \textbf{F-measure}: Harmonic mean of precision and recall. Balances the two metrics but not adjustable.
+        \item \textbf{Cohen's Kappa}: Measures inter-annotator agreement, accounting for chance agreement. Better for evaluating IR systems because it considers agreement beyond chance.
+    \end{itemize}
   \end{solution}
@@
 \begin{exercise}{IR evaluation}
-  Calculate P@5, P@10, MAP for the given IR ranked result list (one for IR system A and one for B. Both go up to ten entries and either have a checkmark or an x in each rank).
+  Calculate P@5, P@10, and MAP for the given IR ranked result list. Analyze both IR system A and system B.
 
   \begin{solution}
-    % TODO
+    For P@5, P@10, and MAP:
+    \begin{itemize}
+        \item \textbf{P@5}: Count relevant documents in the top 5 results and divide by 5.
+        \item \textbf{P@10}: Same calculation for the top 10 results.
+        \item \textbf{MAP}: Average Precision for each query, then average across all queries.
+    \end{itemize}
+    Steps:
+    \begin{itemize}
+        \item Identify positions of relevant documents in ranked lists.
+        \item Calculate precision at each relevant position.
+        \item Take mean of these precision values.
+    \end{itemize}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    Correct statements:
+    \begin{enumerate}
+        \item \textbf{True}: Features for Naive Bayes do not need to be uncorrelated, though it assumes independence for computation.
+        \item \textbf{False}: Naive Bayes is fast to train, not slow.
+        \item \textbf{False}: SVM is not inherently multi-label but can be adapted for multi-label classification.
+        \item \textbf{True}: SVM handles sparse data well due to its kernel functions.
+    \end{enumerate}
   \end{solution}
@@
 \begin{exercise}{Corpus construction}
-  How to construct a corpus for linguistic purposes from web data and how to make it usable?
+  How can a corpus be constructed for linguistic purposes from web data, and how can it be made usable?
 
   \begin{solution}
-    % TODO
+    Steps for constructing a corpus:
+    \begin{enumerate}
+        \item \textbf{Data collection}: Crawl web pages, scrape data, or gather domain-specific text.
+        \item \textbf{Cleaning}: Remove duplicates, irrelevant content, and normalize text.
+        \item \textbf{Annotation}: Add metadata or labels for tasks like classification or parsing.
+        \item \textbf{Usability}: Store in accessible formats (e.g., JSON, XML) and document the corpus for ease of use.
+    \end{enumerate}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    Challenges in pragmatics and discourse analysis:
+    \begin{itemize}
+        \item \textbf{Context dependency}: Understanding depends heavily on prior sentences or external knowledge.
+        \item \textbf{Ambiguity}: Words and phrases can have multiple meanings depending on the situation.
+        \item \textbf{Social norms}: Interpretations vary based on cultural or societal factors.
+    \end{itemize}
   \end{solution}
@@
 \begin{exercise}{Hate speech}
-  Define hate speech and how to construct a hate speech classifier
+  Define hate speech and explain how to construct a hate speech classifier.
 
   \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item \textbf{Definition}: Hate speech involves language that attacks or discriminates against individuals or groups based on attributes like race, religion, ethnicity, gender, or sexual orientation.
+        \item \textbf{Constructing a hate speech classifier}:
+        \begin{enumerate}
+            \item Collect labeled datasets with examples of hate speech and non-hate speech.
+            \item Preprocess text (e.g., tokenization, removing stop words).
+            \item Extract features (e.g., bag-of-words, TF-IDF, embeddings).
+            \item Train a supervised machine learning model (e.g., Naive Bayes, SVM, or deep learning models like LSTMs or transformers).
+            \item Evaluate the classifier using metrics like precision, recall, and F1-score.
+        \end{enumerate}
+    \end{itemize}
   \end{solution}
@@
 \begin{exercise}{Counter hate speech}
-  What is counter hate speech and how to construct such a system
+  What is counter hate speech, and how can a system be constructed for it?
 
   \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item \textbf{Definition}: Counter hate speech is content designed to reduce the impact of hate speech, often by promoting positive, empathetic, or factual responses.
+        \item \textbf{Constructing a counter hate speech system}:
+        \begin{enumerate}
+            \item Collect datasets of effective counter hate responses.
+            \item Train models to generate or retrieve appropriate responses based on the hate speech content.
+            \item Use sentiment analysis and empathy detection to craft responses.
+            \item Deploy the system in platforms like social media or forums, ensuring ethical considerations are addressed.
+        \end{enumerate}
+    \end{itemize}
   \end{solution}
@@
 \begin{exercise}{Crowdsourced annotation}
-  Crowdsourced annotation workflow
+  What is the workflow for crowdsourced annotation?
 
   \begin{solution}
-    % TODO
+    Steps in a crowdsourced annotation workflow:
+    \begin{enumerate}
+        \item \textbf{Task design}: Define the annotation task clearly with examples and instructions.
+        \item \textbf{Platform setup}: Use a crowdsourcing platform (e.g., Amazon Mechanical Turk, Prolific).
+        \item \textbf{Quality control}: Include test questions and redundancy (multiple annotators per item) to ensure data reliability.
+        \item \textbf{Data aggregation}: Aggregate annotations using majority voting or weighted schemes.
+        \item \textbf{Evaluation}: Analyze annotation quality and address inconsistencies.
+    \end{enumerate}
   \end{solution}
@@
 \begin{exercise}{Web corpus}
-  Why are duplicates on the web and how to detect
+  Why do duplicates exist on the web, and how can they be detected?
 
   \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item \textbf{Duplicates}: Duplicates arise from mirrored websites, copied content, or redundant crawls.
+        \item \textbf{Detection methods}:
+        \begin{itemize}
+            \item Hashing (e.g., MD5 or SHA): Compare hash values of documents.
+            \item Similarity measures: Use cosine similarity or Jaccard index on n-grams.
+            \item Structural analysis: Detect similarity in HTML structure.
+        \end{itemize}
+    \end{itemize}
   \end{solution}
@@
 \begin{exercise}{QA vs IR}
-  Differences in QA vs IR
+  What are the differences between Question Answering (QA) and Information Retrieval (IR)?
 
   \begin{solution}
-    % TODO
+    Differences between QA (Question Answering) and IR (Information Retrieval):
+    \begin{itemize}
+        \item \textbf{Goal}:
+        \begin{itemize}
+            \item QA aims to provide direct answers.
+            \item IR retrieves documents containing relevant information.
+        \end{itemize}
+        \item \textbf{Processing}:
+        \begin{itemize}
+            \item QA requires deeper natural language understanding.
+            \item IR focuses on keyword matching and ranking.
+        \end{itemize}
+        \item \textbf{Applications}:
+        \begin{itemize}
+            \item QA: Chatbots, customer support.
+            \item IR: Search engines.
+        \end{itemize}
+    \end{itemize}
   \end{solution}
@@
 \begin{exercise}{Cross-language QA}
-  Explain applications of multi-language QA and difficulties?
+  What are the applications of multi-language QA, and what difficulties arise?
 
   \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item \textbf{Applications}: Multi-language QA systems are used in global customer support, cross-border e-commerce, and multi-language search engines.
+        \item \textbf{Difficulties}:
+        \begin{itemize}
+            \item Translation errors can distort queries or results.
+            \item Cultural and linguistic differences complicate understanding and generating answers.
+            \item Lack of high-quality bilingual corpora for training.
+        \end{itemize}
+    \end{itemize}
   \end{solution}
@@
 \begin{exercise}{Language detection}
-  Explain language detection
+  How does language detection work?
 
   \begin{solution}
-    % TODO
+    Approaches for language detection:
+    \begin{enumerate}
+        \item \textbf{N-gram analysis}: Compare n-grams of text against language-specific frequency profiles.
+        \item \textbf{Machine learning models}: Use classifiers trained on labeled text for different languages.
+        \item \textbf{Lexicon-based methods}: Match words in the text to dictionaries of different languages.
+        \item \textbf{Hybrid approaches}: Combine multiple methods for higher accuracy.
+    \end{enumerate}
   \end{solution}
@@
 \begin{exercise}{Factual QA vs opinionated QA}
-  Differences between factual QA and opinionated QA, and which is easier to evaluate?
+  What are the differences between factual QA and opinionated QA, and which is easier to evaluate?
 
   \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item \textbf{Factual QA}: Deals with objective information (e.g., "What is the capital of France?").
+        \item \textbf{Opinionated QA}: Involves subjective answers (e.g., "Is Paris a good travel destination?").
+        \item \textbf{Evaluation}: Factual QA is easier to evaluate as answers can be verified against ground truth. Opinionated QA requires subjective or aggregated assessments.
+    \end{itemize}
   \end{solution}
@@
 \begin{exercise}{RAG and LangChain}
-  What's RAG and how does LangChain help?
+  What is Retrieval-Augmented Generation (RAG), and how does LangChain assist in implementing it?
 
   \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item \textbf{RAG (Retrieval-Augmented Generation)}: Combines retrieval models with generative models to produce context-aware answers.
+External knowledge is retrieved and fed into the generation model.
+        \item \textbf{LangChain}: A framework for building applications with large language models, offering tools for integrating RAG systems.
+It manages chaining retrieval and generation steps efficiently.
+    \end{itemize}
   \end{solution}
@@
 \begin{exercise}{Sentiment analysis}
-  Discuss: Sentiment analysis using supervised vs. lexicon based approaches.
+  Compare sentiment analysis using supervised approaches and lexicon-based approaches.
 
   \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item \textbf{Supervised learning}: Uses labeled data to train machine learning models.
+        \begin{itemize}
+            \item \textbf{Advantages}: High accuracy with large datasets.
+            \item \textbf{Drawbacks}: Relies on annotated data and is domain-dependent.
+        \end{itemize}
+        \item \textbf{Lexicon-based approaches}: Use sentiment dictionaries to determine polarity.
+        \begin{itemize}
+            \item \textbf{Advantages}: Simplicity and no training requirement.
+            \item \textbf{Drawbacks}: Limited contextual understanding and poor performance on complex text.
+        \end{itemize}
+    \end{itemize}
   \end{solution}
@@
 \begin{exercise}{Edit distance}
-  Calculate the Minimum Edit Distance for "TAC" vs. "CAT" (given is also a table just like on the slides, just without the numbers filled in). List all possible word outcomes (in this case 4).
+  Calculate the Minimum Edit Distance for "TAC" vs. "CAT" and list all possible word outcomes.
 
   \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item \textbf{Minimum Edit Distance for "TAC" and "CAT"}:
+        \[
+        \begin{array}{c|c|c|c|c}
+            &  & C & A & T \\
+            \hline
+            & 0 & 1 & 2 & 3 \\
+            T & 1 & 1 & 2 & 2 \\
+            A & 2 & 2 & 1 & 2 \\
+            C & 3 & 2 & 2 & 2 \\
+        \end{array}
+        \]
+        \item \textbf{Word outcomes}: Possible outcomes from edits:
+        \begin{enumerate}
+            \item CAT
+            \item TAT
+            \item TAC
+            \item CAC
+        \end{enumerate}
+    \end{itemize}
   \end{solution}
@@
 
-\sheet{2022}{1st exam}
+\setcounter{section}{2021}
+\sheet{1st exam}
 
 \begin{exercise}{Corpus construction}
-  How would you construct a corpus for linguistic purposes and how to make it usable?
+  How would you construct a corpus for linguistic purposes and how can it be made usable?
 
   \begin{solution}
-    % TODO
+    \begin{enumerate}
+        \item \textbf{Data collection}: Gather data from relevant sources (e.g., web crawling, domain-specific text).
+        \item \textbf{Cleaning and preprocessing}: Remove duplicates, non-linguistic content, and normalize text (e.g., case normalization, tokenization).
+        \item \textbf{Annotation}: Add metadata or linguistic labels (e.g., POS tags, syntactic trees) to the text.
+        \item \textbf{Storage}: Store the corpus in an accessible format like JSON, XML, or relational databases.
+        \item \textbf{Documentation}: Provide guidelines and metadata about the corpus for user understanding.
+    \end{enumerate}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item \textbf{Context dependency}: Requires understanding of the broader context beyond sentence boundaries.
+        \item \textbf{Ambiguity}: Words or phrases can have multiple meanings depending on the discourse.
+        \item \textbf{Cultural and social norms}: Interpretation is influenced by societal and cultural factors.
+        \item \textbf{Incomplete information}: Missing parts of conversations or text make it harder to analyze.
+    \end{itemize}
   \end{solution}
@@
 \begin{exercise}{Sequence tagging and CRF}
-  Define sequence tagging and what is CRF
+  Define sequence tagging and explain what a Conditional Random Field (CRF) is.
 
   \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item \textbf{Sequence tagging}: The process of assigning labels to each element in a sequence (e.g., part-of-speech tagging, named entity recognition).
+        \item \textbf{Conditional Random Field (CRF)}: A probabilistic model used for structured prediction in sequence data. It considers the context of neighboring labels to improve accuracy.
+    \end{itemize}
   \end{solution}
@@
 \begin{exercise}{Minimum edit distance}
-  Calculate the Minimum Edit Distance for "TAC" vs. "CAT" (given is also a table just like on the slides, just without the numbers filled in). List all possible word outcomes (in this case 4).
+  Calculate the Minimum Edit Distance for "TAC" vs. "CAT" (a table is given). List all possible outcomes.
 
   \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item \textbf{Minimum Edit Distance Table}:
+        \[
+        \begin{array}{c|c|c|c|c}
+            &  & C & A & T \\
+            \hline
+            & 0 & 1 & 2 & 3 \\
+            T & 1 & 1 & 2 & 2 \\
+            A & 2 & 2 & 1 & 2 \\
+            C & 3 & 2 & 2 & 2 \\
+        \end{array}
+        \]
+        \item \textbf{Outcomes}:
+        \begin{enumerate}
+            \item CAT
+            \item TAT
+            \item TAC
+            \item CAC
+        \end{enumerate}
+    \end{itemize}
   \end{solution}
@@
 \begin{exercise}{IR evaluation}
-  Calculate P@5 and MAP for the given IR ranked result list (one for IR system A and one for B. Both go up to ten entries and either have a checkmark or an x in each rank).
+  Calculate P@5 and MAP for given ranked IR results for two systems.
 
   \begin{solution}
-    % TODO
-  \end{solution}
-\end{exercise}
-
-\begin{exercise}{Sequence tagging and CRF}
-  Define sequence tagging and what is CRF
-
-  \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item \textbf{P@5}: Count relevant documents in the top 5 results and divide by 5.
+        \item \textbf{MAP}: For each query, calculate average precision across relevant documents and average these scores.
+    \end{itemize}
   \end{solution}
@@
 \begin{exercise}{Hate speech}
-  Define hate speech and how to construct a hate speech classifier
+  Define hate speech and how to construct a hate speech classifier.
 
   \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item \textbf{Definition}: Hate speech targets individuals or groups based on attributes like race, religion, gender, or orientation.
+        \item \textbf{Classifier construction}:
+        \begin{enumerate}
+            \item Collect labeled datasets of hate and non-hate speech.
+            \item Preprocess data by removing noise and tokenizing.
+            \item Train a machine learning model (e.g., Naive Bayes, SVM, transformers).
+            \item Evaluate using metrics like F1-score, precision, and recall.
+        \end{enumerate}
+    \end{itemize}
   \end{solution}
@@
 \begin{exercise}{Counter hate speech}
-  What is counter hate speech and how to construct such a system
+  What is counter hate speech, and how can a system be constructed for it?
 
   \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item \textbf{Definition}: Counter hate speech neutralizes hate content by promoting positive or factual responses.
+        \item \textbf{System construction}:
+        \begin{enumerate}
+            \item Collect examples of effective counter hate responses.
+            \item Use NLP techniques to classify hate speech and generate counter-responses.
+            \item Train models on empathy and sentiment analysis for response generation.
+        \end{enumerate}
+    \end{itemize}
   \end{solution}
@@
 
+\begin{exercise}{Sentiment lexicon extension}
+  How can sentiment analysis be improved using lexicon extensions from distributed thesauruses?
+
+  \begin{solution}
+    \begin{itemize}
+        \item \textbf{Lexicon extension}: Expand sentiment dictionaries using related terms from distributed word embeddings like Word2Vec or GloVe.
+        \item \textbf{Process}:
+        \begin{enumerate}
+            \item Identify synonyms, antonyms, or similar words in embedding spaces.
+            \item Update lexicons with these terms to improve coverage.
+            \item Validate new terms to ensure sentiment accuracy.
+        \end{enumerate}
+    \end{itemize}
+  \end{solution}
+\end{exercise}
 \begin{exercise}{Sentiment analysis}
@@
   \begin{solution}
-    % TODO
-  \end{solution}
-\end{exercise}
-
-\begin{exercise}{Sentiment lexicon extension}
-  How could you analyse sentiment with lexicon extension using distributed thesauruses.
-
-  \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item \textbf{Lexicon-based approaches}: 
+        \begin{itemize}
+            \item Use pre-built dictionaries of words with sentiment labels (e.g., positive, negative).
+            \item Advantages:
+                \begin{itemize}
+                    \item Easy to implement and interpret.
+                    \item No need for labeled training data.
+                \end{itemize}
+            \item Disadvantages:
+                \begin{itemize}
+                    \item Limited to the coverage of the lexicon.
+                    \item Poor performance on context-dependent sentiment (e.g., sarcasm).
+                \end{itemize}
+        \end{itemize}
+        \item \textbf{Supervised learning approaches}:
+        \begin{itemize}
+            \item Train machine learning models on labeled data to classify sentiment.
+            \item Advantages:
+                \begin{itemize}
+                    \item Can learn from context and handle subtle sentiment expressions.
+                    \item Scalable to large datasets.
+                \end{itemize}
+            \item Disadvantages:
+                \begin{itemize}
+                    \item Requires annotated data, which can be costly to obtain.
+                    \item More computationally intensive than lexicon-based approaches.
+                \end{itemize}
+        \end{itemize}
+    \end{itemize}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item \textbf{Context-dependence}: Lexicons fail to account for how sentiment changes with context.
+        \item \textbf{Negation handling}: Sentences like "not bad" may be incorrectly classified without understanding negation.
+        \item \textbf{Ambiguity}: Words can have different sentiments in different contexts (e.g., "hot" in weather vs. arguments).
+        \item \textbf{Domain-specific issues}: Lexicons often do not cover domain-specific terms or slang.
+    \end{itemize}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item \textbf{Correct}: 
+        \begin{itemize}
+            \item Features for Naive Bayes classifiers do not need to be uncorrelated.
+            \item SVM handles sparse data well.
+        \end{itemize}
+        \item \textbf{Incorrect}:
+        \begin{itemize}
+            \item Naive Bayes is actually fast to train.
+            \item SVM is not inherently multi-label but can be adapted for multi-label classification.
+        \end{itemize}
+    \end{itemize}
   \end{solution}
@@
 \begin{exercise}{Language detection}
-  Explain approaches for language detection
+  Explain approaches for language detection.
 
   \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item \textbf{N-gram analysis}: Compare the frequency of character n-grams against language-specific profiles.
+        \item \textbf{Machine learning models}: Train classifiers (e.g., SVM, neural networks) using labeled text data for various languages.
+        \item \textbf{Lexicon-based methods}: Use dictionaries to match input text to language-specific words.
+        \item \textbf{Hybrid approaches}: Combine n-gram, lexicon, and machine learning techniques for better accuracy.
+    \end{itemize}
   \end{solution}
@@
 \begin{exercise}{Crowdsourced annotation}
-  Explain the crowdsourced annotation workflow
+  Explain the crowdsourced annotation workflow.
 
   \begin{solution}
-    % TODO
+    Steps in a crowdsourced annotation workflow:
+    \begin{enumerate}
+        \item \textbf{Task design}: Define the annotation task with clear instructions and examples.
+        \item \textbf{Platform setup}: Deploy the task on crowdsourcing platforms like Amazon Mechanical Turk.
+        \item \textbf{Quality control}: Use redundancy by assigning multiple annotators to the same task and include test questions.
+        \item \textbf{Data aggregation}: Combine annotations using majority voting or weighted scoring methods.
+        \item \textbf{Evaluation}: Analyze the consistency and reliability of annotations.
+    \end{enumerate}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    Ambiguities in tokenization include:
+    \begin{itemize}
+        \item \textbf{Hyphenation}: Is "state-of-the-art" one token or three?
+        \item \textbf{Punctuation}: Should periods in abbreviations (e.g., "U.S.") be separated?
+        \item \textbf{Contractions}: Should "don't" be split into "do" and "n't"?
+        \item \textbf{Language differences}: Tokenization rules vary by language (e.g., Chinese characters vs. English words).
+    \end{itemize}
   \end{solution}
@@
 \begin{exercise}{QA vs IR}
-  Differences between QA and IR
+  What are the differences between QA and IR?
 
   \begin{solution}
-    % TODO
+    Differences between QA and IR:
+    \begin{itemize}
+        \item \textbf{Goal}:
+        \begin{itemize}
+            \item QA provides direct answers to user queries.
+            \item IR retrieves documents that may contain relevant information.
+        \end{itemize}
+        \item \textbf{Processing}:
+        \begin{itemize}
+            \item QA requires natural language understanding and reasoning.
+            \item IR focuses on keyword matching and ranking algorithms.
+        \end{itemize}
+        \item \textbf{Applications}:
+        \begin{itemize}
+            \item QA is used in chatbots and virtual assistants.
+            \item IR powers search engines like Google.
+        \end{itemize}
+    \end{itemize}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    Challenges in cross-language QA include:
+    \begin{itemize}
+        \item \textbf{Translation errors}: Inaccurate translations can distort queries or answers.
+        \item \textbf{Lack of parallel resources}: Few high-quality bilingual corpora exist for certain language pairs.
+        \item \textbf{Cultural differences}: Questions or answers may need cultural adaptation.
+        \item \textbf{Mismatched structures}: Linguistic differences make aligning queries and answers difficult.
+    \end{itemize}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item \textbf{Stemming}: Reduces a word to its root by chopping off affixes (e.g., "running" → "run"). It is less precise and may produce non-words.
+        \item \textbf{Lemmatization}: Converts a word to its base form (lemma) based on its context (e.g., "better" → "good"). It is more accurate but computationally expensive.
+    \end{itemize}
   \end{solution}
@@
 \begin{exercise}{Web corpus}
-  Why are duplicates on the web and how to detect
+  Why are duplicates on the web, and how can they be detected?
 
   \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item \textbf{Reasons for duplicates}:
+        \begin{itemize}
+            \item Content mirroring across websites.
+            \item Republishing of syndicated articles.
+            \item Crawling the same site multiple times.
+        \end{itemize}
+        \item \textbf{Detection methods}:
+        \begin{itemize}
+            \item Hashing (e.g., MD5, SHA) to compare file contents.
+            \item Cosine similarity or Jaccard index for text comparisons.
+            \item Comparing HTML structure or metadata.
+        \end{itemize}
+    \end{itemize}
   \end{solution}
@@
 \begin{exercise}{Web corpus}
-  What are the 4 challenges when using the web as a corpus
+  What are the 4 challenges when using the web as a corpus?
 
   \begin{solution}
-    % TODO
+    Challenges include:
+    \begin{itemize}
+        \item \textbf{Noisy data}: Web content often includes irrelevant or ungrammatical text.
+        \item \textbf{Duplication}: Large-scale crawls often capture duplicate data.
+        \item \textbf{Bias}: Web data may reflect societal or geographic biases.
+        \item \textbf{Dynamic content}: Websites frequently change, making reproducibility difficult.
+    \end{itemize}
   \end{solution}
@@
 \begin{exercise}{TREC}
-  What are the TREC question types
+  What are the TREC question types?
 
   \begin{solution}
-    % TODO
+    TREC (Text REtrieval Conference) defines these question types:
+    \begin{itemize}
+        \item \textbf{Factoid questions}: Require specific factual answers (e.g., "Who is the president of the U.S.?").
+        \item \textbf{List questions}: Require a list of items (e.g., "List countries in the EU").
+        \item \textbf{Definition questions}: Seek explanations or definitions (e.g., "What is AI?").
+    \end{itemize}
   \end{solution}
@@
 \begin{exercise}{QALD}
-  What is QALD and what are the challenges
+  What is QALD, and what are the challenges?
 
   \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item \textbf{Definition}: QALD (Question Answering over Linked Data) focuses on answering questions using linked data (e.g., RDF datasets).
+        \item \textbf{Challenges}:
+        \begin{itemize}
+            \item Mapping natural language queries to structured data.
+            \item Handling incomplete or ambiguous data in linked datasets.
+            \item Scaling to large and heterogeneous data sources.
+        \end{itemize}
+    \end{itemize}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    Comprehensiveness refers to the extent to which a system, dataset, or model covers all relevant aspects of a task or domain. In NLP:
+    \begin{itemize}
+        \item \textbf{Datasets}: Comprehensive datasets should include diverse and representative samples of the target domain.
+        \item \textbf{Models}: Comprehensive models should handle all variations and exceptions within their scope.
+        \item \textbf{Applications}: Comprehensive applications should meet all user requirements for a given task.
+    \end{itemize}
   \end{solution}
@@
 
-  % TODO: add image
-
   \begin{solution}
-    % TODO
+    To complete the image, analyze the given context and fill in the white boxes based on:
+    \begin{itemize}
+        \item \textbf{Labels}: Assign appropriate titles or labels to the boxes (e.g., steps in a pipeline).
+        \item \textbf{Relationships}: Connect missing elements to ensure the flow aligns with the diagram's logic.
+        \item \textbf{Annotations}: Use any provided hints to accurately reconstruct the intended content.
+    \end{itemize}
+    The exact answer depends on the image provided.
   \end{solution}
@@
 \begin{exercise}{NLTK}
-  Explain two parameters in NLTK that start with a c.
+  Explain two parameters in NLTK that start with a "c."
 
   \begin{solution}
-    % TODO
+    Two common NLTK parameters starting with "c" are:
+    \begin{itemize}
+        \item \textbf{Chunking}: Groups words into meaningful units (e.g., noun phrases) based on part-of-speech tagging.
+        \item \textbf{Concordance}: Displays occurrences of a word within a context window, useful for lexical analysis.
+    \end{itemize}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    For multi-class classification, SVM uses:
+    \begin{itemize}
+        \item \textbf{One-vs-One (OvO)}: Trains a classifier for every pair of classes. Final classification is determined by majority voting.
+        \item \textbf{One-vs-All (OvA)}: Trains one classifier per class against all other classes. The highest-confidence prediction is chosen.
+        \item \textbf{Extensions}: Kernel tricks and soft margins can be applied for better generalization.
+    \end{itemize}
   \end{solution}
@@
 \begin{exercise}{Feature differences}
-  Explain the differences in features between classical ML and deep network based ML.
+  Explain the differences in features between classical ML and deep network-based ML.
 
   \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item \textbf{Classical ML}:
+        \begin{itemize}
+            \item Relies on manually engineered features (e.g., TF-IDF, n-grams).
+            \item Requires domain expertise for feature extraction.
+        \end{itemize}
+        \item \textbf{Deep learning}:
+        \begin{itemize}
+            \item Automatically learns hierarchical feature representations.
+            \item Captures complex patterns (e.g., semantics, dependencies).
+            \item Requires large datasets for training.
+        \end{itemize}
+    \end{itemize}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    Given true positives (\(TP\)), false positives (\(FP\)), and false negatives (\(FN\)):
+    \begin{itemize}
+        \item \textbf{Precision}: \( \frac{TP}{TP + FP} \)
+        \item \textbf{Recall}: \( \frac{TP}{TP + FN} \)
+        \item \textbf{F1-Score}: \( 2 \times \frac{\text{Precision} \times \text{Recall}}{\text{Precision} + \text{Recall}} \)
+    \end{itemize}
+    Plug in the values provided to compute these metrics.
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    Classification is used in IR for:
+    \begin{itemize}
+        \item \textbf{Query classification}: Identify user intent (e.g., navigational vs. informational).
+        \item \textbf{Document classification}: Categorize documents into predefined topics or genres.
+        \item \textbf{Relevance classification}: Predict whether a document is relevant to a query.
+    \end{itemize}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item \textbf{For}:
+        \begin{itemize}
+            \item Expands dataset availability to resource-scarce languages.
+            \item Enables cross-lingual NLP research.
+            \item Saves effort in recreating annotations.
+        \end{itemize}
+        \item \textbf{Against}:
+        \begin{itemize}
+            \item Potential loss of annotation quality in translation.
+            \item Context or cultural relevance may not transfer.
+            \item Requires manual review to ensure correctness.
+        \end{itemize}
+    \end{itemize}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item \textbf{For}:
+        \begin{itemize}
+            \item Large-scale, diverse data sources.
+            \item Includes up-to-date content.
+            \item Covers multiple languages and domains.
+        \end{itemize}
+        \item \textbf{Against}:
+        \begin{itemize}
+            \item Noisy and unstructured data.
+            \item Contains duplicates and redundant information.
+            \item May include biased or low-quality content.
+        \end{itemize}
+    \end{itemize}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    Example:
+    \begin{itemize}
+        \item \textbf{Preprocessing}: Tokenization, normalization, and cleaning.
+        \item \textbf{Feature extraction}: TF-IDF, embeddings.
+        \item \textbf{Modeling}: Training, evaluation, and deployment.
+    \end{itemize}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item \textbf{Mean Reciprocal Rank (MRR)}:
+        \[
+        MRR = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{\text{rank}_i}
+        \]
+        Where \(\text{rank}_i\) is the rank of the first correct answer for query \(i\).
+        \item \textbf{Accuracy}: \(\frac{\text{Number of correct answers}}{\text{Total number of queries}}\)
+    \end{itemize}
+    Calculate based on the provided data.
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    Language detection involves identifying the language of a given text. Common approaches:
+    \begin{itemize}
+        \item \textbf{N-gram analysis}: Compare character n-grams to language profiles.
+        \item \textbf{Lexicon-based methods}: Match text to language-specific dictionaries.
+        \item \textbf{Machine learning}: Train classifiers (e.g., SVM, neural networks) on labeled text.
+        \item \textbf{Hybrid models}: Combine multiple techniques for higher accuracy.
+    \end{itemize}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    Plagiarism detection methods:
+    \begin{itemize}
+        \item \textbf{String matching}: Use algorithms to find exact matches between documents.
+        \item \textbf{Semantic similarity}: Apply embeddings to detect paraphrased or rephrased content.
+        \item \textbf{Citation analysis}: Check for missing or incorrect citations.
+        \item \textbf{Fingerprinting}: Generate hashes of text fragments and compare across sources.
+    \end{itemize}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item \textbf{Pros}:
+        \begin{itemize}
+            \item Fast and scalable data annotation.
+            \item Access to diverse annotators.
+            \item Cost-effective compared to hiring experts.
+        \end{itemize}
+        \item \textbf{Cons}:
+        \begin{itemize}
+            \item Quality control is challenging.
+            \item May introduce noise or bias.
+            \item Difficult to ensure annotator expertise.
+        \end{itemize}
+    \end{itemize}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    Recall can be improved by:
+    \begin{itemize}
+        \item \textbf{Expanding queries}: Use synonyms, related terms, or stemming to match more results.
+        \item \textbf{Reducing filters}: Remove restrictive criteria (e.g., date, location).
+        \item \textbf{Boosting document recall}: Index more documents or reduce relevance thresholds.
+    \end{itemize}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    For non-words, spell checkers:
+    \begin{itemize}
+        \item \textbf{Dictionary lookup}: Identify words not in the dictionary.
+        \item \textbf{Edit distance}: Suggest corrections based on similarity (e.g., Levenshtein distance).
+        \item \textbf{Contextual models}: Use surrounding words to predict the intended word.
+    \end{itemize}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    The IR cycle includes:
+    \begin{itemize}
+        \item \textbf{Indexing}: Process and store data for efficient retrieval.
+        \item \textbf{Querying}: Users submit queries to search the index.
+        \item \textbf{Retrieval}: Return ranked results based on relevance scores.
+        \item \textbf{Evaluation}: Assess performance using metrics like precision and recall.
+    \end{itemize}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    The Vector Space Model represents documents and queries as vectors:
+    \begin{itemize}
+        \item \textbf{Terms}: Each dimension corresponds to a term in the vocabulary.
+        \item \textbf{Weights}: TF-IDF scores are used as weights for each term.
+        \item \textbf{Similarity}: Compute cosine similarity to rank documents by relevance to the query.
+    \end{itemize}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item \textbf{Factual sentiment analysis}: Focuses on extracting objective sentiment (e.g., product ratings).
+        \item \textbf{Opinionated analysis}: Identifies subjective opinions (e.g., customer reviews).
+        \item \textbf{Key difference}: Factual analysis is more structured and easier to evaluate, whereas opinionated analysis is subjective and context-dependent.
+    \end{itemize}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    Sentiment analysis applications:
+    \begin{itemize}
+        \item \textbf{Customer feedback}: Analyze reviews for product improvements.
+        \item \textbf{Social media monitoring}: Detect trends and public sentiment.
+        \item \textbf{Political analysis}: Gauge public opinion on policies or events.
+        \item \textbf{Market research}: Understand consumer preferences.
+    \end{itemize}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    The Teacher/Student approach combines:
+    \begin{itemize}
+        \item \textbf{Teacher signals}: Provide detailed instructions and explanations.
+        \item \textbf{Student feedback}: Allows iterative learning and personalization.
+        \item \textbf{Better than audio}: Interaction and adaptivity enhance understanding and retention compared to static audio content.
+    \end{itemize}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    \begin{itemize}
+        \item \textbf{Representation}: Encodes data (e.g., text or images) into feature vectors.
+        \item \textbf{Fusion}: Combines multiple data modalities (e.g., text and images) to improve performance.
+        \item \textbf{Differences}: Representation focuses on individual modalities, whereas fusion integrates multiple modalities into a unified representation.
+    \end{itemize}
   \end{solution}

### 2024-12-12 SbD/exercises.tex
Added more exercises.

@@
   \begin{solution}
-    % TODO
+    \begin{enumerate}
+      \item $k=k_x \Vert k_y \Vert k_z$. Wenn $k_x$ und $k_y$ kompromittiert sind, muss nur noch $k_z$ geraten werden. Die verbleibende Entropie ist stark verringert.
+      \item $k=k_x \oplus k_y \oplus k_z$. Wenn $k_x$ und $k_y$ kompromittiert sind, bleibt $k_z$ sicher. Die verbleibende Entropie ist unverändert.
+    \end{enumerate}
+    Die zweite Variante ist sicherer, aufgrund der längeren Teilschlüssel. Konkatenation ist allerdings nicht grundsätzlich sicherer als XOR-Verknüpfung.
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    \begin{enumerate}
+      \item Schlüsselzertifizierung dient der Authentifizierung von Schlüsseln und der Sicherstellung ihrer Integrität.
+      \item Skalierbarkeit und Vertrauen bei zentraler Zertifizierung, aber Single Point of Failure.
+Dezentrale Zertifizierung bietet mehr Flexibilität und Unabhängigkeit, aber schwierige Vertrauensbildung.
+      \item Cross Certification ermöglicht die Vertrauensbildung zwischen zwei Zertifizierungsstellen.
+Hierbei werden die Zertifikate der jeweils anderen Stelle gegenseitig akzeptiert.
+Anwendung findet es in hierarchisch-zentralen Modellen.
+      \item Hierarchisch-zentrale Modelle schützen durch ausschließliche Akzeptanz authentischer Zertifikate, solange die CA nicht kompromittiert ist.
+Dezentrale Modelle schützen durch die Vertrauensbildung zwischen den Nutzern solange Zertifikate überprüft werden.
+    \end{enumerate}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    \begin{enumerate}
+      \item D kann A nicht vertrauen, da weder eine direkte noch eine indirekte Zertifizierung von A durch B vorliegt und D nur B vertraut.
+      \item Nun kann D A vertrauen, da B C zertifiziert und C A zertifiziert. Somit liegt eine indirekte Zertifizierung von A durch B vor.
+    \end{enumerate}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    \begin{enumerate}
+      \item Im Schalenmodell müssen alle Zertifikate und Signaturen zum Zeitpunkt der Überprüfung gültig sein. Im Kettenmodell müssen diese lediglich zum Zeitpunkt der Ausstellung gültig sein.
+      \item Das Kettenmodell wird in der Praxis bevorzugt, da es flexibler ist und die Gültigkeit von Zertifikaten und Signaturen nicht permanent überprüft werden muss.
+      \item Schalenmodell: Zertifikatsausstellung bis Ablauf Wurzelzertifikat. Kettenmodell: Zertifikatsausstellung bis Ablauf Alice's Zertifikat.
+      \item In beiden Modellen gültig, da Zertifikatskette und Signatur zum Zeitpunkt der Ausstellung gültig waren.
+      \item Nur noch im Kettenmodell gültig, da im Schalenmodell die Kette nicht mehr gültig ist.
+    \end{enumerate}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    \begin{enumerate}
+      \item \textbf{Home Location Register} (HLR): Speichert Daten von Teilnehmern und wird von Netzbetreibern verwaltet.
+      \item \textbf{Visitor Location Register} (VLR): Speichert Daten von aktuell eingebuchten Teilnehmern und wird von Netzbetreibern verwaltet.
+      \item \textbf{Authentication Center} (AuC): Generiert und speichert Schlüssel und wird von Netzbetreibern verwaltet.
+      \item \textbf{Equipment Identity Register} (EIR): Speichert Daten von Geräten und wird von Netzbetreibern verwaltet.
+    \end{enumerate}
   \end{solution}
@@
   \begin{solution}
-    % TODO
+    \begin{enumerate}
+      \item Das Geheimnis $K_i$ kennen der Teilnehmer (SIM-Karte) und das Authentication Center (AuC).
+      \item Ziele des Angreifers sind IMSI-Catching (Identitätsdiebstahl) und Tracking. Möglichkeiten des Angreifers sind IMSI-Catcher (gefälschte Basisstation) und Analyse der TSMI-Vergabe.
+      \item Die Algorithmen wurden geändert, um die Sicherheit zu erhöhen. A5/1 und A5/2 sind unsicher, Kasumi (UMTS) und AES (LTE) sind sicherer.
+      \item Symmetrische Verfahren sind schneller und benötigen weniger Rechenleistung und verbrauchen somit weniger Energie.
+      \item Ziel ist die Kopie einer SIM-Karte, hierbei werden wiederholte Authentifizierungsanfragen genutzt um $K_i$ zu ermitteln.
+Da die Authentifizierungsanfragen relativ langsam sind, dauert der Angriff lange.
+    \end{enumerate}
   \end{solution}
@@
    \begin{solution}
-     % TODO
+     \begin{enumerate}
+      \item Die Nachrichten sind nicht verschlüsselt, da der Schlüsselaustausch noch nicht abgeschlossen ist und die Nachrichten zur Authentifizierung dienen.
+      \item Einführung eines Pre-Shared Keys (PSK) für die Verschlüsselung der ersten Nachrichten. Der PSK wird bei der SIM-Karten-Aktivierung festgelegt und ist den Netzbetreibern bekannt.
+     \end{enumerate}
    \end{solution}

### 2024-12-16 NLP/exercises.tex
Updated NLP 2021 exam.

@@
 \begin{exercise}{Comprehensiveness}
-  Explain comprehensiveness.
-
-  \begin{solution}
-    Comprehensiveness refers to the extent to which a system, dataset, or model covers all relevant aspects of a task or domain. In NLP:
-    \begin{itemize}
-        \item \textbf{Datasets}: Comprehensive datasets should include diverse and representative samples of the target domain.
-        \item \textbf{Models}: Comprehensive models should handle all variations and exceptions within their scope.
-        \item \textbf{Applications}: Comprehensive applications should meet all user requirements for a given task.
-    \end{itemize}
+  What is comprehensiveness, specifically in the context of NLP?
+
+  \begin{solution}
+    Comprehensiveness refers to the extent to which a system, dataset, or model covers all relevant aspects of a task or domain.
+    In NLP, comprehensiveness can be evaluated based on the coverage of linguistic phenomena, language varieties, or domain-specific terms in a dataset or model.
   \end{solution}
@@
 \begin{exercise}{Complete the picture}
-  Fill in the white boxes in the image.
-
-  \begin{solution}
-    To complete the image, analyze the given context and fill in the white boxes based on:
-    \begin{itemize}
-        \item \textbf{Labels}: Assign appropriate titles or labels to the boxes (e.g., steps in a pipeline).
-        \item \textbf{Relationships}: Connect missing elements to ensure the flow aligns with the diagram's logic.
-        \item \textbf{Annotations}: Use any provided hints to accurately reconstruct the intended content.
-    \end{itemize}
-    The exact answer depends on the image provided.
+  \input{res/qa.fig}
+
+  Fill in the five white boxes in the image.
+
+  \begin{solution}
+    1. query formulation, 2. query classification, 3. indexing, 4. doc retrieval, 5. passage retrieval
   \end{solution}
@@
 \begin{exercise}{NLTK}
-  Explain two parameters in NLTK that start with a "c."
-
-  \begin{solution}
-    Two common NLTK parameters starting with "c" are:
-    \begin{itemize}
-        \item \textbf{Chunking}: Groups words into meaningful units (e.g., noun phrases) based on part-of-speech tagging.
-        \item \textbf{Concordance}: Displays occurrences of a word within a context window, useful for lexical analysis.
+  Explain two parameters in NLTK that start with a "c".
+
+  \begin{solution}
+    \begin{itemize}
+      \item \textbf{Chunking}: Identifies and groups words into syntactic chunks based on their POS tags.
+      \item \textbf{Collocations}: Identifies words that frequently co-occur within a certain span.
+      \item \textbf{Concordance}: Displays a word in context, showing its surrounding words within a specified window.
     \end{itemize}
@@
   \begin{solution}
-    For multi-class classification, SVM uses:
     \begin{itemize}
@@
         \item \textbf{One-vs-All (OvA)}: Trains one classifier per class against all other classes. The highest-confidence prediction is chosen.
-        \item \textbf{Extensions}: Kernel tricks and soft margins can be applied for better generalization.
     \end{itemize}
@@
     \begin{itemize}
-        \item \textbf{Classical ML}:
-        \begin{itemize}
-            \item Relies on manually engineered features (e.g., TF-IDF, n-grams).
-            \item Requires domain expertise for feature extraction.
-        \end{itemize}
-        \item \textbf{Deep learning}:
-        \begin{itemize}
-            \item Automatically learns hierarchical feature representations.
-            \item Captures complex patterns (e.g., semantics, dependencies).
-            \item Requires large datasets for training.
-        \end{itemize}
+        \item \textbf{Classical ML} relies on manually engineered features (like tf-idf or n-grams) and requires domain expertise for feature extraction.
+        \item \textbf{Deep networks} automatically learn hierarchical feature representations from raw data, capturing complex patterns and dependencies, but requiring large datasets for training.
     \end{itemize}
@@
 \begin{exercise}{Cohen's Kappa}
-  Calculate Cohen's Kappa for a given table.
-
-  \begin{solution}
-    Cohen's Kappa is calculated as:
-    \[
-    \kappa = \frac{P_o - P_e}{1 - P_e}
-    \]
-    Where:
-    \begin{itemize}
-        \item \(P_o\): Observed agreement (sum of diagonal elements divided by the total number of items).
-        \item \(P_e\): Expected agreement (sum of product of marginal probabilities for each category).
-    \end{itemize}
-    Substitute the values from the given table into the formula to compute \(\kappa\).
+  \input{res/results.tbl}
+  Calculate Cohen's Kappa for Table \ref{tbl:results}.
+
+  \begin{solution}
+    Cohen's Kappa is calculated as $\kappa = \frac{P_o - P_e}{1 - P_e}$ and describes the observed agreement $P_o$ relative to the expected agreement $P_e$.
+      \[ \kappa = \frac{.7333-.5}{1-.5} = \frac{0.2333}{0.5} = 0.4666 \]
   \end{solution}
@@
 \begin{exercise}{Evaluation metrics}
-  Calculate Precision, Recall, and F1-Score.
-
-  \begin{solution}
-    Given true positives (\(TP\)), false positives (\(FP\)), and false negatives (\(FN\)):
-    \begin{itemize}
-        \item \textbf{Precision}: \( \frac{TP}{TP + FP} \)
-        \item \textbf{Recall}: \( \frac{TP}{TP + FN} \)
-        \item \textbf{F1-Score}: \( 2 \times \frac{\text{Precision} \times \text{Recall}}{\text{Precision} + \text{Recall}} \)
-    \end{itemize}
-    Plug in the values provided to compute these metrics.
+  \input{res/results.tbl}
+  Calculate Precision, Recall, and F1-Score for Table \ref{tbl:results}.
+
+  \begin{solution}
+    \begin{itemize}
+        \item \textbf{Precision} $\frac{TP}{TP + FP} = \frac{10}{10+5} = .6666$
+        \item \textbf{Recall}: $\frac{TP}{TP + FN} = \frac{10}{10+3} = .7692$
+        \item \textbf{F1-Score}: $2 \times \frac{\text{Precision} \times \text{Recall}}{\text{Precision} + \text{Recall}} = 2 \times \frac{.6666 \times .7692}{.6666 + .7692} = .7142$
+    \end{itemize}
   \end{solution}
@@
   \begin{solution}
-    Classification is used in IR for:
     \begin{itemize}
@@
 \begin{exercise}{Translate datasets}
-  List 3 reasons for and against translating an existing annotated dataset from German to another language.
-
-  \begin{solution}
-    \begin{itemize}
-        \item \textbf{For}:
-        \begin{itemize}
-            \item Expands dataset availability to resource-scarce languages.
-            \item Enables cross-lingual NLP research.
-            \item Saves effort in recreating annotations.
-        \end{itemize}
-        \item \textbf{Against}:
-        \begin{itemize}
-            \item Potential loss of annotation quality in translation.
-            \item Context or cultural relevance may not transfer.
-            \item Requires manual review to ensure correctness.
-        \end{itemize}
+  What are the pros and cons of translating datasets for NLP tasks? List 3 for each.
+
+  \begin{solution}
+    \begin{itemize}
+        \item[+] Expands dataset availability to resource-scarce languages.
+        \item[+] Enables cross-lingual NLP research.
+        \item[+] Saves effort in recreating annotations.
+        \item[-] Potential loss of annotation quality in translation.
+        \item[-] Context or cultural relevance may not transfer.
+        \item[-] Requires manual review to ensure correctness.
     \end{itemize}
@@
 \begin{exercise}{Web corpus}
-  List 3 reasons for and against using the web as a corpus.
-
-  \begin{solution}
-    \begin{itemize}
-        \item \textbf{For}:
-        \begin{itemize}
-            \item Large-scale, diverse data sources.
-            \item Includes up-to-date content.
-            \item Covers multiple languages and domains.
-        \end{itemize}
-        \item \textbf{Against}:
-        \begin{itemize}
-            \item Noisy and unstructured data.
-            \item Contains duplicates and redundant information.
-            \item May include biased or low-quality content.
-        \end{itemize}
+  What are opportunities and challenges in using the web as a corpus?
+
+  \begin{solution}
+    \begin{itemize}
+        \item[+] Large-scale, diverse data sources.
+        \item[+] Includes up-to-date content.
+        \item[+] Covers multiple languages and domains.
+        \item[-] Noisy and unstructured data.
+        \item[-] Contains duplicates and redundant information.
+        \item[-] May include biased or low-quality content.
     \end{itemize}
@@
 \begin{exercise}{Pipeline}
-  Given the pipeline from the first slide, assign three words to the individual steps.
-
-  \begin{solution}
-    Example:
-    \begin{itemize}
-        \item \textbf{Preprocessing}: Tokenization, normalization, and cleaning.
-        \item \textbf{Feature extraction}: TF-IDF, embeddings.
+  \input{res/pipeline.fig}
+  Assign three words to each stage of the given NLP pipeline in Figure \ref{fig:pipeline}.
+
+  \begin{solution}
+    \begin{itemize}
+      \item \textbf{Data acquisition}: Crawling, scraping, downloading.
+      \item \textbf{Text cleaning}: Tokenization, normalization, filtering.
+      \item \textbf{Preprocessing}: Lemmatization, POS tagging, parsing.
+      \item \textbf{Feature engineering}: TF-IDF, embeddings, n-grams.
       \item \textbf{Modeling}: Training, evaluation, and deployment.
+      \item \textbf{Evaluation}: Metrics, analysis, feedback.
+      \item \textbf{Deployment}: Integration, monitoring, maintenance.
+      \item \textbf{Monitoring \& Model Updating}: Performance tracking, retraining, versioning.
     \end{itemize}
@@
 \begin{exercise}{MRR and accuracy}
-  Calculate MRR and accuracy for a Question/Answering system.
-
-  \begin{solution}
-    \begin{itemize}
-        \item \textbf{Mean Reciprocal Rank (MRR)}:
-        \[
-        MRR = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{\text{rank}_i}
-        \]
-        Where \(\text{rank}_i\) is the rank of the first correct answer for query \(i\).
-        \item \textbf{Accuracy}: \(\frac{\text{Number of correct answers}}{\text{Total number of queries}}\)
-    \end{itemize}
-    Calculate based on the provided data.
+  \input{res/qa-answers.fig}
+  Calculate MRR and Accuracy for the Question/Answering system in Figure \ref{fig:qa-answers}.
+
+  \begin{solution}
+    \begin{itemize}
+        \item \textbf{Mean Reciprocal Rank} $MRR = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{\text{rank}_i} = \frac{1}{3} (\frac{1}{3}+\frac{1}{1}+0 = .4444)$
+        \item \textbf{Accuracy} $\frac{\text{Number of correct answers}}{\text{Total number of queries}} = \frac{2}{3} = .6666$
+    \end{itemize}
   \end{solution}
@@
 \begin{exercise}{Language detection}
-  Explain language detection.
-
-  \begin{solution}
-    Language detection involves identifying the language of a given text. Common approaches:
+  Explain language detection and common approaches.
+
+  \begin{solution}
+    Language detection involves identifying the language of a given text.
     \begin{itemize}
@@
 \begin{exercise}{Plagiarism detection}
-  Explain how to find plagiarism.
-
-  \begin{solution}
-    Plagiarism detection methods:
+  Explain three methods to detect plagiarism in text.
+
+  \begin{solution}
     \begin{itemize}
@@
 \begin{exercise}{Crowdsourcing}
-  List the pros and cons of crowdsourcing.
-
-  \begin{solution}
-    \begin{itemize}
-        \item \textbf{Pros}:
-        \begin{itemize}
-            \item Fast and scalable data annotation.
-            \item Access to diverse annotators.
-            \item Cost-effective compared to hiring experts.
-        \end{itemize}
-        \item \textbf{Cons}:
-        \begin{itemize}
-            \item Quality control is challenging.
-            \item May introduce noise or bias.
-            \item Difficult to ensure annotator expertise.
-        \end{itemize}
+  What are opportunities and challenges in using crowdsourcing for data annotation?
+
+  \begin{solution}
+    \begin{itemize}
+        \item[+] Fast and scalable data annotation.
+        \item[+] Access to diverse annotators.
+        \item[+] Cost-effective compared to hiring experts.
+        \item[-] Quality control is challenging.
+        \item[-] May introduce noise or bias.
+        \item[-] Difficult to ensure annotator expertise.
     \end{itemize}
@@
   \begin{solution}
-    Recall can be improved by:
     \begin{itemize}
@@
 \begin{exercise}{Spell checker}
-  Explain how a spell checker works with non-words.
-
-  \begin{solution}
-    For non-words, spell checkers:
+  How does a spell checker handle non-words?
+
+  \begin{solution}
     \begin{itemize}
@@
   \begin{solution}
-    The Vector Space Model represents documents and queries as vectors:
     \begin{itemize}
@@
 \begin{exercise}{Sentiment analysis}
-  Explain the areas where sentiment analysis is used.
-
-  \begin{solution}
-    Sentiment analysis applications:
+  Explain three areas where sentiment analysis is used.
+
+  \begin{solution}
     \begin{itemize}
@@
 \begin{exercise}{Teacher/Student}
-  Explain Teacher/Student and why it is better than just audio.
-
-  \begin{solution}
-    The Teacher/Student approach combines:
+  Explain the teacher/student signals approach in educational NLP and why it is better than pure audio content.
+
+  \begin{solution}
     \begin{itemize}
@@
 \begin{exercise}{Representation and fusion}
-  Explain representation and fusion and the differences.
+  What are representation and fusion in multimodal NLP, and how do they differ?
 

### 2024-12-17 NLP/exercises.tex
Merged exercises and removed redundancy.

@@
 
-\begin{exercise}{Web as an application area for NLP}
-  The web is an application area for NLP. Provide examples of applications.
+\begin{exercise}{NLP and the Web}
+  \begin{enumerate}
+    \item The web is an application area for NLP. Provide examples of applications.
+    \item How can the web improve the quality of NLP? Provide examples.
+  \end{enumerate}
 
   \begin{solution}
-    Examples include:
+    \begin{enumerate}
+      \item Information extraction, sentiment analysis, machine translation, question answering, etc.
+      \item Web as a corpus, analyzing web-based knowledge repositories, extracting structured information, using web-scale data for training language models.
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Datasets}
+  \begin{enumerate}
+    \item What datasets are used in NLP research, and why are they important?
+    \item What is the annotation process for a corpus or dataset, and how does it impact machine learning applications?
+    \item What are the main strategies for large-scale dataset collection using crowdsourcing?
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+      \item NLP datasets are collections of annotated or unannotated text data designed for specific NLP tasks (e.g., sentiment analysis, translation).
+They serve as a foundation for training and testing machine learning models, enable reproducibility in research, and facilitate comparisons between models and techniques.
+      \item Annotation process involves adding labels or metadata to text data (e.g., part-of-speech tags, sentiment labels).
+It improves model performance by providing high-quality training data, enables supervised learning by associating input with output labels, and impacts scalability and generalization of models depending on annotation quality.
+      \item Strategies include task design (clearly define annotation tasks with examples and guidelines), platform selection (use platforms like Amazon Mechanical Turk or Prolific), quality control (use redundancy and include test questions to ensure reliability), and data aggregation (combine annotations using majority voting or statistical models).
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Text preprocessing}
+  \begin{enumerate}
+    \item Why are more features not always better for learning a classifier?
+    \item What is the typical type of input data that a sequence tagger training step requires?
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+      \item More features can lead to overfitting, require more training data, introduce noise, and slow down training and inference.
+      \item Labeled training text (e.g., part-of-speech tags, named entity recognition).
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+\setcounter{section}{2022}
+\sheet{1st exam}
+
+\begin{exercise}{Tokenization ambiguity}\label{ex:tokenization}
+  List three types of tokenization ambiguity and explain each.
+
+  \begin{solution}
     \begin{itemize}
-        \item Internet of Services
-        \item Community mining
-        \item Information retrieval
-        \item Machine translation for web-based content
+      \item \textbf{Hyphenation}: Is "state-of-the-art" one token or three?
+      \item \textbf{Punctuation}: Should periods in abbreviations (e.g., "U.S.") be separated?
+      \item \textbf{Contractions}: Should "don't" be split into "do" and "n't"?
+      \item \textbf{Language differences}: Tokenization rules vary by language (e.g., Chinese characters vs. English words).
     \end{itemize}
@@
 
-\begin{exercise}{Web as a resource for NLP}
-  How can the web improve the quality of NLP? Provide examples.
-
-  \begin{solution}
-    Examples include:
-    \begin{itemize}
-        \item Web as Corpus
-        \item Analyzing web-based knowledge repositories
-        \item Extracting structured information from unstructured web data
-        \item Using web-scale data for training language models
-    \end{itemize}
-  \end{solution}
-\end{exercise}
-
-\begin{exercise}{Morphological analysis}
-  Identify all stems and affixes (prefix, suffix, infix, circumfix) in the following words: "index," "incorrect," "interesting."
-
-  \begin{solution}
-    Morphological breakdown:
-    \begin{itemize}
-        \item \textbf{index}: stem = "index"
-        \item \textbf{incorrect}: prefix = "in," stem = "correct"
-        \item \textbf{interesting}: stem = "interest," suffix = "ing"
-    \end{itemize}
-  \end{solution}
-\end{exercise}
-
-\begin{exercise}{Stemming vs lemmatization}
-  In contrast to lemmatization, stemming does not necessarily return a valid word form. Why is stemming still useful?
-
-  \begin{solution}
-    Stemming is still useful because:
-    \begin{itemize}
-        \item It is faster and computationally less expensive.
-        \item It simplifies implementation for applications like Information Retrieval.
-        \item It can still group similar words under a common base, which is often sufficient.
-    \end{itemize}
-  \end{solution}
-\end{exercise}
-
-\begin{exercise}{Corpus definition and usage}
-  What is a corpus, and how can it be built and used for linguistic analysis?
-
-  \begin{solution}
-    \begin{itemize}
-        \item \textbf{Corpus definition}: A corpus is a large and structured set of texts, often used for linguistic research.
-        \item \textbf{How to build}: Collect text data, preprocess it (cleaning, normalization), and annotate it for specific tasks.
-        \item \textbf{Usage}: Analyze linguistic patterns, train NLP models, and evaluate machine learning algorithms.
-    \end{itemize}
-  \end{solution}
-\end{exercise}
-
-\begin{exercise}{NLP datasets}
-  What are NLP datasets, and why are they important?
-
-  \begin{solution}
-    \begin{itemize}
-        \item \textbf{Definition}: NLP datasets are collections of annotated or unannotated text data designed for specific NLP tasks (e.g., sentiment analysis, translation).
-        \item \textbf{Importance}:
-        \begin{itemize}
-            \item Serve as a foundation for training and testing machine learning models.
-            \item Enable reproducibility in research.
-            \item Facilitate comparisons between models and techniques.
-        \end{itemize}
-    \end{itemize}
-  \end{solution}
-\end{exercise}
-
-\begin{exercise}{Corpus and dataset annotation}
-  What is the annotation process for a corpus or dataset, and how does it impact machine learning applications?
-
-  \begin{solution}
-    \begin{itemize}
-        \item \textbf{Annotation process}: Adding labels or metadata to text data (e.g., part-of-speech tags, sentiment labels, named entities).
-        \item \textbf{Impact on machine learning}:
-        \begin{itemize}
-            \item Improves model performance by providing high-quality training data.
-            \item Enables supervised learning by associating input with output labels.
-            \item Impacts scalability and generalization of models depending on annotation quality.
-        \end{itemize}
-    \end{itemize}
-  \end{solution}
-\end{exercise}
-
-\begin{exercise}{Large-scale dataset collection using crowdsourcing}
-  What are the main strategies for large-scale dataset collection using crowdsourcing?
-
-  \begin{solution}
-    Strategies include:
-    \begin{itemize}
-        \item \textbf{Task design}: Clearly define annotation tasks with examples and guidelines.
-        \item \textbf{Platform selection}: Use platforms like Amazon Mechanical Turk or Prolific.
-        \item \textbf{Quality control}: Use redundancy (multiple annotators per task) and include test questions to ensure reliability.
-        \item \textbf{Data aggregation}: Combine annotations using majority voting or statistical models.
-    \end{itemize}
-  \end{solution}
-\end{exercise}
-
-% Q: Why are more features not always better for learning a
-% classifier?
-% A: More training data might be needed; Dependent features
-% might lead to deficient models; Some ML algorithms can only deal
-% with specific data types; Slower.
-% Q: What is the typical type of input data that a sequence tagger
-% training step requires?
-% A: Labeled training text. (e.g. POS, NER)
-
-\setcounter{section}{2022}
-\sheet{1st exam}
-
-\begin{exercise}{Tokenization}
-  Give at least 3 examples of ambiguity in tokenization.
-
-  \begin{solution}
-    \begin{enumerate}
-        \item \textbf{Hyphenated words}: Does "state-of-the-art" count as one token or three?
-        \item \textbf{Contractions}: Should "don't" be treated as one token or split into "do" and "n't"?
-        \item \textbf{Punctuation}: How should punctuation like periods in "U.S." or apostrophes in possessives ("John's book") be handled?
-    \end{enumerate}
-  \end{solution}
-\end{exercise}
-
-\begin{exercise}{Parse trees}
-  "\textit{John shot the man with the gun}". Draw all possible parse trees for this sentence using the English grammar. What types of syntactic ambiguity are present?
+\begin{exercise}{Parsing ambiguity}
+  \begin{quote}\center "\textit{John shot the man with the gun}"\end{quote}
+  Draw all possible parse trees for this sentence using the English grammar. What types of syntactic ambiguity are present?
 
@@
 
-\begin{exercise}{Stemming and lemmatization}
-  What is the difference between stemming and lemmatization?
-
-  \begin{solution}
-    While stemming reduces a word to its base form by removing pre- and suffices ("running" → "run"), lemmatization considers the word's context and grammar to find its dictionary form ("better" → "good").
-Lemmatization is more precise than stemming but computationally more expensive.
-  \end{solution}
-\end{exercise}
-
-\begin{exercise}{Web as corpus}
-  List 4 issues you have to deal with when using the web as a corpus for NLP purposes and explain each.
+\begin{exercise}{Stemming v. Lemmatization}\label{ex:stem-v-lemm}
+  \begin{enumerate}
+    \item What is the difference between stemming and lemmatization?
+    \item In contrast to lemmatization, stemming does not necessarily return a valid word form. Why is stemming still useful?
+    \item Identify all stems and affixes (prefix, suffix, infix, circumfix) in the following words: "index," "incorrect," "interesting."
+  \end{enumerate}
 
@@
     \begin{enumerate}
-        \item \textbf{Noise}: Content may contain errors, irrelevant information, or non-standard grammar.
-        \item \textbf{Duplicates}: Identical or nearly identical data appears multiple times.
-        \item \textbf{Bias}: Web content often reflects societal biases, which can skew analyses.
-        \item \textbf{Reproducability}: Web pages change frequently, making reproducibility challenging.
+        \item Stemming reduces words to their root form by removing affixes. Lemmatization finds the base or dictionary form of a word based on its context.
+        \item Stemming is useful for information retrieval tasks where speed is crucial, and minor inaccuracies are acceptable.
+        \item \begin{itemize}
+            \item \textbf{index}: stem = "index"
+            \item \textbf{incorrect}: prefix = "in," stem = "correct"
+            \item \textbf{interesting}: stem = "interest," suffix = "ing"
+        \end{itemize}
     \end{enumerate}
@@
 
-\begin{exercise}{Cross-language QA}
-  How does cross-language QA work and what are the main issues?
-
-  \begin{solution}
-    Cross-language QA involves answering questions in one language based on information in another.
-This is accomplished by translating the question, retrieving relevant information in the target language, and generating an answer.
-Challenges include translation errors, cultural differences.
-  \end{solution}
-\end{exercise}
-
-\begin{exercise}{Machine learning pipeline}
-  What are the main concepts in a machine learning pipeline?
-
-  \begin{solution}
-    Data Preprocessing $\to$ Feature extraction $\to$ Model training $\to$ Evaluation $\to$ Deployment
-  \end{solution}
-\end{exercise}
-
-\begin{exercise}{IR evaluation}
-  What are the problems with precision, recall, and F-measure in IR? How can these be improved? Why is Cohen's Kappa better?
-
-  \begin{solution}
+\begin{exercise}{Machine learning}\label{ex:ml}
+  \begin{enumerate}
+    \item Which of the following statements are correct? Give a brief explanation for each.
       \begin{itemize}
-        \item \textbf{Precision}: How relevant are the retrieved documents? Favors retrieving fewer documents.
-        \item \textbf{Recall}: How many of the relevant documents were retrieved? Favors retrieving more documents.
-        \item \textbf{F-measure}: Harmonic mean of precision and recall. Balances the two metrics but not adjustable.
-        \item \textbf{Cohen's Kappa}: Measures inter-annotator agreement, accounting for chance agreement. Better for evaluating IR systems because it considers agreement beyond chance.
+        \item[$\square$] Features for Naive Bayes classifiers do not need to be uncorrelated.
+        \item[$\square$] Naive Bayes is slow to train.
+        \item[$\square$] SVM itself is a multi-label classifier.
+        \item[$\square$] SVM handles sparse data well.
       \end{itemize}
-  \end{solution}
-\end{exercise}
-
-\begin{exercise}{IR evaluation}
-  Calculate P@5, P@10, and MAP for the given IR ranked result list. Analyze both IR system A and system B.
-
-  \begin{solution}
-    For P@5, P@10, and MAP:
-    \begin{itemize}
-        \item \textbf{P@5}: Count relevant documents in the top 5 results and divide by 5.
-        \item \textbf{P@10}: Same calculation for the top 10 results.
-        \item \textbf{MAP}: Average Precision for each query, then average across all queries.
-    \end{itemize}
-    Steps:
-    \begin{itemize}
-        \item Identify positions of relevant documents in ranked lists.
-        \item Calculate precision at each relevant position.
-        \item Take mean of these precision values.
-    \end{itemize}
-  \end{solution}
-\end{exercise}
-
-\begin{exercise}{Naive Bayes and SVM}
-  Which of the following statements are correct?
-  \begin{enumerate}
-    \item Features for Naive Bayes classifiers do not need to be uncorrelated.
-    \item Naive Bayes is slow to train.
-    \item SVM itself is a multi-label classifier.
-    \item SVM handles sparse data well.
+      \item Explain Support Vector Machines when you have more than two classes.
+      \item What are the main stages in a machine learning pipeline?
   \end{enumerate}
@@
   \begin{solution}
-    Correct statements:
     \begin{enumerate}
+      \item \begin{itemize}
           \item \textbf{True}: Features for Naive Bayes do not need to be uncorrelated, though it assumes independence for computation.
@@
           \item \textbf{True}: SVM handles sparse data well due to its kernel functions.
+        \end{itemize}
+      \item \begin{itemize}
+          \item \textbf{One-vs-One (OvO)}: Trains a classifier for every pair of classes. Final classification is determined by majority voting.
+          \item \textbf{One-vs-All (OvA)}: Trains one classifier per class against all other classes. The highest-confidence prediction is chosen.
+        \end{itemize}
+      \item Data Preprocessing $\to$ Feature extraction $\to$ Model training $\to$ Evaluation $\to$ Deployment
     \end{enumerate}
@@
 
-\begin{exercise}{Corpus construction}
-  How can a corpus be constructed for linguistic purposes from web data, and how can it be made usable?
+\begin{exercise}{Information retrieval}\label{ex:ir}
+  Information retrieval (IR) is concerned with finding relevant information in large collections of data.
+  \begin{enumerate}
+    \item Describe stages in the IR cycle.
+    \item Where is classification used in IR?
+    \item What are key differences between Question Answering (QA) and IR?
+  \end{enumerate}
 
   \begin{solution}
-    Steps for constructing a corpus:
     \begin{enumerate}
-        \item \textbf{Data collection}: Crawl web pages, scrape data, or gather domain-specific text.
-        \item \textbf{Cleaning}: Remove duplicates, irrelevant content, and normalize text.
-        \item \textbf{Annotation}: Add metadata or labels for tasks like classification or parsing.
-        \item \textbf{Usability}: Store in accessible formats (e.g., JSON, XML) and document the corpus for ease of use.
+        \item \begin{itemize}
+            \item \textbf{Indexing}: Process and store data for efficient retrieval.
+            \item \textbf{Querying}: Users submit queries to search the index.
+            \item \textbf{Retrieval}: Return ranked results based on relevance scores.
+            \item \textbf{Evaluation}: Assess performance using metrics like precision and recall.
+          \end{itemize}
+        \item In query classification, document classification, and relevance classification.
+        \item QA provides direct answers, requires natural language understanding, and is used in chatbots. IR retrieves documents, focuses on keyword matching, and powers search engines.
     \end{enumerate}
@@
 
-\begin{exercise}{Pragmatics and discourse analysis}
+\begin{exercise}{Corpus construction}\label{ex:corpus}
+  \begin{enumerate}
+    \item Define a corpus in the context of NLP.
+    \item How would you construct a corpus usable for linguistic purposes?
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+      \item A corpus is a large collection of text data used for linguistic analysis, machine learning, or NLP research.
+      \item \begin{enumerate}
+          \item \textbf{Data collection}: Gather data from relevant sources (e.g., web crawling, domain-specific text).
+          \item \textbf{Cleaning and preprocessing}: Remove duplicates, non-linguistic content, and normalize text (e.g., case normalization, tokenization).
+          \item \textbf{Annotation}: Add metadata or linguistic labels (e.g., POS tags, syntactic trees) to the text.
+          \item \textbf{Storage}: Store the corpus in an accessible format like JSON, XML, or relational databases.
+          \item \textbf{Documentation}: Provide guidelines and metadata about the corpus for user understanding.
+        \end{enumerate}
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Pragmatics and discourse analysis}\label{ex:pragmatics}
   Why are pragmatics and discourse analysis difficult?
@@
   \begin{solution}
-    Challenges in pragmatics and discourse analysis:
     \begin{itemize}
@@
 
-\begin{exercise}{Hate speech}
-  Define hate speech and explain how to construct a hate speech classifier.
+\begin{exercise}{Hate speech}\label{ex:hate-speech}
+  \begin{enumerate}
+    \item Define hate speech.
+    \item How would you construct a hate speech classifier?
+    \item Explain how to counter hate speech.
+  \end{enumerate}
 
   \begin{solution}
-    \begin{itemize}
-        \item \textbf{Definition}: Hate speech involves language that attacks or discriminates against individuals or groups based on attributes like race, religion, ethnicity, gender, or sexual orientation.
-        \item \textbf{Constructing a hate speech classifier}:
     \begin{enumerate}
-            \item Collect labeled datasets with examples of hate speech and non-hate speech.
-            \item Preprocess text (e.g., tokenization, removing stop words).
-            \item Extract features (e.g., bag-of-words, TF-IDF, embeddings).
-            \item Train a supervised machine learning model (e.g., Naive Bayes, SVM, or deep learning models like LSTMs or transformers).
-            \item Evaluate the classifier using metrics like precision, recall, and F1-score.
+        \item Language that attacks or discriminates against individuals or groups based on attributes.
+        \item Collect labeled datasets, preprocess text, extract features, train a model, and evaluate it.
+        \item Collect examples of effective counter hate responses, classify hate speech, and generate counter-responses using NLP techniques.
     \end{enumerate}
-    \end{itemize}
   \end{solution}
@@
 
-\begin{exercise}{Counter hate speech}
-  What is counter hate speech, and how can a system be constructed for it?
+\begin{exercise}{Crowdsourcing}\label{ex:crowdsourcing}
+  \begin{enumerate}
+    \item What is the workflow for crowdsourced annotation?
+    \item What are opportunities and challenges in using crowdsourcing for data annotation?
+  \end{enumerate}
 
   \begin{solution}
-    \begin{itemize}
-        \item \textbf{Definition}: Counter hate speech is content designed to reduce the impact of hate speech, often by promoting positive, empathetic, or factual responses.
-        \item \textbf{Constructing a counter hate speech system}:
-        \begin{enumerate}
-            \item Collect datasets of effective counter hate responses.
-            \item Train models to generate or retrieve appropriate responses based on the hate speech content.
-            \item Use sentiment analysis and empathy detection to craft responses.
-            \item Deploy the system in platforms like social media or forums, ensuring ethical considerations are addressed.
-        \end{enumerate}
-    \end{itemize}
-  \end{solution}
-\end{exercise}
-
-\begin{exercise}{Crowdsourced annotation}
-  What is the workflow for crowdsourced annotation?
-
-  \begin{solution}
-    Steps in a crowdsourced annotation workflow:
     \begin{enumerate}
@@
     \end{enumerate}
-  \end{solution}
-\end{exercise}
-
-\begin{exercise}{Web corpus}
-  Why do duplicates exist on the web, and how can they be detected?
-
-  \begin{solution}
     \begin{itemize}
-        \item \textbf{Duplicates}: Duplicates arise from mirrored websites, copied content, or redundant crawls.
-        \item \textbf{Detection methods}:
-        \begin{itemize}
-            \item Hashing (e.g., MD5 or SHA): Compare hash values of documents.
-            \item Similarity measures: Use cosine similarity or Jaccard index on n-grams.
-            \item Structural analysis: Detect similarity in HTML structure.
-        \end{itemize}
+        \item[+] Fast and scalable data annotation.
+        \item[+] Access to diverse annotators.
+        \item[+] Cost-effective compared to hiring experts.
+        \item[-] Quality control is challenging.
+        \item[-] May introduce noise or bias.
+        \item[-] Difficult to ensure annotator expertise.
     \end{itemize}
@@
 
-\begin{exercise}{QA vs IR}
-  What are the differences between Question Answering (QA) and Information Retrieval (IR)?
+\begin{exercise}{Web as corpus}\label{ex:web-corpus}
+  In NLP research, the web is often used as a corpus.
+  \begin{enumerate}
+    \item What are some opportunities of using the web as a corpus?
+    \item The web contains many duplicates. How can they be detected?
+    \item Explain other key challenges in using the web as a corpus.
+  \end{enumerate}
 
   \begin{solution}
-    Differences between QA (Question Answering) and IR (Information Retrieval):
-    \begin{itemize}
-        \item \textbf{Goal}:
-        \begin{itemize}
-            \item QA aims to provide direct answers.
-            \item IR retrieves documents containing relevant information.
+    \begin{enumerate}
+      \item Large-scale, diverse data sources; up-to-date content; covers multiple languages and domains.
+      \item Hashing (e.g., MD5, SHA) to compare file contents; cosine similarity or Jaccard index for text comparisons; comparing HTML structure or metadata.
+      \item \begin{itemize}
+          \item \textbf{Noise}: Content may contain errors, irrelevant information, or non-standard grammar.
+          \item \textbf{Duplicates}: Identical or nearly identical data appears multiple times.
+          \item \textbf{Bias}: Web content often reflects societal biases, which can skew analyses.
+          \item \textbf{Reproducibility}: Web pages change frequently, making reproducibility challenging.
         \end{itemize}
-        \item \textbf{Processing}:
-        \begin{itemize}
-            \item QA requires deeper natural language understanding.
-            \item IR focuses on keyword matching and ranking.
-        \end{itemize}
-        \item \textbf{Applications}:
-        \begin{itemize}
-            \item QA: Chatbots, customer support.
-            \item IR: Search engines.
-        \end{itemize}
-    \end{itemize}
-  \end{solution}
-\end{exercise}
-
-\begin{exercise}{Cross-language QA}
-  What are the applications of multi-language QA, and what difficulties arise?
-
-  \begin{solution}
-    \begin{itemize}
-        \item \textbf{Applications}: Multi-language QA systems are used in global customer support, cross-border e-commerce, and multi-language search engines.
-        \item \textbf{Difficulties}:
-        \begin{itemize}
-            \item Translation errors can distort queries or results.
-            \item Cultural and linguistic differences complicate understanding and generating answers.
-            \item Lack of high-quality bilingual corpora for training.
-        \end{itemize}
-    \end{itemize}
-  \end{solution}
-\end{exercise}
-
-\begin{exercise}{Language detection}
-  How does language detection work?
-
-  \begin{solution}
-    Approaches for language detection:
-    \begin{enumerate}
-        \item \textbf{N-gram analysis}: Compare n-grams of text against language-specific frequency profiles.
-        \item \textbf{Machine learning models}: Use classifiers trained on labeled text for different languages.
-        \item \textbf{Lexicon-based methods}: Match words in the text to dictionaries of different languages.
-        \item \textbf{Hybrid approaches}: Combine multiple methods for higher accuracy.
     \end{enumerate}
@@
 
-\begin{exercise}{Factual QA vs opinionated QA}
-  What are the differences between factual QA and opinionated QA, and which is easier to evaluate?
+\begin{exercise}{Classification metrics}\label{ex:class-metrics}
+  Have a look at the classification results in Table \ref{tbl:results}. \input{res/results.tbl}
+
+  \begin{enumerate}
+    \item Calculate Precision, Recall, and F1-Score. What is the problem with these metrics in IR?
+    \item Calculate Cohen's Kappa. Why is it better for evaluating IR systems?
+  \end{enumerate}
 
   \begin{solution}
-    \begin{itemize}
-        \item \textbf{Factual QA}: Deals with objective information (e.g., "What is the capital of France?").
-        \item \textbf{Opinionated QA}: Involves subjective answers (e.g., "Is Paris a good travel destination?").
-        \item \textbf{Evaluation}: Factual QA is easier to evaluate as answers can be verified against ground truth. Opinionated QA requires subjective or aggregated assessments.
+    \begin{enumerate}
+      \item \begin{itemize}
+        \item \textbf{Precision} $\frac{TP}{TP + FP} = \frac{10}{10+5} = .6666$
+        \item \textbf{Recall}: $\frac{TP}{TP + FN} = \frac{10}{10+3} = .7692$
+        \item \textbf{F1-Score}: $2 \times \frac{\text{Precision} \times \text{Recall}}{\text{Precision} + \text{Recall}} = 2 \times \frac{.6666 \times .7692}{.6666 + .7692} = .7142$
         \end{itemize}
+        These metrics are limited as they do not account for chance agreement.
+      \item Cohen's Kappa is calculated as $\kappa = \frac{P_o - P_e}{1 - P_e}$ and describes the observed agreement $P_o$ relative to the expected agreement $P_e$.
+        \[ \kappa = \frac{.7333-.5}{1-.5} = \frac{0.2333}{0.5} = 0.4666 \]
+        This metric is better for evaluating IR systems because it considers agreement beyond chance.
+    \end{enumerate}
   \end{solution}
@@
 
-\begin{exercise}{RAG and LangChain}
+\begin{exercise}{Information retrieval metrics}\label{ex:ir-metrics}
+  Two search engines have produced the following results ($+$ = relevant, $-$ = irrelevant).
+  \begin{itemize}
+    \item \textbf{Engine A}: $+, -, +, -, +, +, -, +, +, +$
+    \item \textbf{Engine B}: $+, +, -, +, +, -, +, +, -, +$
+  \end{itemize}
+
+  \begin{enumerate}
+    \item Calculate P@5, P@10 and MAP for both engines.
+    \item How do these metrics help evaluate the performance of IR systems?
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+      \item \begin{itemize}
+          \item \textbf{Engine A}:
+          \begin{itemize}
+            \item $P@5 = \frac{3}{5} = 0.6$
+            \item $P@10 = \frac{7}{10} = 0.7$
+            \item $MAP = \frac{1}{3} \left( \frac{1}{1} + \frac{2}{4} + \frac{3}{5} \right) = 0.5666$
+          \end{itemize}
+          \item \textbf{Engine B}:
+          \begin{itemize}
+            \item $P@5 = \frac{4}{5} = 0.8$
+            \item $P@10 = \frac{7}{10} = 0.7$
+            \item $MAP = \frac{1}{3} \left( \frac{1}{1} + \frac{2}{4} + \frac{3}{7} \right) = 0.5476$
+          \end{itemize}
+        \end{itemize}
+      \item These metrics help evaluate IR systems by measuring precision at different cutoff points and the average precision across all relevant documents.
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Language and plagiarism detection}\label{ex:detect}
+  \begin{enumerate}
+    \item What are the main approaches for language detection?
+    \item And for plagiarism detection?
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+      \item \begin{itemize}
+          \item \textbf{N-gram analysis}: Compare the frequency of character n-grams against language-specific profiles.
+          \item \textbf{Machine learning models}: Train classifiers (e.g., SVM, neural networks) using labeled text data for various languages.
+          \item \textbf{Lexicon-based methods}: Use dictionaries to match input text to language-specific words.
+          \item \textbf{Hybrid approaches}: Combine n-gram, lexicon, and machine learning techniques for better accuracy.
+        \end{itemize}
+      \item \begin{itemize}
+          \item \textbf{String matching}: Use algorithms to find exact matches between documents.
+          \item \textbf{Semantic similarity}: Apply embeddings to detect paraphrased or rephrased content.
+          \item \textbf{Citation analysis}: Check for missing or incorrect citations.
+          \item \textbf{Fingerprinting}: Generate hashes of text fragments and compare across sources.
+        \end{itemize}
+      \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Question answering}\label{ex:qa}
+  Question Answering (QA) systems aim to provide direct answers to user queries.
+  \begin{enumerate}
+    \item What are the differences between factual QA and opinionated QA? and which is easier to evaluate?
+    \item Briefly explain QALD. What are key challenges?
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+      \item \begin{itemize}
+          \item \textbf{Factual QA}: Answers objective questions with verifiable information.
+          \item \textbf{Opinionated QA}: Provides subjective answers based on opinions or preferences.
+          \item Factual QA is easier to evaluate as answers can be verified against ground truth.
+        \end{itemize}
+      \item QALD (Question Answering over Linked Data) focuses on answering questions using linked data (e.g., RDF datasets).
+Challenges include mapping natural language queries to structured data, handling incomplete or ambiguous data, and scaling to large and heterogeneous sources.
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Cross-language QA}\label{ex:cross-lang-qa}
+  Cross-language Question Answering (QA) is a challenging task.
+  \begin{enumerate}
+    \item Explain the process of cross-language question-answering.
+    \item Where is it applied?
+    \item Explain the challenges faced in cross-language QA.
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+        \item Translate the question, retrieve relevant information in the target language, generate an answer, and translate it back.
+        \item Multi-language QA systems are used in global customer support, cross-border e-commerce, and multi-language search engines.
+        \item Challenges include translation errors, lack of parallel resources, cultural differences, and mismatched linguistic structures.
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{RAG and LangChain}\label{ex:rag}
   What is Retrieval-Augmented Generation (RAG), and how does LangChain assist in implementing it?
@@
 
-\begin{exercise}{Sentiment analysis}
-  Compare sentiment analysis using supervised approaches and lexicon-based approaches.
-
-  \begin{solution}
-    \begin{itemize}
-        \item \textbf{Supervised learning}: Uses labeled data to train machine learning models.
-        \begin{itemize}
-            \item \textbf{Advantages}: High accuracy with large datasets.
-            \item \textbf{Drawbacks}: Relies on annotated data and is domain-dependent.
-        \end{itemize}
-        \item \textbf{Lexicon-based approaches}: Use sentiment dictionaries to determine polarity.
-        \begin{itemize}
-            \item \textbf{Advantages}: Simplicity and no training requirement.
-            \item \textbf{Drawbacks}: Limited contextual understanding and poor performance on complex text.
-        \end{itemize}
-    \end{itemize}
-  \end{solution}
-\end{exercise}
-
-\begin{exercise}{Edit distance}
-  Calculate the Minimum Edit Distance for "TAC" vs. "CAT" and list all possible word outcomes.
-
-  \begin{solution}
-    \begin{itemize}
-        \item \textbf{Minimum Edit Distance for "TAC" and "CAT"}:
-        \[
-        \begin{array}{c|c|c|c|c}
-            &  & C & A & T \\
-            \hline
-            & 0 & 1 & 2 & 3 \\
-            T & 1 & 1 & 2 & 2 \\
-            A & 2 & 2 & 1 & 2 \\
-            C & 3 & 2 & 2 & 2 \\
-        \end{array}
-        \]
-        \item \textbf{Word outcomes}: Possible outcomes from edits:
+\begin{exercise}{Sentiment analysis}\label{ex:sent-analysis}
+  Sentiment analysis aims to determine the sentiment expressed in text.
   \begin{enumerate}
-            \item CAT
-            \item TAT
-            \item TAC
-            \item CAC
+    \item Describe three applications of sentiment analysis.
+    \item How do factual and opinionated sentiment analysis differ?
+    \item Compare sentiment analysis using supervised approaches and lexicon-based approaches.
+    \item How can lexicon-based sentiment analysis be improved using distributed thesauruses?
   \end{enumerate}
-    \end{itemize}
-  \end{solution}
-\end{exercise}
-
-\setcounter{section}{2021}
-\sheet{1st exam}
-
-\begin{exercise}{Corpus construction}
-  How would you construct a corpus for linguistic purposes and how can it be made usable?
 
@@
     \begin{enumerate}
-        \item \textbf{Data collection}: Gather data from relevant sources (e.g., web crawling, domain-specific text).
-        \item \textbf{Cleaning and preprocessing}: Remove duplicates, non-linguistic content, and normalize text (e.g., case normalization, tokenization).
-        \item \textbf{Annotation}: Add metadata or linguistic labels (e.g., POS tags, syntactic trees) to the text.
-        \item \textbf{Storage}: Store the corpus in an accessible format like JSON, XML, or relational databases.
-        \item \textbf{Documentation}: Provide guidelines and metadata about the corpus for user understanding.
+        \item \begin{itemize}
+            \item \textbf{Customer feedback}: Analyze reviews for product improvements.
+            \item \textbf{Social media monitoring}: Detect trends and public sentiment.
+            \item \textbf{Political analysis}: Gauge public opinion on policies or events.
+          \end{itemize}
+        \item \begin{itemize}
+            \item \textbf{Factual sentiment analysis}: Focuses on extracting objective sentiment (e.g., product ratings).
+            \item \textbf{Opinionated analysis}: Identifies subjective opinions (e.g., customer reviews).
+          \end{itemize}
+        \item \begin{itemize}
+            \item \textbf{Supervised learning}: Uses labeled data to train machine learning models.
+            \item \textbf{Lexicon-based approaches}: Use sentiment dictionaries to determine polarity.
+          \end{itemize}
+        \item Extend lexicons by identifying synonyms, antonyms, or similar words in distributed word embeddings like Word2Vec or GloVe.
       \end{enumerate}
-  \end{solution}
-\end{exercise}
-
-\begin{exercise}{Pragmatics and discourse analysis}
-  Why are pragmatics and discourse analysis difficult?
-
-  \begin{solution}
-    \begin{itemize}
-        \item \textbf{Context dependency}: Requires understanding of the broader context beyond sentence boundaries.
-        \item \textbf{Ambiguity}: Words or phrases can have multiple meanings depending on the discourse.
-        \item \textbf{Cultural and social norms}: Interpretation is influenced by societal and cultural factors.
-        \item \textbf{Incomplete information}: Missing parts of conversations or text make it harder to analyze.
-    \end{itemize}
-  \end{solution}
-\end{exercise}
-
-\begin{exercise}{Sequence tagging and CRF}
-  Define sequence tagging and explain what a Conditional Random Field (CRF) is.
-
-  \begin{solution}
-    \begin{itemize}
-        \item \textbf{Sequence tagging}: The process of assigning labels to each element in a sequence (e.g., part-of-speech tagging, named entity recognition).
-        \item \textbf{Conditional Random Field (CRF)}: A probabilistic model used for structured prediction in sequence data. It considers the context of neighboring labels to improve accuracy.
-    \end{itemize}
-  \end{solution}
-\end{exercise}
-
-\begin{exercise}{Minimum edit distance}
-  Calculate the Minimum Edit Distance for "TAC" vs. "CAT" (a table is given). List all possible outcomes.
-
-  \begin{solution}
-    \begin{itemize}
-        \item \textbf{Minimum Edit Distance Table}:
-        \[
-        \begin{array}{c|c|c|c|c}
-            &  & C & A & T \\
-            \hline
-            & 0 & 1 & 2 & 3 \\
-            T & 1 & 1 & 2 & 2 \\
-            A & 2 & 2 & 1 & 2 \\
-            C & 3 & 2 & 2 & 2 \\
-        \end{array}
-        \]
-        \item \textbf{Outcomes}:
-        \begin{enumerate}
-            \item CAT
-            \item TAT
-            \item TAC
-            \item CAC
-        \end{enumerate}
-    \end{itemize}
-  \end{solution}
-\end{exercise}
-
-\begin{exercise}{IR evaluation}
-  Calculate P@5 and MAP for given ranked IR results for two systems.
-
-  \begin{solution}
-    \begin{itemize}
-        \item \textbf{P@5}: Count relevant documents in the top 5 results and divide by 5.
-        \item \textbf{MAP}: For each query, calculate average precision across relevant documents and average these scores.
-    \end{itemize}
-  \end{solution}
-\end{exercise}
-
-\begin{exercise}{Hate speech}
-  Define hate speech and how to construct a hate speech classifier.
-
-  \begin{solution}
-    \begin{itemize}
-        \item \textbf{Definition}: Hate speech targets individuals or groups based on attributes like race, religion, gender, or orientation.
-        \item \textbf{Classifier construction}:
-        \begin{enumerate}
-            \item Collect labeled datasets of hate and non-hate speech.
-            \item Preprocess data by removing noise and tokenizing.
-            \item Train a machine learning model (e.g., Naive Bayes, SVM, transformers).
-            \item Evaluate using metrics like F1-score, precision, and recall.
-        \end{enumerate}
-    \end{itemize}
-  \end{solution}
-\end{exercise}
-
-\begin{exercise}{Counter hate speech}
-  What is counter hate speech, and how can a system be constructed for it?
-
-  \begin{solution}
-    \begin{itemize}
-        \item \textbf{Definition}: Counter hate speech neutralizes hate content by promoting positive or factual responses.
-        \item \textbf{System construction}:
-        \begin{enumerate}
-            \item Collect examples of effective counter hate responses.
-            \item Use NLP techniques to classify hate speech and generate counter-responses.
-            \item Train models on empathy and sentiment analysis for response generation.
-        \end{enumerate}
-    \end{itemize}
-  \end{solution}
-\end{exercise}
-
-\begin{exercise}{Sentiment lexicon extension}
-  How can sentiment analysis be improved using lexicon extensions from distributed thesauruses?
-
-  \begin{solution}
-    \begin{itemize}
-        \item \textbf{Lexicon extension}: Expand sentiment dictionaries using related terms from distributed word embeddings like Word2Vec or GloVe.
-        \item \textbf{Process}:
-        \begin{enumerate}
-            \item Identify synonyms, antonyms, or similar words in embedding spaces.
-            \item Update lexicons with these terms to improve coverage.
-            \item Validate new terms to ensure sentiment accuracy.
-        \end{enumerate}
-    \end{itemize}
-  \end{solution}
-\end{exercise}
-\begin{exercise}{Sentiment analysis}
-  Discuss sentiment detection using lexicon based vs. supervised learning approaches.
-
-  \begin{solution}
-    \begin{itemize}
-        \item \textbf{Lexicon-based approaches}: 
-        \begin{itemize}
-            \item Use pre-built dictionaries of words with sentiment labels (e.g., positive, negative).
-            \item Advantages:
-                \begin{itemize}
-                    \item Easy to implement and interpret.
-                    \item No need for labeled training data.
-                \end{itemize}
-            \item Disadvantages:
-                \begin{itemize}
-                    \item Limited to the coverage of the lexicon.
-                    \item Poor performance on context-dependent sentiment (e.g., sarcasm).
-                \end{itemize}
-        \end{itemize}
-        \item \textbf{Supervised learning approaches}:
-        \begin{itemize}
-            \item Train machine learning models on labeled data to classify sentiment.
-            \item Advantages:
-                \begin{itemize}
-                    \item Can learn from context and handle subtle sentiment expressions.
-                    \item Scalable to large datasets.
-                \end{itemize}
-            \item Disadvantages:
-                \begin{itemize}
-                    \item Requires annotated data, which can be costly to obtain.
-                    \item More computationally intensive than lexicon-based approaches.
-                \end{itemize}
-        \end{itemize}
-    \end{itemize}
-  \end{solution}
-\end{exercise}
-
-\begin{exercise}{Sentiment analysis}
-  Why is lexicon based sentiment analysis not enough?
-
-  \begin{solution}
-    \begin{itemize}
-        \item \textbf{Context-dependence}: Lexicons fail to account for how sentiment changes with context.
-        \item \textbf{Negation handling}: Sentences like "not bad" may be incorrectly classified without understanding negation.
-        \item \textbf{Ambiguity}: Words can have different sentiments in different contexts (e.g., "hot" in weather vs. arguments).
-        \item \textbf{Domain-specific issues}: Lexicons often do not cover domain-specific terms or slang.
-    \end{itemize}
-  \end{solution}
-\end{exercise}
-
-\begin{exercise}{Naive Bayes and SVM}
-  Which of the following 4 are correct:
-  \begin{enumerate}
-    \item Features for Naive Bayes classifiers do not need to be uncorrelated.
-    \item Naive Bayes is slow to train.
-    \item SVM itself is a multi-label classifier.
-    \item SVM handles sparse data well.
-  \end{enumerate}
-
-  \begin{solution}
-    \begin{itemize}
-        \item \textbf{Correct}: 
-        \begin{itemize}
-            \item Features for Naive Bayes classifiers do not need to be uncorrelated.
-            \item SVM handles sparse data well.
-        \end{itemize}
-        \item \textbf{Incorrect}:
-        \begin{itemize}
-            \item Naive Bayes is actually fast to train.
-            \item SVM is not inherently multi-label but can be adapted for multi-label classification.
-        \end{itemize}
-    \end{itemize}
-  \end{solution}
-\end{exercise}
-
-\begin{exercise}{Language detection}
-  Explain approaches for language detection.
-
-  \begin{solution}
-    \begin{itemize}
-        \item \textbf{N-gram analysis}: Compare the frequency of character n-grams against language-specific profiles.
-        \item \textbf{Machine learning models}: Train classifiers (e.g., SVM, neural networks) using labeled text data for various languages.
-        \item \textbf{Lexicon-based methods}: Use dictionaries to match input text to language-specific words.
-        \item \textbf{Hybrid approaches}: Combine n-gram, lexicon, and machine learning techniques for better accuracy.
-    \end{itemize}
-  \end{solution}
-\end{exercise}
-
-\begin{exercise}{Crowdsourced annotation}
-  Explain the crowdsourced annotation workflow.
-
-  \begin{solution}
-    Steps in a crowdsourced annotation workflow:
-    \begin{enumerate}
-        \item \textbf{Task design}: Define the annotation task with clear instructions and examples.
-        \item \textbf{Platform setup}: Deploy the task on crowdsourcing platforms like Amazon Mechanical Turk.
-        \item \textbf{Quality control}: Use redundancy by assigning multiple annotators to the same task and include test questions.
-        \item \textbf{Data aggregation}: Combine annotations using majority voting or weighted scoring methods.
-        \item \textbf{Evaluation}: Analyze the consistency and reliability of annotations.
-    \end{enumerate}
-  \end{solution}
-\end{exercise}
-
-\begin{exercise}{Tokenization ambiguities}
-  What are the ambiguities for tokenization?
-
-  \begin{solution}
-    Ambiguities in tokenization include:
-    \begin{itemize}
-        \item \textbf{Hyphenation}: Is "state-of-the-art" one token or three?
-        \item \textbf{Punctuation}: Should periods in abbreviations (e.g., "U.S.") be separated?
-        \item \textbf{Contractions}: Should "don't" be split into "do" and "n't"?
-        \item \textbf{Language differences}: Tokenization rules vary by language (e.g., Chinese characters vs. English words).
-    \end{itemize}
-  \end{solution}
-\end{exercise}
-
-\begin{exercise}{QA vs IR}
-  What are the differences between QA and IR?
-
-  \begin{solution}
-    Differences between QA and IR:
-    \begin{itemize}
-        \item \textbf{Goal}:
-        \begin{itemize}
-            \item QA provides direct answers to user queries.
-            \item IR retrieves documents that may contain relevant information.
-        \end{itemize}
-        \item \textbf{Processing}:
-        \begin{itemize}
-            \item QA requires natural language understanding and reasoning.
-            \item IR focuses on keyword matching and ranking algorithms.
-        \end{itemize}
-        \item \textbf{Applications}:
-        \begin{itemize}
-            \item QA is used in chatbots and virtual assistants.
-            \item IR powers search engines like Google.
-        \end{itemize}
-    \end{itemize}
-  \end{solution}
-\end{exercise}
-
-\begin{exercise}{Cross-language QA}
-  What are the challenges in cross-language QA?
-
-  \begin{solution}
-    Challenges in cross-language QA include:
-    \begin{itemize}
-        \item \textbf{Translation errors}: Inaccurate translations can distort queries or answers.
-        \item \textbf{Lack of parallel resources}: Few high-quality bilingual corpora exist for certain language pairs.
-        \item \textbf{Cultural differences}: Questions or answers may need cultural adaptation.
-        \item \textbf{Mismatched structures}: Linguistic differences make aligning queries and answers difficult.
-    \end{itemize}
-  \end{solution}
-\end{exercise}
-
-\begin{exercise}{Stemming vs lemmatization}
-  What is the difference between stemming and lemmatization?
-
-  \begin{solution}
-    \begin{itemize}
-        \item \textbf{Stemming}: Reduces a word to its root by chopping off affixes (e.g., "running" → "run"). It is less precise and may produce non-words.
-        \item \textbf{Lemmatization}: Converts a word to its base form (lemma) based on its context (e.g., "better" → "good"). It is more accurate but computationally expensive.
-    \end{itemize}
-  \end{solution}
-\end{exercise}
-
-\begin{exercise}{Web corpus}
-  Why are duplicates on the web, and how can they be detected?
-
-  \begin{solution}
-    \begin{itemize}
-        \item \textbf{Reasons for duplicates}:
-        \begin{itemize}
-            \item Content mirroring across websites.
-            \item Republishing of syndicated articles.
-            \item Crawling the same site multiple times.
-        \end{itemize}
-        \item \textbf{Detection methods}:
-        \begin{itemize}
-            \item Hashing (e.g., MD5, SHA) to compare file contents.
-            \item Cosine similarity or Jaccard index for text comparisons.
-            \item Comparing HTML structure or metadata.
-        \end{itemize}
-    \end{itemize}
-  \end{solution}
-\end{exercise}
-
-\begin{exercise}{Web corpus}
-  What are the 4 challenges when using the web as a corpus?
-
-  \begin{solution}
-    Challenges include:
-    \begin{itemize}
-        \item \textbf{Noisy data}: Web content often includes irrelevant or ungrammatical text.
-        \item \textbf{Duplication}: Large-scale crawls often capture duplicate data.
-        \item \textbf{Bias}: Web data may reflect societal or geographic biases.
-        \item \textbf{Dynamic content}: Websites frequently change, making reproducibility difficult.
-    \end{itemize}
-  \end{solution}
-\end{exercise}
-
-\begin{exercise}{TREC}
-  What are the TREC question types?
-
-  \begin{solution}
-    TREC (Text REtrieval Conference) defines these question types:
-    \begin{itemize}
-        \item \textbf{Factoid questions}: Require specific factual answers (e.g., "Who is the president of the U.S.?").
-        \item \textbf{List questions}: Require a list of items (e.g., "List countries in the EU").
-        \item \textbf{Definition questions}: Seek explanations or definitions (e.g., "What is AI?").
-    \end{itemize}
-  \end{solution}
-\end{exercise}
-
-\begin{exercise}{QALD}
-  What is QALD, and what are the challenges?
-
-  \begin{solution}
-    \begin{itemize}
-        \item \textbf{Definition}: QALD (Question Answering over Linked Data) focuses on answering questions using linked data (e.g., RDF datasets).
-        \item \textbf{Challenges}:
-        \begin{itemize}
-            \item Mapping natural language queries to structured data.
-            \item Handling incomplete or ambiguous data in linked datasets.
-            \item Scaling to large and heterogeneous data sources.
-        \end{itemize}
-    \end{itemize}
   \end{solution}
@@
 
+\begin{exercise}{Sequence tagging and CRF}\label{ex:seq-tagging}
+  \begin{enumerate}
+    \item What is sequence tagging in NLP?
+    \item Explain the role of Conditional Random Fields (CRF) in sequence tagging.
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+        \item Sequence tagging assigns labels to each token in a sequence (e.g., POS tagging, named entity recognition).
+        \item CRF is a probabilistic model that considers the dependencies between labels in a sequence, improving tagging accuracy by modeling label transitions.
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
 \begin{exercise}{NLTK}
@@
       \item \textbf{Concordance}: Displays a word in context, showing its surrounding words within a specified window.
-    \end{itemize}
-  \end{solution}
-\end{exercise}
-
-\begin{exercise}{SVM}
-  Explain Support Vector Machines when you have more than two classes.
-
-  \begin{solution}
-    \begin{itemize}
-        \item \textbf{One-vs-One (OvO)}: Trains a classifier for every pair of classes. Final classification is determined by majority voting.
-        \item \textbf{One-vs-All (OvA)}: Trains one classifier per class against all other classes. The highest-confidence prediction is chosen.
     \end{itemize}
@@
 
-\begin{exercise}{Cohen's Kappa}
-  \input{res/results.tbl}
-  Calculate Cohen's Kappa for Table \ref{tbl:results}.
-
-  \begin{solution}
-    Cohen's Kappa is calculated as $\kappa = \frac{P_o - P_e}{1 - P_e}$ and describes the observed agreement $P_o$ relative to the expected agreement $P_e$.
-      \[ \kappa = \frac{.7333-.5}{1-.5} = \frac{0.2333}{0.5} = 0.4666 \]
-  \end{solution}
-\end{exercise}
-
-\begin{exercise}{Evaluation metrics}
-  \input{res/results.tbl}
-  Calculate Precision, Recall, and F1-Score for Table \ref{tbl:results}.
-
-  \begin{solution}
-    \begin{itemize}
-        \item \textbf{Precision} $\frac{TP}{TP + FP} = \frac{10}{10+5} = .6666$
-        \item \textbf{Recall}: $\frac{TP}{TP + FN} = \frac{10}{10+3} = .7692$
-        \item \textbf{F1-Score}: $2 \times \frac{\text{Precision} \times \text{Recall}}{\text{Precision} + \text{Recall}} = 2 \times \frac{.6666 \times .7692}{.6666 + .7692} = .7142$
-    \end{itemize}
-  \end{solution}
-\end{exercise}
-
-\begin{exercise}{IR and classification}
-  Where is classification used in Information Retrieval?
-
-  \begin{solution}
-    \begin{itemize}
-        \item \textbf{Query classification}: Identify user intent (e.g., navigational vs. informational).
-        \item \textbf{Document classification}: Categorize documents into predefined topics or genres.
-        \item \textbf{Relevance classification}: Predict whether a document is relevant to a query.
-    \end{itemize}
-  \end{solution}
-\end{exercise}
-
 \begin{exercise}{Translate datasets}
@@
         \item[-] Requires manual review to ensure correctness.
-    \end{itemize}
-  \end{solution}
-\end{exercise}
-
-\begin{exercise}{Web corpus}
-  What are opportunities and challenges in using the web as a corpus?
-
-  \begin{solution}
-    \begin{itemize}
-        \item[+] Large-scale, diverse data sources.
-        \item[+] Includes up-to-date content.
-        \item[+] Covers multiple languages and domains.
-        \item[-] Noisy and unstructured data.
-        \item[-] Contains duplicates and redundant information.
-        \item[-] May include biased or low-quality content.
     \end{itemize}
@@
 
-\begin{exercise}{Language detection}
-  Explain language detection and common approaches.
-
-  \begin{solution}
-    Language detection involves identifying the language of a given text.
-    \begin{itemize}
-        \item \textbf{N-gram analysis}: Compare character n-grams to language profiles.
-        \item \textbf{Lexicon-based methods}: Match text to language-specific dictionaries.
-        \item \textbf{Machine learning}: Train classifiers (e.g., SVM, neural networks) on labeled text.
-        \item \textbf{Hybrid models}: Combine multiple techniques for higher accuracy.
-    \end{itemize}
-  \end{solution}
-\end{exercise}
-
-\begin{exercise}{Plagiarism detection}
-  Explain three methods to detect plagiarism in text.
-
-  \begin{solution}
-    \begin{itemize}
-        \item \textbf{String matching}: Use algorithms to find exact matches between documents.
-        \item \textbf{Semantic similarity}: Apply embeddings to detect paraphrased or rephrased content.
-        \item \textbf{Citation analysis}: Check for missing or incorrect citations.
-        \item \textbf{Fingerprinting}: Generate hashes of text fragments and compare across sources.
-    \end{itemize}
-  \end{solution}
-\end{exercise}
-
-\begin{exercise}{Crowdsourcing}
-  What are opportunities and challenges in using crowdsourcing for data annotation?
-
-  \begin{solution}
-    \begin{itemize}
-        \item[+] Fast and scalable data annotation.
-        \item[+] Access to diverse annotators.
-        \item[+] Cost-effective compared to hiring experts.
-        \item[-] Quality control is challenging.
-        \item[-] May introduce noise or bias.
-        \item[-] Difficult to ensure annotator expertise.
-    \end{itemize}
-  \end{solution}
-\end{exercise}
-
 \begin{exercise}{Recall improvement}
@@
 
-\begin{exercise}{IR cycle}
-  Explain the Information Retrieval cycle.
-
-  \begin{solution}
-    The IR cycle includes:
-    \begin{itemize}
-        \item \textbf{Indexing}: Process and store data for efficient retrieval.
-        \item \textbf{Querying}: Users submit queries to search the index.
-        \item \textbf{Retrieval}: Return ranked results based on relevance scores.
-        \item \textbf{Evaluation}: Assess performance using metrics like precision and recall.
-    \end{itemize}
-  \end{solution}
-\end{exercise}
-
 \begin{exercise}{Vector space}
@@
         \item \textbf{Similarity}: Compute cosine similarity to rank documents by relevance to the query.
-    \end{itemize}
-  \end{solution}
-\end{exercise}
-
-\begin{exercise}{Sentiment analysis}
-  Explain the difference between factual sentiment analysis and opinionated analysis.
-
-  \begin{solution}
-    \begin{itemize}
-        \item \textbf{Factual sentiment analysis}: Focuses on extracting objective sentiment (e.g., product ratings).
-        \item \textbf{Opinionated analysis}: Identifies subjective opinions (e.g., customer reviews).
-        \item \textbf{Key difference}: Factual analysis is more structured and easier to evaluate, whereas opinionated analysis is subjective and context-dependent.
-    \end{itemize}
-  \end{solution}
-\end{exercise}
-
-\begin{exercise}{Sentiment analysis}
-  Explain three areas where sentiment analysis is used.
-
-  \begin{solution}
-    \begin{itemize}
-        \item \textbf{Customer feedback}: Analyze reviews for product improvements.
-        \item \textbf{Social media monitoring}: Detect trends and public sentiment.
-        \item \textbf{Political analysis}: Gauge public opinion on policies or events.
-        \item \textbf{Market research}: Understand consumer preferences.
     \end{itemize}
@@
 
+\begin{exercise}{Edit distance}\label{ex:edit-distance}
+  \begin{enumerate}
+    \item Define the Minimum Edit Distance.
+    \item Calculate the Minimum Edit Distance for "TAC" vs. "CAT".
+    \item What are the possible word outcomes?
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+        \item The Minimum Edit Distance is the minimum number of operations (insertion, deletion, substitution) required to transform one string into another.
+        \item \[\begin{array}{c|c|c|c|c}
+            &  & C & A & T \\
+            \hline
+            & 0 & 1 & 2 & 3 \\
+            T & 1 & 1 & 2 & 2 \\
+            A & 2 & 2 & 1 & 2 \\
+            C & 3 & 2 & 2 & 2 \\
+        \end{array}\]
+        \item CAT, TAT, TAC, CAC
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{TREC}\label{ex:trec}
+  What are the TREC question types?
+
+  \begin{solution}
+    \begin{itemize}
+        \item \textbf{Factoid questions}: Require specific factual answers (e.g., "Who is the president of the U.S.?").
+        \item \textbf{List questions}: Require a list of items (e.g., "List countries in the EU").
+        \item \textbf{Definition questions}: Seek explanations or definitions (e.g., "What is AI?").
+    \end{itemize}
+  \end{solution}
+\end{exercise}
+
 \end{document}

### 2024-12-17 BAI/paper/paper.tex
Rewrote the whole paper. now i just need some results.

--- 2024-11-05 BAI/paper/paper.tex
@@
 \begin{abstract}
- This paper is about image captioning.
-We discuss the current state of the art and overview the most important methods.
-We present our results of experiments with different encoder and decoder models.
-We use the flickr8k dataset and evaluate the models using the BLEU score.
-We discuss the results and give an outlook on future work.
+    Image captioning, the task of generating descriptive textual captions for images, bridges the gap between computer vision and natural language processing.
+    This paper investigates the performance of three decoder architectures---Gated Recurrent Units (GRU), Long Short-Term Memory networks (LSTM), and Transformers---for image caption generation.
+    Using the \textbf{Flickr8k} dataset, features are extracted from images using a pre-trained \textbf{EfficientNet B0} model, and captions are generated using the respective decoders.
+    The models are implemented in \textbf{PyTorch}, and their performance is evaluated using standard metrics: \textbf{BLEU}, \textbf{METEOR}, and \textbf{NIST}.
+    % TODO: results
 \end{abstract}
@@
 Image captioning involves generating textual descriptions of given images.
-It is a difficult problem because it combines the visual and textual domains.
-In recent years, deep learning has revolutionized the field of image captioning.
+It is a complex problem because it combines the visual and textual domains, thus bridging computer vision (CV) and natural language processing (NLP).
+Generating natural, descriptive sentences for images involves detecting and interpreting visual elements and their relationships within depicted scenes.
+\par While deep learning with attention mechanisms has long since been the dominant approach in image captioning \cite{xu2015show}, recent advances have been brought about by transformers \cite{vaswani2017attention}.
+These models have shown remarkable performance in various natural language processing tasks, including image captioning.
+\par This paper aims to compare the performance of different decoders directly, while keeping other parameters constant.
+The flickr8k dataset is used and the models are implemented using \textbf{PyTorch}.
+The BLEU, METEOR, and NIST scores are used as the evaluation metric.
 
-This paper presents an overview of the most essential methods in image captioning and some of our results.
+\section{Data}\label{sec:data}
+We focus on the flickr8k dataset \cite{hodosh2013flickr8k}, a pivotal resource in image captioning.
+Comprising roughly 8.000 diverse images, each with five reference captions, this dataset is renowned for its quality and is a common choice for image captioning research due to its manageable size.
 
-Image captioning is a complex and interdisciplinary task bridging computer vision (CV) and natural language processing (NLP).
-The objective is to generate natural, descriptive sentences for images, which involves detecting and interpreting visual elements and their relationships within scenes.
-Recent advancements leverage deep learning, particularly Convolutional Neural Networks (CNNs) for feature extraction and Recurrent Neural Networks (RNNs), especially Long Short-Term Memory (LSTM) networks, for sequential language generation.
-For instance, models such as "Show and Tell" \cite{Vinyals:2015} and "Show, Attend, and Tell" \cite{Xu} utilize CNN-RNN architectures, with the latter integrating attention mechanisms to enhance focus on specific image regions, thereby improving caption relevance and accuracy.
-
-Image captioning models are typically evaluated using metrics like BLEU, METEOR, and CIDEr, which assess linguistic similarity to human references.
-While these metrics are widely accepted, considering the quality of image captions remains challenging due to the subjective nature of language generation.
-Additionally, methods like Reinforcement Learning (RL) have been explored to optimize these models directly on evaluation metrics, addressing issues such as exposure bias by aligning training and inference methods more closely \cite{Rennie}.
-
-Recent innovations in diverse decoding strategies, like Diverse Beam Search, aim to overcome the tendency of beam search to produce near-duplicate captions by enhancing variability without sacrificing quality.
-Diverse Beam Search ensures that generated captions capture a broader range of plausible descriptions, reflecting the inherent ambiguity in image captioning tasks \cite{Vijayakumar}.
-
-\section{Methods}
-\label{sec:methods}
-
-We present the dataset, the encoder-decoder architecture, the encoder models, the decoder models, and the evaluation metrics.
-
-\subsection{Dataset}
 \begin{figure}[H]
     \centering
-    \includegraphics[width=0.8\textwidth]{src/flickr8k.png}
+    \includegraphics[width=.9\textwidth]{res/flickr8k.png}
     \caption{Example images and captions from the flickr8k dataset.}
@@
 \end{figure}
-We focused on the flickr8k dataset, which consists of 8,000 images, each with five reference captions.
-The images are diverse and contain a wide range of objects and scenes.
-This dataset is commonly used for image captioning research due to its quality and limited size.
-The captions are relatively short and straightforward, making them easier to generate.
-\par Other commonly used datasets are flickr30k and MS COCO, but these datasets are larger and cannot be processed due to our limited computational resources.
 
-\subsection{Encoder-Decoder Architecture}
+\subsection{Splitting the Data}\label{sec:splitting}
+A predefined fixed split of the dataset into training, validation, and test sets is used. These datasets are stored in separate \textit{CSV} files.
+\par The training set contains 60\% of the data, the validation set 20\%, and the test set 20\%. This split is fixed to ensure reproducibility and comparability of the results.
+
+\subsection{Preprocessing}\label{sec:preprocessing}
+The models require the images and the captions to be in a suitable format.
+
+\subsubsection{Image Encoder}\label{sec:image-encoder}
+Contrary to the usual approach, the encoding is applied as part of the preprocessing to save computational resources and simplify the model.
+\par EfficientNet is used over other models due to its high performance compared to ResNet, for example.
+The model is pre-trained on the ImageNet dataset and used to extract features from the images.
+We use the B0 variant because it strikes a balance between computational efficiency and accuracy \cite{tan2019efficientnet}.
+
+\subsubsection{Images}\label{sec:images}
+The raw \textit{JPEG} images are loaded using the \textbf{PIL} library and preprocessed as follows:
+\begin{enumerate}
+    \item Resize the images to 224x224 pixels.
+    \item Normalize the pixel values to the range [0, 1].
+    \item Encode the images as tensors of the dimension (224, 224, 3).
+    \item Use a pre-trained \texttt{EfficientNet B0} model to extract features from the images resulting in a tensor of the dimension (1280).
+\end{enumerate}
+
+\subsubsection{Vocabulary}\label{sec:vocabulary}
+\begin{quote}\center Configurable: \texttt{SIZE}, \texttt{THRESHOLD} \end{quote}
+To encode the captions, a vocabulary of words is created from the captions and each word is mapped to an index.
+Special tokens are added for padding, unknown tokens, start, and end of the sequence.
+The vocabulary size is limited to \texttt{SIZE} and words with a frequency below \texttt{THRESHOLD} are replaced by the unknown token.
+
+\begin{table}[H]
+    \begin{tabular}{c|cccccccc}
+        \textbf{Word} & $<$pad$>$ & $<$unknown$>$ & $<$start$>$ & $<$end$>$ & is & and & dog & with \\
+        \textbf{Index} & 0 & 1 & 2 & 3 & 4 & 5 & 6 & 7
+    \end{tabular}
+    \caption{Example vocabulary with special tokens. \label{tab:vocabulary}}
+\end{table}
+
+\subsubsection{Captions}\label{sec:captions}
+\begin{quote}\center Configurable: \texttt{CAPTION\_LEN}\end{quote}
+The raw captions are loaded from the \textit{CSV} file and preprocessed as follows:
+\begin{enumerate}
+    \item Convert the captions to lowercase.
+    \item Tokenize the captions.
+    \item Encode the captions using our Vocabulary \ref{sec:vocabulary}.
+    \item Pad or truncate the captions to a fixed length \texttt{CAPTION\_LEN}.
+    \item Convert the captions to one-dimensional tensors with a specified \texttt{CAPTION\_LEN}.
+\end{enumerate}
+
+\subsection{Dataset and Dataloader}\label{sec:dataset}
+To use the data efficiently within the model, a custom \textbf{PyTorch}-based Dataset with \textbf{PyTorch}'s DataLoader is used
+\par The Dataset loads the images and captions and returns them as a tuple.
+\par The DataLoader then batches the data and, for training data, shuffles the batches.
+
+\section{Models}\label{sec:models}
+\begin{quote}\center Configurable: \texttt{HIDDEN\_DIM}, \texttt{EMBEDDING\_DIM}, \texttt{NUM\_LAYERS}, \texttt{DROPOUT}\end{quote}
+In our paper, different decoder models are used for experiments, while the encoder (refer to \ref{sec:image-encoder}) stays the same.
+Two RNNs, GRU and LSTM, and a Transformer (decoder) for decoding the captions are compared.
+\par We use the \textbf{PyTorch} library to implement the models.
+
+\subsection{GRU}\label{sec:gru}
+Gated recurrent units are a popular simplified alternative to LSTMs \cite{cho2014gru}. The GRU model architecture is shown in Figure \ref{fig:gru}.
+\par The model predicts each token in sequence. The hidden state is updated at each time step and used to predict the next token.
 \begin{figure}[H]
     \centering
-    \includegraphics[width=0.8\textwidth]{src/encoder-decoder.png}
-    \caption{A Survey on Various Deep Learning Models for Automatic Image Captioning \cite{gaurav}}
-    \label{fig:encoder-decoder}
+    \includegraphics[width=.9\textwidth]{res/gru.png}
+    \caption{The GRU model architecture.}\label{fig:gru}
 \end{figure}
-Usually, an encoder-decoder architecture is used to achieve image captioning.
-The encoder is a convolutional neural network (CNN) that extracts features from the image.
-The decoder is a recurrent neural network (RNN) that generates the caption word by word.
 
-\subsection{Encoder Models}
-We experimented with two different models for encoding images: ResNet-50 and VGG-16.
-\par ResNet-50 is a deep residual network that is widely used for image classification.
-It has 50 layers and uses skip connections to avoid the vanishing gradient problem.
-It was introduced in \cite{he}.
-\par VGG-16 is a deep convolutional neural network that was developed by the Visual Geometry Group at the University of Oxford.
-It has 16 layers and is known for its simplicity and effectiveness.
-It was introduced in \cite{simonyan}.
+\subsection{LSTM}\label{sec:lstm}
+Long Short-Term Memory networks are a type of RNN that can learn long-term dependencies \cite{hochreiter1997lstm}. The LSTM model architecture is shown in Figure \ref{fig:lstm}.
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=.9\textwidth]{res/lstm.png}
+    \caption{The LSTM model architecture.}\label{fig:lstm}
+\end{figure}
 
-\subsection{Decoder Models}
-We experimented with two different models for decoding captions: GRU and LSTM.
-\par Gated Recurrent Units (GRUs) are a type of RNN that is similar to LSTMs but has fewer parameters. They are easier to train and faster to compute. They were introduced in \cite{cho}.
-\par Long Short-Term Memory (LSTM) networks are a type of RNN that is designed to avoid the vanishing gradient problem.
-They have a cell state that can store information over long sequences.
-They were introduced in \cite{hochreiter}.
+\subsection{Transformer}\label{sec:transformer}
+The Transformer model architecture is shown in Figure \ref{fig:transformer}. The model is based on the original paper by Vaswani et al. \cite{vaswani2017attention}.
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=.9\textwidth]{res/transformer.png}
+    \caption{The Transformer model architecture.}\label{fig:transformer}
+\end{figure}
 
-\subsection{Evaluation Metrics}
-The accuracy of image captioning is commonly measured using BLEU or CIDEr scores.
-\par BLEU is a precision-based metric that compares n-grams in the generated caption to those in the reference captions.
-It ranges from 0 to 1, with higher values indicating better performance.
-It was introduced in \cite{papineni}.
-\par CIDEr is a recall-based metric that compares the generated caption to the reference captions using cosine similarity.
-It ranges from 0 to 1, with higher values indicating better performance.
-It was introduced in \cite{vedantam}.
-\par For simplicity, we are only using the BLEU score in our experiments.
+\section{Experiments}\label{sec:experiments}
+The research question we are trying to answer is: \textit{Which decoder model performs best for the task of image captioning?}
 
-\section{Approach}
-\label{sec:approach}
+\subsection{Training}\label{sec:training}
+\begin{quote}\center Configurable: \texttt{EPOCHS}, \texttt{PATIENCE}, \texttt{BATCH\_SIZE}, \texttt{LEARNING\_RATE}\end{quote}
+\par The models are trained using the Adam optimizer and the CrossEntropyLoss function.
+\par The models are trained for a number of epochs \texttt{EPOCHS}.
+If the \texttt{PATIENCE} hyperparameter is set and the validation loss does not improve for \texttt{PATIENCE} epochs, the training is stopped early.
+Batching makes training more efficient and can be adjusted with the \texttt{BATCH\_SIZE} hyperparameter.
 
-Our approach begins with data preprocessing, followed by model training and evaluation.
+\subsection{Evaluation metrics}\label{sec:evaluation-metrics}
+To evaluate the models, BLEU, METEOR, and NIST scores are used. These are common metrics used in machine translation and image captioning tasks.
+\begin{enumerate}
+    \item BLEU (Bilingual Evaluation Understudy) is a metric that compares the generated captions to the reference captions.
+It measures the n-gram overlap between the generated and reference captions \cite{papineni2002bleu}.
+    \item METEOR (Metric for Evaluation of Translation with Explicit ORdering) is a metric that considers the generated and reference captions as sequences of words.
+It uses a harmonic mean of precision and recall \cite{banerjee2005meteor}.
+    \item NIST (Normalized Information Retrieval Metric) is a metric that rewards longer n-grams more than shorter ones. It is based on the BLEU metric \cite{doddington2002nist}.
+\end{enumerate}
 
-\subsection{Data Preprocessing}
-\par We preprocess the images by resizing them to 224x224 pixels and normalizing the pixel values and encoding them to tensors of the dimension (224, 224, 3).
-\par We create a vocabulary of words and map each word to an index.
-\par We preprocess the captions by tokenizing them, converting them to lowercase, and padding them to a fixed length.
-Then we encode them using the vocabulary and convert them to one-dimensional tensors with a specified max\_length.
+\subsection{Results}\label{sec:results}
+Our experiments use the hyperparameters shown in Table \ref{tab:hyperparameters}.
+\begin{table}[H]
+    \center
+    \begin{tabular}{r|l}
+        \texttt{CAPTION\_LEN} & 20 \\
+        \texttt{SIZE} & 8000 \\
+        \texttt{THRESHOLD} & 2 \\
+        \texttt{HIDDEN\_DIM} & 512 \\
+        \texttt{EMBEDDING\_DIM} & 256 \\
+        \texttt{NUM\_LAYERS} & 2 \\
+        \texttt{DROPOUT} & .1 \\
+        \texttt{EPOCHS} & 100 \\
+        \texttt{PATIENCE} & None \\
+        \texttt{BATCH\_SIZE} & 64 \\
+        \texttt{LEARNING\_RATE} & 1e-3
+    \end{tabular}
+    \caption{Hyperparameters used in the experiments.}\label{tab:hyperparameters}
+\end{table}
 
-\subsection{Model Training}
+The training and validation loss for the different models are shown in Figure \ref{fig:loss}.
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=.3\textwidth]{res/training-model-gru.png}
+    \includegraphics[width=.3\textwidth]{res/training-model-lstm.png}
+    \includegraphics[width=.3\textwidth]{res/training-model-transformer.png}
+    \caption{Training and validation loss for the different models.}\label{fig:loss}
+\end{figure}
+Training the models took approximately 20 hours per model on a single GPU.
+% TODO: interpret
 
-\subsection{Model Evaluation}
+The results of the evaluation metrics are shown in Table \ref{tab:results} and Figure \ref{fig:results}.
+\begin{table}[H]
+    \center
+    \begin{tabular}{c|c|c|c}
+        \textbf{Model} & \textbf{BLEU} & \textbf{METEOR} & \textbf{NIST} \\
+        GRU &&& \\
+        LSTM &&& \\
+        Transformer &&&
+    \end{tabular}
+    \caption{Results of the evaluation metrics for the different models.}\label{tab:results}
+\end{table}
 
-\section{Results}
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=.9\textwidth]{res/metrics.png}
+    \caption{Results of the evaluation metrics for the different models.}\label{fig:results}
+\end{figure}
+% TODO: insert and interpret results
 
-\section{Results}
-\label{sec:results}
+\section{Conclusion}\label{sec:conclusion}
 
-Comparison of different models, evaluation, and discussion.
-
-\section{Conclusion}
-\label{sec:concl}
-
-Summary of the results, future work, and open questions.
+This paper presents our experiments with different decoder models for the task of image captioning. The flickr8k dataset is used and the models are evaluated using BLEU, METEOR, and NIST scores.
+\par Our results show [...] % TODO
+\par In the future, further experiments should be conducted with
+\begin{enumerate}
+    \item other encoder models, like ResNet or VGG.
+    \item varying the batch size or learning rate to optimize the training process.
+    \item different datasets, like flickr30k and MS COCO.
+\end{enumerate}
+These can build on the code provided in this paper and extend the research on image captioning.
 
@@
 
-\nocite{*}
-\bibliographystyle{apalike}
+\bibliographystyle{acm}
 \bibliography{bib}

### 2024-12-18 BAI/paper/paper.tex
Updated paper.

@@
 
+\setlength{\footskip}{15pt}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
     The models are implemented in \textbf{PyTorch}, and their performance is evaluated using standard metrics: \textbf{BLEU}, \textbf{METEOR}, and \textbf{NIST}.
-    % TODO: results
+    The results show that the Transformer model outperforms the GRU and LSTM models in all metrics.
 \end{abstract}
 
+\setcounter{tocdepth}{1}
 \tableofcontents
@@
 A predefined fixed split of the dataset into training, validation, and test sets is used. These datasets are stored in separate \textit{CSV} files.
-\par The training set contains 60\% of the data, the validation set 20\%, and the test set 20\%. This split is fixed to ensure reproducibility and comparability of the results.
+\par The training set contains 80\% of the data, the validation set 10\%, and the test set 10\%. This split is fixed to ensure reproducibility and comparability of the results.
 
@@
 \par The Dataset loads the images and captions and returns them as a tuple.
-\par The DataLoader then batches the data and, for training data, shuffles the batches.
+\par The DataLoader then batches the data.
 
@@
 Gated recurrent units are a popular simplified alternative to LSTMs \cite{cho2014gru}. The GRU model architecture is shown in Figure \ref{fig:gru}.
-\par The model predicts each token in sequence. The hidden state is updated at each time step and used to predict the next token.
+\par The model predicts each token in sequence.
+The hidden state is updated at each time step and used to predict the next token.
+The model is trained using teacher forcing, where the true target token is used as input to predict the next token.
 \begin{figure}[H]
     \centering
-    \includegraphics[width=.9\textwidth]{res/gru.png}
+    \includegraphics[width=.7\textwidth]{res/gru.png}
     \caption{The GRU model architecture.}\label{fig:gru}
@@
 Long Short-Term Memory networks are a type of RNN that can learn long-term dependencies \cite{hochreiter1997lstm}. The LSTM model architecture is shown in Figure \ref{fig:lstm}.
+\par Like the GRU model, the LSTM model predicts each token in sequence. The hidden state is updated at each time step and used to predict the next token. The model is trained using teacher forcing.
 \begin{figure}[H]
     \centering
-    \includegraphics[width=.9\textwidth]{res/lstm.png}
+    \includegraphics[width=.7\textwidth]{res/lstm.png}
     \caption{The LSTM model architecture.}\label{fig:lstm}
@@
 \subsection{Transformer}\label{sec:transformer}
-The Transformer model architecture is shown in Figure \ref{fig:transformer}. The model is based on the original paper by Vaswani et al. \cite{vaswani2017attention}.
+The Transformer model architecture is shown in Figure \ref{fig:transformer}. The model is based on the original paper \cite{vaswani2017attention}.
+\par Contrary to the RNNs, the Transformer predicts the whole sequence at once.
+The model uses self-attention mechanisms to capture dependencies between words in the sequence.
+The model is trained using teacher forcing.
 \begin{figure}[H]
     \centering
-    \includegraphics[width=.9\textwidth]{res/transformer.png}
+    \includegraphics[width=.7\textwidth]{res/transformer.png}
     \caption{The Transformer model architecture.}\label{fig:transformer}
@@
 It measures the n-gram overlap between the generated and reference captions \cite{papineni2002bleu}.
+We use the BLEU-4 score, meaning up to 4-grams are considered.
     \item METEOR (Metric for Evaluation of Translation with Explicit ORdering) is a metric that considers the generated and reference captions as sequences of words.
 It uses a harmonic mean of precision and recall \cite{banerjee2005meteor}.
-    \item NIST (Normalized Information Retrieval Metric) is a metric that rewards longer n-grams more than shorter ones. It is based on the BLEU metric \cite{doddington2002nist}.
+    \item NIST (Normalized Information Retrieval Metric) is a metric that rewards longer n-grams more than shorter ones.
+It is based on the BLEU metric \cite{doddington2002nist}.
+We use the NIST-4 score, meaning up to 4-grams are considered.
 \end{enumerate}
@@
     \center
+    \fontsize{8}{10}\selectfont
     \begin{tabular}{r|l}
@@
     \centering
-    \includegraphics[width=.3\textwidth]{res/training-model-gru.png}
-    \includegraphics[width=.3\textwidth]{res/training-model-lstm.png}
-    \includegraphics[width=.3\textwidth]{res/training-model-transformer.png}
+    \includegraphics[width=.3\textwidth]{res/training-gru.png}
+    \includegraphics[width=.3\textwidth]{res/training-lstm.png}
+    \includegraphics[width=.3\textwidth]{res/training-transformer.png}
     \caption{Training and validation loss for the different models.}\label{fig:loss}
 \end{figure}
-Training the models took approximately 20 hours per model on a single GPU.
-% TODO: interpret
+Training the models took approximately 20 hours per model on a single GPU. The validation loss is similar for all models, but the LSTM model converges faster than the GRU and Transformer models.
 
 The results of the evaluation metrics are shown in Table \ref{tab:results} and Figure \ref{fig:results}.
-\begin{table}[H]
-    \center
-    \begin{tabular}{c|c|c|c}
-        \textbf{Model} & \textbf{BLEU} & \textbf{METEOR} & \textbf{NIST} \\
-        GRU &&& \\
-        LSTM &&& \\
-        Transformer &&&
-    \end{tabular}
-    \caption{Results of the evaluation metrics for the different models.}\label{tab:results}
-\end{table}
 
@@
     \centering
+    \begin{minipage}{0.4\textwidth}
+        \begin{table}[H]
+            \centering
+            \fontsize{8}{10}\selectfont
+            \begin{tabular}{c|c|c|c}
+                & \textbf{BLEU-4} & \textbf{METEOR} & \textbf{NIST-4} \\
+                GRU & & & \\
+                LSTM & & & \\
+                Transformer & & &
+            \end{tabular}
+            \caption{Evaluation metrics.}\label{tab:results}
+        \end{table}
+    \end{minipage}
+    \hfill
+    \begin{minipage}{0.5\textwidth}
+        \centering
         \includegraphics[width=.9\textwidth]{res/metrics.png}
-    \caption{Results of the evaluation metrics for the different models.}\label{fig:results}
+        \caption{Evaluation metrics.}\label{fig:results}
+    \end{minipage}
 \end{figure}
-% TODO: insert and interpret results
+
+These results show that the Tableransformer model outperforms the other models in all metrics. The LSTM model performs better than the GRU model, but worse than the Transformer model.
+\par Compared to other research, our results are much worse. \cite{vinyals2015show} achieved a BLEU score of  27.7, for example, and \cite{xu2015show} achieved a METEOR score of 24.3.
+\par This has several reasons: \begin{enumerate}
+    \item \textbf{Smaller Dataset}: The flickr8k dataset is smaller than other datasets like MS COCO, which is used in \cite{vinyals2015show} and \cite{xu2015show}.
+    \item \textbf{Encoder}: The EfficientNet B0 model is less powerful than other models like ResNet, for example \cite{he2016deep}.
+    \item \textbf{No attention mechanisms}: The models used in this paper do not use attention mechanisms, which are crucial for image captioning \cite{xu2015show}
+    \item \textbf{Limited training time}: The models were trained for a limited number of epochs due to computational constraints.
+\end{enumerate}
+\par Nonetheless, we were able to show the comparative advantage of Transformers over RNNs for image captioning.
 
@@
 This paper presents our experiments with different decoder models for the task of image captioning. The flickr8k dataset is used and the models are evaluated using BLEU, METEOR, and NIST scores.
-\par Our results show [...] % TODO
+\par Our results show Transformers outperform RNNs for image captioning.
 \par In the future, further experiments should be conducted with
@@
 
+\fontsize{12}{14}\selectfont
 \bibliographystyle{acm}

### 2024-12-19 NLP/exercises.tex
Moved one exercise to 2021 exam to balance the two exams.
Shuffled the exercises to get three comparable exams.

@@
 
-\sheet{Lecture questions}
-
-\begin{exercise}{NLP and the Web}
-  \begin{enumerate}
-    \item The web is an application area for NLP. Provide examples of applications.
-    \item How can the web improve the quality of NLP? Provide examples.
-  \end{enumerate}
-
-  \begin{solution}
-    \begin{enumerate}
-      \item Information extraction, sentiment analysis, machine translation, question answering, etc.
-      \item Web as a corpus, analyzing web-based knowledge repositories, extracting structured information, using web-scale data for training language models.
-    \end{enumerate}
-  \end{solution}
-\end{exercise}
-
-\begin{exercise}{Datasets}
-  \begin{enumerate}
-    \item What datasets are used in NLP research, and why are they important?
-    \item What is the annotation process for a corpus or dataset, and how does it impact machine learning applications?
-    \item What are the main strategies for large-scale dataset collection using crowdsourcing?
-  \end{enumerate}
-
-  \begin{solution}
-    \begin{enumerate}
-      \item NLP datasets are collections of annotated or unannotated text data designed for specific NLP tasks (e.g., sentiment analysis, translation).
-They serve as a foundation for training and testing machine learning models, enable reproducibility in research, and facilitate comparisons between models and techniques.
-      \item Annotation process involves adding labels or metadata to text data (e.g., part-of-speech tags, sentiment labels).
-It improves model performance by providing high-quality training data, enables supervised learning by associating input with output labels, and impacts scalability and generalization of models depending on annotation quality.
-      \item Strategies include task design (clearly define annotation tasks with examples and guidelines), platform selection (use platforms like Amazon Mechanical Turk or Prolific), quality control (use redundancy and include test questions to ensure reliability), and data aggregation (combine annotations using majority voting or statistical models).
-    \end{enumerate}
-  \end{solution}
-\end{exercise}
-
-\begin{exercise}{Text preprocessing}
-  \begin{enumerate}
-    \item Why are more features not always better for learning a classifier?
-    \item What is the typical type of input data that a sequence tagger training step requires?
-  \end{enumerate}
-
-  \begin{solution}
-    \begin{enumerate}
-      \item More features can lead to overfitting, require more training data, introduce noise, and slow down training and inference.
-      \item Labeled training text (e.g., part-of-speech tags, named entity recognition).
-    \end{enumerate}
-  \end{solution}
-\end{exercise}
-
-\setcounter{section}{2022}
-\sheet{1st exam}
+\setcounter{section}{2022}\sheet{1st exam}
 
@@
 
-\begin{exercise}{Pragmatics and discourse analysis}\label{ex:pragmatics}
-  Why are pragmatics and discourse analysis difficult?
-
-  \begin{solution}
-    \begin{itemize}
-      \item \textbf{Context dependency}: Understanding depends heavily on prior sentences or external knowledge.
-      \item \textbf{Ambiguity}: Words and phrases can have multiple meanings depending on the situation.
-      \item \textbf{Social norms}: Interpretations vary based on cultural or societal factors.
-    \end{itemize}
-  \end{solution}
-\end{exercise}
-
 \begin{exercise}{Hate speech}\label{ex:hate-speech}
@@
 
+\begin{exercise}{Question answering}\label{ex:qa}
+  Question Answering (QA) systems aim to provide direct answers to user queries.
+  \begin{enumerate}
+    \item What are the differences between factual QA and opinionated QA? and which is easier to evaluate?
+    \item Briefly explain QALD. What are key challenges?
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+      \item \begin{itemize}
+          \item \textbf{Factual QA}: Answers objective questions with verifiable information.
+          \item \textbf{Opinionated QA}: Provides subjective answers based on opinions or preferences.
+          \item Factual QA is easier to evaluate as answers can be verified against ground truth.
+        \end{itemize}
+      \item QALD (Question Answering over Linked Data) focuses on answering questions using linked data (e.g., RDF datasets).
+Challenges include mapping natural language queries to structured data, handling incomplete or ambiguous data, and scaling to large and heterogeneous sources.
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Cross-language QA}\label{ex:cross-lang-qa}
+  Cross-language Question Answering (QA) is a challenging task.
+  \begin{enumerate}
+    \item Explain the process of cross-language question-answering.
+    \item Where is it applied?
+    \item Explain the challenges faced in cross-language QA.
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+        \item Translate the question, retrieve relevant information in the target language, generate an answer, and translate it back.
+        \item Multi-language QA systems are used in global customer support, cross-border e-commerce, and multi-language search engines.
+        \item Challenges include translation errors, lack of parallel resources, cultural differences, and mismatched linguistic structures.
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+\setcounter{section}{2021}\sheet{1st exam}
+
+\begin{exercise}{NLP and the Web}
+  \begin{enumerate}
+    \item The web is an application area for NLP. Provide at least three examples of applications.
+    \item How can the web improve the quality of NLP? Provide at least three examples.
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+      \item Information extraction, sentiment analysis, machine translation, question answering, …
+      \item Web as a corpus, crowdsourcing annotations, multilingual data, …
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Datasets}
+  \begin{enumerate}
+    \item What datasets are used in NLP research, and why are they important?
+    \item What is the annotation process for a corpus or dataset, and how does it impact machine learning applications?
+    \item What are the main strategies for large-scale dataset collection using crowdsourcing?
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+      \item Collections of (possibly annotated) text data used for training and evaluating NLP models. They are crucial for developing and testing algorithms.
+      \item Add metadata or linguistic labels to the text, which can be used for training supervised machine learning models.
+The quality of annotations directly impacts the performance of machine learning applications.
+      \item Task design (clear instructions, examples), platform selection (e.g., Amazon Mechanical Turk), quality control (test questions, redundancy), data aggregation (majority voting, adjudication), and evaluation (inter-annotator agreement, error analysis).
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Text preprocessing}
+  \begin{enumerate}
+    \item Why are more features not always better for learning a classifier?
+    \item What is the typical type of input data that a sequence tagger training step requires?
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+      \item Overfitting, noise, and computational complexity can increase with the number of features, leading to poor generalization.
+      \item Labeled training text (e.g., part-of-speech tags, named entity recognition).
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
 \begin{exercise}{Information retrieval metrics}\label{ex:ir-metrics}
@@
 
-\begin{exercise}{Question answering}\label{ex:qa}
-  Question Answering (QA) systems aim to provide direct answers to user queries.
-  \begin{enumerate}
-    \item What are the differences between factual QA and opinionated QA? and which is easier to evaluate?
-    \item Briefly explain QALD. What are key challenges?
-  \end{enumerate}
+\begin{exercise}{NLTK}
+  Explain two parameters in NLTK that start with a "c".
 
   \begin{solution}
-    \begin{enumerate}
-      \item \begin{itemize}
-          \item \textbf{Factual QA}: Answers objective questions with verifiable information.
-          \item \textbf{Opinionated QA}: Provides subjective answers based on opinions or preferences.
-          \item Factual QA is easier to evaluate as answers can be verified against ground truth.
+    \begin{itemize}
+      \item \textbf{Chunking}: Identifies and groups words into syntactic chunks based on their POS tags.
+      \item \textbf{Collocations}: Identifies words that frequently co-occur within a certain span.
+      \item \textbf{Concordance}: Displays a word in context, showing its surrounding words within a specified window.
     \end{itemize}
-      \item QALD (Question Answering over Linked Data) focuses on answering questions using linked data (e.g., RDF datasets).
-Challenges include mapping natural language queries to structured data, handling incomplete or ambiguous data, and scaling to large and heterogeneous sources.
-    \end{enumerate}
   \end{solution}
@@
 
-\begin{exercise}{Cross-language QA}\label{ex:cross-lang-qa}
-  Cross-language Question Answering (QA) is a challenging task.
-  \begin{enumerate}
-    \item Explain the process of cross-language question-answering.
-    \item Where is it applied?
-    \item Explain the challenges faced in cross-language QA.
-  \end{enumerate}
+\begin{exercise}{Recall improvement}
+  Explain how to improve recall when using a search engine without changing the algorithm.
 
   \begin{solution}
-    \begin{enumerate}
-        \item Translate the question, retrieve relevant information in the target language, generate an answer, and translate it back.
-        \item Multi-language QA systems are used in global customer support, cross-border e-commerce, and multi-language search engines.
-        \item Challenges include translation errors, lack of parallel resources, cultural differences, and mismatched linguistic structures.
-    \end{enumerate}
+    \begin{itemize}
+        \item \textbf{Expanding queries}: Use synonyms, related terms, or stemming to match more results.
+        \item \textbf{Reducing filters}: Remove restrictive criteria (e.g., date, location).
+        \item \textbf{Boosting document recall}: Index more documents or reduce relevance thresholds.
+    \end{itemize}
   \end{solution}
@@
 
-\begin{exercise}{RAG and LangChain}\label{ex:rag}
-  What is Retrieval-Augmented Generation (RAG), and how does LangChain assist in implementing it?
+\begin{exercise}{Pragmatics and discourse analysis}\label{ex:pragmatics}
+  Why are pragmatics and discourse analysis difficult?
 
@@
     \begin{itemize}
-        \item \textbf{RAG (Retrieval-Augmented Generation)}: Combines retrieval models with generative models to produce context-aware answers.
-External knowledge is retrieved and fed into the generation model.
-        \item \textbf{LangChain}: A framework for building applications with large language models, offering tools for integrating RAG systems.
-It manages chaining retrieval and generation steps efficiently.
+      \item \textbf{Context dependency}: Understanding depends heavily on prior sentences or external knowledge.
+      \item \textbf{Ambiguity}: Words and phrases can have multiple meanings depending on the situation.
+      \item \textbf{Social norms}: Interpretations vary based on cultural or societal factors.
     \end{itemize}
@@
 
-\setcounter{section}{2020}
-\sheet{1st exam}
-
-\begin{exercise}{Comprehensiveness}
-  What is comprehensiveness, specifically in the context of NLP?
+\begin{exercise}{TREC}\label{ex:trec}
+  What are the TREC question types?
 
   \begin{solution}
-    Comprehensiveness refers to the extent to which a system, dataset, or model covers all relevant aspects of a task or domain.
-    In NLP, comprehensiveness can be evaluated based on the coverage of linguistic phenomena, language varieties, or domain-specific terms in a dataset or model.
+    \begin{itemize}
+        \item \textbf{Factoid questions}: Require specific factual answers (e.g., "Who is the president of the U.S.?").
+        \item \textbf{List questions}: Require a list of items (e.g., "List countries in the EU").
+        \item \textbf{Definition questions}: Seek explanations or definitions (e.g., "What is AI?").
+    \end{itemize}
   \end{solution}
@@
     1. query formulation, 2. query classification, 3. indexing, 4. doc retrieval, 5. passage retrieval
+  \end{solution}
+\end{exercise}
+
+\setcounter{section}{2020}\sheet{1st exam}
+
+\begin{exercise}{Comprehensiveness}
+  What is comprehensiveness, specifically in the context of NLP?
+
+  \begin{solution}
+    Comprehensiveness refers to the extent to which a system, dataset, or model covers all relevant aspects of a task or domain.
+    In NLP, comprehensiveness can be evaluated based on the coverage of linguistic phenomena, language varieties, or domain-specific terms in a dataset or model.
   \end{solution}
@@
     \end{enumerate}
-  \end{solution}
-\end{exercise}
-
-\begin{exercise}{NLTK}
-  Explain two parameters in NLTK that start with a "c".
-
-  \begin{solution}
-    \begin{itemize}
-      \item \textbf{Chunking}: Identifies and groups words into syntactic chunks based on their POS tags.
-      \item \textbf{Collocations}: Identifies words that frequently co-occur within a certain span.
-      \item \textbf{Concordance}: Displays a word in context, showing its surrounding words within a specified window.
-    \end{itemize}
   \end{solution}
@@
   \input{res/qa-answers.fig}
-  Calculate MRR and Accuracy for the Question/Answering system in Figure \ref{fig:qa-answers}.
+  Calculate Mean Reciprocal Rank (MRR) and Accuracy for the given query results in Figure \ref{fig:qa-answers} at $k=5$.
 
@@
         \item \textbf{Accuracy} $\frac{\text{Number of correct answers}}{\text{Total number of queries}} = \frac{2}{3} = .6666$
-    \end{itemize}
-  \end{solution}
-\end{exercise}
-
-\begin{exercise}{Recall improvement}
-  Explain how to improve recall when using a search engine without changing the algorithm.
-
-  \begin{solution}
-    \begin{itemize}
-        \item \textbf{Expanding queries}: Use synonyms, related terms, or stemming to match more results.
-        \item \textbf{Reducing filters}: Remove restrictive criteria (e.g., date, location).
-        \item \textbf{Boosting document recall}: Index more documents or reduce relevance thresholds.
     \end{itemize}
@@
 
-\begin{exercise}{TREC}\label{ex:trec}
-  What are the TREC question types?
+\begin{exercise}{RAG and LangChain}\label{ex:rag}
+  What is Retrieval-Augmented Generation (RAG), and how does LangChain assist in implementing it?
 
@@
     \begin{itemize}
-        \item \textbf{Factoid questions}: Require specific factual answers (e.g., "Who is the president of the U.S.?").
-        \item \textbf{List questions}: Require a list of items (e.g., "List countries in the EU").
-        \item \textbf{Definition questions}: Seek explanations or definitions (e.g., "What is AI?").
+        \item \textbf{RAG (Retrieval-Augmented Generation)}: Combines retrieval models with generative models to produce context-aware answers.
+External knowledge is retrieved and fed into the generation model.
+        \item \textbf{LangChain}: A framework for building applications with large language models, offering tools for integrating RAG systems.
+It manages chaining retrieval and generation steps efficiently.
     \end{itemize}

### 2024-12-19 NLP/exercises.tex
Minor updates.

@@
 
-\subject{Natural Language Processsing}
+\subject{Natural Language Processing}
 \semester{Winter 2024}
@@
 
-\setcounter{section}{2022}\sheet{1st exam}
+\setcounter{section}{2022}\sheet{Exam}
 
@@
 
-\setcounter{section}{2021}\sheet{1st exam}
+\setcounter{section}{2021}\sheet{Exam}
 
@@
   \begin{itemize}
-    \item \textbf{Engine A}: $+, -, +, -, +, +, -, +, +, +$
-    \item \textbf{Engine B}: $+, +, -, +, +, -, +, +, -, +$
+    \item \textbf{Engine A}: $+-+-+\quad+-+++$
+    \item \textbf{Engine B}: $---+-\quad-++-+$
   \end{itemize}
@@
   \begin{enumerate}
-    \item Calculate P@5, P@10 and MAP for both engines.
+    \item Calculate P@5, P@10 and MAP@5 for both engines.
     \item How do these metrics help evaluate the performance of IR systems?
@@
             \begin{itemize}
-            \item $P@5 = \frac{3}{5} = 0.6$
-            \item $P@10 = \frac{7}{10} = 0.7$
-            \item $MAP = \frac{1}{3} \left( \frac{1}{1} + \frac{2}{4} + \frac{3}{5} \right) = 0.5666$
+              \item $P@5 = \frac{3}{5} = .6$
+              \item $P@10 = \frac{7}{10} = .7$
+              \item $MA@5 = (\frac{1}{1} + \frac{2}{3} + \frac{3}{5}) / 3 = .756$
             \end{itemize}
@@
             \begin{itemize}
-            \item $P@5 = \frac{4}{5} = 0.8$
-            \item $P@10 = \frac{7}{10} = 0.7$
-            \item $MAP = \frac{1}{3} \left( \frac{1}{1} + \frac{2}{4} + \frac{3}{7} \right) = 0.5476$
+              \item $P@5 = \frac{1}{5} = .2$
+              \item $P@10 = \frac{4}{10} = .4$
+              \item $MAP@5 = (\frac{1}{4}) / 1 = .25$
             \end{itemize}
@@
           \item \textbf{Lexicon-based methods}: Use dictionaries to match input text to language-specific words.
-          \item \textbf{Hybrid approaches}: Combine n-gram, lexicon, and machine learning techniques for better accuracy.
         \end{itemize}
@@
           \item \textbf{Semantic similarity}: Apply embeddings to detect paraphrased or rephrased content.
-          \item \textbf{Citation analysis}: Check for missing or incorrect citations.
           \item \textbf{Fingerprinting}: Generate hashes of text fragments and compare across sources.
@@
     \begin{itemize}
-        \item \textbf{Expanding queries}: Use synonyms, related terms, or stemming to match more results.
         \item \textbf{Reducing filters}: Remove restrictive criteria (e.g., date, location).
         \item \textbf{Boosting document recall}: Index more documents or reduce relevance thresholds.
+
     \end{itemize}
@@
             \item \textbf{Customer feedback}: Analyze reviews for product improvements.
-            \item \textbf{Social media monitoring}: Detect trends and public sentiment.
             \item \textbf{Political analysis}: Gauge public opinion on policies or events.
+            \item \textbf{Stock market prediction}: Predict stock prices based on sentiment in financial news.
           \end{itemize}
@@
         \item \textbf{Factoid questions}: Require specific factual answers (e.g., "Who is the president of the U.S.?").
-        \item \textbf{List questions}: Require a list of items (e.g., "List countries in the EU").
+        \item \textbf{List questions}: Require a list of items (e.g., "What countries are in the EU?").
         \item \textbf{Definition questions}: Seek explanations or definitions (e.g., "What is AI?").
@@
 
-\setcounter{section}{2020}\sheet{1st exam}
+\setcounter{section}{2020}\sheet{Exam}
 

### 2024-12-19 BAI/paper/paper.tex
Paper is ready and presentation is fine.

@@
     The models are implemented in \textbf{PyTorch}, and their performance is evaluated using standard metrics: \textbf{BLEU}, \textbf{METEOR}, and \textbf{NIST}.
-    The results show that the Transformer model outperforms the GRU and LSTM models in all metrics.
+    Unexpectedly, the Transformer model underperforms compared to the LSTM and GRU models across all metrics.
+This underperformance may stem from the limited training time and computational resources, as well as the Transformer's unique architecture.
 \end{abstract}
@@
 
-Image captioning involves generating textual descriptions of given images.
-It is a complex problem because it combines the visual and textual domains, thus bridging computer vision (CV) and natural language processing (NLP).
-Generating natural, descriptive sentences for images involves detecting and interpreting visual elements and their relationships within depicted scenes.
-\par While deep learning with attention mechanisms has long since been the dominant approach in image captioning \cite{xu2015show}, recent advances have been brought about by transformers \cite{vaswani2017attention}.
-These models have shown remarkable performance in various natural language processing tasks, including image captioning.
+Image captioning, the task of generating descriptive textual captions for given images, represents a significant challenge at the intersection of computer vision (CV) and natural language processing (NLP).
+This task requires models to accurately detect and interpret visual elements within an image while synthesizing coherent and contextually appropriate textual descriptions.
+Such a multifaceted problem has broad applications, including assistive technologies for visually impaired individuals, automated metadata generation, and content-based image retrieval.
+\par Deep learning approaches, particularly those employing attention mechanisms, have dominated image captioning research in recent years \cite{xu2015show}.
+These methods leverage the ability of neural networks to focus on relevant parts of an image while generating text.
+Building upon this foundation, Transformer-based architectures \cite{vaswani2017attention} have recently emerged as state-of-the-art in both NLP and CV tasks, owing to their ability to model complex dependencies and process sequences in parallel.
 \par This paper aims to compare the performance of different decoders directly, while keeping other parameters constant.
@@
 \end{figure}
-Training the models took approximately 20 hours per model on a single GPU. The validation loss is similar for all models, but the LSTM model converges faster than the GRU and Transformer models.
+Training the models took approximately 20 hours per model on a single GPU. The validation loss is similar for all models. The weird shape of the Transformer losses is due to interrupted training.
 
 The results of the evaluation metrics are shown in Table \ref{tab:results} and Figure \ref{fig:results}.
-
 \begin{figure}[H]
@@
                 & \textbf{BLEU-4} & \textbf{METEOR} & \textbf{NIST-4} \\
-                GRU & & & \\
-                LSTM & & & \\
-                Transformer & & &
+                GRU & .0511 & .1551 & 1.0202 \\
+                LSTM & .0541 & .1537 & .9137 \\
+                Transformer & .0183 & .0956 & .6013
             \end{tabular}
@@
 
-These results show that the Tableransformer model outperforms the other models in all metrics. The LSTM model performs better than the GRU model, but worse than the Transformer model.
+The results show that the LSTM and GRU models perform roughly the same, while the Transformer model performs worse in all metrics.
+This is contrary to expectations, as Transformers have shown remarkable performance in various NLP tasks.
+\par This is likely due to the limited training time and computational resources. Also, the Transformer predicts the whole sequence at once, which might be suboptimal for image captioning tasks.
 \par Compared to other research, our results are much worse. \cite{vinyals2015show} achieved a BLEU score of  27.7, for example, and \cite{xu2015show} achieved a METEOR score of 24.3.
-\par This has several reasons: \begin{enumerate}
+\par This might have several reasons: \begin{enumerate}
     \item \textbf{Smaller Dataset}: The flickr8k dataset is smaller than other datasets like MS COCO, which is used in \cite{vinyals2015show} and \cite{xu2015show}.
@@
 \end{enumerate}
-\par Nonetheless, we were able to show the comparative advantage of Transformers over RNNs for image captioning.
 
@@
 This paper presents our experiments with different decoder models for the task of image captioning. The flickr8k dataset is used and the models are evaluated using BLEU, METEOR, and NIST scores.
-\par Our results show Transformers outperform RNNs for image captioning.
+\par Our results show that the LSTM and GRU models outperform the Transformer model in all metrics given the limited training time and computational resources.
 \par In the future, further experiments should be conducted with

### 2024-12-19 BAI/paper/presentation.tex
Paper is ready and presentation is fine.

@@
 
-\begin{frame}
+% Slide 1: Introduction
+\begin{frame}{Introduction}
+	\textbf{What is Image Captioning?}
 	\begin{itemize}
-			\item \textbf{Image Captioning}: Generating textual descriptions for images \cite{vinyals:2015}.
-			\item \textbf{Challenges}: Complex problem combining computer vision (CV) and natural language processing (NLP) \cite{xu}.
-			\item \textbf{Advancements}: Deep learning models using CNNs for feature extraction and RNNs for language generation \cite{he, simonyan}.
+			\item Task: Generate descriptive textual captions for images.
+			\item Combines computer vision (CV) and natural language processing (NLP).
+			\item Applications: Assistive technology, automated metadata generation, and more.
 	\end{itemize}
+	% Speaker Notes:
+	% Begin by briefly explaining the task of image captioning. Highlight its interdisciplinary nature and practical applications.
 \end{frame}
 
-\section{Methods}
-\begin{frame}{Methods}
+% Slide 2: Problem Statement
+\begin{frame}{Problem Statement}
+	\textbf{Challenges in Image Captioning}
 	\begin{itemize}
-			\item \textbf{Dataset:} Flickr8k, with 8,000 images and five captions per image \cite{vinyals:2015}.
-			\item \textbf{Architecture:} Encoder-Decoder architecture:
-					\begin{itemize}
-							\item Encoder: CNN-based (ResNet-50, VGG-16) \cite{he, simonyan}.
-							\item Decoder: RNN-based (GRU, LSTM) \cite{hochreiter}.
+			\item Detecting and interpreting visual elements accurately.
+			\item Generating coherent and contextually relevant text.
+			\item Balancing computational efficiency and model performance.
 	\end{itemize}
-			\item \textbf{Evaluation Metrics:} BLEU score for performance \cite{papineni}, with CIDEr as an alternative \cite{vedantam}.
-	\end{itemize}
+	% Speaker Notes:
+	% Emphasize the dual challenges of visual understanding and text generation. Mention the need for scalable solutions in real-world applications.
 \end{frame}
 
-\section{Approach}
-\begin{frame}{Approach}
+% Slide 3: Methodology (Dataset and Features)
+\begin{frame}{Methodology: Dataset and Features}
+	\textbf{Dataset: Flickr8k}
 	\begin{itemize}
-			\item \textbf{Data Preprocessing:} Resize images, normalize, and tokenize captions \cite{vinyals:2015}.
-			\item \textbf{Model Training:} Train decoder using Adam optimizer and monitor BLEU score for validation \cite{papineni}.
-			\item \textbf{Model Evaluation:} Evaluate BLEU score after each epoch for overfitting assessment \cite{rennie}.
+			\item Contains 8,000 images with five captions per image.
 	\end{itemize}
+	\textbf{Feature Extraction}
+	\begin{itemize}
+			\item Pre-trained EfficientNet B0 model.
+			\item Extracts high-level features from images.
+	\end{itemize}
+	% Speaker Notes:
+	% Explain why Flickr8k was chosen and describe its characteristics. Introduce EfficientNet B0 as a state-of-the-art feature extractor.
 \end{frame}
 
-\section{Results}
-\begin{frame}{Results}
+% Slide 4: Methodology (Decoder Architectures)
+\begin{frame}{Methodology: Decoder Architectures}
+	\textbf{Architectures Evaluated}
 	\begin{itemize}
-			\item BLEU scores for encoder-decoder combinations (ResNet-GRU, VGG-GRU, ResNet-LSTM, VGG-LSTM) \cite{he, simonyan}.
-				\item \textbf{Best Performance:} VGG encoder with LSTM decoder, highest BLEU score in the first epoch \cite{vinyals:2015}.
-				\item Observation of \textbf{Overfitting}: Decrease in BLEU scores over epochs \cite{rennie}.
+			\item GRU (Gated Recurrent Units)
+			\item LSTM (Long Short-Term Memory Networks)
+			\item Transformer
 	\end{itemize}
+	\textbf{Implementation}
+	\begin{itemize}
+			\item Implemented in PyTorch.
+			\item Trained using image features and paired captions.
+	\end{itemize}
+	% Speaker Notes:
+	% Briefly describe each architecture and its role in the pipeline. Mention the choice of PyTorch for implementation.
 \end{frame}
 
-\section{Conclusion}
-\begin{frame}{Conclusion}
+% Slide 5: Results (Metrics)
+\begin{frame}{Results: Metrics}
+	\textbf{Evaluation Metrics}
 	\begin{itemize}
-			\item \textbf{Summary:} VGG-LSTM model achieved the best BLEU score and stability across epochs \cite{vinyals:2016}.
-			\item \textbf{Future Work:} Explore other architectures (e.g., transformers), datasets (e.g., Flickr30k, MS COCO), and metrics (e.g., CIDEr) \cite{vedantam}.
+			\item BLEU: Measures n-gram overlap.
+			\item METEOR: Considers semantic matching.
+			\item NIST: Focuses on information content.
 	\end{itemize}
+	\textbf{Key Findings}
+	\begin{itemize}
+			\item Transformer underperforms GRU and LSTM in all metrics.
+	\end{itemize}
+	% Speaker Notes:
+	% Introduce the metrics and explain their relevance. Highlight the surprising result of Transformer underperformance.
 \end{frame}
 
-\begin{frame}[allowframebreaks]
-	\bibliographystyle{apalike}
-	\bibliography{bib}
+% Slide 6: Results (Comparison)
+\begin{frame}{Results: Comparison of Architectures}
+	\begin{itemize}
+			\item GRU and LSTM outperform Transformer.
+			\item Transformer struggles with smaller datasets like Flickr8k.
+	\end{itemize}
+	\textbf{Visual Comparison}
+	\includegraphics[width=.6\textwidth]{res/metrics.png}
+	% Speaker Notes:
+	% Present the comparison visually and discuss why the Transformer might struggle with limited data.
+\end{frame}
+
+% Slide 7: Discussion
+\begin{frame}{Discussion}
+	\textbf{Insights and Implications}
+	\begin{itemize}
+			\item Traditional architectures (GRU, LSTM) remain competitive.
+			\item Transformers may require larger datasets to realize their full potential.
+	\end{itemize}
+	% Speaker Notes:
+	% Reflect on what these results mean for the field and the practical implications of using different architectures.
+\end{frame}
+
+% Slide 8: Conclusion and Future Work
+\begin{frame}{Conclusion and Future Work}
+	\textbf{Summary}
+	\begin{itemize}
+			\item Evaluated GRU, LSTM, and Transformer for image captioning.
+			\item GRU and LSTM outperform Transformer on Flickr8k.
+	\end{itemize}
+	\textbf{Future Work}
+	\begin{itemize}
+			\item Test on larger datasets (e.g., MS COCO).
+			\item Explore hybrid architectures.
+	\end{itemize}
+	% Speaker Notes:
+	% Summarize the study's key points and suggest future directions for improvement and exploration.
 \end{frame}

### 2025-01-08 PbD/paper/summary.tex
Included some feedback.

@@
 
+% TODO: reconsider sectioning?
 \chapter{Introduction}
@@
 The \texttt{Backdoor}, like \texttt{Train}, generates a simple classifier $\widetilde{h}$ which appears similar to $h$, and a backdoor key $bk$.
-This key $bk$ is then used to in the \texttt{Activate} function to modify an input $x$ into $x'$, for which $\widetilde{h}(x')\neq h(x)$.
+This key $bk$ is then used to in the \texttt{Activate} function to modify an input $x$ into a nearly similar $x'$, for which $\widetilde{h}(x')=-\widetilde{h}(x)$.
+This results in a backdoor input $x'$ that triggers the backdoor.
     \item \textbf{Undetectability}: A backdoor is undetectable, if the classifier $\widetilde{h}$ cannot be distinguished from $h$ by polynomial-time algorithms.
@@
 \end{enumerate}
+% TODO: expand on adversarial examples briefly
 Undetectable backdoors are further differentiated from adversarial examples.
@@
 \section{Black-Box Undetectable Backdoors from Digital Signatures}
-The paper presents a method to plant a backdoor in any classifier using digital signatures, with robust properties that make it hard to detect or replicate without the backdoor key.
+% TODO: a little more detail
+A method is presented to plant a backdoor in any classifier using digital signatures, with robust properties that make it hard to detect or replicate without the backdoor key.
 \par By leveraging digital signatures, the backdoor uses a public-private key pair.
@@
 \par The modified classifier includes a verification mechanism for message-signature pairs. When a valid signed input (created with $s_k$) is detected, it overrides the classifier’s natural behavior.
-\par The section culminates in a theorem.
-Assuming the existence of one-way functions, for every training procedure \texttt{Train}, there exists a model backdoor (\texttt{Backdoor}, \texttt{Activate}), which is non-replicable and black-box undetectable \cite{goldwasser2022backdoors}.
+\par The section culminates in a theorem: "Assuming the existence of one-way functions, for every training procedure \texttt{Train}, there exists a model backdoor (\texttt{Backdoor}, \texttt{Activate}), which is non-replicable and black-box undetectable" \cite{goldwasser2022backdoors}.
 
@@
 \section{Persistence Against Post-Processing}
-The authors demonstrate that any neural network $N$ can be efficiently transformed into a functionally identical network $N’$ that is immune to changes from gradient descent.
-This is achieved by constructing some error-correction for the weights of $N’$, making them persistent to iterations of gradient descent.
+The authors demonstrate that any neural network $N$ can be efficiently transformed into a functionally identical network $N'$ that is immune to changes from gradient descent.
+This is achieved by constructing some error-correction for the weights of $N'$, making them persistent to iterations of gradient descent.
 This illustrates the extent of control a malicious trainer can maintain over a model.

### 2025-01-08 PbD/paper/summary.tex
More feedback implemented.

@@
 
-% TODO: reconsider sectioning?
 \chapter{Introduction}
@@
 \end{enumerate}
-% TODO: expand on adversarial examples briefly
 Undetectable backdoors are further differentiated from adversarial examples.
-Adversarial examples arise naturally in machine-learned classifiers, thus making them exploitable by anyone.
-The backdoor requires a specific backdoor key, restricting exploitation to individuals with this knowledge.
-Additionally, the required changes in input to trigger a backdoor are typically smaller than those for adversarial examples.
+Adversarial examples arise naturally in machine-learned classifiers as a result of the inherent complexity of decision boundaries, and can thus be exploited by anyone.
+In contrast, backdoors require a specific backdoor key, restricting their use to those with this knowledge.
+Furthermore, the perturbations needed to activate a backdoor are typically smaller and more precise than those required to generate adversarial examples.
+This subtlety, combined with the targeted nature of backdoors, highlights their malicious potential compared to the broader and naturally occurring phenomenon of adversarial examples.
 
 \section{Black-Box Undetectable Backdoors from Digital Signatures}
-% TODO: a little more detail
-A method is presented to plant a backdoor in any classifier using digital signatures, with robust properties that make it hard to detect or replicate without the backdoor key.
-\par By leveraging digital signatures, the backdoor uses a public-private key pair.
-The public verification key $v_k$ identifies valid inputs, while the secret signing key $s_K$ allows the adversary to create specific inputs that trigger the backdoor, enabling control over the classifier’s output.
-\par The modified classifier includes a verification mechanism for message-signature pairs. When a valid signed input (created with $s_k$) is detected, it overrides the classifier’s natural behavior.
-\par The section culminates in a theorem: "Assuming the existence of one-way functions, for every training procedure \texttt{Train}, there exists a model backdoor (\texttt{Backdoor}, \texttt{Activate}), which is non-replicable and black-box undetectable" \cite{goldwasser2022backdoors}.
+
+A method is presented to plant a backdoor in any classifier using digital signatures, ensuring the backdoor is both hard to detect and replicate without the backdoor key.
+By embedding a public-private key pair, the public verification key $v_k$ identifies valid inputs, while the private signing key $s_k$ allows the adversary to craft inputs that trigger the backdoor, subtly modifying the input $x$ into $x'$.
+This transformation, designed to keep $x'$ similar to $x$, ensures undetectability while overriding the classifier’s normal behavior.
+\par The modified classifier includes a verification mechanism for message-signature pairs.
+When a valid signed input (created with $s_k$) is detected, the classifier outputs the inverse of what it would normally produce.
+This approach ensures that the backdoor remains undetectable to polynomial-time algorithms with oracle access to the classifier, and it is non-replicable without $s_k$.
+\par The section culminates in the theorem: "assuming the existence of one-way functions, for every training procedure \texttt{Train}, there exists a model backdoor (\texttt{Backdoor}, \texttt{Activate}) that is black-box undetectable and non-replicable" \cite{goldwasser2022backdoors}.
 

### 2025-01-09 SbD/exercises.tex
Added some more exercises.
Added exercises into latex.

@@
 
-\usepackage[solutions]{../.template/xrcise}
+\usepackage[]{../.template/xrcise}
 
@@
 
-% \setcounter{subsection}{90}
-% \begin{exercise}{}
-%  % TODO
-%
-%   \begin{solution}
-%     % TODO
-%   \end{solution}
-% \end{exercise}
-
-% \setcounter{subsection}{100}
-% \begin{exercise}{}
-%  % TODO
-%
-%   \begin{solution}
-%     % TODO
-%   \end{solution}
-% \end{exercise}
+\setcounter{subsection}{90}
+\begin{exercise}{Bluetooth-Pairing}
+  Bei Bluetooth erfolgt zur sicheren Gerätekopplung zunächst ein sogenanntes Pairing, das in zwei Phasen abläuft.
+  \begin{enumerate}
+    \item Bitte beschreiben Sie jeweils die Aufgabe der beiden Phasen.
+    \item Wie lauten die jeweiligen Ein- und Ausgabewerte in beiden Phasen?
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+      \item \textbf{Phase 1 (Pairing):} Authentifizierung und Schlüsselaustausch. \textbf{Phase 2 (Bonding):} Speicherung der Schlüssel für zukünftige Verbindungen.
+      \item \textbf{Phase 1:} Eingabe: PIN, Ausgabe: Link Key. \textbf{Phase 2:} Eingabe: Link Key, Ausgabe: Keine.
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{TMSI-Vergabe}
+  Bitte erläutern Sie:
+  \begin{enumerate}
+    \item das Schutzziel
+    \item das Angreifermodell und
+    \item das Verfahren zur Vergabe einer TMSI im GSM.
+    \item Existiert in Bluetooth eine vergleichbare Funktionalität wie die TMSI-Vergabe im GSM? Bitte begründen Sie Ihre Antwort kurz!
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+      \item Schutzziel: Vermeidung von IMSI-Catching und Tracking.
+      \item Angreifermodell: IMSI-Catcher und Analyse der TMSI-Vergabe.
+      \item Verfahren: Zufällige Vergabe einer TMSI durch das VLR.
+      \item Bluetooth: Nein, da Bluetooth keine IMSI verwendet.
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Bruchstückhafter AES}
+  Bei AES werden rundenweise "AddRoundKey", "SubBytes", "ShiftRows" und "MixColumns" angewandt.
+Für diese Aufgabe können Sie eine Blocklänge von 128 Bit und eine Schlüssellänge von 256 Bit annehmen.
+Bitte zeigen Sie unter Angabe der hierfür notwendigen Angriffsart und unter Berechnung der effektiven Schlüssellänge im Kontext dieser Angriffsart die Unsicherheit von AES bei Verzicht pro Runde auf …
+  \begin{enumerate}
+    \item … die Anwendung der Operation "AddRoundKey".
+    \item … die Anwendung der Operation "SubBytes" bei der Schlüsselexpansion sowie der Verschlüsselung.
+    \item … die Anwendung der Operation "SubBytes" bei der Verschlüsselung, mit jedoch weiterbestehender Anwendung dieser Operation bei der Schlüsselexpansion.
+    \item … die Anwendung der Operation "ShiftRows".
+    \item … die Anwendung der Operation "MixColumns".
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+      \item Angriffsart: Known-Plaintext-Angriff. Effektive Schlüssellänge: 128 Bit.
+      \item Angriffsart: Chosen-Plaintext-Angriff. Effektive Schlüssellänge: 128 Bit.
+      \item Angriffsart: Chosen-Plaintext-Angriff. Effektive Schlüssellänge: 128 Bit.
+      \item Angriffsart: Known-Plaintext-Angriff. Effektive Schlüssellänge: 128 Bit.
+      \item Angriffsart: Known-Plaintext-Angriff. Effektive Schlüssellänge: 128 Bit.
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+\setcounter{subsection}{100}
+% Aufgabe 1 Spoofing
+% a) Inwiefern kann ein DNS-Spoofing-Angriff für einen Distributed Denial of Service (DDoS)-Angriff ausge-
+% nutzt werden? Erklären Sie anschaulich.
+% b) Warum funktioniert ARP-Spoofing nur in einem lokalen Netz?
+% c) Sie vermuten, Opfer eines ARP-Spoofing-Angriffs zu sein. Was können Sie an Ihrem Rechner überprü-
+% fen, um einen solchen Angriff erkennen zu können?
+\begin{exercise}{Spoofing}
+  \begin{enumerate}
+    \item Inwiefern kann ein DNS-Spoofing-Angriff für einen Distributed Denial of Service (DDoS)-Angriff genutzt werden? Erklären Sie anschaulich.
+    \item Warum funktioniert ARP-Spoofing nur in einem lokalen Netz?
+    \item Sie vermuten, Opfer eines ARP-Spoofing-Angriffs zu sein. Was können Sie an Ihrem Rechner überprüfen, um einen solchen Angriff erkennen zu können?
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+      \item Ein DNS-Spoofing-Angriff kann genutzt werden, um die IP-Adressen von Opfern auf die IP-Adresse des Opfers umzuleiten.
+Dadurch wird der Traffic auf das Opfer umgeleitet und es kommt zu einem DDoS-Angriff.
+      \item ARP-Spoofing funktioniert nur in einem lokalen Netz, da ARP-Pakete nicht über Router weitergeleitet werden.
+      \item Überprüfen Sie die ARP-Tabelle Ihres Rechners, um verdächtige Einträge zu identifizieren.
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Mehrfachnutzung des IV in einem WEP-verschlüsselten WLAN}
+  \begin{enumerate}
+    \item Der in WEP verwendete Initialisierungsvektor (IV) ist 24 Bit groß und steht in einem Paket $P$ am Paketanfang vor dem Schlüsseltext.
+Wie viele Pakete können maximal erzeugt werden, wenn kein IV mehrfach verwendet werden soll?
+Wie sind die IVs von einem Teilnehmer des WLANs zu erzeugen, sodass eine Kollision möglichst spät auftritt?
+Welches Problem gibt es bei diesem Verfahren, wenn mehrere Teilnehmer kommunizieren?
+    \item Nehmen Sie an, Sie haben die Pakete $P_1$ und $P_2$ empfangen:
+      \begin{align*}
+        P_1&: 00010\ 00111\ 11111\ 11110\ 11001\ 01001\ 00101\ \ldots\\
+        P_2&: 00010\ 00111\ 11111\ 11110\ 11000\ 10011\ 00010\ \ldots
+      \end{align*}
+      Die ersten 10 Klartextbits des ersten Paketes $P_1$ lauten $10010\ 01100$. Bestimmen Sie die ersten 10 Klartextbits des zweiten Paketes $P_2$.
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+      \item Es können maximal $2^{24}$ Pakete erzeugt werden, wenn kein IV mehrfach verwendet werden soll.
+IVs sollten zufällig erzeugt werden, um Kollisionen zu vermeiden.
+Bei mehreren Teilnehmern kann es zu Kollisionen kommen, wenn IVs nicht synchronisiert werden.
+      \item Die ersten 10 Klartextbits des zweiten Paketes $P_2$ lauten $10010\ 01101$.
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Challenge-Response-Authentifikation in GSM, Bluetooth und WEP}
+  \begin{enumerate}
+    \item Wie läuft eine einseitige Challenge-Response-Authentifikation allgemein ab?
+    \item Bitte benennen Sie Gemeinsamkeiten und Unterschiede der Challenge-Response-Authentifikation in Bluetooth, GSM und WEP.
+Beschränken Sie sich dabei auf den Ablauf und die wesentlichen Eigenschaften.
+    \item Die Challenge-Response-Authentifikation bei WEP weist eine Schwäche auf. Wie nutzt ein Angriff diese Schwäche aus? Machen Sie einen Vorschlag, wie man diesen Angriff verhindern könnte.
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+      \item Ein Teilnehmer sendet eine Challenge an einen anderen Teilnehmer, der diese Challenge mit einer Response beantwortet.
+      \item Gemeinsamkeiten: Authentifizierung durch Challenge-Response-Verfahren. Unterschiede: Verwendung in unterschiedlichen Protokollen, z. B. GSM, Bluetooth und WEP.
+      \item Angriff: Wiederholte Authentifizierungsanfragen. Verhinderung: Begrenzung der Authentifizierungsanfragen.
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+% Aufgabe 4 Challenge-Response-Authentifikation überarbeitet
+% In der klassischen beidseitigen Challenge-Response-Authentifikation mittels symmetrischer, deterministi-
+% scher Verschlüsselung sendet ein prüfendes IT-System eine Zufallszahl als Challenge und erwartet deren
+% Verschlüsselung mit einem gemeinsamen Schlüssel K. Nachfolgend findet sich eine formale Darstellung:
+% Lokal für Kanal 1: A wählt und speichert vorläufig RA
+% Kanal 1: A RA −−−−→B
+% Lokal für Kanal 1: B wählt und speichert vorläufig RB
+% Lokal für Kanal 1: B berechnet CA := EK(RA)
+% Kanal 1: A RB,CA ←−−−−B
+% ?
+% Lokal für Kanal 1: A testet CA
+% = EK(RA) (oder: DK(CA) ?
+% = RA) und löscht RA
+% Lokal für Kanal 1: A berechnet CB := EK(RB)
+% Kanal 1: A CB −−−−→B
+% Lokal für Kanal 1: B testet CB
+% ?
+% = EK(RB) (oder: DK(CB) ?
+% = RB) und löscht RB
+% Der verwendete Schlüssel K darf hierbei nur dem Geprüftem und dem prüfenden IT-System bekannt sein.
+% Wo immer Sie im Rahmen dieser Aufgabe ein Protokoll beschreiben sollen, verwenden Sie diese formale
+% Darstellung. Bezeichnen Sie für eine asymmetrische Verschlüsselung mit PK den öffentlichen und mit SK den
+% geheimen Schlüssel, sowie mit EPK(·) die Ver- und mit DSK(·) die Entschlüsselung.
+% a) Wieso werden Zufallszahlen und nicht feste Werte verwendet?
+% b) Gegen welche Art von Angriff, bei welcher der Angreifer nur mit einer Partei kommunizieren muss, ist
+% die klassiche Challenge-Reponse-Authentifikation wie oben dargestellt verwundbar? Beschreiben Sie
+% einen erfolgreichen Angriff.
+% c) Wie lässt sich der Angriff aus b) durch dauerhaftes statt vorläufiges Speichern der versandten Challen-
+% ges und entsprechende Prüfungen vermeiden?
+% d) Beschreiben Sie eine überarbeitete Challenge-Response-Authentifikation mittels symmetrischer Ver-
+% schlüsselung, die gegen Angriffe wie in b) geschützt ist und keine dauerhafte Speicherung von Daten
+% erfordert.
+% e) Die klassische Challenge-Response-Authentifikation mittels symmetrischer Verschlüsselung wendet
+% den Verschlüsselungsalgorithmus zweimal an. Beschreiben Sie ein Protokoll zur gegenseitigen Au-
+% thentifikation mit lediglich einer Anwendung eines symmetrischen Verschlüsselungsalgorithmus.
+% f) Beschreiben Sie ein Challenge-Response-Protokoll zur gegenseitigen Authentifkation mit asymmetri-
+% scher statt symmetrischer Verschlüsselung. Was muss gelten damit das Protokoll gegen Angriffe wie in
+% b) geschützt ist?
+\begin{exercise}{Challenge-Response-Authentifikation überarbeitet}
+  In der klassischen beidseitigen Challenge-Response-Authentifikation mittels symmetrischer, deterministischer Verschlüsselung sendet ein prüfendes IT-System eine Zufallszahl als Challenge und erwartet deren Verschlüsselung mit einem gemeinsamen Schlüssel $K$.
+Nachfolgend findet sich eine formale Darstellung:
+  \begin{align*}
+    \text{Lokal für Kanal 1:} & A \text{ wählt und speichert vorläufig } R_A\\
+    \text{Kanal 1:} & A \xrightarrow{R_A} B\\
+    \text{Lokal für Kanal 1:} & B \text{ wählt und speichert vorläufig } R_B\\
+    \text{Lokal für Kanal 1:} & B \text{ berechnet } C_A := E_K(R_A)\\
+    \text{Kanal 1:} & A \xleftarrow{R_B, C_A} B\\
+    \text{Lokal für Kanal 1:} & A \text{ testet } C_A = E_K(R_A) \text{ und löscht } R_A\\
+    \text{Lokal für Kanal 1:} & A \text{ berechnet } C_B := E_K(R_B)\\
+    \text{Kanal 1:} & A \xrightarrow{C_B} B\\
+    \text{Lokal für Kanal 1:} & B \text{ testet } C_B = E_K(R_B) \text{ und löscht } R_B
+  \end{align*}
+  Der verwendete Schlüssel $K$ darf hierbei nur dem Geprüften und dem prüfenden IT-System bekannt sein.
+Wo immer Sie im Rahmen dieser Aufgabe ein Protokoll beschreiben sollen, verwenden Sie diese formale Darstellung.
+Bezeichnen Sie für eine asymmetrische Verschlüsselung mit $PK$ den öffentlichen und mit $SK$ den geheimen Schlüssel, sowie mit $E_{PK}(\cdot)$ die Ver- und mit $D_{SK}(\cdot)$ die Entschlüsselung.
+  \begin{enumerate}
+    \item Wieso werden Zufallszahlen und nicht feste Werte verwendet?
+    \item Gegen welche Art von Angriff, bei welcher der Angreifer nur mit einer Partei kommunizieren muss, ist die klassische Challenge-Response-Authentifikation wie oben dargestellt verwundbar?
+Beschreiben Sie einen erfolgreichen Angriff.
+    \item Wie lässt sich der Angriff aus b) durch dauerhaftes statt vorläufiges Speichern der versandten Challenges und entsprechende Prüfungen vermeiden?
+    \item Beschreiben Sie eine überarbeitete Challenge-Response-Authentifikation mittels symmetrischer Verschlüsselung, die gegen Angriffe wie in b) geschützt ist und keine dauerhafte Speicherung von Daten erfordert.
+    \item Beschreiben Sie ein Protokoll zur gegenseitigen Authentifikation mit lediglich einer Anwendung eines symmetrischen Verschlüsselungsalgorithmus.
+    \item Beschreiben Sie ein Challenge-Response-Protokoll zur gegenseitigen Authentifikation mit asymmetrischer statt symmetrischer Verschlüsselung.
+Was muss gelten, damit das Protokoll gegen Angriffe wie in b) geschützt ist?
+  \end{enumerate}
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
 

### 2025-01-09 PbD/exercises.tex
Added some more exercises.
Added exercises into latex.

@@
 \begin{document}\createtitle
+
+\section{Assignments}
+\begin{exercise}{Protection Goals}
+  In IT security, the protection goals of confidentiality, availability, and integrity are defined to describe what objectives an attacker aims for or what a technology is designed to protect.
+Provide an appropriate technique for each of these protection goals that can be used to implement or address the respective goal.
+
+  \begin{solution}
+    \begin{itemize}
+      \item \textbf{Confidentiality}: Encryption is used to protect the confidentiality of data.
+      \item \textbf{Availability}: Redundancy is used to ensure the availability of services.
+      \item \textbf{Integrity}: Hash functions are used to ensure the integrity of data.
+    \end{itemize}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Pseudonyms}
+  Pseudonyms are fundamental building blocks of secure and privacy-friendly systems.
+  \begin{enumerate}
+    \item Briefly describe the general purpose of pseudonyms.
+    \item What is the difference between pseudonymity and anonymity?
+    \item Name three applications in which pseudonyms are used. Specify the type of pseudonym used in each case. Explain how pseudonyms are employed in each application.
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+      \item The general purpose of pseudonyms is to protect the identity of users.
+      \item The difference between pseudonymity and anonymity is that pseudonymity uses a pseudonym to protect the identity of users, while anonymity does not require a pseudonym.
+      \item Three applications in which pseudonyms are used are:
+        \begin{itemize}
+          \item E-Mail: Pseudonyms are used to protect the identity of users.
+          \item Social Media: Pseudonyms are used to protect the identity of users.
+          \item Online Shopping: Pseudonyms are used to protect the identity of users.
+        \end{itemize}
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Indeterministic Encryption}
+  \begin{enumerate}
+    \item What does indeterministic encryption mean?
+    \item How do you implement indeterministic encryption in an asymmetric cryptosystem?
+    \item What changes compared to b) when using a symmetric cryptosystem?
+    \item You are tasked with implementing indeterministic encryption. The algorithms RSA and ElGamal are available. How would you proceed with each algorithm?
+    \item Which attack can be prevented by using indeterministic encryption?
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+      \item Indeterministic encryption means that the same plaintext is encrypted to different ciphertexts each time.
+      \item Indeterministic encryption in an asymmetric cryptosystem is implemented by adding randomness to the encryption process.
+      \item In a symmetric cryptosystem, indeterministic encryption is implemented by adding randomness to the encryption process.
+      \item To implement indeterministic encryption with RSA, random padding is added to the plaintext before encryption. With ElGamal, a random number is used to generate the encryption key.
+      \item Indeterministic encryption prevents chosen-ciphertext attacks.
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+% Aufgabe 4 Unobservable retrieval of information
+% Given is a Blind-Message-Service with three servers S1, S2, and S3. The following datasets are stored in the
+% Blind-Message-Service: D0 = 1011, D1 = 0101, D2 = 1100, D3 = 0011, D4 = 0000.
+% a) You want to retrieve dataset D2 without the servers knowing which dataset you are interested in. What
+% messages are transmitted? How do you obtain the requested dataset? Sketch the computational steps.
+% b) You want to ensure, as in task a), that during a query it remains unknown which dataset the client is
+% interested in. However, instead of the Blind-Message-Service, only a single high-bandwidth server is
+% available. The database, however, is relatively small. Which method do you apply?
+\begin{exercise}{Unobservable Retrieval of Information}
+  Given is a Blind-Message-Service with three servers S1, S2, and S3. The following datasets are stored in the Blind-Message-Service: D0 = 1011, D1 = 0101, D2 = 1100, D3 = 0011, D4 = 0000.
+  \begin{enumerate}
+    \item You want to retrieve dataset D2 without the servers knowing which dataset you are interested in.
+What messages are transmitted?
+How do you obtain the requested dataset?
+Sketch the computational steps.
+    \item You want to ensure, as in task a), that during a query it remains unknown which dataset the client is interested in.
+However, instead of the Blind-Message-Service, only a single high-bandwidth server is available.
+The database, however, is relatively small.
+Which method do you apply?
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+      \item To retrieve dataset D2 without the servers knowing which dataset you are interested in, you send a request for all datasets and then filter out the requested dataset.
+The computational steps are as follows:
+        \begin{itemize}
+          \item Send a request for all datasets.
+          \item Receive the datasets from the servers.
+          \item Filter out the requested dataset D2.
+        \end{itemize}
+      \item To ensure that the client's interest remains unknown, you use the Broadcast method.
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Tracking on the Application Layer}
+  \begin{enumerate}
+    \item Name two security goals that are violated during tracking at the application layer.
+    \item What is the difference between tracking at the application layer and tracking at the network layer?
+    \item Name three fundamental mechanisms to prevent or make tracking at the application layer more difficult.
+Explain for each mechanism how it hinders tracking, against which type of attacker, and provide a specific example where the mechanism is implemented.
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+      \item Two security goals that are violated during tracking at the application layer are confidentiality and anonymity.
+      \item The difference between tracking at the application layer and tracking at the network layer is that tracking at the application layer uses browser data, while tracking at the network layer uses IP data.
+      \item Three fundamental mechanisms to prevent or make tracking at the application layer more difficult are:
+        \begin{itemize}
+          \item Block cookies: Prevents tracking by websites. Against advertisers. Example: Browser settings.
+          \item Use VPNs: Prevents tracking by ISPs. Against ISPs. Example: NordVPN.
+          \item Use privacy plugins: Prevents tracking by websites. Against advertisers. Example: uBlock Origin.
+        \end{itemize}
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Surveillance on the Internet}
+  \begin{enumerate}
+    \item Explain a possible approach by which a powerful attacker (e.g., an intelligence agency) can determine the IP addresses of users accessing a specific website through a mix-based anonymization system (e.g., Tor).
+    \item Explain a possible approach by which a powerful attacker (e.g., an intelligence agency) can determine which websites a specific user has accessed via a mix-based anonymization system.
+Note: Describe a different method from the one in task a), even if your method in a) could also be applied to task b).
+    \item Explain three techniques by which a web server can recognize a user during their second (subsequent) visit.
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+      \item A powerful attacker can determine the IP addresses of users accessing a specific website through a mix-based anonymization system by using traffic analysis and traffic correlation.
+      \item A powerful attacker can determine which websites a specific user has accessed via a mix-based anonymization system by using metadata analysis, such as HTTP headers or specific request characteristics.
+      \item Three techniques by which a web server can recognize a user during their second visit are:
+        \begin{itemize}
+          \item Cookies: Store user data in the browser. Against websites. Example: Google Analytics.
+          \item Browser fingerprinting: Collect browser data. Against websites. Example: Panopticlick.
+          \item Tracking pixels: Track user behavior. Against advertisers. Example: Facebook Pixel.
+        \end{itemize}
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
 

### 2025-01-09 SbD/exercises.tex
Updated solutions.

@@
         \item Bei allen Modi führt ein Bitfehler im zweiten Block dazu, dass dieser Block nicht entschlüsselt werden kann.
-Im CBC-Modus wird zusätzlich der Block 3 verändert, da die XOR-Verknüpfung mit dem vorherigen Block nicht mehr korrekt ist.
-        \item Wenn die Blockgrenzen nicht bekannt sind, kann nur der erste Block entschlüsselt werden.
-Im CBC kann zusätzlich selbst bei bekannten Blockgrenzen nichts weiter entschlüsselt werden, da die XOR-Verknüpfung mit dem vorherigen Block nicht mehr korrekt ist.
+Im CBC-Modus wird zusätzlich das relevante Bit im Block 3 verändert, da die XOR-Verknüpfung mit dem vorherigen Block nicht mehr korrekt ist.
+        \item Wenn Blockgrenzen bekannt sind, wie in 3., aber bei CBC wird der dritte Block völlig unbrauchbar.
+Wenn nicht, sind Block 2 und alle nachfolgenden Blöcke nicht entschlüsselbar aufgrund des Synchronisationsverlusts.
         \item Bei ECB und CTR können alle anderen Blöcke korrekt entschlüsselt werden, da sie unabhängig voneinander sind.
@@
     \item Auf welcher mathematischen Schwierigkeit basiert der ECC?
-    \item Berechnen Sie die Punkte auf der folgenden elliptischen Kurve: $a = 6$, $b = 1$, $p = 11$ (allgemeine Formel war gegeben).
+    \item Berechnen Sie die Punkte auf der folgenden elliptischen Kurve: $a = 6$, $b = 1$, $p = 11$.
     \item Nennen Sie zwei Vorteile, die ECC gegenüber sicherheitsmäßig vergleichbaren Verfahren hat.

### 2025-01-14 SbD/exercises.tex
Updated some solutions.

@@
   \begin{enumerate}
-      \item Verschlüsseln Sie den folgenden Klartext mit einer Skytala, die durch eine Matrix mit sieben Zeilen und neun Spalten beschrieben werden kann:
+    \item\label{ex:skytala} Verschlüsseln Sie den folgenden Klartext mit einer Skytala, die durch eine Matrix mit sieben Zeilen und neun Spalten beschrieben werden kann:
       \begin{center}
@@
       \end{center}
-      \item Was ist der geheim zu haltende symmetrische Schlüssel der Skytala aus Teilaufgabe a) und wie hängt dieser mit der ursprünglichen Variante mit Zylinder und Papierstreifen zusammen?
+    \item Was ist der geheim zu haltende symmetrische Schlüssel der Skytala aus Teilaufgabe \ref{ex:skytala} und wie hängt dieser mit der ursprünglichen Variante mit Zylinder und Papierstreifen zusammen?
     \item Erläutern Sie kurz zwei sinnvolle Ansätze zur Kryptanalyse, d. h. Ansätze zum Brechen einer Skytala.
@@
     \begin{enumerate}
-        \item Verschlüsselung mit Skytala: Teilen Sie den Klartext in eine Matrix mit 7 Zeilen und 9 Spalten:
-          \[
-          \texttt{GLUECKLICH IST DER DER DEN DINGEN AUF DEN GRUND GEHEN KONNTE}
-          \]
-          Die Matrix:
+        \item Teilen Sie den Klartext in eine Matrix mit 7 Zeilen und 9 Spalten:
           \[
@@
           \]
-          Der Schlüsseltext wird spaltenweise abgelesen: 
-          \[
-          \texttt{GH\_DDFNLDIRIGTNSEEECEDEINNEK\_CRNUENNUH\_U\_GNGRDKIKAUOE}
-          \]
-
-        \item Der geheime Schlüssel ist die Größe der Matrix: $ 7 \times 9 $ (Zeilen $\times$ Spalten). Beim ursprünglichen Zylinder entspricht dies dem Umfang des Zylinders.
-
-        \item Kryptanalyse:
-        \begin{itemize}
-            \item Häufigkeitsanalyse: Anhand von Buchstabenhäufigkeiten und Mustererkennung können Transpositionsmuster aufgedeckt werden.
+          Der Schlüsseltext wird spaltenweise abgelesen: $\texttt{GH\_DDFNLDIRIGTNSEEECEDEINNEK\_CRNUENNUH\_U\_GNGRDKIKAUOE}$.
+        \item Der geheime Schlüssel ist die Größe der Matrix: $7\times9$ (Zeilen $\times$ Spalten). Beim ursprünglichen Zylinder entspricht dies dem Umfang des Zylinders.
+        \item Kryptanalyse der Skytala.
+          \begin{itemize}
               \item Brute-Force: Alle möglichen Spalten- und Zeilenkombinationen ausprobieren, um den Klartext zu rekonstruieren.
+              \item N-gram-Analyse: Häufige Buchstabenkombinationen im Klartext können auf die Anzahl der Spalten hinweisen und die Wahrscheinlichkeit erhöhen, den Schlüssel zu erraten.
           \end{itemize}
@@
     \begin{enumerate}
-        \item Bei Spaltenvertauschung gibt es $ s!
-$ mögliche Permutationen für $ s $ Spalten.
-Der Schlüsselraum vergrößert sich somit um den Faktor $ s!
-$ im Vergleich zur ursprünglichen Methode ohne Spaltenvertauschung.
-
-        \item Klartext in eine $ 4 \times 3 $-Matrix:
-          \[
-          \texttt{ABCDEFGHIKLMN}
-          \]
-          In die Matrix eingetragen:
+        \item Bei Spaltenvertauschung gibt es $s!$ mögliche Permutationen für $s$ Spalten.
+Der Schlüsselraum vergrößert sich somit um den Faktor $s!$ im Vergleich zur ursprünglichen Methode ohne Spaltenvertauschung.
+        \item Klartext in eine $4\times3$-Matrix:
           \[
             \begin{array}{ccc}
-          A & B & C \\
-          D & E & F \\
-          G & H & I \\
-          K & L & M \\
+            1 & 2 & 3 \\
+            4 & 5 & 6 \\
+            7 & 8 & 9 \\
+            10 & 11 & 12 \\
             \end{array}
           \]
-          Zyklisches Schreiben ergibt:
-          \[
-          \texttt{ADGKBELHCFIM}
-          \]
+          Zyklisches Schreiben ergibt $\texttt{14710 25811 36912}$.
     \end{enumerate}
@@
 \begin{exercise}{Zufallszahlen bei der Schlüsselgenerierung}
-  Warum sind bei der Schlüsselgenerierung (echte) Zufallszahlen nötig?
-Weshalb eignet sich die XOR-Verknüpfung, wenn zur Erhöhung der Sicherheit mehrere Zufallszahlen zu einer einzelnen Zufallszahl verknüpft werden, etwa bei der Schlüsselgenerierung?
+  \begin{enumerate}
+    \item Warum sind bei der Schlüsselgenerierung (echte) Zufallszahlen nötig?
+    \item Weshalb eignet sich die XOR-Verknüpfung, wenn zur Erhöhung der Sicherheit mehrere Zufallszahlen zu einer einzelnen Zufallszahl verknüpft werden, etwa bei der Schlüsselgenerierung?
+  \end{enumerate}
 
@@
 Ein nicht zufälliger Schlüssel kann durch statistische oder deterministische Methoden berechnet werden.
-        \item Die XOR-Verknüpfung sichert, dass ein kombinierter Schlüssel nur dann vorhersagbar ist, wenn alle zufälligen Werte vollständig bekannt sind.
-Dadurch wird ein höherer Grad an Sicherheit erzielt.
+        \item Länge bleibt unverändert, XOR-Operation ist invertierbar durch sich selbst, kommutativ bei der Reihenfolge der Operanden.
     \end{enumerate}
@@
   \begin{solution}
-    Durch die Wiederverwendung des One-Time-Pads $ k $ für mehrere Nachrichten $ m_1, m_2 $ entsteht folgende Gleichung:
-      \[ c_1 \oplus c_2 = (m_1 \oplus k) \oplus (m_2 \oplus k) = m_1 \oplus m_2 \]
-    Dies ergibt die XOR-Verknüpfung der Klartexte $ m_1 \oplus m_2 $, die durch Analyse entschlüsselt werden kann. Beispielsweise:
-      \[ c_1 = \texttt{09 00 04 10}, \quad c_2 = \texttt{10 20 28 09} \]
-    XOR:
-      \[ c_1 \oplus c_2 = \texttt{19 20 2C 19} \]
-    Häufigkeitsanalysen und bekannte Wörter im Klartext ermöglichen die vollständige Entschlüsselung.
+    ???
   \end{solution}
@@
     \begin{enumerate}
-        \item Ursachen für Algorithmuswechsel:
-        \begin{itemize}
-            \item **Technologischer Fortschritt**: Schnellere Computer machen bestimmte Algorithmen unsicher.
-            \item **Neue Angriffsmethoden**: Entdeckte Schwachstellen führen zur Ablösung alter Verfahren.
-        \end{itemize}
-
-        \item Das **One-Time-Pad** bietet langfristige Sicherheit, da es informationstheoretisch sicher ist.
+        \item \begin{itemize}
+            \item Technologischer Fortschritt: Schnellere Computer machen bestimmte Algorithmen unsicher.
+            \item Neue Angriffsmethoden: Entdeckte Schwachstellen führen zur Ablösung alter Verfahren.
+          \end{itemize}
+        \item Das One-Time-Pad bietet langfristige Sicherheit, da es informationstheoretisch sicher ist.
 Es wird nicht überall genutzt, da Schlüssel so lang wie der Klartext sein müssen und sicher verteilt werden müssen.
-
-        \item Bestimmung der Chiffre:
-        \begin{itemize}
-            \item **Häufigkeitsanalyse**: Substitutionschiffren wie Caesar zeigen typische Muster.
-            \item **Permutationsanalyse**: Transpositionen hinterlassen Permutationsmuster.
+        \item \begin{itemize}
+            \item Häufigkeitsanalyse: Substitutionschiffren wie Caesar zeigen typische Muster.
+            \item Permutationsanalyse: Transpositionen hinterlassen Permutationsmuster.
         \end{itemize}
@@
         \begin{itemize}
-            \item **RSA**: Fügen Sie der Nachricht einen zufälligen Wert hinzu (z. B. OAEP-Padding).
-            \item **ElGamal**: Verwenden Sie bei jeder Verschlüsselung einen neuen Zufallswert.
-        \end{itemize}
-        \item Indeterministische Verschlüsselung verhindert einen **Known-Plaintext-Angriff**, bei dem der Angreifer durch die Analyse von wiederholt verschlüsselten Nachrichten auf den Klartext schließen könnte.
+            \item RSA: Fügen Sie der Nachricht einen zufälligen Wert hinzu (z. B. OAEP-Padding).
+            \item ElGamal: Verwenden Sie bei jeder Verschlüsselung einen neuen Zufallswert.
+        \end{itemize}
+        \item Indeterministische Verschlüsselung verhindert einen Known-Plaintext-Angriff, bei dem der Angreifer durch die Analyse von wiederholt verschlüsselten Nachrichten auf den Klartext schließen könnte.
     \end{enumerate}

### 2025-01-14 NLP/exercises.tex
Small exercise updates.

@@
 
-\begin{exercise}{Teacher/Student}
-  Explain the teacher/student signals approach in educational NLP and why it is better than pure audio content.
-
-  \begin{solution}
-    \begin{itemize}
-        \item \textbf{Teacher signals}: Provide detailed instructions and explanations.
-        \item \textbf{Student feedback}: Allows iterative learning and personalization.
-        \item \textbf{Better than audio}: Interaction and adaptivity enhance understanding and retention compared to static audio content.
-    \end{itemize}
-  \end{solution}
-\end{exercise}
-
 \begin{exercise}{Representation and fusion}
@@
     \item Calculate the Minimum Edit Distance for "TAC" vs. "CAT".
-    \item What are the possible word outcomes?
   \end{enumerate}
@@
         \item The Minimum Edit Distance is the minimum number of operations (insertion, deletion, substitution) required to transform one string into another.
-        \item \[\begin{array}{c|c|c|c|c}
-            &  & C & A & T \\
-            \hline
-            & 0 & 1 & 2 & 3 \\
-            T & 1 & 1 & 2 & 2 \\
-            A & 2 & 2 & 1 & 2 \\
-            C & 3 & 2 & 2 & 2 \\
-        \end{array}\]
-        \item CAT, TAT, TAC, CAC
+        \item The Minimum Edit Distance for "TAC" vs. "CAT" is 2.
     \end{enumerate}

### 2025-01-14 SbD/exercises.tex
Small updates to exercises.

@@
 In welchen Zeilen haben Sie die DES-Funktionswerte, die zum Erfolg beim Meet-in-the-Middle Angriff führen, nachgeschlagen?
-      \textit{Hinweis: Beachten Sie das Beispiel zum Ablesen der Tabelle.}
+      \par\textit{Hinweis: Beachten Sie das Beispiel zum Ablesen der Tabelle.}
       \input{res/des.tbl}
@@
     \begin{enumerate}
-        \item 2-DES bietet nur $ 2^{56} $ statt $ 2^{112} $ effektiver Sicherheit, da ein Meet-in-the-Middle-Angriff $ E_{k_1}(m) = D_{k_2}(c) $ nutzt.
+        \item 2-DES bietet nur $2*2^{56}$ statt $2^{112}$ effektiver Sicherheit, da ein Meet-in-the-Middle-Angriff $E_{k_1}(m) = D_{k_2}(c)$ nutzen kann.
+Somit ist die Sicherheit bei doppelter Schlüssellänge lediglich um den Faktor 2 erhöht.
         \item Analyse aus Tabelle 1:
-        \[
-        k_1 = \texttt{e6f4 a450 4898 d7}, \quad k_2 = \texttt{97a4 3741 8086 0b}
-        \]
-        \item Eine vollständige Tabelle umfasst $ 2^{56} $ Einträge, wobei $ E_k(m) $ oder $ D_k(c) $ für alle $ k $ berechnet werden.
+          \[  k_1 = \texttt{e6f4 a450 4898 d7}, \quad k_2 = \texttt{97a4 3741 8086 0b} \]
+        \item Eine vollständige Tabelle umfasst $2*2^{56}$ Einträge, wobei $E_k(m)$ und $D_k(c)$ für alle $k$ berechnet werden.
     \end{enumerate}
@@
 \begin{exercise}{Triple-DES}
-  Beim Triple-DES-Algorithmus (3-DES) handelt es sich um die dreifache Hintereinanderausführung des DES-
-  Algorithmus. Dabei wird mit zwei 56 Bit langen Schlüsseln $ k_1 $ und $ k_2 $ gearbeitet und beim Verschlüsseln
-  entweder $ E(k_1) \to D(k_2) \to E(k_1) $ (EDE-Modus) oder $ E(k_1) \to E(k_2) \to E(k_1) $ (EEE-Modus) ausgeführt.
+  Beim Triple-DES-Algorithmus (3-DES) handelt es sich um die dreifache Hintereinanderausführung des DES-Algorithmus.
+Dabei wird mit zwei 56 Bit langen Schlüsseln $k_1$ und $k_2$ gearbeitet und beim Verschlüsseln entweder $E(k_1) \to D(k_2) \to E(k_1)$ (EDE-Modus) oder $E(k_1) \to E(k_2) \to E(k_1)$ (EEE-Modus) ausgeführt.
   \begin{enumerate}
     \item Warum wird dem EDE-Modus meist der Vorzug vor dem EEE-Modus gegeben?
-    \item Berechnen Sie die theoretische und die effektive Schlüssellänge des Triple-DES-Algorithmus unter ei-
-    nem Known-Plaintext-Angriff mit drei 56 Bit langen Schlüsseln, d.h. $ E(k_1) \to E(k_2) \to E(k_3) $.
-  \end{enumerate}
-
-  \begin{solution}
-    \begin{enumerate}
-        \item Der EDE-Modus (Encrypt-Decrypt-Encrypt) schützt gegen Meet-in-the-Middle-Angriffe, da die Verschlüsselung durch Entschlüsselung unterbrochen wird.
-        \item Die effektive Schlüssellänge beträgt:
-        \[
-        \text{Theoretisch: } 3 \times 56 = 168 \text{ Bit}, \quad \text{Effektiv: } 112 \text{ Bit (wegen Angriffen).}
-        \]
+    \item Berechnen Sie die theoretische und die effektive Schlüssellänge des Triple-DES-Algorithmus unter einem Known-Plaintext-Angriff mit drei 56 Bit langen Schlüsseln, d.h.
+$E(k_1) \to E(k_2) \to E(k_3)$.
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+        \item Abwärtskompatibilität mit DES bei $k_1=k_2$.
+        \item Theoretisch 168 Bit (3*56 Bit), effektiv 112 Bit (2*56) aufgrund der Anfälligkeit gegenüber Meet-in-the-Middle-Angriffen.
     \end{enumerate}
@@
 \begin{exercise}{Cipher-Block-Chaining-Betriebsmodus}
-  Beim Cipher-Block-Chaining-Betriebsmodus wird ein Klartextblock $ M_i $ vor der Anwendung der Verschlüsselungsfunktion mit dem unmittelbar vorher erzeugten Chiffretextblock $ C_{i-1} $ durch die XOR-Operation verknüpft.
+  Beim Cipher-Block-Chaining-Betriebsmodus wird ein Klartextblock $M_i$ vor der Anwendung der Verschlüsselungsfunktion mit dem unmittelbar vorher erzeugten Chiffretextblock $C_{i-1}$ durch die XOR-Operation verknüpft.
   Die erzeugten Chiffretextblöcke hängen dadurch von ihren Vorgängern ab.
-Bei der Verschlüsselung des ersten Klartextblocks $ M_1 $ wird für $ C_0 $ ein zufällig gewählter Initialisierungsvektor (IV) verwendet.
-
-  Hinweis: Gehen Sie davon aus, dass eine moderne Blockchiffre eingesetzt wird, z. B. AES.
+Bei der Verschlüsselung des ersten Klartextblocks $M_1$ wird für $C_0$ ein zufällig gewählter Initialisierungsvektor (IV) verwendet.
+    \par\textit{Hinweis: Gehen Sie davon aus, dass eine moderne Blockchiffre eingesetzt wird, z. B. AES.}
   \begin{enumerate}
@@
         \item Ein verlorenes Bit im ECB-Modus zerstört den betroffenen Block, aber die folgenden Blöcke bleiben lesbar.
-        \item Im CBC-Modus führt der Verlust einzelner Bits dazu, dass alle nachfolgenden Blöcke falsch entschlüsselt werden, da der Modus auf den vorherigen Block angewiesen ist.
-        \item Bei stromorientierten Betriebsarten (z.
-B.
+        \item Im CBC-Modus führt der Verlust einzelner Bits dazu, dass zusätzlich zum fehlerhaften Block der nachfolgenden Block nicht mehr entschlüsselt werden kann.
+        \item Bei stromorientierten Betriebsarten (zB.
 CTR-Modus) ist die Fehlerfortpflanzung auf den aktuellen Block beschränkt, da kein Feedback aus vorherigen Blöcken verwendet wird.
@@
 \begin{exercise}{Verketteter Cipher-Block-Chaining-Betriebsmodus}
-  Beim verketteten Cipher-Block-Chaining-Betriebsmodus wird zur Steigerung des Datendurchsatzes nur für die erste Nachricht $ M_1 $ ein Initialisierungsvektor $ IV $ gewählt.
-Das heißt für den Schlüsseltext $ C_1 = C_1^0 || \ldots || C_1^{n_1} $ der ersten Nachricht $ M_1 = M_1^1 || \ldots || M_1^{n_1} $ gilt wie beim normalen CBC:
-  \[
-      C_1^0 = IV, \quad C_1^i = E(C_1^{i-1} \oplus M_1^i, K) \quad \text{mit } i = 1, \ldots, n_1.
-  \]
-  Für alle folgenden Nachrichten $ M_2, \ldots $ wird anstelle des $ IV $ der letzte Schlüsseltextblock des vorherigen Schlüsseltextes verwendet.
-Das heißt für den Schlüsseltext $ C^j = C^j_1 || \ldots || C^j_{n_j} $ aller weiteren Nachrichten $ M_j = M_j^1 || \ldots || M_j^{n_j} $ mit $ j > 1 $ gilt:
-  \[
-      C^j_1 = E(C^{j-1}_{n_{j-1}} \oplus M_j^1, K), \quad C^j_i = E(C^j_{i-1} \oplus M_j^i, K) \quad \text{mit } i = 2, \ldots, n_j.
-  \]
+  Beim verketteten Cipher-Block-Chaining-Betriebsmodus wird zur Steigerung des Datendurchsatzes nur für die erste Nachricht $M_1$ ein Initialisierungsvektor $IV$ gewählt.
+Das heißt für den Schlüsseltext $C_1 = C_1^0 || \ldots || C_1^{n_1}$ der ersten Nachricht $M_1 = M_1^1 || \ldots || M_1^{n_1}$ gilt wie beim normalen CBC:
+    \[ C_1^0 = IV, \quad C_1^i = E(C_1^{i-1} \oplus M_1^i, K) \quad \text{mit } i = 1, \ldots, n_1. \]
+  Für alle folgenden Nachrichten $M_2, \ldots$ wird anstelle des $IV$ der letzte Schlüsseltextblock des vorherigen Schlüsseltextes verwendet.
+Das heißt für den Schlüsseltext $C^j = C^j_1 || \ldots || C^j_{n_j}$ aller weiteren Nachrichten $M_j = M_j^1 || \ldots || M_j^{n_j}$ mit $j > 1$ gilt:
+    \[ C^j_1 = E(C^{j-1}_{n_{j-1}} \oplus M_j^1, K), \quad C^j_i = E(C^j_{i-1} \oplus M_j^i, K) \quad \text{mit } i = 2, \ldots, n_j. \]
   \begin{enumerate}
@@
 Führen Sie die Berechnungen sowohl für einen Block als auch für vier Blöcke pro Nachricht aus.
-Gehen Sie in beiden Fällen von insgesamt $ 2^{20} $ zu verschlüsselnden Nachrichten vor Neugenerierung eines Schlüssels und Neustart mit einem neuen $ IV $ aus.
-    \item Ihnen ist ein Schlüsseltext $ C^j $ in die Hände gefallen, der nur aus zwei möglichen Klartexten $ M_j $ oder $ \tilde{M}_j $ gebildet worden sein kann.
-Erläutern Sie, wie Sie ermitteln könnten, welcher der beiden Klartexte verschlüsselt wurde, wenn Sie den Klartext $ M_{j+1} $ für die nächste Nachricht selbst wählen dürfen.
+Gehen Sie in beiden Fällen von insgesamt $2^{20}$ zu verschlüsselnden Nachrichten vor Neugenerierung eines Schlüssels und Neustart mit einem neuen $IV$ aus.
+    \item Ihnen ist ein Schlüsseltext $C^j$ in die Hände gefallen, der nur aus zwei möglichen Klartexten $M_j$ oder $\tilde{M}_j$ gebildet worden sein kann.
+Erläutern Sie, wie Sie ermitteln könnten, welcher der beiden Klartexte verschlüsselt wurde, wenn Sie den Klartext $M_{j+1}$ für die nächste Nachricht selbst wählen dürfen.
     \item Beschreiben Sie ein Szenario für einen derartigen Angriff.
@@
     \begin{enumerate}
-        \item Bei einem verketteten CBC-Modus steigt der Datendurchsatz, da kein neuer IV benötigt wird.
-Für $ 2^{20} $ Nachrichten bei einer Blockgröße von $ 128 $ Bit beträgt der Effizienzgewinn ca.
-0,01\%, da der Initialisierungsaufwand entfällt.
-        \item Wenn $ M_j $ oder $ \tilde{M}_j $ verschlüsselt wurde, kann durch die Manipulation des nächsten Klartextblocks $ M_{j+1} $ festgestellt werden, welcher Klartext verwendet wurde.
-Dies ermöglicht einen Known-Plaintext-Angriff.
-        \item Ein mögliches Szenario: Ein Angreifer, der Zugriff auf den Kommunikationskanal hat, modifiziert gezielt den nächsten Klartextblock, um Rückschlüsse auf vorherige Blöcke zu ziehen.
+        \item Bei einem verketteten CBC-Modus steigt der Datendurchsatz, da kein neuer IV benötigt wird. Für einen Block pro Nachricht: $100\%$, für vier Blöcke pro Nachricht: $25\%$.
+        \item Wählen Sie $M_{j+1}^0=M_j^0 \oplus C_j^{-1}$ als Klartext für den ersten Block der nächsten Nachricht.
+Das ergibt $C_{j+1}^0=E(C_j^{-1} \oplus M_{j+1}^0)=E(C_j^{-1} \oplus M_j^0 \oplus C_j^{-1})=E(M_j^0)=C_j^0$, wenn $M_j$ verschlüsselt wurde.
+        \item Ein Angreifer, der Zugriff auf den Kommunikationskanal hat, modifiziert gezielt den nächsten Klartextblock, um Rückschlüsse auf vorherige Blöcke zu ziehen.
         \item Selbst kleinste Änderungen im Verfahren können signifikante Sicherheitsrisiken schaffen.

### 2025-01-15 SbD/exercises.tex
Updated exercises.

@@
   \begin{enumerate}
-    \item Welche Verbindungen könnten von einem sehr starken Angreifer wie einem Nachrichtendienst noch entschlüsselt werden, wenn dieser an $1$ gelangt? Begründen Sie Ihre Antwort.
+    \item Welche Verbindungen könnten von einem sehr starken Angreifer wie einem Nachrichtendienst noch entschlüsselt werden, wenn dieser an $d$ gelangt? Begründen Sie Ihre Antwort.
     \item Skizzieren Sie einen Angriff, der möglich wäre, wenn die Kundin die Signatur nicht prüft.
@@
     \begin{enumerate}
-      \item Ein starker Angreifer kann alle Verbindungen entschlüsseln, wenn er $d$ erhält, da er damit alle geheimen Schlüssel $d_e$ berechnen kann.
-      \item Es könnten gefälschte öffentliche Schlüssel $c_e$ und Signaturen an die Kundin gesendet werden, die nicht von Alice stammen.
-      \item Durch die Verwendung der gleichen Primzahlen $p$ und $q$ für $N$ und $N_e$ kann ein Angreifer $d$ berechnen, wenn er $d_e$ kennt. Dies ermöglicht das Entschlüsseln aller Verbindungen.
+      \item Bei Kenntnis von $d$ können mittels einem Man-in-the-Middle Angriff neue Schlüssel $c_e$ abgefangen und durch eigene ersetzt werden.
+Dadurch kann der Angreifer alle nachfolgenden Verbindungen entschlüsseln, jedoch nicht die alten.
+      \item Es könnten gefälschte öffentliche Schlüssel $c_e$ mit falscher/ohne Signatur an die Kundin gesendet werden.
+Diese würde die Verbindung aufbauen und der Angreifer könnte die Kommunikation mithören.
+      \item Durch die Verwendung der gleichen Primzahlen $p$ und $q$ für $N_e$ kann eine Common Modulus Attack durchgeführt werden.
+Der Angreifer kann die Nachrichten entschlüsseln.
+EEA auf $a\cdot c_{e1}+b\cdot c_{e2} \equiv 1$ anwenden und dann $m=c_{e1}^a\cdot c_{e2}^b \pmod{N_e}$ berechnen.
     \end{enumerate}
@@
           Gemeinsamer Schlüssel: $K\equiv B^a\equiv A^b\equiv 4^2\equiv 3^3\equiv 5 \pmod{11}$
-        \item Sicherheit basiert auf dem diskreten Logarithmusproblem (DLP). Der Angreifer kann $g^a$ und $g^b$ nicht effizient in $a$ und $b$ umwandeln.
+        \item Sicherheit basiert auf dem Problem des diskreten Logarithmus.
+Der Angreifer kann keine diskreten Logarithmen effizient berechnen.
+Annahmen: Passiver Angreifer, der nur den Verkehr mithören kann.
         \item Schutz vor aktiven Angriffen: Vorab-Austausch von $g, p$ und Authentifizierung der öffentlichen Schlüssel (z. B. durch Signaturen).
@@
     \end{enumerate}
-    Die zweite Variante ist sicherer, aufgrund der längeren Teilschlüssel. Konkatenation ist allerdings nicht grundsätzlich sicherer als XOR-Verknüpfung.
+    Die zweite Variante ist sicherer, aufgrund der längeren Teilschlüssel. Konkatenation ist allerdings nicht grundsätzlich unsicherer als XOR-Verknüpfung.
   \end{solution}
@@
     \begin{enumerate}
-      \item Im Schalenmodell müssen alle Zertifikate und Signaturen zum Zeitpunkt der Überprüfung gültig sein. Im Kettenmodell müssen diese lediglich zum Zeitpunkt der Ausstellung gültig sein.
-      \item Das Kettenmodell wird in der Praxis bevorzugt, da es flexibler ist und die Gültigkeit von Zertifikaten und Signaturen nicht permanent überprüft werden muss.
-      \item Schalenmodell: Zertifikatsausstellung bis Ablauf Wurzelzertifikat. Kettenmodell: Zertifikatsausstellung bis Ablauf Alice's Zertifikat.
-      \item In beiden Modellen gültig, da Zertifikatskette und Signatur zum Zeitpunkt der Ausstellung gültig waren.
+      \item Im Schalenmodell müssen alle Zertifikate und Signaturen zum Zeitpunkt der Signatur gültig sein.
+Im Kettenmodell muss lediglich das signierende Zertifikat zum Zeitpunkt der Signatur gültig sein.
+      \item Das Kettenmodell wird in der Praxis bevorzugt, da weniger Zertifikate überprüft werden müssen.
+      \item Schalenmodell: Signatur bis Ablauf Wurzelzertifikat. Kettenmodell: Signatur bis Ablauf Alices Zertifikat.
+      \item In beiden Modellen gültig, da Zertifikatskette zum Zeitpunkt der Signatur gültig waren.
       \item Nur noch im Kettenmodell gültig, da im Schalenmodell die Kette nicht mehr gültig ist.

### 2025-01-15 NLP/exercises.tex
Small update.

@@
               \item $P@10 = \frac{7}{10} = .7$
-              \item $MA@5 = (\frac{1}{1} + \frac{2}{3} + \frac{3}{5}) / 3 = .756$
+              \item $MAP@5 = (\frac{1}{1} + \frac{2}{3} + \frac{3}{5}) / 3 = .756$
             \end{itemize}
@@
         \item \textbf{Boosting document recall}: Index more documents or reduce relevance thresholds.
-
     \end{itemize}
@@
 \begin{exercise}{TREC}\label{ex:trec}
-  What are the TREC question types?
+  What are the 6 TREC question types?
 
   \begin{solution}
-    \begin{itemize}
-        \item \textbf{Factoid questions}: Require specific factual answers (e.g., "Who is the president of the U.S.?").
-        \item \textbf{List questions}: Require a list of items (e.g., "What countries are in the EU?").
-        \item \textbf{Definition questions}: Seek explanations or definitions (e.g., "What is AI?").
-    \end{itemize}
+    Abbreviation, Entity, Description, Human, Location, Numeric.
   \end{solution}

### 2025-01-15 SbD/exercises.tex
Added last exercise sheet.

@@
 \setcounter{subsection}{100}
-% Aufgabe 1 Spoofing
-% a) Inwiefern kann ein DNS-Spoofing-Angriff für einen Distributed Denial of Service (DDoS)-Angriff ausge-
-% nutzt werden? Erklären Sie anschaulich.
-% b) Warum funktioniert ARP-Spoofing nur in einem lokalen Netz?
-% c) Sie vermuten, Opfer eines ARP-Spoofing-Angriffs zu sein. Was können Sie an Ihrem Rechner überprü-
-% fen, um einen solchen Angriff erkennen zu können?
 \begin{exercise}{Spoofing}
@@
 
-% Aufgabe 4 Challenge-Response-Authentifikation überarbeitet
-% In der klassischen beidseitigen Challenge-Response-Authentifikation mittels symmetrischer, deterministi-
-% scher Verschlüsselung sendet ein prüfendes IT-System eine Zufallszahl als Challenge und erwartet deren
-% Verschlüsselung mit einem gemeinsamen Schlüssel K. Nachfolgend findet sich eine formale Darstellung:
-% Lokal für Kanal 1: A wählt und speichert vorläufig RA
-% Kanal 1: A RA −−−−→B
-% Lokal für Kanal 1: B wählt und speichert vorläufig RB
-% Lokal für Kanal 1: B berechnet CA := EK(RA)
-% Kanal 1: A RB,CA ←−−−−B
-% ?
-% Lokal für Kanal 1: A testet CA
-% = EK(RA) (oder: DK(CA) ?
-% = RA) und löscht RA
-% Lokal für Kanal 1: A berechnet CB := EK(RB)
-% Kanal 1: A CB −−−−→B
-% Lokal für Kanal 1: B testet CB
-% ?
-% = EK(RB) (oder: DK(CB) ?
-% = RB) und löscht RB
-% Der verwendete Schlüssel K darf hierbei nur dem Geprüftem und dem prüfenden IT-System bekannt sein.
-% Wo immer Sie im Rahmen dieser Aufgabe ein Protokoll beschreiben sollen, verwenden Sie diese formale
-% Darstellung. Bezeichnen Sie für eine asymmetrische Verschlüsselung mit PK den öffentlichen und mit SK den
-% geheimen Schlüssel, sowie mit EPK(·) die Ver- und mit DSK(·) die Entschlüsselung.
-% a) Wieso werden Zufallszahlen und nicht feste Werte verwendet?
-% b) Gegen welche Art von Angriff, bei welcher der Angreifer nur mit einer Partei kommunizieren muss, ist
-% die klassiche Challenge-Reponse-Authentifikation wie oben dargestellt verwundbar? Beschreiben Sie
-% einen erfolgreichen Angriff.
-% c) Wie lässt sich der Angriff aus b) durch dauerhaftes statt vorläufiges Speichern der versandten Challen-
-% ges und entsprechende Prüfungen vermeiden?
-% d) Beschreiben Sie eine überarbeitete Challenge-Response-Authentifikation mittels symmetrischer Ver-
-% schlüsselung, die gegen Angriffe wie in b) geschützt ist und keine dauerhafte Speicherung von Daten
-% erfordert.
-% e) Die klassische Challenge-Response-Authentifikation mittels symmetrischer Verschlüsselung wendet
-% den Verschlüsselungsalgorithmus zweimal an. Beschreiben Sie ein Protokoll zur gegenseitigen Au-
-% thentifikation mit lediglich einer Anwendung eines symmetrischen Verschlüsselungsalgorithmus.
-% f) Beschreiben Sie ein Challenge-Response-Protokoll zur gegenseitigen Authentifkation mit asymmetri-
-% scher statt symmetrischer Verschlüsselung. Was muss gelten damit das Protokoll gegen Angriffe wie in
-% b) geschützt ist?
 \begin{exercise}{Challenge-Response-Authentifikation überarbeitet}
@@
 
-% \setcounter{subsection}{110}
-% \begin{exercise}{}
-%  % TODO
-%
-%   \begin{solution}
-%     % TODO
-%   \end{solution}
-% \end{exercise}
+\setcounter{subsection}{110}
+\begin{exercise}{Kommutative Verschlüsselung}
+  Kommutative Verschlüsselung erlaubt das mehrfache Ver- und anschließende Entschlüsseln durch unterschiedliche Parteien in beliebiger Reihenfolge, also
+  \begin{align*}
+    D_{K_2}(D_{K_1}(E_{K_2}(E_{K_1}(M)))) &= M\\
+    D_{SK_2}(D_{SK_1}(E_{PK_2}(E_{PK_1}(M)))) &= M
+  \end{align*}
+  für Verschlüsselungen, denen man nicht ansieht, wie oft verschlüsselt wurde. Diese Eigenschaft gilt zuzüglich zur allgemein für Verschlüsselungen gültigen Eigenschaft
+  \begin{align*}
+    D_{K_i}(E_{K_i}(M)) &= M\\
+    D_{SK_i}(E_{PK_i}(M)) &= M
+  \end{align*}
+  \begin{enumerate}
+    \item Zeigen Sie, dass folgende Verschlüsselungen kommutativ entsprechend der obigen Definition sind:
+      \begin{enumerate}
+        \item CTR, wenn Initialisierungsvektoren mit dem geheimen Schlüssel ausgetauscht wurden und daher nicht mit dem Schlüsseltext mitgesendet werden müssen (Sie dürfen der Einfachheit halber annehmen, dass alle Parteien dieselbe Blockgröße verwenden.)
+        \item RSA, wenn alle Parteien denselben Modulus verwenden
+      \end{enumerate}
+    \item Welche weitere Kommutativitätseigenschaft weisen CTR und RSA mit den Spezifikationen aus a) auf?
+    \item Beschreiben Sie eine Möglichkeit, aus ElGamal eine kommutative Verschlüsselung zu machen.
+Überlegen Sie dazu, wie eine Kommutativitätsdefinition für eine Verschlüsselung aussehen kann, der anzusehen ist, wie oft verschlüsselt wurde.
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+      \item
+      \begin{enumerate}
+        \item CTR ist kommutativ, wenn Initialisierungsvektoren mit dem geheimen Schlüssel ausgetauscht wurden, da die Verschlüsselung unabhängig vom Schlüsseltext ist.
+        \item RSA ist kommutativ, wenn alle Parteien denselben Modulus verwenden, da die Verschlüsselung unabhängig vom Schlüsseltext ist.
+      \end{enumerate}
+      \item CTR und RSA sind zusätzlich zu den obigen Eigenschaften auch homomorph.
+      \item ElGamal kann kommutativ gemacht werden, indem der öffentliche Schlüssel $PK$ durch den geheimen Schlüssel $SK$ ersetzt wird.
+Eine mögliche Kommutativitätsdefinition könnte lauten: $EK_1(EK_2(M)) = EK_2(EK_1(M))$.
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Geburtstagsangriff auf CBC}
+  In dieser Aufgabe betrachten Sie einen Geburtstagsangriff auf den Cipher-Block-Chaining-Betriebsmodus, also einen Angriff, der die hohe Wahrscheinlichkeit eines Mehrfachauftreten desselben Wertes bei stochastisch unabhängigen Ziehungen ausnutzt.
+  \par\textit{Hinweis: Bitte gehen Sie im Folgenden davon aus, dass die Blockchiffre von hoher Güte ist, also $E(\cdot, K)$ mit Schlüssel $K$ als eine zufällige Permutation modelliert werden kann.
+Das heißt insbesondere, Sie können annehmen, dass $E(\cdot, K)$ für unterschiedliche Klartexte stochastisch unabhängige Schlüsseltexte ausgibt.}
+
+  \begin{enumerate}
+    \item\label{ex:attack} Bitte entwickeln und beschreiben Sie einen Angriff auf CBC, mit dem ein passiver, beobachtender Angreifer einen Schlüsseltextblock mit ihm zunächst unbekanntem Klartext entschlüsseln kann, wenn alle der folgenden Annahmen gelten:
+      \begin{itemize}
+        \item Der Angreifer kann Schlüsseltexte abhören, welche alle mit demselben Schlüssel generiert wurden.
+        \item Der Angreifer kennt für jeden abgehörten Schlüsseltext den vorderen $p$-ten Anteil des zugrunde liegenden Klartexts.
+        \item Mindestens ein Schlüsseltextblock innerhalb des vorderen $p$-ten Anteils ohne den Initialisierungsvektor gleicht einem Schlüsseltextblock innerhalb des hinteren $(1-p)$-ten Anteils ohne den Initialisierungsvektor.
+      \end{itemize}
+      Nachfolgend findet sich ein Beispiel für drei Schlüsseltexte aus je acht Blöcken ohne den Initialisierungsvektor, die die obige Bedingung mit $p=25\%$ erfüllen:
+      \begin{center}
+        \begin{tabular}{c ccccccccc}
+          $IV_1$ & $C_{1,1}$ & $C_{1,2}$ & $C_{1,3}$ & $C_{1,4}$ & $C_{1,5}$ & $C_{1,6}$ & $C_{1,7}$ & $C_{1,8}$\\
+          & $M_{1,1}$ & $M_{1,2}$ & $X$ & $X$ & $X$ & $X$ & $X$ & $X$\vspace{.5em}\\
+          $IV_2$ & $C_{2,1}$ & $C_{2,2}$ & $C_{2,3}$ & $C_{2,4}$ & $C_{2,5}$ & $C_{2,6}$ & $C_{2,7}$ & $C_{2,8}$\\
+          & $M_{2,1}$ & $M_{2,2}$ & $X$ & $X$ & $X$ & $X$ & $X$ & $X$\vspace{.5em}\\
+          $IV_3$ & $C_{3,1}$ & $C_{3,2}$ & $C_{3,3}$ & $C_{3,4}$ & $C_{3,5}$ & $C_{3,6}$ & $C_{3,7}$ & $C_{3,8}$\\
+          & $M_{3,1}$ & $M_{3,2}$ &  $X$ & $X$ & $X$ & $X$ & $X$ & $X$\\
+        \end{tabular}
+      \end{center}
+    \item Berechnen Sie unter Berücksichtigung der Reihenfolge der Ziehung die Wahrscheinlichkeit, dass bei $m$ zufälligen, stochastisch unabhängigen Ziehungen aus einer Menge $\{1, \ldots, n\}$ mindestens zweimal die gleiche Zahl gezogen wird.
+    \item Geben Sie die Wahrscheinlichkeit an, dass mindestens einmal Gleichheit von unterschiedlichen Schlüsseltextblöcken auftritt, wenn der Angreifer $N$ Schlüsseltexte mit einer Blockgröße von $B$ Bit und $L + 1$ Blöcken pro Schlüsseltext (also $L$ vielen ohne den Initialisierungsvektor) abhört.
+    \item Berechnen Sie eine untere Grenze für die Anzahl an abzuhörenden Schlüsseltexten unter demselben Schlüssel, sodass mindestens einmal Gleichheit von unterschiedlichen Schlüsseltextblöcken mindestens mit Wahrscheinlichkeit $q$ auftritt.
+    \item\label{ex:lowerbound} Berechnen Sie eine untere Grenze für die Anzahl an abzuhörenden (und mit demselben Schlüssel generierten) Schlüsseltexten, sodass mindestens einmal Gleichheit eines vorderen Schlüsseltextblocks und eines hinteren Schlüsseltextblocks (also wie in \ref{ex:attack}) mindestens mit Wahrscheinlichkeit $q$ auftritt.
+    \item Berechnen Sie die untere Grenze aus \ref{ex:lowerbound} für $B = 64$, $L = 4$, $p = 50\%$ und $q = 50\%$. Geben Sie die unteren Grenzen auch als Zweierpotenzen an.
+    \item Berechnen Sie die untere Grenze aus \ref{ex:lowerbound} für $B = 64$, $L = 4$, $p = 50\%$ und $q = 99\%$. Geben Sie die unteren Grenzen auch als Zweierpotenzen an.
+    \item Berechnen Sie jeweils, wie viel Zeit und Speicherplatz der Angriff aus \ref{ex:attack} bei Parameterwahl wie in \ref{ex:lowerbound} benötigt.
+Bewerten Sie darauf aufbauend die realistische Durchführbarkeit des Angriffs aus \ref{ex:attack}.
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+      \item Der Angreifer kann den Klartextblock $M2,1$ entschlüsseln, da $C2,1 = C3,1$ gilt. Der Angreifer berechnet $M2,1 = D_{SK}(C2,1 \oplus IV2) = D_{SK}(C3,1 \oplus IV3)$.
+      \item Die Wahrscheinlichkeit, dass bei $m$ Ziehungen aus einer Menge von $n$ Zahlen mindestens zweimal die gleiche Zahl gezogen wird, beträgt $1 - \frac{n!}{(n-m)!n^m}$.
+      \item Die Wahrscheinlichkeit, dass mindestens einmal Gleichheit von unterschiedlichen Schlüsseltextblöcken auftritt, beträgt $1 - \left(1 - \frac{1}{2^B}\right)^{N \cdot L}$.
+      \item Die untere Grenze für die Anzahl an abzuhörenden Schlüsseltexten beträgt $\lceil \frac{\log(1 - q)}{L \cdot \log(1 - 1/2^B)} \rceil$.
+      \item Die untere Grenze für die Anzahl an abzuhörenden Schlüsseltexten beträgt $\lceil \frac{\log(1 - q)}{2 \cdot L \cdot \log(1 - 1/2^B)} \rceil$.
+      \item Die untere Grenze beträgt $2^{10}$.
+      \item Die untere Grenze beträgt $2^{16}$.
+      \item Der Angriff benötigt $2^{16} \cdot 64$ Bit Speicherplatz und $2^{16} \cdot 64$ Bit Zeit. Der Angriff ist aufgrund des hohen Speicherplatzbedarfs nicht realistisch durchführbar.
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
 

### 2025-01-15 BAI/exercises.tex
Added current state.
Added some lecture questions.

@@
 
-\section[2024]{Mock exam}
+\sheet{Lecture Questions}
+\begin{exercise}{Spiking Neural Networks}
+  \begin{enumerate}
+    \item Explain the concept of spiking neural networks and how they resemble biological information transmission.
+    \item What is the difference between temporal coding and static learning?
+    \item How is learning in spiking neural networks based on synaptic correlations?
+    \item What is Spike-Timing Dependent Plasticity?
+    \item Differentiate rate-coded, integrate-and-fire, and spike-response models.
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+      \item Spiking neural networks are a type of artificial neural network that model the behavior of biological neurons.
+They use spikes, or action potentials, to transmit information between neurons, similar to how biological neurons communicate.
+      \item Temporal coding is a method of encoding information in the timing of spikes, while static learning is a method of encoding information in the strength of synaptic connections.
+      \item Learning in spiking neural networks is based on synaptic correlations, which are the correlations between the timing of pre- and post-synaptic spikes.
+      \item Spike-Timing Dependent Plasticity is a learning rule that strengthens or weakens synaptic connections based on the timing of pre- and post-synaptic spikes.
+      \item Rate-coded models use firing rates to encode information, integrate-and-fire models use membrane potentials to encode information, and spike-response models use spike times to encode information.
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+\sheet[2024]{Mock exam}
 
@@
 
-\section[2023]{1st exam}
+\sheet[2023]{1st exam}
 
@@
 
-\section[2022]{1st exam}
+\sheet[2022]{1st exam}
 
@@
     \item In which models is backpropagation used for learning?
-
       \begin{itemize}
@@
     \item What does "embodied" mean for language?
-
       \begin{itemize}
@@
     \item What are the two streams and their tasks in human vision?
-
       \begin{itemize}

### 2025-01-16 PbD/exercises.tex
Updated assignment.

@@
     \begin{itemize}
-      \item \textbf{Confidentiality}: Encryption is used to protect the confidentiality of data.
-      \item \textbf{Availability}: Redundancy is used to ensure the availability of services.
-      \item \textbf{Integrity}: Hash functions are used to ensure the integrity of data.
+      \item \textbf{Confidentiality}: Encryption is used to protect the confidentiality of data by encoding it so that only authorized parties can access it.
+      \item \textbf{Availability}: Redundancy is used to ensure the availability of services by providing multiple instances of the same service.
+      \item \textbf{Integrity}: Hash functions are used to ensure the integrity of data by comparing the hash of the original data with the hash of the received data.
     \end{itemize}
@@
       \item The general purpose of pseudonyms is to protect the identity of users.
-      \item The difference between pseudonymity and anonymity is that pseudonymity uses a pseudonym to protect the identity of users, while anonymity does not require a pseudonym.
-      \item Three applications in which pseudonyms are used are:
-        \begin{itemize}
-          \item E-Mail: Pseudonyms are used to protect the identity of users.
+      \item In anonymity the identity of the user is not known, while in pseudonymity the identity of the user is known but protected.
+      \item \begin{itemize}
+          \item E-Mail: E-Mail addresses are used to send messages.
           \item Social Media: Pseudonyms are used to protect the identity of users.
-          \item Online Shopping: Pseudonyms are used to protect the identity of users.
+          \item Online Banking: IBAN is used to facilitate transactions.
         \end{itemize}
@@
     \item What does indeterministic encryption mean?
-    \item How do you implement indeterministic encryption in an asymmetric cryptosystem?
-    \item What changes compared to b) when using a symmetric cryptosystem?
+    \item\label{ex:asymmetric} How do you implement indeterministic encryption in an asymmetric cryptosystem?
+    \item What changes compared to \ref{ex:asymmetric} when using a symmetric cryptosystem?
     \item You are tasked with implementing indeterministic encryption. The algorithms RSA and ElGamal are available. How would you proceed with each algorithm?
@@
       \item Indeterministic encryption in an asymmetric cryptosystem is implemented by adding randomness to the encryption process.
-      \item In a symmetric cryptosystem, indeterministic encryption is implemented by adding randomness to the encryption process.
+      \item Nothing changes when using a symmetric cryptosystem compared to an asymmetric cryptosystem.
       \item To implement indeterministic encryption with RSA, random padding is added to the plaintext before encryption. With ElGamal, a random number is used to generate the encryption key.
@@
 
-% Aufgabe 4 Unobservable retrieval of information
-% Given is a Blind-Message-Service with three servers S1, S2, and S3. The following datasets are stored in the
-% Blind-Message-Service: D0 = 1011, D1 = 0101, D2 = 1100, D3 = 0011, D4 = 0000.
-% a) You want to retrieve dataset D2 without the servers knowing which dataset you are interested in. What
-% messages are transmitted? How do you obtain the requested dataset? Sketch the computational steps.
-% b) You want to ensure, as in task a), that during a query it remains unknown which dataset the client is
-% interested in. However, instead of the Blind-Message-Service, only a single high-bandwidth server is
-% available. The database, however, is relatively small. Which method do you apply?
 \begin{exercise}{Unobservable Retrieval of Information}
@@
   \begin{enumerate}
-    \item You want to retrieve dataset D2 without the servers knowing which dataset you are interested in.
+    \item\label{ex:bms} You want to retrieve dataset D2 without the servers knowing which dataset you are interested in.
 What messages are transmitted?
@@
 Sketch the computational steps.
-    \item You want to ensure, as in task a), that during a query it remains unknown which dataset the client is interested in.
+    \item You want to ensure, as in task \ref{ex:bms}, that during a query it remains unknown which dataset the client is interested in.
 However, instead of the Blind-Message-Service, only a single high-bandwidth server is available.
@@
         \end{itemize}
-      \item To ensure that the client's interest remains unknown, you use the Broadcast method.
+      \item Request all datasets one after the other and discard all datasets except the requested one.
     \end{enumerate}
@@
       \item Two security goals that are violated during tracking at the application layer are confidentiality and anonymity.
-      \item The difference between tracking at the application layer and tracking at the network layer is that tracking at the application layer uses browser data, while tracking at the network layer uses IP data.
-      \item Three fundamental mechanisms to prevent or make tracking at the application layer more difficult are:
-        \begin{itemize}
+      \item Application layer uses browser data, while network layer uses IP data.
+      \item \begin{itemize}
           \item Block cookies: Prevents tracking by websites. Against advertisers. Example: Browser settings.
@@
   \begin{enumerate}
-    \item Explain a possible approach by which a powerful attacker (e.g., an intelligence agency) can determine the IP addresses of users accessing a specific website through a mix-based anonymization system (e.g., Tor).
-    \item Explain a possible approach by which a powerful attacker (e.g., an intelligence agency) can determine which websites a specific user has accessed via a mix-based anonymization system.
-Note: Describe a different method from the one in task a), even if your method in a) could also be applied to task b).
+    \item\label{ex:mix-based-ip} Explain a possible approach by which a powerful attacker (e.g., an intelligence agency) can determine the IP addresses of users accessing a specific website through a mix-based anonymization system (e.g., Tor).
+    \item\label{ex:mix-based-history}  Explain a possible approach by which a powerful attacker (e.g., an intelligence agency) can determine which websites a specific user has accessed via a mix-based anonymization system.
+      \par\textit{Note: Describe a different method from the one in task \ref{ex:mix-based-ip}, even if your method in \ref{ex:mix-based-ip} could also be applied to task \ref{ex:mix-based-history}.}
     \item Explain three techniques by which a web server can recognize a user during their second (subsequent) visit.
@@
       \item A powerful attacker can determine which websites a specific user has accessed via a mix-based anonymization system by using metadata analysis, such as HTTP headers or specific request characteristics.
-      \item Three techniques by which a web server can recognize a user during their second visit are:
-        \begin{itemize}
+      \item \begin{itemize}
           \item Cookies: Store user data in the browser. Against websites. Example: Google Analytics.

### 2025-01-16 SbD/exercises.tex
Updated last exercises.

@@
     \item Wieso werden Zufallszahlen und nicht feste Werte verwendet?
-    \item Gegen welche Art von Angriff, bei welcher der Angreifer nur mit einer Partei kommunizieren muss, ist die klassische Challenge-Response-Authentifikation wie oben dargestellt verwundbar?
+    \item\label{ex:challenge-response-attack} Gegen welche Art von Angriff, bei welcher der Angreifer nur mit einer Partei kommunizieren muss, ist die klassische Challenge-Response-Authentifikation wie oben dargestellt verwundbar?
 Beschreiben Sie einen erfolgreichen Angriff.
-    \item Wie lässt sich der Angriff aus b) durch dauerhaftes statt vorläufiges Speichern der versandten Challenges und entsprechende Prüfungen vermeiden?
-    \item Beschreiben Sie eine überarbeitete Challenge-Response-Authentifikation mittels symmetrischer Verschlüsselung, die gegen Angriffe wie in b) geschützt ist und keine dauerhafte Speicherung von Daten erfordert.
+    \item Wie lässt sich der Angriff aus \ref{ex:challenge-response-attack} durch dauerhaftes statt vorläufiges Speichern der versandten Challenges und entsprechende Prüfungen vermeiden?
+    \item Beschreiben Sie eine überarbeitete Challenge-Response-Authentifikation mittels symmetrischer Verschlüsselung, die gegen Angriffe wie in \ref{ex:challenge-response-attack} geschützt ist und keine dauerhafte Speicherung von Daten erfordert.
     \item Beschreiben Sie ein Protokoll zur gegenseitigen Authentifikation mit lediglich einer Anwendung eines symmetrischen Verschlüsselungsalgorithmus.
     \item Beschreiben Sie ein Challenge-Response-Protokoll zur gegenseitigen Authentifikation mit asymmetrischer statt symmetrischer Verschlüsselung.
-Was muss gelten, damit das Protokoll gegen Angriffe wie in b) geschützt ist?
-  \end{enumerate}
-
-  \begin{solution}
-    % TODO
+Was muss gelten, damit das Protokoll gegen Angriffe wie in \ref{ex:challenge-response-attack} geschützt ist?
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+      \item Zufallszahlen verhindern, dass ein Angreifer die Authentifikation durch Kenntnis fester Werte umgehen kann. Außerdem werden Replay-Angriffe verhindert.
+      \item Reflektionsangriff: Angreifer $M$ sendet irgendeine Challenge $R_M$ an $A$.
+$A$ sendet $C_M$ und $R_A$ an $M$ zurück.
+$M$ sendet neue Challenge $R_M'=R_A$ an $A$.
+$A$ sendet $C_M'=C_A$ an $M$ zurück.
+$M$ sendet für ursprüngliche Challenge $C_M'=C_A$ an $A$ zurück.
+      \item Wiederverwendung von Challenges kann verhindert werden.
+      \item $A$ wählt $R_A$, $B$ wählt $R_B$. $A$ sendet $R_A$ an $B$. $B$ berechnet $C_B=E(R_A || R_B)$ und sendet $C_B$ an $A$. $A$ prüft $C_B$ und sendet $R_B$ an $B$. $B$ prüft $R_B$.
+      \item $A$ wählt $R_A$ und $R_B$, berechnet $C=E(R_A || R_B)$ und sendet $R_A, C$ an $B$. $B$ prüft $C=E(R_A || R_B)$ und sendet $R_B$ an $A$. $A$ prüft $R_B$.
+      \item $A$ wählt $R_A$, $B$ wählt $R_B$.
+$A$ sendet $R_A$ an $B$.
+$B$ sendet $R_B$ und $E_{SK}(R_A)$ an $A$.
+$A$ prüft $D_{SK}(E_{SK}(R_A))=R_A$ und sendet $E_{SK}(R_B)$ an $B$.
+$B$ prüft $D_{SK}(E_{SK}(R_B))=R_B$.
+    \end{enumerate}
   \end{solution}
@@
   \begin{enumerate}
-    \item Zeigen Sie, dass folgende Verschlüsselungen kommutativ entsprechend der obigen Definition sind:
-      \begin{enumerate}
-        \item CTR, wenn Initialisierungsvektoren mit dem geheimen Schlüssel ausgetauscht wurden und daher nicht mit dem Schlüsseltext mitgesendet werden müssen (Sie dürfen der Einfachheit halber annehmen, dass alle Parteien dieselbe Blockgröße verwenden.)
+    \item\label{ex:commutative} Zeigen Sie, dass folgende Verschlüsselungen kommutativ entsprechend der obigen Definition sind:
+      \begin{enumerate}
+        \item CTR, wenn Initialisierungsvektoren mit dem geheimen Schlüssel ausgetauscht wurden und daher nicht mit dem Schlüsseltext mitgesendet werden müssen (Sie dürfen der Einfachheit halber annehmen, dass alle Parteien dieselbe Blockgröße verwenden).
         \item RSA, wenn alle Parteien denselben Modulus verwenden
       \end{enumerate}
-    \item Welche weitere Kommutativitätseigenschaft weisen CTR und RSA mit den Spezifikationen aus a) auf?
+    \item Welche weitere Kommutativitätseigenschaft weisen CTR und RSA mit den Spezifikationen aus \ref{ex:commutative} auf?
     \item Beschreiben Sie eine Möglichkeit, aus ElGamal eine kommutative Verschlüsselung zu machen.
@@
       \end{enumerate}
-      \item CTR und RSA sind zusätzlich zu den obigen Eigenschaften auch homomorph.
-      \item ElGamal kann kommutativ gemacht werden, indem der öffentliche Schlüssel $PK$ durch den geheimen Schlüssel $SK$ ersetzt wird.
-Eine mögliche Kommutativitätsdefinition könnte lauten: $EK_1(EK_2(M)) = EK_2(EK_1(M))$.
+      \item CTR und RSA sind zusätzlich zu den obigen Eigenschaften auch homomorph, was bedeutet, dass die Verschlüsselung von zwei Klartexten gleich der Verschlüsselung der Konkatenation der Klartexte ist.
+      \item Die Verschlüsselung müssen folgendermaßen angepasst werden:
+        \begin{align*}
+          c_{1,i}&=c_{1,i-1}\cdot g^{k_i}\pmod{p} \\
+          c_{2,i}&=c_{2,i-1}\cdot h^{k_i} \pmod{p}
+        \end{align*}
+        Und für die Entschlüsselung:
+        \begin{align*}
+          c_{1,i-1}&=c_{1,i}\cdot g^{k_i} \pmod{p} \\
+          c_{2,i-1}&=c_{2,i}\cdot h^{k_i} \pmod{p} \\
+          s_i&=c_{1,i}^{k_i} \pmod{p} \\
+          m_i&=c_{2,i}\cdot s_i^{-1} \pmod{p}
+        \end{align*}
     \end{enumerate}
@@
     \begin{enumerate}
-      \item Der Angreifer kann den Klartextblock $M2,1$ entschlüsseln, da $C2,1 = C3,1$ gilt. Der Angreifer berechnet $M2,1 = D_{SK}(C2,1 \oplus IV2) = D_{SK}(C3,1 \oplus IV3)$.
-      \item Die Wahrscheinlichkeit, dass bei $m$ Ziehungen aus einer Menge von $n$ Zahlen mindestens zweimal die gleiche Zahl gezogen wird, beträgt $1 - \frac{n!}{(n-m)!n^m}$.
-      \item Die Wahrscheinlichkeit, dass mindestens einmal Gleichheit von unterschiedlichen Schlüsseltextblöcken auftritt, beträgt $1 - \left(1 - \frac{1}{2^B}\right)^{N \cdot L}$.
-      \item Die untere Grenze für die Anzahl an abzuhörenden Schlüsseltexten beträgt $\lceil \frac{\log(1 - q)}{L \cdot \log(1 - 1/2^B)} \rceil$.
-      \item Die untere Grenze für die Anzahl an abzuhörenden Schlüsseltexten beträgt $\lceil \frac{\log(1 - q)}{2 \cdot L \cdot \log(1 - 1/2^B)} \rceil$.
-      \item Die untere Grenze beträgt $2^{10}$.
-      \item Die untere Grenze beträgt $2^{16}$.
-      \item Der Angriff benötigt $2^{16} \cdot 64$ Bit Speicherplatz und $2^{16} \cdot 64$ Bit Zeit. Der Angriff ist aufgrund des hohen Speicherplatzbedarfs nicht realistisch durchführbar.
+      \item Zunächst beobachtet der Angreifer die Schlüsseltexte und wartet auf eine Kollision $C_i=C_j$, wobei $C_i$ im vorderen $p$-Teil ist, $C_j$ im hinteren.
+Es folgt $C_{i-1}\oplus M_i = C_{j-1} \oplus M_j$, da die Verschlüsselung deterministisch ist.
+Nun kann der Angreifer $M_i=C_{i-1}\oplus M_j \oplus C_{j-1}$ berechnen.
+      \item P(Kollision)=1-P(keine Kollision)=$1-\prod_{i=0}^{m-1}{(1-\frac{i}{n})}$, was mit $1-e^{-\frac{m(m-1)}{2n}}$ approximiert werden kann für große $n$ und moderate $m$.
+      \item P(Kollision)=$1-\prod_{i=0}^{NL-1}{(1-\frac{i}{2^B})}$.
+      \item $N=\frac{1+\sqrt{2^B\ln(1-q)}}{2L}$.
+      \item $N=\sqrt{\frac{-\ln{1-q}}{\lfloor pL\rfloor\cdot \lfloor (1-p)L \rfloor\cdot \ln{1-\frac{1}{2^B}}}}$.
+      \item $N=2^{31.5}$.
+      \item $N=2^{32.5}$.
+      \item Für $q=50$: $T=2^{65}$ und $S=2^{39.5}$. Für $q=99$: $T=2^{67}$ und $S=2^{40.5}$.
+        Die Ressourcenanforderungen sind sehr hoch selbst bei $q=50$.
+$2^{65}$ Rechenoperationen würden bei modernen GPUs ($10^12$ ops/sec) immer noch länger als ein Jahr dauern.
+$2^{39.5}$ Speicherplatz, oder 97GB, ist ebenfalls recht hoch aber realistisch für einen Angreifer.
     \end{enumerate}

### 2025-01-16 BAI/exercises.tex
Added more lecture questions.

@@
 
-\sheet[2024]{Mock exam}
+\begin{exercise}{Computational Neural Networks}
+  \begin{enumerate}
+    \item Map these biological neural structures to artificial neural networks: neurons/synapses, action potentials, plasticity, and dopamine.
+    \item Explain the concept of backpropagation and how it resembles biological plasticity.
+    \item Describe two different ANN architectures and their applications.
+    \item What are some examples of useful applications of artificial neural networks?
+  \end{enumerate}
 
-% TODO
+  \begin{solution}
+    \begin{enumerate}
+      \item Neurons and synapses in biological neural structures correspond to nodes and weighted connections in artificial neural networks.
+Action potentials in biological neurons correspond to activation values in artificial neural networks.
+Plasticity in biological synapses corresponds to backpropagation in artificial neural networks.
+Dopamine in biological reward systems corresponds to reward signals in artificial neural networks.
+      \item Backpropagation is a learning algorithm used in artificial neural networks to adjust the weights of connections between neurons based on the error in the output.
+This process resembles biological plasticity, where synaptic connections are strengthened or weakened based on the correlation between pre- and post-synaptic activity.
+      \item Two different ANN architectures are convolutional networks, which are used for image processing tasks, and recurrent networks, which are used for sequential data processing tasks.
+      \item Some examples of useful applications of artificial neural networks include speech and language processing, image recognition, and natural language understanding.
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
 

### 2025-01-20 SbD/exercises.tex
Added the mock exam.

@@
 
-\section{Assignments}
+\sheet{Assignments}
 
@@
 
-\section{Catalogue}
+\sheet{Catalogue}
 
@@
 
-\setcounter{section}{2023}
-\section{Mock exam}
-
-% \begin{exercise}{}
-%  % TODO
-%
-%   \begin{solution}
-%     % TODO
-%   \end{solution}
-% \end{exercise}
+\sheet[2024]{Mock exam}
+
+\begin{exercise}{Grundlagen der Kryptographie}
+  \begin{enumerate}
+    \item Erläutern Sie zwei Ursachen, die dafür sorgen, dass fast alle kryptographischen Algorithmen mit der Zeit durch andere ersetzt werden müssen.
+    \item Sie möchten einen Text so verschlüsseln, dass er auch in 96 Jahren nicht ohne Kenntnis des von Ihnen versteckten Schlüssels entschlüsselt werden kann.
+Welches Verfahren wählen Sie dazu?
+Begründen Sie Ihre Entscheidung.
+Warum wird das Verfahren heutzutage nicht überall benutzt?
+    \item In der Vorlesung wurden verschiedene klassische Chiffren behandelt, darunter Substitutionschiffren wie die Caesar-Chiffre (Verschiebechiffre), das Schema von Polybios, das Vigenère-Verfahren und die Vernam-Chiffre.
+Zudem wurden Transpositionschiffren wie die Skytala betrachtet.
+Ihnen liegt ein Geheimtext vor, der mit einer dieser Chiffren erzeugt wurde.
+Ihnen ist bekannt, dass es sich um die Verschlüsselung eines deutschen Textes handelt.
+Wie können Sie - ohne Rechnerunterstützung - vorgehen, um einzugrenzen, welche Chiffre verwendet wurde?
+Erläutern Sie die jeweiligen Ansätze.
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+        \item \begin{itemize}
+            \item \textbf{Entwicklung von Angriffsmethoden:} Fortschritte in der Kryptoanalyse können Algorithmen angreifbar machen.
+            \item \textbf{Steigende Rechenleistung:} Mit der Zeit wird es möglich, Algorithmen durch Brute-Force-Angriffe zu brechen.
+        \end{itemize}
+        \item Ein One-Time-Pad ist sicher, da es perfekte Sicherheit bietet. Es wird jedoch nicht überall benutzt, da es sehr ineffizient ist und der Schlüsselaustausch sehr aufwendig ist.
+        \item Häufigkeitsanalyse und n-Gram-Analyse können helfen Substitutions- von Transpositionschiffren zu unterscheiden.
+Wenn Buchstabenhäufigkeiten mit der deutschen Sprache übereinstimmen, könnte es sich um eine Transpositionschiffre handeln.
+Wenn viele n-Gramme wiederholt auftreten, könnte es sich um eine Substitutionschiffre handeln.
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Konzelationssystem nach ElGamal}
+  Alice möchte Bob die 6 Zeichen lange Nachricht "CYBER!" senden.
+Um ein Abhören durch Eve zu vermeiden, verschlüsselt sie die Nachricht mit dem Verschlüsselungsverfahren nach ElGamal.
+Hierzu ersetzt sie zunächst mit Hilfe der untenstehenden Tabelle jedes Zeichen durch eine Zahl, um diese dann einzeln zu verschlüsseln.
+Alice und Bob einigen sich auf die Primzahl $p=29$ und den Generator $a=2$ (primitive Wurzel von $p$).
+Bobs öffentlicher Schlüssel lautet $y_B=4$.
+  \begin{enumerate}
+    \item Verwenden Sie die folgende Tabelle, um die Zeichen der Nachricht durch Zahlen zu ersetzen:
+      \begin{table}
+        \centering
+        \begin{tabular}{cccccccc}
+          A: 1 & E: 5 & I: 9 & M: 13 & Q: 17 & U: 21 & Y: 25 \\
+          B: 2 & F: 6 & J: 10 & N: 14 & R: 18 & V: 22 & Z: 26 \\
+          C: 3 & G: 7 & K: 11 & O: 15 & S: 19 & W: 23 & !: 27 \\
+          D: 4 & H: 8 & L: 12 & P: 16 & T: 20 & X: 24 & \\
+        \end{tabular}
+      \end{table}
+    \item\label{ex:encrypt} Verschlüsseln Sie jedes einzelne ersetzte Zeichen mit Bobs öffentlichem Schlüssel. Wählen Sie die Zufallszahl $z=5$. Geben Sie den Rechenweg an.
+    \item Auf welcher Annahme basiert die Sicherheit der Verschlüsselung nach ElGamal?
+    \item Bob benötigt zum Entschlüsseln Informationen über die Zufallszahl $z$, die Alice benutzt hat. Nach welcher Formel werden diese Informationen berechnet? Wie werden diese an Bob übermittelt?
+    \item Wenn statt ElGamal ein deterministisches Verschlüsselungsverfahren genutzt würde, wäre das Vorgehen aus Teilaufgabe \ref{ex:encrypt} unsicher.
+Benennen und erläutern Sie den Angriff, der dann aufgrund der Kenntnis des öffentlichen Schlüssels von Bob möglich wäre.
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+        \item Die Nachricht "CYBER!" wird ersetzt durch die Zahlen $3, 25, 2, 5, 18, 27$.
+        \item Die Zufallszahl $z=5$ wird gewählt.
+Die Verschlüsselung erfolgt durch $c_1=2^5 \mod{29}=3$, $c_2=4^5 \cdot 3 \mod{29}=23$, $c_3=4^5 \cdot 25 \mod{29}=24$, $c_4=4^5 \cdot 2 \mod{29}=4$, $c_5=4^5 \cdot 5 \mod{29}=6$, $c_6=4^5 \cdot 18 \mod{29}=5$.
+        \item Die Sicherheit basiert auf der Schwierigkeit des diskreten Logarithmusproblems.
+        \item Die Zufallszahl $z$ wird berechnet durch $z=c_1^{29-1-4} \mod{29}$ und an Bob übermittelt.
+        \item Es würde Anfälligkeit für einen Chosen-Plaintext-Angriff bestehen. Eve könnte $c_1$ berechnen und die Nachrichten entschlüsseln, die Alice an Bob sendet.
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{TLS}
+  Alice möchte eine Internetseite abrufen.
+Durch einen Man-in-the-Middle-Angriff können Angreifende eine Verschlüsselung der Verbindung unterbinden.
+Dabei befindet sich die angreifende Person in der Netzkommunikation zwischen Alice und dem Server der Internetseite.
+  \begin{enumerate}
+    \item Beschreiben Sie einen Angriff, mit dem die Verschlüsselung der Verbindung unterbunden werden kann. Wie kann Alice einen solchen Angriff erkennen?
+    \item Der Angriff soll nun verbessert werden, indem die angreifende Person versucht, ein für die Internetseite gültiges Zertifikat vorzuweisen, statt die Verschlüsselung zu unterbinden.
+Unter welchen Voraussetzungen und wodurch wird dieser Angriff bereits verhindert?
+    \item Zu Audit-Zwecken soll in einem großen Unternehmen auch der via TLS verschlüsselte Datenverkehr zwischen Unternehmensrechnern und externen Servern überwacht werden.
+Die IT-Abteilung des Unternehmens besitzt die vollständige Kontrolle über die innerhalb des Unternehmens installierten Systeme (Computer, Server, Netzinfrastruktur).
+Beschreiben Sie eine Vorgehensweise, die es der IT-Abteilung erlaubt, den gesamten TLS-Datenverkehr einschließlich der Inhaltsdaten auf einem zentralen System innerhalb des Firmennetzes mitzuschneiden.
+Führen Sie dazu die Schritte auf, die erforderlich sind, um auf die Inhaltsdaten des verschlüsselten Datenverkehrs zuzugreifen.
+Dabei soll auch innerhalb des Firmennetzes nicht auf die TLS-Verschlüsselung verzichtet werden.
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+        \item Ausgeben als Server und unverschlüsselte Verbindung anfragen.
+Alice kann den Angriff erkennen, wenn die Verbindung nicht verschlüsselt ist und das Zertifikat der Internetseite nicht gültig ist.
+        \item Der Angriff wird verhindert, wenn der Server der Internetseite ein gültiges Zertifikat besitzt, das von einer vertrauenswürdigen Zertifizierungsstelle ausgestellt wurde.
+        \item Die IT-Abteilung installiert ein Root-Zertifikat auf allen Geräten im Firmennetzwerk.
+Der Datenverkehr wird über einen Proxy-Server geleitet, der mit einem Zertifikat ausgestattet ist, das von dem installierten Root-Zertifikat signiert wurde.
+Der Proxy-Server entschlüsselt den Datenverkehr und verschlüsselt ihn erneut, bevor er an den Empfänger gesendet wird.
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{GSM}
+  \begin{enumerate}
+    \item\label{ex:gsm-attack} Sie befinden sich in der Rolle eines Angreifers und möchten einen Anruf über das GSM-Netz auf Kosten des Mobilfunkkunden A tätigen.
+Obwohl Sie wissen, dass Sie sich oft in der gleichen Funkzelle wie A befinden, sind Sie A nie begegnet.
+Erläutern Sie, welche technischen Maßnahmen und welche Informationen erforderlich sind, um sich dem Mobilfunkprovider gegenüber als Kunde A auszugeben.
+    \item Ist es in der Situation aus Teilaufgabe \ref{ex:gsm-attack} technisch einfacher oder schwieriger, sich nicht dem Mobilfunkprovider gegenüber als anderer Benutzer, sondern gegenüber einem anderen Benutzer als Mobilfunkzelle auszugeben?
+Erläutern Sie das Vorgehen und begründen Sie Ihre Antwort.
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+        \item Erforderliche technische Maßnahmen sind das Abfangen der IMSI von A, das Abfangen des Schlüsselmaterials von A und das Abfangen der Verbindung von A zur Mobilfunkzelle.
+Die Informationen, die erforderlich sind, um sich dem Mobilfunkprovider gegenüber als Kunde A auszugeben, sind die IMSI von A und das Schlüsselmaterial von A.
+        \item Es ist technisch einfacher, sich als Mobilfunkzelle auszugeben, da die IMSI von A nicht benötigt wird.
+Es reicht aus, die Verbindung von A zur Mobilfunkzelle zu unterbrechen und eine eigene Verbindung aufzubauen.
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
 

### 2025-01-21 PbD/exercises.tex
Removed mock exam.
updated some catalogue solutions.

@@
     \begin{itemize}
-      \item **Angriffstypen:** Das System ist sowohl für passive Angreifer (durch Beobachtung des Lieferverkehrs) als auch für aktive Angreifer (Manipulation von Bestellungen) anfällig.
-      \item **Bedrohte Schutzziele:** Vertraulichkeit (Erkenntnisse über geplante Operationen) und Integrität (Manipulation der Bestellungen).
-      \item **Mögliche Gegenmaßnahmen:**
-        \begin{itemize}
-          \item Tarnmaßnahmen wie Bestellungen an Dritte.
-          \item Einsatz von Intermediären zur Verschleierung des Bestellvorgangs.
-          \item Reduzierung der Transparenz durch diversifizierte Bestellziele.
-        \end{itemize}
+      \item Passiver Angreifer: Bestelldatenanalyse können Hinweise auf militärische Operationn geben.
+Vertraulichkeit der operativen Planung wird bedroht.
+Traffic- oder Metadatenanalyse sind mögliche Angriffsarten.
+Als Gegenmaßnahmen können z.B.
+falsche Bestellungen oder Verschleierung der Lieferadresse genannt werden.
+      \item Aktiver Angreifer: Blockade von Lieferungen oder Einschleusung von Schadsoftware (wie QR-Codes) in Lierferungen.
+Integrität der Lieferungen wird bedroht.
+Social Engeineering, Supply Chain Attacks.
+Gegenmaßnahmen könnten z.B.
+die Überprüfung von Lieferungen oder die Verwendung von sicheren Lieferdiensten sein.
     \end{itemize}
@@
     \begin{enumerate}
-      \item **Angriffsmöglichkeiten:** Das System ist sowohl durch passive Angreifer (Abhören von SSIDs) als auch durch aktive Angreifer (Erstellen von Rogue-APs mit gleicher SSID) verwundbar.
-      \item **Bedrohte Schutzziele:** Vertraulichkeit (Tracking von Verbindungen), Integrität (Man-in-the-Middle-Angriffe).
-      \item **Gegenmaßnahmen:**
-        \begin{itemize}
-          \item Deaktivierung der automatischen Verbindung zu bekannten Netzwerken.
-          \item Nutzung von WPA3 oder VPN.
-          \item Vermeidung öffentlicher WLANs.
-        \end{itemize}
+      \item Angriffe sind sowohl durch passive wie aktive Angreifer möglich.
+Passive Angreifer können durch das Abhören von SSIDs Informationen über Nutzer sammeln, während aktive Angreifer gefälschte APs mit bekannten SSIDs aufstellen können, um Nutzer zu täuschen und Daten abzugreifen.
+      \item Passiver Angreifer: Vertraulichkeit; Aktiver Angreifer zusätzlich: Integrität, Verfügbarkeit, Authentizität.
+      \item Mögliche Gegenmaßnahmen sind z.B.
+die Deaktivierung der automatischen WLAN-Verbindung, MAC-Adressen-Änderung um Tracking zu erschweren, die Verwendung von Zertifikaten zur Authentifizierung, oder die Verwendung von VPNs.
     \end{enumerate}
@@
   \begin{enumerate}
-    \item Analysieren Sie die (Un-)Sicherheit eines solchen Protokolls gegenüber einem Angreifer auf der Kommunikationsstrecke.
+    \item Analysieren Sie die Sicherheit eines solchen Protokolls ggü. einem Angreifer auf der Kommunikationsstrecke.
     \item Welche Schutzmaßnahmen können Nutzer und Server umsetzen, um die Sicherheit des Protokolls zu erhöhen?
@@
     \begin{enumerate}
-      \item **Sicherheitsanalyse:** Deterministische Verschlüsselung ermöglicht Replay-Angriffe und das Erraten von PINs durch brute force.
-      \item **Schutzmaßnahmen:** Nutzung eines hybriden Verschlüsselungsverfahrens mit zufälligen Nonces und Transport Layer Security (TLS).
+      \item Der Angreifer kann die PIN zwar nicht direkt entschlüsseln, aber Replay-, Man-in-the-Middle-, und Brute-Force-Angriffe (bei kurzer PIN) sind möglich.
+      \item Server: Challenge-Response-Verfahren, digitale Signaturen, Rate-Limiting um Brute-Force zu verhindern. Nutzer: Lange PINs, regelmäßige Änderung der PIN, TLS verwenden.
     \end{enumerate}
@@
   \begin{solution}
-    Eine mögliche Lösung ist die Nutzung von Threshold-Signaturen, bei denen mehrere Zertifizierungsstellen gemeinsam einen Pseudonym-Schlüssel erzeugen, ohne dass eine einzelne Stelle die Identität zurückverfolgen kann.
-Im Streitfall können die Stellen gemeinsam das Pseudonym aufdecken.
+    Zwei Zertifizierungsstellen $Z_1$ und $Z_2$ zertifizieren jeweils einen Teil des Public Keys.
+Beide Teile zusammen ergeben den vollständigen Public Key.
+Im Streitfall müssen beide Zertifizierungsstellen zusammenarbeiten, um das Pseudonym aufzudecken.
+Denkbar wäre auch, dass eine Stelle die Identität und die andere den Public Key zertifiziert.
   \end{solution}
@@
     \begin{enumerate}
-      \item **Speicherung im GSM:** Ermöglicht effiziente Anrufzustellung. WLAN benötigt dies nicht, da es keine zentrale Infrastruktur nutzt.
-      \item **Datenbanken:** HLR speichert permanent, VLR nur temporär für lokale Effizienz.
-      \item **Schutz:** Insider greifen durch Zugriffsprotokolle erschwert an; Outsider durch Verschlüsselung.
-      \item **TMSI:** Maskiert IMSI, um Aufenthaltsorte besser zu schützen.
+      \item Speicherung in GSM dient der effizienten Zustellung von Anrufen und Nachrichten. Beim WLAN handelt es sich um ein lokales Netzwerk, bei dem keine Standortinformationen benötigt werden.
+      \item Home Location Register (HLR) speichert die Daten dauerhaft, Visitor Location Register (VLR) temporär. Gründe für die Trennung sind Effizienz und Datenschutz.
+      \item Insider: Strikte Zugriffskontrollen und -protokollierung für HLR und VLR, Rollentrennung, sodass kein Mitarbeiter Zugriff auf alle Daten hat, und Verschlüsselung sensibler Informationen können den Aufenthaltsort schützen.
+Outsider: Abfangen von Kommunikation und Abhören von Signalisierungsdaten können den Aufenthaltsort gefährden.
+      \item TMSI schützt den Aufenthaltsort, da sie nicht direkt mit dem IMSI verknüpft ist und regelmäßig geändert wird.
     \end{enumerate}
@@
   \begin{solution}
-    GSM bietet nur beschränkte Sicherheit, da die Verschlüsselung nicht Ende-zu-Ende erfolgt, schwache Verschlüsselungsalgorithmen (z.B.
-A5/1) verwendet werden und Signalisierungsdaten nicht immer geschützt sind.
+    Schwache bzw. veraltete Verschlüsselung, kein Schutz der Signalisierungsdaten, keine Ende-zu-Ende-Verschlüsselung, einseitige Authentifizierung (nur Nutzer zu Netz, nicht umgekehrt), …
   \end{solution}
@@
       \item Offene implizite Adressen sind für jeden sichtbar, verdeckte implizite Adressen verwenden Pseudonyme, um die Identität zu schützen.
-      \item Explizite Adressierung kann verwendet werden, um Nachrichten zu senden, ohne Teilnehmer direkt zu lokalisieren, aber dies erfordert zusätzliche Mechanismen zur Wahrung der Anonymität.
-      \item Ein Satellit kann für variable implizite Adressierung genutzt werden, jedoch sind Bandbreiten- und Latenzanforderungen zu berücksichtigen.
+      \item Grundsätzlich ja, hierbei müsste die Nachricht jedoch an alle Teilnehmer gesendet werden, was ineffizient ist und die Vertraulichkeit gefährdet.
+      \item Variable implizite Adressierung bedeutet, dass die Adresse nicht fest ist, sondern sich ändert.
+Ein Satellit ist grudnsätzlich geeignet, da er große geografische Bereiche gleichzeitig mit einem Broadcast ansprechen könnte.
+Allerdings wäre hohe Bandbreitenanforderungen notwendig, und die Skalierbarkeit des Systems ist beschränkt.
     \end{enumerate}
@@
     \begin{enumerate}
-      \item Der Kollisionsabstand beträgt im Durchschnitt $2^{32} / (100.000 \cdot 50)$ Stunden.
-      \item Um einen mittleren Kollisionsabstand von mindestens 3 Jahren zu erreichen, muss die Adresslänge auf mindestens 48 Bit erhöht werden.
+      \item Es gibt $2^{32}=4.294.967.296$ mögliche Adressen, $100.000 \cdot 50 = 5.000.000$ benötigte Adressen für die Nutzer, und $100.000 \cdot 1 = 100.000$ Nachrichten pro Stunde.
+Der Kollisionsabstand beträgt also
+        \[ d \approx 862h \]
+      \item Um einen mittleren Kollisionsabstand von mindestens 3 Jahren zu erreichen, muss die Adresslänge auf mindestens 41 Bit erhöht werden.
     \end{enumerate}
@@
 
-\setcounter{section}{2023}
-\section{Mock exam}
-
-% \begin{exercise}{}
-%  % TODO
-%
-%   \begin{solution}
-%     % TODO
-%   \end{solution}
-% \end{exercise}
-
-\setcounter{section}{2019}
-\section{1st exam}
+\sheet[2020]{1st exam}
 \begin{exercise}{Broadcast}

### 2025-01-21 BAI/exercises.tex
Updated exam solutions and added some more lecture questions.

@@
 
+\begin{exercise}{Embodied Language Processing}
+  \begin{enumerate}
+    \item What models are suitable for explaining human natural language processing?
+    \item Please explain Hebbian, SOM, and statistical learning.
+    \item What is a grounded language?
+    \item How can bio-inspired grounded language processing improve classic NLP methods?
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+      \item Models suitable for explaining human natural language processing include sensor to actuator mapping, semantic association, and conceptual representations for stimulus prediction.
+      \item Hebbian learning is a learning rule that strengthens synaptic connections between neurons that fire together.
+SOM learning is a learning algorithm used in self-organizing maps to cluster similar input patterns.
+Statistical learning is a method of learning from data by estimating the probability distribution of the data.
+      \item A grounded language is a language that is based on the physical world and is grounded in sensory-motor experiences.
+      \item Bio-inspired grounded language processing can improve classic NLP methods by providing brain-inspired representations and mechanisms that better capture the structure and meaning of language.
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Robot Sound Localization}
+  \begin{enumerate}
+    \item Explain the concepts of ITD and ILD in a hybrid acoustic system.
+    \item How can ITD and ILD be implemented in a hybrid spiking neural network?
+    \item What are the advantages of this approach?
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+      \item Interaural time difference (ITD) is the difference in time it takes for a sound to reach each ear, while interaural level difference (ILD) is the difference in intensity of a sound at each ear.
+In a hybrid acoustic system, ITD and ILD are used to localize sound sources in space.
+      \item ITD and ILD can be implemented in a hybrid spiking neural network by using spiking neurons to model the auditory pathways and integrating ITD and ILD information to localize sound sources.
+      \item The advantages of this approach include the ability to process complex auditory information in real-time, the ability to integrate multiple sensory modalities, and the ability to adapt to changing environments.
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+% Visual information processed in different brain areas
+%  Images projected from retina to striate and extrastriate cortex,
+% filtered according to different features (edges, orientation, colour,
+% motion, …)
+%  Visual Processing inspired
+% hierarchical computational models for:
+% • Object recognition (Fukushima, LeCun)
+% • Deep Learning models
+\begin{exercise}{}
+  % TODO
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+% Models for bio-inspired sound source localization, visual
+% localization and multisensory integration
+%  Multisensory VR environment for neuro-robotic experiments
+%  Effectiveness of bio-inspired models
+%  Demonstrated behavioral, neurophysiological
+% similarity between models and biological systems
+\begin{exercise}{}
+  % TODO
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+% Advantages:
+%  Fast Reaction – Immediate mapping of sensory information onto motor actions
+%  Robustness – If one part fails, robot may retain some behavioral competence
+%  Multiple Goals – Can follow multiple goals simultaneously by combining
+% behaviors
+%  Extensibility – Easy to add new parts on top of previous
+%  Simplicity – Complexity derives from continuous
+% interaction of simple modules with environment
+% and each other
+%  Computational tractability – Usually simple calculations
+% and easy to parallelize
+%  Development of using LLM for planning
+% Disadvantages/Limits:
+%  Limited Information – Agents without environment models must have
+% sufficient information from local environment
+%  Non-Local information – how does agent take into account non-local
+% information (e.g. memory?)
+%  Learning globally – Difficult to make reactive agents
+% that learn global behavior
+%  Complex dynamics – It is hard to engineer agents
+% with large numbers of behaviors (dynamics of
+% interactions become too complex to understand)
+\begin{exercise}{}
+  % TODO
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+% Gestures are an integral part in communication
+% • Important in multimodal systems
+% • Research on autonomous robots and HRI
+%  Model alternatives and choice
+% • Dependent on gesture type → gesture continuum
+% • Deep networks provide partial solution for feature extraction
+% • Recurrent networks in LSTM and GammaGWR
+% • Reservoir Computing models by neurophysiological principles
+% for action selection in prefrontal cortex and striatum
+% • Tradeoff between benchmarking and developing cognitive systems
+\begin{exercise}{}
+  % TODO
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+% Pros:
+% • General heuristic for problem search
+% • Able to find good solutions in feasible computing time
+% • Distributed execution possible
+%  Cons:
+% • Cannot guarantee optimal solutions
+% • Long runtimes, compared to search algorithms
+% • Success depends also on used parameters
+%  Many further EA approaches: NEAT, HyperNEAT…
+\begin{exercise}{}
+  % TODO
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
+%  LMM deliver pointwise understanding for frequent cases
+%  For rare cases LMM often “autocorrect reality” answering often wrong question
+%  Hybrid systems make complementary use of symbolic and connectionist AI
+%  Symbolic rules can be extracted from neural networks
+%  Different architectures for integrating symbolic and neural computing are
+% possible
+%  How images are processed by a neural network can be visualized
+%  Hybrid systems and explainable AI are in the
+% current focus of research
+%  Topics for thesis, independent studies, etc.
+\begin{exercise}{}
+  % TODO
+
+  \begin{solution}
+    % TODO
+  \end{solution}
+\end{exercise}
+
 \sheet[2023]{1st exam}
@@
   \begin{enumerate}
-    \item How is the neighborhood in a self-organizing map (SOM) is computed.
+    \item The neighborhood in a self-organizing map (SOM) is computed as a 
+      \begin{itemize}
+        \item Gaussian
+        \item linear
+        \item step
+        \item constant
+      \end{itemize}
+      function of the distance between the winning neuron and the other neurons.
 
+    \item A ground language is based on the 
       \begin{itemize}
-        \item The neighborhood is computed as a Gaussian function of the distance between the winning neuron and the other neurons.
-        \item The neighborhood is computed as a linear function of the distance between the winning neuron and the other neurons.
-        \item The neighborhood is computed as a step function of the distance between the winning neuron and the other neurons.
-        \item The neighborhood is computed as a constant function of the distance between the winning neuron and the other neurons.
+        \item ground truth.
+        \item physical world.
+        \item ground state.
+        \item ground floor.
       \end{itemize}
 
-    \item What does it mean for a language to be grounded?
-
+    \item The ventriloquism effect is the illusion that a sound is coming from 
       \begin{itemize}
-        \item The language is based on the ground truth.
-        \item The language is based on the physical world.
-        \item The language is based on the ground state.
-        \item The language is based on the ground floor.
+        \item a different location
+        \item the same Location
       \end{itemize}
-
-    \item What is the ventriloquism effect?
-
-      \begin{itemize}
-        \item The ventriloquism effect is the illusion that a sound is coming from a different location than the actual source.
-        \item The ventriloquism effect is the illusion that a sound is coming from the same location as the actual source.
-        \item The ventriloquism effect is the illusion that a sound is coming from a different location than the actual source.
-        \item The ventriloquism effect is the illusion that a sound is coming from the same location as the actual source.
-      \end{itemize}
+      than the actual source.
   \end{enumerate}
@@
     \begin{enumerate}
-      \item The neighborhood is computed as a Gaussian function of the distance between the winning neuron and the other neurons.
-      \item The language is based on the physical world.
-      \item The ventriloquism effect is the illusion that a sound is coming from a different location than the actual source.
+      \item Gaussian.
+      \item Physical world.
+      \item A different location.
     \end{enumerate}
@@
     \begin{enumerate}
-      \item The leaky integrate-and-fire model is an extension of the integrate-and-fire model that includes a leaky term.
-This term allows the membrane potential of the neuron to decay over time, which can help to model more realistic neuron behavior.
+      \item In the integrate-and-fire model, the neuron integrates input signals over time and fires a spike when the membrane potential reaches a threshold.
+In the leaky integrate-and-fire model, the neuron also integrates input signals over time, but the membrane potential decays over time in the absence of input.
+This leads to more realistic behavior and allows the neuron to adapt to changing input conditions.
       \item The similarity angle between two vectors $\mathbf{a}$ and $\mathbf{b}$ is given by
-      \[
-        \cos(\theta) = \frac{\mathbf{a} \cdot \mathbf{b}}{\|\mathbf{a}\| \|\mathbf{b}\|}.
-      \]
+        \[ \cos(\theta) = \frac{\mathbf{a} \cdot \mathbf{b}}{\|\mathbf{a}\| \|\mathbf{b}\|}. \]
         For $\mathbf{w} = [1, 2, 3]$ and $\mathbf{x} = [4, 5, 6]$, we have
-      \[
-        \cos(\theta) = \frac{1 \cdot 4 + 2 \cdot 5 + 3 \cdot 6}{\sqrt{1^2 + 2^2 + 3^2} \sqrt{4^2 + 5^2 + 6^2}} = \frac{32}{\sqrt{14} \sqrt{77}}.
-      \]
+        \[ \cos(\theta) = \frac{1 \cdot 4 + 2 \cdot 5 + 3 \cdot 6}{\sqrt{1^2 + 2^2 + 3^2} \sqrt{4^2 + 5^2 + 6^2}} = \frac{32}{\sqrt{14} \sqrt{77}}. \]
       \item The output of a perceptron is given by
@@
           y = \begin{cases}
-          1 & \text{if } \mathbf{w} \cdot \mathbf{x} + b > 0, \\
+            1 & \text{if } \mathbf{w} \cdot \mathbf{x} > 0, \\
             0 & \text{otherwise}.
@@
       \item The orientation and firing rate of a simple cell for a line of different orientations can be visualized in a diagram with the orientation on the $x$-axis and the firing rate on the $y$-axis.
+It usually shows a peak at the preferred orientation of the cell and decreases for other orientations.
     \end{enumerate}
@@
 \begin{exercise}{Audio Processing}
-  \begin{enumerate}
-    \item Explain interaural time difference (ITD) and interaural level difference (ILD) and which one is used for high and low frequencies.
-  \end{enumerate}
+  Explain interaural time difference (ITD) and interaural level difference (ILD) and which one is used for high and low frequencies.
 
@@
     \item Explain a Braitenberg vehicle with an example and name the behavior of your example vehicle.
-    \item Complete the flow chart of the functional decomposition of the classical approach for a robot controller: sensors
+    \item Complete the flow chart of the functional decomposition of the classical approach for a robot controller: sensors $\rightarrow$ \_ $\rightarrow$ \_ $\rightarrow$ \_ $\rightarrow$ \_ $\rightarrow$ actuators.
     \item Explain the advantage of a behavior-based approach to robotics.
@@
 The behavior of this example vehicle is phototaxis.
-      \item The flow chart of the functional decomposition of the classical approach for a robot controller is: sensors $\rightarrow$ perception $\rightarrow$ cognition $\rightarrow$ planning $\rightarrow$ control $\rightarrow$ actuators.
+      \item The flow chart of the functional decomposition of the classical approach for a robot controller is: sensors $\rightarrow$ perception $\rightarrow$ modelling $\rightarrow$ planning $\rightarrow$ execution $\rightarrow$ actuators.
       \item The advantage of a behavior-based approach to robotics is that it allows for the development of complex behaviors through the combination of simple behaviors.
@@
 \begin{exercise}{Evolutionary Computing}
+  \input{res/tsp.fig}
+
   \begin{enumerate}
-    \item Draw a cycle graph with arrow labels.
-    \item Given an example of finding the shortest distance between cities by only visiting each city once.
+    \item Draw a cycle graph with the nearest-neighbor heuristic for the traveling salesman problem.
+    \item Find the shortest distance between city $1$ and city $4$ in the graph.
     \item Describe a recombination and mutation algorithm for the example you provided.
@@
       \item A cycle graph with arrow labels is a graph where each node is connected to the next node in a cycle, and the arrows indicate the direction of the cycle.
-      \item An example of finding the shortest distance between cities by only visiting each city once is the traveling salesman problem.
-      \item A recombination and mutation algorithm for the traveling salesman problem could involve recombining the order of cities in two different solutions and mutating the order of cities in a single solution.
+      \item The shortest distance between city $1$ and city $4$ in the graph is $1\to3\to4$.
+      \item An approach to solving the traveling salesman problem using recombination and mutation could involve combining the paths of two solutions to create a new solution and randomly swapping cities in the solution to introduce diversity.
       \item Optimization strategies for this algorithm could include using a genetic algorithm to evolve a population of solutions, using local search to improve the quality of solutions, and using elitism to preserve the best solutions.

### 2025-01-21 BAI/paper/presentation.tex
improved presentation a little bit.

@@
 
-% Slide 1: Introduction
 \begin{frame}{Introduction}
@@
 			\item Combines computer vision (CV) and natural language processing (NLP).
-			\item Applications: Assistive technology, automated metadata generation, and more.
+			\item Application examples: Assistive technology, automated metadata generation.
 	\end{itemize}
-	% Speaker Notes:
-	% Begin by briefly explaining the task of image captioning. Highlight its interdisciplinary nature and practical applications.
 \end{frame}
 
-% Slide 2: Problem Statement
-\begin{frame}{Problem Statement}
-	\textbf{Challenges in Image Captioning}
-	\begin{itemize}
-			\item Detecting and interpreting visual elements accurately.
-			\item Generating coherent and contextually relevant text.
-			\item Balancing computational efficiency and model performance.
-	\end{itemize}
-	% Speaker Notes:
-	% Emphasize the dual challenges of visual understanding and text generation. Mention the need for scalable solutions in real-world applications.
+\begin{frame}{Introduction}
+	\begin{figure}[H]
+		\includegraphics[width=.6\textwidth]{res/dog.jpg}
+		\caption{A dog running on a field.}\label{fig:dog}
+	\end{figure}
 \end{frame}
 
-% Slide 3: Methodology (Dataset and Features)
+\begin{frame}{Research Question}
+	How do different decoder architectures (GRU, LSTM, Transformer) perform in the context of image captioning on the Flickr8k dataset?
+\end{frame}
+
 \begin{frame}{Methodology: Dataset and Features}
@@
 	\end{itemize}
-	% Speaker Notes:
-	% Explain why Flickr8k was chosen and describe its characteristics. Introduce EfficientNet B0 as a state-of-the-art feature extractor.
 \end{frame}
 
-% Slide 4: Methodology (Decoder Architectures)
 \begin{frame}{Methodology: Decoder Architectures}
@@
 	\end{itemize}
-	% Speaker Notes:
-	% Briefly describe each architecture and its role in the pipeline. Mention the choice of PyTorch for implementation.
 \end{frame}
 
-% Slide 5: Results (Metrics)
+\begin{frame}{Methodology: Decoder Architectures}
+	\begin{figure}[H]
+		\includegraphics[width=.9\textwidth]{res/gru.png}
+		\caption{GRU Decoder.}\label{fig:gru}
+	\end{figure}
+\end{frame}
+
+\begin{frame}{Methodology: Decoder Architectures}
+	\begin{figure}[H]
+		\includegraphics[width=.9\textwidth]{res/lstm.png}
+		\caption{LSTM Decoder.}\label{fig:lstm}
+	\end{figure}
+\end{frame}
+
+\begin{frame}{Methodology: Decoder Architectures}
+	\begin{figure}[H]
+		\includegraphics[width=.9\textwidth]{res/transformer.png}
+		\caption{Transformer Decoder.}\label{fig:transformer}
+	\end{figure}
+\end{frame}
+
 \begin{frame}{Results: Metrics}
@@
 	\end{itemize}
-	\textbf{Key Findings}
-	\begin{itemize}
-			\item Transformer underperforms GRU and LSTM in all metrics.
-	\end{itemize}
-	% Speaker Notes:
-	% Introduce the metrics and explain their relevance. Highlight the surprising result of Transformer underperformance.
+
+	% TODO: explain these metrics
 \end{frame}
 
-% Slide 6: Results (Comparison)
 \begin{frame}{Results: Comparison of Architectures}
@@
 	\includegraphics[width=.6\textwidth]{res/metrics.png}
-	% Speaker Notes:
-	% Present the comparison visually and discuss why the Transformer might struggle with limited data.
 \end{frame}
 
-% Slide 7: Discussion
 \begin{frame}{Discussion}
@@
 	\end{itemize}
-	% Speaker Notes:
-	% Reflect on what these results mean for the field and the practical implications of using different architectures.
 \end{frame}
 
-% Slide 8: Conclusion and Future Work
 \begin{frame}{Conclusion and Future Work}
@@
 			\item Test on larger datasets (e.g., MS COCO).
-			\item Explore hybrid architectures.
+			\item Finetune pretrained models.
 	\end{itemize}
-	% Speaker Notes:
-	% Summarize the study's key points and suggest future directions for improvement and exploration.
 \end{frame}

### 2025-01-22 PbD/exercises.tex
Updated a few solutions.

@@
     \begin{enumerate}
-      \item Der Blind-Message-Service erfüllt die Schutzziele Vertraulichkeit und Anonymität.
-      \item Verschlüsselte Übertragungen verhindern, dass ein Angreifer Inhalte oder Muster erkennt.
-      \item Die Verschlüsselung der Anfragevektoren genügt, da nur der Inhalt der Anfragen sensibel ist.
-      \item Alternativen sind Onion-Routing und Broadcast-Kommunikation, die jedoch mehr Bandbreite benötigen.
+      \item Vertraulichkeit.
+      \item Verschlüsselte Übertragungen verhindern, dass der Server die Anfragevektoren und die Antworten lesen kann und somit bleibt unbekannt was der Nutzer anfragt.
+      \item Die Antworten sind so gestaltet, dass sie keine direkten Informationen über die angefragten Daten enthalten. Dadurch muss lediglich die Anfrage verschlüsselt werden.
+      \item Alternativ könnte man Broadcast-Kommunikation einsetzen, wobei alle Daten abgefragt werden. Oder man könnte über mehrere Server verteilt Anfragen stellen, und sie dann aggregieren.
     \end{enumerate}
@@
     \begin{enumerate}
-      \item Die Datenbank verhindert Replay-Angriffe, bei denen Nachrichten mehrfach gesendet werden.
-      \item Die Datenbank kann gelöscht werden, sobald alle Nachrichten erfolgreich weitergeleitet wurden.
-      \item Eine mögliche Modifikation ist die Nutzung von Hashes zur Überprüfung statt vollständiger Speicherung, wodurch der Speicherbedarf reduziert wird.
+      \item Replay-Angriffe werden verhindert.
+Hierbei werden alte Nachrichten erneut gesendet, und da der einfache Mix zufällig verschlüsselt, kann der Angreifer feststellen welcher Output unverändert bleibt.
+      \item Bis das Schlüsselpaar des Mixes geändert wird, kann die Datenbank nicht gelöscht werden.
+      \item Hash-basierte Speicherung. Bloom-Filter. Batch-abhängige Verarbeitung. Zeitlich begrenzte Speicherung.
     \end{enumerate}
@@
 \begin{exercise}{DC-Netz}
-  \begin{enumerate}
-    \item Der (entartete) Schlüsselgraph eines DC-Netzes für die Teilnehmer A, B, C und D sei (A-B-C-D).
+  Der (entartete) Schlüsselgraph eines DC-Netzes für die Teilnehmer A, B, C und D sei (A-B-C-D).
 B greift nun A, C und D an.
@@
 Wie sind Sie bei Ihrer Analyse vorgegangen?
-  \end{enumerate}
 
   \begin{solution}
-    Wenn Teilnehmer B A, C und D angreift, bleibt die Anonymität für A, C und D teilweise erhalten. Die Analyse basiert auf der Struktur des Schlüsselgraphs und der Isolation kompromittierter Knoten.
+    Teilnehmer A und C sind direkt gefährdet, da B Zugriff auf ihre Schlüsselanteile hat und diese aus Gesamtsumme isolieren kann.
+D ist relativ sicherer, solange C nicht kompromittiert ist.
+Die Analyse erfolgte durch Berechnung der Nachrichten, die B senden kann.
   \end{solution}

### 2025-01-23 PbD/exercises.tex
Updated all pbd catalogue exercises.

@@
   \begin{enumerate}
-    \item Erläutern Sie eine mögliche Vorgehensweise, mit der ein starker Angreifer (z.B.
+    \item\label{ex:break-mix} Erläutern Sie eine mögliche Vorgehensweise, mit der ein starker Angreifer (z.B.
 ein Nachrichtendienst) die IP-Adressen der Nutzer herausfinden kann, die eine bestimmte Webseite über ein mixbasiertes Anonymisierungssystem (z.B.
 Tor) abrufen.
-    \item Erläutern Sie eine mögliche Vorgehensweise, mit der ein starker Angreifer (z.B.
+    \item\label{ex:break-mix-2} Erläutern Sie eine mögliche Vorgehensweise, mit der ein starker Angreifer (z.B.
 ein Nachrichtendienst) ermitteln kann, welche Webseiten ein bestimmter Nutzer über ein mixbasiertes Anonymisierungssystem abgerufen hat.
-Hinweis: Beschreiben Sie eine andere Vorgehensweise als in Teilaufgabe a), auch wenn Ihre für Teilaufgabe a) gewählte Vorgehensweise für Teilaufgabe b) geeignet ist.
+      \par\textit{Hinweis: Beschreiben Sie eine andere Vorgehensweise als in Teilaufgabe \ref{ex:break-mix}, auch wenn Ihre für Teilaufgabe \ref{ex:break-mix} gewählte Vorgehensweise für Teilaufgabe \ref{ex:break-mix-2} geeignet ist.}
     \item Erläutern Sie drei Techniken, mit denen ein Webserver einen Nutzer bei seinem zweiten (späteren) Besuch wiedererkennen kann.
@@
     \begin{enumerate}
-      \item Ein starker Angreifer kann Timing-Analyse und Traffic-Korrelation verwenden, um IP-Adressen von Nutzern zu identifizieren.
-      \item Ein anderer Ansatz ist die Analyse von Metadaten, wie HTTP-Headern oder spezifischen Anfragecharakteristiken.
-      \item Webserver können Nutzer durch Cookies, Browser-Fingerprinting und IP-Adressen wiedererkennen.
+      \item Timing-Korrelation: Verlauf und Volumen von Anfragen bei der Eingangs-Node und Antworten bei der Exit-Node korrelieren, um Nutzer zu identifizieren.
+      \item Das Endgerät des Nutzers mit Malware infizieren, um Browserverlauf und Cookies zu stehlen.
+      \item Cookies, Browser-Fingerprinting.
     \end{enumerate}
@@
     \begin{enumerate}
-      \item Beim Tagging-Angriff fügt der Angreifer eine markante Änderung in den Datenverkehr ein, um Nutzer zu identifizieren.
-      \item Der Angriff wird durch strikte Prüfung der Datenintegrität und Verschlüsselung verhindert.
+      \item Datenpakete mit Tag versehen (zB. Größe oder Reihenfolge der Pakete), welcher auch bei Verschlüsselung bestehen bleibt. Hiermit kann das Ziel des Pakets identifiziert werden.
+      \item E2E-Verschlüsselung, Prüfsummen, Paketnormalisierung.
     \end{enumerate}
@@
     \begin{enumerate}
-      \item Bedrohte Schutzziele sind Vertraulichkeit und Anonymität.
-      \item Tracking auf Anwendungsebene nutzt Browserdaten, während Netzebenen-Tracking auf IP-Daten basiert.
-      \item Gegenmaßnahmen:
-        \begin{itemize}
-          \item Blockieren von Cookies (z.B. durch Browser-Einstellungen).
-          \item Nutzung von VPNs zur Verschleierung der IP.
-          \item Einsatz von Privacy-Plugins (z.B. uBlock Origin).
+      \item Vertraulichkeit, Anonymität.
+      \item Anwendungsebene verwendet Browserdaten, während die Netzebene IP-Daten verwendet. Tracking innerhalb eines Programms v. Tracking über Netzwerke hinweg.
+      \item \begin{itemize}
+          \item Blockierung von Tracking (zB. Adblock): Tracking-Cookies und Third-Party-Mechanismen werden blockiert. Schutz vor Werbetreibenden.
+          \item Anonymisierungsdienste (zB. Incognito-Modus): Tracking-Cookies und lokale Daten werden nach Sitzung gelöscht. Schutz vor Webseiten und Werbetreibenden.
+          \item Fingerprinting-Schutz (zB.
+Privacy Badger): Standardisierung von Informationen wie Schriftart, Zeitzone, Bildschirmauflösung durch Manipulation der Browser-Umgebung.
+Schutz vor Webseiten.
         \end{itemize}
@@
   \begin{solution}
-    Das Vorgehen besteht in der Analyse der Zeitstempel von Access-Logs und Router-Daten, um Verbindungen herzustellen. Diese Art von Angriff wird als Timing-Korrelation bezeichnet.
+    Hierbei kann ein Timing-Angriff durchgeführt werden.
+Verfügbare Daten sind das Access-Log, DSL-Router-Protokoll und die Latenz zwischen Router und Server.
+Durch Vergleich der Zugriffszeiten im Access-Log und den Latenzen im Router-Protokoll kann die interne IP-Adresse ermittelt werden.
+    \begin{itemize}
+      \item Alle relevanten HTTP-Anfragen aus dem Access-Log identifizieren und die Zeitpunkte notieren.
+      \item Alle ausgehenden Pakete im Router-Protokoll identifizieren und die Zeitpunkte, sowie die zugehörigen IP-Adressen notieren.
+      \item Zeitstempel korrelieren unter Beachtung der Latenz (Anfragen aus Access-Log +-.5 Sekunden später als im Router-Protokoll).
+      \item IP-Adresse mit den korrelierten Zeitstempeln identifizieren.
+    \end{itemize}
   \end{solution}

### 2025-01-23 BAI/exercises.tex
Added lecture questions.

@@
 
-% Visual information processed in different brain areas
-%  Images projected from retina to striate and extrastriate cortex,
-% filtered according to different features (edges, orientation, colour,
-% motion, …)
-%  Visual Processing inspired
-% hierarchical computational models for:
-% • Object recognition (Fukushima, LeCun)
-% • Deep Learning models
-\begin{exercise}{}
-  % TODO
+\begin{exercise}{Hierarchical Vision}
+  \begin{enumerate}
+    \item Which visual information is processed in which brain areas?
+    \item What are the features filtered in the visual processing of the brain?
+    \item Explain a hierarchical computational model for object recognition.
+    \item How can deep learning be utilized for visual processing?
+  \end{enumerate}
 
   \begin{solution}
-    % TODO
+    \begin{enumerate}
+      \item Low-level visual information is processed in the primary visual cortex, while high-level visual information is processed in the ventral and dorsal streams of the brain.
+      \item The features filtered in the visual processing of the brain include edges, textures, shapes, and objects.
+      \item A hierarchical computational model for object recognition involves multiple layers of processing, where each layer extracts increasingly complex features from the input data to recognize objects.
+      \item Deep learning can be utilized for visual processing by using deep neural networks to learn hierarchical representations of visual data and perform tasks like image recognition and object detection.
+    \end{enumerate}
   \end{solution}
@@
 
-% Models for bio-inspired sound source localization, visual
-% localization and multisensory integration
-%  Multisensory VR environment for neuro-robotic experiments
-%  Effectiveness of bio-inspired models
-%  Demonstrated behavioral, neurophysiological
-% similarity between models and biological systems
-\begin{exercise}{}
-  % TODO
+\begin{exercise}{Crossmodal Processing}
+  \begin{enumerate}
+    \item Briefly explain one model for bio-inspired sound source localization, visual localization, and multisensory integration each.
+    \item How effective are bio-inspired models in practice?
+    \item Demonstrate similarities and differences between bio-inspired models and biological systems.
+  \end{enumerate}
 
   \begin{solution}
-    % TODO
+    \begin{enumerate}
+      \item One model for bio-inspired sound source localization is the duplex theory of sound localization, which uses interaural time difference (ITD) and interaural level difference (ILD) to localize sound sources.
+One model for bio-inspired visual localization is the hierarchical processing model, which uses multiple layers of processing to extract features from visual data.
+One model for bio-inspired multisensory integration is the Bayesian integration model, which combines information from multiple sensory modalities to make decisions.
+      \item Bio-inspired models are effective in practice for tasks like sound source localization, visual localization, and multisensory integration.
+These models can capture the complexity and flexibility of biological systems and provide insights into how sensory information is processed in the brain.
+      \item Bio-inspired models demonstrate behavioral and neurophysiological similarities with biological systems, such as the ability to localize sound sources, recognize objects, and integrate information from multiple sensory modalities.
+However, there are also differences between bio-inspired models and biological systems, such as the level of abstraction and complexity of the models.
+    \end{enumerate}
   \end{solution}
@@
 
-% Advantages:
-%  Fast Reaction – Immediate mapping of sensory information onto motor actions
-%  Robustness – If one part fails, robot may retain some behavioral competence
-%  Multiple Goals – Can follow multiple goals simultaneously by combining
-% behaviors
-%  Extensibility – Easy to add new parts on top of previous
-%  Simplicity – Complexity derives from continuous
-% interaction of simple modules with environment
-% and each other
-%  Computational tractability – Usually simple calculations
-% and easy to parallelize
-%  Development of using LLM for planning
-% Disadvantages/Limits:
-%  Limited Information – Agents without environment models must have
-% sufficient information from local environment
-%  Non-Local information – how does agent take into account non-local
-% information (e.g. memory?)
-%  Learning globally – Difficult to make reactive agents
-% that learn global behavior
-%  Complex dynamics – It is hard to engineer agents
-% with large numbers of behaviors (dynamics of
-% interactions become too complex to understand)
-\begin{exercise}{}
-  % TODO
+\begin{exercise}{Behavior-based Robotics}
+  \begin{enumerate}
+    \item Explain the advantages and disadvantages of behavior-based robotics.
+    \item How can behavior-based robotics be used for planning?
+    \item What are the limits of behavior-based robotics?
+  \end{enumerate}
 
   \begin{solution}
-    % TODO
+    \begin{enumerate}
+      \item The advantages of behavior-based robotics include fast reaction times, robustness, the ability to follow multiple goals simultaneously, extensibility, simplicity, and computational tractability.
+The disadvantages of behavior-based robotics include limited information, non-local information, learning globally, and complex dynamics.
+      \item Behavior-based robotics can be used for planning by combining simple behaviors to achieve complex goals.
+This approach allows robots to react to their environment in real-time and adapt to changing conditions.
+      \item The limits of behavior-based robotics include the need for sufficient information from the local environment, the challenge of incorporating non-local information, the difficulty of learning global behavior, and the complexity of interactions in large-scale systems.
+    \end{enumerate}
   \end{solution}
@@
 
-% Gestures are an integral part in communication
-% • Important in multimodal systems
-% • Research on autonomous robots and HRI
-%  Model alternatives and choice
-% • Dependent on gesture type → gesture continuum
-% • Deep networks provide partial solution for feature extraction
-% • Recurrent networks in LSTM and GammaGWR
-% • Reservoir Computing models by neurophysiological principles
-% for action selection in prefrontal cortex and striatum
-% • Tradeoff between benchmarking and developing cognitive systems
-\begin{exercise}{}
-  % TODO
+\begin{exercise}{Gesture recognition}
+  \begin{enumerate}
+    \item Explain the importance of gestures in communication and their role in multimodal systems.
+    \item Which models are suitable for gesture recognition and how do they differ?
+    \item How would you choose a model for gesture recognition?
+  \end{enumerate}
 
   \begin{solution}
-    % TODO
+    \begin{enumerate}
+      \item Gestures are an integral part of communication and play an important role in multimodal systems.
+They provide additional information beyond speech and can convey emotions, intentions, and context.
+      \item Models suitable for gesture recognition include deep networks for feature extraction, recurrent networks like LSTM and GammaGWR for sequence modeling, and reservoir computing models for action selection.
+These models differ in their approach to feature extraction, sequence modeling, and action selection.
+      \item When choosing a model for gesture recognition, it is important to consider the type of gestures being recognized, the complexity of the task, the amount of data available, and the computational resources required.
+Deep networks are suitable for feature extraction, recurrent networks are suitable for sequence modeling, and reservoir computing models are suitable for action selection.
+    \end{enumerate}
   \end{solution}
@@
 
-% Pros:
-% • General heuristic for problem search
-% • Able to find good solutions in feasible computing time
-% • Distributed execution possible
-%  Cons:
-% • Cannot guarantee optimal solutions
-% • Long runtimes, compared to search algorithms
-% • Success depends also on used parameters
-%  Many further EA approaches: NEAT, HyperNEAT…
-\begin{exercise}{}
-  % TODO
+\begin{exercise}{Evolutionary Computing}
+  \begin{enumerate}
+    \item Explain the advantages and disadvantages of evolutionary algorithms.
+    \item What are the limits of evolutionary algorithms?
+    \item Name and describe two evolutionary algorithms.
+  \end{enumerate}
 
   \begin{solution}
-    % TODO
+    \begin{enumerate}
+      \item The advantages of evolutionary algorithms include being a general heuristic for problem search, the ability to find good solutions in feasible computing time, and the possibility of distributed execution.
+The disadvantages of evolutionary algorithms include not guaranteeing optimal solutions, long runtimes compared to search algorithms, and the success depending on the parameters used.
+      \item The limits of evolutionary algorithms include the need for a large number of evaluations, the difficulty of tuning parameters, the risk of premature convergence, and the challenge of scaling to large problems.
+      \item Two evolutionary algorithms are the genetic algorithm, which uses selection, crossover, and mutation operators to evolve a population of solutions, and the genetic programming algorithm, which evolves programs or models to solve a problem.
+    \end{enumerate}
   \end{solution}
@@
 
-%  LMM deliver pointwise understanding for frequent cases
-%  For rare cases LMM often “autocorrect reality” answering often wrong question
-%  Hybrid systems make complementary use of symbolic and connectionist AI
-%  Symbolic rules can be extracted from neural networks
-%  Different architectures for integrating symbolic and neural computing are
-% possible
-%  How images are processed by a neural network can be visualized
-%  Hybrid systems and explainable AI are in the
-% current focus of research
-%  Topics for thesis, independent studies, etc.
-\begin{exercise}{}
-  % TODO
+\begin{exercise}{Neuro-Symbolic and Explainable AI Systems}
+  \begin{enumerate}
+    \item Explain the concept of neuro-symbolic systems and how they combine symbolic and connectionist AI.
+    \item How can symbolic rules be extracted from neural networks?
+    \item Describe different architectures for integrating symbolic and neural computing.
+    \item How can images be processed by a neural network and visualized?
+    \item Why are hybrid systems and explainable AI in the current focus of research?
+  \end{enumerate}
 
   \begin{solution}
-    % TODO
+    \begin{enumerate}
+      \item Neuro-symbolic systems combine symbolic and connectionist AI by integrating symbolic reasoning with neural network processing.
+These systems use symbolic rules to guide neural network learning and decision-making.
+      \item Symbolic rules can be extracted from neural networks by analyzing the learned weights and activations of the network to identify patterns and rules that explain the network's behavior.
+      \item Different architectures for integrating symbolic and neural computing include neural-symbolic integration, where symbolic reasoning is combined with neural network processing, and neural-symbolic learning, where neural networks are used to learn symbolic rules from data.
+      \item Images can be processed by a neural network using convolutional layers to extract features from the input data and fully connected layers to classify the features.
+The processing steps can be visualized by examining the activations and weights of the network at each layer.
+      \item Hybrid systems and explainable AI are in the current focus of research because they provide a way to combine the strengths of symbolic and connectionist AI, improve the interpretability of AI systems, and enable more transparent and accountable decision-making.
+    \end{enumerate}
   \end{solution}
@@
   \begin{enumerate}
-    \item In which models is backpropagation used for learning?
+    \item In 
       \begin{itemize}
@@
         \item Self-organizing map (SOM)
-        \item None of the above
+        \item none of the above
       \end{itemize}
+        is backpropagation used for learning.
 
-    \item What does "embodied" mean for language?
+    \item An embodied language is based on
       \begin{itemize}
-        \item Language is based on the physical world.
-        \item Language is based on the ground truth.
-        \item Language is based on the ground state.
-        \item Language is based on the ground floor.
+        \item the physical world.
+        \item ground truth.
+        \item ground state.
+        \item ground floor.
       \end{itemize}
@@
         \item Ventral stream for spatial vision and dorsal stream for object recognition
-        \item Dorsal stream for object recognition and ventral stream for spatial vision
-        \item Ventral stream for object recognition and dorsal stream for spatial vision
       \end{itemize}

### 2025-01-24 BAI/exercises.tex
Removed code link and added last chapter lecture questions.

@@
       \item Hybrid systems and explainable AI are in the current focus of research because they provide a way to combine the strengths of symbolic and connectionist AI, improve the interpretability of AI systems, and enable more transparent and accountable decision-making.
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Continual Learning for Robotic Action}
+  \begin{enumerate}
+    \item Explain the stability-plasticity dilemma and how it relates to continual learning.
+    \item What is Hebbian learning?
+    \item Outline the problem of catastrophic forgetting in neural networks and how it can be overcome.
+    \item How could continual object recognition be approached?
+    \item What are the applications of continual learning in natural language processing?
+    \item Compare the state-of-the-art models with biological systems.
+    \item How can continual learning be used to model higher-level cognitive functions in artificial systems?
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+      \item The stability-plasticity dilemma is the trade-off between the ability of a neural network to learn new information (plasticity) and the ability of the network to retain previously learned information (stability).
+Continual learning is the ability to incrementally learn over time without forgetting.
+      \item Hebbian learning is a learning rule that strengthens synaptic connections between neurons that fire together.
+      \item Catastrophic forgetting is the phenomenon where a neural network forgets previously learned information when learning new information.
+This problem can be overcome by using regularization, dynamic architectures, and rehearsal/replay.
+      \item Continual object recognition can be approached by using self-organizing networks like Gamma-GWR to learn sequences of objects and by using hybrid models like Growing Dual-Memory (GDM) with memory replay to recognize objects over time.
+      \item Applications of continual learning in natural language processing include continual language learning and open challenges in adapting to new tasks and data.
+      \item State-of-the-art models are far from providing the flexibility, robustness, and scalability of biological systems.
+These models are limited in their ability to learn from new data and adapt to changing environments.
+      \item Continual learning can be used to model higher-level cognitive functions in artificial systems by providing a framework for learning from new data, adapting to new tasks, and retaining previously learned information.
     \end{enumerate}

### 2025-01-28 NLP/exercises.tex
Added quizzes content.
Added quizzes.

@@
 
-\setcounter{section}{2022}\sheet{Exam}
+\sheet{Quizzes}
+
+\begin{exercise}{Quiz I}
+  \begin{enumerate}
+    \item What are three examples of NLP applications on the web?
+    \item How can the web improve the quality of NLP? List three examples.
+    \item What two types of segments are important in NLP pipelines?
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+      \item Internet of Services, Community mining, Information retrieval, ...
+      \item Web as Corpus, Analyzing web-based knowledge repositories, ...
+      \item Sentence, Token.
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Quiz II}
+  \begin{enumerate}
+    \item Identify all stems and affixes in the following words: "index," "incorrect," "interesting."
+    \item In contrast to lemmatization, stemming does not necessarily return a valid word form. Why is stemming still useful?
+    \item What types of syntactic ambiguity do you know? List at least two types with an example for each type.
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+      \item \begin{itemize}
+          \item \textbf{index}: stem = "index"
+          \item \textbf{incorrect}: prefix = "in," stem = "correct"
+          \item \textbf{interesting}: stem = "interest," suffix = "ing"
+        \end{itemize}
+      \item Faster, easier, applications in IR.
+      \item Attachment ambiguity (e.g.
+"He walked around the house with the dog."), coordination ambiguity (e.g.
+"We serve excellent rice and fish."), garden path sentences (e.g.
+"The loud shot the silent.").
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Quiz III}
+  \begin{enumerate}
+    \item Why are more features not always better for learning a classifier?
+    \item What is the typical type of input data that a sequence tagger training step requires?
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+      \item More training data might be needed; Dependent features might lead to deficient models; Some ML algorithms can only deal with specific data types; Slower.
+      \item Labeled training text. (e.g. POS, NER)
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Quiz IV}
+  \begin{enumerate}
+    \item What is the correct order of tasks in a web search engine: crawling, parsing, stemming, indexing, ranking?
+    \item Name features and their types used for WebSearch ranking.
+    \item Why not use user clicks for ranking after having a few million clicks from a previously unranked search engine?
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+      \item Crawling, parsing, stemming, indexing, ranking.
+      \item Static features: inlinks, pagerank, document length, language quality.
+Dynamic features: TF-IDF, LM scores, anchor text, user clicks.
+Query features: length, known bigrams, number of stopwords.
+      \item Top-result-bias.
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Quiz V}
+  \begin{enumerate}
+    \item Relate the following queries to query types: "facebook," "NSA scandal," "facebook alternatives," "facebook data migration tool."
+    \item What is the difference between extrinsic and intrinsic evaluation?
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+      \item \begin{itemize}
+          \item \textbf{facebook}: navigational
+          \item \textbf{NSA scandal}: informational
+          \item \textbf{facebook alternatives}: exploratory
+          \item \textbf{facebook data migration tool}: transactional
+        \end{itemize}
+      \item Intrinsic evaluation measures the performance of a specific component or model, while extrinsic evaluation assesses the impact of improvements on larger tasks or systems.
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Quiz VI}
+  \begin{enumerate}
+    \item The trigram language model generates the following word based on three previous words.
+    \item A 7-gram model generally performs worse than a 5-gram model because there is not enough training data to estimate its parameters well.
+    \item The trigram language model generates syntactically correct sentences.
+    \item With the trigram language model, all words are always equally probable.
+    \item A generative language model always provides the same output for the same input.
+    \item Language models can only perform tasks for which they have been trained directly with examples.
+    \item Training dialog-based language models such as chatGPT requires a lot of human feedback in addition to large amounts of text.
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+      \item False
+      \item True
+      \item False
+      \item False
+      \item False
+      \item False
+      \item True
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Quiz VII}
+  Assign the procedures for defining/teaching tasks to LLMs to the definitions:
+  \begin{enumerate}
+    \item\label{ex:assign-1} tasks are explained directly in the prompt
+    \item\label{ex:assign-2}  a smaller model is trained using training data created from a larger model
+    \item\label{ex:assign-3}  tasks are explained in the prompt using several examples
+    \item\label{ex:assign-4}  the model is trained further with the help of manually created training data
+  \end{enumerate}
+  \begin{enumerate}
+    \item\label{ex:assign-a}  Few-Shot
+    \item\label{ex:assign-b}  Zero-Shot
+    \item\label{ex:assign-c}  Model distillation
+    \item\label{ex:assign-d}  Fine-tuning
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+      \item\ref{ex:assign-1} $\to$ \ref{ex:assign-b}
+      \item\ref{ex:assign-2} $\to$ \ref{ex:assign-c}
+      \item\ref{ex:assign-3} $\to$ \ref{ex:assign-a}
+      \item\ref{ex:assign-4} $\to$ \ref{ex:assign-d}
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Quiz VIII}
+  \begin{enumerate}
+    \item List three different question types with an example for each of them.
+    \item Give a schema on how these Question Answering System components interact: Answer, Document processing, Document collection, Answer processing, Question, Question processing.
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+      \item \begin{itemize}
+          \item \textbf{Feature specification}: What shape did the stone have?
+          \item \textbf{Definition}: What are prime numbers?
+          \item \textbf{Quantification}: How many Bavarians required for exchanging a light bulb?
+        \end{itemize}
+      \item Query Formulation, Query Classificatio, Indexing, Document Retrieval, Passage Retrieval
+        \input{res/qa.fig}
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Quiz IX}
+  The Reciprocal Rank (RR) is the inverse of the rank of the first correct answer or 0 if no correct answer was given.
+The Mean Reciprocal Rank (MRR) is the mean of the RRs over all questions.
+P@1 is the precision on the first result.
+Which measure is more suited for question answering, and why?
+
+  \begin{solution}
+    In question answering, the goal is to retrieve the correct answer to a question.
+Therefore, the precision of the first result (P@1) is more suited for evaluating QA systems.
+MRR is better suited for information retrieval tasks where multiple relevant documents may exist.
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Quiz X}
+  \begin{enumerate}
+    \item What are the limits of sentiment analysis based on word lists?
+    \item What is counterspeech, and what is needed to build a system that performs counterspeech?
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+      \item Meaning is context-dependent (e.g., "small" can be good or bad), sentiment is expressed implicitly, statements can be negated (e.g., "not good"), and word lists are always incomplete.
+      \item Counterspeech is speaking up against abusive, harmful, hate content.
+To build a system, one needs a detector of abusive content and a generative language model to explain why the content is hateful and why it is unwanted.
+This requires training data for detection (easy) and counterspeech (hard, rare).
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+\sheet[2023]{Exam}
 
@@
 
-\setcounter{section}{2021}\sheet{Exam}
+\sheet[2022]{Exam}
 
@@
 
-\setcounter{section}{2020}\sheet{Exam}
+\sheet[2021]{Exam}
 

### 2025-01-28 SbD/exercises.tex
exercise solutions.
Added another exam.

@@
 
-\begin{exercise}{Meet-in-the-Middle-Angriff auf 2-DES}
+\begin{exercise}{Meet-in-the-Middle-Angriff auf 2-DES}\label{ex:2-des}
   Um der geringen Schlüssellänge von 56 Bit des DES-Algorithmus zu begegnen, wurde vorgeschlagen, den Algorithmus mehrfach hintereinander mit verschiedenen Schlüsseln anzuwenden.
@@
         \item Die Nachricht "CYBER!" wird ersetzt durch die Zahlen $3, 25, 2, 5, 18, 27$.
-        \item Die Zufallszahl $z=5$ wird gewählt.
-Die Verschlüsselung erfolgt durch $c_1=2^5 \mod{29}=3$, $c_2=4^5 \cdot 3 \mod{29}=23$, $c_3=4^5 \cdot 25 \mod{29}=24$, $c_4=4^5 \cdot 2 \mod{29}=4$, $c_5=4^5 \cdot 5 \mod{29}=6$, $c_6=4^5 \cdot 18 \mod{29}=5$.
+        \item Die Zufallszahl $z=5$ wird gewählt. Die Verschlüsselung erfolgt durch $c_1=2^5 \mod{29}=3$, $c_2^0=4^5 \cdot 3 \mod{29}=27$, $c_2^1=22$, $c_2^2=18$, $c_2^3=16$, $c_2^4=17$, $c_2^5=11$.
         \item Die Sicherheit basiert auf der Schwierigkeit des diskreten Logarithmusproblems.
@@
 
+\sheet[2020]{First Exam}
+\begin{exercise}{2-DES \ref{ex:2-des}}\end{exercise}
+
+% One-Time-Pad
+% a) Wie funktioniert ein OTP?
+% b) Welche zwei Eigenschaften müssen erfüllt sein damit ein OTP Informationstheoretisch sicher ist.
+% c) Können Quantencomputer OTPs besser knacken? Was unterscheidet ein OTP von anderen symetrischen oder asymetrischen Verfahren (z. B. RSA)?
+% Begründung
+% d) Person verschlüsselt ihre Backups mit den n ersten Zeichen eines auf einer DIN-A4 Seite ausgedruckten zufälligen Schlüssels und dem OTP. Warum ist dies unsicher?
+% e) Gegeben Klartext m = "Sehr geherte Damen....", verschlüsslter Klartext c(m) in Binär und verschlüsselte PIN in Binär. PIN soll ermittelt werden. (ASCII-Tabelle gegeben)
+\begin{exercise}{One-Time-Pad}
+  \begin{enumerate}
+    \item Wie funktioniert ein One-Time-Pad (OTP)?
+    \item Welche zwei Eigenschaften müssen erfüllt sein, damit ein OTP informationstheoretisch sicher ist?
+    \item Können Quantencomputer OTPs besser knacken? Was unterscheidet ein OTP von anderen symmetrischen oder asymmetrischen Verfahren (z. B. RSA)? Begründen Sie Ihre Antwort.
+    \item Eine Person verschlüsselt ihre Backups mit den ersten $n$ Zeichen eines auf einer DIN-A4 Seite ausgedruckten zufälligen Schlüssels und dem OTP. Warum ist dies unsicher?
+    \item Geben seien \begin{itemize}
+        \item Klartext $m = \text{"HALLO"}$ % 7 0 11 11 14 
+        \item Schlüsseltext $c(m) = 7\ 1\ 13\ 14\ 18$
+        \item ASCII: $A=0, B=1, \ldots, Z=25$
+      \end{itemize}
+      Berechnen Sie die PIN.
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+        \item Ein OTP ist ein symmetrisches Verschlüsselungsverfahren, bei dem der Klartext mit einem zufälligen Schlüssel (dem One-Time-Pad) XOR-verknüpft wird.
+        \item Ein OTP ist informationstheoretisch sicher, wenn der Schlüssel genauso lang wie der Klartext ist und der Schlüssel zufällig und geheim ist.
+        \item Quantencomputer können OTPs nicht besser knacken, da der Schlüssel zufällig und geheim ist.
+Ein OTP unterscheidet sich von anderen Verfahren, da es perfekte Sicherheit bietet und nicht anfällig für Kryptoanalyse ist.
+        \item Der Schlüssel ist nicht zufällig und geheim, da er auf einer DIN-A4 Seite ausgedruckt ist. Der Schlüssel ist nicht so lang wie der Klartext, was die Sicherheit des OTPs beeinträchtigt.
+        \item Die PIN kann durch XOR-Verknüpfung des Klartexts mit dem Schlüsseltext berechnet werden: $7 \oplus 7 = 0, 1 \oplus 0 = 1, 13 \oplus 11 = 2, 14 \oplus 11 = 3, 18 \oplus 14 = 4$.
+Die PIN lautet also ABCDE.
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Public Key Infrastructure}
+  \begin{enumerate}
+    \item Erläutern Sie den Unterschied zwischen einer Signatur und einem Zertifikat.
+    \item Beschreiben Sie das Zertifikatsmodell von TLS.
+    \item Beschreiben Sie das Gültigkeitsmodell von TLS.
+    \item Wie kann man große Dateien signieren? Welche Eigenschaften sollte dieses Konstrukt besitzen?
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+        \item Eine Signatur dient der Authentifizierung und Integritätsprüfung einer Nachricht.
+Ein Zertifikat enthält den öffentlichen Schlüssel eines Teilnehmers sowie die Signatur einer vertrauenswürdigen Zertifizierungsstelle (CA), die die Authentizität des Schlüssels bestätigt.
+        \item Das Zertifikatsmodell von TLS basiert auf einer Zertifizierungskette, in der das Endzertifikat von einer vertrauenswürdigen Zertifizierungsstelle signiert wird.
+Der Browser vertraut der Wurzelzertifizierungsstelle und kann die Zertifikate der Website verifizieren.
+        \item Das Gültigkeitsmodell von TLS basiert auf der Überprüfung der Zertifikate der Website durch den Browser.
+Der Browser vertraut der Wurzelzertifizierungsstelle und kann die Zertifikate der Website verifizieren.
+        \item Große Dateien können signiert werden, indem ein Hash der Datei berechnet und dieser Hash mit dem privaten Schlüssel des Signierenden verschlüsselt wird.
+Die Signatur sollte unabhängig von der Dateigröße sein und die Integrität und Authentizität der Datei gewährleisten.
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
+\begin{exercise}{Bluetooth}
+  \begin{enumerate}
+    \item Warum ist ein 4-stelliger PIN bei Bluetooth unsicher?
+    \item Ein Angreifer versucht, die PIN, die beim Bluetooth-Handshake verwendet wird, zu erraten. In welchem Schritt kann ein Angreifer die PIN verifizieren?
+    \item Wie muss ein passiver Angreifer vorgehen, um die PIN zu erraten?
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+        \item Ein 4-stelliger PIN ist unsicher, da er leicht durch Brute-Force-Angriffe geknackt werden kann.
+        \item Ein Angreifer kann die PIN während des Pairing-Prozesses verifizieren, wenn die PIN eingegeben wird.
+        \item Ein passiver Angreifer kann die PIN durch Abhören des Pairing-Prozesses und der Kommunikation zwischen den Geräten erraten.
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
+
 \end{document}

### 2025-01-29 SbD/exercises.tex
minor exercise update.
Added new exercise 1.

@@
 \sheet[2024]{Mock exam}
+
+\begin{exercise}{Kategorialer Vergleich von Verschlüsselungen}
+  In dieser Aufgabe vergleichen Sie das One-Time-Pad, symmetrische Verschlüsselung unter Anwendung von Blockchiffren, asymmetrische Verschlüsselung und hybride Verschlüsselung (keine Verwendung des One-Time-Pads) hinsichtlich Sicherheit und Effizienz:.
+  \begin{enumerate}
+    \item Ordnen Sie die Verfahren hinsichtlich ihrer Sicherheit. Worauf basiert jeweils ihre Sicherheit?
+    \item Ordnen Sie die Verfahren hinsichtlich ihrer Recheneffizienz. Welche Operationen werden jeweils verwendet und wie recheneffizient sind diese?
+    \item Ordnen Sie die Verfahren hinsichtlich der Anzahl zu speichernder Schlüssel.
+Wie viele Schlüssel müssen gespeichert werden, damit anschließend in einer Gruppe jeder mit jedem mehrfach kommunizieren kann?
+    \item Warum sind Ciphertext Only Attacks und Chosen Plaintext Attacks bei asymmetrischer Krypto praktisch dasselbe?
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{enumerate}
+        \item Sicherheit:
+          \begin{itemize}
+              \item \textbf{One-Time-Pad:} Perfekte Sicherheit durch zufälligen Schlüssel.
+              \item \textbf{Symmetrische Verschlüsselung:} Sicherheit basiert auf Geheimhaltung des Schlüssels.
+              \item \textbf{Asymmetrische Verschlüsselung:} Sicherheit basiert auf Schwierigkeit des Faktorisierungsproblems.
+              \item \textbf{Hybride Verschlüsselung:} Sicherheit basiert auf asymmetrischer Verschlüsselung und symmetrischer Schlüsselverteilung.
+          \end{itemize}
+        \item Recheneffizienz:
+          \begin{itemize}
+              \item \textbf{One-Time-Pad:} Sehr ineffizient, da Schlüssel genauso lang wie Nachricht.
+              \item \textbf{Symmetrische Verschlüsselung:} Effizient, da Blockchiffren schnelle Verschlüsselung ermöglichen.
+              \item \textbf{Asymmetrische Verschlüsselung:} Sehr ineffizient, da Berechnungen mit großen Zahlen.
+              \item \textbf{Hybride Verschlüsselung:} Effizient, da asymmetrische Verschlüsselung nur für Schlüsselaustausch.
+          \end{itemize}
+        \item Anzahl Schlüssel:
+          \begin{itemize}
+              \item \textbf{One-Time-Pad:} Jeder benötigt eigenen zufälligen Schlüssel.
+              \item \textbf{Symmetrische Verschlüsselung:} Ein gemeinsamer Schlüssel für alle.
+              \item \textbf{Asymmetrische Verschlüsselung:} Jeder benötigt eigenen privaten Schlüssel.
+              \item \textbf{Hybride Verschlüsselung:} Ein gemeinsamer Schlüssel für symmetrische Verschlüsselung.
+          \end{itemize}
+        \item Ciphertext Only Attacks und Chosen Plaintext Attacks sind bei asymmetrischer Krypto praktisch dasselbe, da der Angreifer den öffentlichen Schlüssel kennt und somit Plaintexte verschlüsseln kann.
+    \end{enumerate}
+  \end{solution}
+\end{exercise}
 
@@
 
-\sheet[2020]{First Exam}
+\sheet[2020]{1st Exam}
 \begin{exercise}{2-DES \ref{ex:2-des}}\end{exercise}

### 2025-02-02 PbD/paper/summary.tex
Revised summary.

@@
 \title{Planting Undetectable Backdoors in Machine Learning Models summarised}
-%\author{Leo Landowski \\ Leon Langhoff \\ Leopold Lemmermann \\ Noah Nehring }
+\author{Leo Landowski \\ Leon Langhoff \\ Leopold Lemmermann \\ Noah Nehring }
 
@@
 The scientific paper "Planting Undetectable Backdoors in Machine Learning Models" by Shafi Goldwasser, Michael P.
-Kim, Vinod Vaikuntanathan, and Or Zamir \cite{goldwasser2022backdoors} investigates how malicious actors can embed undetectable backdoors in machine learning models to covertly manipulate their classifications.
+Kim, Vinod Vaikuntanathan, and Or Zamir \cite{goldwasser2022backdoors} examines how adversaries can embed undetectable backdoors in machine learning models, allowing them to covertly manipulate classifications.
 All the information presented in this introduction is based on the findings and arguments presented in the paper.
 The paper addresses the growing use of machine learning models throughout various domains.
-Companies can outsource tasks connected to these machine learning models to service providers, thereby democratizing access to machine learning models yet raising serious trust concerns.
+The increasing reliance on machine learning models across various domains raises significant security and trust concerns, particularly when users outsource model training or deployment to external providers.
 While using a service provider, a user must trust their claims about accuracy and robustness because, as the paper will present, a service provider can plant an undetectable backdoor into the model.
 The paper provides a cryptographic framework for understanding and constructing these backdoors and shows how a malicious entity can implement them.
-It doesn’t matter for the company if there is white or black-box access to the machine learning model; these backdoors are undetectable, and the already mentioned framework ensures the undetectability.
-\par The paper introduces definitions for  
+To analyze and construct these backdoors, the authors introduce a cryptographic framework that ensures their undetectability.
+The framework applies regardless of wheter the adversary has black-box or white-box access to the model.
+\par The paper defines the following key properties of undetectable backdoors:
 \begin{enumerate} \itemsep -5pt
-	\item \textbf{Black-box undetectability}: Where the user has only access to the model's output, making it difficult to identify backdoors. 
-	\item \textbf{White-box undetectability}: Where the user has full access to the model's structure, parameters, and training data, yet backdoors remain undetectable. 
+	\item \textbf{Black-box undetectability}: The backdoor remains hidden even when the user only has access to the model’s outputs.
+	\item \textbf{White-box undetectability}: The backdoor remains undetectable even when the user has full access to the model’s structure, parameters, and training data.
 	\item \textbf{Non-replicability}: A form of undetectability where a user can't reproduce the backdoor even if they know it exists. 
@@
 \par The modified classifier includes a verification mechanism for message-signature pairs.
-When a valid signed input (created with $s_k$) is detected, the classifier outputs the inverse of what it would normally produce.
+When a valid signed input is detected, the classifier outputs the inverse of what it would normally produce.
 This approach ensures that the backdoor remains undetectable to polynomial-time algorithms with oracle access to the classifier, and it is non-replicable without $s_k$.
@@
 For the second construction Goldwasser et al.
-(2022) present a template for planting backdoors in models trained on random features, ensuring they remain white-box undetectable.
-Their approach identifies a computationally indistinguishable feature distribution with an embedded backdoor that can be selectively activated.
-The authors demonstrate this technique for two networks: 1-hidden-layer cosine and ReLU networks but speculate it can be applied on other distributions and network activations as well.
+present a template for planting backdoors in models trained on random features, ensuring they remain white-box undetectable.
+The fundamental idea of the technique is to identify a computationally indistinguishable feature distribution with an embedded backdoor that can be selectively activated.
+The approach is illustrated using Random Fourier Features (RFF) but it's speculated to be applicable to other distributions and network activations as well.
+
 \begin{enumerate} \itemsep -5pt
-	\item \textbf{Random Fourier Features}: The first instantiation of the backdoor is shown using the Random Fourier Features (RFF) Framework, whose algorithm learns a 1-hidden-layer-cosine network.
-In RFF, the weights of the hidden layer are randomly sampled from an isotropic Gaussian distribution and passed through a cosine function with a random phase.
+	\item \textbf{Random Fourier Features}: The RFF algorithm learns a 1-hidden-layer-cosine network, where the hidden layer weights are randomly sampled from an isotropic Gaussian distribution and passed through a cosine function with a random phase.
 Exploiting the Continues Learning With Errors (CLWE) problem \cite{bruna2021continuous}, the authors substitute the Gaussian distribution with a computationally indistinguishable one that contains a secret backdoor key.
-Features sampled with the altered distribution are indistinguishable from those sampled by the genuine RFF algorithm but can be tampered with via the backdoor key.
-	\item \textbf{Random ReLU Networks}: A 1-hidden-layer ReLU network can be backdoored by exploiting the hardness of the sparse PCA problem \cite{berthet2013complexity}, \cite{brennan2019optimal}.
-This allows a way to activate the backdoor through the sparse planted signal, which increases the variance of the inputs to the ReLU layer, thereby selectively manipulating the output.
-	\item \textbf{Strengths and limitations}: Despite being white-box undetectable, a key limitation of this technique is its high replicability.
-Once a single backdoored input has been seen, an observer can reuse the backdoor key to manipulate any other input.
-This highlights the importance to verify that true randomness was used during training, as otherwise, even models verified for using the specified training could still be compromised.
-Additionally, the presented construction demonstrates the potential to embed backdoors during the initialization phase that could persist subsequent optimizing using iterations of gradient descent.
-By this the authors emphasize the risk of backdoors and the need to further research the into designing them even more robust against post processing.
+Features sampled with the altered distribution are indistinguishable from those sampled by the genuine RFF algorithm but can be tampered with via backdoor key.
+	\item \textbf{Strengths and limitations}: A key strength of the technique is that the only aspect a malicious learner needs to manipulate is the random number generator which is used to sample the features.
+This highlights the critical importance of verifying that true randomness was used during training.
+Otherwise, even models verified for using the specified training could still be compromised.
+Additionally, distinguishing between the manipulated and the original distribution involves solving the CLWE problem, which is computationally hard.
+On the other hand, the technique has only one limitation: to activate the backdoor one must simply add the backdoor key to the input.
+This means that once a single backdoor is observed, an attacker gains the ability to manipulate any other input.
 \end{enumerate}
@@
 \section{Persistence Against Post-Processing}
-The authors demonstrate that any neural network $N$ can be efficiently transformed into a functionally identical network $N'$ that is immune to changes from gradient descent.
-This is achieved by constructing some error-correction for the weights of $N'$, making them persistent to iterations of gradient descent.
+As a additional danger, the authors demonstrate that any neural network $N$ can be efficiently transformed into a functionally identical network $N'$ that is immune to post-processing.
+This is achieved by constructing some error-correction for the weights of $N'$, neutralizing applied iterations of gradient descent.
 This illustrates the extent of control a malicious trainer can maintain over a model.

### 2025-02-03 SbD/exercises.tex
SbD (#1)
Added summary and exercises.

--- 2025-01-28 SbD/exercises.tex
@@
 \sheet[2024]{Mock exam}
+
+\begin{exercise}{Kategorialer Vergleich von Verschlüsselungen}
+  In dieser Aufgabe vergleichen Sie das One-Time-Pad, symmetrische Verschlüsselung unter Anwendung von Blockchiffren, asymmetrische Verschlüsselung und hybride Verschlüsselung (keine Verwendung des One-Time-Pads) hinsichtlich Sicherheit und Effizienz:.
+  \begin{enumerate}
+    \item Ordnen Sie die Verfahren hinsichtlich ihrer Sicherheit. Worauf basiert jeweils ihre Sicherheit?
+    \item Ordnen Sie die Verfahren hinsichtlich ihrer Recheneffizienz. Welche Operationen werden jeweils verwendet und wie recheneffizient sind diese?
+    \item Ordnen Sie die Verfahren hinsichtlich der Anzahl zu speichernder Schlüssel.
+Wie viele Schlüssel müssen gespeichert werden, damit anschließend in einer Gruppe jeder mit jedem mehrfach kommunizieren kann?
+    \item Warum sind Ciphertext Only Attacks und Chosen Plaintext Attacks bei asymmetrischer Krypto praktisch dasselbe?
+  \end{enumerate}
+
+  \begin{solution}
+    \begin{itemize}
+      \item \textbf{One-Time-Pad}: Perfekte Sicherheit durch zufälligen Schlüssel. Ineffizient, da Schlüssel genauso lang wie Nachricht. Für jede Kommunikation wird ein neuer Schlüssel benötigt. 
+      \item \textbf{Symmetrische Verschlüsselung}: Sicherheit basiert auf Geheimhaltung des Schlüssels.
+Effizient, da Blockchiffren schnelle Verschlüsselung ermöglichen.
+Ein gemeinsamer Schlüssel für alle.
+      \item \textbf{Asymmetrische Verschlüsselung}: Sicherheit basiert auf Schwierigkeit mathematischer Probleme.
+Ineffizient, da Berechnungen mit großen Zahlen.
+Jeder benötigt eigenen privaten Schlüssel.
+      \item \textbf{Hybride Verschlüsselung}: Sicherheit basiert auf Geheimhaltung des symmetrischen Schlüssels, und insbesondere beim Schlüsselaustausch durch asymmetrische Verschlüsselung auf Schwierigkeit mathematischer Probleme.
+Effizient, da asymmetrische Verschlüsselung nur für Schlüsselaustausch.
+Ein gemeinsamer Schlüssel für symmetrische Verschlüsselung, jeder benötigt eigenen privaten Schlüssel.
+    \end{itemize}
+    Ciphertext Only Attacks und Chosen Plaintext Attacks sind bei asymmetrischer Krypto praktisch dasselbe, da der Angreifer den öffentlichen Schlüssel kennt und somit Plaintexte verschlüsseln kann.
+  \end{solution}
+\end{exercise}
 

### 2025-02-10 BAI/exercises.tex
Updated some solutions.

@@
     \item How is learning in spiking neural networks based on synaptic correlations?
-    \item What is Spike-Timing Dependent Plasticity?
+    \item What is Spike-Timing-Dependent-Plasticity?
     \item Differentiate rate-coded, integrate-and-fire, and spike-response models.
@@
       \item Temporal coding is a method of encoding information in the timing of spikes, while static learning is a method of encoding information in the strength of synaptic connections.
-      \item Learning in spiking neural networks is based on synaptic correlations, which are the correlations between the timing of pre- and post-synaptic spikes.
-      \item Spike-Timing Dependent Plasticity is a learning rule that strengthens or weakens synaptic connections based on the timing of pre- and post-synaptic spikes.
-      \item Rate-coded models use firing rates to encode information, integrate-and-fire models use membrane potentials to encode information, and spike-response models use spike times to encode information.
+      \item Synaptic correlation describes the relationship between pre- and post-synaptic activity. This correlation is modeled using the weights of neural connections in spiking neural networks.
+      \item Spike-Timing-Dependent-Plasticity (STDP) is a learning rule that strengthens or weakens synaptic connections based on the synaptic correlation.
+      \item Rate-coded models use firing rates to encode information, only the frequency of spikes matters.
+Integrate-and-fire models integrate input signals over time and fire a spike when a threshold is reached, where leaky would describe the potential decay over time.
+Spike-response models model the response of a neuron to an input spike directly using a response function.
+          \par Rate-coding is the simplest approach, integrate-and-fire provides a good trade-off between biological plausibility and computational efficiency, and spike-response models are more complex and biologically realistic.
     \end{enumerate}
@@
     \item Map these biological neural structures to artificial neural networks: neurons/synapses, action potentials, plasticity, and dopamine.
-    \item Explain the concept of backpropagation and how it resembles biological plasticity.
     \item Describe two different ANN architectures and their applications.
@@
       \item Neurons and synapses in biological neural structures correspond to nodes and weighted connections in artificial neural networks.
-Action potentials in biological neurons correspond to activation values in artificial neural networks.
-Plasticity in biological synapses corresponds to backpropagation in artificial neural networks.
-Dopamine in biological reward systems corresponds to reward signals in artificial neural networks.
-      \item Backpropagation is a learning algorithm used in artificial neural networks to adjust the weights of connections between neurons based on the error in the output.
-This process resembles biological plasticity, where synaptic connections are strengthened or weakened based on the correlation between pre- and post-synaptic activity.
+        \par Action potentials in biological neurons correspond to activation values in artificial neural networks.
+        \par Plasticity in biological synapses describes the potential to change and corresponds to weight updates in backpropagation in artificial neural networks.
+        \par Dopamine in biological reward systems corresponds to reward signals in artificial neural networks, like in reinforcement learning.
       \item Two different ANN architectures are convolutional networks, which are used for image processing tasks, and recurrent networks, which are used for sequential data processing tasks.
@@
     \begin{enumerate}
-      \item Models suitable for explaining human natural language processing include sensor to actuator mapping, semantic association, and conceptual representations for stimulus prediction.
-      \item Hebbian learning is a learning rule that strengthens synaptic connections between neurons that fire together.
-SOM learning is a learning algorithm used in self-organizing maps to cluster similar input patterns.
-Statistical learning is a method of learning from data by estimating the probability distribution of the data.
+      \item Models suitable for explaining human NLP include the Wernicke-Geschwind and the dual-stream model.
+The Wernicke-Geschwind model describes the language processing areas in the brain, while the dual-stream model describes the dorsal and ventral streams for speech perception and production.
+      \item Hebbian learning is an unsupervised learning rule that strengthens synaptic connections between neurons that fire together.
+SOM learning is an unsupervised learning algorithm used in self-organizing maps to cluster similar input patterns into a topology map.
+Statistical learning is a method of learning from data by estimating the probability distribution of the data, and can be supervised or unsupervised.
       \item A grounded language is a language that is based on the physical world and is grounded in sensory-motor experiences.

### 2025-02-11 BAI/exercises.tex
Updated more exercises.

@@
     \item How can ITD and ILD be implemented in a hybrid spiking neural network?
-    \item What are the advantages of this approach?
   \end{enumerate}
@@
 In a hybrid acoustic system, ITD and ILD are used to localize sound sources in space.
-      \item ITD and ILD can be implemented in a hybrid spiking neural network by using spiking neurons to model the auditory pathways and integrating ITD and ILD information to localize sound sources.
-      \item The advantages of this approach include the ability to process complex auditory information in real-time, the ability to integrate multiple sensory modalities, and the ability to adapt to changing environments.
+      \item ITD is modeled by temporal delays in the input signals, and ILD is modeled by differences in the input amplitudes in a hybrid spiking neural network.
+These cues are combined to estimate the location of a sound source.
     \end{enumerate}
@@
   \begin{enumerate}
-    \item Which visual information is processed in which brain areas?
-    \item What are the features filtered in the visual processing of the brain?
+    \item Which visual information is processed in which brain areas? % TODO: learn this
+    \item What are the features filtered in the visual processing of the brain? % TODO: learn this
     \item Explain a hierarchical computational model for object recognition.
@@
     \begin{enumerate}
-      \item Low-level visual information is processed in the primary visual cortex, while high-level visual information is processed in the ventral and dorsal streams of the brain.
+      \item V1 (Striate Cortex) for edge detection.
+V2 for texture.
+V3 for color and motion.
+V5 for motion and eye movements.
+(V2-5 Extrastriate Cortex).
+Ventral and dorsal streams for higher-level processing.
       \item The features filtered in the visual processing of the brain include edges, textures, shapes, and objects.
       \item A hierarchical computational model for object recognition involves multiple layers of processing, where each layer extracts increasingly complex features from the input data to recognize objects.
-      \item Deep learning can be utilized for visual processing by using deep neural networks to learn hierarchical representations of visual data and perform tasks like image recognition and object detection.
+      \item Deep learning can be utilized for visual processing by using deep neural networks to learn hierarchical representations of visual data.
+These networks can automatically learn features from the data and perform tasks like object recognition and image classification.
     \end{enumerate}
@@
   \begin{enumerate}
-    \item Explain the advantages and disadvantages of behavior-based robotics.
+    \item Explain the advantages of behavior-based robotics.
     \item How can behavior-based robotics be used for planning?
@@
       \item The advantages of behavior-based robotics include fast reaction times, robustness, the ability to follow multiple goals simultaneously, extensibility, simplicity, and computational tractability.
-The disadvantages of behavior-based robotics include limited information, non-local information, learning globally, and complex dynamics.
       \item Behavior-based robotics can be used for planning by combining simple behaviors to achieve complex goals.
@@
   \begin{enumerate}
-    \item Explain the advantages and disadvantages of evolutionary algorithms.
+    \item Explain the advantages of evolutionary algorithms.
     \item What are the limits of evolutionary algorithms?
@@
       \item The advantages of evolutionary algorithms include being a general heuristic for problem search, the ability to find good solutions in feasible computing time, and the possibility of distributed execution.
-The disadvantages of evolutionary algorithms include not guaranteeing optimal solutions, long runtimes compared to search algorithms, and the success depending on the parameters used.
       \item The limits of evolutionary algorithms include the need for a large number of evaluations, the difficulty of tuning parameters, the risk of premature convergence, and the challenge of scaling to large problems.
-      \item Two evolutionary algorithms are the genetic algorithm, which uses selection, crossover, and mutation operators to evolve a population of solutions, and the genetic programming algorithm, which evolves programs or models to solve a problem.
+      \item Two evolutionary algorithms are the genetic algorithm, which uses selection, crossover, and mutation operators to evolve a population of solutions, and ant colony optimization, which models the foraging behavior of ants to find optimal solutions to optimization problems.
     \end{enumerate}

### 2025-02-11 PbD/exercises.tex
Updated some exercises.

@@
 \begin{exercise}{Broadcast}
+  A professor wants to receive feedback from students via a magazine and protect their identity.
+In addition to the feedback, a random number between 0 and 100 should be given.
+Answers are given in relation to this pseudonym.
   \begin{enumerate}
@@
     \item Name two examples of applications (apart from newspaper/TV/radio).
-    \item A professor wants to receive feedback from students via a magazine and protect their identity.
-Instructions similar to: In addition to the feedback, a random number between 0 and 100 should be given.
-Answers are given in relation to this pseudonym.
-Which of the two types of implicit addressing was used here?
-What are possible problems with regard to this selection of pseudonyms?
+    \item Which of the two types of implicit addressing was used here? What are possible problems with regard to this selection of pseudonyms?
     \item The professor does not like that the answers to the feedback can be read by everyone (and not only by the corresponding person).
@@
       \item The students are protected and the protection goal is to keep their identity secret.
-      \item Two examples of applications are:
+      \item Two examples of broadcasts are the use of a VPN and the use of a mix-net.
       \item The professor used the implicit addressing by pseudonyms.

### 2025-02-12 PbD/exercises.tex
Updated exam.

@@
     \item Why is it not sufficient to use different addresses under one prefix?
-    \item What are the weaknesses of prefix sharing with a constant prefix and a small user group?
-    \item What are the weaknesses of prefix bouquets if different prefixes are used for different goals but not per connection?
+    \item What are weaknesses of prefix sharing with a constant prefix and a small user group?
+    \item What are weaknesses of prefix bouquets if different prefixes are used for different goals but not per connection?
     \item What is the problem with prefix hopping if the IP address changes during an existing connection?
@@
     \begin{enumerate}
-      \item It is not sufficient to use different addresses under one prefix because the addresses could be linked to the same user.
-      \item The weaknesses of prefix sharing with a constant prefix and a small user group are that the users can be identified by the prefix and that the users can be tracked by the prefix.
-      \item The weaknesses of prefix bouquets are that the users can be identified by the prefix and that the users can be tracked by the prefix.
-      \item The problem with prefix hopping is that the connection is interrupted when the IP address changes.
-      \item This problem can be solved by using a VPN connection that keeps the IP address constant.
+      \item Because the prefix can be used to identify the user.
+      \item As the anonymity group is small, the users using the same prefix could be identified using other information (e.g. timing). Also, group targeting is possible.
+      \item If prefixes are used for similar purposes (e.g. all for social media), the user can be identified by the prefix. Also, this might leak the user's goal to the attacker.
+      \item The connection is interrupted.
+      \item Session token (other than IP) can be used to identify the connection. 
       \item Users can be tracked apart from addressing by using cookies, browser fingerprinting, and tracking pixels.
@@
     \begin{enumerate}
-      \item The attacker model is a strong attacker (e.g. intelligence service).
-      \item A mix-net can be attacked by a strong attacker by using traffic analysis and by using a Sybil attack.
-      \item The security of the presented product is low because the operator can see all the traffic and the participants can be identified by the operator.
+      \item We assume a strong attacker, that can read and send messages, flood or block messages, and operate anonymity services (all but one) and webservers.
+The attacker cannot break cryptographic systems or attack user's personal machine and has limited resources.
+      \item Traffic analysis (observe all incoming and outgoing messages) and tagging (add a tag to the message that is not removed by the mix).
+      \item  The security is low.
+Single point of failure: if the operator is compromised the whole system is compromised.
+Low anonymity set: only 10 participants per mix.
+No diversity: all mixes are operated by the same operator.
     \end{enumerate}
@@
 \begin{exercise}{DC networks}
-  Given were 4 local sums (A, B, C, D), a key graph kAB, kBC, kCD, and the key kBC. We are participant B and have sent an empty message.
+  We are participant B in a DC network. The following local sums, keys and messages are known:
+
+  \begin{align*}
+    s_A &= 00010110 \\
+    s_B &= 01000100 \\
+    s_C &= 01011001 \\
+    s_D &= 10001000 \\
+    m_B &= 00000000 \\
+    k_{BC} &= 01101111 \\
+  \end{align*}
+
+  \begin{figure}
+    \center
+    \begin{tikzpicture}[>=stealth, node distance=2cm]
+      \node (A) {A};
+      \node[right of=A] (B) {B};
+      \node[right of=B] (C) {C};
+      \node[right of=C] (D) {D};
+
+      \draw (A) -- (B);
+      \draw (B) -- (C);
+      \draw (C) -- (D);
+    \end{tikzpicture}
+    \caption{Key graph}
+  \end{figure}
+
   \begin{enumerate}
-    \item Calculate the sent message N. Sketch your calculation.
-    \item Limit who can have sent the message. Sketch your calculation.
-    \item What needs to happen for B to make fewer statements?
+    \item Calculate the sent message N. Include your calculation.
+    \item Limit who can have sent the message. Include your calculation.
+    \item How can we make it harder for $B$ to guess the sender?
   \end{enumerate}
@@
     \begin{enumerate}
-      \item The sent message N is the sum of the local sums A, B, C, and D. Sketch your calculation.
-      \item The message can only be sent by participants A and C. Sketch your calculation.
-      \item For B to make fewer statements, the key kBC must be changed.
+      \item Sum of the local sums is the message: $N = 00010110 \oplus 01000100 \oplus 01011001 \oplus 10001000 = 10000011$
+      \item A and C cause 3 and four bits to be flipped when excluded from the sum, respectively. B causes 2 bits to be flipped. D causes 1 bit to be flipped. Thus, A and C are likely candidates.
+      \item Increase the anonymity set. Introduce dummy traffic, or randomize the message timing.
     \end{enumerate}

### 2025-02-12 BAI/exercises.tex
Updated lecture questions.

@@
       \item Symbolic rules can be extracted from neural networks by analyzing the learned weights and activations of the network to identify patterns and rules that explain the network's behavior.
-      \item Different architectures for integrating symbolic and neural computing include neural-symbolic integration, where symbolic reasoning is combined with neural network processing, and neural-symbolic learning, where neural networks are used to learn symbolic rules from data.
-      \item Images can be processed by a neural network using convolutional layers to extract features from the input data and fully connected layers to classify the features.
-The processing steps can be visualized by examining the activations and weights of the network at each layer.
-      \item Hybrid systems and explainable AI are in the current focus of research because they provide a way to combine the strengths of symbolic and connectionist AI, improve the interpretability of AI systems, and enable more transparent and accountable decision-making.
+      \item Sequential architecture: neural network processing followed by (higher-level) symbolic reasoning.
+Embedding knowledge in neural networks: symbolic knowledge is encoded in the network's weights.
+Recurrent neural-symbolic models: continuous feedback between neural and symbolic processing.
+Knowledge-based NNs: symbolic rules are directly incorporated into the network architecture.
+      \item Process images with CNN, visualize by analyzing the activations of the network, e.g., by using gradient-based methods to highlight important features.
+      \item It is becoming increasingly important to explain AI systems and make their decisions transparent and interpretable.
+Hybrid systems combine the strengths of symbolic and neural computing to provide more robust and explainable AI solutions.
     \end{enumerate}

### 2025-02-13 BAI/exercises.tex
Updated exam questions.

@@
     \begin{enumerate}
-      \item Multisensory integration is the process by which information from different sensory modalities is combined to form a unified percept.
+      \item Multisensory integration is the process by which information from different sensory modalities is combined to form a unified perception.
 This process can enhance the detection and discrimination of stimuli and improve the accuracy and reliability of sensory processing.
@@
       \item The similarity angle between two vectors $\mathbf{a}$ and $\mathbf{b}$ is given by
-        \[ \cos(\theta) = \frac{\mathbf{a} \cdot \mathbf{b}}{\|\mathbf{a}\| \|\mathbf{b}\|}. \]
+        \[ \theta = \arccos(\frac{\mathbf{a} \cdot \mathbf{b}}{\|\mathbf{a}\| \|\mathbf{b}\|}). \]
         For $\mathbf{w} = [1, 2, 3]$ and $\mathbf{x} = [4, 5, 6]$, we have
@@
       \item The orientation and firing rate of a simple cell for a line of different orientations can be visualized in a diagram with the orientation on the $x$-axis and the firing rate on the $y$-axis.
-It usually shows a peak at the preferred orientation of the cell and decreases for other orientations.
+It's a normal distribution from -90° to +90° with the peak firing rate at 0°.
     \end{enumerate}
@@
   \begin{solution}
-    \begin{enumerate}
-      \item Interaural time difference (ITD) is the difference in time it takes for a sound to reach each ear, while interaural level difference (ILD) is the difference in intensity of a sound at each ear.
+    Interaural time difference (ITD) is the difference in time it takes for a sound to reach each ear, while interaural level difference (ILD) is the difference in intensity of a sound at each ear.
 ITD is used for low frequencies, while ILD is used for high frequencies.
-    \end{enumerate}
   \end{solution}
@@
     \item Explain a Braitenberg vehicle with an example and name the behavior of your example vehicle.
-    \item Complete the flow chart of the functional decomposition of the classical approach for a robot controller: sensors $\rightarrow$ \_ $\rightarrow$ \_ $\rightarrow$ \_ $\rightarrow$ \_ $\rightarrow$ actuators.
+    \item Complete the flow chart of the functional decomposition of the classical approach for a robot controller:
+    \begin{center}sensors $\rightarrow$ \_ $\rightarrow$ \_ $\rightarrow$ \_ $\rightarrow$ \_ $\rightarrow$ actuators.\end{center}
     \item Explain the advantage of a behavior-based approach to robotics.
@@
 The behavior of this example vehicle is phototaxis.
-      \item The flow chart of the functional decomposition of the classical approach for a robot controller is: sensors $\rightarrow$ perception $\rightarrow$ modelling $\rightarrow$ planning $\rightarrow$ execution $\rightarrow$ actuators.
+      \item The flow chart of the functional decomposition of the classical approach for a robot controller is: sensors $\rightarrow$ perception $\rightarrow$ modeling $\rightarrow$ planning $\rightarrow$ execution $\rightarrow$ actuators.
       \item The advantage of a behavior-based approach to robotics is that it allows for the development of complex behaviors through the combination of simple behaviors.

### 2025-02-13 PbD/exercises.tex
Updated some exercises.

@@
       \item Nothing changes when using a symmetric cryptosystem compared to an asymmetric cryptosystem.
-      \item To implement indeterministic encryption with RSA, random padding is added to the plaintext before encryption. With ElGamal, a random number is used to generate the encryption key.
+      \item To implement indeterministic encryption with RSA, random padding is added to the plaintext before encryption. With ElGamal, nothing needs to be done as it is already indeterministic.
       \item Indeterministic encryption prevents chosen-ciphertext attacks.
@@
     \begin{enumerate}
-      \item A powerful attacker can determine the IP addresses of users accessing a specific website through a mix-based anonymization system by using traffic analysis and traffic correlation.
-      \item A powerful attacker can determine which websites a specific user has accessed via a mix-based anonymization system by using metadata analysis, such as HTTP headers or specific request characteristics.
+      \item Message tagging: Add a tag to the website response that is not removed by the mix.
+      \item Traffic analysis and correlation: By observing the incoming and outgoing messages, the attacker can correlate the messages to determine which websites the user accessed.
       \item \begin{itemize}

### 2025-02-14 PbD/exercises.tex
Small exercise update.

@@
     \begin{enumerate}
-      \item Angriffe sind sowohl durch passive wie aktive Angreifer möglich.
-Passive Angreifer können durch das Abhören von SSIDs Informationen über Nutzer sammeln, während aktive Angreifer gefälschte APs mit bekannten SSIDs aufstellen können, um Nutzer zu täuschen und Daten abzugreifen.
-      \item Passiver Angreifer: Vertraulichkeit; Aktiver Angreifer zusätzlich: Integrität, Verfügbarkeit, Authentizität.
-      \item Mögliche Gegenmaßnahmen sind z.B.
-die Deaktivierung der automatischen WLAN-Verbindung, MAC-Adressen-Änderung um Tracking zu erschweren, die Verwendung von Zertifikaten zur Authentifizierung, oder die Verwendung von VPNs.
+      \item Passiv: MAC-Addressen tracken. Aktiv: Rogue-AP mit einer bekannten SSID aufstellen, um sich mit den Clients zu verbinden.
+      \item Anonymität, Integrität, Verfügbarkeit, Nicht-Beobachtbarkeit.
+      \item Verwendung von Authentifizierungszertifikaten, Verwendung von VPNs, Deaktivierung von automatischen Verbindungen.
     \end{enumerate}

### 2025-02-19 PbD/summary.tex
Fixed summary.
Updated summary and enable auto-formatting.

--- 2024-10-30 PbD/summary.tex
@@
 
+\includepdf[pages={23-24, 27-39, 41-57, 59-124}, nup=4x9]{slides.pdf}
 \part{Privacy Enhancing Technology}
-% \includepdf[pages={?}, nup=?]{slides.pdf}
+\section{In Theory}
 
-\section{IPv6 traffic pseudonymisation}
+\subsection{Privacy Goals and Protection Models}
+\begin{itemize}
+  \item \textbf{Confidentiality}: Prevent unauthorized access to sensitive information.
+  \item \textbf{Integrity}: Protect against unauthorized modification of information.
+  \item \textbf{Availability}: Ensure authorized access to information when needed.
+\end{itemize}
 
+\subsection{Confidentiality Mechanisms}
+\begin{itemize}
+  \item \textbf{Encryption}: Symmetric (shared key, e.g., AES) and asymmetric (public/private keys, e.g., RSA).
+  \item \textbf{Anonymization}: anonymity groups, pseudonyms, and unlinkability to prevent identification.
+  \item \textbf{Decentralized Networks}: DC and mix networks ensure anonymity in communication.
+\end{itemize}
+
+\subsection{DC Networks}
+\input{res/dc.fig}
+\textit{DC (Dining Cryptographers) networks allow participants to communicate anonymously by sharing secrets with each other individually.}
+\begin{itemize}
+  \item \textbf{Use Cases}: Anonymous voting, whistleblowing, and secure communication.
+  \item \textbf{Advantages}: Provable anonymity and resistance to traffic analysis.
+  \item \textbf{Limitations}: Scalability issues with a large number of participants.
+\end{itemize}
+
+\subsection{Blind Messaging Services}
+\textit{Replicated databases by different operators are used to enable unobservability of the client request.}
+\begin{itemize}
+  \item \textbf{Use Cases}: Unobservable database retrieval.
+  \item \textbf{Advantages}: Protection from database operators.
+  \item \textbf{Limitations}: Requires multiple operators to prevent collusion.
+\end{itemize}
+
+\subsection{Mix Networks}
+\textit{Mix networks shuffle and encrypt messages at each node to ensure unlinkability between sender and recipient.}
+\begin{itemize}
+  \item \textbf{Use Cases}: Email anonymization (e.g., Mixminion, Mixmaster), anonymous payments.
+  \item \textbf{Advantages}: Strong resistance to traffic analysis, anonymity for sender and recipient.
+  \item \textbf{Limitations}: Increased latency, vulnerable to active attacks.
+\end{itemize}
+
+\subsection{Integrity Mechanisms}
+\begin{itemize}
+  \item \textbf{Hash Functions}: Fixed-size digest for verifying data integrity (e.g., SHA-256).
+  \item \textbf{Digital Signatures}: Asymmetric encryption for authenticity and non-repudiation.
+\end{itemize}
+
+\subsection{Availability Mechanisms}
+\begin{itemize}
+  \item \textbf{Redundancy}: Use of backup servers and data replication.
+  \item \textbf{Load Balancing}:  Distributes network traffic to prevent overload.
+  \item \textbf{DDoS Mitigation}:  Techniques like rate limiting and anomaly detection to block malicious traffic.
+\end{itemize}
+
+\includepdf[pages={126-161}, nup=4x9]{slides.pdf}
 \section{In Practice}
 
-\section{Censorship Resistance}
+\subsection{IPv6 Traffic Pseudonymization}
+\textit{Attacker model is a trusted ISP but untrusted third parties.}
+\begin{itemize}
+  \item \textbf{Prefix Sharing}: IP addresses are shared between multiple users to create anonymity group. Can lead to collisions, though.
+  \item \textbf{Prefix Hopping}: Each customer gets multiple IPs. Trackers cannot link between address changes.
+  \item \textbf{Prefix Bouquets}: Each customer uses a different IP for each destination. Trackers cannot link across connections.
+\end{itemize}
 
-\section{Criminal Misuse and Data Retention}
+\subsection{Anonymity Networks in Practice}
+\begin{itemize}
+  \item \textbf{Tor}: Relays user traffic through multiple nodes to obscure origin and destination. Vulnerable to global adversaries and end-node monitoring.
+  \item \textbf{I2P}: Optimized for anonymous communication within the network. Use cases: Peer-to-peer file sharing, messaging, and hosting services.
+  \item \textbf{Mix Networks}: Provide stronger anonymity by shuffling and re-encrypting messages at each node. Used in email anonymization and anonymous transactions.
+\end{itemize}
+\subsection{Other noteworthy topics}
+\begin{itemize}
+  \item \textbf{Censorship Resistance}: Techniques to bypass network restrictions and surveillance.
+  \item \textbf{Misuse of Anonymity}: Use of privacy-enhancing technologies for illicit activities (e.g., illegal marketplaces, malware distribution).
+  \item \textbf{Data Retention Policies}: Mandated storage of communication metadata for investigative purposes. Conflict with privacy-focused systems and anonymization techniques.
+\end{itemize}
 
+\includepdf[pages={161-216}, nup=4x9]{slides.pdf}
 \part{Location Privacy}
+\section{Location Privacy \& Mobile Networks}
+
+\subsection{Overview}
+\begin{itemize}
+  \item \textbf{Objective}: Protect users' location information from being tracked or monitored by adversaries or unauthorized entities.
+  \item \textbf{Challenges}: Balancing privacy and utility in location-based services.
+\end{itemize}
+
+\subsection{Systematic Protection Methods}
+\begin{enumerate}
+  \item \textbf{Broadcasting}: Randomized transmission of location data.
+  \item \textbf{Group Pseudonyms}: Shared identifiers among a group of users to obscure individual location.
+  \item \textbf{Temporal Pseudonyms}: Changing pseudonyms periodically to reduce traceability.
+  \item \textbf{Mixing Networks}: Use of cascaded nodes to anonymize the source and destination of location data.
+\end{enumerate}
+
+\subsection{Broadcasting Techniques}
+\begin{itemize}
+  \item \textbf{Implicit Addressing}: Indirectly target devices by broadcasting generalized location updates.
+  \item \textbf{Variable Broadcasts}: Adjust broadcast size dynamically to obscure exact location.
+\end{itemize}
+
+\subsection{Pseudonyms}
+\begin{itemize}
+  \item \textbf{Group Pseudonyms}: Shared identifiers among a group of users within region to obscure individual location.
+  \item \textbf{Temporal Pseudonyms}: Pseudonyms change over time to prevent long-term tracking.
+\end{itemize}
+
+\subsection{Mobile Communication-Mixing}
+\begin{itemize}
+  \item \textbf{Centralized Approach}: Trusted third party manages pseudonym and location updates. High scalability but susceptible to single points of failure.
+  \item \textbf{Decentralized Approach}: Users and mix cascades collaborate to anonymize location updates. Improves privacy but increases coordination complexity.
+\end{itemize}
+
+\subsection{Anonymity Techniques in Mobile Networks}
+\begin{itemize}
+  \item \textbf{Mix Networks}: Shuffle and re-encrypt location data to prevent tracking.
+  \item \textbf{Blind Signatures}: Sign messages without revealing the signer's identity.
+\end{itemize}
+
+\subsection{Mobile Internet Protocol (IP)}
+\begin{itemize}
+  \item \textbf{Location Hiding}: Use of proxies or agents to hide the user's exact location.
+  \item \textbf{Challenges}: Higher latency due to intermediate nodes. Trade-off between performance and privacy.
+\end{itemize}
+
+\subsection{Performance Analysis of Privacy Methods}
+\begin{itemize}
+  \item \textbf{Trust Levels}: Varying levels of trust required in centralized vs. decentralized systems.
+  \item \textbf{Efficiency Loss}: Privacy-preserving methods incur bandwidth and processing overhead (1-10\%).
+  \item \textbf{Protection Metrics}: Degree of protection against tracking and profiling varies by method.
+\end{itemize}
 

### 2025-02-27 PbD/paper/template.tex
PbD (#5)
Added summary, exercises, and paper.

--- 2024-11-05 PbD/templates/latex/stylesvs.tex
@@
 % Kodierung, Sprache, Patches {{{
-\usepackage[T1]{fontenc}    % Ausgabekodierung; ermöglicht Akzente und Umlaute
-                            %  sowie korrekte Silbentrennung.
+\usepackage[T1]{fontenc}
 \usepackage[utf8]{inputenc} % Erlaubt die direkte Eingabe spezieller Zeichen;
                             %  utf8 muss die Eingabekodierung des Editors sein.
-\usepackage[ngerman]{babel} % Deutsche Sprachanpassungen (z.B. Überschriften).
+\usepackage{babel} % Deutsche Sprachanpassungen (z.B. Überschriften).
 \usepackage{microtype}      % Optimale Randausrichtung und Skalierung.
-\usepackage[
-    autostyle,
-    ]{csquotes}             % Korrekte Anführungszeichen in der Literaturliste.
-%\usepackage{fixltx2e}      % Patches fuer LaTeX2e - seit 2015 nicht mehr nötig
+\usepackage[autostyle]{csquotes}             % Korrekte Anführungszeichen in der Literaturliste.
 \usepackage{scrhack}        % Verhindert Warnungen mit älteren Paketen.
-\usepackage[
-  newcommands
-]{ragged2e}                 % Verbesserte \ragged...Befehle
-\PassOptionsToPackage{
-  hyphens
-}{url}                      % Sorgt für URL-Umbrüche in Fußzeilen u. Literatur
+\usepackage[newcommands]{ragged2e}                 % Verbesserte \ragged...Befehle
+\PassOptionsToPackage{hyphens}{url}% Sorgt für URL-Umbrüche in Fußzeilen u. Literatur
 % }}}
@@
 % Biblatex {{{
-\usepackage[
-    style=alphabetic,
-    backend=biber,
-    %backref=true
-    ]{biblatex}             % Biblatex mit alphabetischem Style und biber.
-\usepackage[realmainfile]{currfile} % Overleaf unterstuetzt \jobname nicht richtig.
-\bibliography{\currfilebase.bib} % Dateiname der bib-Datei (funktioniert auch mit Overleaf)
-%\bibliography{\jobname.bib} % Dateiname der bib-Datei (funktioniert nicht mit Overleaf)
-\DeclareFieldFormat*{title}{
-    \mkbibemph{#1}}         % Make titles italics
-% }}}
+% \usepackage[
+%     style=alphabetic,
+%     backend=biber,
+%     %backref=true
+%     ]{biblatex}             % Biblatex mit alphabetischem Style und biber.
+% \usepackage[realmainfile]{currfile} % Overleaf unterstuetzt \jobname nicht richtig.
+% \bibliography{\currfilebase.bib} % Dateiname der bib-Datei (funktioniert auch mit Overleaf)
+% %\bibliography{\jobname.bib} % Dateiname der bib-Datei (funktioniert nicht mit Overleaf)
+% \DeclareFieldFormat*{title}{
+%     \mkbibemph{#1}}         % Make titles italics
+% % }}}
 
